### 2017-04-09 02-synth.tex
Initial ShareLaTeX Import

\chapter{Evaluating the Effect of Homicide Prevention Strategies in São Paulo, Brazil: A Synthetic Control Approach}
\label{chap:synth}

\section{Introduction}

Brazil has long been ravaged by an undeclared civil war. According to the Citizen Council on Public Security and Criminal Justice, a Mexican think-tank, 19 of the 50 most violent cities in the world are located in Brazil \citep{mexico2014}.\footnote{The study disregards war zones and cities with unavailable data.} The 2014 Violence Map survey shows that 56,337 people were murdered in Brazil in 2012 alone, the highest incidence rates of intentional homicides on the planet \citep{mapa2014, unodc2013}. Paradoxically, the sharp rise in lethal violence has occurred during Brazil's longest period of political openness \citep{ahnen2003, pinheiro2000, pinheiro2001}. Murder rates have almost doubled over three decades of democracy, jumping from 15 homicides per 100,000 people in 1985 to roughly 29 per 100,000 in 2012 \citep{mapa2014}.\footnote{\citet{cerqueira2013} argues that the actual rates may be different from the official statistics. He states that many homicides from 1996 to 2010 were (intentionally or not) misclassified as `death by undetermined causes.' After performing data correction procedures, the author estimates that the number of homicides in Brazil during that period should be 18.3\% higher than the reported figures. Recent criticism about the quality of São Paulo homicide data can also be found at \url{http://goo.gl/x0pHac} (in Portuguese). Access: January, 2016. In this article, I avoid these issues by using obituary data instead of police records.}

São Paulo has traditionally occupied a key position in Brazil's violence statistics. It is the country's richest and most-densely populated state, and in the 1990s its homicide rate was roughly 50\% higher than the national average \citep[p. 120]{barata2000}. Some areas of the namesake capital city had even worse numbers. Between 1996 and 1999, the ramshackle districts of Jardim São Luiz and Jardim Ângela had respectively 103 and 116 violent deaths per 100,000 residents \citep[p. 8]{cardia2003}, figures that placed them amongst the deadliest neighbourhoods on the globe \citep{who2015}.

Nevertheless, the state of São Paulo has experienced a drastic reduction in homicides during the last years \citep{camargo2007}. The decline is so remarkable that some authors have called it `the great homicide drop' \citep{goertzel2009}. The city of São Paulo, which is currently home to about 11 million inhabitants, provides a telling example. Over a span of only seven years (2000--2007), the number of annual violent deaths in the capital fell from 5,979 to 1,311, a 78\% decrease.\footnote{The homicide statistics cited in this paragraph come from the Centre for the Study of Violence, a research group of the University of São Paulo. Their dataset can be found at the following electronic address: \url{http://nevusp.org/downloads/bancodedados/homicidios/distritossp/num-homicidios-distritos-2000-2007.htm}. Access: March, 2016.} Significantly, São Paulo city became the safest state capital in Brazil \citep{mapa2011}.

São Paulo's success should be attributed to local factors. From 1999 onwards, the state government created or expanded a number of policies that have arguably contributed to the decrease in criminality. In a move coherent with the basic tenets of the economics of crime \citep[e.g.][]{becker1968crime, cornish2014reasoning}, the administration increased the certainty and the intensity of punishment to discourage potential offenders. Amongst other measures, the government implemented strict gun control policies \citep{goertzel2009}, raised incarceration rates \citep{salla2007}, and imposed harsher sentences on those convicted of a crime \citep{carvalho2005}.

But whereas several authors acknowledge the effectiveness of these policies, few quantitative studies have gone beyond statistical correlations to justify their arguments. In the case of São Paulo, a major difficulty is separating the state's particular time trend to that of Brazil. Ideally, one should compare São Paulo to a control case that shares the same characteristics of the existing state, except that it has not been subjected to the specific set of policies implemented by the São Paulo state government. This thought exercise, which emulates the logic of a controlled experiment \citep{angrist2008mostly, imbens2015causal, holland1986, morgan2014counterfactuals}, would allow practitioners to untangle the effects of homicide reduction programmes from other potential confounders.

In this paper, I employ the synthetic control method (henceforth SCM) to approximate this experimental ideal and measure the total causal effect of post-1999 public policies on São Paulo homicide rates. The method consists of creating an artificial counterfactual to estimate the impact of a given intervention on a unit of interest. SCM has gained widespread acceptance in many fields, having been successfully applied in political science \citep{abadie2014, montalvo2011}, economics \citep{billmeier2013, coffman2012, jinjarak2013}, education studies \citep{hinrichs2012}, and public health science \citep{heim2014}. However, SCM has rarely, if ever, been used to evaluate homicide prevention strategies in São Paulo, despite being a useful tool for this particular type of question. SCM was specifically designed for situations where there is only one treated unit of interest, no readily-available counterfactual, and no certainty as to whether the treated and the control units follow parallel trends after the intervention \citep{abadie2003, abadie2010, abadie2014}. Moreover, SCM also has some of the desirable properties of other popular causal inference tools such as differences-in-differences \citep{angrist2008mostly, bertrand2004much} and matching estimators \citep{dehejia2002propensity, ho2007matching, rubin1973matching, stuart2010matching}.

I find that from 1999 to 2009, about 20,000 lives were saved in São Paulo. When compared to a synthetic counterfactual, São Paulo's actual homicide rates were less than 50\% of what would be expected in the absence of policy implementation (15 versus 32 homicides per 100,000 people). Additional tests confirm the robustness of the results and indicate a 96.3\% chance of a causal effect in the intervention period.

The article is structured as follows. Section~\ref{sec:theoretical_background} discusses how deterrence provides a useful framework to understand the reduction in homicide rates in the state. I also examine an alternative hypothesis for the drop in crime in São Paulo -- the rise of the \emph{Primeiro Comando da Capital} -- and argue that the prison gang should be regarded as a moderator, but probably not as an independent cause of homicide reduction. Section \ref{sec:methods} presents a justification for, and a technical explanation of, the synthetic control method. Section \ref{sec:data} describes the data used in this paper. Section \ref{sec:analysis} discusses the results of the models and presents several robustness tests. Section \ref{sec:conclusion} offers some concluding remarks.

\section{Theoretical Background}
\label{sec:theoretical_background}

\subsection{Deterrence, Information and the Drop in Homicides}
\label{sub:deterrence_and_the_drop_in_homicides}

A myriad of explanations have been proposed for the fall in homicide rates in São Paulo. Some authors have stressed the importance of long-term factors on local levels of violence. \citet{mello2010} claim that the shrinking of the proportion of males in the 15--25 age bracket has led to fewer violent deaths at both state and city levels. \citet{hughes2004} argues that São Paulo's spatial segregation patterns have had a lasting impact on murder rates. \citet{barata2000}, in turn, posits that macroeconomic conditions, mainly inequality indicators, are positively correlated with violent crime in São Paulo.

Structural variables have likely been important in reducing violence, but the role of public policies should not be underestimated. The Brazilian Social Democracy Party (\emph{Partido da Social Democracia Brasileira}, PSDB), which has ruled São Paulo since 1995, has repeatedly asserted its commitment to reducing urban crime throughout the state \citep{bueno2014}. In 1998, former governor Mário Covas -- then running for re-election -- set the ambitious goal of ``slashing criminality rates in half'' during his second term in office \citep{santos2008}. This commitment was then followed by his vice-governor and successor, Geraldo Alckmin, who has expanded those measures and taken a notoriously tough stance on crime \citep{de2012governo}.

Methods of crime prevention have received considerable attention from the authorities. Firstly, the São Paulo government significantly increased incarceration rates in the past decade \citep{salla2007}. The state currently holds around 200,000 convicts in prison (35\% of Brazil's inmate population) and adds another 15,000 inmates to the official statistics every year \citep{brasildefato2013}. Furthermore, prisoners have also become subject to harsher legal punishments. The São Paulo administration has also been making large use of the \emph{Regime Disciplinar Diferenciado} (Special Disciplinary Regime), which provides for up to 360 days of solitary confinement for disobeying the law \citep{carvalho2005}.

Secondly, the state government has successfully enforced a ban on gun possession in São Paulo. Studies show that this policy has been effective in reducing homicides resulting from both drug-related crimes and domestic disputes \citep{goertzel2009, kahn2005papel}. Furthermore, the impact of the Brazil's 2003 National Disarmament Act was particularly pronounced in São Paulo. \citet{cerqueira2013evaluating} argue that between 2005 and 2007 the enforcement of the anti-firearm legislation was responsible for saving between 2,000 to 2,750 lives in cities with more than half a million inhabitants in the state of São Paulo.

This set of policies is largely in line with the rational choice theory of crime \citep[e.g.][]{becker1968crime, ehrlich1973participation,levitt1996effect, levitt1997using, paternoster2010much}. The rational choice school posits that criminals are motivated by utilitarian cost-benefit analysis. Individuals calculate what the possible trade-offs are between the benefit of the committing a crime and the risk of being punished for it. Criminal offenders, therefore, are in no way different from non-criminals: the only difference between them is their \emph{choices} \citep{nagin2007moving}. To reduce criminality, policy-makers have to ensure that the costs of committing a crime outweigh the eventual utility an individual derives from it.

Deterrence measures have been complemented by investments in police intelligence. In 1999, the state administration created a new system for crime prevention, Infocrim \citep[p. 3]{risso2014intentional} The system gathers geo-coded information on homicides and maps the most important `hot spots' of criminal activity in the state. The government has also developed a new photo database, Fotocrim, to speed up the process of facial recognition of criminals \citep[p. 3]{mello2010}.

More information improves the effectiveness of police strategies via two mechanisms. On the one hand, police forces can be quickly moved to where they are most needed. This reinforces the role of deterrence as it increases the likelihood of punishment for criminals. On the other hand, the system also makes clear what regions are making progress in reducing crime. This allows police chiefs to monitor local personnel and take measures to improve performance if required.\footnote{See: \url{http://goo.gl/kqLhYb} (in Portuguese). Access: August 2016.}

Recent evidence shows that the intelligence system has effectively lowered the crime statistics in São Paulo. Using a spatial differences-in-differences estimator, \citet{cabral2016infocrim} argues that Infocrim has had a large negative impact on homicide rates in the municipalities where it was implemented. The author also notes that the effect remains important even after accounting for possible displacement effects. As expected, some criminals did take their activities elsewhere after the creation of Infocrim, but this movement has not offset the benefits of the system.

How well have these policies performed over time? The results suggest a favourable outlook. Compared to other Brazilian states, São Paulo is an outlier when it comes to homicide rates. Despite the fact that crimes against property have declined little over the last decades,\footnote{Recent data on property crimes in São Paulo can be seen at \url{http://www.ssp.sp.gov.br/novaestatistica/Pesquisa.aspx} (in Portuguese). Access: July 2016.} the number of violent deaths per 100,000 inhabitants shows a steep downward trend. Figure \ref{fig:figure1} presents the evolution of homicide rates in São Paulo in comparison with the Brazilian average.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/br.eps}
    \caption{Homicide Rates per 100,000 Population -- São Paulo and Brazil (Excluding São Paulo)}
    \label{fig:figure1}
\end{figure}

The trends are even more striking if we consider that deterrence policies are still controversial in the literature. \citet{barbarino2014incapacitation}, \citet{buonanno2013incarceration}, \citet{levitt1996effect,levitt2004understanding} and \citet{owens2009more} claim that incapacitation measures effectively reduce crime, but \citet{eck2006have} and \citet{beattie2007police} suggest that increases in police forces and incarceration rates in the United States and in Canada did not lead to expected outcomes.

There is good evidence that incapacitation measures have worked well in São Paulo during the last decade. Gun-related homicides have declined about 74\% from 2001 to 2008 \citep{peres2011queda} at the same time when São Paulo has experienced an increase of 770\% in the arrests of repeated murderers \citep[p. 36]{manso2012homicidos}. Although there are studies that indicate possible `hardening effects' of imprisonment, that is, longer sentences may positively affect an individual's tendency to commit further crimes \citep[e.g.][]{chen2007harsher,glaeser1996crime,western2001labor}, the São Paulo case appears to suggest otherwise. Moreover, violent deaths have decreased in all population strata, but especially amongst males (-74.5\%), 15 to 24 year-old men (-78,0\%) and those who live in extreme poverty (-79,3\%), groups that are generally associated with criminality \citep{peres2011queda}.

Nevertheless, it is difficult to know which of the policies have contributed more to this large homicide reduction. Not only we do not have disaggregated data to test preliminary hypotheses, but there may be large interaction effects amongst different public security measures. Therefore, at the moment it is not possible to disentangle micro-level causes from macro effects. But the aggregated impact of the anti-crime policies can be correctly identified if there is no other variable in the causal path leading from the policies mentioned above to our dependent variable (state homicide rates). I argue below that this type of estimation is feasible for the São Paulo case. To back this claim, I suggest that a competing explanation for the homicide drop in São Paulo -- the rise of the PCC -- interferes only with the direct effect of the policies on crime, but not with their \emph{total} effect. In this sense, the synthetic control method provides a plausible identification strategy for my question of interest.

\subsection{Alternative Explanation: The Emergence of the PCC}
\label{sub:alternative_explanation_the_emergence_of_the_pcc}

A recent hypothesis attributes the decrease in violent deaths in São Paulo to the \emph{Primeiro Comando da Capital} (First Command of the Capital, henceforth PCC) \citep{biondi2010, dias2009, dias2011pulverizaccao, de2010crime, de2012governo, willis2015killing}. The PCC is a prison gang that emerged in the early 1990s as a response to the demands of a growing prison population. The PCC provides personal security and financial assistance to their members and affiliates. The gang's internal statute clearly declares that ``[\dots] those who are in liberty [must contribute] to the brothers inside prisons [PCC members] through lawyers, money, help to family members and prison outbreak operations'' \citep{folha2001estatutopcc}.

A group of scholars argue that the PCC significantly contributed to the reduction in violence mainly through the São Paulo prison system. At least since the mid-2000s, these authors argue that the PCC has been able to emerge as an undisputed mediator and solve conflicts between inmates. \citet[p. 83]{dias2009ocupando} writes that ``[\dots] when unable to constitute a universal source of regulation, the official law leaves gaps which are filled by informal instances -- such as the Primeiro Comando da Capital (PCC), in the prisons of São Paulo.'' The gang has implemented informal courts that resemble state institutions, and those meetings have progressively replaced other forms of popular justice such as lynchings or the hiring of target killers \citep[p. 3]{feltran2012metodos}. Moreover, the \emph{Comando} has developed a series of assertive ways to terrorise inmates. Since the PCC's threats are credible, the group is able to impose discipline within the São Paulo prison system \citep{biondi2010,dias2009}.

Paradoxically, the PCC might have also helped to reduce crime in São Paulo by collaborating with street-level police. The Brazilian state does not hold a perfect monopoly of force in many areas of the country \citep{arias2009drugs,de2012governo,hughes2004,pinheiro2000}, thus access to local knowledge may prove vital for the success of a given operation. In this regard, the PCC and the state may collude if the situation is beneficial to both, and as such there is an informal--but potentially unstable--``killing consensus'' in the state \citep{willis2014antagonistic, willis2015killing}.

There has been a vigorous debate over whether the PCC has had a significant impact on violence rates. A few authors see the PCC as the sufficient condition behind the homicide rates decline across São Paulo state \citep{biondi2010, dias2009, dias2011pulverizaccao}, whilst others take a more nuanced view of the role of the prison gang \citep[e.g.][]{willis2015killing}. But both groups of scholars affirm that, based on their first-hand experience, the PCC is the key explanatory variable behind the drop in murders in São Paulo.

Recent econometric works, however, do not seem to confirm that argument. Marcelo Nery has found no convincing results in favour of the `PCC hypothesis' using geo-referenced data for São Paulo \citep{bbc2016pcc}. \citet{biderman2015pax} use anonymous calls to a crime hotline as a proxy for PCC presence in São Paulo city favelas. The authors suggest there is some support for the idea that the criminal syndicate reduces lethal violence in areas under its control, but PCC presence corresponds to only a minor drop in violent crime. Although the PCC impact is not negligible, the gang is not a sufficient condition for the homicide decline.

Another counter-argument to the PCC thesis is that homicides also decreased in areas and groups over which the PCC does not exert control. Firstly, descriptive statistics show that the decline in violent deaths started \emph{before} the PCC's expansion period.\footnote{As shown in figure \ref{fig:figure1}, São Paulo's homicide rates started to drop in 1999. The PCC consolidated their power in the prison system only in the mid-2000s \citep{dias2011pulverizaccao}.} Secondly, the drop in crime was evenly distributed throughout the state: urban and rural areas, small and large cities alike experienced fewer murders.\footnote{See: \url{http://www.fenapef.org.br/27764/} (in Portuguese). Access: July 2016.} Finally, as noted above, \citet{peres2011queda} point out that violent death rates decreased in \emph{all} age groups and social classes in the city São Paulo. Hence, cohorts that do not correspond to typical PCC members (such as the elderly or middle-age females) are also less affected by violence.

It seems that the influence of the PCC on physical violence has been overstated. It is unlikely that the PCC -- which is underfunded for its size\footnote{A Parliamentary Commission of Inquiry has stated that the PCC earns about 16 million Brazilian Reals per month, which amounts to approximately 60 million US dollars per year (see: \url{http://goo.gl/FwhPa3} (in Portuguese). Access: July 2016. Given the size of the organisation and its undisputed position as the leading crime syndicate in São Paulo, the figures are rather small. As a comparison, Mexico's Sinaloa Cartel profits about 3 billion dollars per year, a sum comparable to the annual earnings of Netflix or Facebook. See: \url{http://nyti.ms/1B09qyV}. Access: July 2016.} -- could have achieved such deep penetration into society and lowered the violence levels across all population groups in the whole state.

Yet, the group's importance cannot be fully dismissed either. Data on PCC-controlled areas are likely to contain measurement errors that may bias the coefficients, thus caution is required before making strong causal claims on this discussion. Despite mounting observational evidence that the PCC may not provide a complete explanation to São Paulo's lower crime rates, the argument could only be comprehensively tested in a counterfactual case in which the PCC is present and the state policies are not.\footnote{I would like to thank an anonymous reviewer for highlighting this point.} Currently-available data do not allow us to evaluate such scenario.

\subsection{Causal Paths, Moderators, and Total Effects}
\label{sub:causal_paths_moderators_and_total_effects}

A methodological issue remains. If we are to estimate the causal effect of the public measures on the crime rates, how should we proceed? I have noted above that the specific impact of micro-level policies cannot be evaluated due to lack of data. Nonetheless, it is theoretically possible to estimate the \emph{total} effect of policies on crime.

The difference between direct and total effects can be understood as follows. The direct effect captures the sensitivity of a dependent variable $Y$ to changes in $X$ when this relationship is not mediated by any other variables in the model. Holding all factors constant, the direct effect is a causal chain of length one \citep[p. 160]{sobel1987direct} and could be described simply as $X \rightarrow Y$. In turn, the total effect can be defined as $P(Y_{x} = y)$, that is, ``the probability that response variable $Y$ would take on the value $y$ when $X$ is set to $x$ by external intervention'' \citep[p. 1572]{pearl2001direct}. The total effect is the sum of direct and indirect (or mediated) effects.

In our case, gun control, incarceration, and police intelligence have likely had a direct effect on homicides. Combined, these variables comprise a direct aggregate policy effect. The omission of a variable measuring the impact of the PCC could bias such an effect, but not interfere with the \emph{total policy effect}. This point is worthy of further consideration. The total policy effect would be unbiased under the assumption that the PCC is in fact a \emph{moderator} between the public policies and the homicide rates, even if the gang's impact over the violence levels is not particularly large.

Although this argument has rarely been posited in such terms, this position is largely supported by the qualitative literature on the PCC. Fieldwork research generally traces the group's origins and growth to the rising incarceration rates in São Paulo and the need for protection amongst prisoners \citep{dias2011pulverizaccao, manso2014}. Like other prison groups, the PCC would only mobilise resources to provide welfare and act as an arbitrator under the condition that the certainty of punishment by the state is high \citep{skarbek2011, freire2014}. Had the state not increased the costs associated with crime, the prison gang would not have expanded their reach, or even been created in the first place. Hence, the impact of the PCC on street-level violent deaths -- if it exists -- can be safely assumed to be a moderator effect.

Whereas it would be interesting for researchers to separate these types of effects and isolate the PCC from the other causal outcomes, such estimation is not possible at the state level. However, as these measures were implemented throughout São Paulo state at roughly the same time, their combined effect is computable even though their individual direct effects are not. To do so, it is only necessary to contrast the treated unit (São Paulo) with a counterfactual without the time-assigned treatment (1999 onwards) and evaluate the aggregated effect of the public policies.

This analysis can be estimated in a consistent manner with the synthetic control method. In the following sections I describe how the method creates a valid counterfactual case under a certain set of assumptions. The assumptions are: 1) the PCC is an outcome, not a cause, of the crime-targeting policies; 2) the model does not include unnecessary control variables; 3) interpolation bias is not very severe because the cases in the `donor pool' are relatively similar to the treated unit.

\section{Methods}
\label{sec:methods}

The synthetic control approach provides an adequate solution for two enduring problems in the social sciences: the arbitrary selection of comparative cases and the poor estimation of causal effects when few pre-treatment observations are available \citep{abadie2003, abadie2010}. With respect to the first issue, scholars often resort to ambiguous criteria in their choice of control units. This practice ends up casting doubts over the validity of their selected counterfactual \citep{abadie2011}. The synthetic method provides a reliable comparative case by adopting a purely data-driven process in order to select a counterfactual. Also, the researcher can still specify what control cases enter the `donor pool.' In this sense, qualitative expert knowledge can be incorporated in the estimation via the selection of cases.

Regarding the second issue, the accurate estimation of coefficients from a small number of cases, SCM employs a consistent statistical solution to problems of incorrect data extrapolation and model dependence. SCM can be understood as a combination of matching with differences-in-differences. SCM uses matching as a flexible pre-processing tool to reduce imbalance between treated and control units \citep{ho2007matching, rubin1973matching, rubin2006matched}. But unlike matching, SCM deals with only one treated unit over time. Therefore, the method can also be interpreted as a semi-parametric extension to differences-in-differences estimators in which both treated and control units are not required to follow parallel trends in the whole period. \citep{abadie2005semiparametric}. By combining semi-parametric matching with differences-in-differences, SCM provides a rigorous yet versatile method to evaluate time-dependent treatment effects.

The method works as follows.\footnote{Please refer to Appendix~\ref{sec:appendix_a} for a formal presentation of the synthetic control method.} SCM starts with the assumption that one case in the sample has received a treatment. The treatment is defined as a time-delimited event that affects the unit of interest, such as the implementation of a new policy or the outbreak of a conflict. SCM also requires a series of control cases to estimate the models, that is, units that did not receive the treatment during the same period. These cases are often related to the treatment case in some meaningful way, and natural choices for the donor pool are provinces within the same country, or states that share important characteristics. These traits can also be more specifically defined and included as quantitative variables in the estimation models.

SCM then selects a few cases from the donor pool to create a new, artificial control for the treated unit of interest. The main goal of SCM is to construct a counterfactual that resembles the treatment unit more closely than any individual control in the donor pool. Cases are combined in way similar to a weighted average, in which controls that are more similar to the treated unit receive more weights. The weights make explicit the contribution of each separate case to the synthetic control, what also increases the transparency and reliability of the method \citep{abadie2014}. The closer the synthetic control matches the original treated unit before the assignment of the period, the better the quality of the counterfactual.

The method uses an algorithm to minimise the difference between the control cases and the treated unit before the intervention. The authors adopt the mean squared prediction error (MSPE) as a measure of fit \citep{abadie2003}. MSPE is simply the difference between the fitted and the observed trends of the treatment case. A small value means that the two lines are highly correlated and the artificial control is a good approximation of the missing counterfactual in the post-intervention period. In our case, the counterfactual would be São Paulo from 1999 to 2010 without the crime-reducing policies.

SCM has an intuitive interpretation. Although numeric summaries and other statistics can be obtained from the model, a simple time series graph is usually enough to assess the results. The causal effect is the difference between the treated and the synthetic cohort. The larger the post-treatment gap, the stronger the treatment impact.

As with all types of observational studies, SCM can also suffer from omitted variable bias. One can never be sure whether all required confounders have been included in a given model. However, the graphical output of the SCM helps diagnose the presence of large disparities between treatment and control cases. If the trends follow similar paths during the control period, it provides some indication -- albeit only informally -- that omitted variable biases are not driving the output. This bias can also be mitigated with expert knowledge. Econometric studies show that the inclusion of a large number of covariates and post-treatment variables to correct for omitted variables bias can actually \emph{worsen} the problem \citep{achen1992social, achen2002toward, clarke2005phantom, clarke2009return, pearl2009causality}. This is particularly true for matching methods. Authors have noted that `over-matching' can lead to severe statistical bias \citep{baser2006too, brookhart2006variable, marsh2002removal}. In this regard, the most plausible solution seems to be attention to the trends and sensible selection of control variables. As I discuss below, the covariates included in this paper are some of the most robust quantitative predictors of homicides.

Furthermore, placebo tests can be run to test the robustness of the findings. For instance, researchers can include `in-time placebos,' dates under which the treatment \emph{did not} occur. Results should change only in the period when the treatment starts and not at any other point in time. Moreover, scholars can also add `in-space placebos' to their models. This test consists of adding different members of the donor pools into the models to see if the estimation varies \citep{abadie2014}. Finally, one can also compare the effects of the treatment of interest by creating a distribution of synthetic cohorts, where every unit (treated or not) is matched with a specific synthetic control case. The parameter of interest should still be relevant. I employ all of these tests in this article and the results can be seen in the following sections.

\section{Data}
\label{sec:data}

I build panel data for the variables \emph{Homicide Rate}, \emph{State GDP per Capita}, \emph{State GDP Growth}, \emph{Years of Schooling}, \emph{Gini Index}, \emph{Natural Logarithm of Population} and \emph{Population Living in Extreme Poverty}. These variables are very common in the specialised literature\footnote{For overviews of cross-national studies of homicide, see \citet{lafree1999summary}, \citet{nivette2011cross} and \citet{trent2012review}.} and represent important social and economic factors I wish to control for.

The unit of analysis is State-Year. I have data from all of the 26 states plus the capital city (Distrito Federal), ranging from 1990 to 2009. The data for years prior to 1990 are scarce and for years after 2009 have not yet been published. All data used in this paper come from the same source, the \emph{Instituto de Pesquisa Econômica e Aplicada} (IPEA), a government-led research group.\footnote{The data are publicly available at \url{http://www.ipeadata.gov.br/}. The original data files have also been added to \url{https://github.com/danilofreire/homicides-sp-synth} for reproducibility purposes.}

My dependent variable measures the number of homicides per 100,000 inhabitants, which is the most commonly used unit of analysis for lethal violence. This variable was coded by the Brazilian Health Ministry from obituary records, therefore it is less likely than police files to suffer from intentional misrepresentation.

There are six control variables in the models. \emph{State GDP per Capita} is adjusted in 2010 Brazilian Reals (at the time 1 Brazilian Real bought roughly 0.5 U.S. dollars). \emph{State GDP Growth} is measured in constant 2010 Brazilian Reals and varies by percentage points. \emph{Years of Schooling} describes the average number of years of formal instruction at educational facilities (males and females, 25 years old or more.) \emph{Gini Index} is a measure of inequality, ranging from 0 to 1 where 0 is the most equal and 1 the most unequal. \emph{Natural Logarithm of Population} represents yearly projections of the state population. Since Brazil only runs a census every 10 years, these projections represent the most accurate data available. I have taken the natural logarithm of this variable to account for size effects. Finally, \emph{Population Living in Extreme Poverty} describes the percentage of the state population which do not meet the minimum intake of 2,000 calories per day. This is the only variable that I created specifically for this study. It was coded by simply taking the number of individuals classified as extremely poor by the IPEA and dividing this number by the state's total population.\footnote{\emph{Years of Schooling} and \emph{Gini Index} had a small number of missing observations (about 15 percent) and those cases were imputed with linear interpolation. Both original and imputed variables are available online. See the supplementary appendix for further details on how to replicate this study.}

\section{Analysis}
\label{sec:analysis}

\subsection{Main Model}

I construct the synthetic cohort (\emph{Synthetic São Paulo}) by imputing information from all of the Brazilian states plus the Federal District. The synthetic control method outputs a set of weights for states and variables such that the treatment state is approximated optimally by these weighted components. This method not only provides a quantitative way of selecting comparison cases but also gives us a much better baseline to compare with the treatment unit. Synthetic São Paulo is constructed using six states, \emph{i.e.}, the six out of the 27 possible cases that received non-zero weights. Table 1 shows that the states that best synthesize São Paulo are, respectively, Santa Catarina (0.274), Distrito Federal (Brasília) (0.210), Espírito Santo (0.209), Rio de Janeiro (0.169), Roraima (0.137) and Pernambuco, which only accounts for 0.01 of the weights. In this regard the state selection does not appear as a complete surprise. Apart from Roraima, the other members of the federation are richer, more densely populated and better schooled than the country average, thus being indeed similar to São Paulo.

\begin{table*}[htp!]
\caption{Synthetic Weights for São Paulo}
\begin{tabular*}{\hsize}
{@{\extracolsep{\fill}}lccc}
\hline
\multicolumn1c{\emph{State}}&\multicolumn1c{\emph{Synthetic Control Weights}}&
\multicolumn1c{\emph{Predictor}}&\multicolumn1c{\emph{Weights}}
\cr
\hline
\emph{Santa Catarina}&0.274
&\emph{Years of Schooling}&0.469
\cr
\emph{Distrito Federal}&0.210
&\emph{State GDP per Capita}&0.275
\cr
\emph{Esp\'{i}rito Santo}&0.209
&\emph{Homicide Rate}&0.241
 \cr
\emph{Rio de Janeiro}&0.169
&\emph{Population Living in Extreme Poverty}&0.009
 \cr
\emph{Roraima}&0.137
&\emph{Gini Index}&0.005
 \cr
 \emph{Pernambuco}&0.001
&\emph{Ln Population}&0.001
 \cr
\hline
\end{tabular*}
\end{table*}
\label{table1}

Among the independent variables, only three out of six receive substantial weights. Given the data I could obtain, the predictors that receive more weight are Years of Schooling (0.469), State GDP per Capita (0.275) and past Homicide Rate (0.241). The three remaining variables are much less relevant to the model. They are, respectively, the Population Living in Extreme Poverty (0.009), Gini Index (0.005) and Natural Logarithm of the Population (0.001). Table 2 compares characteristics of São Paulo and its synthetic control prior to policy implementation. We see that Synthetic São Paulo has very similar coefficients to those of the treatment unit. Moreover, the synthetic control clearly outperforms the sample means in all of the three relevant predictors. The worst measure is State GDP Growth, whose mean is about 2.6 whereas the figure for São Paulo is roughly 1.3 during that period. However, this outcome does not affect the results since the variables that received zero weight were discarded from the models.

\begin{table*}[htp!]
\caption{Homicide Rate Predictor Means Before Policy Implementation}
\begin{tabular*}{\hsize}
{@{\extracolsep{\fill}}lccc}
\hline
\multicolumn1c{\emph{Predictor}}&\multicolumn1c{\emph{São Paulo}}&
\multicolumn1c{\emph{Synthetic São Paulo}}&
\multicolumn1c{\emph{Sample Mean}}
\cr
\hline
\emph{Years of Schooling} & 6.089 & 6.110 & 4.963
\cr
\emph{State GDP Per Capita} & 23.285 & 23.079 & 11.830
\cr
\emph{Homicide Rate} & 32.672 & 32.479 & 21.843
 \cr
\emph{Population Living in Extreme Poverty} & 0.054 & 0.082 & 0.185
 \cr
\emph{Gini Index} & 0.536 & 0.561 & 0.578
 \cr
\emph{Ln Population} & 17.335 & 14.838 & 14.867
\cr
\emph{State GDP Growth} & 1.330 & 2.585 & 3.528
\cr
\hline
\end{tabular*}
\end{table*}
\label{table2}

The results show that the synthetic control method has successfully created a valid counterfactual to our case of interest. Figure \ref{fig:figure2} depicts the evolution of the dependent variable for the treatment and synthetic control cases. We can see that São Paulo and synthetic São Paulo have very close homicide rates series for the period ranging from 1990 until 1998. From 1999 onwards we observe the trajectories departing sharply from each other. The increase in homicide rates shown in the graph is consistent with previous statistical evidence. It indeed confirms that São Paulo had higher than expected levels of lethal violence, which I noted in the first part of this text.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/trends.eps}
    \caption{Trends in Homicide Rates: São Paulo versus Synthetic São Paulo}
    \label{fig:figure2}
\end{figure}

Despite the high levels of violence in 1999 -- when the new crime-reducing programme was implemented -- the number of homicides consistently declined until 2009. The trend is indeed monotonic and there is not a single peak in homicide rates after the policies have been put into practice. I interpret that as strong evidence in favour of the public policies.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/gaps.eps}
    \caption{Homicide Rates Gap between São Paulo and Synthetic São Paulo}
    \label{fig:figure3}
\end{figure}

With respect to the size of the effect, in 1998 the homicide rate in São Paulo was around 40 deaths per 100,000 inhabitants. In 2009 -- the last year for which data are available -- the rate dropped to 15, whereas synthetic São Paulo observed above 30 deaths per 100,000. That means a gap of $-20$ deaths for every 100,000 people in São Paulo in 2009, as can be seen in Figure \ref{fig:figure3}. I estimate that the new policies implemented in São Paulo saved roughly 20,300 lives in the period from 1999 to 2009.\footnote{My estimate of lives saved by the policies implemented in São Paulo is done as follows. I consider the years after policy implementation (1999--2009), then I sum the number of homicides in São Paulo in that period. This gives us 124,077 homicides between 1999 and 2009. I do the same procedure for the synthetic São Paulo; I sum the number of homicides in each state that makes the synthetic control in the period, while adjusting the contribution of each of these states by their respective weights in the synthesis. The number of homicides in synthetic São Paulo between 1999 and 2009 is 144,408. Finally, I subtract the number of homicides in the control by the number of homicides in the treatment. The result is 20,331 lives saved.} It is important to mention that the homicide rate in São Paulo continues to drop by the year, while the same is not happening in the rest of the country.

\subsection{Robustness Checks}

To further analyse the findings, I run five robustness tests. I first create an `in-time placebo' synthetic control. This test consists of creating a false starting date for the intervention period to check if one could observe false treatment effects in the pre-treatment years \citep{abadie2014}. If that were to be the case, the validity of the main results could be put into question. The result of this placebo test can be seen in Figure \ref{fig:figure4}. When I run the model with 1994 as the year when there was a supposed policy change, the result shows that there is only a minor gap between both lines. In other words, the method does not indicate a definite departure of trends between treatment and control cases.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/placebo.eps}
    \caption{Placebo Policy Implementation in 1994: São Paulo versus Synthetic São Paulo}
    \label{fig:figure4}
\end{figure}

I also conducted a leave-one-out robustness test. In this test I drop the states composing the synthetic control one at a time. The main goal of this analysis is to evaluate whether a single control state is driving the results. This would suggest that the original synthetic control -- which is composed of five states at a time -- is probably not a reasonable counterfactual. The results of this analysis can be found in Figure \ref{fig:figure5}. We see that the synthetic control (dashed line) is a reasonable amalgam of cases. Also, because the relative positions of treatment and controls are stable across controls, we observe that no control state is biasing the estimates.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/leave-one-out.eps}
    \caption{Leave-One-Out Distribution of the synthetic Control for São Paulo}
    \label{fig:figure5}
\end{figure}

Figure \ref{fig:figure6} shows the difference in homicide rates between the treated units and their synthetic controls. Here I estimate a synthetic control case for São Paulo and for each of the other 26 Brazilian states. This test assesses whether there is any previously unobserved national or regional trend that explains the original results. We observe that in São Paulo the homicide rate gap increases consistently during the treatment period, whereas the lines for the other states are moving randomly. Several lines fail to show any substantial difference between the state line and that of its synthetic counterfactual case. This indicates the results for São Paulo are unlikely to be a result of broader trends.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/permutation-gaps2.eps}
    \caption{Permutation Test: Homicide Rate Gaps in São Paulo and 26 Control States}
    \label{fig:figure6}
\end{figure}

Figure \ref{fig:figure7} presents the same test displayed in figure \ref{fig:figure6}, but it uses a stricter threshold for the simulated synthetic controls. The graph features cases in which the mean squared prediction error, a measure of goodness-of-fit, is no higher than twice that of São Paulo. That is, only placebos that have good synthetic matches were selected for the analysis \citep[p. 503]{abadie2010}.  In this group, the negative gap for the homicide rate São Paulo is by far the most relevant, providing further evidence for the original findings.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/low-mspe.eps}
    \caption{Permutation Test: Homicide Rate Gaps in São Paulo and Selected Control States}
    \label{fig:figure7}
\end{figure}

Lastly, I estimate another synthetic control using another approach. Here, I employ a Bayesian structural time-series model to verify the stability of the previous results \citep{brodersen2015inferring}. This inference procedure is similar to that described in section \ref{sec:methods} and it also consists of matching pre-treatment values of the unit of interest, São Paulo, to other potential control states. However, in this model only the time trends of the dependent variable are matched. In a sense, this is closer to a traditional differences-in-differences approach, but without the restrictive assumption that the treated and the control cases would follow parallel trends over time \citep[p. 494]{abadie2010}.

\begin{figure}[H]
    \centering
    \includegraphics[height=7.5cm]{images/causal-impact.eps}
    \caption{Bayesian Structural Time Series Model: São Paulo and Synthetic São Paulo}
    \label{fig:figure8}
\end{figure}

The model shows that in 2009 we should have expected São Paulo to have a homicide rate equal to 32.3 deaths per 100,000, but we observe only 15.2. Thus, the actual rate in São Paulo corresponds to only 47\% of the expected counterfactual. The method also generates an estimate for the probability of causal effect. The calculations indicate a 96.3\% chance of a causal impact in the period. In this sense, it is unlikely that the results are a statistical fluke.

\section{Conclusion}
\label{sec:conclusion}

As I have hopefully demonstrated, when compared to a synthetic control case, homicide rates were drastically reduced in São Paulo. Although it is not possible to estimate the treatment effect of each specific policy implemented during the 1990s and 2000s, I suggest that their aggregate impact is surely not negligible. If the estimation strategy employed in this paper is correct, the state of São Paulo offers an example that it is feasible to fight crime with targeted policies. This as an encouraging result, as it suggests that governments can make progress in reducing crime with the resources they already have at hand and need not rely exclusively upon structural conditions that are largely beyond their control, such as unemployment, per capita income and inequality. Robustness tests provide further evidence for my findings.

I also argue in favour of the synthetic control method as a tool to evaluate government policies. This approach offers an intuitive way to assess causality claims when there is only a single treated unit and it can be easily applied in a great number of situations. Assuming that there is a reasonable number of potential cases in the `donor pool,' a synthetic control can be meaningfully compared to the actual case. In this way, the technique allows the researcher to use the potential outcomes framework even in unusual conditions.

Future research can extend the present findings in a number of ways. First, it would be interesting to test whether other criminal activities have been affected by the state government policies I mentioned previously. Since property crimes are pervasive in São Paulo, scholars could evaluate the causal link (or lack thereof) between public policies and the incidence of theft or robberies. Unfortunately, several states in Brazil do not publish time-series data for property crime, so I could not use the synthetic control method for that dependent variable. As more data become available, this will create an interesting opportunity for investigation. Secondly, micro-level studies are needed to clarify the mechanisms behind São Paulo's homicide reduction, and isolate direct from indirect effects of each individual policies. Due to the shortage of data on targeted policies, qualitative research may explain what the motivations, successes and shortcomings of São Paulo's recent security measures were. Finally, there are still unresolved questions with regards to the `PCC hypothesis', and this is a promising avenue for future academic work. New research could provide insights into how public policies work and, hopefully, help public authorities to design more effective policies against crime.

\newpage

\section{Appendix} 
\label{sec:appendix_a}

This appendix presents a formal presentation of the synthetic control estimator. Let $j = 1, \dots, J + 1$ be a series of units in periods $t = 1, \dots, T$. In our case, the units are the 27 Brazilian federal states and the time period spans from 1990 to 2009. Assuming that the first unit, São Paulo, has been exposed to the treatment, we have $J$ control units to be included in the case studies donor pool, \emph{i.e.} the 26 remaining states. We define treatment as the series of post-1999 government anti-crime policies implemented in the São Paulo.

Let $Y_{it}^N$ be the homicide rate that would be observed for unit $i$, São Paulo, at time $t$ with no treatment (1990--1998). Conversely, let $Y_{it}^I$ be the observable outcome for unit $i$ at time $t$ had it been subjected to the treatment in periods $T_{0} + 1$ to $T$ (1999--2009). An important assumption is that the treatment has no effect on unit $i$ before the date of intervention, therefore, the values for São Paulo with and without the policy interventions are the same for the pre-treatment period (1990--1998). In formal terms, $Y_{it}^I = Y_{it}^N$ $\forall t < T_{0}$. The observed outcome is defined by $Y_{it}^I = Y_{it}^N + \alpha_{it}D_{it}$, where $\alpha_{it}$ is the effect of crime-reducing policies on homicide rates, and $D_{it}$ is a binary variable that takes the value of $1$ if we refer to post-intervention period (after 1999) and $0$ otherwise. The goal of this paper is to estimate $\alpha_{it}$, the effect of the treatment (homicide reduction policies), for the state of São Paulo for all $t \geq T_0$, that is, from 1999 to 2009. However, we cannot observe São Paulo \emph{without} those policies, as there is no way for the state to have and not have the intervention at the same time. This is what \citet{holland1986} calls the ``fundamental problem of causal inference'': only one of the outcomes of interest is measurable at any given time.

But although we cannot accurately know how São Paulo would be without the treatment, we can approximate it by using a weighted average of the remaining Brazilian states such that $Y_{it}^N = \delta_{t} + \theta_{t}Z_{i} + \lambda_{t}\mu_{i} + \epsilon_{it}$. In this model, $\delta_{t}$ is an unobserved time-dependent factor common to all cases, $Z_{i}$ is a $(1 \times r)$ vector of observed control variables not affected by the policy, $\theta_{t}$ is a $(r \times 1)$ vector of unknown time-specific parameters, $\lambda_{t}$ is a $(1 \times F)$ vector of unknown common factors to all states, $\mu_{i}$ is a state-specific unobservable variable and $\epsilon_{it}$ represents unobserved transitory shocks with mean $0$ for all units (error term). Basically, what SCM tries to do is to match $Z_{i}$, the control variables, and the pre-treatment $Y_{it}$ of São Paulo (1990--1998) so that $\mu_{i}$ is matched as a result.

Synthetic São Paulo is the weighted average of the other 26 Brazilian states. Thus, it is a $(J \times 1)$ vector of weights $W = (w_2 , \dots, w_{J+1})'$ with $w_j \geq 0$ for $j = 2, \dots, J + 1$ and $w_2 + \dots + w_{J+1} = 1$. Each of the elements included in $W$ represents a specific weighted average of control states, that is, a potential synthetic control for São Paulo. The idea is to select a case that resembles São Paulo as closely as possible. Let $X_{1}$ be a $(k \times 1)$ vector of pre-1999 predictor variables for São Paulo and let $X_{0}$ be a $(k \times J)$ matrix containing the predictor variables for the potential control states. Let $\bar{Y}_{i}^{K_{1}}, \dots, \bar{Y}_{i}^{K_{M}}$ be $M$ linear functions of pre-treatment outcomes $(M \geq F)$. One can choose $w^*$ such that:

$$\sum_{j = 2}^{J + 1} w_{j}^{*} Z_{j} = Z_{1}, \sum_{j = 2}^{J + 1} w_{j}^{*}\bar{Y}_{j}^{K_{1}} = \bar{Y}_{1}^{K_{1}}, \dots, \sum_{j = 2}^{J + 1} w_{j}^{*}\bar{Y}_{j}^{K_{M}} = \bar{Y}_{1}^{K_{M}}$$

Consequently, as noted by Abadie and his collaborators (\citeyear{abadie2010}), if $T_{0}$ is sufficiently large when compared to the scale of $\epsilon_{it}$, an approximately unbiased estimator for $\alpha_{1t}$, the effect of public security policies in São Paulo, can be described by:

$$\hat{\alpha}_{1t} = Y_{1t} - \sum_{j = 2}^{J + 1} w_{j}^{*} Y_{jt}$$

for all $t \in \left\{T_{0} + 1, \dots, T \right\}$, that is, after the intervention period (1999--2009). In practice, $W^{*}$ is chosen non-parametrically as to minimise $||X_{1} - X_{0}W||$, subject to the weight constrains. Consider $||X_{1} - X_{0}W||v = \sqrt{(X_{1} - X_{0}W)'V (X_{1} - X_{0}W)}$, where $V$ is a $(k \times k)$ symmetric and semi-definite positive matrix with the relative importance of each assigned homicide rate predictor. From various possible ways of choosing $V$, in this paper I follow the recommendation of \citet{abadie2003} and choose $V^{*}$ as the value of $V$ that minimises the root mean squared prediction error (RMSPE) for homicide rates in the entire pre-treatment period (1990-1998).

### 2017-04-09 03-bicho.tex
Initial ShareLaTeX Import

\chapter{Beasts of Prey or Rational Animals? Private Governance in Brazil's \emph{Jogo do Bicho}}
\label{chap:bicho}

\section{Introduction}
\label{sec:intro}

In 1892, Baron João Batista de Viana Drummond came up with a new idea to fund his cash-strapped zoo. Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \emph{Jardim Zoológico,} or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city. But it lacked visitors. As an experienced businessman, Drummond soon realised the zoo would have to provide other kinds of entertainment to keep itself afloat. Among his suggestions, one seemed particularly promising: a lottery raffle.

The rules were straightforward. In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance. Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, making the game similar to a common raffle. But it did not take long until visitors could name their animals of choice. This change made the game considerably more appealing and lasts until this day \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.\footnote{The amount was higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.} The lottery was labelled as the \emph{jogo do bicho}, or the animal game, and it was immediately adopted by the public. Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but in stores across Rio de Janeiro. He rightly predicted that this small change would increase profits, but there was one thing the Baron did not foresee. He unleashed a new gambling market.

The \emph{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace \citep{magalhaes2005ganhou, soares1993jogo}. A network of street bookmakers, called \emph{bicheiros}, expanded spontaneously the original game in innovative ways. Evading state regulations, \emph{bicheiros} made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the clandestine numbers game  \citep[37]{chazkel2011laws}. The \emph{jogo do bicho} became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today {[}1895{]}, in Rio de Janeiro, the game is everything. {[}\ldots{}{]} Nobody works! Everybody plays' \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.} But this tolerant state of affairs did not last. Civil servants and police officers criminalised the \emph{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos, krelling2014jogos, villar2008contravencao}.\footnote{The National Lottery Company (\emph{Companhia das Loterias Nacionaes do Brazil}), a public-private partnership founded four years after the creation of the \emph{jogo do bicho}, also lobbied actively for a hard-line stance against the \emph{bicheiros} \citep[82]{da1999aguias}.}

Yet the game has survived. The animal game has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{gaspari2002ditadura, jupiara2015poroes}. But more than a act of defiance, the \emph{jogo do bicho} is a successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}. A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \emph{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \url{http://goo.gl/9kNeX8} and \url{http://goo.gl/8FSAZl} (in Portuguese). Access: December 2016.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \emph{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the Brazilian workforce. See \url{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}. Access: December 2016.}

\begin{figure}[!htbp]
	\centering
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho01.jpg}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho02.jpg}
	\end{minipage}
	\caption{Left: cartoon of the Baron of Drummond and the animals of the \emph{jogo do bicho} (1896). Right: entry ticket to Rio's Zoological Garden that allowed the bearer to join the raffle. Sources: Instituto Histórico e Geográfico Brasileiro, Rio de Janeiro, Revista Illustrada, ano 21, no. 718 (1896) and Museu da Imagem e do Som, Rio de Janeiro. Reproduced in \citet[35--36]{chazkel2011laws}.}
	\label{fig:barao}
\end{figure}

Moreover, the animal game plays a crucial role in the expansion of Rio de Janeiro's Carnival Parade, a popular festivity synonymous with Brazil at home and abroad \citep{araujo2003carnaval,costa2001100,da1973carnaval, da1979carnavais,vianna1995misterio}. \emph{Bicheiros} donate hefty sums to `samba schools' to gather support of poor communities and, no less importantly, to co-opt local politicians attracted by the financial and electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}. This patron-client relationship has been proven effective: in 2016, the Carnival generated about USD 900 million in revenue and the Rio de Janeiro state received more than one million tourists.\footnote{Data provided by the Brazilian government. See \url{https://goo.gl/XMcbTM} (in Portuguese). Access: December 2016.}

In this article I offer a rational choice interpretation of the \emph{jogo do bicho} and discuss how \emph{bicheiros} promote social order, solve information asymmetries, and reduce negative externalities. My analysis discusses three strands of academic literature. First, this work contributes to the scholarship on extra-legal institutions, mainly to the literature on collective action within criminal organisations. For instance, \citet{gambetta1996sicilian} examines the strategies used by the Sicilian Mafia to settle disputes among their members and enforce rules in the areas they exercise control. \citet{leeson2007arrgh,leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations. \citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking and coordinating productive activities in California prison gangs. I argue that \emph{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and foster trust among criminals. Moreover, I also describe how \emph{bicheiros} have developed sophisticated financial mechanisms, such as informal hedging operations and risk-sharing contracts, to prevent predatory behaviour in their community.

Second, this work relates to the literature on repugnant transactions and the relationship between morality and the market \citep{boettke1995morality, roth2007repugnance, sandel2012money, satz2010some, simmel1900geldes, zelizer1979morals}. In the following sections, I claim that the Brazilian elites have attached pejorative meaning to the \emph{jogo do bicho} to constrain the gambling market. I provide evidence that \emph{bicheiros} were aware of this problem, and as a response, they devised a series of rules aimed at reducing the costs associated with repugnance \citep{labronici2014sorteio, magalhaes2005ganhou}. \emph{Bicheiros} have made considerable efforts to increase the levels of trust in the system and distance themselves from other types of illegal activities. Their main tool to increase credibility was costly signalling, that is, the \emph{bicheiros} hoped the public would see them as credible brokers by sacrificing their immediate interests  \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}.

Lastly, this work connects to the literature on state capture, which is among the most important topics in public choice theory \citep{hellman2003seize, rose1978corruption, rose1999corruption, shleifer2002grabbing, tollison1982rent}. More specifically, I use the Brazilian case to illustrate how politicians and civil servants can be co-opted by criminal groups and produce sub-optimal social outcomes. \citet{queiroz1992carnaval} explored why \emph{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities. \citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested that the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s. In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \emph{jogo do bicho} and the military regime in Brazil (1964--1985). I supplement this literature by highlighting how asymmetrical information, agency dilemmas, and rent-seeking behaviour offer convincing explanations to the issues presented above. Although those concepts have a long tradition in public choice, scholars have not applied those ideas thus far to understand the dynamics of the \emph{jogo do bicho}. By doing so, I integrate seemingly contradictory historical facts into a single narrative that connects micro-level decisions to macro-level outcomes.

The remainder of this article proceeds as follows. Section \ref{sec:organisation} presents a brief historical overview of the \emph{jogo do bicho}. It describes the necessary conditions for the emergence of the game and presents its basic organisational structure. Section \ref{sec:governance} details the \emph{jogo do bicho}'s governance mechanisms, particularly the strategies employed by vendors to increase trust in markets that operate at the margins of the law. Section \ref{sec:capture} discusses the links between illegal gambling markets, samba schools and the Brazilian state. Section \ref{sec:conclusion} offers some concluding remarks.

\section{\emph{Jogo do Bicho} as an Emergent Institution}
\label{sec:organisation}

\subsection{Historical Background}
\label{sub:background}

The early history of the \emph{jogo do bicho} is a textbook example of spontaneous order. Spontaneous orders are emergent macro-level phenomena that result from voluntary actions of purposive, self-interested individuals utilising their contextual knowledge \citep{boettke1990theory, boettke2005methodological, hayek1945use, hayek1960constitution, hayek1973law, leeson2008coordination, menger1871grundsatze, polanyi1948planning, polanyi1951logic}. Drummond, the Zoological Garden's original owner, designed the basic framework for the \emph{jogo do bicho}; but independent bookmakers were the ones who popularised the game \citep[77]{magalhaes2005ganhou}. Ticket sellers could quickly respond to market signals and then allocate their products where they were more valuable because of the lack of central coordination. Moreover, competition among sellers fostered innovation, and the \emph{bicheiros} invented new game rules to make the lottery more appealing to their customers \citep[61]{mello1989historia}. In this sense, the animal game is the materialisation of an evolutionary process of entrepreneurial discovery in which the interactions that provided the highest value to consumers were preserved over time \citep{boettke2008gordon, boettke2014entrepreneurship, buchanan1964should, hayek1978competition, kirzner1997entrepreneurial}.

However, the \emph{jogo do bicho} only emerged because of historically contingent circumstances. The late nineteenth-century Brazil had four characteristics that explain how the animal game came to being: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement. Figure \ref{fig:dag} presents a simple directed acyclic graph (DAG) \citep{pearl2009causality} that shows the relationships between these explanatory variables and the development of the \emph{jogo do bicho}.\footnote{The main purpose of direct acyclic graphs is to graphically display the possible links among the exposure variables, confounders, and outcomes \citep{morgan2014counterfactuals, pearl2009causality}. DAGs are transparent by definition, as all theoretical choices made by the researcher are stated explicitly in the model. Each single-headed arrow in a DAG indicates that the variable at the origin causes the variable at the end of the directed edge. Dashed edges suggest that two variables are jointly dependent on unobserved common causes. There are no assumptions regarding the functional form of the relationships, and unless mentioned otherwise, the arrows represent fully non-parametric associations. Variables between two nodes are mediators, and variables pointed at by two or more factors have multiple causes. They are called \emph{colliders}. For the sake of clarity, errors are assumed independent and often excluded from the graphs. For an accessible introduction to DAGs see \citet[chap. 3--4]{morgan2014counterfactuals} and \citet{pearl2016causal}.}

\begin{figure}[!htbp]
	\begin{tikzpicture}[->,>=stealth',shorten >=4pt,auto,node distance=3cm, semithick][h]
		% nodes %
		\node[circle,fill,inner sep=3pt,label=above right: \shortstack{Urban \\ Poverty}] (p) {};
		\node[circle,fill,inner sep=3pt,label=right: \shortstack{Jogo \\ do Bicho}, right = 8 of p] (y) {};
		\node[circle,fill,inner sep=3pt,label=below left: {Immigration}, below = 3 of p] (i) {};
		\node[circle,fill,inner sep=3pt,label=above:{Money Supply}, above = 3 of p] (m) {};
		\node[circle,fill,inner sep=3pt,label=below right:{Social Ties}, right = 3 of i] (s) {};
		\node[circle,fill,inner sep=3pt,label=above:{Weak Law Enforcement}, above = 3 of y] (l) {};
		\node[circle,fill=white, draw, outer sep=0pt, inner sep=3pt,label= left: \shortstack{Correlated \\ Background Factors}, above left = 2 of p] (u) {};
		% edges %
		\draw[->, line width = 1.2] (p) -- (y);
		\draw[->, line width = 1] (s) -- (y);
		\draw[->, line width = 1] (i) -- (s);
		\draw[->, line width = 1.2] (l) -- (y);
		\draw[->, line width = 1] (m) -- (y);
		\draw[->, line width = 1.2] (i) -- (p);
		\draw[->, dashed, line width = 1] (u) -- (p);
		\draw[->, dashed, line width = 1] (u) -- (m);
	\end{tikzpicture}
	\caption{Directed Acyclic Graph -- Explanatory Variables for the \emph{Jogo do Bicho}}
	\label{fig:dag}
\end{figure}

I start with the impact of urban poverty on the animal game. Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, fausto2014concise, naro1992revision, skidmore1993black}. The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration, smith1979ethnic}. Nevertheless, the job market tightened considerably after the \emph{Encilhamento} financial crisis of 1891 \citep{topik2014political, triner2005baring}. During the economic downturn, the informal economy was an obvious destination for the urban poor. Given its widespread popularity, the \emph{jogo do bicho} attracted hopeful entrepreneurs, either Brazilian or foreign-born, who could not enter the formal labour force.\footnote{The underground economy was also more democratic than the formal sector. As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor. These vendors used to sell different types of merchandise and many of them would later offer \emph{jogo do bicho} tickets.}

The immigration also influenced the \emph{jogo do bicho} via social ties. Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{klein1994imigraccao, lobo2001imigraccao, trento1989outro}. Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}. Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, torcato2011repressao, villar2008contravencao}. Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \emph{jogo do bicho}'s formative years.

Next is the impact of expanded monetary supply. The abolition of slavery and the growing industrialisation of Brazil increased the amount of capital available in the country \citep{franco1987reformas, schulz2008financial}. The 1888 Banking Act gave extra liquidity to local financial markets, and the \emph{jogo do bicho} entrepreneurs utilised that increase in the monetary base to extend the scope of their business. Some years later, the animal game would be available not only across the city of Rio de Janeiro but throughout Brazil \citep[76]{da1999aguias}.

The country's lax financial policy might correlate with poverty through unspecified factors. For instance, political decisions may have caused inadvertently both poverty and the expansion of the monetary base \citep{mattos2013shantytown, schmidt1982modernization}; alternatively, external events such as institutional instability \citep{costantini2014index, fausto2014concise, luna2014economic} or commodity shocks \citep{musacchio2014colonial} could be the cause of those two variables. There is not enough evidence to discard such scenarios. To illustrate this uncertainty, the two nodes, namely, money supply and urban poverty, are connected with a dashed edge in Figure \ref{fig:dag}.

The last necessary condition for the emergence of the \emph{jogo do bicho} is weak law enforcement. \citet[69--100]{chazkel2011laws} notes that until the 1940s police district chiefs operated within a large margin of discretion and repression against bookmakers was idiosyncratic. Prosecution against the \emph{bicheiros} had hardened in 1917, but only in 1946, when the federal government banned all gambling activities in the country \citep[155--156]{magalhaes2005ganhou}, the law was consistently enforced.

\subsection{Organisational Structure}
\label{sub:organisation}

The animal game has three levels of hierarchy. At the bottom are the \emph{bicheiros}, who are those in charge of selling \emph{jogo do bicho} tickets \citep{chazkel2007beyond, chazkel2011laws, da1999aguias, labronici2014sorteio, magalhaes2005ganhou, misse2007illegal}. \emph{Bicheiros} are the most visible part of the \emph{jogo do bicho} structure. The bookmakers often build their vending stands inside the premises of a local shop, such as a small grocery store or a pub, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}. \emph{Bicheiros} usually work alone, but they may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.

The \emph{gerentes} (managers) oversee all \emph{jogo do bicho} stands in a given area. Their task is akin to that of a firm accountant. Gerentes control the cash flow between the \emph{bicheiros} and the bankers, manage the payroll of the employees, and provide financial information to the top members of the organisation. They also supervise individuals who carry menial tasks in the business, transfer money to other gambling branches and double-check the balance sheets of the betting sites \citetext{\citealp[71]{labronici2012paratodos}; \citealp[142]{misse2007illegal}}.

\begin{figure}[!htbp]
	\centering
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho03.jpg}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth, height=6cm]{images/bicho04.jpg}
	\end{minipage}
	\caption{\emph{Jogo do bicho} betting sites in 1917 and in 2011. The picture on the right shows a \emph{bicheiro}, a street-corner vendor. Sources: \citet{alecrim2012bicho} and \citet{ferrarini2011bicho}.}
	\label{fig:banca1917}
\end{figure}

The \emph{banqueiros}, or the Portuguese for bankers, occupy the top position in the \emph{jogo do bicho} hierarchy, comprising the small financial elite of the game. A 2012 report by the Brazilian Federal Police affirmed that 10 \emph{banqueiros} controlled the market throughout the country; five of them based in the state of Rio de Janeiro \citep{globo2012contraventores}. Apart from funding the game, the bankers provide support for the employees to undertake their activities. The \emph{banqueiros}' main attributions include paying bribes to police personnel, bailing out sellers arrested by security forces, and offering judicial assistance to employees in case of legal persecution \citep[75]{labronici2012paratodos}.

\emph{Banqueiros} run their businesses from fortified houses in unknown locations, the \emph{fortalezas} (`forts'). The first \emph{fortalezas} likely appeared in the 1950s, when the animal game was already well-established across the Brazilian territory. The period coincides with a time when the \emph{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}. Due to the growing size of the \emph{jogo do bicho} economy, \emph{banqueiros} decided to move their operations away from the public to avoid police persecution and reduce coordination costs.

Although the forts provided safety to the bankers, the existence of those hideouts posed a challenge to the organisation. Bankers removed from the public view are not accountable to players and booking agents. Similarly, bankers and managers working in the \emph{fortalezas} cannot oversee their employees as effectively as before. Considering that the animal game itself is illegal and the amount of money involved in the bets is often substantial, both \emph{banqueiros} and booking agents have strong incentives to defect. Players, in turn, have no evident reason to trust \emph{banqueiros} or \emph{bicheiros}. How do \emph{jogo do bicho} agents overcome trust issues and cooperate under uncertainty?

I argue that the \emph{jogo do bicho} solves problems of internal cooperation by providing club goods \citep{buchanan1965economic, berman2008religion, berman2009radical, leeson2011government, roth2014prison} while simultaneously shunning cheaters through punishments and appeals to `the shadow of the future' \citep{axelrod1984evolution, axelrod1985achieving, bo2005cooperation, roth1978equilibrium}. Clients and \emph{bicheiros} cooperate based on trust-enhancing mechanisms, most of them devised specifically for the \emph{jogo do bicho} \citep{da1999aguias, magalhaes2005ganhou}. Such mechanisms are relevant because they have allowed the \emph{jogo do bicho} to distance itself from other shadow markets and become a profitable enterprise in the long run.

\section{Governance of the \emph{Jogo do Bicho}}
\label{sec:governance}

\subsection{Gambling Markets and Repugnant Transactions}
\label{sub:repugnance}

The \emph{jogo do bicho} is a repugnant market. Individuals that like to gamble cannot do so because of strong moral objections from outsiders \citep{brisset2016marche, roth2007repugnance, satz2010some, zelizer1979morals}. As early as in 1890, Brazilian public authorities positioned themselves against the \emph{jogo do bicho} arguing that `{[}\ldots{}{]} this type of amusement is prejudicial to the interests of the unwise, who are naively seduced by the deceptive hope of uncertain lucre' \citep[544]{chazkel2007beyond}. In 1941, the government banned the animal game;\footnote{See: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm} (in Portuguese). Access: December 2016.} five years later, it prohibited all games of chance.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'. The text can be read at: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm} (in Portuguese). Access: December 2016.} The \emph{jogo do bicho}, casinos and bingos remain illegal in the country. Recent estimations show that the prohibition of the \emph{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players. \citep{congressoemfoco2015bicho, fsp2016legalizarbicho}.

In contrast with the official statements, the noxious element of the \emph{jogo do bicho} does not come from its inherent randomness. The game is `repugnant' precisely because it is \emph{a market}, a setting in which individuals can monetise the entertainment for private profit \citep{chazkel2007beyond, chazkel2011laws}. The Brazilian state has never seen any contradiction between banning games of chance and running a national lottery company of its own; even the Catholic Church, which has long condemned the \emph{jogo do bicho}, frequently organises raffles to fund its activities \citep[49]{abreu1996imperio, magalhaes2005ganhou}. Only after the introduction of private money that Brazilians objected the idea of benefiting from someone else's bad luck.

In this sense, the main obstacle that confronted \emph{bicheiros} was to convince others that the animal game would not cause the Brazilian society `to slide down a slippery slope to genuinely repugnant transactions' \citep[45]{roth2007repugnance} such as prostitution or debt bondage. As the century-old history of the game can attest, \emph{bicheiros} have succeeded in this task. But how? The literature on repugnant costs tell us little about how markets transition from noxious to tolerated. Here I posit two mechanisms that reduced the stigma associated with the game: 1) \emph{a strong reputation of honesty} expressed by costly signals from sellers, and 2) the provision of \emph{selective incentives} for both clients and booking agents. Below, I offer evidence that these two factors allowed the animal game to reach its current semi-legal status in Brazil.

\subsection{External Cooperation}
\label{sub:external}

Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and experimental studies \citep{dawes1977behavior, isaac1984divergent, kim1984free, marwell1981economists} have both demonstrated that long-term cooperation is possible whenever players expect future pay-offs to be higher than present ones. Fear of retaliation induces individuals not to cheat. Nevertheless, illegal organisations tend to discount the future even more heavily than the other groups, what makes cooperative behaviour among criminals uncommon \citep{gambetta2009codes, skarbek2011governance,skarbek2012prison,skarbek2014social}. The \emph{jogo do bicho} is an exception to this rule. The market properties of the game and inconsistent repression by Brazilian authorities have permitted \emph{bicheiros} to overcome the stigma of repugnance and improve the game's long-term profitability.

The \emph{jogo do bicho} entrepreneurs have made considerable efforts to present themselves as honest brokers. The first trust-enhancing mechanism they have employed to foster external cooperation was the use of a \emph{fixed-multiplier formula} for pay-outs. It works as follows. If a player wins the lowest prize of the animal game, he or she receives 18 times his/her investment regardless of the size of the bet. Bigger prizes naturally offer higher returns; a lucky winner of the top prize wins up to 4,000 times the value of his/her bet \citetext{\citealp[89]{labronici2012paratodos}; \citealp[20]{magalhaes2005ganhou}}.

This stands in sharp contrast to the common practice of sharing a prize among winners. Lottery pay-outs demand high levels of interpersonal trust: players rely on unverifiable information about the total funds collected by the lottery, and they can never be sure whether the payments are evenly distributed. The fixed-multiplier formula alleviates such problems of adverse selection \citep{akerlof1970market, cohen2010testing, levin2001information}. As players and vendors known the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \emph{bicheiros} to a contract that can be easily enforced. This technique offers buyers a simple yet effective screening strategy that induces \emph{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.

\emph{Bicheiros} have addressed information asymmetries in another ways. Since the 1950s, when the \emph{jogo do bicho} bankers had moved their operations to the \emph{fortalezas}, the public could not oversee the lottery draws \citep[259]{chazkel2011laws}. This could lead to a decline in trust among buyers and vendors of lottery tickets and, as a result, to reduced profits. \emph{Bicheiros} have mitigated this problem with a two-pronged strategy. First, they started to utilise the winning numbers from the licit government-run lottery, the \emph{Loteria Federal}, instead of their own draws \citetext{\citealp[546]{chazkel2007beyond}; \citealp[89]{labronici2012paratodos}; \citealp[39-40]{mello1989historia}}. The federal lottery numbers are public information. The media broadcasts the draws on radio and TV, so any interested player can verify the selected numbers. The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public; hence, \emph{bicheiros} can free ride on the lottery's long-standing reputation of credibility.\footnote{As of April 2016, the lottery was audited by the \emph{Controladoria Geral da União} (Comptroller General of Brazil), the \emph{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young. The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute. See \url{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm} (in Portuguese). Access: December 2016.}

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=\textwidth, height=6cm]{images/bicho06.jpg}
	\caption{Results of a \emph{jogo do bicho} draw from 09 January 2016. `Federal' means that the winning numbers were drawn by the federal government lottery. The first prize was group 14, the cat. Source: Unknown. Available at: \url{https://goo.gl/6PHV8u}. Access: December 2016.}
	\label{fig:federal}
\end{figure}

Second, they included representatives of all major \emph{jogo do bicho} bankers in every draw and independently publicise the game results. Certain \emph{bicheiros} went as far as publishing the numbers in Rio's newspapers. In the early twentieth century, some tabloids were entirely dedicated to the game \citep[60]{magalhaes2005ganhou}. Booking agents see this strategy as a credible signal from the game financiers, as providing contrasting information would indicate game manipulation. Moreover, collusion can also be spotted if the draws show repeated numbers or unusual patterns.

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=\textwidth, height=6cm]{images/bicho05.jpg}
	\caption{\emph{Jogo do bicho} results are fixed on light poles in Rio de Janeiro. Source: \citet{gomes1998bicho}.}
	\label{fig:poste}
\end{figure}

These efforts have proved popular with the game enthusiasts. One often-repeated saying about the \emph{jogo do bicho} is that `in the \emph{jogo do bicho}, what is written down counts' \citep[159]{chazkel2011laws}, that is, buyers and sellers do fulfill their informal obligations without third-party enforcement. Such mutual confidence reduces the potential for conflict in the game. As the public does not see the \emph{jogo do bicho} as violent or harmful, the stigma of repugnance associated with gambling becomes less pervasive. By reducing the possibilities of cheating and putting long-term interests first, the \emph{jogo do bicho} bankers have avoided the fate of other repugnant markets and run their business relatively undisturbed for decades \citep[20]{da1999aguias}.

\subsection{Internal Governance}
\label{sub:internal}

Individuals working at different levels of hierarchy often have non-aligned interests. As a result, it may occur that one party (the agent) behaves rationally in a manner that maximises his/her benefits, but that is contrary to the interests of his/her superior (the principal). This dilemma is pervasive in formal organisations \citep{holmstrom1979moral,jensen1976theory,moe1984new,shapiro2005agency,spence1971insurance}; in illegal markets perhaps it is even more so \citep{campana2013cooperation,gambetta2009codes,skarbek2011governance,skarbek2014social}. As monitoring costs in criminal businesses are higher than in formal ones, principals face considerable difficulties to induce cooperation from agents. Moreover, criminals often engage in opportunistic behaviour and `hidden actions', that is, they do not put the required levels of effort if they know they are not being monitored \citep[38--42]{arrow1985agency}.

In the \emph{jogo do bicho} setting, one such problems concerns the trade-off between short- and long-term incentives for managers and bankers on the one side and bookmakers on the other. Managers have a permanent interest in the long-run profitability of the game, whereas street-corner booking agents tend to discount the future more heavily because their financial gains are small compared to that of their superiors. Additionally, bookmakers may denounce their employers to the police if they feel threatened.

One way by which the \emph{jogo do bicho} principals solve the agency dilemma is by supplying club goods and selective incentives for low-tier members. Club goods are goods that can be simultaneously enjoyed by more than one individual but where exclusion mechanisms prevent consumption by non-members \citep{buchanan1965economic, cornes1996theory, olson1965logic, sandler1980economic, sandler1997club}. Basically, club goods are `public goods \emph{sans} non-excludability' \citep[928]{mcnutt1999public}. The first club good offered to \emph{bicheiros} by their bosses is private security. The game bankers have built an extensive network of gunmen and bribed police officers to protect their employees (and their profits) from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}. The \emph{jogo do bicho} network has a powerful deterrence effect and lethal force is rarely employed. However, threats are constant. `Zé' (Little Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \emph{jogo do bicho} informal security personnel:

\begin{quote}
	[\dots] bums are scared and they don't mess around with us; they think there's a guard nearby or something like that. Look at all this money here! [shows the interviewer a handful of cash] It's not ours [referring to street-corner bookmakers]. And if it's not ours, it's someone else's. When I worked in Penha (\emph{a low middle-class neighbourhood in the city of Rio de Janeiro -- translator's note}), the owner of a pub close to where I used to work always asked me to stay at the front door of his pub. People know that bums are afraid of \emph{bicheiros}.
\end{quote}

Apart from guaranteeing the physical integrity of the \emph{bicheiros}, bankers and managers also provide financial incentives for the bookmakers. \emph{Bicheiros} are allowed to receive tips from players, often have small expenses paid by managers, and may even request interest-free loans to cover unexpected costs such as illness-related expenses \citep{labronici2012paratodos}.

However, the most important financial mechanism implemented by bankers to help \emph{bicheiros} is the \emph{descarga}, which is loosely translated as `the unloading'. The descarga is the \emph{jogo do bicho}'s main hedging technique and its purpose is to insure bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}; \citealp[16]{misse2007illegal}; \citealp[75]{soares1993jogo}}. Booking agents are sometimes unable to honour expensive bets. As mentioned above, the top prize in the animal game pays up to 4,000 times the amount invested, thus \emph{bicheiros} may have to raise thousands of Brazilian Reals in a single day. To prevent the \emph{quebra da banca} (`bust of the bank'), \emph{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}. The \emph{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \emph{jogo do bicho}.

The descarga has played an important role in reducing individual risk; nevertheless, it has also changed the distribution of resources in the \emph{jogo do bicho}. Simple probability dictates that a booking agent rarely pays the highest prize in the \emph{jogo do bicho}; in contrast, the bankers receive a commission for \emph{every game} they hedge. Over time, there is a transfer of income from the bottom to the top of the animal game structure led by this constant inflow of fees. This accumulation of capital is probably one of the reasons why bankers were able to diversify their businesses and offer other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}. The descarga has made the game more resilient at the aggregated level, although it increased profits for the richest financiers at the expense of small bookmakers.

\section{Tropical State Capture: \emph{Jogo do Bicho}, Samba and Politics}
\label{sec:capture}

The impact of the \emph{jogo do bicho} is not restricted to the Brazilian economy. Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural and social festivity, the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,chinelli1993vazio,queiroz1992carnaval}. The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that `without the \emph{jogo do bicho} the Carnival would have ended' \citep{odia2016aniz}. Owing to that support, \emph{bicheiros} have established an extensive patronage network with samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}. Although that network brings large material benefits to their members, the patronage system has created perverse incentives for government officials.

The \emph{jogo do bicho}'s clientelism is more evident in the state of Rio de Janeiro than in other parts of the country. Historical factors explain why this is the case. Firstly, Rio de Janeiro city was the capital of Brazil for almost 200 years; despite losing the position to Brasília in 1960, it remains one of the country's main cultural and financial centres. Secondly, \emph{jogo do bicho} operators had historical ties with popular movements, which they eventually exploited to their advantage. Thirdly, the emergence of state-sponsored Carnival parades created a window of opportunity for \emph{bicheiros} to expand their influence over public authorities, either via bribing or by funding political campaigns. In this regard, Rio provided a suitable environment for self-interested politicians, community leaders and animal game financiers to collaborate. These illegal networks are crucial to understand why samba and Carnival became constituent features of Brazil's national identity, and how the festival has contributed to Rio's high levels of state corruption.

\subsection{The `Medici of Samba': \emph{Bicheiros} as Patrons of Carnival}
\label{sub:patrons}

In 1930, opposition leader Getúlio Vargas led a bloodless coup d'état that brought Brazil's First Republic to an end. During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by dismantling effectively federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,desouza1983estado,fausto1972revoluccao,fausto2014concise,skidmore1967politics}. In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,nava1998lessons,williams2001culture}. Popular music, in turn, occupied an important place in Vargas's project of `brazilianing Brazil'. Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{avelar2011brazilian,mccann2004hello,stockler2011samba,vassberg1969villa,vassberg1975villa}.

By the late 1930s, samba reached a unique position in Brazil's cultural identity. In a period when civil and political rights were limited \citep{de2001cidadania,duarte1993vicissitudes}, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}. Patriotic sambas exalted the country's natural beauties and the figure of the `friendly, happy, cordial and industrious' mulatto\footnote{A mulatto is a person of mixed white and black ancestry. The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of the male donkey and a female horse. However, in the 1930s the word loses its pejorative connotation in Brazil. Mainly due to the work of sociologist \citet{freyre1933casa}, the idea of a racial democracy becomes pervasive in the government discourse, and as a result the word gains a positive tone \citep[4]{reiter2009brazil}.} \citetext{\citealp[47]{dangelo2016samba}; \citealp[51]{vianna1995misterio}}. The institutionalisation of the Carnival parade in 1935, and the subsequent increases in public funding to the festival, cemented the relationship between politicians and samba groups \citep{almeida2017carnaval,cabral2016escolas,soihet1998subversao}.

However, the samba groups were not passive members in this process. Since the 1960s, the Rio Carnival expanded in scope and, stimulated by growing numbers of spectators, the parades became more elaborate \citetext{\citealp{cabral2016escolas}; \citealp[214]{chinelli1993vazio}; \citealp[240]{hertzman2013making}}. Unable to cope with the rising costs of the show, the `samba schools', which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}. This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,farias2013carnival,misse2011crime,queiroz1992carnaval}.

As I have mentioned in the previous section, the animal game at times faced opposition by the local population. The public often perceived the game as immoral and repugnant. Moreover, even after the bicho was well-established in Rio de Janeiro, the transition from a competitive betting market to an oligopoly involved the threat and often the use of physical violence against bookmakers who resisted the change \citetext{\citealp[143]{bezerra2009mecenato}, \citealp[52]{labronici2012paratodos}}. \emph{Bicheiros} were aware of the reputation costs their strategy entailed. They decided to finance samba schools hoping to win `the hearts and minds' of the population and attach a more positive image of the game among urban classes. Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but only as individuals who had a private interest in samba \citep[209]{chinelli1993vazio}. In 1984, a group of rich \emph{jogo do bicho} financiers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro. The LIESA marked a shift in the Carnival. For the first time, \emph{bicheiros} decided to act as a group rather than individuals. The organisation consolidated the power of \emph{bicheiros} over the parade and provided a formal mechanism to solve disputes among the samba school patrons \citetext{\citealp[43]{cavalcanti2006carnaval}; \citealp[171]{farias2013carnival}; \citealp[55]{labronici2012paratodos}}.

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=.8\textwidth, height=8cm]{images/bicho07.jpg}
	\caption{Castor de Andrade is shown celebrating after the samba school he sponsored, Mocidade Independente, won the Carnival parade in 1996. He was surrounded by colleagues and police officers. Andrade was the founding president of LIESA (1984--1985) and the wealthiest \emph{bicheiro} of Rio de Janeiro at the time. Source: Folha Imagem. Reproduced in \citet[139]{misse2007illegal}.}
	\label{fig:castor}
\end{figure}

The funding of the samba schools had an indirect effect to the animal game. The patronage also reduced agent-principal problems within the \emph{jogo do bicho}. \emph{Bicheiros} donate to samba school to gather support of the communities, and by doing so they gain access to local information on their business. Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring is cost-effective for animal game managers. Thus, street bookmakers have fewer incentives to cheat. In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}'s donations \citep{bbc2012aniz}. Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation. The patronage can be interpreted as an illegal version of `profit-sharing', a mechanism which has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.

The samba schools have profited from this association too. First, they have gained autonomy from the government. The samba schools do not need to rely exclusively on public funds to organise the parade, and money from the \emph{jogo do bicho} permitted the schools to act independently \citep[209]{chinelli1993vazio}. Second, the support of the \emph{jogo do bicho} has increased the political and social clout of the samba schools. In a country where the state is not present throughout the territory and human right abuses are frequent \citep{ahnen2003,odonnell1993state,pinheiro2000,pinheiro2001}, \emph{jogo do bicho} bankers, and more recently drug traffickers, have provided private governance to poor areas of Rio de Janeiro by enforcing property rights, mediating disputes, and preventing police abuse in the favelas \citep{arias2006dynamics,goldstein2013laughter,leeds1996cocaine}. In return for funds and protection from the \emph{bicheiros}, samba schools have served as intermediaries between the underworld and the political system. Although the \emph{banqueiros} are interested in weak law enforcement against the animal game, politicians have resorted to samba schools to contact \emph{bicheiros} and use their financial and electoral influence in the shanty towns \citep[17]{misse2011crime}. The samba schools, therefore, have increased their bargaining power in the political sphere and extended their reach within Rio's poor communities \citep[215]{chinelli1993vazio}.

\subsection{Political Support}

If politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, their relationship with the animal game bankers have become more ambivalent in the last decades. The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citetext{\citealp{gaspari2002ditadura}; \citealp{jupiara2015poroes}; \citealp[39]{zaluar2007democratizaccao}}. Given the absence of democratic checks and balances, paramilitaries and police forces colluded to repress potential dissidents of the regime and, frequently, to extort civilians \citep{gorender1999combate,magalhaes1997logica,misse2009acumulaccao,skidmore1990politics}. \emph{Bicheiros} saw the corruption of some members of the military as an opportunity. Wealthy \emph{jogo do bicho} bankers hired rogue police officers not only to work as security guards but to threaten eventual competitors in their regions of influence. The agreement between \emph{bicheiros} and corrupt members of the military was the ultimate responsible for the transformation of the \emph{jogo do bicho} into a `coercive oligopoly' \citep{jupiara2015poroes}. The support of the armed forces meant that new groups would be prohibited from entering the market and that the illegal lottery could operate undisturbed by the government.

The links between \emph{bicheiros} and the public authorities changed after Brazil became a democracy in 1985. In the military regime, government officials were mainly interested in bribes from the animal game. But in the democratic period, votes became a sought-after political resource. \emph{Bicheiros} are important in this sense as they have direct influence over a number of poor communities. Their patronage networks ensure that candidates supported by \emph{bicheiros} receive a substantial amount of votes from areas where campaigning is too difficult or too costly \citep[17]{misse2011crime}.

The Brazilian political system is particularly conductive to clientelistic practices. Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}. In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp{mainwaring1992brazilian}; \citealp[483]{samuels2000ambition}}. These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence. Nevertheless, that independence means candidates rely mostly on themselves to raise funds and establish communication with potential voters. Hence, political campaigns in Brazil tend to be expensive and personality-centred.

The support from the \emph{jogo do bicho} mitigates both problems. With respect to the financial costs of campaigns, illegal donations from \emph{bicheiros} help to cover advertising expenses while having the additional benefit of not appearing in the official records of the candidates \citep{congressoemfoco2007bicho,gazetadopovo2007bicho,globo2012bicheiro}. This suggests that \emph{jogo do bicho}-funded politicians can circumvent spending limits and have an electoral advantage over their competitors. As candidates do not know whether their competitors receive funding from the \emph{jogo do bicho} nor the amount each one was paid, their dominant position is to contact the \emph{bicheiros} and join their networks. The situation is a prisoner's dilemma in which candidates would be better off by running cheaper campaigns and not being dependent of \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching an optimal solution.

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
	\caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right). The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'. Source: \citet{extra2012aniz}.}
	\label{fig:aniz}
\end{figure}

The votes from poor communities are instrumental for aspiring politicians. Brazil has an enforced compulsory voting system; therefore, turnout rates tend to be higher than in other democracies. Consequently, votes have high marginal utility for politicians. As elections may be decided by a small difference, the \emph{bicheiros}' clientelistic ties guarantee a minimum number of votes that politicians can rely upon on election day. Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process per se. Individuals may be punished if the candidate does not receive the expected number of votes, and are often compelled to vote for politicians that have only loose connections with their communities. Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil’s lower classes.

Finally, the \emph{jogo do bicho} patronage highlights a crucial social dilemma within the Brazilian public law. Even though federal judges have prosecuted \emph{bicheiros}, politicians and police forces have no incentives to enforce the punishment. Although Brazilian judges enjoy job stability, the latter groups constantly require local-level support from the \emph{bicheiros}. Politicians and police officers may have accurate information on \emph{jogo do bicho} operations and \emph{bicheiros}' whereabouts, but the federal government cannot rely upon their cooperation. That can be one of the reasons why even after many attempts to arrest \emph{bicheiros}, there has been little progress in that regard in Brazil's latest democratic period (1985--present).

\section{Conclusion}
\label{sec:conclusion}

Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}. Yet, the \emph{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years. \emph{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers. Furthermore, by investing in the Carnival parade \emph{bicheiros} have been able to gather popular and government support. Poor communities have associated with the \emph{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \emph{jogo do bicho}'s networks can provide.

Nevertheless, the \emph{jogo do bicho} has also created negative externalities. Violence is used to punish defectors and to constrain competitors. The clientelistic relationship that \emph{bicheiros} have with local politicians have lead to sub-optimal outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations. These negative externalities have long-term effects and still impact the Brazilian public sphere.

Although the \emph{jogo do bicho} has received an increasing attention from scholars, much of its inner workings remain poorly understood. First, the relationship between \emph{bicheiros} and drug dealers is a topic that deserves attention. Brazil has become one of the world's largest consumers of illicit drugs and South America's principal drug trafficking transit route \citep{miraglia2015drugs,misse2011crime}. The question whether \emph{bicheiros} collaborated or opposed the emergent drug dealing business is still unclear. Second, the extent to which \emph{bicheiros} use other businesses, such as hotels or factories, to laundry money has been mentioned by members of the Brazilian judiciary \citep{globo2012bicheiro,globo2015cacaniquel}; however, there is no reliable estimate on its size. Lastly, more research is required to clarify how \emph{bicheiros} from different parts of Brazil coordinate their activities and prevent large-scale conflicts. Cases studies are usually focused on Rio de Janeiro's \emph{bicheiros}, but scholars would benefit from comparative analyses with a larger number of states. This is an important step to elucidate how \emph{bicheiros} continue to influence politics and the public across Brazil.

### 2017-04-09 thesis.tex
Initial ShareLaTeX Import

\documentclass[a4paper,12pt]{report}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% University of Sussex thesis template
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Modification History
%
% Based on usthesis.cls by Jonathon Read
% http://www.cogs.susx.ac.uk/users/jlr24/latex.html
% Modified by Anthony Smith, Feb 2007
% Incorporated into single thesis.tex file, Anthony Smith, 30 June 2008
% Minor alterations to page numbering, AJS, 25 July 2008
% New alternative hyperref options for print version, AJS, 11 Sep 2008
% "DRAFT" on header, AJS, 12 Sep 2008
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LINE SPACING
\newcommand{\linespacing}{1.5}
\renewcommand{\baselinestretch}{\linespacing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BIBLIOGRAPHY STYLE
\usepackage[authoryear]{natbib}
% \bibliographystyle{plain} for [1], [2] etc.
\bibliographystyle{apalike}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% OTHER FORMATTING/LAYOUT DECLARATIONS
% Graphics
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{amssymb,amsmath}
\usepackage{lmodern}
\usepackage{libertine}
\usepackage[libertine]{newtxmath}
\usepackage[scaled=0.95]{inconsolata}
\usepackage{xcolor}
\definecolor{darkblue}{rgb}{0.0,0.0,0.55}
\usepackage{setspace}
\usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
\usepackage[backref,pagebackref]{hyperref}
\usepackage{graphicx}
\usepackage{float}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usepackage{mathtools}
\usepackage{caption}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\cleanlookdateon
\exhyphenpenalty=1000
\hyphenpenalty=1000
\widowpenalty=1000
\clubpenalty=1000
\flushbottom
% Pages should be numbered consecutively thorugh the main text.  Page numbers
% should be located centrally at the top of the page.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANY OTHER DECLARATIONS HERE:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% HYPERREF
\renewcommand*{\backref}[1]{}
\renewcommand*{\backrefalt}[4]{%
	\ifcase #1 (Not cited.)%
	\or        Cited on page~#2.%
	\else      Cited on pages~#2.%
	\fi}
\renewcommand{\backreftwosep}{ and~}
\renewcommand{\backreflastsep}{ and~}
%\urlstyle{same}  % don't use monospace font for urls

\hypersetup{pdftitle={Essays in Political and Criminal Violence},
	pdfauthor={Danilo Freire},
	pdfborder={0 0 0},
	breaklinks=true,
	linkcolor=black,
	citecolor=darkblue,
	urlcolor=darkblue,
	colorlinks=true}
% For print version, use this instead:
%\usepackage[pdfusetitle,bookmarksnumbered,plainpages=false]{hyperref}
%\usepackage{backref}
%\renewcommand{\backrefpagesname}{Cited on}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BEGIN DOCUMENT
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE: roman page numbering i, ii, iii, ...
\pagenumbering{roman}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% TITLE PAGE: 
\thispagestyle{empty}
\begin{flushright}
\includegraphics[width=6cm]{kcl.eps}
\end{flushright}	
\vskip40mm
\begin{center}
% TITLE
\huge\textbf{Essays on Political and Criminal Violence}
\vskip2mm
% SUBTITLE (optional)
%\LARGE\textit{How it all works}
\vskip5mm
% AUTHOR
\Large Danilo Freire
\normalsize
\vfill
\large
% QUALIFICATION
Submitted for the degree of Doctor of Philosophy \\
Department of Political Economy \\
King's College London	\\
% DATE OF SUBMISSION
April 2017
\end{center}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DECLARATIONS
\chapter*{Declaration}
\noindent 
I hereby declare that except where specific reference is made to the work of others, the contents of this dissertation are original and have not been submitted in whole or in part for consideration for any other degree or qualification in this, or any other university. This dissertation is my own work and contains nothing which is the outcome of work done in collaboration with others, except as specified in the text and Acknowledgements.

% ADDITIONAL DECLARATIONS HERE (IF ANY)

\vskip20mm
\noindent
Signature:
\vskip20mm
% AUTHOR
\noindent 
Danilo Freire
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SUMMARY PAGE
\chapter*{Abstract}
\renewcommand{\baselinestretch}{\linespacing}
\small\normalsize
% SUMMARY HERE (300 word limit for most subjects):

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGEMENTS
\chapter*{Acknowledgements}
\renewcommand{\baselinestretch}{\linespacing}
\small\normalsize
% ACKNOWLEDGEMENTS HERE:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLE OF CONTENTS, LISTS OF TABLES & FIGURES
\newpage
\pdfbookmark[0]{Contents}{contents_bookmark}
\tableofcontents
\listoftables
\phantomsection
\addcontentsline{toc}{chapter}{List of Tables}
\listoffigures
\phantomsection
\addcontentsline{toc}{chapter}{List of Figures}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAIN THESIS TEXT: arabic page numbering 1, 2, 3, ...
\newpage
\pagenumbering{arabic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-----------------------------------------------------
% Chapter 1: Introduction
%-----------------------------------------------------

\input{01-intro}

%-----------------------------------------------------
% Chapter 2: Synthetic Control
%-----------------------------------------------------

\input{02-synth}

%-----------------------------------------------------
% Chapter 3: Jogo do bicho
%-----------------------------------------------------

\input{03-bicho}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BIBLIOGRAPHY
\clearpage
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% START APPENDICES
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-----------------------------------------------------
% Appendix: Code
%-----------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% END DOCUMENT
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-04-09 01-intro.tex
Upgrade: first draft

\chapter{Introduction}
\label{chap:intro}

The literature on political and criminal violence has increased exponentially over the last decades. Although interstate wars have traditionally occupied a privileged position in this research agenda, scholars have broadened their analyses and considered a myriad of hitherto understudied phenomena. Civil wars \citep[][]{collier2004greed,fearon2003ethnicity,kalyvas2006logic}, genocides \citep[][]{mamdani2014victims, power2013problem}, ethnic conflicts \citep[][]{kaufmann1996possible, montalvo2005ethnic, sambanis2001ethnic}, wartime sexual abuse \citep[][]{cohen2013explaining,wood2006variation,wood2009armed}, electoral violence \citep[][]{hoglund2009electoral,wilkinson2006votes}, state-sponsored killings \citep[][]{harff1988toward, krain1997state,krain2005international,uzonyi2014unpacking}, terrorism \citep[][]{de2005quality,bueno2007propaganda,pape2003strategic}, drug-related violence \citep[][]{holmes2006drugs,lessing2015logics,richani2013systems, shirk2010drug}, street gangs \citep[][]{franzese2016youth,jones2009youth,rodgers2006living,sobel1987direct}, and prison gangs \citep[][]{dias2011pulverizaccao,freire2014,skarbek2011governance,skarbek2012prison,skarbek2014social} have moved to the centre stage of the discipline. This dissertation contributes to this growing literature.

The following chapters discuss different types of organised violence. In order to clarify specific aspects of my research questions, I employ an eclectic combination of research designs, ranging from statistical analysis to field experiments and qualitative case studies. This diversity of methods not only reflects the multiple angles by which violence can be analysed, but it is also a pragmatic response to problems which are common to researchers in this area, such as incomplete data, biased information, and the practical impossibility of random assignment of treatment. By using a number of methodological tools, I hope to overcome some of these difficulties and offer compelling evidence for the results presented below.

Regarding the geographical scope of this dissertation, most of the essays deals with issues of violence in Latin America, specially in Brazil. According to the World Bank, Latin America is home to about 8\% of the global population, yet it accounts for more than 30\% of the world's homicides.\footnote{See: \url{https://goo.gl/d2WC3V}. Access: April 2017.} Moreover, the yearly ranking by the Citizen's Council for Public Security and Criminal Justice (Consejo Ciudadano para la Seguridad Pública y la Justicia Penal), a Mexican non-governmental organisation, shows that 43 of the 50 most violent cities in the world are located in Latin America, including all in the top 10.\footnote{For the complete ranking, see \url{http://www.seguridadjusticiaypaz.org.mx/biblioteca/prensa/summary/6-prensa/239-las-50-ciudades-mas-violentas-del-mundo-2016-metodologia}. Access: April 2017.} Given the severity of violence in the continent, Latin America was a natural candidate to be the primary focus of this study.

Brazil provides a major example of the challenges of fighting violence in Latin America. Brazil has the highest absolute number of homicides in the world, about 56,000 per year, and the country hosts 19 of the 50 world's deadliest cities according to the above-mentioned ranking \citep{mapa2014, mexico2014,unodc2013}. Homicide rates have increased markedly after democratisation (1985), and whereas the country has tried several policies to reduce violence, the results are yet to be evaluated in a consistent fashion.

The first chapter attempts to address that issue. Although Brazil remains notably affected by civil violence, the state of São Paulo has made significant inroads into fighting criminality. In the last decade, São Paulo has witnessed a 70\% decline in homicide rates, a result that policy-makers attribute to a series of crime-reducing measures implemented by the state government \citep{goertzel2009,kahn2005papel}. While recent academic studies seem to confirm this downward trend, no estimation of the total impact of state policies on homicide rates currently exists. I fill this gap by employing the synthetic control method \citep{abadie2003,abadie2010,abadie2014}, a generalisation of differences-in-differences \citep{angrist2008mostly,bertrand2004much,imbens2009recent}, to compare these measures against an artificial São Paulo. The results indicate a large drop in homicide rates in actual São Paulo when contrasted with the synthetic counterfactual, with about 20,000 lives saved during the period. The theoretical usefulness of the synthetic control method for public policy analysis, the role of the \textit{Primeiro Comando da Capital}, a local prison gang, as a moderating variable, and the practical implications of the security measures taken by the São Paulo state government are also discussed.

The second chapter offers a rational choice account for the \textit{jogo do bicho}, the animal game, possibly the largest illegal gambling game in the world. The lottery has been running for over 120 years and according to estimations of Fundação Getúlio Vargas, a Brazilian think tank, it profits up to 800 million dollars per year.\footnote{See \url{http://goo.gl/9kNeX8} and \url{http://goo.gl/8FSAZl} (in Portuguese). Access: April 2017} The \emph{jogo do bicho} has exerted a significant impact on the Brazilian society. The lottery has been a major sponsor of the Carnival Parade in Rio de Janeiro, which is among the world's most famous popular festivals, and it has remained an important driver of state corruption in the country \citep{bezerra2009mecenato,chazkel2011laws,da1999aguias,labronici2012paratodos,magalhaes2005ganhou,soares1993jogo}. I investigate the institutions that have caused the \emph{jogo do bicho}'s notable growth and long-term survival outside the boundaries of the Brazilian law. I show how \textit{bicheiros} or bookmakers promote social order, solve information asymmetries, and reduce negative externalities via costly signalling and the provision of club goods. I also explain the emergence of the informal rules that govern the game as well as their enforcement mechanisms.

### 2017-04-10 01-intro.tex
Upgrade: second draft

@@
 The literature on political and criminal violence has increased exponentially over the last decades.
-Although interstate wars have traditionally occupied a privileged position in this research agenda, scholars have broadened their analyses and considered a myriad of hitherto understudied phenomena.
-Civil wars \citep[][]{collier2004greed,fearon2003ethnicity,kalyvas2006logic}, genocides \citep[][]{mamdani2014victims, power2013problem}, ethnic conflicts \citep[][]{kaufmann1996possible, montalvo2005ethnic, sambanis2001ethnic}, wartime sexual abuse \citep[][]{cohen2013explaining,wood2006variation,wood2009armed}, electoral violence \citep[][]{hoglund2009electoral,wilkinson2006votes}, state-sponsored killings \citep[][]{harff1988toward, krain1997state,krain2005international,uzonyi2014unpacking}, terrorism \citep[][]{de2005quality,bueno2007propaganda,pape2003strategic}, drug-related violence \citep[][]{holmes2006drugs,lessing2015logics,richani2013systems, shirk2010drug}, street gangs \citep[][]{franzese2016youth,jones2009youth,rodgers2006living,sobel1987direct}, and prison gangs \citep[][]{dias2011pulverizaccao,freire2014,skarbek2011governance,skarbek2012prison,skarbek2014social} have moved to the centre stage of the discipline.
+Although interstate wars have traditionally occupied a privileged position in this research agenda, scholars have broadened their scope to include a myriad of hitherto understudied phenomena into their research endeavours.
+Civil wars \citep[][]{collier2004greed,fearon2003ethnicity,kalyvas2006logic}, genocides \citep[][]{mamdani2014victims, power2013problem}, ethnic conflicts \citep[][]{kaufmann1996possible, montalvo2005ethnic, sambanis2001ethnic}, wartime sexual abuse \citep[][]{cohen2013explaining,wood2006variation,wood2009armed}, electoral violence \citep[][]{hoglund2009electoral,wilkinson2006votes}, state-sponsored killings \citep[][]{harff1988toward, krain1997state,krain2005international,uzonyi2014unpacking}, terrorism \citep[][]{de2005quality,bueno2007propaganda,pape2003strategic}, drug-related violence \citep[][]{holmes2006drugs,lessing2015logics,richani2013systems, shirk2010drug}, street gangs \citep[][]{franzese2016youth,jones2009youth,rodgers2006living,sobel1987direct}, and prison gangs \citep[][]{dias2011pulverizaccao,freire2014,skarbek2011governance,skarbek2012prison,skarbek2014social} have recently moved to the centre stage of the discipline.
 This dissertation contributes to this growing literature.
 
-The following chapters discuss different types of organised violence.
-In order to clarify specific aspects of my research questions, I employ an eclectic combination of research designs, ranging from statistical analysis to field experiments and qualitative case studies.
-This diversity of methods not only reflects the multiple angles by which violence can be analysed, but it is also a pragmatic response to problems which are common to researchers in this area, such as incomplete data, biased information, and the practical impossibility of random assignment of treatment.
-By using a number of methodological tools, I hope to overcome some of these difficulties and offer compelling evidence for the results presented below.
+In order to clarify crucial aspects of my research topic, I employ an eclectic combination of research designs, ranging from statistical analysis to field experiments and qualitative case studies.
+This diversity of methods not only reflects the multiple aspects of organised violence, but it is also a pragmatic response to problems which are common in this area, such as incomplete data, biased information, and the practical impossibility of random assignment of treatment.
+By using an array of methodological tools, I hope to overcome some of these inherent difficulties.
 
-Regarding the geographical scope of this dissertation, most of the essays deals with issues of violence in Latin America, specially in Brazil.
+Regarding the geographical scope of this dissertation, most of the following chapters deal with issues of violence in Latin America, specially in Brazil.
 According to the World Bank, Latin America is home to about 8\% of the global population, yet it accounts for more than 30\% of the world's homicides.\footnote{See: \url{https://goo.gl/d2WC3V}.
-Access: April 2017.} Moreover, the yearly ranking by the Citizen's Council for Public Security and Criminal Justice (Consejo Ciudadano para la Seguridad Pública y la Justicia Penal), a Mexican non-governmental organisation, shows that 43 of the 50 most violent cities in the world are located in Latin America, including all in the top 10.\footnote{For the complete ranking, see \url{http://www.seguridadjusticiaypaz.org.mx/biblioteca/prensa/summary/6-prensa/239-las-50-ciudades-mas-violentas-del-mundo-2016-metodologia}.
-Access: April 2017.} Given the severity of violence in the continent, Latin America was a natural candidate to be the primary focus of this study.
+Access: April 2017.} Moreover, the yearly ranking by the Citizen's Council for Public Security and Criminal Justice (Consejo Ciudadano para la Seguridad Pública y la Justicia Penal), a Mexican non-governmental organisation, shows that 43 of the 50 most violent cities in the world are located in Latin America, including all of the top 10.\footnote{For the complete ranking, see \url{http://www.seguridadjusticiaypaz.org.mx/biblioteca/prensa/summary/6-prensa/239-las-50-ciudades-mas-violentas-del-mundo-2016-metodologia}.
+Access: April 2017.} Given the severity of violence in the continent, Latin America was the natural candidate to be the primary focus of this study.
 
-Brazil provides a major example of the challenges of fighting violence in Latin America.
+Brazil exemplifies many of the challenges of fighting violence in Latin America.
 Brazil has the highest absolute number of homicides in the world, about 56,000 per year, and the country hosts 19 of the 50 world's deadliest cities according to the above-mentioned ranking \citep{mapa2014, mexico2014,unodc2013}.
@@
 
-The first chapter attempts to address that issue.
+The first chapter attempts to address this issue.
 Although Brazil remains notably affected by civil violence, the state of São Paulo has made significant inroads into fighting criminality.
@@
 
-The second chapter offers a rational choice account for the \textit{jogo do bicho}, the animal game, possibly the largest illegal gambling game in the world.
+The second chapter offers a rational choice account for the \textit{jogo do bicho}, or the `animal game', possibly the largest illegal gambling game in the world.
 The lottery has been running for over 120 years and according to estimations of Fundação Getúlio Vargas, a Brazilian think tank, it profits up to 800 million dollars per year.\footnote{See \url{http://goo.gl/9kNeX8} and \url{http://goo.gl/8FSAZl} (in Portuguese).

### 2017-04-10 thesis.tex
Upgrade: first draft
Upgrade: second draft
Upgrade: third version

@@
 % LINE SPACING
-\newcommand{\linespacing}{1.5}
+\newcommand{\linespacing}{2}
 \renewcommand{\baselinestretch}{\linespacing}
@@
 \hyphenpenalty=1000
-\widowpenalty=1000
-\clubpenalty=1000
-\flushbottom
-% Pages should be numbered consecutively thorugh the main text.  Page numbers
+\widowpenalty=10000
+\clubpenalty=10000
+%\flushbottom
+% Pages should be numbered consecutively through the main text.  Page numbers
 % should be located centrally at the top of the page.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% ANY OTHER DECLARATIONS HERE:
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \thispagestyle{empty}
-\begin{flushright}
+\begin{center}
 \includegraphics[width=6cm]{kcl.eps}
-\end{flushright}	
+\end{center}	
 \vskip40mm
@@
 % SUMMARY HERE (300 word limit for most subjects):
+This dissertation addresses four questions on political and criminal violence.
+The first essay is an empirical evaluation of a broad set of homicide reduction policies implemented in the state of São Paulo, Brazil.
+I employ the synthetic control method, a generalisation of differences-in-differences, to compare these measures against an artificial São Paulo.
+The results indicate a large drop in homicide rates in actual São Paulo when contrasted with the synthetic counterfactual, with about 20,000 lives saved during the period.
 
+The second essay offers a rational choice account for the Brazil's \textit{jogo do bicho}, or the `animal game', possibly the largest illegal gambling game in the world.
+I investigate the institutions that have caused the \textit{jogo do bicho}'s notable growth and long-term survival outside the boundaries of the Brazilian law.
+I show how \textit{bicheiros} or bookmakers promote social order, solve information asymmetries, and reduce negative externalities via costly signalling and the provision of club goods.
+I also explain the emergence of the informal rules that govern the game as well as their enforcement mechanisms.
+
+The third essay, co-authored with Gary Uzonyi, presents a meta-analysis of the determinants of government-sponsored violence.
+We employ traditional meta-analysis estimators and Bayesian meta-regressions with random and fixed effects to a set of 48 articles published in the last 20 years to evaluate the main predictors of mass killing onset.
+
+The fourth essay, co-authored with Umberto Mignozzetti, examines the specific mechanisms by which the \textit{Primeiro Comando da Capital} (PCC), a Brazilian prison gang,  has helped reduce violence in São Paulo.
+We address this gap by proposing a survey experiment to evaluate three competing mechanisms that might have plausibly caused the drop in homicides, namely economic welfare, protection against police harassment, and the local provision of law and order.Our main hypothesis is that the PCC proposed a cohesive system of law and order by informally securing property rights and solving conflicts through credible threats and a local monopoly of the use of force.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-04-22 chapters/01-intro.tex
Add folders

--- 2017-04-10 01-intro.tex
@@
 
-The second chapter offers a rational choice account for the \textit{jogo do bicho}, or the `animal game', possibly the largest illegal gambling game in the world.
+The second chapter offers a rational choice account for Brazil's \textit{jogo do bicho}, or the `animal game', possibly the largest illegal gambling game in the world.
 The lottery has been running for over 120 years and according to estimations of Fundação Getúlio Vargas, a Brazilian think tank, it profits up to 800 million dollars per year.\footnote{See \url{http://goo.gl/9kNeX8} and \url{http://goo.gl/8FSAZl} (in Portuguese).

### 2017-04-22 chapters/02-synth.tex
Add folders

--- 2017-04-09 02-synth.tex
@@
 SCM was specifically designed for situations where there is only one treated unit of interest, no readily-available counterfactual, and no certainty as to whether the treated and the control units follow parallel trends after the intervention \citep{abadie2003, abadie2010, abadie2014}.
-Moreover, SCM also has some of the desirable properties of other popular causal inference tools such as differences-in-differences \citep{angrist2008mostly, bertrand2004much} and matching estimators \citep{dehejia2002propensity, ho2007matching, rubin1973matching, stuart2010matching}.
+Moreover, SCM also has some of the desirable properties of popular causal inference tools such as differences-in-differences \citep{angrist2008mostly, bertrand2004much} and matching estimators \citep{dehejia2002propensity, ho2007matching, rubin1973matching, stuart2010matching}.
 
@@
 It seems that the influence of the PCC on physical violence has been overstated.
-It is unlikely that the PCC -- which is underfunded for its size\footnote{A Parliamentary Commission of Inquiry has stated that the PCC earns about 16 million Brazilian Reals per month, which amounts to approximately 60 million US dollars per year (see: \url{http://goo.gl/FwhPa3} (in Portuguese).
+It is unlikely that the PCC -- which is underfunded for its size\footnote{A Parliamentary Commission of Inquiry has stated that the PCC earns about 16 million Brazilian Reals per month, which amounts to approximately 60 million US dollars per year.
+See: \url{http://goo.gl/FwhPa3} (in Portuguese).
 Access: July 2016.

### 2018-02-08 chapters/01-intro.tex
add R code to synth

@@
 The literature on political and criminal violence has increased exponentially over the last decades.
-Although interstate wars have traditionally occupied a privileged position in this research agenda, scholars have broadened their scope to include a myriad of hitherto understudied phenomena into their research endeavours.
-Civil wars \citep[][]{collier2004greed,fearon2003ethnicity,kalyvas2006logic}, genocides \citep[][]{mamdani2014victims, power2013problem}, ethnic conflicts \citep[][]{kaufmann1996possible, montalvo2005ethnic, sambanis2001ethnic}, wartime sexual abuse \citep[][]{cohen2013explaining,wood2006variation,wood2009armed}, electoral violence \citep[][]{hoglund2009electoral,wilkinson2006votes}, state-sponsored killings \citep[][]{harff1988toward, krain1997state,krain2005international,uzonyi2014unpacking}, terrorism \citep[][]{de2005quality,bueno2007propaganda,pape2003strategic}, drug-related violence \citep[][]{holmes2006drugs,lessing2015logics,richani2013systems, shirk2010drug}, street gangs \citep[][]{franzese2016youth,jones2009youth,rodgers2006living,sobel1987direct}, and prison gangs \citep[][]{dias2011pulverizaccao,freire2014,skarbek2011governance,skarbek2012prison,skarbek2014social} have recently moved to the centre stage of the discipline.
-This dissertation contributes to this growing literature.
+Although interstate wars have traditionally occupied a privileged position in political science, scholars have broadened their scope to include a myriad of hitherto understudied phenomena into their research agendas.
+Civil wars \citep{collier2004greed,fearon2003ethnicity,kalyvas2006logic}, genocides \citep{mamdani2014victims, power2013problem}, ethnic conflicts \citep{kaufmann1996possible, montalvo2005ethnic, sambanis2001ethnic}, wartime sexual abuse \citep{cohen2013explaining,wood2006variation,wood2009armed}, electoral violence \citep{hoglund2009electoral,wilkinson2006votes}, state-sponsored killings \citep{harff1988toward, krain1997state,krain2005international,uzonyi2014unpacking}, terrorism \citep{de2005quality,bueno2007propaganda,pape2003strategic}, drug-related violence \citep{holmes2006drugs,lessing2015logics,richani2013systems, shirk2010drug}, street gangs \citep{franzese2016youth,jones2009youth,rodgers2006living,sobel1987direct}, and prison gangs \citep{dias2011pulverizaccao,freire2014,skarbek2011governance,skarbek2012prison,skarbek2014social} have recently moved from the margins to the centre stage of the discipline.
+The present dissertation contributes to this expanding field.
 
-In order to clarify crucial aspects of my research topic, I employ an eclectic combination of research designs, ranging from statistical analysis to field experiments and qualitative case studies.
-This diversity of methods not only reflects the multiple aspects of organised violence, but it is also a pragmatic response to problems which are common in this area, such as incomplete data, biased information, and the practical impossibility of random assignment of treatment.
-By using an array of methodological tools, I hope to overcome some of these inherent difficulties.
+In order to clarify crucial aspects of my research topic, I employ an eclectic combination of research designs.
+The methods range from qualitative case studies to randomised experiments and machine learning algorithms.
+This diversity not only reflects the multiple aspects of organised violence, but it is also a pragmatic response to problems which are common in this area, such as incomplete data, reporting bias, and model uncertainty.
+By using an array of methodological tools, I hope to overcome some of these challenges.
 
-Regarding the geographical scope of this dissertation, most of the following chapters deal with issues of violence in Latin America, specially in Brazil.
+Regarding the geographical scope of this dissertation, two of the following chapters deal with issues of violence in Latin America, specially in Brazil.
 According to the World Bank, Latin America is home to about 8\% of the global population, yet it accounts for more than 30\% of the world's homicides.\footnote{See: \url{https://goo.gl/d2WC3V}.
 Access: April 2017.} Moreover, the yearly ranking by the Citizen's Council for Public Security and Criminal Justice (Consejo Ciudadano para la Seguridad Pública y la Justicia Penal), a Mexican non-governmental organisation, shows that 43 of the 50 most violent cities in the world are located in Latin America, including all of the top 10.\footnote{For the complete ranking, see \url{http://www.seguridadjusticiaypaz.org.mx/biblioteca/prensa/summary/6-prensa/239-las-50-ciudades-mas-violentas-del-mundo-2016-metodologia}.
-Access: April 2017.} Given the severity of violence in the continent, Latin America was the natural candidate to be the primary focus of this study.
+Access: February 2018.} Given the severity of violence in the continent, Latin America was expected to be an important part of this study.
 

### 2018-02-08 chapters/02-synth.tex
add R code to synth

@@
-\chapter{Evaluating the Effect of Homicide Prevention Strategies in São Paulo, Brazil: A Synthetic Control Approach}
+\chapter{Evaluating the Effect of Homicide Prevention Strategies in São Paulo, Brazil: A Synthetic Control Approach\footnote{Forthcoming in \textit{Latin American Research Review}, 53(2), 2018.
+All data, code, and information required to replicate this study are available at \href{https://github.com/danilofreire/homicides-sp-synth}{https://github.com/danilofreire/homicides-sp-synth}.}}
 \label{chap:synth}
@@
 São Paulo has traditionally occupied a key position in Brazil's violence statistics.
-It is the country's richest and most-densely populated state, and in the 1990s its homicide rate was roughly 50\% higher than the national average \citep[p.
-120]{barata2000}.
+It is the country's richest and most-densely populated state, and in the 1990s its homicide rate was roughly 50\% higher than the national average \citep[120]{barata2000}.
 Some areas of the namesake capital city had even worse numbers.
-Between 1996 and 1999, the ramshackle districts of Jardim São Luiz and Jardim Ângela had respectively 103 and 116 violent deaths per 100,000 residents \citep[p.
-8]{cardia2003}, figures that placed them amongst the deadliest neighbourhoods on the globe \citep{who2015}.
+Between 1996 and 1999, the ramshackle districts of Jardim São Luiz and Jardim Ângela had respectively 103 and 116 violent deaths per 100,000 residents \citep[8]{cardia2003}, figures that placed them amongst the deadliest neighbourhoods on the globe \citep{who2015}.
 
@@
 Deterrence measures have been complemented by investments in police intelligence.
-In 1999, the state administration created a new system for crime prevention, Infocrim \citep[p.
-3]{risso2014intentional} The system gathers geo-coded information on homicides and maps the most important `hot spots' of criminal activity in the state.
-The government has also developed a new photo database, Fotocrim, to speed up the process of facial recognition of criminals \citep[p.
-3]{mello2010}.
+In 1999, the state administration created a new system for crime prevention, Infocrim \citep[3]{risso2014intentional} The system gathers geo-coded information on homicides and maps the most important `hot spots' of criminal activity in the state.
+The government has also developed a new photo database, Fotocrim, to speed up the process of facial recognition of criminals \citep[3]{mello2010}.
 
@@
 There is good evidence that incapacitation measures have worked well in São Paulo during the last decade.
-Gun-related homicides have declined about 74\% from 2001 to 2008 \citep{peres2011queda} at the same time when São Paulo has experienced an increase of 770\% in the arrests of repeated murderers \citep[p.
-36]{manso2012homicidos}.
+Gun-related homicides have declined about 74\% from 2001 to 2008 \citep{peres2011queda} at the same time when São Paulo has experienced an increase of 770\% in the arrests of repeated murderers \citep[36]{manso2012homicidos}.
 Although there are studies that indicate possible `hardening effects' of imprisonment, that is, longer sentences may positively affect an individual's tendency to commit further crimes \citep[e.g.][]{chen2007harsher,glaeser1996crime,western2001labor}, the São Paulo case appears to suggest otherwise.
@@
 At least since the mid-2000s, these authors argue that the PCC has been able to emerge as an undisputed mediator and solve conflicts between inmates.
-\citet[p.
-83]{dias2009ocupando} writes that ``[\dots] when unable to constitute a universal source of regulation, the official law leaves gaps which are filled by informal instances -- such as the Primeiro Comando da Capital (PCC), in the prisons of São Paulo.'' The gang has implemented informal courts that resemble state institutions, and those meetings have progressively replaced other forms of popular justice such as lynchings or the hiring of target killers \citep[p.
-3]{feltran2012metodos}.
+\citet[83]{dias2009ocupando} writes that ``[\dots] when unable to constitute a universal source of regulation, the official law leaves gaps which are filled by informal instances -- such as the Primeiro Comando da Capital (PCC), in the prisons of São Paulo.'' The gang has implemented informal courts that resemble state institutions, and those meetings have progressively replaced other forms of popular justice such as lynchings or the hiring of target killers \citep[3]{feltran2012metodos}.
 Moreover, the \emph{Comando} has developed a series of assertive ways to terrorise inmates.
@@
 The direct effect captures the sensitivity of a dependent variable $Y$ to changes in $X$ when this relationship is not mediated by any other variables in the model.
-Holding all factors constant, the direct effect is a causal chain of length one \citep[p.
-160]{sobel1987direct} and could be described simply as $X \rightarrow Y$.
-In turn, the total effect can be defined as $P(Y_{x} = y)$, that is, ``the probability that response variable $Y$ would take on the value $y$ when $X$ is set to $x$ by external intervention'' \citep[p.
-1572]{pearl2001direct}.
+Holding all factors constant, the direct effect is a causal chain of length one \citep[160]{sobel1987direct} and could be described simply as $X \rightarrow Y$.
+In turn, the total effect can be defined as $P(Y_{x} = y)$, that is, ``the probability that response variable $Y$ would take on the value $y$ when $X$ is set to $x$ by external intervention'' \citep[1572]{pearl2001direct}.
 The total effect is the sum of direct and indirect (or mediated) effects.
@@
 
-The method works as follows.\footnote{Please refer to Appendix~\ref{sec:appendix_a} for a formal presentation of the synthetic control method.} SCM starts with the assumption that one case in the sample has received a treatment.
+The method works as follows.\footnote{Please refer to Appendix~\ref{sec:synth-appendix} for a formal presentation of the synthetic control method.} SCM starts with the assumption that one case in the sample has received a treatment.
 The treatment is defined as a time-delimited event that affects the unit of interest, such as the implementation of a new policy or the outbreak of a conflict.
@@
 The graph features cases in which the mean squared prediction error, a measure of goodness-of-fit, is no higher than twice that of São Paulo.
-That is, only placebos that have good synthetic matches were selected for the analysis \citep[p.
-503]{abadie2010}.
+That is, only placebos that have good synthetic matches were selected for the analysis \citep[503]{abadie2010}.
 In this group, the negative gap for the homicide rate São Paulo is by far the most relevant, providing further evidence for the original findings.
@@
 However, in this model only the time trends of the dependent variable are matched.
-In a sense, this is closer to a traditional differences-in-differences approach, but without the restrictive assumption that the treated and the control cases would follow parallel trends over time \citep[p.
-494]{abadie2010}.
+In a sense, this is closer to a traditional differences-in-differences approach, but without the restrictive assumption that the treated and the control cases would follow parallel trends over time \citep[494]{abadie2010}.
 
@@
 
-\newpage
+\section{Appendix} 
+\label{sec:synth-appendix}
 
-\section{Appendix} 
-\label{sec:appendix_a}
+\subsection{The Synthetic Control Estimator}
+\label{sec:synth-estimator}
 
@@
 Consider $||X_{1} - X_{0}W||v = \sqrt{(X_{1} - X_{0}W)'V (X_{1} - X_{0}W)}$, where $V$ is a $(k \times k)$ symmetric and semi-definite positive matrix with the relative importance of each assigned homicide rate predictor.
-From various possible ways of choosing $V$, in this paper I follow the recommendation of \citet{abadie2003} and choose $V^{*}$ as the value of $V$ that minimises the root mean squared prediction error (RMSPE) for homicide rates in the entire pre-treatment period (1990-1998).
+From various possible ways of choosing $V$, here I follow the recommendation of \citet{abadie2003} and choose $V^{*}$ as the value of $V$ that minimises the root mean squared prediction error (RMSPE) for homicide rates in the entire pre-treatment period (1990-1998).
+
+\subsection{\texttt{R} Code}
+\label{sec:synth-code}
+
+The \texttt{R} code below replicates all statistical analyses and graphs included in this chapter.
+The original data files as well as the final dataset are available at \href{https://github.com/danilofreire/homicides-sp-synth}{https://github.com/danilofreire/homicides-sp-synth}.
+
+\singlespacing
+\small
+\begin{verbatim}
+######################
+### Data Wrangling ###
+######################
+
+# Please set your working directory to the data/ folder
+
+# Clear the workspace
+rm(list = ls())
+
+# Load necessary packages
+library(reshape2)  # data manipulation
+
+# Dependent variable:
+dep <- read.csv("homicide-rates.csv", header = TRUE, skip = 1)
+
+dep.molten <- melt(dep,
+                   id.vars = c("Sigla",
+                               "Código",
+                               "Estado")
+                    )
+
+colnames(dep.molten) <- c("abbreviation",
+                          "code",
+                          "state",
+                          "year",
+                          "homicide.rates")
+
+dep.molten$year <- as.numeric(substring(dep.molten$year, 2))
+\end{verbatim}
+
+\newpage
+
+\begin{verbatim}
+# Independent variables
+ind1 <- read.csv("state-gdp-capita.csv", header = TRUE, skip = 1)
+
+ind1.molten <- melt(ind1,
+                    id.vars = c("Sigla",
+                                "Código",
+                                "Estado")
+                    )
+
+colnames(ind1.molten) <- c("abbreviation",
+                           "code",
+                           "state",
+                           "year",
+                           "state.gdp.capita")
+
+ind1.molten$year <- as.numeric(substring(ind1.molten$year, 2))
+
+ind2 <- read.csv("state-gdp-growth-percentage.csv", header = TRUE, skip = 1)
+
+ind2.molten <- melt(ind2,
+                    id.vars = c("Sigla",
+                                "Código",
+                                "Estado")
+                    )
+
+colnames(ind2.molten) <- c("abbreviation",
+                           "code",
+                           "state",
+                           "year",
+                           "state.gdp.growth.percent")
+
+ind2.molten$year <- as.numeric(substring(ind2.molten$year, 2))
+
+ind3 <- read.csv("gini.csv", header = TRUE, skip = 1)
+
+ind3.molten <- melt(ind3,
+                    id.vars = c("Sigla",
+                                "Código",
+                                "Estado")
+                    )
+
+colnames(ind3.molten) <- c("abbreviation",
+                           "code",
+                           "state",
+                           "year",
+                           "gini")
+
+ind3.molten$year <- as.numeric(substring(ind3.molten$year, 2))
+
+ind4 <- read.csv("population-projection.csv",
+                 header = TRUE,
+                 skip   = 1)
+
+ind4.molten <- melt(ind4,
+                    id.vars = c("Sigla",
+                                "Código",
+                                "Estado")
+                    )
+
+colnames(ind4.molten) <- c("abbreviation",
+                           "code",
+                           "state",
+                           "year",
+                           "population.projection")
+
+ind4.molten$year <- as.numeric(substring(ind4.molten$year, 2))
+
+ind5 <- read.csv("population-extreme-poverty.csv", header = TRUE, skip = 1)
+
+ind5.molten <- melt(ind5,
+                    id.vars = c("Sigla",
+                                "Código",
+                                "Estado")
+                    )
+
+colnames(ind5.molten) <- c("abbreviation",
+                           "code",
+                           "state",
+                           "year",
+                           "population.extreme.poverty")
+
+ind5.molten$year <- as.numeric(substring(ind5.molten$year, 2))
+
+ind6 <- read.csv("years-schooling.csv", header = TRUE, skip = 1)
+
+ind6.molten <- melt(ind6,
+                    id.vars = c("Sigla",
+                                "Código",
+                                "Estado")
+                    )
+
+colnames(ind6.molten) <- c("abbreviation",
+                           "code",
+                           "state",
+                           "year",
+                           "years.schooling")
+
+ind6.molten$year <- as.numeric(substring(ind6.molten$year, 2))
+
+# Merges files
+data.list <- list(dep.molten,
+                  ind1.molten,
+                  ind2.molten,
+                  ind3.molten,
+                  ind4.molten,
+                  ind5.molten,
+                  ind6.molten)
+
+data1 <- Reduce(function(...) merge(..., all = TRUE), data.list)
+
+# Subset and sort
+data2 <- subset(data1, year >= 1990 & year <= 2009)
+data2 <- data2[order(data2$state), ]
+rownames(data2) <- NULL
+
+# Count missing observations, calculate their percentage
+round(sapply(data2, function(x) length(which(is.na(x)))), 2)
+round(sapply(data2, function(x) length(which(is.na(x)))/length(x)), 2)
+
+# Linear imputation of missing values.
+data2$gini.imp <- approxfun(seq_along(data2$gini), data2$gini)(seq_along(data2$gini))
+
+data2$population.extreme.poverty.imp <- approxfun(seq_along(data2$population.extreme.poverty),
+data2$population.extreme.poverty)(seq_along(data2$population.extreme.poverty))
+
+data2$years.schooling.imp <- approxfun(seq_along(data2$years.schooling),
+data2$years.schooling)(seq_along(data2$years.schooling))
+
+# Create proportion.extreme.poverty
+data2$proportion.extreme.poverty <- data2$population.extreme.poverty.imp / data2$population.projection
+
+# Transform variables to improve interpretation
+data2$population.projection.ln <- log(data2$population.projection)
+
+# Save data as df.csv
+write.table(data2,
+            "df.csv",
+            row.names = FALSE,
+            col.names = TRUE,
+            sep       = ",")
+
+#####################
+### Data Analysis ###
+#####################
+
+# Load necessary packages
+library(dplyr) # data manipulation
+library(Synth) # models
+
+# Load data
+df <- read.csv("df.csv", header = TRUE)
+
+# Prepare dataset
+df$state <- as.character(df$state) # required by dataprep()
+
+# Plot: Homicide rates for Sao Paulo and Brazil (average)
+df1 <- df %>%
+        mutate(homicide.sp = ifelse(homicide.rates & state == "São Paulo", homicide.rates, NA)) %>%
+        select(year, homicide.sp)
+
+df2 <- df %>%
+        mutate(homicide.rates1 = ifelse(homicide.rates & state != "São Paulo", homicide.rates, NA)) %>%
+        group_by(year) %>%
+        summarise(homicide.br = mean(homicide.rates1, na.rm = TRUE))
+
+setEPS()
+postscript(file    = "br.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,     # 17.8 cm
+           height  = 5.25)  # 13.3 cm
+
+plot(x = df1$year,
+     y = df1$homicide.sp,
+     type = "l",
+     ylim = c(0, 60),
+     xlim = c(1990, 2009),
+     xlab = "Year",
+     ylab = "Homicide Rates",
+     cex = 3,
+     lwd = 2,
+     xaxs = "i",
+     yaxs = "i"
+)
+
+lines(df2$year,
+      df2$homicide.br,
+      lty = 2,
+      cex = 3,
+      lwd = 2)
+
+arrows(1997, 50, 1999, 50,
+       col    = "black",
+       length = .1)
+
+text(1995, 50,
+     "Policy Change",
+     cex = .8)
+
+abline(v   = 1999,
+       lty = 2)
+
+legend(x = "bottomleft",
+       legend = c("São Paulo",
+                  "Brazil (average)"),
+       lty    = c("solid", "dashed"),
+       cex    = .8,
+       bg     = "white",
+       lwdc(2, 2)
+)
+
+invisible(dev.off())
+
+# Prepare data for synth
+dataprep.out <-
+        dataprep(df,
+                 predictors = c("state.gdp.capita",
+                                "state.gdp.growth.percent",
+                                "population.projection.ln",
+                                "years.schooling.imp"
+                                ),
+                 special.predictors = list(
+                         list("homicide.rates", 1990:1998, "mean"),
+                         list("proportion.extreme.poverty", 1990:1998, "mean"),
+                         list("gini.imp", 1990:1998, "mean")
+                         ),
+                 predictors.op = "mean",
+                 dependent     = "homicide.rates",
+                 unit.variable = "code",
+                 time.variable = "year",
+                 unit.names.variable   = "state",
+                 treatment.identifier  = 35,
+                 controls.identifier   = c(11:17, 21:27, 31:33, 41:43, 50:53),
+                 time.predictors.prior = c(1990:1998),
+                 time.optimize.ssr     = c(1990:1998),
+                 time.plot             = c(1990:2009)
+                 )
+
+# Run synth
+synth.out <- synth(dataprep.out)
+
+# Get result tables
+print(synth.tables   <- synth.tab(
+        dataprep.res = dataprep.out,
+        synth.res    = synth.out)
+      )
+
+# Plot: Main model
+setEPS()
+postscript(file    = "trends.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,     # 17.8 cm
+           height  = 5.25)  # 13.3 cm
+
+path.plot(synth.res    = synth.out,
+          dataprep.res = dataprep.out,
+          Ylab         = c("Homicide Rates"),
+          Xlab         = c("Year"),
+          Legend       = c("São Paulo","Synthetic São Paulo"),
+          Legend.position = c("bottomleft")
+)
+
+abline(v   = 1999,
+       lty = 2)
+
+arrows(1997, 50, 1999, 50,
+       col    = "black",
+       length = .1)
+
+text(1995, 50,
+     "Policy Change",
+     cex = .8)
+
+invisible(dev.off())
+
+# Main model: gaps plot
+setEPS()
+postscript(file    = "gaps.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,
+           height  = 5.25)
+
+gaps.plot(synth.res    = synth.out,
+          dataprep.res = dataprep.out,
+          Ylab         = c("Gap in Homicide Rates"),
+          Xlab         = c("Year"),
+          Ylim         = c(-30, 30),
+          Main         = ""
+)
+
+abline(v   = 1999,
+       lty = 2)
+
+arrows(1997, 20, 1999, 20,
+       col    = "black",
+       length = .1)
+
+text(1995, 20,
+     "Policy Change",
+     cex = .8)
+
+invisible(dev.off())
+
+## Calculating how many lives were saved during the treatment period
+
+# Weights below retrieved form dataprep.out
+# State Code  State Weight  State Name        State Abbreviation
+# 42          0.274         Santa Catarina    SC
+# 53          0.210         Distrito Federal  DF
+# 32          0.209         Espirito Santo    ES
+# 33          0.169         Rio de Janeiro    RJ
+# 14          0.137         Roraima           RR
+# 14          0.001         Pernambuco        PB
+# 35          treat         Sao Paulo         SP
+\end{verbatim}
+
+\begin{verbatim}
+# Get years after policy change
+df.2 <- df[which(df$year >= 1999),]
+
+# Calculate total number of deaths in SP
+num.deaths.sp <- sum( (df.2$homicide.rates[which(df.2$abbreviation == "SP")])/100000 *
+    (df.2$population.projection[which(df.2$abbreviation == "SP")]))
+
+# Calculate estimated number of deaths in Synthetic São Paulo
+num.deaths.synthetic.sp <- sum( (0.274 * (df.2$homicide.rates[which(df.2$abbreviation == "SC")])/100000 *
+(df.2$population.projection[which(df.2$abbreviation == "SP")]))
+                                + (0.210 * (df.2$homicide.rates[which(df.2$abbreviation == "DF")])/100000 *
+                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
+                                + (0.209 * (df.2$homicide.rates[which(df.2$abbreviation == "ES")])/100000 *
+                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
+                                + (0.169 * (df.2$homicide.rates[which(df.2$abbreviation == "RJ")])/100000 *
+                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
+                                + (0.137 * (df.2$homicide.rates[which(df.2$abbreviation == "RR")])/100000 *
+                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
+                                + (0.001 * (df.2$homicide.rates[which(df.2$abbreviation == "PB")])/100000 *
+                                (df.2$population.projection[which(df.2$abbreviation == "SP")]))
+                                )
+
+lives.saved <- num.deaths.synthetic.sp - num.deaths.sp
+lives.saved # Between 1999 and 2009
+
+########################
+### Robustness Tests ###
+########################
+
+# Prepare dataset
+df$state <- as.character(df$state) # required by dataprep()
+
+## Placebo Test -- Control ends in 1994
+dataprep.out1 <-
+        dataprep(df,
+                 predictors = c("state.gdp.capita",
+                                "state.gdp.growth.percent",
+                                "population.projection.ln",
+                                "years.schooling.imp"
+                 ),
+                 special.predictors = list(
+                         list("homicide.rates", 1990:1994, "mean"),
+                         list("proportion.extreme.poverty", 1990:1994, "mean"),
+                         list("gini.imp", 1990:1994, "mean")
+                 ),
+                 predictors.op = "mean",
+                 dependent     = "homicide.rates",
+                 unit.variable = "code",
+                 time.variable = "year",
+                 unit.names.variable   = "state",
+                 treatment.identifier  = 35,
+                 controls.identifier   = c(11:17, 21:27, 31:33, 41:43, 50:53),
+                 time.predictors.prior = c(1990:1994),
+                 time.optimize.ssr     = c(1990:1994),
+                 time.plot             = c(1990:1998))
+
+# Run synth
+synth.out1 <- synth(dataprep.out1)
+
+# Get result tables
+print(synth.tables   <- synth.tab(
+        dataprep.res = dataprep.out1,
+        synth.res    = synth.out1)
+      )
+
+# Placebo test: graph
+setEPS()
+postscript(file    = "placebo.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,
+           height  = 5.25)
+
+path.plot(synth.res       = synth.out1,
+          dataprep.res    = dataprep.out1,
+          Ylab            = c("Homicide Rates"),
+          Xlab            = c("Year"),
+          Legend          = c("São Paulo","Synthetic São Paulo"),
+          Legend.position = c("bottomleft"),
+          Ylim            = c(0, 50)
+)
+
+abline(v   = 1995,
+       lty = 2)
+
+arrows(1994, 40, 1995, 40,
+       col    = "black",
+       length = .1)
+
+text(1993, 40,
+     "Placebo \nPolicy Change",
+     cex = .8)
+
+invisible(dev.off())
+
+## Leave-one-out
+
+# Loop over leave one outs
+storegaps <- matrix(NA, length(1990:2009), 4)
+
+colnames(storegaps) <- c(14, 33, 42, 53) # RR, RJ, SC, DF
+co <- unique(df$code)
+co <- co[-25]
+
+for(k in 1:4){
+
+        # Data prep for training model
+        omit <- c(14, 33, 42, 53)[k]
+
+        # Prepare data for synth
+        dataprep.out2 <-
+                dataprep(df,
+                         predictors = c("state.gdp.capita",
+                                        "state.gdp.growth.percent",
+                                        "population.projection.ln",
+                                        "years.schooling.imp"
+                         ),
+                         special.predictors = list(
+                                 list("homicide.rates", 1990:1998, "mean"),
+                                 list("proportion.extreme.poverty", 1990:1998, "mean"),
+                                 list("gini.imp", 1990:1998, "mean")
+                         ),
+                         predictors.op = "mean",
+                         dependent     = "homicide.rates",
+                         unit.variable = "code",
+                         time.variable = "year",
+                         unit.names.variable   = "state",
+                         treatment.identifier  = 35,
+                         controls.identifier   = co[-which(co==omit)],
+                         time.predictors.prior = c(1990:1998),
+                         time.optimize.ssr     = c(1990:1998),
+                         time.plot             = c(1990:2009)
+                )
+
+        # Run synth
+        synth.out2 <- synth(dataprep.out2)
+
+        storegaps[,k] <- (dataprep.out2$Y0%*%synth.out2$solution.w)
+} # Close loop over leave one outs
+
+# Leave-one-out: graph
+setEPS()
+postscript(file    = "leave-one-out.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,
+           height  = 5.25)
+
+path.plot(synth.res    = synth.out,
+          dataprep.res = dataprep.out,
+          Ylab         = c("Homicide Rates"),
+          Xlab         = c("Year"),
+          Legend       = c("São Paulo","Synthetic São Paulo"),
+          Legend.position = c("bottomleft"))
+
+abline(v   = 1999,
+       lty = 2)
+
+arrows(1997, 50, 1999, 50,
+       col    = "black",
+       length = .1)
+
+text(1995, 50,
+     "Policy Change",
+     cex = .8)
+
+for(i in 1:4){
+        lines(1990:2009,
+              storegaps[,i],
+              col = "darkgrey",
+              lty = "solid")
+}
+
+lines(1990:2009,
+      dataprep.out$Y0plot %*% synth.out$solution.w,
+      col = "black",
+      lty = "dashed",
+      lwd = 2)
+
+legend(x = "bottomleft",
+       legend = c("São Paulo",
+                  "Synthetic São Paulo",
+                  "Synthetic São Paulo (leave-one-out)"
+       ),
+       lty    = c("solid", "dashed", "solid"),
+       col    = c("black", "black", "darkgrey"),
+       cex    = .8,
+       bg     = "white",
+       lwdc(2, 2, 1)
+)
+
+invisible(dev.off())
+
+## Permutation test
+states <- c(11:17, 21:27, 31:33, 35, 41:43, 50:53)
+
+# Prepare data for synth
+results <- list()
+results_synth <- list()
+gaps <- list()
+
+for (i in states) {
+    dataprep.out <-
+            dataprep(df,
+                     predictors = c("state.gdp.capita",
+                                    "state.gdp.growth.percent",
+                                    "population.projection.ln",
+                                    "years.schooling.imp"
+                                    ),
+                     special.predictors = list(
+                             list("homicide.rates", 1990:1998, "mean"),
+                             list("proportion.extreme.poverty", 1990:1998, "mean"),
+                             list("gini.imp", 1990:1998, "mean")
+                             ),
+                     predictors.op = "mean",
+                     dependent     = "homicide.rates",
+                     unit.variable = "code",
+                     time.variable = "year",
+                     unit.names.variable   = "state",
+                     treatment.identifier  = i,
+                     controls.identifier   = states[which(states!=i)],
+                     time.predictors.prior = c(1990:1998),
+                     time.optimize.ssr     = c(1990:1998),
+                     time.plot             = c(1990:2009)
+                     )
+    results[[as.character(i)]] <- dataprep.out
+    results_synth[[as.character(i)]] <- synth(results[[as.character(i)]])
+    gaps[[as.character(i)]] <- results[[as.character(i)]]$Y1plot - (results[[as.character(i)]]$Y0plot %*%
+    results_synth[[as.character(i)]]$solution.w)
+
+}
+
+## Permutation test
+setEPS()
+postscript(file    = "permutation-gaps2.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,
+           height  = 5.25)
+
+plot(1990:2009,
+     ylim = c(-30, 30),
+     xlim = c(1990,2009),
+     ylab = "Gap in Homicide Rates",
+     xlab = "Year"
+)
+
+for (i in states) {
+        lines(1990:2009,
+              gaps[[as.character(i)]],
+              col = "lightgrey",
+              lty = "solid",
+              lwd = 2
+        )
+}
+
+lines(1990:2009,
+      gaps[["35"]], # São Paulo
+      col = "black",
+      lty = "solid",
+      lwd = 2
+)
+
+abline(v   = 1999,
+       lty = 2)
+\end{verbatim}
+
+\newpage
+
+\begin{verbatim}
+abline(h   = 0,
+       lty = 1,
+       lwd = 1)
+
+arrows(1997, 25, 1999, 25,
+       col    = "black",
+       length = .1)
+
+text(1995, 25,
+     "Policy Change",
+     cex = .8)
+
+legend(x = "bottomleft",
+       legend = c("São Paulo",
+                  "Control States"),
+       lty  = c("solid", "solid"),
+       col  = c("black", "darkgrey"),
+       cex  = .8,
+       bg  = "white",
+       lwdc(2, 2, 1)
+)
+
+invisible(dev.off())
+
+# Permutation graph: states with MSPE no higher than 2x São Paulo's
+low.mspe <- c(13, 15, 17, 21, 23, 24, 25, 31, 41:43, 53)
+
+setEPS()
+postscript(file    = "low-mspe.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,
+           height  = 5.25)
+
+plot(1990:2009,
+     ylim = c(-30, 30),
+     xlim = c(1990,2009),
+     ylab = "Gap in Homicide Rates",
+     xlab = "Year"
+)
+
+for (i in low.mspe) {
+lines(1990:2009,
+      gaps[[as.character(i)]],
+      col = "lightgrey",
+      lty = "solid",
+      lwd = 2
+      )
+}
+
+lines(1990:2009,
+      gaps[["35"]], # São Paulo
+      col = "black",
+      lty = "solid",
+      lwd = 2
+      )
+
+abline(v   = 1999,
+       lty = 2)
+
+abline(h   = 0,
+       lty = 1,
+       lwd = 1)
+
+arrows(1997, 25, 1999, 25,
+       col    = "black",
+       length = .1)
+
+text(1995, 25,
+     "Policy Change",
+     cex = .8)
+
+legend(x = "bottomleft",
+       legend = c("São Paulo",
+                  "Control States (MSPE Less Than Two Times That of São Paulo)"),
+       lty    = c("solid", "solid"),
+       col    = c("black", "darkgrey"),
+       cex    = .8,
+       bg     = "white",
+       lwdc(2, 2, 1)
+)
+
+invisible(dev.off())
+
+## CausalImpact
+# Uncomment the lines below to install necessary packages
+# install.packages(c("devtools", "dtw"))
+# library(devtools)
+# install_github("google/CausalImpact")
+# install_github("klarsen1/MarketMatching", build_vignettes=TRUE)
+
+# Load packages
+library(CausalImpact)
+library(MarketMatching)
+
+# Prepare data
+df$year2 <- as.Date(paste(df$year, sep = "", "-01-01"))
+
+# Estimate model
+mm <- best_matches(data=df,
+                   id_variable="code",
+                   date_variable="year2",
+                   matching_variable="homicide.rates",
+                   parallel=TRUE,
+                   warping_limit=1, # warping limit=1
+                   dtw_emphasis=1, # rely only on dtw for pre-screening
+                   matches=5, # request 5 matches
+                   start_match_period="1990-01-01",
+                   end_match_period="1998-01-01")
+
+# View best matches
+subset(mm$BestMatches, code == 35) # SP
+
+# Results
+results <- MarketMatching::inference(matched_markets = mm,
+                                     test_market = "35",
+                                     end_post_period = "2009-01-01")
+
+# Predictions
+results$Predictions
+
+# Plot results
+results$PlotActualVersusExpected +
+        ggtitle("São Paulo versus Synthetic São Paulo") + theme_bw() +
+        geom_line(aes(results$PlotActualVersusExpected$data$test_market),colour="#000099")
+results$PlotCumulativeEffect
+
+# Graph
+setEPS()
+postscript(file    = "causal-impact.eps",
+           horiz   = FALSE,
+           onefile = FALSE,
+           width   = 7,     # 17.8 cm
+           height  = 5.25)  # 13.3 cm
+
+plot(x = (1990:2009),
+     y = as.numeric(results$Predictions$Response),
+     type = "l",
+     ylim = c(0, 60),
+     xlim = c(1990, 2009),
+     xlab = "Year",
+     ylab = "Homicide Rates",
+     cex = 3,
+     lwd = 2)
+
+lines(x = (1990:2009),
+      y = as.numeric(results$Predictions$Predicted),
+      type = "l",
+      lty = 2,
+      cex = 3,
+      lwd = 2)
+
+arrows(1997, 50, 1999, 50,
+       col    = "black",
+       length = .1)
+
+text(1995, 50,
+     "Policy Change",
+     cex = .8)
+
+abline(v   = 1999,
+       lty = 2)
+
+legend(x = "bottomleft",
+       legend = c("São Paulo",
+                  "Brazil (average)"),
+       lty    = c("solid", "dashed"),
+       cex    = .8,
+       bg     = "white",
+       lwdc(2, 2)
+)
+
+invisible(dev.off())
+\end{verbatim}
+
+\doublespacing
+\normalsize

### 2018-02-08 dissertation.tex
add R code to synth

--- 2017-04-10 thesis.tex
@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% DOCUMENT CLASS
 \documentclass[a4paper,12pt]{report}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% University of Sussex thesis template
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Modification History
-%
-% Based on usthesis.cls by Jonathon Read
-% http://www.cogs.susx.ac.uk/users/jlr24/latex.html
-% Modified by Anthony Smith, Feb 2007
-% Incorporated into single thesis.tex file, Anthony Smith, 30 June 2008
-% Minor alterations to page numbering, AJS, 25 July 2008
-% New alternative hyperref options for print version, AJS, 11 Sep 2008
-% "DRAFT" on header, AJS, 12 Sep 2008
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \usepackage[authoryear]{natbib}
-% \bibliographystyle{plain} for [1], [2] etc.
 \bibliographystyle{apalike}
+
+% REMOVE COMMA
+\setcitestyle{aysep={}}
+
+% HYPERLINK YEAR
+\usepackage{etoolbox}
+
+\makeatletter
+
+% Patch case where name and year are separated by aysep
+\patchcmd{\NAT@citex}
+  {\@citea\NAT@hyper@{%
+     \NAT@nmfmt{\NAT@nm}%
+     \hyper@natlinkbreak{\NAT@aysep\NAT@spacechar}{\@citeb\@extra@b@citeb}%
+     \NAT@date}}
+  {\@citea\NAT@nmfmt{\NAT@nm}%
+   \NAT@aysep\NAT@spacechar\NAT@hyper@{\NAT@date}}{}{}
+
+% Patch case where name and year are separated by opening bracket
+\patchcmd{\NAT@citex}
+  {\@citea\NAT@hyper@{%
+     \NAT@nmfmt{\NAT@nm}%
+     \hyper@natlinkbreak{\NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi}%
+       {\@citeb\@extra@b@citeb}%
+     \NAT@date}}
+  {\@citea\NAT@nmfmt{\NAT@nm}%
+   \NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi\NAT@hyper@{\NAT@date}}
+  {}{}
+
+\makeatother
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % OTHER FORMATTING/LAYOUT DECLARATIONS
-% Graphics
 \usepackage[utf8]{inputenc}
@@
 \usepackage[libertine]{newtxmath}
-\usepackage[scaled=0.95]{inconsolata}
-\usepackage{xcolor}
+\usepackage[scaled=.95]{inconsolata}
+\usepackage[stable]{footmisc}
+\usepackage[dvipsnames]{xcolor}
 \definecolor{darkblue}{rgb}{0.0,0.0,0.55}
@@
 \clubpenalty=10000
-%\flushbottom
-% Pages should be numbered consecutively through the main text.  Page numbers
-% should be located centrally at the top of the page.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \renewcommand{\backreflastsep}{ and~}
-%\urlstyle{same}  % don't use monospace font for urls
+\urlstyle{same}  % don't use monospace font for urls
 
 \hypersetup{pdftitle={Essays in Political and Criminal Violence},
-	pdfauthor={Danilo Freire},
+	pdfauthor={Danilo Alves M. Freire},
 	pdfborder={0 0 0},
@@
 	linkcolor=black,
-	citecolor=darkblue,
+	citecolor=Mahogany,
 	urlcolor=darkblue,
 	colorlinks=true}
-% For print version, use this instead:
-%\usepackage[pdfusetitle,bookmarksnumbered,plainpages=false]{hyperref}
-%\usepackage{backref}
-%\renewcommand{\backrefpagesname}{Cited on}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\includegraphics[width=6cm]{kcl.eps}
+\includegraphics[width=6cm]{images/kcl.eps}
 \end{center}	
@@
 \begin{center}
+
 % TITLE
@@
 \vskip5mm
+
 % AUTHOR
-\Large Danilo Freire
+\Large Danilo Alves M. Freire
 \normalsize
@@
 \large
+
 % QUALIFICATION
@@
 King's College London	\\
+
 % DATE OF SUBMISSION
-April 2017
+May 2018
 \end{center}	
@@
 % ADDITIONAL DECLARATIONS HERE (IF ANY)
-
 \vskip20mm
@@
 Signature:
-\vskip20mm
+\vskip10mm
+\begin{flushleft}
+\includegraphics[scale=.25]{images/sig.pdf}
+\end{flushleft}
 % AUTHOR
 \noindent 
-Danilo Freire
+Danilo Alves M. Freire
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \small\normalsize
+
 % SUMMARY HERE (300 word limit for most subjects):
@@
 
-The third essay, co-authored with Gary Uzonyi, presents a meta-analysis of the determinants of government-sponsored violence.
-We employ traditional meta-analysis estimators and Bayesian meta-regressions with random and fixed effects to a set of 48 articles published in the last 20 years to evaluate the main predictors of mass killing onset.
+In the third essay, co-authored with Gary Uzonyi, we employ extreme bounds analysis and two machine learning methods -- distributed random forests and gradient boosting machines -- to identify the key determinants of state-sponsored violence.
+We test the sensitivity of 35 variables on a sample of 177 countries from 1945 to 2013.
+Our results show that GDP per capita and the post-Cold War period are negatively associated with mass killings, whereas previous political turmoils increase the risk of state-led violence.
+Years since the last episode of mass violence, GDP per capita, urban population, openness to trade, and the number of military personnel make the greatest contribution to our models' out-of-sample predictive power.
 
@@
 \small\normalsize
+
 % ACKNOWLEDGEMENTS HERE:
+Testing
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MAIN THESIS TEXT: arabic page numbering 1, 2, 3, ...
+% MAIN THESIS TEXT: Arabic page numbering 1, 2, 3, ...
 \newpage
@@
 
-\input{01-intro}
+\input{chapters/01-intro}
 
@@
 
-\input{02-synth}
+\input{chapters/02-synth}
 
 %-----------------------------------------------------
-% Chapter 3: Jogo do bicho
+% Chapter 3: Jogo do Bicho
 %-----------------------------------------------------
 
-\input{03-bicho}
+\input{chapters/03-bicho}
 
@@
 \addcontentsline{toc}{chapter}{Bibliography}
-\bibliography{bib}
+\bibliography{references/bibliography}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%-----------------------------------------------------
-% Appendix: Code
-%-----------------------------------------------------
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-02-14 chapters/02-synth.tex
Edit jogo do bicho chapter

@@
 The \texttt{R} code below replicates all statistical analyses and graphs included in this chapter.
-The original data files as well as the final dataset are available at \href{https://github.com/danilofreire/homicides-sp-synth}{https://github.com/danilofreire/homicides-sp-synth}.
+The original data files as well as the final dataset are available at \url{https://github.com/danilofreire/homicides-sp-synth}.
 

### 2018-02-14 chapters/03-bicho.tex
add R code to synth
Edit jogo do bicho chapter

--- 2017-04-09 03-bicho.tex
@@
 \section{Introduction}
-\label{sec:intro}
+\label{sec:intro3}
 
@@
 
-\begin{figure}[!htbp]
-	\centering
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho01.jpg}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho02.jpg}
-	\end{minipage}
-	\caption{Left: cartoon of the Baron of Drummond and the animals of the \emph{jogo do bicho} (1896).
+% \begin{figure}[!htbp]
+%  \centering
+%  \begin{minipage}[b]{0.45\textwidth}
+%   \includegraphics[width=\textwidth, height=6cm]{images/bicho01.jpg}
+%  \end{minipage}
+%  \hfill
+%  \begin{minipage}[b]{0.45\textwidth}
+%   \includegraphics[width=\textwidth, height=6cm]{images/bicho02.jpg}
+%  \end{minipage}
+%  \caption{Left: cartoon of the Baron of Drummond and the animals of the \emph{jogo do bicho} (1896).
 Right: entry ticket to Rio's Zoological Garden that allowed the bearer to join the raffle.
@@
 Reproduced in \citet[35--36]{chazkel2011laws}.}
-	\label{fig:barao}
-\end{figure}
+%  \label{fig:barao}
+% \end{figure}
 
@@
 This patron-client relationship has been proven effective: in 2016, the Carnival generated about USD 900 million in revenue and the Rio de Janeiro state received more than one million tourists.\footnote{Data provided by the Brazilian government.
-See \url{https://goo.gl/XMcbTM} (in Portuguese).
+See \href{https://goo.gl/XMcbTM}{https://goo.gl/XMcbTM} (in Portuguese).
 Access: December 2016.}
@@
 \section{\emph{Jogo do Bicho} as an Emergent Institution}
-\label{sec:organisation}
+\label{sec:institution}
 
@@
   % edges %
-		\draw[->, line width = 1.2] (p) -- (y);
+  \draw[->, line width = 1] (p) -- (y);
   \draw[->, line width = 1] (s) -- (y);
   \draw[->, line width = 1] (i) -- (s);
-		\draw[->, line width = 1.2] (l) -- (y);
+  \draw[->, line width = 1] (l) -- (y);
   \draw[->, line width = 1] (m) -- (y);
-		\draw[->, line width = 1.2] (i) -- (p);
+  \draw[->, line width = 1] (i) -- (p);
   \draw[->, dashed, line width = 1] (u) -- (p);
@@
 
-\begin{figure}[!htbp]
-	\centering
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho03.jpg}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho04.jpg}
-	\end{minipage}
-	\caption{\emph{Jogo do bicho} betting sites in 1917 and in 2011.
+% \begin{figure}[!htbp]
+%  \centering
+%  \begin{minipage}[b]{0.45\textwidth}
+%   \includegraphics[width=\textwidth, height=6cm]{images/bicho03.jpg}
+%  \end{minipage}
+%  \hfill
+%  \begin{minipage}[b]{0.45\textwidth}
+%   \includegraphics[width=\textwidth, height=6cm]{images/bicho04.jpg}
+%  \end{minipage}
+%  \caption{\emph{Jogo do bicho} betting sites in 1917 and in 2011.
 The picture on the right shows a \emph{bicheiro}, a street-corner vendor.
 Sources: \citet{alecrim2012bicho} and \citet{ferrarini2011bicho}.}
-	\label{fig:banca1917}
-\end{figure}
+%  \label{fig:banca1917}
+% \end{figure}
 
@@
 As early as in 1890, Brazilian public authorities positioned themselves against the \emph{jogo do bicho} arguing that `{[}\ldots{}{]} this type of amusement is prejudicial to the interests of the unwise, who are naively seduced by the deceptive hope of uncertain lucre' \citep[544]{chazkel2007beyond}.
-In 1941, the government banned the animal game;\footnote{See: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm} (in Portuguese).
+In 1941, the government banned the animal game;\footnote{See: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm}{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del3688.htm} (in Portuguese).
 Access: December 2016.} five years later, it prohibited all games of chance.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.
-The text can be read at: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm} (in Portuguese).
+The text can be read at: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm} (in Portuguese).
 Access: December 2016.} The \emph{jogo do bicho}, casinos and bingos remain illegal in the country.
@@
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=\textwidth, height=6cm]{images/bicho06.jpg}
-	\caption{Results of a \emph{jogo do bicho} draw from 09 January 2016.
+% \begin{figure}[!htbp]
+%  \centering
+%  \includegraphics[width=\textwidth, height=6cm]{images/bicho06.jpg}
+%  \caption{Results of a \emph{jogo do bicho} draw from 09 January 2016.
 `Federal' means that the winning numbers were drawn by the federal government lottery.
@@
 Source: Unknown.
-Available at: \url{https://goo.gl/6PHV8u}.
+Available at: \href{https://goo.gl/6PHV8u}{https://goo.gl/6PHV8u}.
 Access: December 2016.}
-	\label{fig:federal}
-\end{figure}
+%  \label{fig:federal}
+% \end{figure}
 
@@
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=\textwidth, height=6cm]{images/bicho05.jpg}
-	\caption{\emph{Jogo do bicho} results are fixed on light poles in Rio de Janeiro. Source: \citet{gomes1998bicho}.}
-	\label{fig:poste}
-\end{figure}
+% \begin{figure}[!htbp]
+%  \centering
+%  \includegraphics[width=\textwidth, height=6cm]{images/bicho05.jpg}
+%  \caption{\emph{Jogo do bicho} results are fixed on light poles in Rio de Janeiro. Source: \citet{gomes1998bicho}.}
+%  \label{fig:poste}
+% \end{figure}
 
@@
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=.8\textwidth, height=8cm]{images/bicho07.jpg}
-	\caption{Castor de Andrade is shown celebrating after the samba school he sponsored, Mocidade Independente, won the Carnival parade in 1996.
+% \begin{figure}[!htbp]
+%  \centering
+%  \includegraphics[width=.8\textwidth, height=8cm]{images/bicho07.jpg}
+%  \caption{Castor de Andrade is shown celebrating after the samba school he sponsored, Mocidade Independente, won the Carnival parade in 1996.
 He was surrounded by colleagues and police officers.
@@
 Reproduced in \citet[139]{misse2007illegal}.}
-	\label{fig:castor}
-\end{figure}
+%  \label{fig:castor}
+% \end{figure}
 
@@
 \subsection{Political Support}
+\label{sub:support}
 
@@
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
-	\caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right).
+% \begin{figure}[!htbp]
+%  \centering
+%  \includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
+%  \caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right).
 The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'.
 Source: \citet{extra2012aniz}.}
-	\label{fig:aniz}
-\end{figure}
+%  \label{fig:aniz}
+% \end{figure}
 

### 2018-03-25 dissertation.tex
Edit jogo do bicho chapter
add acknowledgements

@@
 \usepackage[libertine]{newtxmath}
-\usepackage[scaled=.95]{inconsolata}
+\usepackage{inconsolata}
 \usepackage[stable]{footmisc}
@@
 
-The fourth essay, co-authored with Umberto Mignozzetti, examines the specific mechanisms by which the \textit{Primeiro Comando da Capital} (PCC), a Brazilian prison gang,  has helped reduce violence in São Paulo.
-We address this gap by proposing a survey experiment to evaluate three competing mechanisms that might have plausibly caused the drop in homicides, namely economic welfare, protection against police harassment, and the local provision of law and order.Our main hypothesis is that the PCC proposed a cohesive system of law and order by informally securing property rights and solving conflicts through credible threats and a local monopoly of the use of force.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % ACKNOWLEDGEMENTS HERE:
-Testing
+I would like to extend thanks to the many people, in many countries, who so generously contributed to the work presented in this dissertation.
+First and foremost, I want to thank my supervisor, Professor David Skarbek.
+It was a great privilege and honour to work and study under his guidance, and he has taught me more than I could ever give him credit for here.
+David was not only an excellent supervisor but also a great friend; his empathy, enthusiasm, and sense of humour made my stay at King's and in London a very pleasant one.
+I simply could not have wished for a better mentor for my PhD studies.
+Thank you very much.
+
+Similarly, profound gratitude goes to Professor Gabriel Leon, who gave me an incredible amount of support and guidance in all stages of this project.
+His help in the last semester of my PhD studies was invaluable and I am very thankful for his generosity.
+
+The Department of Political Economy at King's College London has provided me with a very stimulating environment in what concerns the extraordinary quality of its academic staff, and that experience will leave marks beyond this thesis.
+I express my thanks to Professor Florian Foos, Professor Ruben Rufino and all my PhD colleagues for the inspirational discussions and suggestions.
+Professors Kostas Matakos, John Meadowcroft, Mark Pennington, and Emily Skarbek have supported me in so many ways that I cannot even count.
+Thank you very much for the opportunities you provided, for your time and skills, and for having faith in me.
+
+I have been very fortunate to work with bright and motivated co-authors during this and other related projects.
+I am grateful to Rodolpho Bernabel, Gabriel Cepaluni, Diogo Costa, Malte Hendrickx, Robert McDonnell, Umberto Mignozzetti, and Gary Uzonyi for having taught me so much about both scientific research and life in general.
+
+One of the best experiences I have had in the previous years was to become affiliated with the Mercatus Center and the Institute for Humane Studies.
+There, I have met a community of smart, hard-working and innovative scholars that have in many ways contributed to this research.
+I would like to express my gratitude to Paul Dragos Aligica, Nigel Ashford, Peter Boettke, Don Boudreaux, Brandon Brice, John Buchmann, Jeff Carroll, Chris Coyne, Courtney Dunn, Bobbi Herzberg, Marcus Hunt, Nick Jacobs, Isaac Jilbert, Peter Lipsey, Ian Madison, Kelly Nelson, Gabriel Oliva, Raj Patel, Blaž Remic, Virgil Storr, Alienor van den Bosch, Richard Wagner, and Larry White for all the engaging conversations.
+I have learned so much from you.
+
+Completing this work would have been all the more difficult were it not for the support and constructive criticism provided by my friends and colleagues.
+Specially, I would like to thank Veronica Adaeva, André Amaro, Aline Bagatin, Fábio Barros, Michel Hulmann, Guilherme Kerr, Maurício Pantaleão, Letícia Sene, and the participants of academic seminars at King's College London, the Mercatus Center at George Mason University, and the Prometheus Institute in Berlin.
+I am deeply indebted to them for their encouragement, insightful comments, and hard questions.
+
+I am deeply thankful to several present and past members of the Friedrich Naumann Foundation for Freedom.
+Many years ago, I was received a scholarship to attend a short course in Germany, and that first experience abroad transformed my life ever since.
+I then decided to pursue my post-graduate studies in Europe and it was one of the best choices I have made.
+The present work is, to a large extent, a result of the push I received from the Naumann Foundation almost 10 years ago.
+I owe a great debt of gratitude to Cidália Achten, Gulmina Bilal, Rainer Erkers, Beate Forbriger, Verena Gierszewski, Birgit Lamm, Monica Lehmann, Stefan Melnik, and Wulf Pabst for their support.
+As time goes on, it is easy to see the huge impact they have had on my academic career.
+
+A very special word of thanks goes for my mother Rosali and my grandmother Maria.
+They have been absolutely wonderful over the years and have gone above and beyond to provide me with a good education.
+This dissertation is dedicated to them.
+\textit{Nós conseguimos}.
 

### 2018-04-23 chapters/02-synth.tex
add chapter 4; small changes

@@
 Over a span of only seven years (2000--2007), the number of annual violent deaths in the capital fell from 5,979 to 1,311, a 78\% decrease.\footnote{The homicide statistics cited in this paragraph come from the Centre for the Study of Violence, a research group of the University of São Paulo.
-Their dataset can be found at the following electronic address: \url{http://nevusp.org/downloads/bancodedados/homicidios/distritossp/num-homicidios-distritos-2000-2007.htm}.
+Their data set can be found at the following electronic address: \url{http://nevusp.org/downloads/bancodedados/homicidios/distritossp/num-homicidios-distritos-2000-2007.htm}.
 Access: March, 2016.} Significantly, São Paulo city became the safest state capital in Brazil \citep{mapa2011}.
@@
 
-Lastly, I estimate another synthetic control using another approach.
-Here, I employ a Bayesian structural time-series model to verify the stability of the previous results \citep{brodersen2015inferring}.
+Lastly, I estimate another synthetic control using a different approach.
+I employ a Bayesian structural time-series model to verify the stability of the previous results \citep{brodersen2015inferring}.
 This inference procedure is similar to that described in section \ref{sec:methods} and it also consists of matching pre-treatment values of the unit of interest, São Paulo, to other potential control states.
@@
 The \texttt{R} code below replicates all statistical analyses and graphs included in this chapter.
-The original data files as well as the final dataset are available at \url{https://github.com/danilofreire/homicides-sp-synth}.
+The original data files as well as the final data set are available at \url{https://github.com/danilofreire/homicides-sp-synth}.
 
@@
 
-# Prepare dataset
+# Prepare data set
 df$state <- as.character(df$state) # required by dataprep()

### 2018-04-23 chapters/03-bicho.tex
add chapter 4; small changes

@@
 The remainder of this article proceeds as follows.
-Section \ref{sec:organisation} presents a brief historical overview of the \emph{jogo do bicho}.
+Section \ref{sec:institution} presents a brief historical overview of the \emph{jogo do bicho}.
 It describes the necessary conditions for the emergence of the game and presents its basic organisational structure.
@@
 Section \ref{sec:capture} discusses the links between illegal gambling markets, samba schools and the Brazilian state.
-Section \ref{sec:conclusion} offers some concluding remarks.
+Section \ref{sec:conclusion3} offers some concluding remarks.
 
@@
 \section{Conclusion}
-\label{sec:conclusion}
+\label{sec:conclusion3}
 

### 2018-04-23 chapters/04-killings.tex
add chapter 4; small changes

\chapter{What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models}
\label{chap:killings}

\section{Introduction}
\label{sec:intro4}

Since the end of World War II, mass killings, genocides, and politicides have claimed over 34.5 million lives \citep{marshall2017pitf}.\footnote{Genocide and politicide are the attempted intentional destruction of communal or political groups, respectively \citep[see][]{harff1988toward}. Mass killing includes these atrocities, as well as attacks against civilians that result in at least 1,000 deaths but are not intended to destroy a particular group \citep[see][]{ulfelder2008assessing}. While some conflate the logic of these types of atrocities \citep[e.g.,][]{stanton2015regulating, valentino2004draining}, others claim genocide and politicide follow a different logic from other forms of government violence \citep{kalyvas2006logic,stanton2015regulating}. Note that these arguments focus on the logic of the violence, not legal definitions that separate and prioritise genocide as the crime of crimes \citep[see][]{schabas2000genocide}.} The international community has responded with an effort to prevent further state-sponsored mass murder by strengthening laws against crimes against humanity, genocide, and war crimes. It formed \textit{ad hoc}, hybrid, and a permanent criminal court to punish and deter future atrocities. Furthermore, the United Nations established a Special Adviser on the Prevention of Genocide and recognised its members’ responsibility to protect civilian populations within and outside their own borders. Yet, such atrocities still occur. Recently, President al-Assad of Syria has massacred tens of thousands of civilians during the Syrian Civil War \citep{goldman2017nyt}. Similarly, South Sudan's President Kiir is actively starving and killing civilians from dissident and rival tribes \citep{nichols2017reuters}. While there is some evidence that such atrocities may be declining since the Cold War \citep{valentino2014we}, the international community has been far from successful in realising slogans like ``Never Again'' and ``Not on My Watch'' \citep{cheadle2007not}.

Ultimately, successful prevention requires us to understand why these atrocities occur. In this vein, the academic community has laboured tremendously to establish evidence-based theories as to why governments engage in brutality against their civilian populations. Indeed, since 1995, there have been over 45 quantitative political science articles focused on explaining government-sponsored killing of civilians. Note, this number does not consider the myriad other books, qualitative articles, and non-political science research conducted into this matter. Overall, the mass violence literature generally agrees that as threat increases, so does the likelihood of atrocity, if the costs to such violence are not prohibitive. However, there is little consensus on what factors influence the level of threat or costs a regime faces. Part of the reason for this uncertainty is that scholars use very different model specifications when testing their arguments. Examining the quantitative literature on government violence against civilians, we found that in 45 studies, scholars used nearly 180 measurements to capture roughly 30 key concepts related to threat and costs. While models should be constructed specifically to test particular arguments, one concern is that small changes in model specification could in influence the robustness of empirical results and the inferences that we can draw from these results.

To overcome these limitations and provide a better understanding of government atrocity, we employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities. Our approach is similar to \cite{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but we provide additional tests to check which variables are able to predict out-of-sample cases of mass violence during both peacetime and wartime. In conducting this analysis, we address three debates in the mass violence literature:

\begin{enumerate}
    \item Why do some governments engage in mass killings, genocides, or politicides? This is the primary question asked by advocates, policymakers, and scholars in this field of research. It is also the question that the 45 studies mentioned above each attempt to answer.
    \item Does the logic underpinning government decision-making follow different patterns during peacetime and wartime. Recent research suggests that government atrocity occurs predominantly during periods of civil war \citep{harff2003no} which has led some scholars to restrict their analyses to only periods of civil war \citep[e.g.,][]{colaresi2008kill, valentino2004draining} or concentrate on predicting both the onset of civil war and atrocity \citep{goldsmith2013forecasting}. Yet, others estimate models of all country-year \citep[e.g.,][]{krain1997state, montalvo2008discrete}, raising questions of how well these studies speak to each other.
    \item Is there a difference in logic between those atrocities labelled as genocide or politicide, compared to other mass killings? While the Political Instability Task Force \citep{marshall2017pitf} provides the most widely used data on government atrocity, these data are limited only to cases of genocide and politicide. Others provide data with much more lenient inclusion criteria \citep[e.g.,][]{eck2007one, rummel1995democracy, stanton2015regulating, ulfelder2012forecasting}. These differences in definition of atrocity have led to divergent results, raising questions about important determinants of government behaviour \citep[for discussion, see][]{uzonyi2016domestic, wayman2010explaining}.
\end{enumerate}

Our analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013. We find that GDP \textit{per capita}, democratic regimes, and the post-Cold War period are negatively associated with mass killings, whereas previous political turmoil increases the risk of such violence. In contrast, the distributed random forest algorithm indicates that years since the last episode of mass violence, GDP \textit{per capita}, urban population, openness to trade, and the number of military personnel make the greatest contribution to our models’ out-of-sample predictive power. While we discuss the importance of these findings in more detail later in this article, a few notes are imperative to make here. First, in some ways, these findings confirm previous research–unstable countries are more likely to witness the regime employ atrocity \citep[e.g.,]{goldsmith2013forecasting,harff2003no,krain1997state}.  However, many of the factors scholars often cite as observable indicators of such instability--regime transitions, coups d'état, the presence of militias, etc.--are not good proxies for instability. Thus, policymakers may be looking for the incorrect signs of impending atrocity when seeking to prevent its onset. Second, these findings raise concerns about policy options for preventing violence against civilians. If our conclusion is that unstable countries are violent, then preventing atrocity likely requires significant investments of time and resources in state-building, which is often politically and practically unfeasible \citep{doyle2006making}. Lastly, our findings are robust to a myriad of specifications, functional forms, and additional tests as we describe below. This analysis contributes significantly to the political violence literature by highlighting the parsimonious nature of the logic behind government atrocity and clearing away much of the empirical clutter surrounding this conclusion.

\section{Empirical Methods}
\label{sec:methods4}

We employ two methods to test the robustness of the potential determinants of state-led violence. First, we use a variant of extreme bounds analysis (EBA) to check the statistical significance of 35 explanatory variables cited in the academic literature.\footnote{The complete list of variables included in this article is available in the online appendix. Our selection procedure was as follows. We included all variables that appeared in at least two two quantitative papers on mass killings which employed logit or probit models in a global sample. We considered only published articles in our list.} Researchers have employed the EBA to assess the sensitivity of the determinants of civil war \citep{hegre2006sensitivity}, coups d'état \citep{gassebner2016expect}, democratisation \citep{gassebner2013extreme}, economic growth \citep{levine1992sensitivity, sala1997just, sturm2005determinants}, nuclear deterrence \citep{bell2015examining}, political repression \citep{hafner2005right}, and state corruption \citep{serra2006empirical}. The method is particularly useful when there is no consensus about which covariates belong in the ``true'' regression model \citep[178]{sala1997just} and scholars worry that omitted or unnecessary predictors could bias the parameter estimates \citep[60]{angrist2008mostly, clarke2005phantom, elwert2014endogenous, spector2011methodological}. By combining a large set of regression models into a single posterior distribution, the EBA reduces the uncertainty associated with the \textit{ad hoc} selection of predictors and provides an intuitive way of assessing the relative strength of the statistical findings \citep{leamer1985sensitivity, sala1997just}.

Second, we employ distributed random forest (DRF) \citep{breiman2001random, h2o2017} to measure the predictive power of our set of independent variables. Random forest has been widely used in the machine learning community and consistently ranks among the best algorithms for predicting structured data.\footnote{Random forests, gradient boosting machines, and neural networks have won the highest number of competitions on Kaggle, a crowd-sourced platform for predictive modelling \citep{carpenter2011may}. While data analysts prefer neural networks for problems with a high number of hidden features such as image recognition, random forest and gradient boosting are the top choices for dealing with tabular data. See: \href{https://www.kaggle.com/antgoldbloom/what-algorithms-are-most-successful-on-kaggle}{https://www.kaggle.com/antgoldbloom/what-algorithms-are-most-successful-on-kaggle} (access: December 2017).} DRF does not require distribution assumptions, can be used with any type of response variables, and are able to automatically detect nonlinear relationships between correlates \citep{fernandez2014we, hill2014empirical, jones2015exploratory, muchlinski2015comparing}. Moreover, unlike neural networks and other deep learning algorithms \citep{castelvecchi2016can,rojas2013neural,shwartz2017opening}, the output of DRF models can be interpreted meaningfully. DRF offers precise estimates of the predictive ability of each individual variable, and it also allows researchers to graphically visualise the marginal effect of the covariates on the response variable \citep{friedman2001greedy,friedman2001elements,goldstein2015peeking}.

\subsection{Extreme Bounds Analysis}
\label{subsec:eba}

The main purpose of the extreme bounds analysis is to estimate the distribution of coefficients of each predictor $x$ in an exhaustive combination of regression models with $y$ as a dependent variable. \cite[308]{leamer1985sensitivity} argued that the EBA could certify scholars that ``minor changes in the list of variables do not alter fundamentally the conclusions, nor does a slight re-weighting of observations, nor correction for dependence among observations, etcetera, etcetera.'' He proposed that ``sturdy'' variables are those whose minimum and maximum of their coefficient distribution have the same sign and are situated at a distance from zero. If we are to use the conventional value of $p < 0.05$, the mean of the variable coefficients' distribution should be located at least $1.96$ standard deviations away from zero.

Leamer's criterion is intuitive, but other authors contend it is too strict for most social science applications. \cite{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response. If a given variable of interest appears as both positive and negative in the literature, it is likely that after a large a number of regressions the predictor will change signs and therefore be deemed fragile \citep[179]{sala1997just}. Sala-i-Martin then offered a less stringent version of the EBA in which researchers analyse the full distribution of point estimates instead of relying only on the coefficients' extreme bounds.

In this paper, we follow his advice and consider the whole range of values of $CDF(0)$. We choose to use the whole distribution because the aggregate $CDF(0)$ allows us to move away from a binary indicator of robustness, the one we obtain with Leamer's extreme bounds, and present the estimations with their appropriate degrees of confidence \citep[179]{sala1997just}\footnote{As \citet[179]{sala1997just} describes, ``if 95 percent of the density function for the estimates of $\beta_{1}$ lies to the right of zero and only 52 percent of the density function for $\beta_{2}$  lies to the right of zero, one will probably think of variable 1 as being more likely to be correlated with [a dependent variable $Y$] than variable 2.''}. Our main focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero. We do not assume that the CDF has a normal distribution: Many of our coefficients do not follow a normal distribution, so the generic model provides a better fit to our data. We specify our models as follows:

\begin{equation}
\text{\textit{Mass Killing Onset}}_{it} = \beta_{M}M_{it} + \beta_{F}F_{it} + \beta_{Z}Z_{it} + v_{it}
\end{equation}

Our main dependent variable is \textit{Mass Killing Onset}, which denotes the onset of government-sponsored killings. It was coded by \cite{ulfelder2008assessing}. The authors define a mass killing as ``any event in which the actions of state agents result in the intentional death of at least 1,000 noncombatants from a discrete group in a period of sustained violence'' \citep[2]{ulfelder2008assessing}. In our additional tests, we also estimate the models with an indicator of Genocide/Politicide Onset by \cite{harff2003no}. We prefer the first measure not only because it provides a clear numerical threshold that separates mass killings from low-level violence, but also because it focuses on the killing of members of a given community, not on the absolute number of deaths.

The indices $i$ and $t$ indicate country and year, respectively. $M$ is a set of three covariates that are included in every model due to their prominence in the literature \citep{levine1992vale}. In our analysis, we have added the natural logarithm of the GDP \textit{per capita} to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence. $F$ denotes a vector of variables of interest, and $Z$ is a vector of other control variables in addition to those included in $M$. Our first model includes the variable of interest, 3 additional control variables, and 3 covariates that appear in every regression. $v$ is the error term. In practice, however, since we are interested in the effect of all variables in the data set and we do not have true control variables except from $M$, in our case $F$ and $Z$ are interchangeable. Following \citet[514]{hegre2006sensitivity}, the independent variables have been lagged one year to reduce the risk of endogeneity. Although our dependent variable is dichotomous, we use linear probability models in our main analysis. As argued by \citet[298]{gassebner2016expect}, linear probability models are less prone to convergence problems, their estimation is faster, and their results can be readily interpreted. While we do employ logistic and probit models as robustness tests, we report the coefficients of the linear probability models as our primary estimates. Moreover, since the data are grouped into countries, we use cluster-robust standard errors to control for within-cluster error correlation.

One of the shortcomings of our data is that some variables are strongly correlated. For instance, we include the Correlates of War's Composite Index of National Capability (\textit{CINC}), which is composed by six different variables \citep{cow2017cinc,singer1988reconstructing}.\footnote{For more information about the CINC index, please refer to the Correlates of War website: \href{http://www.correlatesofwar.org/data-sets/national-material-capabilities/national-material-capabilities-v4-0}{http://www.correlatesofwar.org/data-sets/national-material-capabilities/national-material-capabilities-v4-0} (access: December 2017).} Multicollinearity would induce bias to our estimates, so we add CINC and three of its components\footnote{The variables correspond to the ratio between the national and the global values.} -- military expenditure, military personnel, and total population -- as mutually exclusive variables. These predictors are not included in the same regression model to avoid collinearity issues.

We add another set of mutually exclusive variables to reduce collinearity. There are three covariates that denote ongoing civil conflicts: one measured by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, another coded by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}. We include only one of these measures at a time.

As a last precaution against collinearity, we place a limit on the Variance Inflation Factor (\textit{VIF}) of all regression coefficients. The VIF estimates how much of the variance of each predictor is dependent on the other covariates in a model. A VIF of 1 indicates that the predictor is uncorrelated with the remaining covariates. The VIF limits are often arbitrary \citep{bell2015examining,o2007caution}, thus we use a moderately conservative VIF of 7 in our estimates. As robustness tests, we run the same models with different VIF cut-offs and without restriction.

Two variables were omitted from the EBA models but included in the machine learning models below. The first is \textit{democracy}, a dummy variable that indicates whether the country is a has a Polity IV score equal or higher than 5. The second is \textit{interstate war}, a binary covariate measuring if the country is at war in a given year \citep{sarkees2010resort}. We have decided to omit democracy because of its obvious correlation with the Polity measure and interstate war due to its correlation with our dependent variables. The EBA models do not converge if those variables are added. Since this problem does not affect machine learning algorithms, the two variables were included in the second set of estimations.

Lastly, we depart slightly from Sala-i-Martin's suggested method and do not assign weights to the EBA. Although he recommends using goodness-of-fit measures to construct regression weights, we agree with \citet{sturm2002robust} and \citet[299]{gassebner2016expect} and use the unweighted version of the CDF instead. Goodness-of-fit indicators are not equivalent to the probability of a given model being true \citep{achen1977measuring,anscombe1973graphs,king1986not}, and the weights constructed this way are not invariant to transformations in the dependent variable. Moreover, our data set has a number of missing observations, so that model comparison would be misleading \citep{lall2016multiple}. Thus, the results below are for unweighted estimations.

\subsection{Distributed Random Forest}
\label{sub:drf}

Random forest is a machine learning algorithm that consists of a combination of individual decision trees. In a classification problem, each decision tree uses a vector of covariates to split the dependent variable into two increasingly homogeneous parts \citep{breiman2001statistical}. However, decision trees are prone to overfitting, i.e., they match the original data set so closely that they tend to perform poorly with new data \citep{dietterich1995comparison,ho1998random}. Random forest, in contrast, avoids this issue by growing a decision tree only to a bootstrap sample of the original data, selecting random features at each split, then aggregating the different trees into a single prediction. If the independent variable is continuous, the algorithm will simply choose the average value of the predictions as the best candidate; if the covariate is discrete, the majority class will be employed. The simple procedure of leaving out some data points and growing separate trees with a random subset of covariates is sufficient to eliminate the risk of overfitting \citep[9-10]{jones2015exploratory}.

Random forest has many desirable properties, such as ``highly accurate predictions, robustness to noise and outliers, internally unbiased estimate of the generalisation error, efficient computation, and the ability to handle large dimensions and many predictors'' \citep[7]{muchlinski2015comparing}. Thus, random forest allows the researcher to estimate very flexible models with minimal assumptions. Unlike parametric methods such as ordinary least squares or logistic regressions, the analyst does not have to impose any distributional form to the data-generating process. As a result, random forest is able to effectively uncover complex, nonlinear interaction effects in the data without prespecification \citep{jones2015exploratory,strobl2007bias}.

The algorithm can also deal with heavily imbalanced data. Researchers can balance the classes of their response variable by assigning different sampling probability to each category. This enables the algorithm to make accurate prediction for rare events without adding prior distributions or resorting to subjective modelling choices \citep{chen2004using,del2014use,muchlinski2015comparing}.

In this paper we use distributed random forest (DRF) to model our data \citep{h2o2017}. The DRF is essentially identical to original random forest algorithm, but it has two additional features that are useful for our purposes. Firstly, DRF is optimised for big data, as it grows decision trees on separate cores to speed up computation time. Secondly, in DRF, non-observed cases are not assumed to be missing at random, but rather as values that contain information in themselves. When building decision trees, DRF treats the missing observations as a separate category that can go either left or right. This is a more conservative approach than assuming that missing cases fit into an underlying parametric distribution.\footnote{For more information about how the distributed random forest algorithm deals with missing observations, please refer to: \href{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html}{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html} (access: December 2017).}

The DRF has a series of hyperparameters that can be tuned to improve the algorithm's predictive performance. For instance, users can control the number of decision trees in each iteration, how deep trees should grow, which performance metric to adopt, whether to balance classes or to use cross-validation, and many other options. The interaction between parameters is generally complex and may involve thousands of potential combinations. As an example, a researcher interested in four parameters with 10 possible values each would have to estimate 10,000 models before deciding which is the most efficient one. Also, machine learning parameters are sensitive to the data at hand; so what seems an optimal solution for one problem cannot be readily implemented in another data set \citep{genuer2008random,goldstein2010application,jones2015exploratory}.

To address this issue, we have adopted an automated procedure to select the model parameters. We perform a grid search where the algorithm starts with a random combination of parameters, jumps to another randomly-chosen set, and then stops after it reaches a certain threshold \citep[123]{cook2017h2o}. We follow the literature on predictive political science and use the area under the ROC curve (AUC) as our model evaluation metric \citep[e.g.,][]{clayton2014will,hill2014empirical,ward2010perils,ward2013learning,weidmann2010predicting}. We set the metric as follows: If five random models had not increased the AUC by at least in 0.1\% comparing to the previous ones, the algorithm considers the result to be optimal. We set the maximum number of models to 1,000.

We have added several parameters to the grid search. The first is the number of independent trees to grow in each random forest model. The starting values are 256, 512, and 1024 trees. The machine learning literature does not provide a heuristic on how large a random forest should be, but \citet[166]{oshiro2012many} affirm that ``from 128 trees there is no more significant difference between the forests using 256, 512, 1024, 2048 and 4096 trees.'' We employ a more conservative approach and start from a higher value that the authors suggest as adding more trees do not reduce prediction accuracy \citep[7]{breiman2001statistical}. However, we find little evidence that the number of trees makes a noticeable difference to the estimations.

The depth of each decision tree also influences the algorithm performance. Deeper trees indicate more complex models, and in general they provide a better fit to the data. Nevertheless, this complexity comes at the risk of overfitting, so deeper trees are not necessarily the most adequate solution for every model \citep[596]{friedman2001greedy,segal2004machine}. In this article, we let the algorithm decide among using 10, 20, or 40 levels for each tree. In our view, these number offer a good balance between parsimony and complexity.

We test whether having balanced classes of our dependent variable (mass killing onset) affects the predictive ability of the model. Since the response measure is heavily imbalanced, oversampling the positive responses could potentially improve our results \citep{chawla2004special,del2014use,japkowicz2002class}. Somewhat unexpectedly, DRF had a better fit using unbalanced classes, which is the result we report in the next section.

We also vary how many variables should be considered for each split in the data. The DRF's default is to use $\sqrt{p}$, where p is the number of columns in the data set. As we have 34 covariates of interest, we have selected 5, 6 and 7 variables per split. The DRF uses a majority voting procedure to select which variable is most important. Additionally, the algorithm chooses the percentage of the training set to be modelled by each tree. The default option is 63.2\%, but we include the options of using 50\% and 100\% of the data. Similarly, we give a range of options for choosing how many columns will be included in each tree. The algorithm can randomly choose among 50\%, 90\% or 100\% of the independent variables when estimating a decision tree.

Finally, we use four types of histogram to find optimal split points for each independent variable. Decision trees consider every value of a given independent variable as a potential candidate for a split in the training data. This process is notably time-consuming, and computation time can be significantly reduced at little loss of precision by taking discrete values of the predictor distribution. The DRF algorithm offers four choices of histogram selection and we include all of them in our estimations.

\section{Results}
\label{sec:results4}

Table \ref{tab:eba1} summarises our main EBA results with Ulfelder and Valentino's \citeyear{ulfelder2008assessing} \textit{Mass Killing Onset} as the dependent variable. The table shows the average coefficient estimate of all regressions for each robust variable along with their mean standard deviations.\footnote{A list of all independent variables and coding rules are available in the online appendix.} The table also displays the percentage of regressions that are statistically significant at the 90\% level. $CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{We show whichever area is the largest. The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is our main statistic of interest, and we consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}. Lastly, we report the number of estimated regressions models which included each variable.

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & -0.0091 & 0.0052 & 76.055 & 0.9335 & 226707 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.0133 & 0.0085 & 72.845 & 0.9472 & 35614 \\
UCDP civil war onset & 0.0529 & 0.0321 & 52.378 & 0.9441 & 20854 \\
Previous riots & 0.0140 & 0.0100 & 56.242 & 0.9216 & 35614 \\
UCDP ongoing civil war & 0.0172 & 0.0115 & 65.652 & 0.9092 & 20854 \\
Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.674 & 0.9050 & 35614 \\
Polity IV squared & -0.0002 & 0.0001 & 61.206 & 0.9031 & 35614 \\ \hline
\end{tabular}
\caption{Extreme Bounds Analysis -- Mass Killings (robust variables only)}
\label{tab:eba1}
\end{table}

Seven variables pass our EBA criteria. Three variables decrease the likelihood of mass killings. First, as widely suggested in the literature, the natural logarithm of GDP per capita is robustly associated with the onset of mass killings 
\citep[e.g.,][]{besanccon2005relative, easterly2006development,esteban2015strategic}. In our models, GDP per capita has a negative effect on onset risk is statistically significant in 76\% of the models. Second, the post-Cold War years are correlated with lower levels of government violence. Despite eventual outbreaks of large-scale attacks against civilians \citep[e.g.,][]{prunier2005darfur,tatum2010genocide,waldorf2009revisiting}, after 1991 these events have become less prevalent. Indeed, this finding is in line with several studies that point to a general decline in violence over the last decades, including riots, civil wars, and urban crime \citep{eisner2003long,pinker2011better}. The third variable negatively correlated with mass killings is the squared term of the Polity IV political regime index. Although the linear Polity IV indicator does not reach our threshold for significance, about 90\% of the distribution area of Polity IV squared is located below zero. This finding points to a nonlinear relationship between political regime and mass killings, thus providing further evidence that democracy reduces state-sponsored violence \citep{rost2013will,rummel1995democracy} and that regimes that mix democratic with autocratic features have the highest risk of conflict \citep{hegre2001toward,mitchell2013domestic,muchlinski2014grievances,regan2010changing}.

Four variables are positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of state-sponsored violence. The onset and continuation of civil wars are correlated with mass killings, but only when we employ the UCDP measures of violent conflict. We find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}. Coding choices likely account for this difference, and the result is robust to a myriad of specifications. Former instances of political turmoil also have a positive coefficient in our models. Countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent 
\citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}. Our results also show that a higher levels of ethnic diversity increase the likelihood of atrocities against civilians. Nevertheless, the variable does not pass all additional tests we implement below and the sturdiness of this finding remains open to question. The uncertainty about the impact of ELF on mass killings is also reflected in the literature. As \citet[237]{hoeffler2016development} points out, \citet{rummel1995democracy} reports that ELF is not statistically significant in his models, \citet{wood2014opportunities} finds a negative relationship and \citet{querido2009state} finds a positive one. Our analysis does not offer a definitive answer for this question either.

A different set of variables reach significance if we only consider mass killings that occur during civil wars. We ran the models on three indicators of violent armed conflict. The first is the civil war measure by the Uppsala Conflict Database Program  (\citeyear{allansson2017organized,gleditsch2002armed}). Here, two variables meet our criteria for robustness, Post-Cold War years and conflicts with territorial aims, both negatively correlated with mass killing onset. When we limit our analysis to cases marked as civil wars by the Correlates of War project, years since the last mass killing episode, previous riots, former human rights abuses, and ethnic diversity appear have a positive effect on the outcome. Rather surprisingly, the presence of militias has a negative effect on genocides and politicides. Finally, our model using ethnic civil war shows similar results: Militias have a negative impact on the likelihood of mass killings during ethnic conflict, as do conflicts over territorial aims. This finding is in contradiction with previous research \citep{koren2017means} and is deserving of further exploration. Overall, these results suggest two takeaways. First, the determinants of mass killing onset during peacetime and wartime are different from each other. While there are more restraints on atrocities during peacetime, these restraints fail once fighting begins. This raises particularly interesting questions about the pacifying nature of democracy. Second, how civil wars are classified significantly alters the results produced. This raises concerns for scholars about making claims based on only one data set.

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & -0.0091 & 0.0052 & 76.055 & 0.9335 & 226707 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.0133 & 0.0085 & 72.845 & 0.9472 & 35614 \\
UCDP civil war onset & 0.0529 & 0.0321 & 52.378 & 0.9441 & 20854 \\
Previous riots & 0.0140 & 0.0100 & 56.242 & 0.9216 & 35614 \\
UCDP ongoing civil war & 0.0172 & 0.0115 & 65.652 & 0.9092 & 20854 \\
Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.674 & 0.9050 & 35614 \\
Polity IV squared & -0.0002 & 0.0001 & 61.206 & 0.9031 & 35614 \\ \hline
\end{tabular}
\caption{Extreme Bounds Analysis -- Mass Killings (robust variables only)}
\label{tab:eba1}
\end{table}

The random forest models confirm some of the main findings of the EBA, yet they also show some interesting prediction patterns. Figure \ref{fig:drfuv} presents the six most important predictors of state-sponsored violence in out-of-sample tests. Overall, the machine learning estimations have a good fit, with an AUC of about 0.8 in the test samples.

However, the findings should be interpreted with caution since variable importance metrics are sensitive to the choice of random seed numbers. But despite that expected variation, the results in the random forest models are quite stable even when estimated with different random seeds.

\begin{figure}[h]
\includegraphics[width=.9\textwidth, height=8cm]{images/drf.pdf}
\caption{Distributed Random Forest -- Variable Importance (scaled)}
\label{fig:drfuv}
\end{figure}

We see that the most important predictor in our models is the natural log of GDP per capita. This confirms the findings of the EBA model. Years since the last mass killing episode has a nonlinear relationship with large-scale violence: The likelihood of renewed genocides is small in the first years, although it increases slightly over the next decades. The percentage of urban population exhibits a similar pattern. Rural areas are less prone to mass killings, yet the probability of such events rise according to a country’s level of urbanisation. We understand both GDP per capita and percentage of urban population as proxies for state capacity. Conducting a mass killing campaign demands some minimal level of military hierarchy and resources, and countries that are overwhelmingly rural are probably less able to amass the human and physical capital required to conduct such task. Furthermore, urban population tend to be less dispersed, what facilitates a large attack by government forces.

\vspace{.5cm}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth, height=9cm]{images/drfdpp.pdf}
\caption{Distributed Random Forest -- Partial Dependence Plots}
\label{fig:drfdpp}
\end{center}
\end{figure}

The ethnic polarisation variable indicates that countries in which a large share of the population is excluded from politically relevant positions, the lower the probability of a mass killing. Groups that manage to control the state apparatus rarely face dangerous opposition movements, and as such they rarely resort to mass violence to impose their rule over the population. Next, the size of the military personnel substantially increases the likelihood of state-sponsored violence. While small and weak states tend to be more prone to civil war and other types of conflict \citep{collier2004greed,fearon2003ethnicity}, they are less likely to experience one-sided state violence. Lastly, the Polity IV index shows that mixed regimes have a higher likelihood of atrocity, and democracies tend to have lower levels of state-led violence.

With respect to mass atrocities that occur during armed conflicts, we see that log GDP per capita and years since mass killings are again relevant predictors. Additionally, both covariates are among the 10 most important predictors in the three models. The results obtained with the UCDP civil war data are broadly similar to those of the previous model, but it highlights that total battle deaths is the predictor that has a noticeable impact on mass killings onset. The partial dependence graph shows a large positive effect along the curve, with a pronounced jump at around 200,000 deaths.

\vspace{1cm}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth, height=9cm]{images/drfdpp2.pdf}
\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (UCDP data)}
\label{fig:drfdpp2}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth, height=9cm]{images/drfdpp3.pdf}
\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (COW data)}
\label{fig:drfdpp3}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth, height=9cm]{images/drfdpp4.pdf}
\caption{ Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. data)}
\label{fig:drfdpp4}
\end{center}
\end{figure}

\newpage

When using the COW data, we find that GDP per capita and total battle deaths have little influence in the likelihood of mass killings, and years since the last state-led atrocity increases the chances of a future event, a result we have also seen in the EBA models. As expected, higher levels of abuse against the physical integrity of citizens are associated with mass killings. Ethnic polarisation and the threshold of excluded population have a somewhat constant effect in our estimations, yet both show a sharp decline in probability when they reach their maximum levels. This provides more evidence to the idea that when certain groups are able to concentrate power, they tend not to resort to mass-scale violence as a political means.

Our last model concerns the onset of ethnic conflicts as defined by \citet{cederman2010ethnic}. The graphs show ethnic wars operate with a different dynamics than other armed conflicts. We see that an increase in all the most important predictors have a large negative impact on the likelihood of mass killings, thus we conclude that poor, unstable countries are more prone mass killings during ethnic conflicts.

\section{Additional Tests}
\label{sec:additional-tests4}

We estimate a set of additional regressions to assess the robustness of our main findings. In regard to the EBA, we include 10 variants of our original model. All of them largely confirm our prior results. First, we varied the number of covariates included in each regression to 3 and 5 while keeping the $M$ set of 3 variables included in every model, that is, the natural logarithm of GDP per capita, the Polity IV index, and a linear time trend since the last episode of state-sponsored violence. The results are identical to the main model when we add 3 control variables at a time. The same covariates discussed above remain significant and with the same sign: log of GDP per capita (negative), post-Cold War period (negative), previous riots (positive), civil war onset and ongoing indicators as measured by the UCDP data set (both positive), Polity IV squared (negative), and ethnic diversity (negative). When 5 control variables are added to every model, three variables have a CDF(0) of 0.9 or higher. They are the natural logarithm of GDP per capita, civil war onset (UCDP), previous riots, and the dummy indicating the post-Cold War period. Ethnic fractionalisation and Polity IV squared become marginally significant with a CDF(0) of about 0.88.

Second, we place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving our results. The test is similar to that employed by Bell (2015). While in the main analysis we use a VIF of 7, in the additional analyses we include a conservative VIF of 2.5, a less strict value of 10, and a model without any restriction. The results are unchanged. The two models with different values of VIF show identical results to the results displayed in table \ref{tab:eba1}. In the model with no VIF restriction, however, ethnic fractionalisation fails to meet our threshold by a very small margin. The CDF(0) of that covariate is 0.897, very close to the required value of 0.9.

We also estimate our models using logit and probit regressions. In order to deal with the issue of complete separation \citep{bell2015questioning,zorn2005solution} we follow \citet{gelman2008weakly} and add a weakly informative prior distribution to the coefficients. We first scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then we added a Cauchy distribution with center 0 and scale 2.5. The probit regressions use a scale of $2.5 \times 1.6$, which is also recommended by the authors \citep{arm2017rpackage}.

In both cases, the logarithm of GDP per capita, post-Cold War period, previous riots, and Polity IV squared remain significant. In contrast, ethnic fractionalisation and former conflict as measured by the UCDP data set are again close to the 0.9 significance level.

In regard to random forests, grid searches are themselves a data-driven selection of many possible machine learning models, thus it is not strictly necessary to run a batch of additional tests. Nevertheless, we performed a series of grid searches using three different seeds obtained from \texttt{Random.Org} to estimate how the choice of different starting numbers would influence the model outcomes. The output of those models are largely comparable.

As a last set of robustness models, we estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide. No variable appear as significant in our EBA models for genocide or politicide onset in peacetime. When we limit our sample to civil war years, the Post-Cold War period is again negatively correlated with the outcome when using the Correlates of War data set. Additionally, excluded population has a negative sign in more than 90\% of the models using both Correlates of War’s and Cederman et al's (\citeyear{cederman2010ethnic}) indicators of conflict. Displaced population also has a negative effect in the Correlates of War data set. During ethnic conflicts, our dummy variable for political assassinations has a negative impact on the onset of genocides. Overall, we conclude then that the significant covariate of genocide and politicide onset differ significantly from those of more general forms of government mass violence. Finally, the machine learning models are comparable to the ones we present above, with a similar set of variables appearing in the random forest estimations. The complete results are available in the appendix.

\section{Conclusion}
\label{sec:conclusion4}

In this paper, we apply extreme bounds analysis and distributed random forests to estimate the robustness and predictive ability of 40 variables that have been pointed out as potential determinants of mass killings. We find that from all variables we employ, GDP per capita is the covariate that appears statistically significant more often in our models. In the EBA estimations, we see that the level of democracy, and the post-Cold War period are negatively associated with our dependent variable, whereas ethnic diversity, civil wars, and previous political turmoils increase the likelihood of those events. When we run a machine learning algorithm to estimate out-of-sample predictive performance, GDP per capita and years since last mass killings again appear as important predictors, and variables related to military capabilities and the composition of the population are included in most models. In a nutshell, our tests provide strong evidence in favour of two well-established findings in the specialised literature: Mass killings are unlikely to happen in rich, stable countries and there is a positive association between several types of political conflict, such as previous riots, civil wars and government atrocities.

Nevertheless, there is considerable heterogeneity in some of our findings. Mass killings are rare outcomes, so it is possible that our analysis does not capture all variables that explain the onset of large-scale violence, or that the effect of predictors vary across time and space \citep[8]{bell2015examining}. Additionally, the findings point out that mass killings may have different causes according to the context in which they erupt, so a general theory of state atrocities may obscure important details in our understanding of the root causes of state killings.

Yet we see this diversity of outcomes under a positive light. The findings presented in this text suggest new avenues for research, and we believe they also highlight the importance of scholars moving from purely correlational, cross-country regressions to other methods that can yield more robust predictions and causal explanations. For instance, why are mass killings in ethnic conflicts correlated with a different set of variables than in armed conflicts in general? Would the results remain robust had scholars decided to code ethnic conflicts in another way? More theoretical advancement would also be welcome. Given that GDP per capita is negatively correlated to state atrocities in virtually every model, it would be interesting to unpack the causal mechanisms by which it operates by testing more specific mechanisms. Disaggregated data can also be of great help to further current research on genocides, politicide, and mass killings.

In terms of practical implications, the results indicate that regime change and pro-growth economic policies are the most efficient ways to prevent mass killings. The international community can therefore play a role in deterring leaders from using force against their own population, either by offering support for domestic opposition groups, intervening, or by fostering economic development. Although costly in the short run, these measures would substantially decrease the likelihood of state violence by breaking the ``conflict trap'' in which past conflicts create the condition for new ones \citep{collier2003breaking}.

\section{Appendix} 
\label{sec:mk-appendix}

\subsection{Variable Selection}
\label{sec:mk-vs}

This appendix contains all required information to replicate the numerical analyses presented in the paper. We present a series of histograms with the coefficient distribution of every variable included in our EBA models. Robust variables  (CDF(0) $> 0.9$) are described in further detail in the tables below.

The complete results can also be retrieved from data included at the following GitHub repository: \href{https://github.com/danilofreire/mass-killings}{https://github.com/danilofreire/mass-killings}. \textt{R} code is available in section \ref{sec:mk-code}. The analyses were conducted on \texttt{R} version 3.4.4 (15-03-2018) and Ubuntu 16.04.4 LTS.

\subsection{Descriptive Statistics}
\label{sec:mk-ds}

\begin{table}[!htbp] \centering 
  \caption{Descriptive Statistics} 
  \label{tab:mk-ds} 
\footnotesize 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Country code & 9,162 & 452.84 & 247.74 & 2 & 950 \\ 
Year & 9,162 & 1,983.56 & 18.77 & 1,945 & 2,013 \\ 
Mass killing onset & 9,162 & 0.01 & 0.11 & 0 & 1 \\ 
Years since last mass killing & 9,162 & 23.81 & 17.71 & 0 & 68 \\ 
Years since last mass killing squared & 9,162 & 880.44 & 1,086.13 & 0 & 4,624 \\ 
Years since last mass killing cubed & 9,162 & 39,339.21 & 65,025.39 & 0 & 314,432 \\ 
UCDP civil war onset & 8,733 & 0.02 & 0.14 & 0 & 1 \\ 
UCDP civil war ongoing & 8,733 & 0.15 & 0.36 & 0 & 1 \\ 
COW civil war onset & 8,160 & 0.01 & 0.12 & 0 & 1 \\ 
COW civil war ongoing & 8,160 & 0.07 & 0.25 & 0 & 1 \\ 
Ethnic war start & 7,760 & 0.01 & 0.12 & 0 & 1 \\ 
Ethnic war ongoing & 7,760 & 0.11 & 0.31 & 0 & 1 \\ 
Assassination dummy & 8,991 & 0.08 & 0.27 & 0 & 1 \\ 
Democracy dummy & 8,991 & 0.37 & 0.48 & 0 & 1 \\ 
Ethnic diversity (ELF) & 6,981 & 0.41 & 0.31 & 0 & 1 \\ 
Mountainous Terrain & 7,358 & 2.14 & 1.43 & 0.00 & 4.56 \\ 
Population & 8,293 & 32,993.61 & 112,886.40 & 118.21 & 1,324,353.00 \\ 
Real GDP & 8,293 & 215,317.70 & 804,827.20 & 129.68 & 13,193,478.00 \\ 
Ln of real GDP per capita & 8,293 & 8.25 & 1.20 & 4.89 & 13.36 \\ 
Real GDP per capita & 8,293 & 8,104.20 & 18,376.73 & 132.82 & 632,239.50 \\ 
Polity IV & 8,558 & 0.42 & 7.50 & $-$10 & 10 \\ 
Polity IV squared & 8,558 & 56.35 & 32.59 & 0 & 100 \\ 
Excluded population & 6,981 & 0.16 & 0.22 & 0.00 & 0.98 \\ 
Discriminated population & 6,981 & 0.06 & 0.15 & 0.00 & 0.98 \\ 
Polarisation (all groups/main group) & 6,981 & 0.70 & 0.26 & 0.05 & 1 \\ 
Polarisation (all groups/population) & 6,981 & 0.63 & 0.32 & 0 & 1 \\ 
Polarisation (included groups/population) & 5,610 & 0.64 & 0.32 & 0 & 1 \\ 
Polarisation (included groups/main group) & 6,981 & 0.23 & 0.35 & 0 & 1 \\ 
Discriminated dummy & 6,981 & 0.35 & 0.48 & 0 & 1 \\ 
Interstate war & 8,159 & 0.04 & 0.19 & 0 & 1 \\ 
Military expenditure & 8,290 & 4,607,120 & 27,785,906 & 0 & 693,600,000 \\ 
Military personnel & 8,620 & 176.70 & 520.90 & 0 & 12,500 \\ 
Percent urban population & 8,767 & 0.22 & 0.17 & 0.00 & 1.51 \\ 
Post-Cold War & 8,991 & 0.40 & 0.49 & 0 & 1 \\ 
Coup dummy & 8,587 & 0.05 & 0.21 & 0 & 1 \\ 
Riot dummy & 8,991 & 0.16 & 0.36 & 0 & 1 \\ 
War with territory aims & 8,924 & 0.07 & 0.26 & 0 & 1 \\ 
Total trade & 8,174 & 53,804.01 & 222,209.90 & 0.80 & 4,825,363.00 \\ 
Trade dependence & 7,670 & 0.26 & 0.69 & 0.0001 & 22.11 \\ 
Militias & 4,097 & 0.22 & 0.42 & 0 & 1 \\ 
Physical integrity & 4,499 & 4.73 & 2.31 & 0 & 8 \\ 
CINC & 8,767 & 0.01 & 0.02 & 0.00 & 0.38 \\ 
Total battle deaths & 714 & 6,050.86 & 24,404.78 & 100 & 350,000 \\  
Guerrilla dummy & 714 & 0.81 & 0.40 & 0 & 1 \\ 
Regime transition & 1,221 & $-$4.24 & 41.50 & $-$77 & 99 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\raggedright{\newline \textit{Note}: All independent variables were lagged one year.}
\end{table} 
\normalsize

\newpage

\subsection{Extreme Bounds Analysis Extensions}
\label{sec:mk-ebae}

\subsubsection{Complete Results}

Here we present a series of histograms with the distribution of coefficients of all variables of the main EBA model. There are 36 variables in total, of which seven are robust: Log GDP per capita, post-Cold War period, onset and ongoing civil wars (measured by the UCDP), previous riots, ethnic diversity and the squared term of the Polity IV index.

\vspace{1cm}

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & -0.0091 & 0.0052 & 76.055 & 0.9335 & 226707 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.0133 & 0.0085 & 72.845 & 0.9472 & 35614 \\
UCDP civil war onset & 0.0529 & 0.0321 & 52.378 & 0.9441 & 20854 \\
Previous riots & 0.0140 & 0.0100 & 56.242 & 0.9216 & 35614 \\
UCDP ongoing civil war & 0.0172 & 0.0115 & 65.652 & 0.9092 & 20854 \\
Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.674 & 0.9050 & 35614 \\
Polity IV squared & -0.0002 & 0.0001 & 61.206 & 0.9031 & 35614 \\ \hline
\end{tabular}
\caption{Extreme Bounds Analysis -- Mass killings (robust variables only)}
\label{tab:eba1}
\end{table}

\clearpage
\begin{sidewaysfigure}
    \centering
    \includegraphics[width=\textwidth]{images/mk.pdf}
    \caption{Extreme Bounds Analysis -- Mass killings}
    \label{fig:hist-mk}
\end{sidewaysfigure}
\clearpage

\newpage

%%%%% Add models during civil wars

\subsubsection{Alternative Number of Variables}

The following table shows the results of an EBA model using 3 variable combinations at a time. The results are very similar to those reported in the paper, in which we used 4 variables in each model.

To speed up computational time, the models included in this appendix are based on 50,000 random draws from the full set of all possible regression models. \citet[819]{salaimartin2004determinants} argue that random sampling produces unbiased estimates of the regression coefficients, and as such can accelerate convergence of the results. The models presented in the paper, however, include the full set of regressions.

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & 0.0082 & 0.0043 & 81.439 & 0.9504 & 40677 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.0121 & 0.0069 & 77.804 & 0.9609 & 5064 \\
UCDP civil war onset & 0.0523 & 0.0292 & 62.561 & 0.9574 & 3304 \\
Previous riots &0.0134 & 0.0084 & 65.936 & 0.9401 & 5064 \\
UCDP ongoing civil war & 0.0177 & 0.0094 & 72.367 & 0.9372 & 3304 \\
Polity IV squared & -0.0002 & 0.0001 & 66.035 & 0.9268 & 5064 \\ 
Ethnic diversity (ELF) & 0.0162 & 0.0110 & 70.794 & 0.9266 & 5064 \\\hline
\end{tabular}
\caption{EBA -- 3 variables in each regression model}
\label{tab:mk-3vars}
\end{table}

\clearpage
\begin{sidewaysfigure}
    \centering
    \includegraphics[width=\textwidth]{images/mk-3vars.pdf}
    \caption{EBA -- 3 variables in each regression model}
    \label{fig:hist-mk-3vars}
\end{sidewaysfigure}
\clearpage

Table \ref{tab:mk-5vars} presents the results for models with up 5 variables in each regressions. In contrast with the main EBA model, the indicators of UCDP ongoing civil wars, ethnic diversity, and Polity IV square drop out of significance. Their CDF(0) is 0.88, just marginally below our threshold of 0.9.

\vspace{1cm}

\begin{table}[!htpb]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & -0.010 & 0.006 & 70.806 & 0.9161 & 50000 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.014 & 0.010 & 68.496 & 0.9336 & 9532 \\
UCDP civil war onset & 0.053 & 0.035 & 44.784 & 0.9308 & 5100 \\
Previous riots & 0.015 & 0.012 & 47.988 & 0.9047 & 9569 \\\hline
\end{tabular}
\caption{EBA -- 5 variables in each regression model}
\label{tab:mk-5vars}
\end{table}

\clearpage
\begin{sidewaysfigure}
    \centering
    \includegraphics[width=\textwidth]{images/mk-5vars.pdf}
    \caption{EBA -- 5 variables in each regression model}
    \label{fig:hist-mk-5vars}
\end{sidewaysfigure}
\clearpage

\subsubsection{Alternative Variance Inflation Factors}

In this section, we estimate the EBA model with different values of Variance Inflation Factor (VIF), a measure of multicollinearity for linear models. There is no standard definition about what constitutes an acceptable VIF value, although researchers often use a value of 10 as rule of thumb to indicate strong multicollinearity \citep[674]{o2007caution}. Our original model used 7 as a cutoff, and here we employ 10 (less strict), a lower value of 2.5 (more conservative), and a model without any VIF value (no restriction). The results are essentially identical to those in the main model. In the model with no VIF restriction, however, ethnic fractionalisation fails to meet our threshold by a very small margin. The CDF(0) of that covariate is 0.897, very close to the required value of 0.9.

\vspace{1cm}

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & -0.0091 & 0.0052 & 76.354 & 0.9343 & 50000 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.0134 & 0.0084 & 73.540 & 0.9495 & 7929 \\
UCDP civil war onset & 0.0529 & 0.0322 & 52.141 & 0.9438 & 4553 \\
Previous riots & 0.0140 & 0.0100 & 56.433 & 0.9216 & 7772 \\
UCDP ongoing civil war & 0.0172 & 0.0113 & 66.013 & 0.9113 & 4587 \\
Ethnic diversity (ELF) & 0.0182 & 0.0136 & 56.872 & 0.9056 & 8076 \\
Polity IV squared & -0.0002 & 0.0001 & 60.791 & 0.9021 & 7835 \\ \hline
\end{tabular}
\caption{EBA -- VIF 10 (less strict)}
\label{tab:eba1}
\end{table}

\clearpage
\begin{sidewaysfigure}
    \centering
    \includegraphics[width=\textwidth]{images/mk-high-vif.pdf}
    \caption{EBA -- VIF 10 (less strict)}
    \label{fig:hist-mk-high-vif}
\end{sidewaysfigure}
\clearpage

\vspace{1cm}

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & -0.0090 & 0.0051 & 76.055 & 0.9343 & 49620 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.0132 & 0.0084 & 72.845 & 0.9490 & 7929 \\
UCDP civil war onset & 0.0529 & 0.0322 & 52.378 & 0.9438 & 4553 \\
Previous riots & 0.0141 & 0.0101 & 56.242 & 0.9199 & 7772 \\
UCDP ongoing civil war & 0.0174 & 0.0114 & 65.652 & 0.9103 & 4587 \\
Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.674 & 0.9054 & 8076 \\
Polity IV squared & -0.0002 & 0.0001 & 61.206 & 0.90267 & 7835 \\ \hline
\end{tabular}
\caption{EBA -- VIF 2.5 (more conservative)}
\label{tab:eba1}
\end{table}

\clearpage
\begin{sidewaysfigure}
    \centering
    \includegraphics[width=\textwidth]{images/mk-high-vif.pdf}
    \caption{EBA -- VIF 2.5 (more conservative)}
    \label{fig:hist-mk-high-vif}
\end{sidewaysfigure}
\clearpage

\newpage

\subsubsection{Generalised Linear Models}

We estimate the same EBA with logit and probit in addition to the linear probability model reported above. The results are

\subsection{Random Forest Extensions}
\label{sec:mk-rfe}

\subsection{\texttt{R} Code}
\label{sec:mk-code}

The code below reproduces the analyses included here.

\singlespacing
\small
\begin{verbatim}
######################
### Data Wrangling ###
######################

\end{verbatim}

\doublespacing
\normalsize

### 2018-04-23 dissertation.tex
add chapter 4; small changes

@@
 \usepackage{float}
+\usepackage{rotating}
+\usepackage{adjustbox}
 \usepackage{pgf}
@@
 \usepackage{caption}
+\usepackage{dcolumn}
 \usepackage[UKenglish]{babel}
@@
 \clubpenalty=10000
+\usepackage[fencedCode,inlineFootnotes,citations,definitionLists,hashEnumerators,smartEllipses,hybrid]{markdown}
+\markdownSetup{rendererPrototypes={
+   link = {\href{#2}{#1}},
+   image = {\begin{figure}[hbt!]
+     \centering
+     \includegraphics{#3}%
+     \ifx\empty#4\empty\else
+     \caption{#4}%
+     \fi
+     \label{fig:#1}%
+     \end{figure}}
+}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \noindent 
-I hereby declare that except where specific reference is made to the work of others, the contents of this dissertation are original and have not been submitted in whole or in part for consideration for any other degree or qualification in this, or any other university.
-This dissertation is my own work and contains nothing which is the outcome of work done in collaboration with others, except as specified in the text and Acknowledgements.
+I hereby declare that except where specific reference is made to the work of others, the contents of this thesis are original and have not been submitted in whole or in part for consideration for any other degree or qualification in this, or any other university.
+This thesis is my own work and contains nothing which is the outcome of work done in collaboration with others, except as specified in the text and Acknowledgements.
 
@@
 % SUMMARY HERE (300 word limit for most subjects):
-This dissertation addresses four questions on political and criminal violence.
+This thesis addresses three questions on political and criminal violence.
 The first essay is an empirical evaluation of a broad set of homicide reduction policies implemented in the state of São Paulo, Brazil.
@@
 
-In the third essay, co-authored with Gary Uzonyi, we employ extreme bounds analysis and two machine learning methods -- distributed random forests and gradient boosting machines -- to identify the key determinants of state-sponsored violence.
-We test the sensitivity of 35 variables on a sample of 177 countries from 1945 to 2013.
-Our results show that GDP per capita and the post-Cold War period are negatively associated with mass killings, whereas previous political turmoils increase the risk of state-led violence.
-Years since the last episode of mass violence, GDP per capita, urban population, openness to trade, and the number of military personnel make the greatest contribution to our models' out-of-sample predictive power.
-
+In the third essay, co-authored with Gary Uzonyi, we employ extreme bounds analysis and distributed random forests to identify the key determinants of state-sponsored violence.
+We test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+The results show that GDP per capita, democracy, and the post-Cold War period are negatively associated with mass killings, whereas ethnic diversity, civil wars, and previous political turmoils increase the risk of state-led violence.
+GDP per capita, years since the last episode of mass violence, urban population, ethnic polarisation, and the number of military personnel make the greatest contribution to the models' out-of-sample predictive power.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % ACKNOWLEDGEMENTS HERE:
-I would like to extend thanks to the many people, in many countries, who so generously contributed to the work presented in this dissertation.
+I would like to extend thanks to the many people, in many countries, who so generously contributed to the work presented in this thesis.
 First and foremost, I want to thank my supervisor, Professor David Skarbek.
 It was a great privilege and honour to work and study under his guidance, and he has taught me more than I could ever give him credit for here.
-David was not only an excellent supervisor but also a great friend; his empathy, enthusiasm, and sense of humour made my stay at King's and in London a very pleasant one.
+David was not only an excellent supervisor but also a great friend; his empathy, enthusiasm, and sense of humour made my stay at King's College a very pleasant one.
 I simply could not have wished for a better mentor for my PhD studies.
@@
 
-The Department of Political Economy at King's College London has provided me with a very stimulating environment in what concerns the extraordinary quality of its academic staff, and that experience will leave marks beyond this thesis.
-I express my thanks to Professor Florian Foos, Professor Ruben Rufino and all my PhD colleagues for the inspirational discussions and suggestions.
+The Department of Political Economy at King's College London has provided me with a very stimulating environment in what concerns the extraordinary quality of its students and  academic staff, and that experience will leave marks beyond this thesis.
+I express my thanks to Professor Florian Foos, Professor Rubén Ruiz-Rufino, and all my PhD colleagues for the inspirational discussions and suggestions.
+David Muchlinski has taught me many things about machine learning and I am very thankful for that.
 Professors Kostas Matakos, John Meadowcroft, Mark Pennington, and Emily Skarbek have supported me in so many ways that I cannot even count.
@@
 I have been very fortunate to work with bright and motivated co-authors during this and other related projects.
-I am grateful to Rodolpho Bernabel, Gabriel Cepaluni, Diogo Costa, Malte Hendrickx, Robert McDonnell, Umberto Mignozzetti, and Gary Uzonyi for having taught me so much about both scientific research and life in general.
+I am grateful to Rodolpho Bernabel, Gabriel Cepaluni, Diogo Costa, Guilherme Jardim Duarte, Malte Hendrickx, Robert McDonnell, Umberto Mignozzetti, and Gary Uzonyi for having taught me so much about both scientific research and life in general.
 
@@
 There, I have met a community of smart, hard-working and innovative scholars that have in many ways contributed to this research.
-I would like to express my gratitude to Paul Dragos Aligica, Nigel Ashford, Peter Boettke, Don Boudreaux, Brandon Brice, John Buchmann, Jeff Carroll, Chris Coyne, Courtney Dunn, Bobbi Herzberg, Marcus Hunt, Nick Jacobs, Isaac Jilbert, Peter Lipsey, Ian Madison, Kelly Nelson, Gabriel Oliva, Raj Patel, Blaž Remic, Virgil Storr, Alienor van den Bosch, Richard Wagner, and Larry White for all the engaging conversations.
+I would like to express my gratitude to Paul Dragos Aligica, Nigel Ashford, Peter Boettke, Don Boudreaux, Brandon Brice, John Buchmann, Jeff Carroll, Chris Coyne, Courtney Dunn, Luke Foster, Bobbi Herzberg, Marcus Hunt, Nick Jacobs, Isaac Jilbert, Peter Lipsey, Ian Madison, Kelly Nelson, Gabriel Oliva, Raj Patel, Pablo Prieto, Blaž Remic, Virgil Storr, Alienor van den Bosch, Richard Wagner, and Larry White for all the engaging conversations.
 I have learned so much from you.
 
-Completing this work would have been all the more difficult were it not for the support and constructive criticism provided by my friends and colleagues.
-Specially, I would like to thank Veronica Adaeva, André Amaro, Aline Bagatin, Fábio Barros, Michel Hulmann, Guilherme Kerr, Maurício Pantaleão, Letícia Sene, and the participants of academic seminars at King's College London, the Mercatus Center at George Mason University, and the Prometheus Institute in Berlin.
-I am deeply indebted to them for their encouragement, insightful comments, and hard questions.
+Completing this work would have been all the more difficult were it not for the support and constructive criticism provided by my friends.
+Specially, I would like to thank Veronica Adaeva, André Amaro, Paulo Roberto Araujo, Aline Bagatin, Fábio Barros, Michel Hulmann, Guilherme Kerr, Maurício Pantaleão, Letícia Sene, and the participants of academic seminars at King's College London, the Mercatus Center at George Mason University, and the Prometheus Institute in Berlin.
+I am deeply indebted to them for their insightful comments and hard questions.
 
-I am deeply thankful to several present and past members of the Friedrich Naumann Foundation for Freedom.
-Many years ago, I was received a scholarship to attend a short course in Germany, and that first experience abroad transformed my life ever since.
-I then decided to pursue my post-graduate studies in Europe and it was one of the best choices I have made.
+I am also deeply thankful to several present and past members of the Friedrich Naumann Foundation for Freedom.
+Some years ago, I had received a scholarship to attend a short course in Germany, and that first experience abroad has transformed my life ever since.
 The present work is, to a large extent, a result of the push I received from the Naumann Foundation almost 10 years ago.
@@
 They have been absolutely wonderful over the years and have gone above and beyond to provide me with a good education.
-This dissertation is dedicated to them.
+This thesis is dedicated to them.
 \textit{Nós conseguimos}.
@@
 
+%-----------------------------------------------------
+% Chapter 4: Mass Killings
+%-----------------------------------------------------
+
+\input{chapters/04-killings}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-05-04 chapters/01-intro.tex
add chapter 4; small changes
appdx 3; uamk models

@@
 In order to clarify crucial aspects of my research topic, I employ an eclectic combination of research designs.
-The methods range from qualitative case studies to randomised experiments and machine learning algorithms.
+The methods range from qualitative case studies to machine learning algorithms.
 This diversity not only reflects the multiple aspects of organised violence, but it is also a pragmatic response to problems which are common in this area, such as incomplete data, reporting bias, and model uncertainty.
@@
 
-Brazil exemplifies many of the challenges of fighting violence in Latin America.
-Brazil has the highest absolute number of homicides in the world, about 56,000 per year, and the country hosts 19 of the 50 world's deadliest cities according to the above-mentioned ranking \citep{mapa2014, mexico2014,unodc2013}.
-Homicide rates have increased markedly after democratisation (1985), and whereas the country has tried several policies to reduce violence, the results are yet to be evaluated in a consistent fashion.
+Brazil exemplifies many of the challenges of fighting violence in developing nations.
+The country has the highest absolute number of homicides in the world, about 56,000 per year, and it hosts 19 of the 50 world's deadliest cities according to the above-mentioned ranking \citep{mapa2014, mexico2014,unodc2013}.
+Homicide rates have increased markedly after democratisation (1985), and whereas Brazil has tried several policies to reduce violence, the results are yet to be evaluated in a consistent fashion.
 
-The first chapter attempts to address this issue.
+The next chapter attempts to address this issue.
 Although Brazil remains notably affected by civil violence, the state of São Paulo has made significant inroads into fighting criminality.
@@
 
-The second chapter offers a rational choice account for Brazil's \textit{jogo do bicho}, or the `animal game', possibly the largest illegal gambling game in the world.
+Chapter three offers a rational choice account for Brazil's \textit{jogo do bicho}, or the `animal game', possibly the largest illegal gambling game in the world.
 The lottery has been running for over 120 years and according to estimations of Fundação Getúlio Vargas, a Brazilian think tank, it profits up to 800 million dollars per year.\footnote{See \url{http://goo.gl/9kNeX8} and \url{http://goo.gl/8FSAZl} (in Portuguese).
@@
 I also explain the emergence of the informal rules that govern the game as well as their enforcement mechanisms.
+
+The last chapter presents an empirical evaluation of explanations for genocides and politicides.
+Although the literature on state-sponsored killings has grown significantly over the last decades, it remains unclear whether estimates are robust to different model specifications, or which variables accurately predict the onset of large-scale violence.
+I employ extreme bounds analysis and distributed random forests to test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+The results show that GDP per capita, the post-Cold War period, and stable political regimes are negatively associated with mass killings.
+In contrast, ethnic diversity, civil wars, and previous political turmoils increase the risk of state-led violence.
+Years since the last episode of mass violence, GDP per capita, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to the models' out-of-sample predictive power.

### 2018-05-04 chapters/04-killings.tex
appdx 3; uamk models

@@
 Our analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
-We find that GDP \textit{per capita}, democratic regimes, and the post-Cold War period are negatively associated with mass killings, whereas previous political turmoil increases the risk of such violence.
-In contrast, the distributed random forest algorithm indicates that years since the last episode of mass violence, GDP \textit{per capita}, urban population, openness to trade, and the number of military personnel make the greatest contribution to our models’ out-of-sample predictive power.
+We find that GDP \textit{per capita}, stable political regimes, and the post-Cold War period are negatively associated with mass killings.
+Conversely, previous political turmoil, ethnic diversity, and civil wars increase the risk of such violence.
+The distributed random forest algorithm indicates that years since the last episode of mass violence, GDP \textit{per capita}, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to our models' out-of-sample predictive power.
 While we discuss the importance of these findings in more detail later in this article, a few notes are imperative to make here.
@@
 We employ two methods to test the robustness of the potential determinants of state-led violence.
-First, we use a variant of extreme bounds analysis (EBA) to check the statistical significance of 35 explanatory variables cited in the academic literature.\footnote{The complete list of variables included in this article is available in the online appendix.
+First, we use a variant of extreme bounds analysis (EBA) to check the statistical significance of 35 explanatory variables cited in the academic literature.\footnote{The complete list of variables included in this article is available in appendix \ref{sec:mk-appendix}.
 Our selection procedure was as follows.
@@
 Our main focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero.
-We do not assume that the CDF has a normal distribution: Many of our coefficients do not follow a normal distribution, so the generic model provides a better fit to our data.
-We specify our models as follows:
+We do not assume that the CDF has a normal distribution: Many of our coefficients do not follow a standard distribution, so the generic model provides a better fit to our data.\footnote{Histograms for all coefficients are available in appendix \ref{sec:mk-appendix}.} We specify our models as follows:
 
@@
 Lastly, we report the number of estimated regressions models which included each variable.
+
+\vspace{1cm}
 
@@
 Here, two variables meet our criteria for robustness, Post-Cold War years and conflicts with territorial aims, both negatively correlated with mass killing onset.
-When we limit our analysis to cases marked as civil wars by the Correlates of War project, years since the last mass killing episode, previous riots, former human rights abuses, and ethnic diversity appear have a positive effect on the outcome.
+When we limit our analysis to cases marked as civil wars by the Correlates of War project \citep{sarkees2010resort}, years since the last mass killing episode, previous riots, former human rights abuses, and ethnic diversity appear have a positive effect on the outcome.
 Rather surprisingly, the presence of militias has a negative effect on genocides and politicides.
-Finally, our model using ethnic civil war shows similar results: Militias have a negative impact on the likelihood of mass killings during ethnic conflict, as do conflicts over territorial aims.
+Finally, our model using ethnic civil war \citep{cederman2010ethnic} shows similar results: Militias have a negative impact on the likelihood of mass killings during ethnic conflict, as do conflicts over territorial aims.
 This finding is in contradiction with previous research \citep{koren2017means} and is deserving of further exploration.
@@
 
+\vspace{1cm}
+
 \begin{table}[H]
@@
 \textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
-\textit{Base variables} &  &  &  &  &  \\
-Log GDP per capita & -0.0091 & 0.0052 & 76.055 & 0.9335 & 226707 \\
+\textit{UCDP data} &  &  &  &  &  \\
+Territory aims & -0.044 & 0.019 & 74.997 & 0.9804 & 17902 \\
+Post-Cold War years & -0.038 & 0.019 & 66.574 & 0.9222 & 17902 \\
  &  &  &  &  &  \\
-\textit{Additional variables} &  &  &  &  &  \\
-Post-Cold War years & -0.0133 & 0.0085 & 72.845 & 0.9472 & 35614 \\
-UCDP civil war onset & 0.0529 & 0.0321 & 52.378 & 0.9441 & 20854 \\
-Previous riots & 0.0140 & 0.0100 & 56.242 & 0.9216 & 35614 \\
-UCDP ongoing civil war & 0.0172 & 0.0115 & 65.652 & 0.9092 & 20854 \\
-Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.674 & 0.9050 & 35614 \\
-Polity IV squared & -0.0002 & 0.0001 & 61.206 & 0.9031 & 35614 \\ \hline
+\textit{COW data} &  &  &  &  &  \\
+Physical integrity & 0.024 & 0.013 & 66.674 & 0.9564 & 17902 \\
+Militias & -0.099 & 0.048 & 73.104 & 0.9490 & 17902 \\
+Years since last mass killing & 0.006 & 0.002 & 88.208 & 0.9472 & 101583 \\
+Previous riots & 0.078 & 0.041 & 65.412 & 0.9348 & 17902 \\
+Ethnic diversity (ELF) & 0.095 & 0.062 & 48.615 & 0.9000 & 17902 \\
+ &  &  &  &  &  \\
+\textit{Cederman et al. data} &  &  &  &  &  \\
+Territory aims & -0.051 & 0.026 & 74.288 & 0.9167 & 17902 \\
+Militias & -0.050 & 0.035 & 52.240 & 0.9101 & 17902 \\ \hline
 \end{tabular}
-\caption{Extreme Bounds Analysis -- Mass Killings (robust variables only)}
-\label{tab:eba1}
+\caption{EBA -- Mass killings during civil wars (robust variables only)}
+\label{tab:ucdp}
 \end{table}
@@
 \includegraphics[width=\textwidth, height=9cm]{images/drfdpp4.pdf}
-\caption{ Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. data)}
+\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. data)}
 \label{fig:drfdpp4}
@@
 Finally, the machine learning models are comparable to the ones we present above, with a similar set of variables appearing in the random forest estimations.
-The complete results are available in the appendix.
+The results are available in appendix \ref{sec:mk-appendix}.
 
@@
 
-\subsubsection{Complete Results}
-
-Here we present a series of histograms with the distribution of coefficients of all variables of the main EBA model.
-There are 36 variables in total, of which seven are robust: Log GDP per capita, post-Cold War period, onset and ongoing civil wars (measured by the UCDP), previous riots, ethnic diversity and the squared term of the Polity IV index.
+\subsubsection{Main Model}
+
+Here we present a series of histograms with the coefficients' distribution of all variables in the main EBA model.
+There are 36 variables in total, seven of which are robust: Log GDP per capita, post-Cold War period, onset and ongoing civil wars (measured by the UCDP), previous riots, ethnic diversity and the squared term of the Polity IV index.
 
@@
 \end{tabular}
-\caption{Extreme Bounds Analysis -- Mass killings (robust variables only)}
-\label{tab:eba1}
+\caption{Extreme Bounds Analysis -- Mass killings}
+\label{tab:mk}
 \end{table}
@@
     \caption{Extreme Bounds Analysis -- Mass killings}
-    \label{fig:hist-mk}
+    \label{fig:mk}
 \end{sidewaysfigure}
@@
 
-\newpage
-
-%%%%% Add models during civil wars
+\subsection{Genocides during Civil Wars}
+\label{sec:civil-wars}
+
+Next, we discuss genocides that occur during wartime.
+We use three covariates that denote ongoing civil conflicts: one by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, another by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}.
+The significant variables in the models below is notably different from the previous ones.
+We see that as evidence that genocides that take place during conflicts follow a particular logic and are not caused by the same factors that foster mass killings in peacetime.
+
+\vspace{1cm}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{UCDP data} &  &  &  &  &  \\
+Territory aims & -0.044 & 0.019 & 74.997 & 0.9804 & 17902 \\
+Post-Cold War years & -0.038 & 0.019 & 66.574 & 0.9222 & 17902 \\
+ &  &  &  &  &  \\
+\textit{COW data} &  &  &  &  &  \\
+Physical integrity & 0.024 & 0.013 & 66.674 & 0.9564 & 17902 \\
+Militias & -0.099 & 0.048 & 73.104 & 0.9490 & 17902 \\
+Years since last mass killing & 0.006 & 0.002 & 88.208 & 0.9472 & 101583 \\
+Previous riots & 0.078 & 0.041 & 65.412 & 0.9348 & 17902 \\
+Ethnic diversity (ELF) & 0.095 & 0.062 & 48.615 & 0.9000 & 17902 \\
+ &  &  &  &  &  \\
+\textit{Cederman et al. data} &  &  &  &  &  \\
+Territory aims & -0.051 & 0.026 & 74.288 & 0.9167 & 17902 \\
+Militias & -0.050 & 0.035 & 52.240 & 0.9101 & 17902 \\ \hline
+\end{tabular}
+\caption{EBA -- Mass killings during civil wars}
+\label{tab:ucdp1}
+\end{table}
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/mk-ucdp.pdf}
+    \caption{EBA -- Mass killings during civil wars (UCDP data)}
+    \label{fig:mk-ucdp}
+\end{sidewaysfigure}
+\clearpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/mk-ucdp.pdf}
+    \caption{EBA -- Mass killings during civil wars (COW data)}
+    \label{fig:mk-cow}
+\end{sidewaysfigure}
+\clearpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/mk-ucdp.pdf}
+    \caption{EBA -- Mass killings during ethnic civil wars (Cederman et al. data)}
+    \label{fig:mk-eth}
+\end{sidewaysfigure}
+\clearpage
 
@@
 
-The following table shows the results of an EBA model using 3 variable combinations at a time. The results are very similar to those reported in the paper, in which we used 4 variables in each model.
-
-To speed up computational time, the models included in this appendix are based on 50,000 random draws from the full set of all possible regression models.
-\citet[819]{salaimartin2004determinants} argue that random sampling produces unbiased estimates of the regression coefficients, and as such can accelerate convergence of the results.
-The models presented in the paper, however, include the full set of regressions.
+The following table shows the results of an EBA model using 3 variable combinations per model. The results are very similar to those reported in the paper, in which we used 4 variables in each model.
+
+The models below are based on 50,000 random draws from the full set of all possible regression models.
+\citet[819]{salaimartin2004determinants} argue that random sampling produces unbiased estimates of the regression coefficients with low computational time.
+The models presented in the manuscript, however, include the full set of possible regressions.
+
+\vspace{1cm}
 
@@
     \caption{EBA -- 3 variables in each regression model}
-    \label{fig:hist-mk-3vars}
+    \label{fig:mk-3vars}
 \end{sidewaysfigure}
@@
 In contrast with the main EBA model, the indicators of UCDP ongoing civil wars, ethnic diversity, and Polity IV square drop out of significance.
-Their CDF(0) is 0.88, just marginally below our threshold of 0.9.
+Their individual CDFs(0) are about 0.88, just marginally below our specified threshold of 0.9.
 
@@
     \caption{EBA -- 5 variables in each regression model}
-    \label{fig:hist-mk-5vars}
+    \label{fig:mk-5vars}
 \end{sidewaysfigure}
@@
 There is no standard definition about what constitutes an acceptable VIF value, although researchers often use a value of 10 as rule of thumb to indicate strong multicollinearity \citep[674]{o2007caution}.
-Our original model used 7 as a cutoff, and here we employ 10 (less strict), a lower value of 2.5 (more conservative), and a model without any VIF value (no restriction).
+Our original model used a slightly more conservative value of 7 as a cutoff.
+Here we test the same model with VIF $=$ 10 (less strict), 2.5 (more conservative), and a model without any VIF value (no restriction).
 The results are essentially identical to those in the main model.
@@
 \end{tabular}
-\caption{EBA -- VIF 10 (less strict)}
-\label{tab:eba1}
+\caption{EBA -- VIF 10}
+\label{tab:mk-high-vif}
 \end{table}
@@
     \includegraphics[width=\textwidth]{images/mk-high-vif.pdf}
-    \caption{EBA -- VIF 10 (less strict)}
-    \label{fig:hist-mk-high-vif}
+    \caption{EBA -- VIF 10}
+    \label{fig:mk-high-vif}
 \end{sidewaysfigure}
@@
 \end{tabular}
-\caption{EBA -- VIF 2.5 (more conservative)}
+\caption{EBA -- VIF 2.5}
+\label{tab:low-vif}
+\end{table}
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/mk-high-vif.pdf}
+    \caption{EBA -- VIF 2.5}
+    \label{fig:mk-low-vif}
+\end{sidewaysfigure}
+\clearpage
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{Base variables} &  &  &  &  &  \\
+Log GDP per capita & -0.0091 & 0.0052 & 75.940 & 0.9343 & 50000 \\
+ &  &  &  &  &  \\
+\textit{Additional variables} &  &  &  &  &  \\
+Post-Cold War years & -0.0133 & 0.0085 & 72.756 & 0.9469 & 7800 \\
+UCDP civil war onset & 0.0531 & 0.0321 & 53.068 & 0.9452 & 4596 \\
+Previous riots & 0.0140 & 0.0101 & 56.139 & 0.9200 & 7811 \\
+UCDP ongoing civil war & 0.0170 & 0.0116 & 64.487 & 0.9057 & 4497 \\
+Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.814 & 0.9056 & 7808 \\
+Polity IV squared & -0.0002 & 0.0001 & 60.825 & 0.9009 & 7903 \\ \hline
+\end{tabular}
+\caption{EBA -- No VIF Restriction}
+\label{tab:mk-no-vif}
+\end{table}
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/mk-no-vif.pdf}
+    \caption{EBA -- No VIF restriction}
+    \label{fig:mk-no-vif}
+\end{sidewaysfigure}
+\clearpage
+
+\subsubsection{Generalised Linear Models}
+
+We estimate the same EBA with logit and probit in addition to the linear probability models reported above.
+Logistic and probit regressions may have issues of complete separation, that is, when some covariates perfectly separates zeros and ones in the outcome variable.\footnote{We thank Mark Bell for sharing \texttt{R} code to estimate penalised-likelihood models.} In that case, the estimations fail to converge.
+We addressed this problem by adding a weak prior to the regression coefficients as suggested by \citet{gelman2008weakly}.
+We first scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then we added a Cauchy distribution with centre 0 and scale 2.5.
+The probit regressions use a scale of $2.5 \times 1.6$, which is also recommended by the authors \citep{arm2017rpackage}.
+Ethnic diversity and ongoing civil wars come close to meeting our threshold values (0.88 and 0.84, respectively), and civil war onset (UCDP) has a higher percentage of significant coefficients and a high CDF(0) area than in the linear probability models.
+
+\vspace{1cm}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{Base variables} &  &  &  &  &  \\
+Log GDP per capita & 0.434 & 0.223 & 75.570 & 0.9267 & 50000 \\
+ &  &  &  &  &  \\
+\textit{Additional variables} &  &  &  &  &  \\
+UCDP civil war onset & 1.308 & 0.530 & 87.261 & 0.9742 & 4506 \\
+Post-Cold War years & -0.911 & 0.428 & 70.456 & 0.9448 & 7890 \\
+Previous riots & 0.744 & 0.38 & 66.778 & 0.9383 & 7805 \\
+Polity IV squared & -0.015 & 0.008 & 68.038 & 0.9285 & 7975 \\ \hline
+\end{tabular}
+\caption{EBA -- Logistic model}
+\label{tab:mk-logit}
+\end{table}
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/mk-logit.pdf}
+    \caption{EBA -- Logistic model}
+    \label{fig:mk-logit}
+\end{sidewaysfigure}
+\clearpage
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{Base variables} &  &  &  &  &  \\
+Log GDP per capita & -0.1924 & 0.1031 & 76.118 & 0.9258 & 50000 \\
+ &  &  &  &  &  \\
+\textit{Additional variables} &  &  &  &  &  \\
+UCDP civil war onset & 0.6422 & 0.2582 & 89.225 & 0.9772 & 4501 \\
+Previous riots & 0.3367 & 0.1743 & 71.813 & 0.9436 & 7851 \\
+Post-Cold War years & -0.3709 & 0.1830 & 71.465 & 0.9404 & 7836 \\
+Polity IV squared & -0.0061 & 0.0032 & 70.155 & 0.9315 & 7931 \\ \hline
+\end{tabular}
+\caption{EBA -- Probit model}
 \label{tab:eba1}
@@
     \centering
-    \includegraphics[width=\textwidth]{images/mk-high-vif.pdf}
-    \caption{EBA -- VIF 2.5 (more conservative)}
-    \label{fig:hist-mk-high-vif}
+    \includegraphics[width=\textwidth]{images/mk-probit.pdf}
+    \caption{EBA -- Probit model}
+    \label{fig:mk-probit}
 \end{sidewaysfigure}
@@
 
-\newpage
-
-\subsubsection{Generalised Linear Models}
-
-We estimate the same EBA with logit and probit in addition to the linear probability model reported above. The results are
-
 \subsection{Random Forest Extensions}
@@
 
+In the manuscript, we perform a grid search to optimise the hyperparameters of the random forest models.
+Because the grid search evaluates a wide range of parameter values at once, it is often unnecessary to run additional tests to assess the robustness of the results.
+Nevertheless, as random forests themselves are an approximation to a myriad of possible parameter combinations, changes in seed numbers may influence model output.
+Thus, we start the models with different random seed numbers to evaluate how study are our original results.\footnote{The numbers were generated at \href{https://www.random.org/}{https://www.random.org/}.} The findings holds quite well: Although variable importance chances from one model to another, the most significant variables appear repeatedly in the estimations.
+The marginal plots also show that their effects on the outcome variable are similar despite the nonlinearities.
+Below we show the 6 most significant variables and their respective partial dependence plots.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics{images/drf-mk.pdf}
+    \caption{Distributed Random Forest -- Variable importance}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp.pdf}
+    \caption{Distributed Random Forest -- Partial plots}
+    \label{fig:my_label}
+\end{figure}
+
+\subsubsection{Alternative Random Seeds}
+
+We tested the same model using different random seeds.
+Although not identical, the results bear strong similarities with the previous random forest estimation.
+A certain amount of variation is expected in machine learning algorithms; each model produces slightly different results as they are sensitive to the choice of seed.
+But as we see below, the results are broadly comparable, what confirms the robustness of the previous findings.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics{images/drf-mk2.pdf}
+    \caption{Variable importance -- Seed 44849999}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp2.pdf}
+    \caption{Partial plots -- Seed 44849999}
+    \label{fig:my_label}
+\end{figure}
+
+\newpage
+
+Third model -- seed 1502436:
+
+\begin{figure}[H]
+    \centering
+    \includegraphics{images/drf-mk3.pdf}
+    \caption{Variable importance -- Seed 1502436}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp3a.pdf}
+    \caption{Partial plots -- Seed 1502436}
+    \label{fig:my_label}
+\end{figure}
+
+\newpage
+
+\subsection{Genocide/Politicide Onset}
+\label{sec:mk-other-variable}
+
+In this section, we evaluate the models presented above with a measure of genocide and politicide by \citet{harff2003no}. The results show important contrasts with the previous analyses.
+
+\newpage
+\subsubsection{Distributed Random Forests}
+
+Lastly, we present below three models using the distributed random forest algorithm \citep{h2o2017}.
+The results are in line with those obtained with the mass killing variable \citep{ulfelder2008assessing}.
+Again we see that CINC, the percentage of urban population, and variables concerning the military are some of the most important predictors of state-led violence.
+According to this measure, poor countries are at a much higher risk of genocide that its counterparts.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics{images/drf-gp.pdf}
+    \caption{Variable importance -- Genocide/Politicide}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp4a.pdf}
+    \caption{Partial plots -- Genocide/Politicide}
+    \label{fig:my_label}
+\end{figure}
+
+\newpage
+
+The next graphs show the most important predictors of genocides that occur during civil wars and their respective partial dependence plots.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics{images/drf-gp1.pdf}
+    \caption{Variable importance -- Genocide/Politicide (UCDP data)}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp5a.pdf}
+    \caption{Partial plots -- Genocide/Politicide (UCDP data)}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics{images/drf-gp2.pdf}
+    \caption{Variable importance -- Genocide/Politicide (COW data)}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp6a.pdf}
+    \caption{Partial plots -- Genocide/Politicide (COW data)}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics{images/drf-gp3.pdf}
+    \caption{Variable importance -- Genocide/Politicide (Cederman et al. data)}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp7a.pdf}
+    \caption{Partial plots -- Genocide/Politicide (Cederman et al. data)}
+    \label{fig:my_label}
+\end{figure}
+
+\newpage
+
 \subsection{\texttt{R} Code}
@@
 
-The code below reproduces the analyses included here.
+The \texttt{R} code below reproduces the analyses included in the manuscript.
 
@@
 
+## Install and load required packages 
+if (!require("tidyverse")) {
+        install.packages("tidyverse")
+}
+if (!require("data.table")) {
+        install.packages("data.table")
+}
+if (!require("ExtremeBounds")) {
+        install.packages("ExtremeBounds")
+}
+if (!require("h2o")) {
+        install.packages("h2o")
+}
+if (!require("sandwich")) {
+        install.packages("sandwich")
+}
+if (!require("arm")) {
+        install.packages("arm")
+}
+if (!require("stargazer")) {
+        install.packages("stargazer")
+}
+
+## Load dataset
+df <- haven::read_dta("data/base variables.dta") %>% setDT()
+
+## Select and lag variables
+sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
+             "COWcivilwarongoing", "ethnowarstart", "ethnowarongoing",
+             "assdummy", "demdummy", "elf", "lmtnest", "pop", "realgdp",
+             "rgdppc", "polity2", "exclpop", "discpop", "polrqnew",
+             "poltrqnew", "egiptpolrqnew", "egippolrqnew", "discrim",
+             "elf2", "interstatewar", "milex", "milper", "percentpopurban",
+             "postcoldwar", "coupdummy", "riotdummy", "territoryaims",
+             "totaltrade", "tradedependence", "militias", "physint", "cinc",
+             "totalbeaths", "change", "guerrilladummy", "sf", "regtrans")
+
+df1 <- cbind(df, df[, shift(.SD, 1, give.names = TRUE),
+                    by = ccode, .SDcols = sd.cols])
+
+# Remove the second `ccode` variable
+df1 <- as.data.frame(df1[, -c(70)])
+
+# Add new variables
+df1$logrgdppc_lag_1 <- log(df1$rgdppc_lag_1)
+df1$polity2sq_lag_1 <- df1$polity2_lag_1^2
+
+# UCDP civil war == 1
+df.ucdp <- df1 %>% filter(UCDPcivilwarongoing == 1)
+df.ucdp <- as.data.frame(df.ucdp[, c(1:7, 76:111)])
+names(df.ucdp) <- sub("_.*","", names(df.ucdp))
+
+# COW civil war == 1
+df.cow <- df1 %>% filter(COWcivilwarongoing == 1)
+df.cow <- as.data.frame(df.cow[, c(1:7, 76:111)])
+names(df.cow) <- sub("_.*","", names(df.cow))
+
+# Ethnic civil war == 1
+df.eth <- df1 %>% filter(ethnowarongoing == 1)
+df.eth <- as.data.frame(df.eth[, c(1:7, 76:111)])
+names(df.eth) <- sub("_.*","", names(df.eth))
+
+# Regular model
+df2 <- as.data.frame(df1[, c(1:7, 70:111)])
+names(df2) <- sub("_.*","", names(df2))
+
+###############################
+### Extreme Bounds Analyses ###
+###############################
+
+## Classifying a few variables as mutually exclusive.
+## "Change" was removed because it was correlated at 0.99 with "regtrans". 
+free.variables <- c("logrgdppc", "polity2", "mksyr")
+civilwar.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart")
+doubtful.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart", "assdummy",
+                        "totaltrade", "tradedependence", "milper", "milex",
+                        "pop", "totalbeaths", "guerrilladummy", "regtrans",
+                        "riotdummy", "territoryaims", "militias",
+                        "physint", "percentpopurban", "coupdummy",
+                        "postcoldwar",  "lmtnest", "realgdp", "discrim",
+                        "exclpop", "discpop", "elf",  "polrqnew",
+                        "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                        "polity2sq")
+
+# Cluster-robust standard errors
+se.clustered.robust <- function(model.object){
+        model.fit <- vcovHC(model.object, type = "HC", cluster = "country")
+        out <- sqrt(diag(model.fit))
+        return(out)
+}
+
+# Main model
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, vif = 7, level = 0.9,
+          se.fun = se.clustered.robust)
+
+summary(m1)
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
+                       "UCDPcivilwarongoing",
+                       "UCDPcivilwarstart", "COWcivilwarongoing",
+                       "COWcivilwarstart", "ethnowarongoing", "ethnowarstart",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
+              "UCDP ongoing", "UCDP onset", "COW ongoing", "COW onset", 
+              "Ethnic ongoing", "Ethnic onset", "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+# Mass killings during civil war
+# UCDP civil conflicts == 1
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop",
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "MKstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.ucdp, vif = 7,
+          level = 0.9, se.fun = se.clustered.robust)
+
+summary(m1)
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
+              "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+# COW civil wars == 1
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop",
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "MKstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.cow, vif = 7,
+          level = 0.9, se.fun = se.clustered.robust)
+
+summary(m1)
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
+              "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+# Ethnic civil war == 1
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop", 
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "MKstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.eth, vif = 7,
+          level = 0.9, se.fun = se.clustered.robust)
+
+summary(m1)
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
+              "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+## Different values of k
+## Code for the histogram not included as it is the same as that of the main model.
+
+# 3 variables per model
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:3,
+          data = df2, vif = 7, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+
+# 5 variables per model
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:5,
+          data = df2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+
+## Alternative VIFs
+
+# VIF = 10
+# Low VIF
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, vif = 10, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+
+# VIF = 2.5
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, vif = 2.5, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+
+# No VIF
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+
+## Generalised linear models
+
+# Logit
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, level = 0.9, vif = 7, draws = 50000,
+          reg.fun = bayesglm, family = binomial(link = "logit"))
+
+# Probit
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, level = 0.9, vif = 7, draws = 50000,
+          reg.fun = bayesglm, family = binomial(link = "probit"))
+
+##################################
+### Distributed Random Forests ###
+##################################
+
+## Random Forests
+
+## Prepare the data set
+h2o.init(nthreads = -1, max_mem_size = "20g") # memory size
+
+df2a <- as.h2o(df2)
+
+df2a$MKstart <- as.factor(df2a$MKstart)  # encode the binary response as a factor
+h2o.levels(df2a$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df2), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf01",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf01",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+# Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf01_model_21")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Polity IV", 
+                      "Military personnel",
+                      "Ethnic polarisation", 
+                      "% Urban pop.",
+                      "Years mass killing",
+                      "Log GDP per capita"),
+        main = "")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
+p1 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
+p2 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Years since mass killing") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p3 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + ylim(0, 0.05) + xlab("% Urban") + ylab("Mean response")
+
+egiptpolrqnew <- h2o.partialPlot(object = a, data = train, cols = c("egiptpolrqnew"))
+p4 <- qplot(egiptpolrqnew$egiptpolrqnew, egiptpolrqnew$mean_response) + geom_line() +
+        theme_classic() + ylim(0, 0.05) + xlab("Ethnic polarisation") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p5 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Military personnel") + ylab("Mean response")
+
+polity2 <- h2o.partialPlot(object = a, data = train, cols = c("polity2"))
+p6 <- qplot(polity2$polity2, polity2$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Polity IV") + ylab("Mean response")
+
+# Multiplot function: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+        library(grid)
+
+        # Make a list from the ... arguments and plotlist
+        plots <- c(list(...), plotlist)
+
+        numPlots = length(plots)
+
+        # If layout is NULL, then use 'cols' to determine layout
+        if (is.null(layout)) {
+                # Make the panel
+                # ncol: Number of columns of plots
+                # nrow: Number of rows needed, calculated from # of cols
+                layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                                 ncol = cols, nrow = ceiling(numPlots/cols))
+        }
+
+        if (numPlots==1) {
+                print(plots[[1]])
+
+        } else {
+                # Set up the page
+                grid.newpage()
+                pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+
+                # Make each plot, in the correct location
+                for (i in 1:numPlots) {
+                        # Get the i,j matrix positions of the regions that contain this subplot
+                        matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+
+                        print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                                        layout.pos.col = matchidx$col))
+                }
+        }
+}
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+## Different seeds
+
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf01b",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 44849999))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf01b",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+# Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf01b_model_8")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Population", 
+                      "Military personnel",
+                      "Trade dependence", 
+                      "Log GDP per capita",
+                      "% Urban pop.",
+                      "Years mass killing"),
+        main = "")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
+p1 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Years since mass killing") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + ylim(0, 0.05) + xlab("% Urban") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
+p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
+p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+        theme_classic() + ylim(0, 0.05) + xlab("Trade dependence") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p5 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Military personnel") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"))
+p6 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Population") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf01c",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 1502436))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf01c",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+# Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf01c_model_58")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Population", 
+                      "Military personnel",
+                      "Trade dependence", 
+                      "Log GDP per capita",
+                      "% Urban pop.",
+                      "Years mass killing"),
+        main = "")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
+p1 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Years since mass killing") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + ylim(0, 0.05) + xlab("% Urban") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
+p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
+p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+        theme_classic() + ylim(0, 0.05) + xlab("Trade dependence") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p5 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Military personnel") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"))
+p6 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
+        xlab("Population) + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+## Genocides in civil wars
+
+## UCDP data
+df.ucdpa <- as.h2o(df.ucdp)
+
+df.ucdpa$MKstart <- as.factor(df.ucdpa$MKstart)  # encode the binary response as a factor
+h2o.levels(df.ucdpa$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.ucdpa, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf02",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf02",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+# Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf02_model_34")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Total battle deaths", 
+                      "Log GDP per capita",
+                       "Trade dependence",
+                      "Military personnel",
+                      "% Urban pop.",
+                      "Years mass killing"),
+        main = "")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
+p1 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + 
+        ylim(0, 0.1) + xlab("Years since mass killing") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        ylim(0, 0.1) + theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p3 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+        ylim(0, 0.25) + xlab("Military personnel") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
+p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() + theme_classic() +
+        ylim(0, 0.1) + xlab("Trade dependence") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
+p5 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+        ylim(0, 0.1) + xlab("Log GDP per capita") + ylab("Mean response")
+
+totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"))
+p6 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() + theme_classic() +
+        ylim(0, 0.35) + xlab("Total battle deaths") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+## COW data
+df.cowa <- as.h2o(df.cow)
+
+df.cowa$MKstart <- as.factor(df.cowa$MKstart)
+h2o.levels(df.cowa$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.cowa, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
+                               "mksyr2", "mksyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf03",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf03",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## Graphs
+a <- h2o.loadModel("/root/Documents/mk/gridrf03_model_3")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6)) 
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Total battle deaths", 
+                      "Excluded population",
+                      "Yrs since mass killing",
+                      "Log GDP per capita",
+                      "Ethnic polarisation",
+                      "Physical integrity"),
+        main = "")
+
+physint <- h2o.partialPlot(object = a, data = train, cols = c("physint"))
+p1 <- qplot(physint$physint, physint$mean_response) + geom_line() + theme_classic() + 
+        xlab("Physical integrity") +  ylab("Mean response")
+
+egiptpolrqnew <- h2o.partialPlot(object = a, data = train, cols = c("egiptpolrqnew"))
+p2 <- qplot(egiptpolrqnew$egiptpolrqnew, egiptpolrqnew$mean_response) + geom_line() +
+        theme_classic() + xlab("Ethnic polarisation") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
+p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+        ylim(0, 0.1) + xlab("Log GDP per capita") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
+p4 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + 
+        ylim(0, 0.1) + xlab("Years since mass killing") +  ylab("Mean response")
+
+exclpop <- h2o.partialPlot(object = a, data = train, cols = c("exclpop"))
+p5 <- qplot(exclpop$exclpop, exclpop$mean_response) + geom_line() + theme_classic() +
+        ylim(0, 0.1) + xlab("Excluded population") + ylab("Mean response")
+
+totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"))
+p6 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() + theme_classic() +
+        ylim(0, 0.1) + xlab("Total battle deaths") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+## Ethnic war
+df.etha <- as.h2o(df.eth)
+
+df.etha$MKstart <- as.factor(df.etha$MKstart) 
+h2o.levels(df.etha$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.etha, 
+                         ratios = c(0.7, 0.15), 
+                         seed = 42)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.eth), c(y, "ccode", "year", "rgdppc",
+                               "mksyr2", "mksyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf04",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf04",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+# Graphs
+a <- h2o.loadModel("/root/Documents/mk/gridrf04_model_14")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Military expenditure",
+                      "Total trade", 
+                      "% Urban",
+                      "Trade dependence",
+                      "Military personnel",
+                      "CINC"),
+        main = "")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"))
+p1 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
+        xlab("CINC") +  ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p2 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+        theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
+p3 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() + theme_classic() +
+        xlab("Trade dependence") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p4 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+       theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"))
+p5 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() +
+        theme_classic() + xlab("Total trade") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
+p6 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+####################################
+### Genocide/Politicide Variable ###
+####################################
+
+## The code below replicates the same analyses presented above
+## but using a measure of genocide/politicide coded by Harff (2003).
+
+## Data wrangling
+df3 <- haven::read_dta("data/uamkstart.dta") %>% setDT()
+
+sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
+             "COWcivilwarongoing", "ethnowarstart", "ethnowarongoing",
+             "assdummy", "demdummy", "elf", "lmtnest", "pop", "realgdp",
+             "rgdppc", "polity2", "exclpop", "discpop", "polrqnew",
+             "poltrqnew", "egiptpolrqnew", "egippolrqnew", "discrim",
+             "elf2", "interstatewar", "milex", "milper", "percentpopurban",
+             "postcoldwar", "coupdummy", "riotdummy", "territoryaims",
+             "totaltrade", "tradedependence", "militias", "physint", "cinc",
+             "totalbeaths", "change", "guerrilladummy", "sf", "regtrans")
+
+df4 <- cbind(df3, df3[, shift(.SD, 1, give.names = TRUE),
+                      by = ccode, .SDcols = sd.cols])
+
+# Remove the second `ccode` variable
+df4 <- as.data.frame(df4[, -c(75)])
+
+# Add new variables
+df4$logrgdppc_lag_1 <- log(df4$rgdppc_lag_1)
+df4$polity2sq_lag_1 <- df4$polity2_lag_1^2
+
+# Renaming variables
+df5 <- as.data.frame(df4[, c(1:4, 72:116)])
+names(df5) <- sub("_.*","", names(df5))
+
+#################################################
+### Distributed Random Forests - Harff (2003) ###
+#################################################
+
+df5a <- as.h2o(df5)
+
+df5a$uamkstart <- as.factor(df5a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df5a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df5a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df5), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq")) 
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, 
+               grid_id = "gridrf05",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf05",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf05_model_27")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Real GDP", 
+                      "Military expenditure",
+                      "Total trade", 
+                      "Military personnel",
+                      "CINC", 
+                      "% Urban"),
+        main = "")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p1 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"))
+p2 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
+        xlab("CINC") +  ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p3 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+        theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"))
+p4 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+        xlab("Total trade") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
+p5 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"))
+p6 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() +
+        theme_classic() + xlab("Real GDP") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+## UCDP == 1
+df.ucdp2 <- df5 %>% filter(UCDPcivilwarongoing == 1)
+df.ucdp2a <- as.h2o(df.ucdp2)
+
+df.ucdp2a$uamkstart <- as.factor(df.ucdp2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.ucdp2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.ucdp2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.ucdp2), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf06",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf06",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf06_model_43")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Excluded population",
+                      "% Urban pop.", 
+                      "Trade dependence",
+                      "Log GDP per capita",
+                      "Years since genocide", 
+                      "Military personnel"),
+        main = "")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p1 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+        theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+uamkyr <- h2o.partialPlot(object = a, data = train, cols = c("uamkyr"))
+p2 <- qplot(uamkyr$uamkyr, uamkyr$mean_response) + geom_line() + theme_classic() + 
+        xlab("Years since genocide") +  ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
+p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
+p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+        theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p5 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+exclpop <- h2o.partialPlot(object = a, data = train, cols = c("exclpop"))
+p6 <- qplot(exclpop$exclpop, exclpop$mean_response) + geom_line() +
+        theme_classic() + xlab("Excluded population") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+## COW == 1
+df.cow2 <- df5 %>% filter(COWcivilwarongoing == 1)
+
+df.cow2a$uamkstart <- as.factor(df.cow2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.cow2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.cow2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.cow2), c(y, "ccode", "year", "rgdppc",
+                               "uamkyr2", "uamkyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+## Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf07",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf07",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf07_model_27")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Military expenditure",
+                      "% Urban pop.", 
+                      "Total trade",
+                      "Years since genocide", 
+                      "Military personnel", 
+                      "Trade dependence"),
+        main = "")
+
+uamkyr <- h2o.partialPlot(object = a, data = train, cols = c("uamkyr"))
+p3 <- qplot(uamkyr$uamkyr, uamkyr$mean_response) + geom_line() + theme_classic() + 
+        xlab("Years since genocide") +  ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p2 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+        theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"))
+p4 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+        xlab("Total trade") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p5 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
+p1 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+        theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
+p6 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+## ETHONSET == 1
+df.eth2 <- df5 %>% filter(ethnowarongoing == 1)
+
+df.eth2a <- as.h2o(df.eth2)
+
+df.eth2a$uamkstart <- as.factor(df.eth2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.eth2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.eth2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.eth2), c(y, "ccode", "year", "rgdppc",
+                               "uamkyr2", "uamkyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf08",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 1000, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf08",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## Graphs
+a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf08_model_55")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[6:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Total battle deaths", 
+                      "Military personnel",
+                      "Log GDP per capita", 
+                      "CINC",
+                      "% Urban",
+                      "Military expenditure"),
+        main = "")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"))
+p3 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
+        xlab("CINC") +  ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
+p5 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+        theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
+p4 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
+p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"))
+p6 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() +
+        theme_classic() + xlab("Total battle deaths") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
+p1 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
 \end{verbatim}

### 2018-05-04 dissertation.tex
appdx 3; uamk models

@@
 	breaklinks=true,
-	linkcolor=black,
+	linkcolor=Mahogany,
 	citecolor=Mahogany,
@@
 
-In the third essay, co-authored with Gary Uzonyi, we employ extreme bounds analysis and distributed random forests to identify the key determinants of state-sponsored violence.
-We test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
-The results show that GDP per capita, democracy, and the post-Cold War period are negatively associated with mass killings, whereas ethnic diversity, civil wars, and previous political turmoils increase the risk of state-led violence.
-GDP per capita, years since the last episode of mass violence, urban population, ethnic polarisation, and the number of military personnel make the greatest contribution to the models' out-of-sample predictive power.
+In the third essay, I employ extreme bounds analysis and distributed random forests to identify the key determinants of state-sponsored violence.
+Although the literature on mass killings has grown significantly over the last decades, it remains unclear whether estimates are robust to different model specifications or which variables accurately predict the onset of genocides and politicides.
+To address this issue, I test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+The results show that GDP per capita, the post-Cold War period, and stable political regimes are negatively associated with mass killings.
+In contrast, ethnic diversity, civil wars, and previous political turmoils increase the risk of state-led violence.
+Mass killings that occur during civil wars are correlated with a different set of variables, and I find that territorial conflicts and the use of militias are negatively associated with the outcome.
+The random forests estimations indicate that years since the last episode of mass violence, GDP per capita, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to the models' out-of-sample predictive power.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 I would like to extend thanks to the many people, in many countries, who so generously contributed to the work presented in this thesis.
-First and foremost, I want to thank my supervisor, Professor David Skarbek.
-It was a great privilege and honour to work and study under his guidance, and he has taught me more than I could ever give him credit for here.
+First and foremost, I want to thank my supervisor, Professor David Skarbek, for the guidance and encouragement he provided me.
 David was not only an excellent supervisor but also a great friend; his empathy, enthusiasm, and sense of humour made my stay at King's College a very pleasant one.
-I simply could not have wished for a better mentor for my PhD studies.
+I could not have wished for a better mentor for my PhD studies.
 Thank you very much.
@@
 His help in the last semester of my PhD studies was invaluable and I am very thankful for his generosity.
+
+I am heartily grateful to my examiners, Professor Toke Aidt and Professor Graham Denyer Willis, who accepted to dedicate their precious time to evaluate this work.
 
@@
 \newpage
+{\hypersetup{linkcolor=black}
 \pdfbookmark[0]{Contents}{contents_bookmark}
@@
 \addcontentsline{toc}{chapter}{List of Figures}
+}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-05-07 chapters/04-killings.tex
beta version :)

@@
 Our analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
-We find that GDP \textit{per capita}, stable political regimes, and the post-Cold War period are negatively associated with mass killings.
+We find that GDP per capita, stable political regimes, and the post-Cold War period are negatively associated with mass killings.
 Conversely, previous political turmoil, ethnic diversity, and civil wars increase the risk of such violence.
-The distributed random forest algorithm indicates that years since the last episode of mass violence, GDP \textit{per capita}, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to our models' out-of-sample predictive power.
+The distributed random forest algorithm indicates that years since the last episode of mass violence, GDP per capita, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to our models' out-of-sample predictive power.
 While we discuss the importance of these findings in more detail later in this article, a few notes are imperative to make here.
@@
 $M$ is a set of three covariates that are included in every model due to their prominence in the literature \citep{levine1992vale}.
-In our analysis, we have added the natural logarithm of the GDP \textit{per capita} to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence.
+In our analysis, we have added the natural logarithm of the GDP per capita to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence.
 $F$ denotes a vector of variables of interest, and $Z$ is a vector of other control variables in addition to those included in $M$.
@@
 
+This appendix contains all required information to replicate the numerical analyses presented in sections \ref{sec:results4} and \ref{sec:additional-tests4}.
+\textt{R} code can be found in subsection \ref{sec:mk-code} and the data are available on the following GitHub repository: \href{https://github.com/danilofreire/mass-killings}{https://github.com/danilofreire/mass-killings}.
+We used \texttt{R} version 3.4.4 (15-03-2018) and Ubuntu 16.04.4 LTS to perform all statistical calculations.
+
 \subsection{Variable Selection}
@@
 
-This appendix contains all required information to replicate the numerical analyses presented in the paper.
-We present a series of histograms with the coefficient distribution of every variable included in our EBA models.
-Robust variables  (CDF(0) $> 0.9$) are described in further detail in the tables below.
-
-The complete results can also be retrieved from data included at the following GitHub repository: \href{https://github.com/danilofreire/mass-killings}{https://github.com/danilofreire/mass-killings}.
-\textt{R} code is available in section \ref{sec:mk-code}.
-The analyses were conducted on \texttt{R} version 3.4.4 (15-03-2018) and Ubuntu 16.04.4 LTS.
+We employ some criteria to select our explanatory variables.
+First, we included only published articles in our sample.
+Although working papers and policy may also provide important insights about the onset of mass killings, we believe that peer-reviewed research is probably better suited for our purposes.
+Also, we included only papers that use regression methods on a global sample and were published from 1995 to 2015.
+Our final sample comprises 45 articles: \citet{anderton2015new}, \citet{balcells2010rivalry, balcells2011continuation}, \citet{besanccon2005relative}, \citet{bulutgil2015social}, \citet{bundervoet2009livestock}, \citet{clayton2016civilianizing}, \citet{colaresi2008kill}, \citet{downes2006desperate, downes2007restraint},  \citet{easterly2006development}, \citet{eck2007one}, \citet{esteban2015strategic}, \citet{fazal2015particular}, \citet{fjelde2014weakening}, \citet{goldsmith2013forecasting}, \citet{harff2003no}, \citet{joshi2017kills}, \citet{kim2010makes}, \citet{kim2016revolutionary}, \citet{kisangani2007political}, \citet{koren2017means}, \citet{krain1997state}, \citet{manekin2013violence}, \citet{mcdoom2013killed,mcdoom2014predicting}, \citet{melander2009new}, \citet{montalvo2008discrete}, \citet{pilster2016differentiation}, \citet{querido2009state}, \citet{raleigh2012violence}, \citet{rost2013will}, \citet{rummel1995democracy}, \citet{schneider2013accounting}, \citet{siroky2015empire}, \citet{stanton2015regulating}, \citet{sullivan2012blood}, \citet{tir2008domestic}, \citet{ulfelder2008assessing}, \citet{ulfelder2012forecasting}, \citet{uzonyi2015civil, uzonyi2016domestic} \citet{valentino2004draining}, \citet{valentino2006covenants}, \citet{verpoorten2012leave}, \citet{wayman2010explaining}, \citet{wig2016local}, and \citet{yanagizawa2014propaganda}.
+
+We find that in those 45 studies scholars made use of nearly 180 measurements to capture roughly 30 key concepts related to threat and costs of mass killings.
+To be added to our models, a variable should appear in at least two articles.
+The covariates are summarised in table \ref{tab:mk-vs}.
+A complete list of variables is available at \href{https://github.com/danilofreire/mass-killings}{https://github.com/danilofreire/mass-killings}.
+
+\begin{table}[!htbp] \centering 
+  \caption{Independent Variables} 
+  \label{tab:mk-vs} 
+\footnotesize
+\begin{tabular}{@{\extracolsep{5pt}}lcc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] {Variable} & \multicolumn{1}{c}{Coded} & \multicolumn{1}{c}{Source}\\ 
+\hline \\[-1.8ex] 
+Assassination & Dichotomous & \citet{banks1999cross} \\ 
+CINC & Continuous & \citet{cow2017cinc}\\ 
+Coup d'état & Dichotomous & \citet{marshall2017pitf}  \\ 
+COW civil war onset & Dichotomous & \citet{cow2017cinc,singer1988reconstructing} \\ 
+COW civil war ongoing & Dichotomous & \citet{cow2017cinc,singer1988reconstructing} \\ 
+Democracy (Polity IV $\geq 6$) & Dichotomous  & Authors' own calculations \\ 
+Discriminated dummy & Dichotomous & \citet{cederman2010ethnic}\\ 
+Discriminated population & Continuous & \citet{cederman2010ethnic} \\ 
+Ethnic diversity (ELF) & Continuous & \citet{fearon2003ethnicity} \\ 
+Ethnic war start & Dichotomous & \citet{cederman2010ethnic} \\ 
+Ethnic war ongoing & Dichotomous & \citet{cederman2010ethnic} \\ 
+Excluded population & Continuous & \citet{cederman2010ethnic} \\ 
+Interstate war & Dichotomous & \citet{singer1988reconstructing,cow2017cinc} \\ 
+Guerrilla & Dichotomous & \citet{balcells2014does}\\ 
+Military expenditure & Continuous & \citet{cow2017cinc} \\ 
+Military personnel & Continuous & \citet{cow2017cinc} \\ 
+Militias & Dichotomous & \citet{carey2013states} \\ 
+Mountainous Terrain & Continuous & \citet{fearon2003ethnicity} \\ 
+Physical integrity & Continuous & \citet{cingranelli2010cingranelli}\\ 
+Polarisation (all groups/main group) & Continuous &  Authors' own calculations \\ 
+Polarisation (all groups/population) & Continuous &  Authors' own calculations  \\ 
+Polarisation (included groups/population) & Continuous &  Authors' own calculations  \\ 
+Polarisation (included groups/main group) & Continuous &  Authors' own calculations  \\ 
+Polity IV & Continuous & \citet{marshall2017pitf}\\ 
+Polity IV squared & Continuous & Authors' own calculations \\ 
+Population & Continuous & \citet{gleditsch2002expanded} \\
+Post-Cold War & Dichotomous & Authors' own calculations \\ 
+Real GDP & Continuous & \citet{gleditsch2002expanded} \\ 
+Real GDP per capita & Continuous & \citet{gleditsch2002expanded} \\ 
+Real GDP per capita (log) & Continuous & Authors' own calculations  \\ 
+Regime transition & Continuous & Authors' own calculations \\ 
+Riot & Dichotomous & \citet{banks1999cross}\\ 
+Total battle deaths & Continuous & \citet{lacina2005monitoring} \\  
+Total trade & Continuous & \citet{cow2017cinc} \\ 
+Trade dependence (total trade/real GDP) & Continuous & Authors' own calculations \\ 
+UCDP civil war onset & Dichotomous & \citet{allansson2017organized,gleditsch2002armed} \\ 
+UCDP civil war ongoing & Dichotomous & \citet{allansson2017organized,gleditsch2002armed} \\ 
+Urban population (percentage) & Continuous & \citet{cow2017cinc} \\ 
+Years since last mass killing & Continuous & Authors' own calculations \\ 
+War with territory aims & Dichotomous & \citet{allansson2017organized,gleditsch2002armed} \\ 
+\hline \\[-1.8ex] 
+\end{tabular} 
+\end{table}
+
+\newpage
 
@@
 Year & 9,162 & 1,983.56 & 18.77 & 1,945 & 2,013 \\ 
+Genocide/politicide onset & 8,933 & 0.005 & 0.07 & 0 & 1\\ 
 Mass killing onset & 9,162 & 0.01 & 0.11 & 0 & 1 \\ 
-Years since last mass killing & 9,162 & 23.81 & 17.71 & 0 & 68 \\ 
-Years since last mass killing squared & 9,162 & 880.44 & 1,086.13 & 0 & 4,624 \\ 
-Years since last mass killing cubed & 9,162 & 39,339.21 & 65,025.39 & 0 & 314,432 \\ 
-UCDP civil war onset & 8,733 & 0.02 & 0.14 & 0 & 1 \\ 
-UCDP civil war ongoing & 8,733 & 0.15 & 0.36 & 0 & 1 \\ 
+&&&&&\\
+\textit{Independent Variables} & & & & \\
+&&&&&\\
+Assassination dummy & 8,991 & 0.08 & 0.27 & 0 & 1 \\ 
+CINC & 8,767 & 0.01 & 0.02 & 0.00 & 0.38 \\ 
+Coup dummy & 8,587 & 0.05 & 0.21 & 0 & 1 \\ 
 COW civil war onset & 8,160 & 0.01 & 0.12 & 0 & 1 \\ 
 COW civil war ongoing & 8,160 & 0.07 & 0.25 & 0 & 1 \\ 
+Democracy dummy & 8,991 & 0.37 & 0.48 & 0 & 1 \\ 
+Discriminated dummy & 6,981 & 0.35 & 0.48 & 0 & 1 \\ 
+Discriminated population & 6,981 & 0.06 & 0.15 & 0.00 & 0.98 \\ 
+Ethnic diversity (ELF) & 6,981 & 0.41 & 0.31 & 0 & 1 \\ 
 Ethnic war start & 7,760 & 0.01 & 0.12 & 0 & 1 \\ 
 Ethnic war ongoing & 7,760 & 0.11 & 0.31 & 0 & 1 \\ 
-Assassination dummy & 8,991 & 0.08 & 0.27 & 0 & 1 \\ 
-Democracy dummy & 8,991 & 0.37 & 0.48 & 0 & 1 \\ 
-Ethnic diversity (ELF) & 6,981 & 0.41 & 0.31 & 0 & 1 \\ 
+Excluded population & 6,981 & 0.16 & 0.22 & 0.00 & 0.98 \\ 
+Interstate war & 8,159 & 0.04 & 0.19 & 0 & 1 \\ 
+Guerrilla dummy & 714 & 0.81 & 0.40 & 0 & 1 \\ 
+Military expenditure & 8,290 & 4,607,120 & 27,785,906 & 0 & 693,600,000 \\ 
+Military personnel & 8,620 & 176.70 & 520.90 & 0 & 12,500 \\ 
+Militias & 4,097 & 0.22 & 0.42 & 0 & 1 \\ 
 Mountainous Terrain & 7,358 & 2.14 & 1.43 & 0.00 & 4.56 \\ 
-Population & 8,293 & 32,993.61 & 112,886.40 & 118.21 & 1,324,353.00 \\ 
-Real GDP & 8,293 & 215,317.70 & 804,827.20 & 129.68 & 13,193,478.00 \\ 
-Ln of real GDP per capita & 8,293 & 8.25 & 1.20 & 4.89 & 13.36 \\ 
-Real GDP per capita & 8,293 & 8,104.20 & 18,376.73 & 132.82 & 632,239.50 \\ 
-Polity IV & 8,558 & 0.42 & 7.50 & $-$10 & 10 \\ 
-Polity IV squared & 8,558 & 56.35 & 32.59 & 0 & 100 \\ 
-Excluded population & 6,981 & 0.16 & 0.22 & 0.00 & 0.98 \\ 
-Discriminated population & 6,981 & 0.06 & 0.15 & 0.00 & 0.98 \\ 
+Physical integrity & 4,499 & 4.73 & 2.31 & 0 & 8 \\ 
 Polarisation (all groups/main group) & 6,981 & 0.70 & 0.26 & 0.05 & 1 \\ 
@@
 Polarisation (included groups/main group) & 6,981 & 0.23 & 0.35 & 0 & 1 \\ 
-Discriminated dummy & 6,981 & 0.35 & 0.48 & 0 & 1 \\ 
-Interstate war & 8,159 & 0.04 & 0.19 & 0 & 1 \\ 
-Military expenditure & 8,290 & 4,607,120 & 27,785,906 & 0 & 693,600,000 \\ 
-Military personnel & 8,620 & 176.70 & 520.90 & 0 & 12,500 \\ 
-Percent urban population & 8,767 & 0.22 & 0.17 & 0.00 & 1.51 \\ 
+Polity IV & 8,558 & 0.42 & 7.50 & $-$10 & 10 \\ 
+Polity IV squared & 8,558 & 56.35 & 32.59 & 0 & 100 \\ 
+Population & 8,293 & 32,993.61 & 112,886.40 & 118.21 & 1,324,353.00 \\
 Post-Cold War & 8,991 & 0.40 & 0.49 & 0 & 1 \\ 
-Coup dummy & 8,587 & 0.05 & 0.21 & 0 & 1 \\ 
+Real GDP & 8,293 & 215,317.70 & 804,827.20 & 129.68 & 13,193,478.00 \\ 
+Real GDP per capita & 8,293 & 8,104.20 & 18,376.73 & 132.82 & 632,239.50 \\ 
+Real GDP per capita (log) & 8,293 & 8.25 & 1.20 & 4.89 & 13.36 \\ 
+Regime transition & 1,221 & $-$4.24 & 41.50 & $-$77 & 99 \\ 
 Riot dummy & 8,991 & 0.16 & 0.36 & 0 & 1 \\ 
-War with territory aims & 8,924 & 0.07 & 0.26 & 0 & 1 \\ 
+Total battle deaths & 714 & 6,050.86 & 24,404.78 & 100 & 350,000 \\  
 Total trade & 8,174 & 53,804.01 & 222,209.90 & 0.80 & 4,825,363.00 \\ 
 Trade dependence & 7,670 & 0.26 & 0.69 & 0.0001 & 22.11 \\ 
-Militias & 4,097 & 0.22 & 0.42 & 0 & 1 \\ 
-Physical integrity & 4,499 & 4.73 & 2.31 & 0 & 8 \\ 
-CINC & 8,767 & 0.01 & 0.02 & 0.00 & 0.38 \\ 
-Total battle deaths & 714 & 6,050.86 & 24,404.78 & 100 & 350,000 \\  
-Guerrilla dummy & 714 & 0.81 & 0.40 & 0 & 1 \\ 
-Regime transition & 1,221 & $-$4.24 & 41.50 & $-$77 & 99 \\ 
+UCDP civil war onset & 8,733 & 0.02 & 0.14 & 0 & 1 \\ 
+UCDP civil war ongoing & 8,733 & 0.15 & 0.36 & 0 & 1 \\ 
+Urban population (percentage) & 8,767 & 0.22 & 0.17 & 0.00 & 1.51 \\ 
+Years since last mass killing & 9,162 & 23.81 & 17.71 & 0 & 68 \\ 
+War with territory aims & 8,924 & 0.07 & 0.26 & 0 & 1 \\ 
 \hline \\[-1.8ex] 
@@
 
-Here we present a series of histograms with the coefficients' distribution of all variables in the main EBA model.
+We present a series of histograms with the coefficients' distribution of all variables in the main EBA model.
 There are 36 variables in total, seven of which are robust: Log GDP per capita, post-Cold War period, onset and ongoing civil wars (measured by the UCDP), previous riots, ethnic diversity and the squared term of the Polity IV index.
@@
 
-\subsection{Genocides during Civil Wars}
+\subsubsection{Genocides during Civil Wars}
 \label{sec:civil-wars}
@@
 We use three covariates that denote ongoing civil conflicts: one by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, another by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}.
-The significant variables in the models below is notably different from the previous ones.
-We see that as evidence that genocides that take place during conflicts follow a particular logic and are not caused by the same factors that foster mass killings in peacetime.
+The variables that reach significance in this set of models below are notably different from those obtained in the main estimation.
+This result provides evidence that mass violence during wartime time follows a separate logic from state killings in peacetime.
 
@@
 
-The following table shows the results of an EBA model using 3 variable combinations per model. The results are very similar to those reported in the paper, in which we used 4 variables in each model.
-
 The models below are based on 50,000 random draws from the full set of all possible regression models.
 \citet[819]{salaimartin2004determinants} argue that random sampling produces unbiased estimates of the regression coefficients with low computational time.
-The models presented in the manuscript, however, include the full set of possible regressions.
+The models presented in section \ref{sec:results4}, however, include the full set of possible regressions.
+
+The following table shows the results of an EBA with 3 variable combinations per model. The results are very similar to those reported above.
 
@@
 Table \ref{tab:mk-5vars} presents the results for models with up 5 variables in each regressions.
-In contrast with the main EBA model, the indicators of UCDP ongoing civil wars, ethnic diversity, and Polity IV square drop out of significance.
+In contrast with our main EBA model, the indicators of UCDP ongoing civil wars, ethnic diversity, and Polity IV square drop out of significance.
 Their individual CDFs(0) are about 0.88, just marginally below our specified threshold of 0.9.
@@
 
-In this section, we estimate the EBA model with different values of Variance Inflation Factor (VIF), a measure of multicollinearity for linear models.
-There is no standard definition about what constitutes an acceptable VIF value, although researchers often use a value of 10 as rule of thumb to indicate strong multicollinearity \citep[674]{o2007caution}.
+In this subsection, we estimate EBA models with different values of Variance Inflation Factor (VIF), which is a measure of multicollinearity.
+There is no standard definition about what constitutes an acceptable VIF value, although researchers often use 10 as rule of thumb to indicate strong multicollinearity \citep[674]{o2007caution}.
 Our original model used a slightly more conservative value of 7 as a cutoff.
-Here we test the same model with VIF $=$ 10 (less strict), 2.5 (more conservative), and a model without any VIF value (no restriction).
-The results are essentially identical to those in the main model.
-In the model with no VIF restriction, however, ethnic fractionalisation fails to meet our threshold by a very small margin.
+Here, we test the same model with VIF $=$ 10 (less strict), 2.5 (more conservative), and a model without VIF restrictions.
+The results are essentially identical to those of the main model.
+In the model with no VIF restriction, however, ethnic fractionalisation fails to meet the threshold by a very small margin.
 The CDF(0) of that covariate is 0.897, very close to the required value of 0.9.
@@
 
-We estimate the same EBA with logit and probit in addition to the linear probability models reported above.
-Logistic and probit regressions may have issues of complete separation, that is, when some covariates perfectly separates zeros and ones in the outcome variable.\footnote{We thank Mark Bell for sharing \texttt{R} code to estimate penalised-likelihood models.} In that case, the estimations fail to converge.
-We addressed this problem by adding a weak prior to the regression coefficients as suggested by \citet{gelman2008weakly}.
-We first scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then we added a Cauchy distribution with centre 0 and scale 2.5.
+We reestimate the main EBA model with logit and probit models.
+Nevertheless, logistic and probit regressions may have issues of complete separation, that is, some covariates may perfectly separate zeros and ones in the outcome variable.
+In that case, the estimations fail to converge.
+We address this problem by adding a weak prior to the regression coefficients as suggested by \citet{gelman2008weakly}.\footnote{We thank Mark Bell for sharing \texttt{R} code to estimate penalised-likelihood models.} First, we scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then added a Cauchy distribution with centre 0 and scale 2.5.
 The probit regressions use a scale of $2.5 \times 1.6$, which is also recommended by the authors \citep{arm2017rpackage}.
@@
 
-In the manuscript, we perform a grid search to optimise the hyperparameters of the random forest models.
-Because the grid search evaluates a wide range of parameter values at once, it is often unnecessary to run additional tests to assess the robustness of the results.
-Nevertheless, as random forests themselves are an approximation to a myriad of possible parameter combinations, changes in seed numbers may influence model output.
-Thus, we start the models with different random seed numbers to evaluate how study are our original results.\footnote{The numbers were generated at \href{https://www.random.org/}{https://www.random.org/}.} The findings holds quite well: Although variable importance chances from one model to another, the most significant variables appear repeatedly in the estimations.
-The marginal plots also show that their effects on the outcome variable are similar despite the nonlinearities.
-Below we show the 6 most significant variables and their respective partial dependence plots.
-
-\begin{figure}[H]
-    \centering
-    \includegraphics{images/drf-mk.pdf}
-    \caption{Distributed Random Forest -- Variable importance}
-    \label{fig:my_label}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp.pdf}
-    \caption{Distributed Random Forest -- Partial plots}
-    \label{fig:my_label}
-\end{figure}
-
 \subsubsection{Alternative Random Seeds}
 
-We tested the same model using different random seeds.
-Although not identical, the results bear strong similarities with the previous random forest estimation.
-A certain amount of variation is expected in machine learning algorithms; each model produces slightly different results as they are sensitive to the choice of seed.
-But as we see below, the results are broadly comparable, what confirms the robustness of the previous findings.
+As noted in section \ref{sec:methods4}, we perform a grid search to optimise the hyperparameters of the random forest models.
+The grid search evaluates a wide range of parameter values at once, therefore it is generally unnecessary to run additional tests to assess the robustness of the results.
+Nevertheless, as random forests themselves are an approximation to a number of possible parameter combinations, changes in seed numbers may influence the output.
+Thus, we start the models with different random seed numbers to evaluate how sturdy are our original results.\footnote{The numbers were generated at \href{https://www.random.org/}{https://www.random.org/}.} The findings holds quite well: Although variable importance changes from one model to another, the most significant variables appear repeatedly in the estimations.
+The marginal plots also show that their effects on the outcome variable remains similar despite eventual nonlinearities.
+We show the six most significant predictors of mass killings and their respective partial dependence plots.
 
@@
 
-Third model -- seed 1502436:
+Second model -- seed 1502436:
 
@@
 
-\subsection{Genocide/Politicide Onset}
+\subsection{Genocide/Politicide}
 \label{sec:mk-other-variable}
 
-In this section, we evaluate the models presented above with a measure of genocide and politicide by \citet{harff2003no}. The results show important contrasts with the previous analyses.
+In this section, we evaluate the models presented above with a measure of genocide and politicide by \citet{harff2003no}.
+The results show important contrasts with the previous analyses.
+First, no variable appear as significant in the main extreme bounds analysis.
+That is, none of the 36 predictors reached the threshold of CDF(0) $> 0.9$.
+The variable that came closest to significance was a dummy indicator of coups d'état, which has a CDF(0) of 0.897 and, as expected, is positively correlated with the onset of genocides.
+The distibution of the covariates' coefficients are available in figure \ref{fig:uamk}.
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/uamk.pdf}
+    \caption{EBA -- Genocides and politicides}
+    \label{fig:uamk}
+\end{sidewaysfigure}
+\clearpage
+
+\subsubsection{Genocide/Politicide during Civil Wars}
+
+Next, we evaluate what covariates are robust when considering only genocides and politicide that occur during civil conflicts.
+Post-Cold War years again appear as a significant variable and with a negative sign; excluded population also has a negative impact on the outcome variable in two analyses.
+
+\vspace{1cm}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{UCDP data} &  &  &  &  &  \\
+Excluded population & -0.037 & 0.022 & 64.524 & 0.9176 & 8758 \\
+ &  &  &  &  &  \\
+\textit{COW data} &  &  &  &  &  \\
+Excluded population & -0.057 & 0.031 & 65.703 & 0.9570 & 8820 \\
+Discriminated population & -0.050 & 0.029 & 53.850 & 0.93.67 & 8767 \\
+Post-Cold War years & -0.019 & 0.013 & 42.531 & 0.9203 & 8904 \\
+ &  &  &  &  &  \\
+\textit{Cederman et al. data} &  &  &  &  &  \\
+Assassination dummy & -0.009 & 0.006 & 47.723 & 0.9232 & 8828 \\ \hline
+\end{tabular}
+\caption{EBA -- Genocides and politicides}
+\label{tab:uamk1}
+\end{table}
 
 \newpage
-\subsubsection{Distributed Random Forests}
-
-Lastly, we present below three models using the distributed random forest algorithm \citep{h2o2017}.
-The results are in line with those obtained with the mass killing variable \citep{ulfelder2008assessing}.
-Again we see that CINC, the percentage of urban population, and variables concerning the military are some of the most important predictors of state-led violence.
-According to this measure, poor countries are at a much higher risk of genocide that its counterparts.
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/uamk-ucdp.pdf}
+    \caption{EBA -- Genocides and politicides during civil wars (UCDP data)}
+    \label{fig:uamk-ucdp}
+\end{sidewaysfigure}
+\clearpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/uamk-ucdp.pdf}
+    \caption{EBA -- Genocides and politicides during civil wars (COW data)}
+    \label{fig:uamk-cow}
+\end{sidewaysfigure}
+\clearpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/uamk-ucdp.pdf}
+    \caption{EBA -- Genocides and politicides during ethnic civil wars (Cederman et al. data)}
+    \label{fig:uamk-eth}
+\end{sidewaysfigure}
+\clearpage
+
+\newpage
+\subsection{Genocide/Politicide -- Random Forests}
+
+Lastly, we present three models using the distributed random forest algorithm \citep{h2o2017}.
+The results are in line with those obtained with the mass killing variable by  \citet{ulfelder2008assessing}.
+Again, we see that CINC, the percentage of urban population, and variables concerning the military are some of the most important predictors of state-led violence.
+The results confirm the overall finding of the chapter: Poor countries are more likely to experience mass killings, and states with a stronger army see a significant upward shift in genocide risk.
 
@@
 
-The \texttt{R} code below reproduces the analyses included in the manuscript.
+The \texttt{R} code below reproduces the analyses presented in chapter \ref{chap:killings}.
 
@@
 multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
+
+##############################################
+### Genocides and Politicides (Harff 2003) ###
+##############################################
+
+# Preparing the dataset
+df3 <- haven::read_dta("data/uamkstart.dta") %>% setDT()
+sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
+             "COWcivilwarongoing", "ethnowarstart", "ethnowarongoing",
+             "assdummy", "demdummy", "elf", "lmtnest", "pop", "realgdp",
+             "rgdppc", "polity2", "exclpop", "discpop", "polrqnew",
+             "poltrqnew", "egiptpolrqnew", "egippolrqnew", "discrim",
+             "elf2", "interstatewar", "milex", "milper", "percentpopurban",
+             "postcoldwar", "coupdummy", "riotdummy", "territoryaims",
+             "totaltrade", "tradedependence", "militias", "physint", "cinc",
+             "totalbeaths", "change", "guerrilladummy", "sf", "regtrans")
+
+df4 <- cbind(df3, df3[, shift(.SD, 1, give.names = TRUE),
+                    by = ccode, .SDcols = sd.cols])
+
+# Remove the second `ccode` variable
+df4 <- as.data.frame(df4[, -c(75)])
+
+# Add new variables
+df4$logrgdppc_lag_1 <- log(df4$rgdppc_lag_1)
+df4$polity2sq_lag_1 <- df4$polity2_lag_1^2
+
+# Renaming variables
+df5 <- as.data.frame(df4[, c(1:4, 72:116)])
+names(df5) <- sub("_.*","", names(df5))
+
+## Extreme Bounds
+free.variables <- c("logrgdppc", "polity2", "uamkyr")
+civilwar.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart")
+doubtful.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart", "assdummy",
+                        "totaltrade", "tradedependence", "milper", "milex",
+                        "pop", "totalbeaths", "guerrilladummy", "regtrans",
+                        "riotdummy", "territoryaims", "militias",
+                        "physint", "percentpopurban", "coupdummy",
+                        "postcoldwar",  "lmtnest", "realgdp", "discrim",
+                        "exclpop", "discpop", "elf",  "polrqnew",
+                        "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                        "polity2sq")
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df5, vif = 7, level = 0.9, 
+          se.fun = se.clustered.robust)
+
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
+                       "UCDPcivilwarongoing",
+                       "UCDPcivilwarstart", "COWcivilwarongoing",
+                       "COWcivilwarstart", "ethnowarongoing", "ethnowarstart",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last genocide",
+              "UCDP ongoing", "UCDP onset", "COW ongoing", "COW onset", 
+              "Ethnic ongoing", "Ethnic onset", "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+### Ongoing Civil Wars
+
+# UCDPcivilwarongoing == 1
+df.ucdp2 <- df5 %>% filter(UCDPcivilwarongoing == 1)
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop",
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "uamkstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.ucdp2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last genocide",
+              "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Groups/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+# COWcivilwarongoing == 1
+df.cow2 <- df5 %>% filter(COWcivilwarongoing == 1)
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop",
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "uamkstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.cow2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last genocide",
+              "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Groups/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+# ethnic civil war == 1
+df.eth2 <- df5 %>% filter(ethnowarongoing == 1)
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop", 
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "uamkstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.eth2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last genocide",
+              "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Groups/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+########################################################
+### Genocide/Politicide -- Distributed Random Forest ###
+########################################################
+
+df5a <- as.h2o(df5)
+
+df5a$uamkstart <- as.factor(df5a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df5a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df5a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df5), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, 
+               grid_id = "gridrf05",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 500, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf05",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## UCDP == 1
+df.ucdp2a <- as.h2o(df.ucdp2)
+
+df.ucdp2a$uamkstart <- as.factor(df.ucdp2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.ucdp2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.ucdp2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.ucdp2), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf06",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 100, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf06",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## COW == 1
+df.cow2a <- as.h2o(df.cow2)
+
+df.cow2a$uamkstart <- as.factor(df.cow2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.cow2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.cow2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.cow2), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf07",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 100, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf07",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
+## ETHONSET == 1
+df.eth2a <- as.h2o(df.eth2)
+
+df.eth2a$uamkstart <- as.factor(df.eth2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.eth2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.eth2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.eth2), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf08",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
+               search_criteria = list(strategy = "RandomDiscrete", 
+                                      max_models = 100, 
+                                      stopping_metric = "auc", 
+                                      stopping_tolerance = 0.01, 
+                                      stopping_rounds = 5, 
+                                      seed = 26227709))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf08",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/root/Documents/mk/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+h2o.performance(rf2, newdata = test)
+
 \end{verbatim}

### 2018-05-07 dissertation.tex
beta version :)

@@
 % SUMMARY HERE (300 word limit for most subjects):
-This thesis addresses three questions on political and criminal violence.
+This thesis addresses three topics in political and criminal violence.
 The first essay is an empirical evaluation of a broad set of homicide reduction policies implemented in the state of São Paulo, Brazil.
@@
 Although the literature on mass killings has grown significantly over the last decades, it remains unclear whether estimates are robust to different model specifications or which variables accurately predict the onset of genocides and politicides.
-To address this issue, I test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+To address these issues, I test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
 The results show that GDP per capita, the post-Cold War period, and stable political regimes are negatively associated with mass killings.
 In contrast, ethnic diversity, civil wars, and previous political turmoils increase the risk of state-led violence.
-Mass killings that occur during civil wars are correlated with a different set of variables, and I find that territorial conflicts and the use of militias are negatively associated with the outcome.
+Mass killings that occur during civil wars are correlated with a different set of covariates, and I find that territorial conflicts and the use of militias are negatively associated with the outcome variable.
 The random forests estimations indicate that years since the last episode of mass violence, GDP per capita, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to the models' out-of-sample predictive power.
@@
 I would like to extend thanks to the many people, in many countries, who so generously contributed to the work presented in this thesis.
-First and foremost, I want to thank my supervisor, Professor David Skarbek, for the guidance and encouragement he provided me.
-David was not only an excellent supervisor but also a great friend; his empathy, enthusiasm, and sense of humour made my stay at King's College a very pleasant one.
-I could not have wished for a better mentor for my PhD studies.
+First and foremost, I want to thank my supervisor, Professor David Skarbek, for the guidance and encouragement he provided me throughout this research.
+David has been very supportive since the first time we met, and his dynamism, vision, empathy, and great sense of humour have deeply inspired me.
+I could not have wished for a better supervisor for my PhD studies.
 Thank you very much.
@@
 Completing this work would have been all the more difficult were it not for the support and constructive criticism provided by my friends.
-Specially, I would like to thank Veronica Adaeva, André Amaro, Paulo Roberto Araujo, Aline Bagatin, Fábio Barros, Michel Hulmann, Guilherme Kerr, Maurício Pantaleão, Letícia Sene, and the participants of academic seminars at King's College London, the Mercatus Center at George Mason University, and the Prometheus Institute in Berlin.
+Specially, I would like to thank Veronica Adaeva, André Amaro, Paulo Roberto Araujo, Aline Bagatin, Fábio Barros, Michel Hulmann, Guilherme Kerr, Maurício Pantaleão, Letícia Sene, and the participants of academic seminars at King's College London and the Prometheus Institute in Berlin.
 I am deeply indebted to them for their insightful comments and hard questions.
 
-I am also deeply thankful to several present and past members of the Friedrich Naumann Foundation for Freedom.
-Some years ago, I had received a scholarship to attend a short course in Germany, and that first experience abroad has transformed my life ever since.
+I am also sincerely thankful to several present and past members of the Friedrich Naumann Foundation for Freedom.
+In 2009, I received a scholarship to attend a short course in Germany, and that first experience abroad has transformed my life ever since.
 The present work is, to a large extent, a result of the push I received from the Naumann Foundation almost 10 years ago.

### 2018-05-07 chapters/04-killings.tex
beta 2! :)

@@
 \end{tabular}
-\caption{Extreme Bounds Analysis -- Mass Killings (robust variables only)}
+\caption{Extreme Bounds Analysis -- Mass Killings (Robust Variables Only)}
 \label{tab:eba1}
@@
 \end{tabular}
-\caption{EBA -- Mass killings during civil wars (robust variables only)}
+\caption{EBA -- Mass Killings during Civil Wars (Robust Variables Only)}
 \label{tab:ucdp}
@@
 \includegraphics[width=.9\textwidth, height=8cm]{images/drf.pdf}
-\caption{Distributed Random Forest -- Variable Importance (scaled)}
+\caption{Distributed Random Forest -- Variable Importance (Scaled)}
 \label{fig:drfuv}
@@
 Although costly in the short run, these measures would substantially decrease the likelihood of state violence by breaking the ``conflict trap'' in which past conflicts create the condition for new ones \citep{collier2003breaking}.
+
+\newpage
 
@@
     \includegraphics[width=\textwidth]{images/mk.pdf}
-    \caption{Extreme Bounds Analysis -- Mass killings}
+    \caption{Extreme Bounds Analysis -- Mass Killings}
     \label{fig:mk}
@@
 \end{tabular}
-\caption{EBA -- Mass killings during civil wars}
+\caption{EBA -- Mass Killings during Civil Wars}
 \label{tab:ucdp1}
@@
     \includegraphics[width=\textwidth]{images/mk-ucdp.pdf}
-    \caption{EBA -- Mass killings during civil wars (UCDP data)}
+    \caption{EBA -- Mass Killings during Civil Wars (UCDP Data)}
     \label{fig:mk-ucdp}
@@
     \centering
-    \includegraphics[width=\textwidth]{images/mk-ucdp.pdf}
-    \caption{EBA -- Mass killings during civil wars (COW data)}
+    \includegraphics[width=\textwidth]{images/mk-cow.pdf}
+    \caption{EBA -- Mass Killings during Civil Wars (COW Data)}
     \label{fig:mk-cow}
@@
     \centering
-    \includegraphics[width=\textwidth]{images/mk-ucdp.pdf}
-    \caption{EBA -- Mass killings during ethnic civil wars (Cederman et al. data)}
+    \includegraphics[width=\textwidth]{images/mk-eth.pdf}
+    \caption{EBA -- Mass Killings during ethnic civil wars (Cederman et al. Data)}
     \label{fig:mk-eth}
@@
 \end{tabular}
-\caption{EBA -- 3 variables in each regression model}
+\caption{EBA -- 3 Variables}
 \label{tab:mk-3vars}
@@
     \includegraphics[width=\textwidth]{images/mk-3vars.pdf}
-    \caption{EBA -- 3 variables in each regression model}
+    \caption{EBA -- 3 Variables}
     \label{fig:mk-3vars}
@@
 \end{tabular}
-\caption{EBA -- 5 variables in each regression model}
+\caption{EBA -- 5 Variables}
 \label{tab:mk-5vars}
@@
     \includegraphics[width=\textwidth]{images/mk-5vars.pdf}
-    \caption{EBA -- 5 variables in each regression model}
+    \caption{EBA -- 5 Variables}
     \label{fig:mk-5vars}
@@
     \centering
-    \includegraphics[width=\textwidth]{images/mk-high-vif.pdf}
+    \includegraphics[width=\textwidth]{images/mk-low-vif.pdf}
     \caption{EBA -- VIF 2.5}
@@
 \end{tabular}
-\caption{EBA -- Logistic model}
+\caption{EBA -- Logistic Regression}
 \label{tab:mk-logit}
@@
     \includegraphics[width=\textwidth]{images/mk-logit.pdf}
-    \caption{EBA -- Logistic model}
+    \caption{EBA -- Logistic Regression}
     \label{fig:mk-logit}
@@
 \end{tabular}
-\caption{EBA -- Probit model}
+\caption{EBA -- Probit Regression}
 \label{tab:eba1}
@@
     \includegraphics[width=\textwidth]{images/mk-probit.pdf}
-    \caption{EBA -- Probit model}
+    \caption{EBA -- Probit Regression}
     \label{fig:mk-probit}
@@
     \includegraphics{images/drf-mk2.pdf}
-    \caption{Variable importance -- Seed 44849999}
+    \caption{Variable Importance -- Seed 44849999}
     \label{fig:my_label}
@@
     \includegraphics[width=\textwidth, height=9cm]{images/drfdpp2.pdf}
-    \caption{Partial plots -- Seed 44849999}
+    \caption{Partial Plots -- Seed 44849999}
     \label{fig:my_label}
@@
     \includegraphics{images/drf-mk3.pdf}
-    \caption{Variable importance -- Seed 1502436}
+    \caption{Variable Importance -- Seed 1502436}
     \label{fig:my_label}
@@
     \includegraphics[width=\textwidth, height=9cm]{images/drfdpp3a.pdf}
-    \caption{Partial plots -- Seed 1502436}
+    \caption{Partial Plots -- Seed 1502436}
     \label{fig:my_label}
@@
 
-\subsection{Genocide/Politicide}
+\subsection{Genocides/Politicides}
 \label{sec:mk-other-variable}
@@
     \includegraphics[width=\textwidth]{images/uamk.pdf}
-    \caption{EBA -- Genocides and politicides}
+    \caption{EBA -- Genocides/Politicides}
     \label{fig:uamk}
@@
 
-\subsubsection{Genocide/Politicide during Civil Wars}
+\subsubsection{Genocides/Politicides during Civil Wars}
 
@@
 \end{tabular}
-\caption{EBA -- Genocides and politicides}
+\caption{EBA -- Genocides/Politicides}
 \label{tab:uamk1}
@@
     \includegraphics[width=\textwidth]{images/uamk-ucdp.pdf}
-    \caption{EBA -- Genocides and politicides during civil wars (UCDP data)}
+    \caption{EBA -- Genocides and Politicides during Civil Wars (UCDP Data)}
     \label{fig:uamk-ucdp}
@@
     \centering
-    \includegraphics[width=\textwidth]{images/uamk-ucdp.pdf}
-    \caption{EBA -- Genocides and politicides during civil wars (COW data)}
+    \includegraphics[width=\textwidth]{images/uamk-cow.pdf}
+    \caption{EBA -- Genocides and Politicides during Civil Wars (COW Data)}
     \label{fig:uamk-cow}
@@
     \centering
-    \includegraphics[width=\textwidth]{images/uamk-ucdp.pdf}
-    \caption{EBA -- Genocides and politicides during ethnic civil wars (Cederman et al. data)}
+    \includegraphics[width=\textwidth]{images/uamk-eth.pdf}
+    \caption{EBA -- Genocides and Politicides during Ethnic Civil Wars (Cederman et al. Data)}
     \label{fig:uamk-eth}
@@
 \newpage
-\subsection{Genocide/Politicide -- Random Forests}
+\subsection{Genocides/Politicides -- Random Forests}
 
@@
     \includegraphics{images/drf-gp.pdf}
-    \caption{Variable importance -- Genocide/Politicide}
+    \caption{Variable Importance -- Genocides/Politicides}
     \label{fig:my_label}
@@
     \includegraphics[width=\textwidth, height=9cm]{images/drfdpp4a.pdf}
-    \caption{Partial plots -- Genocide/Politicide}
+    \caption{Partial Plots -- Genocides/Politicides}
     \label{fig:my_label}
@@
     \includegraphics{images/drf-gp1.pdf}
-    \caption{Variable importance -- Genocide/Politicide (UCDP data)}
+    \caption{Variable Importance -- Genocides/Politicides (UCDP Data)}
     \label{fig:my_label}
@@
     \includegraphics[width=\textwidth, height=9cm]{images/drfdpp5a.pdf}
-    \caption{Partial plots -- Genocide/Politicide (UCDP data)}
+    \caption{Partial Plots -- Genocides/Politicides (UCDP Data)}
     \label{fig:my_label}
@@
     \includegraphics{images/drf-gp2.pdf}
-    \caption{Variable importance -- Genocide/Politicide (COW data)}
+    \caption{Variable Importance -- Genocides/Politicides (COW Data)}
     \label{fig:my_label}
@@
     \includegraphics[width=\textwidth, height=9cm]{images/drfdpp6a.pdf}
-    \caption{Partial plots -- Genocide/Politicide (COW data)}
+    \caption{Partial Plots -- Genocides/Politicides (COW Data)}
     \label{fig:my_label}
@@
     \includegraphics{images/drf-gp3.pdf}
-    \caption{Variable importance -- Genocide/Politicide (Cederman et al. data)}
+    \caption{Variable Importance -- Genocides/Politicides (Cederman et al. Data)}
     \label{fig:my_label}
@@
     \includegraphics[width=\textwidth, height=9cm]{images/drfdpp7a.pdf}
-    \caption{Partial plots -- Genocide/Politicide (Cederman et al. data)}
+    \caption{Partial Plots -- Genocides/Politicides (Cederman et al. Data)}
     \label{fig:my_label}

### 2018-05-07 chapters/04-killings.tex
beta 3! :)

@@
 \textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
-\textit{UCDP data} &  &  &  &  &  \\
+\textit{UCDP Data} &  &  &  &  &  \\
 Territory aims & -0.044 & 0.019 & 74.997 & 0.9804 & 17902 \\
@@
  &  &  &  &  &  \\
-\textit{COW data} &  &  &  &  &  \\
+\textit{COW Data} &  &  &  &  &  \\
 Physical integrity & 0.024 & 0.013 & 66.674 & 0.9564 & 17902 \\
@@
  &  &  &  &  &  \\
-\textit{Cederman et al. data} &  &  &  &  &  \\
+\textit{Cederman et al. Data} &  &  &  &  &  \\
 Territory aims & -0.051 & 0.026 & 74.288 & 0.9167 & 17902 \\
@@
 \includegraphics[width=\textwidth, height=9cm]{images/drfdpp2.pdf}
-\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (UCDP data)}
+\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (UCDP Data)}
 \label{fig:drfdpp2}
@@
 \includegraphics[width=\textwidth, height=9cm]{images/drfdpp3.pdf}
-\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (COW data)}
+\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (COW Data)}
 \label{fig:drfdpp3}
@@
 \includegraphics[width=\textwidth, height=9cm]{images/drfdpp4.pdf}
-\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. data)}
+\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. Data)}
 \label{fig:drfdpp4}

### 2018-05-07 dissertation.tex
beta 3! :)

@@
 	pdfauthor={Danilo Alves M. Freire},
+	pdfsubject={Political Economy},
+	pdfkeywords={King's College London, Latin America, PhD Thesis, Political Economy, Violence},
 	pdfborder={0 0 0},

### 2018-09-15 paper-bicho/bicho.tex
add files

--- 2018-04-23 chapters/03-bicho.tex
@@
-\chapter{Beasts of Prey or Rational Animals? Private Governance in Brazil's \emph{Jogo do Bicho}}
-\label{chap:bicho}
+\documentclass[a4paper,12pt]{article}
+\usepackage[utf8]{inputenc}
+\usepackage{microtype}
+\usepackage{libertine}
+\usepackage{inconsolata}
+\usepackage{xcolor}
+\definecolor{darkblue}{rgb}{0.0,0.0,0.55}
+\usepackage{setspace}
+\usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
+\usepackage[backref,pagebackref]{hyperref}
+\usepackage{graphicx}
+\usepackage{float}
+\usepackage{mathtools}
+\usepackage{caption}
+\usepackage[UKenglish]{babel}
+\usepackage[UKenglish]{isodate}
+\cleanlookdateon
+\usepackage[authoryear]{natbib}
+\usepackage{babelbib}
+\exhyphenpenalty=1000
+\hyphenpenalty=1000
+\widowpenalty=1000
+\clubpenalty=1000
+\renewcommand*{\backref}[1]{}
+\renewcommand*{\backrefalt}[4]{%
+	\ifcase #1 (Not cited.)%
+	\or        Cited on page~#2.%
+	\else      Cited on pages~#2.%
+	\fi}
+\renewcommand{\backreftwosep}{ and~}
+\renewcommand{\backreflastsep}{ and~}
+\urlstyle{same}  % don't use monospace font for urls
+
+\hypersetup{pdftitle={Beasts of Prey or Rational Animals? Private Governance in Brazil's Jogo do bicho},
+	pdfauthor={},
+	pdfborder={0 0 0},
+	breaklinks=true,
+	linkcolor=darkblue,
+	citecolor=darkblue,
+	urlcolor=darkblue,
+	colorlinks=true}
+
+\doublespacing
+
+\title{\textbf{Beasts of Prey or Rational Animals?\\ Private Governance in Brazil's \emph{Jogo do Bicho}}}
+
+\author{Danilo Freire\thanks{Postdoctoral Research Associate, Political Theory Project, Brown University, 8 Fones Alley, Providence, RI 02912, \href{mailto:danilofreire@brown.edu}{\texttt{danilofreire@brown.edu}}, \href{http://danilofreire.com}{\texttt{http://danilofreire.com}}.
+I would like to thank Paulo Roberto Araujo, Jose Bolanos, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, the editor, and four reviewers for their helpful comments.
+I gratefully acknowledge the support of the National Council for Scientific and Technological Development.}
+}
+
+\date{\today}
+
+\begin{document}
+\maketitle
+
+\begin{abstract}
+ \noindent
+ This work presents a rational choice account for the \emph{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
+Over 120 years, the \emph{jogo do bicho} has grown into a multimillion-dollar business and exerted a significant impact on the Brazilian society.
+The lottery has been a major sponsor of the Carnival Parade in Rio de Janeiro, which is among the world's most famous popular festivals, and it has remained an important driver of state corruption in the country.
+This work investigates the institutions that have caused the \emph{jogo do bicho}'s notable growth and long-term survival outside the boundaries of the Brazilian law.
+It also explains the emergence of the informal rules that govern the game as well as their enforcement mechanisms.
+
+ \vspace{.5cm}
+ \noindent
+ \textbf{Keywords}: Brazil; criminal organisations; gambling; \emph{jogo do bicho}; private governance
+
+ \vspace{.25cm}
+ \noindent
+ \textbf{JEL Classification Codes}: D72, K42, Z00
+\end{abstract}
+
+\newpage
 
 \section{Introduction}
-\label{sec:intro3}
+\label{sec:intro}
 
@@
 Access: December 2016.}
-
-% \begin{figure}[!htbp]
-%  \centering
-%  \begin{minipage}[b]{0.45\textwidth}
-%   \includegraphics[width=\textwidth, height=6cm]{images/bicho01.jpg}
-%  \end{minipage}
-%  \hfill
-%  \begin{minipage}[b]{0.45\textwidth}
-%   \includegraphics[width=\textwidth, height=6cm]{images/bicho02.jpg}
-%  \end{minipage}
-%  \caption{Left: cartoon of the Baron of Drummond and the animals of the \emph{jogo do bicho} (1896).
-Right: entry ticket to Rio's Zoological Garden that allowed the bearer to join the raffle.
-Sources: Instituto Histórico e Geográfico Brasileiro, Rio de Janeiro, Revista Illustrada, ano 21, no.
-718 (1896) and Museu da Imagem e do Som, Rio de Janeiro.
-Reproduced in \citet[35--36]{chazkel2011laws}.}
-%  \label{fig:barao}
-% \end{figure}
 
@@
 
-\begin{figure}[!htbp]
- \begin{tikzpicture}[->,>=stealth',shorten >=4pt,auto,node distance=3cm, semithick][h]
-  % nodes %
-  \node[circle,fill,inner sep=3pt,label=above right: \shortstack{Urban \\ Poverty}] (p) {};
-  \node[circle,fill,inner sep=3pt,label=right: \shortstack{Jogo \\ do Bicho}, right = 8 of p] (y) {};
-  \node[circle,fill,inner sep=3pt,label=below left: {Immigration}, below = 3 of p] (i) {};
-  \node[circle,fill,inner sep=3pt,label=above:{Money Supply}, above = 3 of p] (m) {};
-  \node[circle,fill,inner sep=3pt,label=below right:{Social Ties}, right = 3 of i] (s) {};
-  \node[circle,fill,inner sep=3pt,label=above:{Weak Law Enforcement}, above = 3 of y] (l) {};
-  \node[circle,fill=white, draw, outer sep=0pt, inner sep=3pt,label= left: \shortstack{Correlated \\ Background Factors}, above left = 2 of p] (u) {};
-  % edges %
-  \draw[->, line width = 1] (p) -- (y);
-  \draw[->, line width = 1] (s) -- (y);
-  \draw[->, line width = 1] (i) -- (s);
-  \draw[->, line width = 1] (l) -- (y);
-  \draw[->, line width = 1] (m) -- (y);
-  \draw[->, line width = 1] (i) -- (p);
-  \draw[->, dashed, line width = 1] (u) -- (p);
-  \draw[->, dashed, line width = 1] (u) -- (m);
- \end{tikzpicture}
- \caption{Directed Acyclic Graph -- Explanatory Variables for the \emph{Jogo do Bicho}}
- \label{fig:dag}
-\end{figure}
-
 I start with the impact of urban poverty on the animal game.
@@
 They also supervise individuals who carry menial tasks in the business, transfer money to other gambling branches and double-check the balance sheets of the betting sites \citetext{\citealp[71]{labronici2012paratodos}; \citealp[142]{misse2007illegal}}.
-
-% \begin{figure}[!htbp]
-%  \centering
-%  \begin{minipage}[b]{0.45\textwidth}
-%   \includegraphics[width=\textwidth, height=6cm]{images/bicho03.jpg}
-%  \end{minipage}
-%  \hfill
-%  \begin{minipage}[b]{0.45\textwidth}
-%   \includegraphics[width=\textwidth, height=6cm]{images/bicho04.jpg}
-%  \end{minipage}
-%  \caption{\emph{Jogo do bicho} betting sites in 1917 and in 2011.
-The picture on the right shows a \emph{bicheiro}, a street-corner vendor.
-Sources: \citet{alecrim2012bicho} and \citet{ferrarini2011bicho}.}
-%  \label{fig:banca1917}
-% \end{figure}
 
@@
 
-% \begin{figure}[!htbp]
-%  \centering
-%  \includegraphics[width=\textwidth, height=6cm]{images/bicho06.jpg}
-%  \caption{Results of a \emph{jogo do bicho} draw from 09 January 2016.
-`Federal' means that the winning numbers were drawn by the federal government lottery.
-The first prize was group 14, the cat.
-Source: Unknown.
-Available at: \href{https://goo.gl/6PHV8u}{https://goo.gl/6PHV8u}.
-Access: December 2016.}
-%  \label{fig:federal}
-% \end{figure}
-
 Second, they included representatives of all major \emph{jogo do bicho} bankers in every draw and independently publicise the game results.
@@
 Moreover, collusion can also be spotted if the draws show repeated numbers or unusual patterns.
-
-% \begin{figure}[!htbp]
-%  \centering
-%  \includegraphics[width=\textwidth, height=6cm]{images/bicho05.jpg}
-%  \caption{\emph{Jogo do bicho} results are fixed on light poles in Rio de Janeiro. Source: \citet{gomes1998bicho}.}
-%  \label{fig:poste}
-% \end{figure}
 
@@
 
-% \begin{figure}[!htbp]
-%  \centering
-%  \includegraphics[width=.8\textwidth, height=8cm]{images/bicho07.jpg}
-%  \caption{Castor de Andrade is shown celebrating after the samba school he sponsored, Mocidade Independente, won the Carnival parade in 1996.
-He was surrounded by colleagues and police officers.
-Andrade was the founding president of LIESA (1984--1985) and the wealthiest \emph{bicheiro} of Rio de Janeiro at the time.
-Source: Folha Imagem.
-Reproduced in \citet[139]{misse2007illegal}.}
-%  \label{fig:castor}
-% \end{figure}
-
 The funding of the samba schools had an indirect effect to the animal game.
@@
 
-% \begin{figure}[!htbp]
-%  \centering
-%  \includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
-%  \caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right).
-The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'.
-Source: \citet{extra2012aniz}.}
-%  \label{fig:aniz}
-% \end{figure}
-
 The votes from poor communities are instrumental for aspiring politicians.
@@
 This is an important step to elucidate how \emph{bicheiros} continue to influence politics and the public across Brazil.
+
+\newpage
+
+\bibliography{references}
+\bibliographystyle{apalike}
+\end{document}

### 2018-09-19 paper-bicho/bicho.tex
rewrite intro1

@@
 \usepackage{libertine}
-\usepackage{inconsolata}
-\usepackage{xcolor}
+\usepackage[scaled=.9]{inconsolata}
+\usepackage[usenames,dvipsnames]{xcolor}
 \definecolor{darkblue}{rgb}{0.0,0.0,0.55}
 \usepackage{setspace}
-\usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
+\usepackage[top=2.0cm,bottom=2.0cm,left=2.2cm,right=2.2cm]{geometry}
 \usepackage[backref,pagebackref]{hyperref}
@@
 \urlstyle{same}  % don't use monospace font for urls
-
+\setcitestyle{aysep={}} 
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd{\NAT@citex}
+	  {\@citea\NAT@hyper@{%
+		 \NAT@nmfmt{\NAT@nm}%
+		 \hyper@natlinkbreak{\NAT@aysep\NAT@spacechar}{\@citeb\@extra@b@citeb}%
+		 \NAT@date}}
+	  {\@citea\NAT@nmfmt{\NAT@nm}%
+	   \NAT@aysep\NAT@spacechar\NAT@hyper@{\NAT@date}}{}{}
+	\patchcmd{\NAT@citex}
+	  {\@citea\NAT@hyper@{%
+		 \NAT@nmfmt{\NAT@nm}%
+		 \hyper@natlinkbreak{\NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi}%
+		   {\@citeb\@extra@b@citeb}%
+		 \NAT@date}}
+	  {\@citea\NAT@nmfmt{\NAT@nm}%
+	   \NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi\NAT@hyper@{\NAT@date}}
+	  {}{}
+\makeatother
 \hypersetup{pdftitle={Beasts of Prey or Rational Animals? Private Governance in Brazil's Jogo do bicho},
-	pdfauthor={},
+	pdfauthor={Danilo Freire},
 	pdfborder={0 0 0},
 	breaklinks=true,
-	linkcolor=darkblue,
-	citecolor=darkblue,
+	linkcolor=Mahogany,
+	citecolor=Mahogany,
 	urlcolor=darkblue,
@@
 \author{Danilo Freire\thanks{Postdoctoral Research Associate, Political Theory Project, Brown University, 8 Fones Alley, Providence, RI 02912, \href{mailto:danilofreire@brown.edu}{\texttt{danilofreire@brown.edu}}, \href{http://danilofreire.com}{\texttt{http://danilofreire.com}}.
-I would like to thank Paulo Roberto Araujo, Jose Bolanos, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, the editor, and four reviewers for their helpful comments.
+I would like to thank Toke Aidt, Paulo Roberto Araujo, Jose Bola\~{n}os, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, Graham Denyer Willis, the editors of Latin American Perspectives, and four reviewers for their valuable comments.
 I gratefully acknowledge the support of the National Council for Scientific and Technological Development.}
@@
  \noindent
- This work presents a rational choice account for the \emph{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
-Over 120 years, the \emph{jogo do bicho} has grown into a multimillion-dollar business and exerted a significant impact on the Brazilian society.
-The lottery has been a major sponsor of the Carnival Parade in Rio de Janeiro, which is among the world's most famous popular festivals, and it has remained an important driver of state corruption in the country.
-This work investigates the institutions that have caused the \emph{jogo do bicho}'s notable growth and long-term survival outside the boundaries of the Brazilian law.
-It also explains the emergence of the informal rules that govern the game as well as their enforcement mechanisms.
+This article presents a rational choice account of Brazil's \emph{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
+Over 120 years, the \emph{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and has been a major driver of the persistent political corruption in the country.
+First, I examine the institutions that fostered \emph{jogo do bicho}'s notable growth outside of the boundaries of the law.
+Second, I explain how the lottery financiers combine costly signals, selective incentives, and innovative uses of government institutions to establish their reputation as honest brokers.
+Lastly, I analyse the relationship between the game sponsors and political actors and discuss how the game financiers influence the Brazilian state.
 
@@
  \noindent
- \textbf{Keywords}: Brazil; criminal organisations; gambling; \emph{jogo do bicho}; private governance
+ \textbf{Keywords}: Brazil; criminal organisations; \emph{jogo do bicho}; political corruption; private governance
 
@@
 In 1892, Baron João Batista de Viana Drummond came up with a new idea to fund his cash-strapped zoo.
-Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \emph{Jardim Zoológico,} or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city.
+Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \emph{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city.
 But it lacked visitors.
@@
 Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, making the game similar to a common raffle.
-But it did not take long until visitors could name their animals of choice.
-This change made the game considerably more appealing and lasts until this day \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.\footnote{The amount was higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.} The lottery was labelled as the \emph{jogo do bicho}, or the animal game, and it was immediately adopted by the public.
-Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but in stores across Rio de Janeiro.
-He rightly predicted that this small change would increase profits, but there was one thing the Baron did not foresee.
-He unleashed a new gambling market.
+But soon visitors could name their animals of choice.
+This change made the game more appealing and lasts until this day \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.
+The lottery was labelled as the \emph{jogo do bicho}, or the animal game, and it was immediately adopted by the public.
+Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in stores across Rio de Janeiro.
+Unexpectedly, the Baron created a whole new gambling market.
 
-The \emph{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace \citep{magalhaes2005ganhou, soares1993jogo}.
-A network of street bookmakers, called \emph{bicheiros}, expanded spontaneously the original game in innovative ways.
-Evading state regulations, \emph{bicheiros} made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the clandestine numbers game  \citep[37]{chazkel2011laws}.
-The \emph{jogo do bicho} became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today {[}1895{]}, in Rio de Janeiro, the game is everything.
+The \emph{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace.
+A network of street bookmakers, called \emph{bicheiros}, made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the clandestine numbers game  \citep[37]{chazkel2011laws}.
+The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today {[}1895{]}, in Rio de Janeiro, the game is everything.
 {[}\ldots{}{]} Nobody works!
-Everybody plays' \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.} But this tolerant state of affairs did not last.
-Civil servants and police officers criminalised the \emph{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos, krelling2014jogos, villar2008contravencao}.\footnote{The National Lottery Company (\emph{Companhia das Loterias Nacionaes do Brazil}), a public-private partnership founded four years after the creation of the \emph{jogo do bicho}, also lobbied actively for a hard-line stance against the \emph{bicheiros} \citep[82]{da1999aguias}.}
+Everybody plays' \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.}
 
-Yet the game has survived.
-The animal game has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{gaspari2002ditadura, jupiara2015poroes}.
+But this tolerant state of affairs did not last.
+Civil servants and police officers criminalised the \emph{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}.
+The campaign extended for several decades and received considerable support from the \emph{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after the creation of the animal game \citep[82]{da1999aguias}.
+The game was officially banned in 1946 and remains illegal to this day.
+
+Yet, the \emph{jogo do bicho} has survived.
+The animal game has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{jupiara2015poroes}.
 But more than a act of defiance, the \emph{jogo do bicho} is a successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
@@
 
-Moreover, the animal game plays a crucial role in the expansion of Rio de Janeiro's Carnival Parade, a popular festivity synonymous with Brazil at home and abroad \citep{araujo2003carnaval,costa2001100,da1973carnaval, da1979carnavais,vianna1995misterio}.
-\emph{Bicheiros} donate hefty sums to `samba schools' to gather support of poor communities and, no less importantly, to co-opt local politicians attracted by the financial and electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
-This patron-client relationship has been proven effective: in 2016, the Carnival generated about USD 900 million in revenue and the Rio de Janeiro state received more than one million tourists.\footnote{Data provided by the Brazilian government.
-See \href{https://goo.gl/XMcbTM}{https://goo.gl/XMcbTM} (in Portuguese).
-Access: December 2016.}
+In this article I offer a rational choice interpretation of the \emph{jogo do bicho}.
+More specifically, I discuss how \emph{bicheiros} adopt rational strategies to solve collective action problems and to maximise their political leverage.
+I use the case of Rio de Janeiro to illustrate the lottery's profit abilities and its effect in political corruption.
 
-In this article I offer a rational choice interpretation of the \emph{jogo do bicho} and discuss how \emph{bicheiros} promote social order, solve information asymmetries, and reduce negative externalities.
 My analysis discusses three strands of academic literature.

### 2018-09-19 paper-bicho/bicho.tex
update intro2

@@
 Over 120 years, the \emph{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and has been a major driver of the persistent political corruption in the country.
-First, I examine the institutions that fostered \emph{jogo do bicho}'s notable growth outside of the boundaries of the law.
+First, I examine the mechanisms that fostered \emph{jogo do bicho}'s notable growth outside of the boundaries of the law.
 Second, I explain how the lottery financiers combine costly signals, selective incentives, and innovative uses of government institutions to establish their reputation as honest brokers.
@@
  \noindent
- \textbf{Keywords}: Brazil; criminal organisations; \emph{jogo do bicho}; political corruption; private governance
+ \textbf{Keywords}: Brazil; criminal organisations; corruption; \emph{jogo do bicho}; private governance
 
@@
 In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
-Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, making the game similar to a common raffle.
-But soon visitors could name their animals of choice.
-This change made the game more appealing and lasts until this day \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.
-The lottery was labelled as the \emph{jogo do bicho}, or the animal game, and it was immediately adopted by the public.
+Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{The zoo staff first distributed the tickets at random, but soon visitors were allowed to select their animals of choice.
+This change made the game considerably more appealing to bettors \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.\footnote{The amount was higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.} The lottery was labelled as the \emph{jogo do bicho}, or the animal game, and it was immediately adopted by the public.
 Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in stores across Rio de Janeiro.
@@
 Civil servants and police officers criminalised the \emph{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}.
-The campaign extended for several decades and received considerable support from the \emph{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after the creation of the animal game \citep[82]{da1999aguias}.
+The campaign extended for several decades and received considerable support from the \emph{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}.
 The game was officially banned in 1946 and remains illegal to this day.
@@
 But more than a act of defiance, the \emph{jogo do bicho} is a successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
-A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \emph{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \url{http://goo.gl/9kNeX8} and \url{http://goo.gl/8FSAZl} (in Portuguese).
-Access: December 2016.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \emph{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the Brazilian workforce.
-See \url{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}.
-Access: December 2016.}
+A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \emph{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
+Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \emph{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the Brazilian workforce.
+See \href{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}{\texttt{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}}.
+Access: August 2018.}
 
 In this article I offer a rational choice interpretation of the \emph{jogo do bicho}.
-More specifically, I discuss how \emph{bicheiros} adopt rational strategies to solve collective action problems and to maximise their political leverage.
+More specifically, I discuss how \emph{bicheiros} adopt rational strategies to solve collective action problems and maximise their political leverage.
 I use the case of Rio de Janeiro to illustrate the lottery's profit abilities and its effect in political corruption.
+The lottery financiers have long been sponsors of Rio de Janeiro's world-famous Carnival Parade, and several of the country's most high-profile \emph{bicheiros} hail from that state.
+Furthermore, in no other part of Brazil the patron-client relationship between \emph{bicheiros} and politicians has been so widely publicised, with game financiers openly funding cultural activities to co-opt local politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
 
@@
 For instance, \citet{gambetta1996sicilian} examines the strategies used by the Sicilian Mafia to settle disputes among their members and enforce rules in the areas they exercise control.
-\citet{leeson2007arrgh,leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations.
+\citet{leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations.
 \citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking and coordinating productive activities in California prison gangs.
@@
 
-Second, this work relates to the literature on repugnant transactions and the relationship between morality and the market \citep{boettke1995morality, roth2007repugnance, sandel2012money, satz2010some, simmel1900geldes, zelizer1979morals}.
-In the following sections, I claim that the Brazilian elites have attached pejorative meaning to the \emph{jogo do bicho} to constrain the gambling market.
-I provide evidence that \emph{bicheiros} were aware of this problem, and as a response, they devised a series of rules aimed at reducing the costs associated with repugnance \citep{labronici2014sorteio, magalhaes2005ganhou}.
-\emph{Bicheiros} have made considerable efforts to increase the levels of trust in the system and distance themselves from other types of illegal activities.
-Their main tool to increase credibility was costly signalling, that is, the \emph{bicheiros} hoped the public would see them as credible brokers by sacrificing their immediate interests  \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}.
+Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,connelly2011signaling,spence1973job}.
+I provide evidence that \emph{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and reduce the uncertainty associated with clandestine markets.
+Their main tool to increase credibility was costly signalling \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}.
+\emph{Bicheiros} believed that by sacrificing their immediate interests they could gain a reputation of honesty that would benefit them in the long run.
 
-Lastly, this work connects to the literature on state capture, which is among the most important topics in public choice theory \citep{hellman2003seize, rose1978corruption, rose1999corruption, shleifer2002grabbing, tollison1982rent}.
+Lastly, this work connects to the literature on state capture, which is among the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}.
 More specifically, I use the Brazilian case to illustrate how politicians and civil servants can be co-opted by criminal groups and produce sub-optimal social outcomes.
@@
 
-\bibliography{references}
+\bibliography{bibliography}
 \bibliographystyle{apalike}

### 2018-09-20 paper-bicho/bicho.tex
update intro3

@@
 
-Yet, the \emph{jogo do bicho} has survived.
+Yet the \emph{jogo do bicho} has survived.
 The animal game has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{jupiara2015poroes}.
@@
 I use the case of Rio de Janeiro to illustrate the lottery's profit abilities and its effect in political corruption.
-The lottery financiers have long been sponsors of Rio de Janeiro's world-famous Carnival Parade, and several of the country's most high-profile \emph{bicheiros} hail from that state.
-Furthermore, in no other part of Brazil the patron-client relationship between \emph{bicheiros} and politicians has been so widely publicised, with game financiers openly funding cultural activities to co-opt local politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
+The lottery financiers have long been major sponsors of Rio de Janeiro's world-famous Carnival Parade, and several of the country's most high-profile \emph{jogo do bicho} entrepreneurs hail from that state.
+Furthermore, in no other part of Brazil the patron-client relationship between \emph{bicheiros} and politicians has been so widely publicised.
+
+ to co-opt local politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
 

### 2018-09-20 paper-bicho/bicho.tex
changes sec1

@@
 In this article I offer a rational choice interpretation of the \emph{jogo do bicho}.
-More specifically, I discuss how \emph{bicheiros} adopt rational strategies to solve collective action problems and maximise their political leverage.
+More specifically, I discuss how \emph{bicheiros} adopt rational, yet often repressive, strategies to solve collective action problems and maximise their political leverage.
 I use the case of Rio de Janeiro to illustrate the lottery's profit abilities and its effect in political corruption.
-The lottery financiers have long been major sponsors of Rio de Janeiro's world-famous Carnival Parade, and several of the country's most high-profile \emph{jogo do bicho} entrepreneurs hail from that state.
-Furthermore, in no other part of Brazil the patron-client relationship between \emph{bicheiros} and politicians has been so widely publicised.
-
- to co-opt local politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
+In no other part of Brazil the patron-client relationship among \emph{bicheiros}, the civil society, and politicians has been so widely publicised.
+The \emph{jogo do bicho} entrepreneurs have long been involved in Rio's cultural and financial activities, mainly through the funding of `samba schools' and the Carnival Parade.
+As I describe below, the parade has played a significant role not only in strenghtening the ties between the \emph{jogo do bicho} and poor urban communities, but also in coopting politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
+Furthermore, Brazil's fragmented electoral system creates strong incentives for candidates to use resources from, and ultimately depend on, the lottery's rich sponsors.
 
@@
 
-\section{\emph{Jogo do Bicho} as an Emergent Institution}
+\section{Historical Background}
 \label{sec:institution}
 
-\subsection{Historical Background}
-\label{sub:background}
-
-The early history of the \emph{jogo do bicho} is a textbook example of spontaneous order.
-Spontaneous orders are emergent macro-level phenomena that result from voluntary actions of purposive, self-interested individuals utilising their contextual knowledge \citep{boettke1990theory, boettke2005methodological, hayek1945use, hayek1960constitution, hayek1973law, leeson2008coordination, menger1871grundsatze, polanyi1948planning, polanyi1951logic}.
-Drummond, the Zoological Garden's original owner, designed the basic framework for the \emph{jogo do bicho}; but independent bookmakers were the ones who popularised the game \citep[77]{magalhaes2005ganhou}.
-Ticket sellers could quickly respond to market signals and then allocate their products where they were more valuable because of the lack of central coordination.
-Moreover, competition among sellers fostered innovation, and the \emph{bicheiros} invented new game rules to make the lottery more appealing to their customers \citep[61]{mello1989historia}.
-In this sense, the animal game is the materialisation of an evolutionary process of entrepreneurial discovery in which the interactions that provided the highest value to consumers were preserved over time \citep{boettke2008gordon, boettke2014entrepreneurship, buchanan1964should, hayek1978competition, kirzner1997entrepreneurial}.
-
-However, the \emph{jogo do bicho} only emerged because of historically contingent circumstances.
 The late nineteenth-century Brazil had four characteristics that explain how the animal game came to being: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
-Figure \ref{fig:dag} presents a simple directed acyclic graph (DAG) \citep{pearl2009causality} that shows the relationships between these explanatory variables and the development of the \emph{jogo do bicho}.\footnote{The main purpose of direct acyclic graphs is to graphically display the possible links among the exposure variables, confounders, and outcomes \citep{morgan2014counterfactuals, pearl2009causality}.
-DAGs are transparent by definition, as all theoretical choices made by the researcher are stated explicitly in the model.
-Each single-headed arrow in a DAG indicates that the variable at the origin causes the variable at the end of the directed edge.
-Dashed edges suggest that two variables are jointly dependent on unobserved common causes.
-There are no assumptions regarding the functional form of the relationships, and unless mentioned otherwise, the arrows represent fully non-parametric associations.
-Variables between two nodes are mediators, and variables pointed at by two or more factors have multiple causes.
-They are called \emph{colliders}.
-For the sake of clarity, errors are assumed independent and often excluded from the graphs.
-For an accessible introduction to DAGs see \citet[chap.
-3--4]{morgan2014counterfactuals} and \citet{pearl2016causal}.}
 
 I start with the impact of urban poverty on the animal game.
-Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, fausto2014concise, naro1992revision, skidmore1993black}.
+Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, skidmore1993black}.
 The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration, smith1979ethnic}.
 Nevertheless, the job market tightened considerably after the \emph{Encilhamento} financial crisis of 1891 \citep{topik2014political, triner2005baring}.
-During the economic downturn, the informal economy was an obvious destination for the urban poor.
-Given its widespread popularity, the \emph{jogo do bicho} attracted hopeful entrepreneurs, either Brazilian or foreign-born, who could not enter the formal labour force.\footnote{The underground economy was also more democratic than the formal sector.
+During the economic downturn, the informal economy was the obvious destination for the urban poor.
 As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor.
-These vendors used to sell different types of merchandise and many of them would later offer \emph{jogo do bicho} tickets.}
+Given the \emph{jogo do bicho} rising popularity, many of those vendors would sell lottery tickets along other types of merchandise.
 
 The immigration also influenced the \emph{jogo do bicho} via social ties.
-Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{klein1994imigraccao, lobo2001imigraccao, trento1989outro}.
+Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}.
 Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}.
-Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, torcato2011repressao, villar2008contravencao}.
+Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, villar2008contravencao}.
 Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \emph{jogo do bicho}'s formative years.
@@
 Some years later, the animal game would be available not only across the city of Rio de Janeiro but throughout Brazil \citep[76]{da1999aguias}.
-
-The country's lax financial policy might correlate with poverty through unspecified factors.
-For instance, political decisions may have caused inadvertently both poverty and the expansion of the monetary base \citep{mattos2013shantytown, schmidt1982modernization}; alternatively, external events such as institutional instability \citep{costantini2014index, fausto2014concise, luna2014economic} or commodity shocks \citep{musacchio2014colonial} could be the cause of those two variables.
-There is not enough evidence to discard such scenarios.
-To illustrate this uncertainty, the two nodes, namely, money supply and urban poverty, are connected with a dashed edge in Figure \ref{fig:dag}.
 

### 2018-09-21 paper-bicho/bicho.tex
update sec2

@@
 \author{Danilo Freire\thanks{Postdoctoral Research Associate, Political Theory Project, Brown University, 8 Fones Alley, Providence, RI 02912, \href{mailto:danilofreire@brown.edu}{\texttt{danilofreire@brown.edu}}, \href{http://danilofreire.com}{\texttt{http://danilofreire.com}}.
-I would like to thank Toke Aidt, Paulo Roberto Araujo, Jose Bola\~{n}os, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, Graham Denyer Willis, the editors of Latin American Perspectives, and four reviewers for their valuable comments.
-I gratefully acknowledge the support of the National Council for Scientific and Technological Development.}
+I would like to thank Toke Aidt, Paulo Roberto Araujo, Jose Bola\~{n}os, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, Graham Denyer Willis, the participants of academic seminars at King's College London, the University of Cambridge and the Prometheus Institut in Berlin, the editors of Latin American Perspectives, and four reviewers for their valuable comments.
+I gratefully acknowledge the financial support of the National Council for Scientific and Technological Development.}
 }
@@
 
-The late nineteenth-century Brazil had four characteristics that explain how the animal game came to being: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
+The late nineteenth-century Brazil had four characteristics that explain why the animal game became so popular: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
 
@@
 Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, skidmore1993black}.
-The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration, smith1979ethnic}.
-Nevertheless, the job market tightened considerably after the \emph{Encilhamento} financial crisis of 1891 \citep{topik2014political, triner2005baring}.
-During the economic downturn, the informal economy was the obvious destination for the urban poor.
-As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor.
-Given the \emph{jogo do bicho} rising popularity, many of those vendors would sell lottery tickets along other types of merchandise.
+The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration}.
+However, their hopes would soon be frustrated.
+The Brazilian job market suffered a severe contraction due to the \emph{Encilhamento} financial crisis of 1891, and the economic downturn aggravated the already difficult conditions of the working classes \citep{topik2014political, triner2005baring}.
 
-The immigration also influenced the \emph{jogo do bicho} via social ties.
+In that regard, the informal economy became the only source of income for large swathes of the urban population.
+As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor, as it had low barriers of entry and required little technical skills.
+These vendors also sold \emph{jogo do bicho} tickets to complement their income, and as the game gained a following in Brazil's First Republic it became relatively more important in the extra-legal economy.
+
+Immigration also influenced the \emph{jogo do bicho} via social ties.
 Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}.
 Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}.
-Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, villar2008contravencao}.
+Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, villar2008contravencao}.
 Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \emph{jogo do bicho}'s formative years.
@@
 The abolition of slavery and the growing industrialisation of Brazil increased the amount of capital available in the country \citep{franco1987reformas, schulz2008financial}.
-The 1888 Banking Act gave extra liquidity to local financial markets, and the \emph{jogo do bicho} entrepreneurs utilised that increase in the monetary base to extend the scope of their business.
-Some years later, the animal game would be available not only across the city of Rio de Janeiro but throughout Brazil \citep[76]{da1999aguias}.
+Moreover, the 1888 Banking Act gave extra liquidity to local financial markets, and the \emph{jogo do bicho} entrepreneurs used that increase in the monetary base to expand their businesses.
+Credit became more widely available in urban areas, and it gave a temporary boost to individual incomes in cities like Rio de Janeiro and São Paulo.
+Therefore, in a few years the animal game would be available not only in Rio but throughout Brazil \citep[76]{da1999aguias}.
 
@@
 \citet[69--100]{chazkel2011laws} notes that until the 1940s police district chiefs operated within a large margin of discretion and repression against bookmakers was idiosyncratic.
-Prosecution against the \emph{bicheiros} had hardened in 1917, but only in 1946, when the federal government banned all gambling activities in the country \citep[155--156]{magalhaes2005ganhou}, the law was consistently enforced.
+In the early years of \emph{jogo do bicho}, lottery `bankers' were allowed to operate virtually free from police interference, what surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}.
+Prosecution against the \emph{bicheiros} hardened in 1917 after the promulgation of the Civil Code, but it was only in 1946 that the federal government effectively banned gambling in the country \citep[155--156]{magalhaes2005ganhou}.
+
+% Add paragraph about state coercion.
 

### 2018-10-11 paper-bicho/bicho.tex
update

--- 2018-09-20 paper-bicho/bicho.tex
@@
 \usepackage{libertine}
+\usepackage[libertine]{newtxmath}
 \usepackage[scaled=.9]{inconsolata}
@@
 \usepackage{setspace}
-\usepackage[top=2.0cm,bottom=2.0cm,left=2.2cm,right=2.2cm]{geometry}
+\usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
 \usepackage[backref,pagebackref]{hyperref}
@@
 
-\title{\textbf{Beasts of Prey or Rational Animals?\\ Private Governance in Brazil's \emph{Jogo do Bicho}}}
+\title{\textbf{Beasts of Prey or Rational Animals?\\ Private Governance in Brazil's \textit{Jogo do Bicho}}}
 
 \author{Danilo Freire\thanks{Postdoctoral Research Associate, Political Theory Project, Brown University, 8 Fones Alley, Providence, RI 02912, \href{mailto:danilofreire@brown.edu}{\texttt{danilofreire@brown.edu}}, \href{http://danilofreire.com}{\texttt{http://danilofreire.com}}.
-I would like to thank Toke Aidt, Paulo Roberto Araujo, Jose Bola\~{n}os, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, Graham Denyer Willis, the editors of Latin American Perspectives, and four reviewers for their valuable comments.
-I gratefully acknowledge the support of the National Council for Scientific and Technological Development.}
+I would like to thank Toke Aidt, Paulo Roberto Araujo, Jose Bolaños, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, Graham Denyer Willis, the participants of academic seminars at King's College London, the University of Cambridge and the Prometheus Institut in Berlin, the editors of Latin American Perspectives, and four reviewers for their valuable comments.
+I gratefully acknowledge the financial support of the National Council for Scientific and Technological Development.}
 }
@@
 \noindent
-This article presents a rational choice account of Brazil's \emph{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
-Over 120 years, the \emph{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and has been a major driver of the persistent political corruption in the country.
-First, I examine the mechanisms that fostered \emph{jogo do bicho}'s notable growth outside of the boundaries of the law.
-Second, I explain how the lottery financiers combine costly signals, selective incentives, and innovative uses of government institutions to establish their reputation as honest brokers.
-Lastly, I analyse the relationship between the game sponsors and political actors and discuss how the game financiers influence the Brazilian state.
+This article presents a rational choice account of Brazil's \textit{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
+Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and the game has been a major driver of political corruption in Brazil.
+First, I examine the mechanisms that fostered the \textit{jogo do bicho}'s notable growth outside of the boundaries of the law.
+Second, I explain how the lottery financiers combine costly signals and selective incentives to foster external cooperation from members of community.
+Lastly, I analyse the relationship between the lottery sponsors and Brazilian representatives, specially the ways by which the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term goals.
 
@@
  \noindent
- \textbf{Keywords}: Brazil; criminal organisations; corruption; \emph{jogo do bicho}; private governance
+ \textbf{Keywords}: Brazil; criminal organisations; corruption; \textit{jogo do bicho}; private governance
 
@@
 In 1892, Baron João Batista de Viana Drummond came up with a new idea to fund his cash-strapped zoo.
-Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \emph{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city.
+Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \textit{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city.
 But it lacked visitors.
-As an experienced businessman, Drummond soon realised the zoo would have to provide other kinds of entertainment to keep itself afloat.
-Among his suggestions, one seemed particularly promising: a lottery raffle.
+An experienced businessman, Drummond realised the zoo would have to provide other kinds of entertainment to keep itself afloat.
+One of his plans seemed particularly promising: a lottery raffle.
 
-The rules were straightforward.
+The rules were simple.
 In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
-Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{The zoo staff first distributed the tickets at random, but soon visitors were allowed to select their animals of choice.
-This change made the game considerably more appealing to bettors \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.\footnote{The amount was higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.} The lottery was labelled as the \emph{jogo do bicho}, or the animal game, and it was immediately adopted by the public.
-Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in stores across Rio de Janeiro.
-Unexpectedly, the Baron created a whole new gambling market.
+Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, but they soon allowed participants to choose the animals they preferred.
+This small change made the game considerably more appealing to the public, who often resorted to clairvoyants and fortune tellers to guess the winning animals \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.
+The Baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
+Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in many stores across Rio de Janeiro.
+What was once a small raffle soon became a large gambling market of its own.
 
-The \emph{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace.
-A network of street bookmakers, called \emph{bicheiros}, made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the clandestine numbers game  \citep[37]{chazkel2011laws}.
-The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today {[}1895{]}, in Rio de Janeiro, the game is everything.
-{[}\ldots{}{]} Nobody works!
+A \textit{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace.
+A network of street bookmakers, called \textit{bicheiros}, made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the numbers game \citep[37]{chazkel2011laws}.
+The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today $[1985]$ in Rio de Janeiro, the game is everything.
+$[\dots]$  Nobody works!
 Everybody plays' \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.}
@@
 But this tolerant state of affairs did not last.
-Civil servants and police officers criminalised the \emph{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}.
-The campaign extended for several decades and received considerable support from the \emph{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}.
-The game was officially banned in 1946 and remains illegal to this day.
+Civil servants and police officers criminalised the \textit{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}.
+The campaign extended for several decades and received considerable support from the \textit{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}.
+The Brazilian government officially banned the \textit{jogo do bicho} in 1946 and it remains illegal until this day.
 
-Yet the \emph{jogo do bicho} has survived.
-The animal game has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{jupiara2015poroes}.
-But more than a act of defiance, the \emph{jogo do bicho} is a successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
-A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \emph{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
-Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \emph{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the Brazilian workforce.
+Yet the game has survived.
+The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{jupiara2015poroes}.
+But more than a act of defiance, the \textit{jogo do bicho} is a successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
+A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
+Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the Brazilian workforce.
 See \href{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}{\texttt{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}}.
@@
 
-In this article I offer a rational choice interpretation of the \emph{jogo do bicho}.
-More specifically, I discuss how \emph{bicheiros} adopt rational, yet often repressive, strategies to solve collective action problems and maximise their political leverage.
-I use the case of Rio de Janeiro to illustrate the lottery's profit abilities and its effect in political corruption.
-In no other part of Brazil the patron-client relationship among \emph{bicheiros}, the civil society, and politicians has been so widely publicised.
-The \emph{jogo do bicho} entrepreneurs have long been involved in Rio's cultural and financial activities, mainly through the funding of `samba schools' and the Carnival Parade.
-As I describe below, the parade has played a significant role not only in strenghtening the ties between the \emph{jogo do bicho} and poor urban communities, but also in coopting politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
-Furthermore, Brazil's fragmented electoral system creates strong incentives for candidates to use resources from, and ultimately depend on, the lottery's rich sponsors.
+In this article I offer a rational choice interpretation of the \textit{jogo do bicho}.
+More specifically, I use an array of bibliographical sources to show how the game operators, called \textit{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.
+The tactics are unusual and often repressive, such as bribing corrupt police officers to threaten undisciplined employees.
+Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive at first, they do address the long-term needs of the \textit{jogo do bicho} financiers.
+
+Like any other business manager, \textit{bicheiros} have to run their firm with low costs to increase profits.
+However, the fact that the \textit{jogo do bicho} is clandestine imposes additional difficulties for its operation.
+In particular,  \textit{bicheiros} face two main challenges to keep the lottery running.
+First, they need to gather public support so that bettors are not discouraged to engage in the lottery despite it being illegal.
+Second, \textit{bicheiros} also have to ensure that the state repression is not prohibitively costly to their business, otherwise they would be better off by shutting it down.
+I argue below that the \textit{bicheiros} have succeeded in both by using carefully-designed reputation strategies and employing costly signals to the communities they serve.
+
+I use the case of Rio de Janeiro to illustrate how the \textit{jogo do bicho} has overcome the obstacles to its expansion.
+Rio is a particularly interesting case because in no other part of Brazil the game financiers established such an effective patronage network.
+\textit{Bicheiros} have sponsored political campaigns, financed cultural activities and football teams, and sometimes even run in local elections themselves.
+I discuss the ways by which the \textit{bicheiros} have exploited fragilities of the Brazilian political system to their advantage and how those practices have weakened Brazilian democracy.
 
@@
 \citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking and coordinating productive activities in California prison gangs.
-I argue that \emph{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and foster trust among criminals.
-Moreover, I also describe how \emph{bicheiros} have developed sophisticated financial mechanisms, such as informal hedging operations and risk-sharing contracts, to prevent predatory behaviour in their community.
+I argue that \textit{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and foster trust in the community.
 
 Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,connelly2011signaling,spence1973job}.
-I provide evidence that \emph{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and reduce the uncertainty associated with clandestine markets.
+I provide evidence that \textit{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and reduce the uncertainty associated with clandestine markets.
 Their main tool to increase credibility was costly signalling \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}.
-\emph{Bicheiros} believed that by sacrificing their immediate interests they could gain a reputation of honesty that would benefit them in the long run.
+\textit{Bicheiros} believed that by sacrificing their immediate interests they could gain a reputation of honesty that would benefit them in the long run.
 
-Lastly, this work connects to the literature on state capture, which is among the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}.
+Lastly, this work connects to the literature on state capture, which is one of the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}.
 More specifically, I use the Brazilian case to illustrate how politicians and civil servants can be co-opted by criminal groups and produce sub-optimal social outcomes.
-\citet{queiroz1992carnaval} explored why \emph{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities.
+\citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities.
 \citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested that the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s.
-In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \emph{jogo do bicho} and the military regime in Brazil (1964--1985).
+In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \textit{jogo do bicho} and the military regime in Brazil (1964--1985).
 I supplement this literature by highlighting how asymmetrical information, agency dilemmas, and rent-seeking behaviour offer convincing explanations to the issues presented above.
-Although those concepts have a long tradition in public choice, scholars have not applied those ideas thus far to understand the dynamics of the \emph{jogo do bicho}.
+Although those concepts have a long tradition in public choice, scholars have not applied those ideas thus far to understand the dynamics of the \textit{jogo do bicho}.
 By doing so, I integrate seemingly contradictory historical facts into a single narrative that connects micro-level decisions to macro-level outcomes.
 
-The remainder of this article proceeds as follows.
-Section \ref{sec:institution} presents a brief historical overview of the \emph{jogo do bicho}.
-It describes the necessary conditions for the emergence of the game and presents its basic organisational structure.
-Section \ref{sec:governance} details the \emph{jogo do bicho}'s governance mechanisms, particularly the strategies employed by vendors to increase trust in markets that operate at the margins of the law.
-Section \ref{sec:capture} discusses the links between illegal gambling markets, samba schools and the Brazilian state.
-Section \ref{sec:conclusion3} offers some concluding remarks.
+\section{An Overview of the \textit{Jogo do Bicho}}
+\label{sec:overview}
 
-\section{Historical Background}
-\label{sec:institution}
+\subsection{Historical Background}
+\label{sub:historical_background}
 
-The late nineteenth-century Brazil had four characteristics that explain how the animal game came to being: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
+The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
+I discuss each of these elements below.
 
 I start with the impact of urban poverty on the animal game.
-Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, skidmore1993black}.
-The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration, smith1979ethnic}.
-Nevertheless, the job market tightened considerably after the \emph{Encilhamento} financial crisis of 1891 \citep{topik2014political, triner2005baring}.
-During the economic downturn, the informal economy was the obvious destination for the urban poor.
-As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor.
-Given the \emph{jogo do bicho} rising popularity, many of those vendors would sell lottery tickets along other types of merchandise.
+Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising.
+Brazil's growing cities offered a number of new occupations for freed slaves, and many were eager to live far from their former masters \citep{andrews1991blacks, skidmore1993black}.
+The freed slaves were joined by increasing numbers of Asian and European immigrants who were promised access to farming lands in the country \citep{hall1969origins, lesser2013immigration}.
+However, the hopes of the African-Brazilians and the new settlers would soon be frustrated by a series of economic downturns.
+The Brazilian labour market suffered a severe contraction in the wake of the \textit{Encilhamento} financial crisis of 1891, and the economic instability aggravated the already difficult conditions of the working classes \citep{topik2014political, triner2005baring}.
 
-The immigration also influenced the \emph{jogo do bicho} via social ties.
+In that regard, large swathes of the urban population turned to the informal economy.
+As \citet[115]{chazkel2011laws} observes, there were few occupations available to poor women and foreigners in the 1890s, and many of them became street vendors.
+The profession requires little technical skills and has low barriers of entry, but it can be very profitable if for whatever reason there is a strong demand for particular product.
+The \textit{jogo do bicho} was one of those products that offered a high rate of return.
+The game was simple to operate and that simplicity attracted more people willing to try their luck.
+As the game gained a following in Brazil's First Republic, it comprised an important share of the extra-legal economy in Rio de Janeiro.
+
+Immigration also influenced the \textit{jogo do bicho} via social ties.
 Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}.
 Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}.
-Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, villar2008contravencao}.
-Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \emph{jogo do bicho}'s formative years.
+Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil} and in the \textit{jogo do bicho} in particular \citep{magalhaes2005ganhou, villar2008contravencao}.
+Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \textit{jogo do bicho}'s formative years.
 
@@
 The abolition of slavery and the growing industrialisation of Brazil increased the amount of capital available in the country \citep{franco1987reformas, schulz2008financial}.
-The 1888 Banking Act gave extra liquidity to local financial markets, and the \emph{jogo do bicho} entrepreneurs utilised that increase in the monetary base to extend the scope of their business.
-Some years later, the animal game would be available not only across the city of Rio de Janeiro but throughout Brazil \citep[76]{da1999aguias}.
+Moreover, the 1888 Banking Act gave extra liquidity to local financial markets, what made credit more widely available in cities like São Paulo and Rio de Janeiro.
+Individuals received a temporary boost in personal income, a part of which they spent on leisure activities such as the \textit{jogo do bicho}.
+Moreover, the lottery attracted new entrants as it became more profitable, and in only a few years similar versions of the animal game were available throughout Brazil \citep[79]{da1999aguias}.
 
-The last necessary condition for the emergence of the \emph{jogo do bicho} is weak law enforcement.
+The last necessary condition for the emergence of the \textit{jogo do bicho} is weak law enforcement.
 \citet[69--100]{chazkel2011laws} notes that until the 1940s police district chiefs operated within a large margin of discretion and repression against bookmakers was idiosyncratic.
-Prosecution against the \emph{bicheiros} had hardened in 1917, but only in 1946, when the federal government banned all gambling activities in the country \citep[155--156]{magalhaes2005ganhou}, the law was consistently enforced.
+In the early years of \textit{jogo do bicho}, lottery `bankers' were allowed to operate virtually free from police interference, what surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}.
+Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code and in 1941 the animal game was banned.\footnote{See: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del3688.htm}} (in Portuguese).
+Access: September 2017.} Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.
+The text can be read at: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm}} (in Portuguese).
+Access: September 2017.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players.
+\citep{fsp2016legalizarbicho}.
+
+Since the mid-twentieth century, the \textit{jogo do bicho} has been in a state of `semi-legality' in Brazil.
+For all purposes, the \textit{jogo do bicho}  remains illegal in the country, yet there is no Brazilian city which does not have its own group of local \textit{bicheiros}.
+In that regard, the \textit{jogo} does not operate exclusively at the margins of the Brazilian law, but it consciously exploits and subsidises large sectors of the formal economy according to the interests of wealthy game sponsors.
+For instance, the \textit{jogo do bicho} has benefited from the
 
@@
 
-The animal game has three levels of hierarchy.
-At the bottom are the \emph{bicheiros}, who are those in charge of selling \emph{jogo do bicho} tickets \citep{chazkel2007beyond, chazkel2011laws, da1999aguias, labronici2014sorteio, magalhaes2005ganhou, misse2007illegal}.
-\emph{Bicheiros} are the most visible part of the \emph{jogo do bicho} structure.
-The bookmakers often build their vending stands inside the premises of a local shop, such as a small grocery store or a pub, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}.
-\emph{Bicheiros} usually work alone, but they may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
+The animal game operates with three levels of hierarchy.
+At the bottom level are the \textit{bicheiros}, those in charge of selling \textit{jogo do bicho} tickets \citep{chazkel2007beyond, da1999aguias}.
+\textit{Bicheiros} are the most visible part of the \textit{jogo do bicho} structure.
+The bookmakers usually build their vending stands inside the premises of a local shop, such as a small grocery store, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}.
+\textit{Bicheiros} usually work alone, but may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
 
-The \emph{gerentes} (managers) oversee all \emph{jogo do bicho} stands in a given area.
+The \textit{gerentes} (managers) oversee all \textit{jogo do bicho} stands in a given area.
 Their task is akin to that of a firm accountant.
-Gerentes control the cash flow between the \emph{bicheiros} and the bankers, manage the payroll of the employees, and provide financial information to the top members of the organisation.
+Gerentes control the cash flow between the \textit{bicheiros} and the bankers, manage the payroll of the employees, and provide financial information to the top members of the organisation.
 They also supervise individuals who carry menial tasks in the business, transfer money to other gambling branches and double-check the balance sheets of the betting sites \citetext{\citealp[71]{labronici2012paratodos}; \citealp[142]{misse2007illegal}}.
 
-The \emph{banqueiros}, or the Portuguese for bankers, occupy the top position in the \emph{jogo do bicho} hierarchy, comprising the small financial elite of the game.
-A 2012 report by the Brazilian Federal Police affirmed that 10 \emph{banqueiros} controlled the market throughout the country; five of them based in the state of Rio de Janeiro \citep{globo2012contraventores}.
+The \textit{banqueiros}, or the Portuguese for bankers, occupy the top position in the \textit{jogo do bicho} hierarchy.
+They comprise the small financial elite of the game.
+A 2012 report by the Brazilian Federal Police affirmed that 10 \textit{banqueiros} controlled the market throughout the country; five of them based in the state of Rio de Janeiro \citep{globo2012contraventores}.
 Apart from funding the game, the bankers provide support for the employees to undertake their activities.
-The \emph{banqueiros}' main attributions include paying bribes to police personnel, bailing out sellers arrested by security forces, and offering judicial assistance to employees in case of legal persecution \citep[75]{labronici2012paratodos}.
+The \textit{banqueiros}' main attributions include paying bribes to police personnel, bailing out sellers arrested by security forces, and offering judicial assistance to employees in case of legal persecution \citep[75]{labronici2012paratodos}.
 
-\emph{Banqueiros} run their businesses from fortified houses in unknown locations, the \emph{fortalezas} (`forts').
-The first \emph{fortalezas} likely appeared in the 1950s, when the animal game was already well-established across the Brazilian territory.
-The period coincides with a time when the \emph{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}.
-Due to the growing size of the \emph{jogo do bicho} economy, \emph{banqueiros} decided to move their operations away from the public to avoid police persecution and reduce coordination costs.
+\textit{Banqueiros} run their businesses from fortified houses in unknown locations, the \textit{fortalezas} (`forts').
+The first \textit{fortalezas} likely appeared in the 1950s, when the animal game was already well-established across the Brazilian territory.
+The period coincides with a time when the \textit{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}.
+Due to the growing size of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and reduce coordination costs.
 
-Although the forts provided safety to the bankers, the existence of those hideouts posed a challenge to the organisation.
-Bankers removed from the public view are not accountable to players and booking agents.
-Similarly, bankers and managers working in the \emph{fortalezas} cannot oversee their employees as effectively as before.
-Considering that the animal game itself is illegal and the amount of money involved in the bets is often substantial, both \emph{banqueiros} and booking agents have strong incentives to defect.
-Players, in turn, have no evident reason to trust \emph{banqueiros} or \emph{bicheiros}.
-How do \emph{jogo do bicho} agents overcome trust issues and cooperate under uncertainty?
+\textit{Banqueiros} solve problems of internal cooperation by providing club goods \citep{buchanan1965economic} while simultaneously shunning cheaters through selective punishments \citep{bo2005cooperation, roth1978equilibrium}.
+The first club good offered to \textit{bicheiros} by their bosses is private security.
+As the game is illegal, street sellers cannot rely on official institutions to protect themselves.
+Thus, the game bankers have built an extensive network of gunmen and bribed police officers to protect their employees from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
 
-I argue that the \emph{jogo do bicho} solves problems of internal cooperation by providing club goods \citep{buchanan1965economic, berman2008religion, berman2009radical, leeson2011government, roth2014prison} while simultaneously shunning cheaters through punishments and appeals to `the shadow of the future' \citep{axelrod1984evolution, axelrod1985achieving, bo2005cooperation, roth1978equilibrium}.
-Clients and \emph{bicheiros} cooperate based on trust-enhancing mechanisms, most of them devised specifically for the \emph{jogo do bicho} \citep{da1999aguias, magalhaes2005ganhou}.
-Such mechanisms are relevant because they have allowed the \emph{jogo do bicho} to distance itself from other shadow markets and become a profitable enterprise in the long run.
-
-\section{Governance of the \emph{Jogo do Bicho}}
-\label{sec:governance}
-
-\subsection{Gambling Markets and Repugnant Transactions}
-\label{sub:repugnance}
-
-The \emph{jogo do bicho} is a repugnant market.
-Individuals that like to gamble cannot do so because of strong moral objections from outsiders \citep{brisset2016marche, roth2007repugnance, satz2010some, zelizer1979morals}.
-As early as in 1890, Brazilian public authorities positioned themselves against the \emph{jogo do bicho} arguing that `{[}\ldots{}{]} this type of amusement is prejudicial to the interests of the unwise, who are naively seduced by the deceptive hope of uncertain lucre' \citep[544]{chazkel2007beyond}.
-In 1941, the government banned the animal game;\footnote{See: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm}{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del3688.htm} (in Portuguese).
-Access: December 2016.} five years later, it prohibited all games of chance.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.
-The text can be read at: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm} (in Portuguese).
-Access: December 2016.} The \emph{jogo do bicho}, casinos and bingos remain illegal in the country.
-Recent estimations show that the prohibition of the \emph{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players.
-\citep{congressoemfoco2015bicho, fsp2016legalizarbicho}.
-
-In contrast with the official statements, the noxious element of the \emph{jogo do bicho} does not come from its inherent randomness.
-The game is `repugnant' precisely because it is \emph{a market}, a setting in which individuals can monetise the entertainment for private profit \citep{chazkel2007beyond, chazkel2011laws}.
-The Brazilian state has never seen any contradiction between banning games of chance and running a national lottery company of its own; even the Catholic Church, which has long condemned the \emph{jogo do bicho}, frequently organises raffles to fund its activities \citep[49]{abreu1996imperio, magalhaes2005ganhou}.
-Only after the introduction of private money that Brazilians objected the idea of benefiting from someone else's bad luck.
-
-In this sense, the main obstacle that confronted \emph{bicheiros} was to convince others that the animal game would not cause the Brazilian society `to slide down a slippery slope to genuinely repugnant transactions' \citep[45]{roth2007repugnance} such as prostitution or debt bondage.
-As the century-old history of the game can attest, \emph{bicheiros} have succeeded in this task.
-But how?
-The literature on repugnant costs tell us little about how markets transition from noxious to tolerated.
-Here I posit two mechanisms that reduced the stigma associated with the game: 1) \emph{a strong reputation of honesty} expressed by costly signals from sellers, and 2) the provision of \emph{selective incentives} for both clients and booking agents.
-Below, I offer evidence that these two factors allowed the animal game to reach its current semi-legal status in Brazil.
-
-\subsection{External Cooperation}
-\label{sub:external}
-
-Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and experimental studies \citep{dawes1977behavior, isaac1984divergent, kim1984free, marwell1981economists} have both demonstrated that long-term cooperation is possible whenever players expect future pay-offs to be higher than present ones.
-Fear of retaliation induces individuals not to cheat.
-Nevertheless, illegal organisations tend to discount the future even more heavily than the other groups, what makes cooperative behaviour among criminals uncommon \citep{gambetta2009codes, skarbek2011governance,skarbek2012prison,skarbek2014social}.
-The \emph{jogo do bicho} is an exception to this rule.
-The market properties of the game and inconsistent repression by Brazilian authorities have permitted \emph{bicheiros} to overcome the stigma of repugnance and improve the game's long-term profitability.
-
-The \emph{jogo do bicho} entrepreneurs have made considerable efforts to present themselves as honest brokers.
-The first trust-enhancing mechanism they have employed to foster external cooperation was the use of a \emph{fixed-multiplier formula} for pay-outs.
-It works as follows.
-If a player wins the lowest prize of the animal game, he or she receives 18 times his/her investment regardless of the size of the bet.
-Bigger prizes naturally offer higher returns; a lucky winner of the top prize wins up to 4,000 times the value of his/her bet \citetext{\citealp[89]{labronici2012paratodos}; \citealp[20]{magalhaes2005ganhou}}.
-
-This stands in sharp contrast to the common practice of sharing a prize among winners.
-Lottery pay-outs demand high levels of interpersonal trust: players rely on unverifiable information about the total funds collected by the lottery, and they can never be sure whether the payments are evenly distributed.
-The fixed-multiplier formula alleviates such problems of adverse selection \citep{akerlof1970market, cohen2010testing, levin2001information}.
-As players and vendors known the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \emph{bicheiros} to a contract that can be easily enforced.
-This technique offers buyers a simple yet effective screening strategy that induces \emph{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.
-
-\emph{Bicheiros} have addressed information asymmetries in another ways.
-Since the 1950s, when the \emph{jogo do bicho} bankers had moved their operations to the \emph{fortalezas}, the public could not oversee the lottery draws \citep[259]{chazkel2011laws}.
-This could lead to a decline in trust among buyers and vendors of lottery tickets and, as a result, to reduced profits.
-\emph{Bicheiros} have mitigated this problem with a two-pronged strategy.
-First, they started to utilise the winning numbers from the licit government-run lottery, the \emph{Loteria Federal}, instead of their own draws \citetext{\citealp[546]{chazkel2007beyond}; \citealp[89]{labronici2012paratodos}; \citealp[39-40]{mello1989historia}}.
-The federal lottery numbers are public information.
-The media broadcasts the draws on radio and TV, so any interested player can verify the selected numbers.
-The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public; hence, \emph{bicheiros} can free ride on the lottery's long-standing reputation of credibility.\footnote{As of April 2016, the lottery was audited by the \emph{Controladoria Geral da União} (Comptroller General of Brazil), the \emph{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
-The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute.
-See \url{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm} (in Portuguese).
-Access: December 2016.}
-
-Second, they included representatives of all major \emph{jogo do bicho} bankers in every draw and independently publicise the game results.
-Certain \emph{bicheiros} went as far as publishing the numbers in Rio's newspapers.
-In the early twentieth century, some tabloids were entirely dedicated to the game \citep[60]{magalhaes2005ganhou}.
-Booking agents see this strategy as a credible signal from the game financiers, as providing contrasting information would indicate game manipulation.
-Moreover, collusion can also be spotted if the draws show repeated numbers or unusual patterns.
-
-These efforts have proved popular with the game enthusiasts.
-One often-repeated saying about the \emph{jogo do bicho} is that `in the \emph{jogo do bicho}, what is written down counts' \citep[159]{chazkel2011laws}, that is, buyers and sellers do fulfill their informal obligations without third-party enforcement.
-Such mutual confidence reduces the potential for conflict in the game.
-As the public does not see the \emph{jogo do bicho} as violent or harmful, the stigma of repugnance associated with gambling becomes less pervasive.
-By reducing the possibilities of cheating and putting long-term interests first, the \emph{jogo do bicho} bankers have avoided the fate of other repugnant markets and run their business relatively undisturbed for decades \citep[20]{da1999aguias}.
-
-\subsection{Internal Governance}
-\label{sub:internal}
-
-Individuals working at different levels of hierarchy often have non-aligned interests.
-As a result, it may occur that one party (the agent) behaves rationally in a manner that maximises his/her benefits, but that is contrary to the interests of his/her superior (the principal).
-This dilemma is pervasive in formal organisations \citep{holmstrom1979moral,jensen1976theory,moe1984new,shapiro2005agency,spence1971insurance}; in illegal markets perhaps it is even more so \citep{campana2013cooperation,gambetta2009codes,skarbek2011governance,skarbek2014social}.
-As monitoring costs in criminal businesses are higher than in formal ones, principals face considerable difficulties to induce cooperation from agents.
-Moreover, criminals often engage in opportunistic behaviour and `hidden actions', that is, they do not put the required levels of effort if they know they are not being monitored \citep[38--42]{arrow1985agency}.
-
-In the \emph{jogo do bicho} setting, one such problems concerns the trade-off between short- and long-term incentives for managers and bankers on the one side and bookmakers on the other.
-Managers have a permanent interest in the long-run profitability of the game, whereas street-corner booking agents tend to discount the future more heavily because their financial gains are small compared to that of their superiors.
-Additionally, bookmakers may denounce their employers to the police if they feel threatened.
-
-One way by which the \emph{jogo do bicho} principals solve the agency dilemma is by supplying club goods and selective incentives for low-tier members.
-Club goods are goods that can be simultaneously enjoyed by more than one individual but where exclusion mechanisms prevent consumption by non-members \citep{buchanan1965economic, cornes1996theory, olson1965logic, sandler1980economic, sandler1997club}.
-Basically, club goods are `public goods \emph{sans} non-excludability' \citep[928]{mcnutt1999public}.
-The first club good offered to \emph{bicheiros} by their bosses is private security.
-The game bankers have built an extensive network of gunmen and bribed police officers to protect their employees (and their profits) from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
-The \emph{jogo do bicho} network has a powerful deterrence effect and lethal force is rarely employed.
-However, threats are constant.
-`Zé' (Little Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \emph{jogo do bicho} informal security personnel:
-
-\begin{quote}
- [\dots] bums are scared and they don't mess around with us; they think there's a guard nearby or something like that.
-Look at all this money here!
-[shows the interviewer a handful of cash] It's not ours [referring to street-corner bookmakers].
-And if it's not ours, it's someone else's.
-When I worked in Penha (\emph{a low middle-class neighbourhood in the city of Rio de Janeiro -- translator's note}), the owner of a pub close to where I used to work always asked me to stay at the front door of his pub.
-People know that bums are afraid of \emph{bicheiros}.
-\end{quote}
-
-Apart from guaranteeing the physical integrity of the \emph{bicheiros}, bankers and managers also provide financial incentives for the bookmakers.
-\emph{Bicheiros} are allowed to receive tips from players, often have small expenses paid by managers, and may even request interest-free loans to cover unexpected costs such as illness-related expenses \citep{labronici2012paratodos}.
-
-However, the most important financial mechanism implemented by bankers to help \emph{bicheiros} is the \emph{descarga}, which is loosely translated as `the unloading'.
-The descarga is the \emph{jogo do bicho}'s main hedging technique and its purpose is to insure bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}; \citealp[16]{misse2007illegal}; \citealp[75]{soares1993jogo}}.
-Booking agents are sometimes unable to honour expensive bets.
-As mentioned above, the top prize in the animal game pays up to 4,000 times the amount invested, thus \emph{bicheiros} may have to raise thousands of Brazilian Reals in a single day.
-To prevent the \emph{quebra da banca} (`bust of the bank'), \emph{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
-The \emph{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \emph{jogo do bicho}.
-
-The descarga has played an important role in reducing individual risk; nevertheless, it has also changed the distribution of resources in the \emph{jogo do bicho}.
-Simple probability dictates that a booking agent rarely pays the highest prize in the \emph{jogo do bicho}; in contrast, the bankers receive a commission for \emph{every game} they hedge.
-Over time, there is a transfer of income from the bottom to the top of the animal game structure led by this constant inflow of fees.
-This accumulation of capital is probably one of the reasons why bankers were able to diversify their businesses and offer other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
-The descarga has made the game more resilient at the aggregated level, although it increased profits for the richest financiers at the expense of small bookmakers.
-
-\section{Tropical State Capture: \emph{Jogo do Bicho}, Samba and Politics}
-\label{sec:capture}
-
-The impact of the \emph{jogo do bicho} is not restricted to the Brazilian economy.
-Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural and social festivity, the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,chinelli1993vazio,queiroz1992carnaval}.
-The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that `without the \emph{jogo do bicho} the Carnival would have ended' \citep{odia2016aniz}.
-Owing to that support, \emph{bicheiros} have established an extensive patronage network with samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}.
-Although that network brings large material benefits to their members, the patronage system has created perverse incentives for government officials.
-
-The \emph{jogo do bicho}'s clientelism is more evident in the state of Rio de Janeiro than in other parts of the country.
-Historical factors explain why this is the case.
-Firstly, Rio de Janeiro city was the capital of Brazil for almost 200 years; despite losing the position to Brasília in 1960, it remains one of the country's main cultural and financial centres.
-Secondly, \emph{jogo do bicho} operators had historical ties with popular movements, which they eventually exploited to their advantage.
-Thirdly, the emergence of state-sponsored Carnival parades created a window of opportunity for \emph{bicheiros} to expand their influence over public authorities, either via bribing or by funding political campaigns.
-In this regard, Rio provided a suitable environment for self-interested politicians, community leaders and animal game financiers to collaborate.
-These illegal networks are crucial to understand why samba and Carnival became constituent features of Brazil's national identity, and how the festival has contributed to Rio's high levels of state corruption.
-
-\subsection{The `Medici of Samba': \emph{Bicheiros} as Patrons of Carnival}
-\label{sub:patrons}
-
-In 1930, opposition leader Getúlio Vargas led a bloodless coup d'état that brought Brazil's First Republic to an end.
-During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by dismantling effectively federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,desouza1983estado,fausto1972revoluccao,fausto2014concise,skidmore1967politics}.
-In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,nava1998lessons,williams2001culture}.
-Popular music, in turn, occupied an important place in Vargas's project of `brazilianing Brazil'.
-Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{avelar2011brazilian,mccann2004hello,stockler2011samba,vassberg1969villa,vassberg1975villa}.
-
-By the late 1930s, samba reached a unique position in Brazil's cultural identity.
-In a period when civil and political rights were limited \citep{de2001cidadania,duarte1993vicissitudes}, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}.
-Patriotic sambas exalted the country's natural beauties and the figure of the `friendly, happy, cordial and industrious' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
-The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of the male donkey and a female horse.
-However, in the 1930s the word loses its pejorative connotation in Brazil.
-Mainly due to the work of sociologist \citet{freyre1933casa}, the idea of a racial democracy becomes pervasive in the government discourse, and as a result the word gains a positive tone \citep[4]{reiter2009brazil}.} \citetext{\citealp[47]{dangelo2016samba}; \citealp[51]{vianna1995misterio}}.
-The institutionalisation of the Carnival parade in 1935, and the subsequent increases in public funding to the festival, cemented the relationship between politicians and samba groups \citep{almeida2017carnaval,cabral2016escolas,soihet1998subversao}.
-
-However, the samba groups were not passive members in this process.
-Since the 1960s, the Rio Carnival expanded in scope and, stimulated by growing numbers of spectators, the parades became more elaborate \citetext{\citealp{cabral2016escolas}; \citealp[214]{chinelli1993vazio}; \citealp[240]{hertzman2013making}}.
-Unable to cope with the rising costs of the show, the `samba schools', which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}.
-This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,farias2013carnival,misse2011crime,queiroz1992carnaval}.
-
-As I have mentioned in the previous section, the animal game at times faced opposition by the local population.
-The public often perceived the game as immoral and repugnant.
-Moreover, even after the bicho was well-established in Rio de Janeiro, the transition from a competitive betting market to an oligopoly involved the threat and often the use of physical violence against bookmakers who resisted the change \citetext{\citealp[143]{bezerra2009mecenato}, \citealp[52]{labronici2012paratodos}}.
-\emph{Bicheiros} were aware of the reputation costs their strategy entailed.
-They decided to finance samba schools hoping to win `the hearts and minds' of the population and attach a more positive image of the game among urban classes.
-Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but only as individuals who had a private interest in samba \citep[209]{chinelli1993vazio}.
-In 1984, a group of rich \emph{jogo do bicho} financiers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
-The LIESA marked a shift in the Carnival.
-For the first time, \emph{bicheiros} decided to act as a group rather than individuals.
-The organisation consolidated the power of \emph{bicheiros} over the parade and provided a formal mechanism to solve disputes among the samba school patrons \citetext{\citealp[43]{cavalcanti2006carnaval}; \citealp[171]{farias2013carnival}; \citealp[55]{labronici2012paratodos}}.
-
-The funding of the samba schools had an indirect effect to the animal game.
-The patronage also reduced agent-principal problems within the \emph{jogo do bicho}.
-\emph{Bicheiros} donate to samba school to gather support of the communities, and by doing so they gain access to local information on their business.
-Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring is cost-effective for animal game managers.
-Thus, street bookmakers have fewer incentives to cheat.
-In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}'s donations \citep{bbc2012aniz}.
-Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation.
-The patronage can be interpreted as an illegal version of `profit-sharing', a mechanism which has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.
-
-The samba schools have profited from this association too.
-First, they have gained autonomy from the government.
-The samba schools do not need to rely exclusively on public funds to organise the parade, and money from the \emph{jogo do bicho} permitted the schools to act independently \citep[209]{chinelli1993vazio}.
-Second, the support of the \emph{jogo do bicho} has increased the political and social clout of the samba schools.
-In a country where the state is not present throughout the territory and human right abuses are frequent \citep{ahnen2003,odonnell1993state,pinheiro2000,pinheiro2001}, \emph{jogo do bicho} bankers, and more recently drug traffickers, have provided private governance to poor areas of Rio de Janeiro by enforcing property rights, mediating disputes, and preventing police abuse in the favelas \citep{arias2006dynamics,goldstein2013laughter,leeds1996cocaine}.
-In return for funds and protection from the \emph{bicheiros}, samba schools have served as intermediaries between the underworld and the political system.
-Although the \emph{banqueiros} are interested in weak law enforcement against the animal game, politicians have resorted to samba schools to contact \emph{bicheiros} and use their financial and electoral influence in the shanty towns \citep[17]{misse2011crime}.
-The samba schools, therefore, have increased their bargaining power in the political sphere and extended their reach within Rio's poor communities \citep[215]{chinelli1993vazio}.
-
-\subsection{Political Support}
-\label{sub:support}
-
-If politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, their relationship with the animal game bankers have become more ambivalent in the last decades.
-The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citetext{\citealp{gaspari2002ditadura}; \citealp{jupiara2015poroes}; \citealp[39]{zaluar2007democratizaccao}}.
-Given the absence of democratic checks and balances, paramilitaries and police forces colluded to repress potential dissidents of the regime and, frequently, to extort civilians \citep{gorender1999combate,magalhaes1997logica,misse2009acumulaccao,skidmore1990politics}.
-\emph{Bicheiros} saw the corruption of some members of the military as an opportunity.
-Wealthy \emph{jogo do bicho} bankers hired rogue police officers not only to work as security guards but to threaten eventual competitors in their regions of influence.
-The agreement between \emph{bicheiros} and corrupt members of the military was the ultimate responsible for the transformation of the \emph{jogo do bicho} into a `coercive oligopoly' \citep{jupiara2015poroes}.
-The support of the armed forces meant that new groups would be prohibited from entering the market and that the illegal lottery could operate undisturbed by the government.
-
-The links between \emph{bicheiros} and the public authorities changed after Brazil became a democracy in 1985.
-In the military regime, government officials were mainly interested in bribes from the animal game.
-But in the democratic period, votes became a sought-after political resource.
-\emph{Bicheiros} are important in this sense as they have direct influence over a number of poor communities.
-Their patronage networks ensure that candidates supported by \emph{bicheiros} receive a substantial amount of votes from areas where campaigning is too difficult or too costly \citep[17]{misse2011crime}.
-
-The Brazilian political system is particularly conductive to clientelistic practices.
-Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}.
-In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp{mainwaring1992brazilian}; \citealp[483]{samuels2000ambition}}.
-These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence.
-Nevertheless, that independence means candidates rely mostly on themselves to raise funds and establish communication with potential voters.
-Hence, political campaigns in Brazil tend to be expensive and personality-centred.
-
-The support from the \emph{jogo do bicho} mitigates both problems.
-With respect to the financial costs of campaigns, illegal donations from \emph{bicheiros} help to cover advertising expenses while having the additional benefit of not appearing in the official records of the candidates \citep{congressoemfoco2007bicho,gazetadopovo2007bicho,globo2012bicheiro}.
-This suggests that \emph{jogo do bicho}-funded politicians can circumvent spending limits and have an electoral advantage over their competitors.
-As candidates do not know whether their competitors receive funding from the \emph{jogo do bicho} nor the amount each one was paid, their dominant position is to contact the \emph{bicheiros} and join their networks.
-The situation is a prisoner's dilemma in which candidates would be better off by running cheaper campaigns and not being dependent of \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching an optimal solution.
-
-The votes from poor communities are instrumental for aspiring politicians.
-Brazil has an enforced compulsory voting system; therefore, turnout rates tend to be higher than in other democracies.
-Consequently, votes have high marginal utility for politicians.
-As elections may be decided by a small difference, the \emph{bicheiros}' clientelistic ties guarantee a minimum number of votes that politicians can rely upon on election day.
-Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process per se.
-Individuals may be punished if the candidate does not receive the expected number of votes, and are often compelled to vote for politicians that have only loose connections with their communities.
-Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil’s lower classes.
-
-Finally, the \emph{jogo do bicho} patronage highlights a crucial social dilemma within the Brazilian public law.
-Even though federal judges have prosecuted \emph{bicheiros}, politicians and police forces have no incentives to enforce the punishment.
-Although Brazilian judges enjoy job stability, the latter groups constantly require local-level support from the \emph{bicheiros}.
-Politicians and police officers may have accurate information on \emph{jogo do bicho} operations and \emph{bicheiros}' whereabouts, but the federal government cannot rely upon their cooperation.
-That can be one of the reasons why even after many attempts to arrest \emph{bicheiros}, there has been little progress in that regard in Brazil's latest democratic period (1985--present).
-
-\section{Conclusion}
+\section{Concluding Remarks}
 \label{sec:conclusion3}
@@
 Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}.
-Yet, the \emph{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
-\emph{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers.
-Furthermore, by investing in the Carnival parade \emph{bicheiros} have been able to gather popular and government support.
-Poor communities have associated with the \emph{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \emph{jogo do bicho}'s networks can provide.
+Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
+\textit{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers.
+Furthermore, by investing in the Carnival parade \textit{bicheiros} have been able to gather popular and government support.
+Poor communities have associated with the \textit{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.
 
-Nevertheless, the \emph{jogo do bicho} has also created negative externalities.
+Nevertheless, the \textit{jogo do bicho} has also created negative externalities.
 Violence is used to punish defectors and to constrain competitors.
-The clientelistic relationship that \emph{bicheiros} have with local politicians have lead to sub-optimal outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
+The clientelistic relationship that \textit{bicheiros} have with local politicians have lead to sub-optimal outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
 These negative externalities have long-term effects and still impact the Brazilian public sphere.
 
-Although the \emph{jogo do bicho} has received an increasing attention from scholars, much of its inner workings remain poorly understood.
-First, the relationship between \emph{bicheiros} and drug dealers is a topic that deserves attention.
+Although the \textit{jogo do bicho} has received an increasing attention from scholars, much of its inner workings remain poorly understood.
+First, the relationship between \textit{bicheiros} and drug dealers is a topic that deserves attention.
 Brazil has become one of the world's largest consumers of illicit drugs and South America's principal drug trafficking transit route \citep{miraglia2015drugs,misse2011crime}.
-The question whether \emph{bicheiros} collaborated or opposed the emergent drug dealing business is still unclear.
-Second, the extent to which \emph{bicheiros} use other businesses, such as hotels or factories, to laundry money has been mentioned by members of the Brazilian judiciary \citep{globo2012bicheiro,globo2015cacaniquel}; however, there is no reliable estimate on its size.
-Lastly, more research is required to clarify how \emph{bicheiros} from different parts of Brazil coordinate their activities and prevent large-scale conflicts.
-Cases studies are usually focused on Rio de Janeiro's \emph{bicheiros}, but scholars would benefit from comparative analyses with a larger number of states.
-This is an important step to elucidate how \emph{bicheiros} continue to influence politics and the public across Brazil.
+The question whether \textit{bicheiros} collaborated or opposed the emergent drug dealing business is still unclear.
+Second, the extent to which \textit{bicheiros} use other businesses, such as hotels or factories, to laundry money has been mentioned by members of the Brazilian judiciary \citep{globo2012bicheiro,globo2015cacaniquel}; however, there is no reliable estimate on its size.
+Lastly, more research is required to clarify how \textit{bicheiros} from different parts of Brazil coordinate their activities and prevent large-scale conflicts.
+Cases studies are usually focused on Rio de Janeiro's \textit{bicheiros}, but scholars would benefit from comparative analyses with a larger number of states.
+This is an important step to elucidate how \textit{bicheiros} continue to influence politics and the public across Brazil.
 
 \newpage
-
 \bibliography{bibliography}

### 2018-10-13 paper-bicho/bicho.tex
small update

@@
 This article presents a rational choice account of Brazil's \textit{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
-Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and the game has been a major driver of political corruption in Brazil.
+Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and it has played a major role in Brazil's cultural and political spheres.
 First, I examine the mechanisms that fostered the \textit{jogo do bicho}'s notable growth outside of the boundaries of the law.
-Second, I explain how the lottery financiers combine costly signals and selective incentives to foster external cooperation from members of community.
-Lastly, I analyse the relationship between the lottery sponsors and Brazilian representatives, specially the ways by which the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term goals.
+Second, I explain how the lottery financiers combine costly signals and selective incentives to induce cooperation from members of community.
+Lastly, I analyse the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term goals.
 
@@
 
-Like any other business manager, \textit{bicheiros} have to run their firm with low costs to increase profits.
+Like any business manager, \textit{bicheiros} have to run their firm with low costs to increase profits.
 However, the fact that the \textit{jogo do bicho} is clandestine imposes additional difficulties for its operation.
@@
 
-Since the mid-twentieth century, the \textit{jogo do bicho} has been in a state of `semi-legality' in Brazil.
-For all purposes, the \textit{jogo do bicho}  remains illegal in the country, yet there is no Brazilian city which does not have its own group of local \textit{bicheiros}.
-In that regard, the \textit{jogo} does not operate exclusively at the margins of the Brazilian law, but it consciously exploits and subsidises large sectors of the formal economy according to the interests of wealthy game sponsors.
-For instance, the \textit{jogo do bicho} has benefited from the
+Since the mid-twentieth century, the \textit{jogo do bicho} has been both iniquitous and ubiquitous in Brazil.
+For all purposes, the \textit{jogo do bicho} remains illegal, yet there is almost no Brazilian city which does not have its own local \textit{bicheiros}.
+In that regard, the \textit{jogo} does not operate exclusively at the margins of the Brazilian law; given its size, it is clear that the \textit{jogo} consciously exploits and subsidises large sectors of the formal economy.
+One good example is the relationship between the \textit{jogo} and the informal workforce.
+The \textit{jogo do bicho} gives a boost to the Brazilian economy by providing jobs for unskilled workers who cannot easily join the labour market.
+By doing so, the \textit{jogo} indirectly prevents some of poorest members of the Brazilian society from demanding more inclusive government policies, although they are those who would benefit the most from public assistance.
+Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without having to resort to government funding, which can be spent with other, probably wealthier, sectors of the population.
+I describe the animal game organisation structure and their impact in the Brazilian economy in further detail below.
 
@@
 At the bottom level are the \textit{bicheiros}, those in charge of selling \textit{jogo do bicho} tickets \citep{chazkel2007beyond, da1999aguias}.
-\textit{Bicheiros} are the most visible part of the \textit{jogo do bicho} structure.
-The bookmakers usually build their vending stands inside the premises of a local shop, such as a small grocery store, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}.
-\textit{Bicheiros} usually work alone, but may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
+\textit{Bicheiros} are the most visible part of the \textit{jogo do bicho}, and the name loosely describes all those involved in the lottery organisation.
+Yet their meaning in the \textit{jogo do bicho} structure is more particular and refers to street-level ticker sellers.
+The \textit{bicheiros}  usually build their vending stands inside the premises of a local shop, such as a small grocery store, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}.
+The street bookmakers usually work alone, but may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
 
@@
 The period coincides with a time when the \textit{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}.
-Due to the growing size of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and reduce coordination costs.
+Due to the growing scope of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and reduce coordination costs.
 

### 2018-10-15 paper-bicho/bicho.tex
Update 2.3

@@
 Thus, the game bankers have built an extensive network of gunmen and bribed police officers to protect their employees from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
+`Zé' (Little Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \emph{jogo do bicho} informal security personnel:
+
+\begin{quote}
+ [\dots] bums are scared and they don't mess around with us; they think there's a guard nearby or something like that.
+Look at all this money here!
+[shows the interviewer a handful of cash] It's not ours [referring to street-corner bookmakers].
+And if it's not ours, it's someone else's.
+When I worked in Penha (\emph{a low middle-class neighbourhood in the city of Rio de Janeiro -- translator's note}), the owner of a pub close to where I used to work always asked me to stay at the front door of his pub.
+People know that bums are afraid of \emph{bicheiros}.
+\end{quote}
+
+But the \textit{banqueiros} did not use violence only against other criminals.
+They often employ violent methods against competitors and their own staff, too.
+\citet{jupiara2015poroes} argue that Ailton Guimarães Jorge, a former Army officer, tortured and murdered rival lottery bosses in the late 1970s.
+One of his former allies, Army Colonel Paulo Malhães, told the Rio de Janeiro State Truth Commission how Guimarães consolidated his power: `he went on a rampage' \citep{belem2015guimaraes}.
+$Castor d$e Andrade, Rio's most influential animal game \textit{banqueiro}, also employed similar methods to run his business.
+Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors.
+In a famous case, Andrade shot Euclides Ponar, an old \textit{banqueiro} known as `Grey-Headed Chinese',\footnote{In Portuguese, \textit{China Cabe\c{c}a Branca}.} after Ponar denounced a fraud in lottery draws in 1976.
+Andrade directly planned the assassination of at least other four competitors in the 1980s \citep{globo2017castor}.
 
@@
 \newpage
-\bibliography{bibliography}
+\bibliography{bibliography.bib}
 \bibliographystyle{apalike}

### 2018-10-15 paper-bicho/bicho.tex
Update bicho.tex

@@
 
-But the \textit{banqueiros} did not use violence only against other criminals.
+But the \textit{banqueiros} do not use violence only against other criminals.
 They often employ violent methods against competitors and their own staff, too.
 \citet{jupiara2015poroes} argue that Ailton Guimarães Jorge, a former Army officer, tortured and murdered rival lottery bosses in the late 1970s.
-One of his former allies, Army Colonel Paulo Malhães, told the Rio de Janeiro State Truth Commission how Guimarães consolidated his power: `he went on a rampage' \citep{belem2015guimaraes}.
-$Castor d$e Andrade, Rio's most influential animal game \textit{banqueiro}, also employed similar methods to run his business.
+One of his former allies, Army Colonel Paulo Malhães, told to the Rio de Janeiro State Truth Commission that Guimarães `went on a rampage' to consolidate his power \citep{belem2015guimaraes}.
+Castor de Andrade, Rio's most influential animal game \textit{banqueiro}, also employed similar methods to run his business.
 Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors.
-In a famous case, Andrade shot Euclides Ponar, an old \textit{banqueiro} known as `Grey-Headed Chinese',\footnote{In Portuguese, \textit{China Cabe\c{c}a Branca}.} after Ponar denounced a fraud in lottery draws in 1976.
+In a famous case, Andrade shot Euclides Ponar, an old \textit{jogo do bicho} boss known as `Grey-Headed Chinese',\footnote{In Portuguese, \textit{China Cabe\c{c}a Branca}.} after Ponar denounced a fraud in lottery draws in 1976.
 Andrade directly planned the assassination of at least other four competitors in the 1980s \citep{globo2017castor}.
+
+A more puzzling question, however, is how the animal game bosses elicit cooperation from \textit{external} members, such as bettors, community leaders, or public officers.
+It is puzzling because \textit{bicheiros} do not use violence to induce individuals to play the lottery, nor have they ever clashed with the Brazilian government.
+Precisely because violence would shun off profits, \textit{bicheiros} devised other mechanisms to create a friendly environment for the illegal lottery.
+In the next section, I investigate two of them, costly signalling and reputation building.
 

### 2018-10-16 paper-bicho/bicho.tex
update 2.4

@@
 This article presents a rational choice account of Brazil's \textit{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
-Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and it has played a major role in Brazil's cultural and political spheres.
+Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and it has played a major role in Brazil's cultural and political life.
+My analysis has three goals.
 First, I examine the mechanisms that fostered the \textit{jogo do bicho}'s notable growth outside of the boundaries of the law.
-Second, I explain how the lottery financiers combine costly signals and selective incentives to induce cooperation from members of community.
-Lastly, I analyse the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term goals.
+Second, I investigate how the lottery financiers combine costly signals and selective incentives to induce cooperation from members of community.
+Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term goals.
 
@@
 In particular,  \textit{bicheiros} face two main challenges to keep the lottery running.
-First, they need to gather public support so that bettors are not discouraged to engage in the lottery despite it being illegal.
+First, they need to gather public support so that gamblers are not discouraged to engage in the lottery despite it being illegal.
 Second, \textit{bicheiros} also have to ensure that the state repression is not prohibitively costly to their business, otherwise they would be better off by shutting it down.
@@
 Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising.
-Brazil's growing cities offered a number of new occupations for freed slaves, and many were eager to live far from their former masters \citep{andrews1991blacks, skidmore1993black}.
-The freed slaves were joined by increasing numbers of Asian and European immigrants who were promised access to farming lands in the country \citep{hall1969origins, lesser2013immigration}.
-However, the hopes of the African-Brazilians and the new settlers would soon be frustrated by a series of economic downturns.
+Brazil's growing cities offered new occupations for former slaves who desired to move away from their former masters \citep{andrews1991blacks, skidmore1993black}.
+Increasing numbers of Asian and European immigrants joined the freed slaves and moved to the cities soon after they arrived in Brazil \citep{hall1969origins, lesser2013immigration}.
+However, the hopes of the African-Brazilians and the new foreign settlers would soon be frustrated by a series of economic downturns.
 The Brazilian labour market suffered a severe contraction in the wake of the \textit{Encilhamento} financial crisis of 1891, and the economic instability aggravated the already difficult conditions of the working classes \citep{topik2014political, triner2005baring}.
@@
 In that regard, large swathes of the urban population turned to the informal economy.
-As \citet[115]{chazkel2011laws} observes, there were few occupations available to poor women and foreigners in the 1890s, and many of them became street vendors.
+As \citet[115]{chazkel2011laws} observes, there were few occupations available to lower-class women and foreigners in the 1890s, and a large number of poor workers became street vendors.
 The profession requires little technical skills and has low barriers of entry, but it can be very profitable if for whatever reason there is a strong demand for particular product.
@@
 
-A more puzzling question, however, is how the animal game bosses elicit cooperation from \textit{external} members, such as bettors, community leaders, or public officers.
+The threat of violence is balanced with financial benefits to low-rank members of the \textit{jogo do bicho} organisation.
+
+A more puzzling question, however, is how the animal game bosses elicit cooperation from \textit{external} members, such as gamblers, community leaders, or public officers.
 It is puzzling because \textit{bicheiros} do not use violence to induce individuals to play the lottery, nor have they ever clashed with the Brazilian government.
@@
 In the next section, I investigate two of them, costly signalling and reputation building.
+
+\section{External Cooperation}%
+\label{sec:external}
+
+Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and experimental studies \citep{isaac1984divergent, marwell1981economists} have both demonstrated that long-term cooperation is possible whenever players expect future pay-offs to be higher than present ones.
+Fear of retaliation induces individuals not to cheat.
+Nevertheless, illegal organisations tend to discount the future more heavily than the other groups, what makes cooperative behaviour uncommon in criminal groups \citep{skarbek2011governance,skarbek2012prison,skarbek2014social}.
+
+The \emph{jogo do bicho} is an exception to this rule.
+Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community, costly signalling and reputation strategies.
+In the following section I discuss how the \textit{jogo} colludes with corrupt politicians and civil servants for mutual benefit.
+
+\subsection{}%
+\label{sub:}
+
+The \emph{jogo do bicho} entrepreneurs have made considerable efforts to present themselves as honest brokers.
+The first trust-enhancing mechanism they have employed to foster external cooperation was the use of a \emph{fixed-multiplier formula} for pay-outs.
+It works as follows.
+If a player wins the lowest prize of the animal game, he or she receives 18 times his/her investment regardless of the size of the bet.
+Bigger prizes naturally offer higher returns; a lucky winner of the top prize wins up to 4,000 times the value of his/her bet \citetext{\citealp[89]{labronici2012paratodos}; \citealp[20]{magalhaes2005ganhou}}.
+
+This stands in sharp contrast to the common practice of sharing a prize among winners.
+Lottery pay-outs demand high levels of interpersonal trust: players rely on unverifiable information about the total funds collected by the lottery, and they can never be sure whether the payments are evenly distributed.
+The fixed-multiplier formula alleviates such problems of adverse selection \citep{akerlof1970market, cohen2010testing, levin2001information}.
+As players and vendors known the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \emph{bicheiros} to a contract that can be easily enforced.
+This technique offers buyers a simple yet effective screening strategy that induces \emph{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.
+
+\emph{Bicheiros} have addressed information asymmetries in another ways.
+Since the 1950s, when the \emph{jogo do bicho} bankers had moved their operations to the \emph{fortalezas}, the public could not oversee the lottery draws \citep[259]{chazkel2011laws}.
+This could lead to a decline in trust among buyers and vendors of lottery tickets and, as a result, to reduced profits.
+\emph{Bicheiros} have mitigated this problem with a two-pronged strategy.
+First, they started to utilise the winning numbers from the licit government-run lottery, the \emph{Loteria Federal}, instead of their own draws \citetext{\citealp[546]{chazkel2007beyond}; \citealp[89]{labronici2012paratodos}; \citealp[39-40]{mello1989historia}}.
+The federal lottery numbers are public information.
+The media broadcasts the draws on radio and TV, so any interested player can verify the selected numbers.
+The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public; hence, \emph{bicheiros} can free ride on the lottery's long-standing reputation of credibility.\footnote{As of April 2016, the lottery was audited by the \emph{Controladoria Geral da União} (Comptroller General of Brazil), the \emph{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
+The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute.
+See \url{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm} (in Portuguese).
+Access: December 2016.}
+
+Second, they included representatives of all major \emph{jogo do bicho} bankers in every draw and independently publicise the game results.
+Certain \emph{bicheiros} went as far as publishing the numbers in Rio's newspapers.
+In the early twentieth century, some tabloids were entirely dedicated to the game \citep[60]{magalhaes2005ganhou}.
+Booking agents see this strategy as a credible signal from the game financiers, as providing contrasting information would indicate game manipulation.
+Moreover, collusion can also be spotted if the draws show repeated numbers or unusual patterns.
+
+These efforts have proved popular with the game enthusiasts.
+One often-repeated saying about the \emph{jogo do bicho} is that `in the \emph{jogo do bicho}, what is written down counts' \citep[159]{chazkel2011laws}, that is, buyers and sellers do fulfill their informal obligations without third-party enforcement.
+Such mutual confidence reduces the potential for conflict in the game.
+As the public does not see the \emph{jogo do bicho} as violent or harmful, the stigma of repugnance associated with gambling becomes less pervasive.
+By reducing the possibilities of cheating and putting long-term interests first, the \emph{jogo do bicho} bankers have avoided the fate of other repugnant markets and run their business relatively undisturbed for decades \citep[20]{da1999aguias}.
 

### 2018-10-16 paper-bicho/bicho.tex
small edits sec 2

@@
 
-\subsection{Historical Background}
+\subsection{Historical Background: How the \textit{Bicheiros} Avoided Extinction}
 \label{sub:historical_background}
@@
 The text can be read at: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm}} (in Portuguese).
-Access: September 2017.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players.
+Access: September 2018.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players.
 \citep{fsp2016legalizarbicho}.
@@
 
-\subsection{Organisational Structure}
+\subsection{A Hierarchical Organisational Structure}
 \label{sub:organisation}
@@
 The threat of violence is balanced with financial benefits to low-rank members of the \textit{jogo do bicho} organisation.
+\emph{Bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.
 
-A more puzzling question, however, is how the animal game bosses elicit cooperation from \textit{external} members, such as gamblers, community leaders, or public officers.
+The most important financial mechanism implemented by bankers to help \emph{bicheiros} is the \emph{descarga}, loosely translated as `the unloading'.
+The descarga is the \emph{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}; }; \citealp[75]{soares1993jogo}}.
+Booking agents are sometimes unable to honour expensive bets.
+The top prize in the animal game pays up to 4,000 times the amount invested, thus \emph{bicheiros} may have to raise thousands of Brazilian Reals in a single day to pay the lucky winners.
+To prevent the \emph{quebra da banca} (`bust of the bank'), \emph{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
+The \emph{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \emph{jogo do bicho}.
+
+The descarga has significantly changed the distribution of resources in the \emph{jogo do bicho}, and the richest bankers benefited the most from it.
+Simple probability dictates that a booking agent rarely pays the highest lottery prize, and yet the bankers receive a commission for \emph{every game} they hedge.
+Over time, there is a transfer of income from the bottom to the top of the animal game structure due to the fees.
+This accumulation of capital is probably the reason why in the 1990s bankers started offering other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
+They simply had more capital to invest.
+In sum, while the descarga has made the game more resilient at the aggregated level, it increased profits for the richest financiers at the expense of small bookmakers.
+
+By combining both carrots and sticks, the \textit{jogo do bicho} bosses are able to effectively coordinate their employees.
+A more difficult question, however, is how the bankers elicit cooperation from \textit{external} members, such as gamblers, community leaders, or public officers.
 It is puzzling because \textit{bicheiros} do not use violence to induce individuals to play the lottery, nor have they ever clashed with the Brazilian government.
-Precisely because violence would shun off profits, \textit{bicheiros} devised other mechanisms to create a friendly environment for the illegal lottery.
+Precisely because violence could shun off profits, \textit{bicheiros} devised other mechanisms to create a friendly environment for the illegal lottery.
 In the next section, I investigate two of them, costly signalling and reputation building.
@@
 
-\subsection{}%
-\label{sub:}
+\subsection{Winning Hearts, Minds, and Pockets: Market Dynamics}%
+\label{sub:pockets}
 

### 2018-10-16 paper-bicho/bicho.tex
add text sec 2

@@
 
-The threat of violence is balanced with financial benefits to low-rank members of the \textit{jogo do bicho} organisation.
-\emph{Bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.
+Despite these serious events, killings are rare in the animal game.
+Since the lottery bosses can credibly indicate that violence is a low-cost option for them, the mere threat of punishment is enough to induce cooperation.
+This is a good strategy for the \textit{banqueiros}.
+Interestingly, the fact that they had committed violent crimes in the past reduces the need to commit them in the present, as and a result they can spend less money on security and increase profits.
+As it happens in many traditional markets, if a group is able to form a cartel, they can increase the price of their services without fearing immediate competition.
+The same logic is valid for the \textit{jogo do bicho}, although the means it employs are unconventional.
+
+The threat of violence is not the only means the \textit{bicheiros} have at hand.
+They balance it  with financial benefits to low-rank members of the organisation.
+For instance, street \emph{bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.
 
@@
 
-Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and experimental studies \citep{isaac1984divergent, marwell1981economists} have both demonstrated that long-term cooperation is possible whenever players expect future pay-offs to be higher than present ones.
-Fear of retaliation induces individuals not to cheat.
-Nevertheless, illegal organisations tend to discount the future more heavily than the other groups, what makes cooperative behaviour uncommon in criminal groups \citep{skarbek2011governance,skarbek2012prison,skarbek2014social}.
+Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible in a great number of situations.
+The main requirement for sustained cooperation is that players should believe that future pay-offs will be higher than present ones.
+If that condition is true, fear of retaliation will induce individuals not to cheat.
+
+In theory, the same should also apply to illegal organisations.
+Yet in practice we see that criminal groups tend to be short-term oriented, or, in more technical terms, they tend to discount the future more heavily that most people.
+This makes cooperative behaviour rather uncommon among criminal organisations, and there is considerable evidence suggesting that illegal groups have considerable difficulty to maintain themselves in the long run.
 
 The \emph{jogo do bicho} is an exception to this rule.
+The game has been running for more than a century with no considerable interruption and with moderately low levels of violence -- at least when compared to other illegal activities such as drug trafficking.
+Moreover,
+
  Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community, costly signalling and reputation strategies.

### 2018-10-17 paper-bicho/bicho.tex
moar edits

@@
 \usepackage[libertine]{newtxmath}
-\usepackage[scaled=.9]{inconsolata}
+\usepackage[scaled=.95]{inconsolata}
 \usepackage[usenames,dvipsnames]{xcolor}
@@
 
-Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,connelly2011signaling,spence1973job}.
+Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,spence1973job}.
 I provide evidence that \textit{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and reduce the uncertainty associated with clandestine markets.
@@
 Thus, the game bankers have built an extensive network of gunmen and bribed police officers to protect their employees from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
-`Zé' (Little Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \emph{jogo do bicho} informal security personnel:
+`Zé' (Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \emph{jogo do bicho} informal security personnel:
 
@@
 Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors.
-In a famous case, Andrade shot Euclides Ponar, an old \textit{jogo do bicho} boss known as `Grey-Headed Chinese',\footnote{In Portuguese, \textit{China Cabe\c{c}a Branca}.} after Ponar denounced a fraud in lottery draws in 1976.
-Andrade directly planned the assassination of at least other four competitors in the 1980s \citep{globo2017castor}.
+In a famous case, Andrade shot Euclides Ponar, an old \textit{jogo do bicho} boss known as `Grey-Headed Chinese' (\textit{China Cabe\c{c}a Branca}) after Ponar denounced a fraud in lottery draws in 1976.
+Andrade planned the assassination of at least four other competitors \citep{globo2017castor}.
 
@@
 
-The threat of violence is not the only means the \textit{bicheiros} have at hand.
+The threat of violence is not the only means the \textit{bicheiros} have at their disposal.
 They balance it with financial benefits to low-rank members of the organisation.
@@
 The most important financial mechanism implemented by bankers to help \emph{bicheiros} is the \emph{descarga}, loosely translated as `the unloading'.
-The descarga is the \emph{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}; }; \citealp[75]{soares1993jogo}}.
+The descarga is the \emph{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}}.
 Booking agents are sometimes unable to honour expensive bets.
@@
 Over time, there is a transfer of income from the bottom to the top of the animal game structure due to the fees.
-This accumulation of capital is probably the reason why in the 1990s bankers started offering other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
+This accumulation of capital is one of the reasons why in the 1990s bankers started offering other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
 They simply had more capital to invest.
-In sum, while the descarga has made the game more resilient at the aggregated level, it increased profits for the richest financiers at the expense of small bookmakers.
+In sum, while the \textit{descarga} has made the game more resilient at the aggregated level, it increased profits for the richest financiers at the expense of small bookmakers.
+
+This has brought broader consequences to the Brazilian illegal economy.
+As the number of slot machines increased in Brazil, the opportunity for criminals to use them for money laundering are higher as well.
+The scheme is easy to carry out.
+The owner of a slot machine issues a ticket with a winning prize, and the criminal declares the prize as his legitimate wealth.
+Then he can legally use that money for any purpose without raising suspicion from the authorities.
+The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called `the slot machine mafia', including relatives of Castor de Andrade \citep{estado2011cacaniquel}.
+In 2012, President Dilma Rousseff sanctioned a law that classified slot machines and the \textit{jogo do bicho} as money-laundering activities \citep{agenciabrasil2012dilma}.
 
@@
 
-Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible in a great number of situations.
+Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible even in difficult situations.
 The main requirement for sustained cooperation is that players should believe that future pay-offs will be higher than present ones.
@@
 In theory, the same should also apply to illegal organisations.
-Yet in practice we see that criminal groups tend to be short-term oriented, or, in more technical terms, they tend to discount the future more heavily that most people.
-This makes cooperative behaviour rather uncommon among criminal organisations, and there is considerable evidence suggesting that illegal groups have considerable difficulty to maintain themselves in the long run.
+Yet in practice we see that criminal groups tend to be short-term oriented, that is, they tend to discount the future more heavily that most people.
+This makes cooperative behaviour among criminal rather uncommon, and there is substantial evidence suggesting that illegal groups face problems to maintain themselves in the long run \citep[e.g.,][]{gambetta2009codes, leeson2010pirational, skarbek2011governance, skarbek2012prison, varese2001russian}.
 
 The \emph{jogo do bicho} is an exception to this rule.
-The game has been running for more than a century with no considerable interruption and with moderately low levels of violence -- at least when compared to other illegal activities such as drug trafficking.
-Moreover,
+The game has been running for more than a century with no considerable interruption.
+Importantly, the \textit{jogo} involves moderately low levels of violence -- at least when compared to other illegal activities such as drug trafficking.
+Moreover, Brazilians widely consider the \textit{jogo do bicho} an honest lottery, and reports of cheating are uncommon.
+The popular motto associated with the game testifies in its favour: `\textit{vale o escrito; ganhou, leva}', or `what is written is what counts; if you win it, you take it' \citep{magalhaes2005ganhou}.
+Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community, costly signalling and reputation strategies.
 
- Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community, costly signalling and reputation strategies.
-In the following section I discuss how the \textit{jogo} colludes with corrupt politicians and civil servants for mutual benefit.
-
-\subsection{Winning Hearts, Minds, and Pockets: Market Dynamics}%
+\subsection{Winning Hearts, Minds, and Pockets: Illegal Market Dynamics}%
 \label{sub:pockets}
@@
 The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute.
-See \url{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm} (in Portuguese).
+See \href{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm}{\texttt{https://bit.ly/2q3XnAV}} (in Portuguese).
 Access: December 2016.}
@@
 These efforts have proved popular with the game enthusiasts.
-One often-repeated saying about the \emph{jogo do bicho} is that `in the \emph{jogo do bicho}, what is written down counts' \citep[159]{chazkel2011laws}, that is, buyers and sellers do fulfill their informal obligations without third-party enforcement.
 Such mutual confidence reduces the potential for conflict in the game.

### 2018-10-17 paper-bicho/bicho.tex
update

@@
 \makeatother
-\hypersetup{pdftitle={Beasts of Prey or Rational Animals? Private Governance in Brazil's Jogo do bicho},
+\hypersetup{pdftitle={Beasts of Prey or Rational Animals? Private Governance in Brazil's Jogo do Bicho},
 	pdfauthor={Danilo Freire},
@@
 
-\section{Introduction}
+za\section{Introduction}
 \label{sec:intro}
@@
 Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, but they soon allowed participants to choose the animals they preferred.
-This small change made the game considerably more appealing to the public, who often resorted to clairvoyants and fortune tellers to guess the winning animals \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.
-The Baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
+This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.
+It is still common for gamblers to consult fortune tellers about the next results of the lottery.} The Baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
 Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in many stores across Rio de Janeiro.
@@
 The campaign extended for several decades and received considerable support from the \textit{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}.
-The Brazilian government officially banned the \textit{jogo do bicho} in 1946 and it remains illegal until this day.
+The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and it remains illegal until this day.
 
 Yet the game has survived.
-The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{jupiara2015poroes}.
-But more than a act of defiance, the \textit{jogo do bicho} is a successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
+The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under both military regimes and democratic governments \citep{jupiara2015poroes}.
+But more than a act of civic disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
 A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
-Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the Brazilian workforce.
+Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce.
 See \href{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}{\texttt{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}}.
@@
 More specifically, I use an array of bibliographical sources to show how the game operators, called \textit{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.
-The tactics are unusual and often repressive, such as bribing corrupt police officers to threaten undisciplined employees.
+This does not mean that their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals.
 Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive at first, they do address the long-term needs of the \textit{jogo do bicho} financiers.
@@
 Lastly, this work connects to the literature on state capture, which is one of the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}.
-More specifically, I use the Brazilian case to illustrate how politicians and civil servants can be co-opted by criminal groups and produce sub-optimal social outcomes.
+More specifically, I use the Brazilian case to illustrate how politicians and civil servants can be co-opted by criminal groups and produce bad social outcomes.
 \citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities.
@@
 
-The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
+The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}.
+They are: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
 I discuss each of these elements below.
@@
 The last necessary condition for the emergence of the \textit{jogo do bicho} is weak law enforcement.
-\citet[69--100]{chazkel2011laws} notes that until the 1940s police district chiefs operated within a large margin of discretion and repression against bookmakers was idiosyncratic.
+\citet[69--100]{chazkel2011laws} notes that until the 1940s, police district chiefs operated within a large margin of discretion, so official repression against bookmakers was notably inconsistent.
 In the early years of \textit{jogo do bicho}, lottery `bankers' were allowed to operate virtually free from police interference, what surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}.
-Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code and in 1941 the animal game was banned.\footnote{See: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del3688.htm}} (in Portuguese).
+Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code, and in 1941 the animal game was banned.\footnote{See \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del3688.htm}} (in Portuguese).
 Access: September 2017.} Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.
-The text can be read at: \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm}} (in Portuguese).
-Access: September 2018.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players.
-\citep{fsp2016legalizarbicho}.
+The text can be read at \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm}} (in Portuguese).
+Access: September 2018.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
 
-Since the mid-twentieth century, the \textit{jogo do bicho} has been both iniquitous and ubiquitous in Brazil.
-For all purposes, the \textit{jogo do bicho} remains illegal, yet there is almost no Brazilian city which does not have its own local \textit{bicheiros}.
-In that regard, the \textit{jogo} does not operate exclusively at the margins of the Brazilian law; given its size, it is clear that the \textit{jogo} consciously exploits and subsidises large sectors of the formal economy.
+Since the mid-twentieth century, the \textit{jogo do bicho} has been illegal but ubiquitous in Brazil.
+For all purposes, the \textit{jogo} remains criminalised, yet there is virtually no Brazilian city which does not have its local \textit{bicheiros}.
+In that regard, the \textit{jogo} does not operate exclusively at the margins of the Brazilian law; it is clear that the \textit{jogo} consciously exploits and subsidises large sectors of the formal economy, otherwise it would not have lasted for so long.
 One good example is the relationship between the \textit{jogo} and the informal workforce.
@@
 By doing so, the \textit{jogo} indirectly prevents some of poorest members of the Brazilian society from demanding more inclusive government policies, although they are those who would benefit the most from public assistance.
-Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without having to resort to government funding, which can be spent with other, probably wealthier, sectors of the population.
+Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without having to resort to government assistance, thus public funds be spent on other, probably wealthier, sectors of the population.
 I describe the animal game organisation structure and their impact in the Brazilian economy in further detail below.
@@
 They comprise the small financial elite of the game.
-A 2012 report by the Brazilian Federal Police affirmed that 10 \textit{banqueiros} controlled the market throughout the country; five of them based in the state of Rio de Janeiro \citep{globo2012contraventores}.
+A 2012 report by the Brazilian Federal Police affirmed that 10 \textit{banqueiros} controlled the market throughout the country, and five of them were based in the state of Rio de Janeiro \citep{globo2012contraventores}.
 Apart from funding the game, the bankers provide support for the employees to undertake their activities.
-The \textit{banqueiros}' main attributions include paying bribes to police personnel, bailing out sellers arrested by security forces, and offering judicial assistance to employees in case of legal persecution \citep[75]{labronici2012paratodos}.
+The \textit{banqueiros}' attributions include paying bribes to police personnel, bailing out sellers arrested by security forces, and offering judicial assistance to employees in case of legal persecution \citep[75]{labronici2012paratodos}.
 
-\textit{Banqueiros} run their businesses from fortified houses in unknown locations, the \textit{fortalezas} (`forts').
+The lottery bosses run their businesses from fortified houses in unknown locations, called the \textit{fortalezas} (`forts').
 The first \textit{fortalezas} likely appeared in the 1950s, when the animal game was already well-established across the Brazilian territory.
 The period coincides with a time when the \textit{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}.
-Due to the growing scope of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and reduce coordination costs.
+Due to the growing scope of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and make coordination easier.
 
@@
 \begin{quote}
- [\dots] bums are scared and they don't mess around with us; they think there's a guard nearby or something like that.
+ [\dots] bums are scared and they don't mess around with us; they think there's an officer nearby or something like that.
 Look at all this money here!
@@
 And if it's not ours, it's someone else's.
-When I worked in Penha (\emph{a low middle-class neighbourhood in the city of Rio de Janeiro -- translator's note}), the owner of a pub close to where I used to work always asked me to stay at the front door of his pub.
+When I worked in Penha\footnote{Penha is a low middle-class neighbourhood in the city of Rio de Janeiro.}, the owner of a pub close by always asked me to stay at the front door of his pub.
 People know that bums are afraid of \emph{bicheiros}.
@@
 One of his former allies, Army Colonel Paulo Malhães, told to the Rio de Janeiro State Truth Commission that Guimarães `went on a rampage' to consolidate his power \citep{belem2015guimaraes}.
-Castor de Andrade, Rio's most influential animal game \textit{banqueiro}, also employed similar methods to run his business.
+Castor de Andrade, Rio's most influential animal game banker, also employed similar methods to run his business.
 Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors.
 In a famous case, Andrade shot Euclides Ponar, an old \textit{jogo do bicho} boss known as `Grey-Headed Chinese' (\textit{China Cabe\c{c}a Branca}) after Ponar denounced a fraud in lottery draws in 1976.
-Andrade planned the assassination of at least four other competitors \citep{globo2017castor}.
+Andrade was likely involved in other assassination plots in the 1990s  \citep{globo2017castor}.
 
 Despite these serious events, killings are rare in the animal game.
-Since the lottery bosses can credibly indicate that violence is a low-cost option for them, the mere threat of punishment is enough to induce cooperation.
+Since the lottery bosses can credibly indicate that violence is a low-cost option for them, the mere threat of punishment is enough to discourage defectors.
 This is a good strategy for the \textit{banqueiros}.
-Interestingly, the fact that they had committed violent crimes in the past reduces the need to commit them in the present, as and a result they can spend less money on security and increase profits.
+The fact that they had committed violent crimes in the past reduces the need to commit them in the present, as and a result the bosses can spend less money on security and increase profits.
 As it happens in many traditional markets, if a group is able to form a cartel, they can increase the price of their services without fearing immediate competition.
-The same logic is valid for the \textit{jogo do bicho}, although the means it employs are unconventional.
+The same logic is valid for the \textit{jogo do bicho}, although through unconventional means.
 
-The threat of violence is not the only means the \textit{bicheiros} have at their disposal.
-They balance it with financial benefits to low-rank members of the organisation.
+The threat of violence is not the only tool the \textit{bicheiros} have at their disposal.
+They balance the use of violence with financial benefits to low-rank members of the organisation.
 For instance, street \emph{bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.
@@
 This accumulation of capital is one of the reasons why in the 1990s bankers started offering other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
-They simply had more capital to invest.
+They simply had more capital to invest, and this eventually helped to compensate the downturn in the \textit{jogo do bicho} markets in the last years \citep{globo2017castor}.
 In sum, while the \textit{descarga} has made the game more resilient at the aggregated level, it increased profits for the richest financiers at the expense of small bookmakers.

### 2018-10-18 paper-bicho/bicho.tex
macbook

@@
 
-za\section{Introduction}
+\section{Introduction}
 \label{sec:intro}
@@
 Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible even in difficult situations.
-The main requirement for sustained cooperation is that players should believe that future pay-offs will be higher than present ones.
+The main requirement for sustained cooperation is that players believe that future pay-offs will be higher than present ones.
 If that condition is true, fear of retaliation will induce individuals not to cheat.
@@
 In theory, the same should also apply to illegal organisations.
-Yet in practice we see that criminal groups tend to be short-term oriented, that is, they tend to discount the future more heavily that most people.
-This makes cooperative behaviour among criminal rather uncommon, and there is substantial evidence suggesting that illegal groups face problems to maintain themselves in the long run \citep[e.g.,][]{gambetta2009codes, leeson2010pirational, skarbek2011governance, skarbek2012prison, varese2001russian}.
+Yet in practice we see that criminal groups are generally short-term oriented, that is, they tend to discount the future more heavily that most people.
+This makes cooperative behaviour among criminal rather uncommon, and there is substantial evidence suggesting that illegal groups face serious collective action problems \citep[e.g.,][]{gambetta2009codes, leeson2010pirational, skarbek2011governance, skarbek2012prison, varese2001russian}.
 
-The \emph{jogo do bicho} is an exception to this rule.
-The game has been running for more than a century with no considerable interruption.
+The \textit{jogo do bicho} is an exception to this rule.
+The game has been running for more than a century without considerable interruption, what attests that \textit{bicheiros} have managed to solve collective action issues in one way or another.
 Importantly, the \textit{jogo} involves moderately low levels of violence -- at least when compared to other illegal activities such as drug trafficking.
-Moreover, Brazilians widely consider the \textit{jogo do bicho} an honest lottery, and reports of cheating are uncommon.
+Moreover, Brazilians widely consider the \textit{jogo do bicho} an honest lottery, and reports of cheating are surprisingly uncommon.
 The popular motto associated with the game testifies in its favour: `\textit{vale o escrito; ganhou, leva}', or `what is written is what counts; if you win it, you take it' \citep{magalhaes2005ganhou}.
@@
 \label{sub:pockets}
+
+Signalling theory predicts that when one person cannot obtain a
 

### 2018-10-23 paper-bicho/bicho.tex
add sections

@@
 \maketitle
-
 \begin{abstract}
@@
 Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and it has played a major role in Brazil's cultural and political life.
-My analysis has three goals.
-First, I examine the mechanisms that fostered the \textit{jogo do bicho}'s notable growth outside of the boundaries of the law.
-Second, I investigate how the lottery financiers combine costly signals and selective incentives to induce cooperation from members of community.
-Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term goals.
+My analysis of the \textit{jogo do bicho} has three goals.
+First, I examine the mechanisms that fostered the lottery's notable growth outside of the boundaries of Brazilian law.
+Second, I investigate how the animal game financiers combine costly signals and selective incentives to induce cooperation from members of community.
+Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term interests.
 
@@
 
-\section{Introduction}
+\section{Introduction}%
 \label{sec:intro}
@@
 
-The rules were simple.
+The rules were straightforward.
 In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
 Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, but they soon allowed participants to choose the animals they preferred.
-This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.
-It is still common for gamblers to consult fortune tellers about the next results of the lottery.} The Baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
+This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.} The lucky winner would take home a prize worth 20 times the ticket price, an amount higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.
+The Baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
 Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in many stores across Rio de Janeiro.
@@
 This does not mean that their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals.
-Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive at first, they do address the long-term needs of the \textit{jogo do bicho} financiers.
+Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive, they do address the long-term needs of the \textit{jogo do bicho} financiers.
 
@@
 
-\section{An Overview of the \textit{Jogo do Bicho}}
+\section{An Overview of the \textit{Jogo do Bicho}}%
 \label{sec:overview}
 
-\subsection{Historical Background: How the \textit{Bicheiros} Avoided Extinction}
+\subsection{Historical Background: How the \textit{Bicheiros} Avoided Extinction}%
 \label{sub:historical_background}
@@
 For all purposes, the \textit{jogo} remains criminalised, yet there is virtually no Brazilian city which does not have its local \textit{bicheiros}.
-In that regard, the \textit{jogo} does not operate exclusively at the margins of the Brazilian law; it is clear that the \textit{jogo} consciously exploits and subsidises large sectors of the formal economy, otherwise it would not have lasted for so long.
-One good example is the relationship between the \textit{jogo} and the informal workforce.
-The \textit{jogo do bicho} gives a boost to the Brazilian economy by providing jobs for unskilled workers who cannot easily join the labour market.
-By doing so, the \textit{jogo} indirectly prevents some of poorest members of the Brazilian society from demanding more inclusive government policies, although they are those who would benefit the most from public assistance.
-Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without having to resort to government assistance, thus public funds be spent on other, probably wealthier, sectors of the population.
+The \textit{jogo} does not operate exclusively at the margins of the Brazilian law; it is clear that it consciously exploits and subsidises large sectors of the formal economy.
 I describe the animal game organisation structure and their impact in the Brazilian economy in further detail below.
 
-\subsection{A Hierarchical Organisational Structure}
+\subsection{A Hierarchical Organisational Structure}%
 \label{sub:organisation}
@@
 Thus, the game bankers have built an extensive network of gunmen and bribed police officers to protect their employees from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
-`Zé' (Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \emph{jogo do bicho} informal security personnel:
+`Zé' (Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \textit{jogo do bicho} informal security personnel:
 
@@
 When I worked in Penha\footnote{Penha is a low middle-class neighbourhood in the city of Rio de Janeiro.}, the owner of a pub close by always asked me to stay at the front door of his pub.
-People know that bums are afraid of \emph{bicheiros}.
+People know that bums are afraid of \textit{bicheiros}.
 \end{quote}
@@
 They balance the use of violence with financial benefits to low-rank members of the organisation.
-For instance, street \emph{bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.
+For instance, street \textit{bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.
 
-The most important financial mechanism implemented by bankers to help \emph{bicheiros} is the \emph{descarga}, loosely translated as `the unloading'.
-The descarga is the \emph{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}}.
+The most important financial mechanism implemented by bankers to help \textit{bicheiros} is the \textit{descarga}, loosely translated as `the unloading'.
+The descarga is the \textit{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}}.
 Booking agents are sometimes unable to honour expensive bets.
-The top prize in the animal game pays up to 4,000 times the amount invested, thus \emph{bicheiros} may have to raise thousands of Brazilian Reals in a single day to pay the lucky winners.
-To prevent the \emph{quebra da banca} (`bust of the bank'), \emph{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
-The \emph{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \emph{jogo do bicho}.
+The top prize in the animal game pays up to 4,000 times the amount invested, thus \textit{bicheiros} may have to raise thousands of Brazilian Reals in a single day to pay the lucky winners.
+To prevent the \textit{quebra da banca} (`bust of the bank'), \textit{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
+The \textit{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \textit{jogo do bicho}.
 
-The descarga has significantly changed the distribution of resources in the \emph{jogo do bicho}, and the richest bankers benefited the most from it.
-Simple probability dictates that a booking agent rarely pays the highest lottery prize, and yet the bankers receive a commission for \emph{every game} they hedge.
+The \textit{descarga} has significantly changed the distribution of resources in the \textit{jogo do bicho}, and the richest bankers benefited the most from it.
+Simple probability dictates that a booking agent rarely pays the highest lottery prize, and yet the bankers receive a commission for \textit{every game} they hedge.
 Over time, there is a transfer of income from the bottom to the top of the animal game structure due to the fees.
@@
 
-This has brought broader consequences to the Brazilian illegal economy.
+This has brought broad consequences to the Brazilian illegal economy.
 As the number of slot machines increased in Brazil, the opportunity for criminals to use them for money laundering are higher as well.
@@
 The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called `the slot machine mafia', including relatives of Castor de Andrade \citep{estado2011cacaniquel}.
-In 2012, President Dilma Rousseff sanctioned a law that classified slot machines and the \textit{jogo do bicho} as money-laundering activities \citep{agenciabrasil2012dilma}.
+In 2012, President Dilma Rousseff sanctioned a law that considered slot machines and the \textit{jogo do bicho} as money laundering \citep{agenciabrasil2012dilma}.
 
-By combining both carrots and sticks, the \textit{jogo do bicho} bosses are able to effectively coordinate their employees.
-A more difficult question, however, is how the bankers elicit cooperation from \textit{external} members, such as gamblers, community leaders, or public officers.
+Money from the \textit{jogo} also has other effects on the economy.
+Some of these effects are indirect.
+For instance, the \textit{jogo do bicho} gives a boost to the Brazilian economy by providing jobs for unskilled workers who cannot easily join the labour market.
+By doing so, the \textit{jogo} prevents some of poorest members of the Brazilian society from demanding more inclusive government policies, although they are the ones who would benefit the most from public assistance.
+Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population.
+
+Another indirect economic effect of the \textit{jogo} is the rise in inequality.
+While formal workers can -- at least in theory -- demand higher compensations during a market upswing, the same is not true for the animal game employees.
+The threat of violence reduces the space for collective bargaining with the lottery bosses, and increasing profits at the top of the \textit{jogo do bicho} structure do not trickle down to higher salaries at the bottom.
+This is one of the reasons why the game bankers turned the game into an oligopoly: the use of violence guarantees not only that new entrants will not be allowed to join the market, but also that profits are concentrated in the hands of very few individuals.
+
+The \textit{jogo do bicho} market formation closely resembles what \citet{fligstein1996markets} calls `markets as politics', in which firms create institutions to restrain the competition and organise the labour force.
+At the formation of the lottery market, competition is fierce and incumbents an challengers are akin to social movements.
+They are constantly trying to convince others of the viability of their ideas.
+When markets stabilise, however, incumbents collude to impose their conditions of control to other players and to workers.
+In the case of the \textit{jogo}, this implies in a mix of financial incentives and violence threats carried by corrupt state agents or private bodyguards.
+
+A more difficult question is how the bankers elicit cooperation from \textit{external} members, such as gamblers, community leaders, or public officers.
 It is puzzling because \textit{bicheiros} do not use violence to induce individuals to play the lottery, nor have they ever clashed with the Brazilian government.
@@
 
-\section{External Cooperation}%
-\label{sec:external}
+\section{Winning Hearts, Minds, and Pockets: Illegal Market Dynamics}%
+\label{sub:pockets}
 
-Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible even in difficult situations.
+Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical case studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible even in difficult situations.
 The main requirement for sustained cooperation is that players believe that future pay-offs will be higher than present ones.
@@
 
-\subsection{Winning Hearts, Minds, and Pockets: Illegal Market Dynamics}%
-\label{sub:pockets}
+Signalling theory predicts that when someone cannot easily observe a characteristic she is interested in, she searches for signals that credibly conveys pieces of that required information.
+In the case of an illegal lottery, the main signal a gambler is looking for is \textit{honesty}, that is, that she has a fair chance of winning the game if she makes the right bet.
 
-Signalling theory predicts that when one person cannot obtain a
+Legal lotteries employ many techniques to show that this is the case.
+For instance, the Brazilian official lottery, run by the federal government, are regularly audited by the \textit{Controladoria Geral da União} (Comptroller General of Brazil), the \textit{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
+The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute.\footnote{See \href{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm}{\texttt{https://bit.ly/2q3XnAV}} (in Portuguese).
+Access: December 2016.} A clandestine lottery, in contrast, cannot provide the same signals.
+Thus, the game providers need to assure gamblers that their business is honest although it is illegal.
+\textit{Bicheiros} addressed this issue using a traditional commercial practice: they invested on their reputations.
 
-The \emph{jogo do bicho} entrepreneurs have made considerable efforts to present themselves as honest brokers.
-The first trust-enhancing mechanism they have employed to foster external cooperation was the use of a \emph{fixed-multiplier formula} for pay-outs.
+The \textit{jogo do bicho} entrepreneurs have made considerable efforts to present themselves as honest brokers.
+The first trust-enhancing mechanism they have employed to foster external cooperation was the use of a \textit{fixed-multiplier formula} for pay-outs.
 It works as follows.
@@
 The fixed-multiplier formula alleviates such problems of adverse selection \citep{akerlof1970market, cohen2010testing, levin2001information}.
-As players and vendors known the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \emph{bicheiros} to a contract that can be easily enforced.
-This technique offers buyers a simple yet effective screening strategy that induces \emph{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.
+As players and vendors known the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \textit{bicheiros} to a contract that can be easily enforced.
+This technique offers buyers a simple yet effective screening strategy that induces \textit{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.
 
-\emph{Bicheiros} have addressed information asymmetries in another ways.
-Since the 1950s, when the \emph{jogo do bicho} bankers had moved their operations to the \emph{fortalezas}, the public could not oversee the lottery draws \citep[259]{chazkel2011laws}.
+\textit{Bicheiros} have addressed information asymmetries in another ways.
+Since the 1950s, when the \textit{jogo do bicho} bankers had moved their operations to the \textit{fortalezas}, the public could not oversee the lottery draws \citep[259]{chazkel2011laws}.
 This could lead to a decline in trust among buyers and vendors of lottery tickets and, as a result, to reduced profits.
-\emph{Bicheiros} have mitigated this problem with a two-pronged strategy.
-First, they started to utilise the winning numbers from the licit government-run lottery, the \emph{Loteria Federal}, instead of their own draws \citetext{\citealp[546]{chazkel2007beyond}; \citealp[89]{labronici2012paratodos}; \citealp[39-40]{mello1989historia}}.
+\textit{Bicheiros} have mitigated this problem with a two-pronged strategy.
+First, they started to utilise the winning numbers from the licit government-run lottery, the \textit{Loteria Federal}, instead of their own draws \citetext{\citealp[546]{chazkel2007beyond}; \citealp[89]{labronici2012paratodos}; \citealp[39-40]{mello1989historia}}.
 The federal lottery numbers are public information.
 The media broadcasts the draws on radio and TV, so any interested player can verify the selected numbers.
-The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public; hence, \emph{bicheiros} can free ride on the lottery's long-standing reputation of credibility.\footnote{As of April 2016, the lottery was audited by the \emph{Controladoria Geral da União} (Comptroller General of Brazil), the \emph{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
-The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute.
-See \href{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm}{\texttt{https://bit.ly/2q3XnAV}} (in Portuguese).
-Access: December 2016.}
+The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public; hence, \textit{bicheiros} can free ride on the lottery's long-standing reputation of credibility.
 
-Second, they included representatives of all major \emph{jogo do bicho} bankers in every draw and independently publicise the game results.
-Certain \emph{bicheiros} went as far as publishing the numbers in Rio's newspapers.
+Second, they included representatives of all major \textit{jogo do bicho} bankers in every draw and independently publicise the game results.
+Certain \textit{bicheiros} went as far as publishing the numbers in Rio's newspapers.
 In the early twentieth century, some tabloids were entirely dedicated to the game \citep[60]{magalhaes2005ganhou}.
@@
 Such mutual confidence reduces the potential for conflict in the game.
-As the public does not see the \emph{jogo do bicho} as violent or harmful, the stigma of repugnance associated with gambling becomes less pervasive.
-By reducing the possibilities of cheating and putting long-term interests first, the \emph{jogo do bicho} bankers have avoided the fate of other repugnant markets and run their business relatively undisturbed for decades \citep[20]{da1999aguias}.
+As the public does not see the \textit{jogo do bicho} as violent or harmful, the stigma of repugnance associated with gambling becomes less pervasive.
+By reducing the possibilities of cheating and putting long-term interests first, the \textit{jogo do bicho} bankers have avoided the fate of other repugnant markets \citep[20]{da1999aguias}.
+
+This ability to elicit cooperation from external actors caused the \textit{jogo do bicho} to last longer than collection action theories would anticipate.
+In that regard, the bankers' investment in costly signalling and reputation has largely paid off: the lottery's continuous profits are a proof of their success.
+With that continuous flow of income, \textit{banqueiros} were able to extend their influence well beyond the poor communities in which they operate.
+One of these areas is politics.
+
+\section{Tropical State Capture: \emph{Jogo do Bicho}, Samba and Politics}
+\label{sec:capture}
+
+The impact of the \emph{jogo do bicho} is not restricted to the Brazilian economy.
+Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural and social festivity, the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,chinelli1993vazio,queiroz1992carnaval}.
+The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that `without the \emph{jogo do bicho} the Carnival would have ended' \citep{odia2016aniz}.
+Owing to that support, \emph{bicheiros} have established an extensive patronage network with samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}.
+Although that network brings large material benefits to their members, the patronage system has created perverse incentives for government officials.
+
+The \emph{jogo do bicho}'s clientelism is more evident in the state of Rio de Janeiro than in other parts of the country.
+Historical factors explain why this is the case.
+Firstly, Rio de Janeiro city was the capital of Brazil for almost 200 years; despite losing the position to Brasília in 1960, it remains one of the country's main cultural and financial centres.
+Secondly, \emph{jogo do bicho} operators had historical ties with popular movements, which they eventually exploited to their advantage.
+Thirdly, the emergence of state-sponsored Carnival parades created a window of opportunity for \emph{bicheiros} to expand their influence over public authorities, either via bribing or by funding political campaigns.
+In this regard, Rio provided a suitable environment for self-interested politicians, community leaders and animal game financiers to collaborate.
+These illegal networks are crucial to understand why samba and Carnival became constituent features of Brazil's national identity, and how the festival has contributed to Rio's high levels of state corruption.
+
+\subsection{The `Medici of Samba': \emph{Bicheiros} as Patrons of Carnival}
+\label{sub:patrons}
+
+In 1930, opposition leader Getúlio Vargas led a bloodless coup d'état that brought Brazil's First Republic to an end.
+During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by dismantling effectively federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,desouza1983estado,fausto1972revoluccao,fausto2014concise,skidmore1967politics}.
+In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,nava1998lessons,williams2001culture}.
+Popular music, in turn, occupied an important place in Vargas's project of `brazilianing Brazil'.
+Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{avelar2011brazilian,mccann2004hello,stockler2011samba,vassberg1969villa,vassberg1975villa}.
+
+By the late 1930s, samba reached a unique position in Brazil's cultural identity.
+In a period when civil and political rights were limited \citep{de2001cidadania,duarte1993vicissitudes}, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}.
+Patriotic sambas exalted the country's natural beauties and the figure of the `friendly, happy, cordial and industrious' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
+The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of the male donkey and a female horse.
+However, in the 1930s the word loses its pejorative connotation in Brazil.
+Mainly due to the work of sociologist \citet{freyre1933casa}, the idea of a racial democracy becomes pervasive in the government discourse, and as a result the word gains a positive tone \citep[4]{reiter2009brazil}.} \citetext{\citealp[47]{dangelo2016samba}; \citealp[51]{vianna1995misterio}}.
+The institutionalisation of the Carnival parade in 1935, and the subsequent increases in public funding to the festival, cemented the relationship between politicians and samba groups \citep{almeida2017carnaval,cabral2016escolas,soihet1998subversao}.
+
+However, the samba groups were not passive members in this process.
+Since the 1960s, the Rio Carnival expanded in scope and, stimulated by growing numbers of spectators, the parades became more elaborate \citetext{\citealp{cabral2016escolas}; \citealp[214]{chinelli1993vazio}; \citealp[240]{hertzman2013making}}.
+Unable to cope with the rising costs of the show, the `samba schools', which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}.
+This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,farias2013carnival,misse2011crime,queiroz1992carnaval}.
+
+As I have mentioned in the previous section, the animal game at times faced opposition by the local population.
+The public often perceived the game as immoral and repugnant.
+Moreover, even after the bicho was well-established in Rio de Janeiro, the transition from a competitive betting market to an oligopoly involved the threat and often the use of physical violence against bookmakers who resisted the change \citetext{\citealp[143]{bezerra2009mecenato}, \citealp[52]{labronici2012paratodos}}.
+\emph{Bicheiros} were aware of the reputation costs their strategy entailed.
+They decided to finance samba schools hoping to win `the hearts and minds' of the population and attach a more positive image of the game among urban classes.
+Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but only as individuals who had a private interest in samba \citep[209]{chinelli1993vazio}.
+In 1984, a group of rich \emph{jogo do bicho} financiers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
+The LIESA marked a shift in the Carnival.
+For the first time, \emph{bicheiros} decided to act as a group rather than individuals.
+The organisation consolidated the power of \emph{bicheiros} over the parade and provided a formal mechanism to solve disputes among the samba school patrons \citetext{\citealp[43]{cavalcanti2006carnaval}; \citealp[171]{farias2013carnival}; \citealp[55]{labronici2012paratodos}}.
+
+The funding of the samba schools had an indirect effect to the animal game.
+The patronage also reduced agent-principal problems within the \emph{jogo do bicho}.
+\emph{Bicheiros} donate to samba school to gather support of the communities, and by doing so they gain access to local information on their business.
+Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring is cost-effective for animal game managers.
+Thus, street bookmakers have fewer incentives to cheat.
+In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}'s donations \citep{bbc2012aniz}.
+Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation.
+The patronage can be interpreted as an illegal version of `profit-sharing', a mechanism which has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.
+
+The samba schools have profited from this association too.
+First, they have gained autonomy from the government.
+The samba schools do not need to rely exclusively on public funds to organise the parade, and money from the \emph{jogo do bicho} permitted the schools to act independently \citep[209]{chinelli1993vazio}.
+Second, the support of the \emph{jogo do bicho} has increased the political and social clout of the samba schools.
+In a country where the state is not present throughout the territory and human right abuses are frequent \citep{ahnen2003,odonnell1993state,pinheiro2000,pinheiro2001}, \emph{jogo do bicho} bankers, and more recently drug traffickers, have provided private governance to poor areas of Rio de Janeiro by enforcing property rights, mediating disputes, and preventing police abuse in the favelas \citep{arias2006dynamics,goldstein2013laughter,leeds1996cocaine}.
+In return for funds and protection from the \emph{bicheiros}, samba schools have served as intermediaries between the underworld and the political system.
+Although the \emph{banqueiros} are interested in weak law enforcement against the animal game, politicians have resorted to samba schools to contact \emph{bicheiros} and use their financial and electoral influence in the shanty towns \citep[17]{misse2011crime}.
+The samba schools, therefore, have increased their bargaining power in the political sphere and extended their reach within Rio's poor communities \citep[215]{chinelli1993vazio}.
+
+\subsection{Political Support}
+\label{sub:support}
+
+If politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, their relationship with the animal game bankers have become more ambivalent in the last decades.
+The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citetext{\citealp{gaspari2002ditadura}; \citealp{jupiara2015poroes}; \citealp[39]{zaluar2007democratizaccao}}.
+Given the absence of democratic checks and balances, paramilitaries and police forces colluded to repress potential dissidents of the regime and, frequently, to extort civilians \citep{gorender1999combate,magalhaes1997logica,misse2009acumulaccao,skidmore1990politics}.
+\emph{Bicheiros} saw the corruption of some members of the military as an opportunity.
+Wealthy \emph{jogo do bicho} bankers hired rogue police officers not only to work as security guards but to threaten eventual competitors in their regions of influence.
+The agreement between \emph{bicheiros} and corrupt members of the military was the ultimate responsible for the transformation of the \emph{jogo do bicho} into a `coercive oligopoly' \citep{jupiara2015poroes}.
+The support of the armed forces meant that new groups would be prohibited from entering the market and that the illegal lottery could operate undisturbed by the government.
+
+The links between \emph{bicheiros} and the public authorities changed after Brazil became a democracy in 1985.
+In the military regime, government officials were mainly interested in bribes from the animal game.
+But in the democratic period, votes became a sought-after political resource.
+\emph{Bicheiros} are important in this sense as they have direct influence over a number of poor communities.
+Their patronage networks ensure that candidates supported by \emph{bicheiros} receive a substantial amount of votes from areas where campaigning is too difficult or too costly \citep[17]{misse2011crime}.
+
+The Brazilian political system is particularly conductive to clientelistic practices.
+Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}.
+In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp{mainwaring1992brazilian}; \citealp[483]{samuels2000ambition}}.
+These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence.
+Nevertheless, that independence means candidates rely mostly on themselves to raise funds and establish communication with potential voters.
+Hence, political campaigns in Brazil tend to be expensive and personality-centred.
+
+The support from the \emph{jogo do bicho} mitigates both problems.
+With respect to the financial costs of campaigns, illegal donations from \emph{bicheiros} help to cover advertising expenses while having the additional benefit of not appearing in the official records of the candidates \citep{congressoemfoco2007bicho,gazetadopovo2007bicho,globo2012bicheiro}.
+This suggests that \emph{jogo do bicho}-funded politicians can circumvent spending limits and have an electoral advantage over their competitors.
+As candidates do not know whether their competitors receive funding from the \emph{jogo do bicho} nor the amount each one was paid, their dominant position is to contact the \emph{bicheiros} and join their networks.
+The situation is a prisoner's dilemma in which candidates would be better off by running cheaper campaigns and not being dependent of \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching an optimal solution.
+
+% \begin{figure}[!htbp]
+%  \centering
+%  \includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
+%  \caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right).
+The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'.
+Source: \citet{extra2012aniz}.}
+%  \label{fig:aniz}
+% \end{figure}
+
+The votes from poor communities are instrumental for aspiring politicians.
+Brazil has an enforced compulsory voting system; therefore, turnout rates tend to be higher than in other democracies.
+Consequently, votes have high marginal utility for politicians.
+As elections may be decided by a small difference, the \emph{bicheiros}' clientelistic ties guarantee a minimum number of votes that politicians can rely upon on election day.
+Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process per se.
+Individuals may be punished if the candidate does not receive the expected number of votes, and are often compelled to vote for politicians that have only loose connections with their communities.
+Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil’s lower classes.
+
+Finally, the \emph{jogo do bicho} patronage highlights a crucial social dilemma within the Brazilian public law.
+Even though federal judges have prosecuted \emph{bicheiros}, politicians and police forces have no incentives to enforce the punishment.
+Although Brazilian judges enjoy job stability, the latter groups constantly require local-level support from the \emph{bicheiros}.
+Politicians and police officers may have accurate information on \emph{jogo do bicho} operations and \emph{bicheiros}' whereabouts, but the federal government cannot rely upon their cooperation.
+That can be one of the reasons why even after many attempts to arrest \emph{bicheiros}, there has been little progress in that regard in Brazil's latest democratic period (1985--present).
 
@@
 
+Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}.
+Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
+\textit{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers.
+Furthermore, by investing in the Carnival parade \textit{bicheiros} have been able to gather popular and government support.
+Poor communities have associated with the \textit{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.
 Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}.

### 2018-10-26 paper-bicho/bicho.tex
final text?

@@
 \usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
-\usepackage[backref,pagebackref]{hyperref}
+\usepackage[backref,pagebackref,breaklinks=true]{hyperref}
 \usepackage{graphicx}
@@
 \author{Danilo Freire\thanks{Postdoctoral Research Associate, Political Theory Project, Brown University, 8 Fones Alley, Providence, RI 02912, \href{mailto:danilofreire@brown.edu}{\texttt{danilofreire@brown.edu}}, \href{http://danilofreire.com}{\texttt{http://danilofreire.com}}.
-I would like to thank Toke Aidt, Paulo Roberto Araujo, Jose Bolaños, Diogo Costa, Guilherme Duarte, Robert McDonnell, David Skarbek, Graham Denyer Willis, the participants of academic seminars at King's College London, the University of Cambridge and the Prometheus Institut in Berlin, the editors of Latin American Perspectives, and four reviewers for their valuable comments.
+I would like to thank Toke Aidt, Paulo Roberto Araujo, Jose Bolaños, Diogo Costa, Robert McDonnell, David Skarbek, Graham Denyer Willis, and the participants of academic seminars at King's College London, the University of Cambridge, and the Prometheus Institut in Berlin for their comments on earlier drafts of this paper.
+I also thank the editors of Latin American Perspectives and four reviewers for their valuable suggestions.
 I gratefully acknowledge the financial support of the National Council for Scientific and Technological Development.}
@@
 This article presents a rational choice account of Brazil's \textit{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
-Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and it has played a major role in Brazil's cultural and political life.
+Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and the game has played a major role in Brazil's cultural and political life.
 My analysis of the \textit{jogo do bicho} has three goals.
@@
 Like any business manager, \textit{bicheiros} have to run their firm with low costs to increase profits.
-However, the fact that the \textit{jogo do bicho} is clandestine imposes additional difficulties for its operation.
+However, the fact that the \textit{jogo do bicho} is clandestine imposes additional difficulties to its managers.
 In particular, \textit{bicheiros} face two main challenges to keep the lottery running.
@@
 
-In that regard, large swathes of the urban population turned to the informal economy.
+As a result, large swathes of the urban population turned to the informal economy.
 As \citet[115]{chazkel2011laws} observes, there were few occupations available to lower-class women and foreigners in the 1890s, and a large number of poor workers became street vendors.
@@
 
-Since the mid-twentieth century, the \textit{jogo do bicho} has been illegal but ubiquitous in Brazil.
+Since the mid-twentieth century, the \textit{jogo do bicho} has been illegal but pervasive in Brazil.
 For all purposes, the \textit{jogo} remains criminalised, yet there is virtually no Brazilian city which does not have its local \textit{bicheiros}.
-The \textit{jogo} does not operate exclusively at the margins of the Brazilian law; it is clear that it consciously exploits and subsidises large sectors of the formal economy.
-I describe the animal game organisation structure and their impact in the Brazilian economy in further detail below.
+\citet{cross2006risk} suggest a distinction between informal and illegal markets that may be useful here.
+The animal game started as an informal activity, in which the Baron of Drummond and his associates sold lottery tickets to the public.
+Although the state did not regulate the lottery market, the product itself was not illegal.
+After 1941, however, selling \textit{jogo do bicho} tickets became a criminal offence.
+Since then, the \textit{jogo} has moved to the underground and consciously evaded state regulations.
+Moreover, the game has also exploited and subsidised large sectors of Brazil's formal economy, mainly in poor urban areas.
+I describe the animal game organisation structure and their impact in the Brazilian economic activity in further detail below.
 
@@
 The most important financial mechanism implemented by bankers to help \textit{bicheiros} is the \textit{descarga}, loosely translated as `the unloading'.
-The descarga is the \textit{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}}.
+The \textit{descarga} is the \textit{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}}.
 Booking agents are sometimes unable to honour expensive bets.
@@
 While formal workers can -- at least in theory -- demand higher compensations during a market upswing, the same is not true for the animal game employees.
-The threat of violence reduces the space for collective bargaining with the lottery bosses, and increasing profits at the top of the \textit{jogo do bicho} structure do not trickle down to higher salaries at the bottom.
+The threat of violence reduces the space for collective bargaining with the lottery bosses, and higher profits at the top of the \textit{jogo do bicho} structure do not trickle down to those at the bottom.
 This is one of the reasons why the game bankers turned the game into an oligopoly: the use of violence guarantees not only that new entrants will not be allowed to join the market, but also that profits are concentrated in the hands of very few individuals.
 
-The \textit{jogo do bicho} market formation closely resembles what \citet{fligstein1996markets} calls `markets as politics', in which firms create institutions to restrain the competition and organise the labour force.
+The \textit{jogo do bicho's} market formation closely resembles what \citet{fligstein1996markets} calls `markets as politics', in which firms create institutions to restrain the competition and organise the labour force.
 At the formation of the lottery market, competition is fierce and incumbents an challengers are akin to social movements.
@@
 Precisely because violence could shun off profits, \textit{bicheiros} devised other mechanisms to create a friendly environment for the illegal lottery.
-In the next section, I investigate two of them, costly signalling and reputation building.
+I investigate two of them, costly signalling and reputation building.
 
@@
 Signalling theory predicts that when someone cannot easily observe a characteristic she is interested in, she searches for signals that credibly conveys pieces of that required information.
-In the case of an illegal lottery, the main signal a gambler is looking for is \textit{honesty}, that is, that she has a fair chance of winning the game if she makes the right bet.
+In the case of an illegal lottery, the main signal a gambler is looking for is \textit{honesty}, that is, that she will receive her prize if she buys the winning ticket.
 
@@
 In 1930, opposition leader Getúlio Vargas led a bloodless coup d'état that brought Brazil's First Republic to an end.
-During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by dismantling effectively federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,desouza1983estado,fausto1972revoluccao,fausto2014concise,skidmore1967politics}.
-In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,nava1998lessons,williams2001culture}.
+During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by dismantling effectively federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,fausto1972revoluccao,skidmore1967politics}.
+In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,williams2001culture}.
 Popular music, in turn, occupied an important place in Vargas's project of `brazilianing Brazil'.
-Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{avelar2011brazilian,mccann2004hello,stockler2011samba,vassberg1969villa,vassberg1975villa}.
+Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{mccann2004hello,stockler2011samba}.
 
 By the late 1930s, samba reached a unique position in Brazil's cultural identity.
-In a period when civil and political rights were limited \citep{de2001cidadania,duarte1993vicissitudes}, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}.
+In a period when civil and political rights were limited, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}.
 Patriotic sambas exalted the country's natural beauties and the figure of the `friendly, happy, cordial and industrious' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
@@
 However, in the 1930s the word loses its pejorative connotation in Brazil.
-Mainly due to the work of sociologist \citet{freyre1933casa}, the idea of a racial democracy becomes pervasive in the government discourse, and as a result the word gains a positive tone \citep[4]{reiter2009brazil}.} \citetext{\citealp[47]{dangelo2016samba}; \citealp[51]{vianna1995misterio}}.
-The institutionalisation of the Carnival parade in 1935, and the subsequent increases in public funding to the festival, cemented the relationship between politicians and samba groups \citep{almeida2017carnaval,cabral2016escolas,soihet1998subversao}.
+Mainly due to the work of sociologist \citet{freyre1933casa}, the idea of a racial democracy becomes pervasive in the government discourse, and as a result the word gains a positive tone \citep[4]{reiter2009brazil}.} \citep[51]{vianna1995misterio}.
+The institutionalisation of the Carnival parade in 1935, and the subsequent increases in public funding to the festival, cemented the relationship between politicians and samba groups \citep{cabral2016escolas,soihet1998subversao}.
 
@@
 Unable to cope with the rising costs of the show, the `samba schools', which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}.
-This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,farias2013carnival,misse2011crime,queiroz1992carnaval}.
+This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,misse2011crime,queiroz1992carnaval}.
 
-As I have mentioned in the previous section, the animal game at times faced opposition by the local population.
+The animal game at times faced opposition by the local population.
 The public often perceived the game as immoral and repugnant.
-Moreover, even after the bicho was well-established in Rio de Janeiro, the transition from a competitive betting market to an oligopoly involved the threat and often the use of physical violence against bookmakers who resisted the change \citetext{\citealp[143]{bezerra2009mecenato}, \citealp[52]{labronici2012paratodos}}.
-\emph{Bicheiros} were aware of the reputation costs their strategy entailed.
-They decided to finance samba schools hoping to win `the hearts and minds' of the population and attach a more positive image of the game among urban classes.
-Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but only as individuals who had a private interest in samba \citep[209]{chinelli1993vazio}.
-In 1984, a group of rich \emph{jogo do bicho} financiers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
-The LIESA marked a shift in the Carnival.
-For the first time, \emph{bicheiros} decided to act as a group rather than individuals.
-The organisation consolidated the power of \emph{bicheiros} over the parade and provided a formal mechanism to solve disputes among the samba school patrons \citetext{\citealp[43]{cavalcanti2006carnaval}; \citealp[171]{farias2013carnival}; \citealp[55]{labronici2012paratodos}}.
+\emph{Bicheiros} were aware of that problem.
+They decided to finance samba schools hoping to win the support of the population and attach a more positive image of the game among urban classes.
+Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but in 1984, a group of rich \emph{jogo do bicho} financiers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
+The LIESA marked a shift in the history of the Carnival Parade.
+For the first time, the \emph{bicheiros} decided to act as a group rather than individuals who shared an interest in popular festivals.
+The institution was very effective in expanding the parade, but it also provided other benefits to the \textit{jogo do bicho} bankers.
+The organisation consolidated their power over the Carnival and provided a formal mechanism for the \textit{banqueiros} to solve disputes \citetext{\citealp[43]{cavalcanti2006carnaval}; \citealp[171]{farias2013carnival}; \citealp[55]{labronici2012paratodos}}.
 
@@
 Second, the support of the \emph{jogo do bicho} has increased the political and social clout of the samba schools.
-In a country where the state is not present throughout the territory and human right abuses are frequent \citep{ahnen2003,odonnell1993state,pinheiro2000,pinheiro2001}, \emph{jogo do bicho} bankers, and more recently drug traffickers, have provided private governance to poor areas of Rio de Janeiro by enforcing property rights, mediating disputes, and preventing police abuse in the favelas \citep{arias2006dynamics,goldstein2013laughter,leeds1996cocaine}.
+In a country where the state is not present throughout the territory and human right abuses are frequent \citep{odonnell1993state,pinheiro2000}, \emph{jogo do bicho} bankers, and more recently drug traffickers, have provided private governance to poor areas of Rio de Janeiro by enforcing property rights, mediating disputes, and preventing police abuse in the favelas \citep{arias2006dynamics,goldstein2013laughter}.
 In return for funds and protection from the \emph{bicheiros}, samba schools have served as intermediaries between the underworld and the political system.
@@
 If politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, their relationship with the animal game bankers have become more ambivalent in the last decades.
-The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citetext{\citealp{gaspari2002ditadura}; \citealp{jupiara2015poroes}; \citealp[39]{zaluar2007democratizaccao}}.
-Given the absence of democratic checks and balances, paramilitaries and police forces colluded to repress potential dissidents of the regime and, frequently, to extort civilians \citep{gorender1999combate,magalhaes1997logica,misse2009acumulaccao,skidmore1990politics}.
+The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citep{gaspari2002ditadura,jupiara2015poroes}.
+Given the absence of democratic checks and balances, paramilitaries and police forces colluded to repress potential dissidents of the regime and, frequently, to extort civilians \citep{magalhaes1997logica,misse2009acumulaccao}.
 \emph{Bicheiros} saw the corruption of some members of the military as an opportunity.
@@
 
+Politicians from all spheres of government are involved with \textit{jogo do bicho} bankers.
+Recent investigations have shown that from local politicians to senators receive money from the lottery to fund their campaigns.
+Carlinhos Cachoeira, a famous animal game banker from the state of Goiás established a large patronage network that included mayors, deputies, senators, judges, and businessmen, many of them linked to the federal government.
+His brother Marcos ironically noted that Carlinhos was `too dedicated to politics' and that he illegally donated about USD 300 million to political candidates in his home state \citep{estadao2012gene}.
+
 The Brazilian political system is particularly conductive to clientelistic practices.
 Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}.
-In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp{mainwaring1992brazilian}; \citealp[483]{samuels2000ambition}}.
+In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}.
 These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence.
@@
 The support from the \emph{jogo do bicho} mitigates both problems.
-With respect to the financial costs of campaigns, illegal donations from \emph{bicheiros} help to cover advertising expenses while having the additional benefit of not appearing in the official records of the candidates \citep{congressoemfoco2007bicho,gazetadopovo2007bicho,globo2012bicheiro}.
+With respect to the financial costs of campaigns, illegal donations from \emph{bicheiros} help to cover advertising expenses while having the additional benefit of not appearing in the official records of the candidates \citep{congressoemfoco2007bicho,globo2012bicheiro}.
 This suggests that \emph{jogo do bicho}-funded politicians can circumvent spending limits and have an electoral advantage over their competitors.
 As candidates do not know whether their competitors receive funding from the \emph{jogo do bicho} nor the amount each one was paid, their dominant position is to contact the \emph{bicheiros} and join their networks.
-The situation is a prisoner's dilemma in which candidates would be better off by running cheaper campaigns and not being dependent of \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching an optimal solution.
+The situation is a prisoner's dilemma in which candidates would be better off by running cheaper campaigns and not being dependent of \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching a cheaper solution.
 
-% \begin{figure}[!htbp]
-%  \centering
-%  \includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
-%  \caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right).
-The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'.
-Source: \citet{extra2012aniz}.}
-%  \label{fig:aniz}
-% \end{figure}
-
-The votes from poor communities are instrumental for aspiring politicians.
+Votes from poor communities are instrumental for aspiring politicians.
 Brazil has an enforced compulsory voting system; therefore, turnout rates tend to be higher than in other democracies.
@@
 
-Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}.
-Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
-\textit{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers.
-Furthermore, by investing in the Carnival parade \textit{bicheiros} have been able to gather popular and government support.
-Poor communities have associated with the \textit{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.
 Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}.

### 2018-11-05 paper-bicho/bicho.tex
update

@@
 Second, I investigate how the animal game financiers combine costly signals and selective incentives to induce cooperation from members of community.
-Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of the Brazilian political system to advance their long-term interests.
+Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of Brazil's political system to advance their long-term interests.
 
@@
 A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
-Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce.
+Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce.
 See \href{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}{\texttt{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}}.
@@
 
-In this article I offer a rational choice interpretation of the \textit{jogo do bicho}.
-More specifically, I use an array of bibliographical sources to show how the game operators, called \textit{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.
+In this article I offer a rational choice interpretation of the \textit{jogo do bicho}.\footnote{I adopt a very broad definition of rationality in this paper.
+In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information when making their choices.
+Thus, my analysis employs a `thin' notion of rationality and a `thick' description of social institutions \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, called \textit{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.
 This does not mean that their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals.
@@
 Lastly, this work connects to the literature on state capture, which is one of the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}.
-More specifically, I use the Brazilian case to illustrate how politicians and civil servants can be co-opted by criminal groups and produce bad social outcomes.
+More specifically, I use the Brazilian case to illustrate how politicians and civil servants are co-opted by criminal groups, and how this collusion distorts the electoral process and benefits wealthy members of the \textit{jogo do bicho} network.
 \citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities.
@@
 Then he can legally use that money for any purpose without raising suspicion from the authorities.
-The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called `the slot machine mafia', including relatives of Castor de Andrade \citep{estado2011cacaniquel}.
+The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called `the slot machine mafia'.
+Relatives of Castor de Andrade were also involved in the scheme \citep{estado2011cacaniquel}.
 In 2012, President Dilma Rousseff sanctioned a law that considered slot machines and the \textit{jogo do bicho} as money laundering \citep{agenciabrasil2012dilma}.
@@
 Patriotic sambas exalted the country's natural beauties and the figure of the `friendly, happy, cordial and industrious' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
-The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of the male donkey and a female horse.
+The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of a male donkey and a female horse.
 However, in the 1930s the word loses its pejorative connotation in Brazil.
@@
 They decided to finance samba schools hoping to win the support of the population and attach a more positive image of the game among urban classes.
-Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but in 1984, a group of rich \emph{jogo do bicho} financiers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
+Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but in 1984, a group of rich bankers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
 The LIESA marked a shift in the history of the Carnival Parade.
@@
 Politicians from all spheres of government are involved with \textit{jogo do bicho} bankers.
-Recent investigations have shown that from local politicians to senators receive money from the lottery to fund their campaigns.
+Recent investigations have shown that from local representatives to senators, politicians of every level receive illegal money to fund their campaigns.
 Carlinhos Cachoeira, a famous animal game banker from the state of Goiás established a large patronage network that included mayors, deputies, senators, judges, and businessmen, many of them linked to the federal government.
@@
 These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence.
-Nevertheless, that independence means candidates rely mostly on themselves to raise funds and establish communication with potential voters.
+Nevertheless, that independence means candidates rely mostly on themselves to raise funds and mobilise potential voters.
 Hence, political campaigns in Brazil tend to be expensive and personality-centred.
@@
 Cases studies are usually focused on Rio de Janeiro's \textit{bicheiros}, but scholars would benefit from comparative analyses with a larger number of states.
-This is an important step to elucidate how \textit{bicheiros} continue to influence politics and the public across Brazil.
+This is an important step to elucidate how \textit{bicheiros} continue to influence politics and the public throughout Brazil.
 

### 2018-11-05 paper-bicho/response-reviewers.tex
response reviewers

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{libertine}
\usepackage[libertine]{newtxmath}
\usepackage[scaled=.95]{inconsolata}	
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{darkblue}{rgb}{0.0,0.0,0.55}
\usepackage{setspace}
\usepackage[top=2.3cm,bottom=2.3cm,left=2.3cm,right=2.3cm]{geometry}
\usepackage[breaklinks=true]{hyperref}
\usepackage{graphicx}
\usepackage{float}
\usepackage{mathtools}
\usepackage{caption}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage[authoryear]{natbib}
\usepackage{babelbib}
\cleanlookdateon
\exhyphenpenalty=1000
\hyphenpenalty=1000
\widowpenalty=1000
\clubpenalty=1000
% \urlstyle{same}  % don't use monospace font for urls
\setcitestyle{aysep={}} 
\usepackage{etoolbox}
\makeatletter
\patchcmd{\NAT@citex}
	  {\@citea\NAT@hyper@{%
		 \NAT@nmfmt{\NAT@nm}%
		 \hyper@natlinkbreak{\NAT@aysep\NAT@spacechar}{\@citeb\@extra@b@citeb}%
		 \NAT@date}}
	  {\@citea\NAT@nmfmt{\NAT@nm}%
	   \NAT@aysep\NAT@spacechar\NAT@hyper@{\NAT@date}}{}{}
	\patchcmd{\NAT@citex}
	  {\@citea\NAT@hyper@{%
		 \NAT@nmfmt{\NAT@nm}%
		 \hyper@natlinkbreak{\NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi}%
		   {\@citeb\@extra@b@citeb}%
		 \NAT@date}}
	  {\@citea\NAT@nmfmt{\NAT@nm}%
	   \NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi\NAT@hyper@{\NAT@date}}
	  {}{}
\makeatother

\hypersetup{pdftitle={Beasts of Prey or Rational Animals? Private Governance in Brazil's Jogo do Bicho},
	pdfauthor={Danilo Freire},
	pdfborder={0 0 0},
	breaklinks=true,
	linkcolor=Mahogany,
	citecolor=Mahogany,
	urlcolor=darkblue,
	colorlinks=true}

\doublespacing

\title{Beasts of Prey or Rational Animals? \\Private Governance in Brazil's \textit{Jogo do Bicho}}

\author{Response to the Reviewers}

\date{\today}

\begin{document} %
\maketitle

I would like to thank the four reviewers for their thoughtful criticisms and suggestions. Their comments have allowed me to reconsider important aspects of my manuscript and, I believe, improve it. Below, I describe in detail the changes I have made in response to the reviewers' individual comments.

\subsection*{Reviewer 1: G. Leddy} % (fold)
\label{sub:reviewer_1_g_leddy}

\noindent 1) \textit{In one case the vagueness of the theory is clear ``civil servants can be co-opted by criminal groups and produce sub-optimal social outcomes.'' Sub-optimal is vague. [\dots] The problem I have with the theoretical approaches used in this text is that they are neutralistic, evading discussions of social conflict in favor of an amalgam of theories based on rational choice and functional aspects of government.}

\vspace{.25cm}

I completely agree with the comment. I have made substantial changes to the manuscript in order to better illustrate the social tensions inherent in the \textit{jogo do bicho}. For instance, in paragraph 6 of section 2.2, I highlight that ``[\dots] \textit{the banqueiros do not use violence only against other criminals. They often employ violent methods against competitors and their own staff, too}", and give some examples of how \textit{jogo do bicho} bankers employ violence to run their business. Paragraphs 1 and 2 in section 4.2 describe other ways by which the \textit{banqueiros} enforce their rule over employees and members of civil society.

I also adopt a more nuanced approach when analysing the relationship between the \textit{jogo} and the public. In the last 5 paragraphs of section 2.2, I point out that not only the animal game has negatively affected the formal labour market, but it has also reduced the demand for social rights in Brazil. The \textit{jogo} provides jobs for those would could not easily enter the labour force, and as result ``[\dots] \textit{poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population}.'' I also mention how the \textit{jogo} has collaborated to increase inequality levels in the country.

I believe the text is not as neutralistic as it was before, as social tensions now deservedly occupy a more prominent role in the manuscript.

\vspace{.5cm}

\noindent 2) \textit{I think this could have been a more LAP style paper if it had expanded on the role of gambling rackets in the economy, the specific types of political candidates supported by the jogo, the overall impact of this much cash circulating and how it is harnessed by the formal economy, and the manipulation of poor district votes. I find much of the discussion on how the game is perceived and the theories of choice to be less interesting. What welfare provisions are provided by the jogo? Why is the only discussion of violence saved for the conclusion? The most interesting questions (as far as the VCC issue is concerned) are found in the conclusion.}

\vspace{.25cm}

The reviewer is correct to point out that those topics should have been described in further detail. I discuss the role of gambling rackets in the economy in the last paragraphs of section 2. They address specifically that issue. In the case of the political candidates supported by the \textit{jogo}, I note in paragraph 3, section 4.2, that `[\dots] \textit{Politicians from all spheres of government are involved with jogo do bicho bankers. Recent investigations have shown that from local representatives to senators, politicians of every level receive illegal money to fund their campaigns.}.' The claim is backed by newspaper articles describing Carlinhos Cachoeira's wide political patronage network. Cachoeira is a well-known \textit{jogo do bicho} boss in his home state of Goiás. I discuss the overall impact of the \textit{jogo} in the formal economy in the last paragraphs of section 2.2.

The welfare provisions offered by the \textit{jogo} have received a longer treatment in this new version. I argue that the \textit{banqueiros} provide a mix of carrots and sticks to enforce compliance from bookmakers, and describe how the \textit{jogo} elite give cash benefits and risk-sharing contracts as incentives to their employees. The parts on the funding of samba schools have been revised and are available in section 4.1.

\vspace{.5cm}

\noindent 3) \textit{Footnotes show go at the end. Spelling should be US and not UK. Links do not work and should not be in the footnotes.}

\vspace{.25cm}

I have addressed these issues in the revised version of the manuscript. I have checked the internet links, moved the footnotes to the end, and adopted American spelling in the manuscript.

\vspace{.5cm}

I would like to thank G. Leddy for their valuable comments. I believe I have answered each one in turn, and that the manuscript has improved as a result.

% subsection reviewer_1_g_leddy (end)

\newpage

\subsection*{Reviewer 2: Paulo Simões} % (fold)
\label{sub:reviewer_2_paulo_simoes}

\noindent 1) \textit{It is written in (British) English, and there are minor grammatical problems that can be easily corrected. It should not be too difficult to edit. Also, though the actual article is 27 pages, the extensive bibliography (mostly made up of secondary sources) takes up another 19 pages! I wonder how much of this is necessary.}

\vspace{.25cm}

I would like to express my gratitude to Paulo Simões for his excellent comments, which encouraged me to improve the article along the lines suggested.

The article has been proofread and the text now uses American spelling as requested. With regards to the reference list, I fully agree that it was too extensive. The bibliography is about 50\% shorter, and in my view the text remains as informative as before. Hopefully the manuscript is now easier to read.

\vspace{.5cm}

\noindent 2) \textit{Though I believe the article should be published, I think it would be good to insert more specific examples (both historic and current) taken from newspapers and other media of how the jogo do bicho has and continues to affect politics and society. Right now the article is too academic. A little journalism might make it more effective.}

\vspace{.25cm}

I have included several newspaper stories to the article with the aim of making it more appealing to a wider audience. For instance, in section 2.2 I mention a story about two bankers and their rise to power in Rio de Janeiro:

\begin{quote}
	\citet{jupiara2015poroes} argue that Ailton Guimarães Jorge, a former Army officer, tortured and murdered rival lottery bosses in the late 1970s. One of his former allies, Army Colonel Paulo Malha ̃es, told to the Rio de Janeiro State Truth Commission that Guimarães `went on a rampage' to consolidate his power \citep{belem2015guimaraes}. Castor de Andrade, Rio's most influential animal game banker, also employed similar methods to run his business. Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors. In a famous case, Andrade shot Euclides Ponar, an old jogo do bicho boss known as `Grey-Headed Chinese' (\textit{China Cabeça Branca}) after Ponar denounced a fraud in lottery draws in 1976. Andrade was likely involved in other assassination plots in the 1990s \citep{globo2017castor}.
\end{quote}

There are a number of similar excerpts in other parts of the manuscript. For instance, in the paragraph below I write about how the animal game profit from slot machines in Brazil:

\begin{quote}
	The owner of a slot machine issues a ticket with a winning prize, and the criminal declares the prize as his legitimate wealth. Then he can legally use that money for any purpose without raising suspicion from the authorities. The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called `the slot machine mafia'. Relatives of Castor de Andrade were involved in the scheme \citep{estado2011cacaniquel}. In 2012, President Dilma Rousseff sanctioned a law that considered slot machines and the \textit{jogo do bicho} as money laundering \citep{agenciabrasil2012dilma}.
\end{quote}

In another example, I allude to the history of Carlinhos Cachoeira to show how pervasive is the animal game in Brazil, an argument that the reviewer also makes in his referee report.

\vspace{.5cm}

Once again, I thank the referee for his interesting suggestions.

% subsection reviewer_2_paulo_simoes (end)
\newpage

\subsection*{Reviewer 3: Matthew Lorenzen}%
\label{sub:reviewer_3_matthew_lorenzen}

\noindent 1) \textit{The author's methodology is, at first, not clear. It eventually becomes apparent that the article is based on an analysis of bibliographical sources, and not on any primary data or secondary statistical data. The author should make this clear in the introduction.}

\vspace{.25cm}

I have inserted one line in the introduction to clarify that I am dealing with secondary sources in this article. Indeed this was not obvious to readers.I now state the following in the introduction: ``[\dots] \textit{In this article, I offer a rational choice interpretation of the \emph{jogo do bicho}. More specifically, I use an array of bibliographical sources to show how the game operators, called \emph{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.}''

\vspace{.5cm}

\noindent 2) \textit{For example, is there a connection between internal governance in the jogo do bicho and cooptation of civil servants?}

\vspace{.25cm}

The reviewer is right to suggest that the connection between those two sections was unclear. As mentioned above, I now emphasise two theoretical arguments -- costly signalling and reputation strategies -- as the main tools to understand the behaviour of the \textit{jogo do bicho} elite. These two methods have been instrumental to the profitability of their business, and these large profits have allowed the group to venture into politics. In the revised version of the manuscript I try to emphasise this connection, which I agree was not fully developed in the previous text.

\vspace{.5cm}

\noindent 3) \textit{In addition, the long examination of the historical background, including the directed acyclic graph, does not seem to be crucial for the rest of the article. Moreover, by attempting to address too many issues, the author does not focus on a theoretical perspective (or contrasts theoretical perspectives), but instead briefly mentions different perspectives for each separate issue (repugnant markets, game theory, state capture...). I suggest the author to focus more keenly on one or two issues and one or two different theoretical perspectives, rather than attempting to address all topics in an overarching way.}

\vspace{.25cm}

This is an important point. The article deals with a number of topics that could indeed have been better connected. In that sense, I have removed the discussion about repugnant markets from the manuscript. The section on repugnance added little to the main ideas of the text and it is not required for readers to understand the remaining of the article. Additionally, I have stressed the role of costly signalling and reputation strategies as governance instruments of the \textit{jogo}, and they now appear more prominently in the text.

Section 2 (Historical Background) has also been significantly reduced and it now describes only the essential conditions for the emergence of the animal game. Moreover, I agree that the acyclical graph does not add much to readers, so I have removed it along with the extensive footnote that followed it. I hope this change meets the expectations of the reviewer.

\vspace{.5cm}

I would like to express my sincere thanks to the referee for his excellent comments. I honestly appreciate the constructive feedback.

% subsection reviewer_3_matthew_lorenzen (end)
\newpage

\subsection*{Reviewer 4: Andrew Smolski}%
\label{sub:reviewer_4_andrew_smolski}

\noindent 1) \textit{Agents are differentiated by their role (bookmaker, banker, etc.), yet there is little discussion of how they form a hierarchy of power to affect. It appears as if each agent has a similar power to affect, which muddles the understanding of the ability of bicheiros to implement their strategies. For instance, in the discussion of state capture through the samba schools, a prior understanding of the hierarchy of roles would help the reader to understand the ability to impose oligopolistic conditions, as well as strengthen the argument concerning the role of bicheiros in market formation. See: Neil Fligstein's work on incumbent-challengers in capitalist markets, a political-cultural theory of market formation and alteration}.

\vspace{.25cm}

The referee rightly points out that I did not emphasise strongly enough the power structures of the animal game. This criticism also appeared in other referee reports and I am happy to address it here. As previously mentioned, section 2.2 has been largely rewritten and it now included a few paragraphs stressing the social conflicts within the \textit{jogo}. I have also changed the subtitle to `A Hierarchical Organisational Structure' to make this point clear. The indication of Neil Fligstein's work was very appropriate and I sincerely thank the referee for it. I have added one paragraph about Fligstein's concept of `markets as politics' to conclude section 2.2.

\vspace{.5cm}

\noindent 2) \textit{Further, and following from this concern over hierarchy, at the beginning of section 2.1, the author writes, \emph{``Spontaneous orders are emergent macro-level phenomena that result from voluntary actions of purposive, self-interested individuals utilizing their contextual knowledge}'' (emphasis mine). What does the author mean by voluntary? For instance, urban poverty as a structural factor places severe limitations on what actions can be taken, and this goes beyond just a matter of knowledge. So, while neo-institutionalists like North would reduce institutions to schemas for action, others in the institutional political economy and neo-Marxist traditions, argue that institutions are also very much a matter of power (see Portes 2010 for instance). Considering LAP is a critical journal, the author would do well to explain the rational choice sense of agency, and contrast it with more critical accounts.}

\vspace{.25cm}

In the revised version of the manuscript, I have removed the first part of section 2 -- the one quoted by the reviewer -- because I do agree that it is misleading. The quote downplays the violent aspects of the animal game's internal governance, and as such I do not believe it should be included. In contrast, I have added examples of how the \textit{jogo} bosses employ violence, or the threat thereof, against their own employees to enforce compliance to their decisions. 
,
Nevertheless, I argue that violence can be used in rational terms, and I believe it is the case with the \textit{jogo}. However, I adopt a model of rationality which also assumes humans are embedded in a myriad of formal and informal institutions that design and shape their choices. It is unrealistic to assume otherwise. Apart from institutional constrains, individuals have limited access to information, what limits their abilities to correctly evaluate alternative scenarios. But in the \textit{jogo}, individuals generally behave in ways that maximise their own interest. This `thin rationality' conforms well to a wide range of the literature on crime and violence, such as \citet{balcells2010rivalry}, \citet{gambetta1996sicilian,gambetta2009codes}, \citet{kalyvas2006logic}, \citet{leeson2010pirational}, or \citet{skarbek2011governance,skarbek2012prison}.

I did not include a lengthier discussion about the rational choice sense of agency because I thought it would be slightly too abstract for my purposes. The article is mainly focused on the bankers' strategies in the long run, and I believe readers will be more interested in more concrete aspects of the animal game. But I did add a footnote to the introduction describing the notion of rationality I employ in the text. The footnote reads as follows:

\begin{quote}
	I adopt a very broad definition of rationality in this paper. In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information when making their choices. Thus, my analysis employs a ‘thin’ notion of rationality and a ‘thick’ description of social institutions (Boettke 2001, 253).
\end{quote}

I hope this answers the question posed by the referee.

\vspace{.5cm}

\noindent 3) \textit{The author does not conceptualize the informal (extra-legal) economy in contrast to the formal economy. In the literature, the informal economy is not just an economy absent state regulation, although this is of course important. Rather, it is wrapped up in questions of exploitation and subsidizing the formal economy, and thus once again questions of power. See for instance Centeno and Portes 2006. Also, informal and extra-legal are not always the same thing, as pointed out by Cross and Peña 2006. There is a difference in selling an illegal product illegally and selling a legal product illegally. This gets at the idea of the repugnant market and morality of a market arguments, and could help to clarify further the exact definition of an extra-legal market.}

\vspace{.25cm}

This is also a very interesting suggestion. I have followed the reviewer's advice and included a new paragraph citing \citet{cross2006risk} and showed how the \textit{jogo} takes advantage of the abundance of unskilled workers in the formal labour force. I quote a few excerpts of the text below:

\begin{quote}
	[\dots] \citet{cross2006risk} suggest a distinction between informal and illegal markets that may be useful here. The animal game started as an informal activity, in which the Baron of Drummond and his associates sold lottery tickets to the public. Although the state did not regulate the lottery market, the product itself was not illegal. After 1941, however, selling \textit{jogo do bicho} tickets became a criminal offence. Since then, the \textit{jogo} has moved to the underground and consciously evaded state regulations. Moreover, the game has also exploited and subsidised large sectors of Brazil's formal economy, mainly in poor urban areas. [\dots]

	[\dots] For instance, the \textit{jogo do bicho} gives a boost to the Brazilian economy by providing jobs for unskilled workers who cannot easily join the labour market. By doing so, the \textit{jogo} prevents some of poorest members of the Brazilian society from demanding more inclusive government policies, although they are the ones who would benefit the most from public assistance. Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population.

	[\dots] Another indirect economic effect of the \textit{jogo} is the rise in inequality. While formal workers can -- at least in theory -- demand higher compensations during a market upswing, the same is not true for the animal game employees. The threat of violence reduces the space for collective bargaining with the lottery bosses, and higher profits at the top of the \textit{jogo do bicho} structure do not trickle down to those at the bottom. This is one of the reasons why the game bankers turned the game into an oligopoly: the use of violence guarantees not only that new entrants will not be allowed to join the market, but also that profits are concentrated in the hands of very few individuals.  
\end{quote}

\vspace{.5cm}

\noindent 4) \textit{The point made by Arrow used in section 3.3 that criminals ``do not put the required levels of effort if they know they are not being monitored'', lacks empirical backing in the literature. It is to take a doxa about criminals and treat it as true. I suggest looking at contemporary work on pirates for instance, and other extra-legal organizations, instead of using moral judgements as theoretical arguments. Remember, who is and who is not a criminal is a social matter, not a universal. Police invading a favela is a criminal act, but we do not label them criminals. Or, we can state that criminals are always those that follow Arrow's definition, but then you would have to demonstrate that the agents in the jogo do bicho act this way. Otherwise, it substitutes empirical reality with a possibly unfounded theoretical conjecture. This also returns us to point 3, whereby clarifying extra-legal economy would help to understand extra-legal actors.}

\vspace{.25cm}

I think this is an important criticism. Due to poor phrasing, the sentence highlighted by the referee does suggest a moralistic undertone. Yet I would like to affirm that it was not the case. My intention was not to use a moral judgements to describe the character or motivation of individuals, but only to draw attention to potential agent-principal dilemmas. This problem is widely discussed in the economics literature and it not ascribed only to informal or illegal activities, but pervasive in firms big or small and in government agencies \citep[e.g.,][]{grossman1983analysis,laffont2009theory,sappington1991incentives}. I sincerely apologise for the confusion.

I have removed the mention to Arrow's work to avoid further misunderstandings. As I have mentioned previously, I believe the revised manuscript exposes more clearly the social tensions of the animal game, and this has improved the quality of the text.

\vspace{.5cm}

\noindent 5) \textit{Lastly, I suggest the author look at work by Peter Evans, and other sociological institutional political economists, who have worked on State embedding. The State itself is embedded within a social structure that dictates why it acts the way it does. That structure is cultural, political, and economic, producing social path dependence for the institutions that operate within it. Instead, the State is here left undefined as an entity. As such, all the agents operate the same despite their institutional and structural context, which makes overly rational agents that are actually bounded.}

\vspace{.25cm}

Thank you very much for another valuable recommendation. I have added two paragraphs to section 4.2 framing the state as an embedded entity as suggested by the reviewer.

\vspace{.5cm}

I am grateful to the Andrew Smolski for the detailed comments and constructive criticism.

% subsection reviewer_4_andrew_smolski (end)

\newpage
\bibliography{bibliography.bib}
\bibliographystyle{apalike}
\end{document} 
% end of fold

### 2018-11-06 paper-bicho/response-reviewers.tex
update response reviewers

@@
 I completely agree with the comment.
-I have made substantial changes to the manuscript in order to better illustrate the social tensions inherent in the \textit{jogo do bicho}.
-For instance, in paragraph 6 of section 2.2, I highlight that ``[\dots] \textit{the banqueiros do not use violence only against other criminals.
-They often employ violent methods against competitors and their own staff, too}", and give some examples of how \textit{jogo do bicho} bankers employ violence to run their business.
-Paragraphs 1 and 2 in section 4.2 describe other ways by which the \textit{banqueiros} enforce their rule over employees and members of civil society.
+I have made substantial changes to the manuscript to better illustrate the social tensions inherent in the \textit{jogo do bicho}.
+For instance, in section 2.2 I highlight that ``[\dots] \textit{the banqueiros do not use violence only against other criminals.
+They often employ violent methods against competitors and their own staff, too}." Then I give examples of how the \textit{jogo do bicho} elites frequently exploit their subordinates.
+Similarly, in section 4.2 I describe other ways by which the \textit{banqueiros} enforce their rule over employees and members of civil society.
+In this sense, social conflicts are more prominent in my description of the lottery.
 
@@
 The \textit{jogo} provides jobs for those would could not easily enter the labour force, and as result ``[\dots] \textit{poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population}.'' I also mention how the \textit{jogo} has collaborated to increase inequality levels in the country.
+Discussions about how the animal game distorts the allocation of public funds and corrupts political representation are also included in the last two sections.
 
-I believe the text is not as neutralistic as it was before, as social tensions now deservedly occupy a more prominent role in the manuscript.
+I believe the text is not as neutralistic as before, and hopefully these changes have illustrated how pervasive are social tensions in the animal game.
 
@@
 
-The reviewer is correct to point out that those topics should have been described in further detail.
-I discuss the role of gambling rackets in the economy in the last paragraphs of section 2.
-They address specifically that issue.
-In the case of the political candidates supported by the \textit{jogo}, I note in paragraph 3, section 4.2, that `[\dots] \textit{Politicians from all spheres of government are involved with jogo do bicho bankers.
-Recent investigations have shown that from local representatives to senators, politicians of every level receive illegal money to fund their campaigns.}.' The claim is backed by newspaper articles describing Carlinhos Cachoeira's wide political patronage network.
+The reviewer is correct to point out that I should have described those topics should have in further detail.
+I now dedicate the last paragraphs of section 2 to a discussion about the role of gambling rackets in Brazil's formal economy.
+The paragraphs address specifically the issue raised by the reviewer.
+In the case of the political candidates supported by the \textit{jogo}, I note in paragraph 3 of  section 4.2 that `[\dots] \textit{Politicians from all spheres of government are involved with jogo do bicho bankers.
+Recent investigations have shown that from local representatives to senators, politicians of every level receive illegal money to fund their campaigns.}.' The claim is backed by a description of Carlinhos Cachoeira's wide political patronage network.
 Cachoeira is a well-known \textit{jogo do bicho} boss in his home state of Goiás.
-I discuss the overall impact of the \textit{jogo} in the formal economy in the last paragraphs of section 2.2.
+As mentioned in the previous response, I discuss the overall impact of the \textit{jogo} in the formal economy in the last paragraphs of section 2.2.
 
 The welfare provisions offered by the \textit{jogo} have received a longer treatment in this new version.
-I argue that the \textit{banqueiros} provide a mix of carrots and sticks to enforce compliance from bookmakers, and describe how the \textit{jogo} elite give cash benefits and risk-sharing contracts as incentives to their employees.
+I argue that the \textit{banqueiros} provide a mix of carrots and sticks to enforce compliance from bookmakers, and describe how the \textit{jogo} elite offers cash benefits and risk-sharing contracts as incentives to their employees.
 The parts on the funding of samba schools have been revised and are available in section 4.1.
@@
 
-I have addressed these issues in the revised version of the manuscript. I have checked the internet links, moved the footnotes to the end, and adopted American spelling in the manuscript.
+I have addressed these issues in the revised version of the manuscript.
+I have checked the internet links, moved the footnotes to the end, and adopted American spelling in the manuscript as per the reviewer's request.
 
@@
 
-I would like to thank G. Leddy for their valuable comments. I believe I have answered each one in turn, and that the manuscript has improved as a result.
+I would like to thank George Leddy for his very valuable comments. I believe I have answered each one in turn, and that the manuscript has improved as a result.
 
@@
 With regards to the reference list, I fully agree that it was too extensive.
-The bibliography is about 50\% shorter, and in my view the text remains as informative as before.
+The bibliography is now about 50\% shorter, and in my view the text remains as informative as before.
 Hopefully the manuscript is now easier to read.
@@
 
-There are a number of similar excerpts in other parts of the manuscript. For instance, in the paragraph below I write about how the animal game profit from slot machines in Brazil:
+There are a number of similar excerpts in other parts of the manuscript. For instance, I also add more details about how the animal game profits from slot machines in Brazil:
 
@@
 
-In another example, I allude to the history of Carlinhos Cachoeira to show how pervasive is the animal game in Brazil, an argument that the reviewer also makes in his referee report.
+In another example, I allude to the history of Carlinhos Cachoeira to show how widespread is the animal game in Brazil, giving more evidence to an argument the reviewer makes in his report.
 
@@
 
-Once again, I thank the referee for his interesting suggestions.
+Once again, I thank Paulo Simões for his interesting suggestions.
 
@@
 
-I have inserted one line in the introduction to clarify that I am dealing with secondary sources in this article.
-Indeed this was not obvious to readers.I now state the following in the introduction: ``[\dots] \textit{In this article, I offer a rational choice interpretation of the \emph{jogo do bicho}.
-More specifically, I use an array of bibliographical sources to show how the game operators, called \emph{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.}''
+I have inserted one line in the introduction to clarify that I am dealing with secondary sources in this article. Indeed this was not obvious to readers. I now state the following in the introduction:
+
+\begin{quote}
+	In this article, I offer a rational choice interpretation of the \emph{jogo do bicho}.
+More specifically, I use an array of bibliographical sources to show how the game operators, called \emph{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.
+\end{quote}
 

### 2018-11-06 paper-bicho/response-reviewers.tex
edit

@@
 
-I have inserted one line in the introduction to clarify that I am dealing with secondary sources in this article. Indeed this was not obvious to readers. I now state the following in the introduction:
+I have mentioned in the introduction that the article uses secondary sources.
+Indeed this was not obvious to readers and I thank the reviewer for suggesting the clarification.
+The introduction now reads as follows:
 

### 2018-11-06 paper-bicho/response-reviewers.tex
update

@@
 
-The reviewer is right to suggest that the connection between those two sections was unclear.
-As mentioned above, I now emphasise two theoretical arguments -- costly signalling and reputation strategies -- as the main tools to understand the behaviour of the \textit{jogo do bicho} elite.
-These two methods have been instrumental to the profitability of their business, and these large profits have allowed the group to venture into politics.
-In the revised version of the manuscript I try to emphasise this connection, which I agree was not fully developed in the previous text.
+The reviewer is right that the connection between those two sections was unclear.
+In the revised version of the manuscript, I emphasise how two theories -- costly signalling and reputation strategies -- largely explain the behaviour of the \textit{jogo do bicho} elite.
+Signalling and reputation have been instrumental to the profitability of the lottery, and such capital accumulation has allowed the \textit{banqueiros} to extend their influence in the political realm.
+I agree this connection between capital accumulation and political power was not fully developed in the previous version of the text, and I hope readers can understand my theoretical argument more clearly.
 
@@
 This is an important point.
-The article deals with a number of topics that could indeed have been better connected.
-In that sense, I have removed the discussion about repugnant markets from the manuscript.
-The section on repugnance added little to the main ideas of the text and it is not required for readers to understand the remaining of the article.
-Additionally, I have stressed the role of costly signalling and reputation strategies as governance instruments of the \textit{jogo}, and they now appear more prominently in the text.
+The article deals with a number of topics that seemed only loosely associated.
+The section on repugnance, for instance, added little to the main ideas of the text and is not required for readers to understand the remaining of the article.
+That section has been removed from the manuscript and the subsections on costly signalling and reputation strategies now appear more prominently in the text.
+The article is, in my view, more focused and concise as a result.
 
@@
 The referee rightly points out that I did not emphasise strongly enough the power structures of the animal game.
-This criticism also appeared in other referee reports and I am happy to address it here.
-As previously mentioned, section 2.2 has been largely rewritten and it now included a few paragraphs stressing the social conflicts within the \textit{jogo}.
-I have also changed the subtitle to `A Hierarchical Organisational Structure' to make this point clear.
-The indication of Neil Fligstein's work was very appropriate and I sincerely thank the referee for it.
+This criticism also appeared in other referee reports and I am happy to address it.
+As previously mentioned, section 2.2 has been largely rewritten and it now includes new paragraphs stressing the social conflicts within the \textit{jogo}.
+I have also changed the section title to `A Hierarchical Organisational Structure' to make this point clear.
+The indication of Neil Fligstein's work is also very appropriate and I sincerely thank the referee for it.
 I have added one paragraph about Fligstein's concept of `markets as politics' to conclude section 2.2.
@@
 
-Thank you very much for another valuable recommendation. I have added two paragraphs to section 4.2 framing the state as an embedded entity as suggested by the reviewer.
+Thank you very much for another valuable recommendation. The last three paragraphs of section 4.2 frame the state as an embedded entity as suggested by the reviewer.
 

### 2018-11-06 paper-bicho/bicho.tex
update

@@
 As elections may be decided by a small difference, the \emph{bicheiros}' clientelistic ties guarantee a minimum number of votes that politicians can rely upon on election day.
-Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process per se.
+Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process \textit{per se}.
 Individuals may be punished if the candidate does not receive the expected number of votes, and are often compelled to vote for politicians that have only loose connections with their communities.
@@
 
-Finally, the \emph{jogo do bicho} patronage highlights a crucial social dilemma within the Brazilian public law.
-Even though federal judges have prosecuted \emph{bicheiros}, politicians and police forces have no incentives to enforce the punishment.
-Although Brazilian judges enjoy job stability, the latter groups constantly require local-level support from the \emph{bicheiros}.
-Politicians and police officers may have accurate information on \emph{jogo do bicho} operations and \emph{bicheiros}' whereabouts, but the federal government cannot rely upon their cooperation.
-That can be one of the reasons why even after many attempts to arrest \emph{bicheiros}, there has been little progress in that regard in Brazil's latest democratic period (1985--present).
+The difficulty in permanently outlawing the \textit{jogo} showcases how the Brazilian state itself is deeply embedded within criminal sectors of the society.
+While some sectors of the Brazilian bureaucracy may be described as `islands of excellence' and closely resemble the Weberian ideal of public administration \citep{bersch2017state}, local politicians remain dependent of unstable, and often unlawful, connections with social elites.
+This system of `partial embedded autonomy' provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour \cite{evans1995embedded}.
+This structure tends to perpetuate itself as it brings benefits for both parts, the animal game bankers and politicians.
+The partial autonomy of state agents also highlight a social dilemma within the Brazilian public law.
+Although federal judges have successfully prosecuted \textit{bicheiros}, local state agents are not willing to comply with judicial decisions.
+The lottery elites are soon able to return to their positions in little time.
+As a result,
 

### 2018-11-07 paper-bicho/bicho.tex
pr sec 3

@@
 The impact of the \emph{jogo do bicho} is not restricted to the Brazilian economy.
-Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural and social festivity, the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,chinelli1993vazio,queiroz1992carnaval}.
+Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural festivity, the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,queiroz1992carnaval}.
 The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that `without the \emph{jogo do bicho} the Carnival would have ended' \citep{odia2016aniz}.
-Owing to that support, \emph{bicheiros} have established an extensive patronage network with samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}.
-Although that network brings large material benefits to their members, the patronage system has created perverse incentives for government officials.
+Based on that support, \emph{bicheiros} have established an extensive patronage network over samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}.
+Such clientelistic network brings large material benefits to their members, yet it has created perverse incentives for government officials and causes several distortions to the Brazilian democratic system.
 
-The \emph{jogo do bicho}'s clientelism is more evident in the state of Rio de Janeiro than in other parts of the country.
+The \emph{jogo do bicho}'s clientelism is most evident in the state of Rio de Janeiro.
 Historical factors explain why this is the case.
-Firstly, Rio de Janeiro city was the capital of Brazil for almost 200 years; despite losing the position to Brasília in 1960, it remains one of the country's main cultural and financial centres.
-Secondly, \emph{jogo do bicho} operators had historical ties with popular movements, which they eventually exploited to their advantage.
+First, Rio de Janeiro city was the capital of Brazil for almost 200 years, and despite losing the position to Brasília in 1960, it has remained one of the country's main cultural and financial centres.
+Secondly, \emph{jogo do bicho} operators had historical ties with popular movements, especially samba groups, and the animal game elites eventually exploited these connections to their advantage.
 Thirdly, the emergence of state-sponsored Carnival parades created a window of opportunity for \emph{bicheiros} to expand their influence over public authorities, either via bribing or by funding political campaigns.

### 2018-11-08 paper-bicho/response-reviewers.tex
more rev

@@
 \begin{quote}
-	[\dots] \citet{cross2006risk} suggest a distinction between informal and illegal markets that may be useful here.
+	[\dots]  \citet{cross2006risk} suggest a distinction between informal and illegal markets that is useful to understand the current `semi-legal' status of the \textit{jogo}.
 The animal game started as an informal activity, in which the Baron of Drummond and his associates sold lottery tickets to the public.
-Although the state did not regulate the lottery market, the product itself was not illegal.
-After 1941, however, selling \textit{jogo do bicho} tickets became a criminal offence.
+Although the state did not regulate the lottery market, the product itself was not illegal, allegedly not even immoral.
+In fact, many organisations, including the Catholic Church, sold raffles and lottery tickets to sponsor their activities \citep[49]{torcato2011repressao}.
+Moreover, the state initially found no contradiction between the animal game and its own lottery, and both coexisted for about 15 years \citep[559]{chazkel2007beyond}.
+After 1941, however, selling \textit{jogo do bicho} tickets became a legal offence after a long campaign asking for its criminalisation.
 Since then, the \textit{jogo} has moved to the underground and consciously evaded state regulations.
-Moreover, the game has also exploited and subsidised large sectors of Brazil's formal economy, mainly in poor urban areas.
-[\dots]
+This shift from informal to illegal has brought important changes to the whole dynamics of the \textit{jogo}.
+As an example, wealthy animal game bosses have colluded with shady sectors of the armed forces for protection, and this violence has also used against potential competitors.
+Moreover, this move to illegality enabled the game to exploit and subsidise large sectors of Brazil's formal economy, mainly in poor urban areas.
 

### 2018-11-08 paper-bicho/bicho.tex
more rev

@@
 In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
-Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, but they soon allowed participants to choose the animals they preferred.
+Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, so the game consisted in a simple raffle.
+However, the zoo soon allowed participants to choose the animals they preferred.
 This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.} The lucky winner would take home a prize worth 20 times the ticket price, an amount higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.
@@
 The campaign extended for several decades and received considerable support from the \textit{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}.
-The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and it remains illegal until this day.
+The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and the lottery remains illegal until this day.
 
@@
 But more than a act of civic disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
-A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
-Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce.
+A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
+Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce.
 See \href{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}{\texttt{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}}.
@@
 In this article I offer a rational choice interpretation of the \textit{jogo do bicho}.\footnote{I adopt a very broad definition of rationality in this paper.
-In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information when making their choices.
-Thus, my analysis employs a `thin' notion of rationality and a `thick' description of social institutions \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, called \textit{bicheiros}, have developed unique strategies to solve collective action problems and maximise their political strategies.
+In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information at the moment of their choices.
+Thus, my analysis employs a `thin' notion of rationality and a `thick' description of social institutions, where individual action can only be understood with reference to the social environment \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, commonly called \textit{bicheiros},\footnote{As I describe in the next section, the \textit{jogo do bicho} structure can be broadly divided into three occupations: the rich game financiers (\textit{banqueiros}), mid-level managers (\textit{gerentes}), and street bookmakers (\textit{bicheiros}).
+While there are important hierarchical differences among these groups, all members of the \textit{jogo} are collectively called \textit{bicheiros} in Brazil.
+I follow the same practice here, and only refer to their specific role within the animal game structure when necessary for clarification.} have developed unique strategies to solve collective action problems and maximise their political strategies.
 This does not mean that their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals.
@@
 In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \textit{jogo do bicho} and the military regime in Brazil (1964--1985).
-I supplement this literature by highlighting how asymmetrical information, agency dilemmas, and rent-seeking behaviour offer convincing explanations to the issues presented above.
-Although those concepts have a long tradition in public choice, scholars have not applied those ideas thus far to understand the dynamics of the \textit{jogo do bicho}.
+I supplement this literature by highlighting how asymmetrical information and rent-seeking behaviour offer convincing explanations to the issues presented above.
+Although those concepts have a long tradition in public choice, scholars have not applied those ideas to understand the dynamics of the \textit{jogo do bicho}.
 By doing so, I integrate seemingly contradictory historical facts into a single narrative that connects micro-level decisions to macro-level outcomes.
@@
 Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising.
-Brazil's growing cities offered new occupations for former slaves who desired to move away from their former masters \citep{andrews1991blacks, skidmore1993black}.
-Increasing numbers of Asian and European immigrants joined the freed slaves and moved to the cities soon after they arrived in Brazil \citep{hall1969origins, lesser2013immigration}.
-However, the hopes of the African-Brazilians and the new foreign settlers would soon be frustrated by a series of economic downturns.
-The Brazilian labour market suffered a severe contraction in the wake of the \textit{Encilhamento} financial crisis of 1891, and the economic instability aggravated the already difficult conditions of the working classes \citep{topik2014political, triner2005baring}.
+Cities like São Paulo and Rio de Janeiro offered a number of occupations for former slaves who wanted to move away from their former rural masters \citep{andrews1991blacks, skidmore1993black}.
+Increasing numbers of Asian and European immigrants also moved to large cities after arriving in Brazil, as urban areas usually provided better standards of living than the countryside \citep{hall1969origins, lesser2013immigration}.
+However, the hopes of the African-Brazilians and the new foreign settlers would be frustrated by a series of downturns in the Brazilian economy.
+Brazil's labour markets suffered a severe contraction in the wake of the \textit{Encilhamento} financial crisis of 1891, and the economic instability aggravated the already difficult conditions of the working classes \citep{topik2014political, triner2005baring}.
 
@@
 As \citet[115]{chazkel2011laws} observes, there were few occupations available to lower-class women and foreigners in the 1890s, and a large number of poor workers became street vendors.
-The profession requires little technical skills and has low barriers of entry, but it can be very profitable if for whatever reason there is a strong demand for particular product.
-The \textit{jogo do bicho} was one of those products that offered a high rate of return.
-The game was simple to operate and that simplicity attracted more people willing to try their luck.
-As the game gained a following in Brazil's First Republic, it comprised an important share of the extra-legal economy in Rio de Janeiro.
+The profession requires little technical skills and has low barriers of entry, but it can be profitable if vendors are able to quickly identify a growing demand for a particular product.
+The \textit{jogo do bicho} was one such case; not only the game had a surge in popularity in the 1890s, but it also offered a high rate of return.
+As the demand for lottery tickets grew, the \textit{jogo} comprised an important share of the Brazilian extra-legal economy.
 
@@
 The text can be read at \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm}} (in Portuguese).
-Access: September 2018.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
+Access: September 2018.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning up to BRL 20 billion (USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
 
-Since the mid-twentieth century, the \textit{jogo do bicho} has been illegal but pervasive in Brazil.
-For all purposes, the \textit{jogo} remains criminalised, yet there is virtually no Brazilian city which does not have its local \textit{bicheiros}.
-\citet{cross2006risk} suggest a distinction between informal and illegal markets that may be useful here.
+Since the mid-twentieth century, the \textit{jogo do bicho} has been outlawed but it continues to be ubiquitous in Brazil.
+There is virtually no Brazilian city which does not have its local group of \textit{bicheiros}.
+\citet{cross2006risk} suggest a distinction between informal and illegal markets that is useful to understand the current `semi-legal' status of the \textit{jogo}.
 The animal game started as an informal activity, in which the Baron of Drummond and his associates sold lottery tickets to the public.
-Although the state did not regulate the lottery market, the product itself was not illegal.
-After 1941, however, selling \textit{jogo do bicho} tickets became a criminal offence.
+Although the state did not regulate the lottery market, the product itself was not illegal, allegedly not even immoral.
+In fact, many organisations, including the Catholic Church, sold raffles and lottery tickets to sponsor their activities \citep[49]{torcato2011repressao}.
+Moreover, the state initially found no contradiction between the animal game and its own lottery, and both coexisted for about 15 years \citep[559]{chazkel2007beyond}.
+After 1941, however, selling \textit{jogo do bicho} tickets became a legal offence after a long campaign asking for its criminalisation.
 Since then, the \textit{jogo} has moved to the underground and consciously evaded state regulations.
-Moreover, the game has also exploited and subsidised large sectors of Brazil's formal economy, mainly in poor urban areas.
+This shift from informal to illegal has brought important changes to the whole dynamics of the \textit{jogo}.
+As an example, wealthy animal game bosses have colluded with shady sectors of the armed forces for protection, and this violence has also used against potential competitors.
+Moreover, this move to illegality enabled the game to exploit and subsidise large sectors of Brazil's formal economy, mainly in poor urban areas.
 I describe the animal game organisation structure and their impact in the Brazilian economic activity in further detail below.
@@
 
-If politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, their relationship with the animal game bankers have become more ambivalent in the last decades.
+Politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, but their relationship with the animal game bankers later became more ambivalent.
 The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citep{gaspari2002ditadura,jupiara2015poroes}.
-Given the absence of democratic checks and balances, paramilitaries and police forces colluded to repress potential dissidents of the regime and, frequently, to extort civilians \citep{magalhaes1997logica,misse2009acumulaccao}.
-\emph{Bicheiros} saw the corruption of some members of the military as an opportunity.
+The regime effectively dismantled the few checks and balances implemented in the Second Republic (1945--1964), and paramilitaries and police forces had considerable discretion to repress potential dissidents.
+Extortion of civilians was also widespread \citep{magalhaes1997logica,misse2009acumulaccao}.
+But \emph{bicheiros} saw the corruption of some members of the military as an opportunity to increase profits.
 Wealthy \emph{jogo do bicho} bankers hired rogue police officers not only to work as security guards but to threaten eventual competitors in their regions of influence.
@@
 While some sectors of the Brazilian bureaucracy may be described as `islands of excellence' and closely resemble the Weberian ideal of public administration \citep{bersch2017state}, local politicians remain dependent of unstable, and often unlawful, connections with social elites.
-This system of `partial embedded autonomy' provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour \cite{evans1995embedded}.
+This system of `partial embedded autonomy' provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour.
+In contrast with the  \cite{evans1995embedded}.
 This structure tends to perpetuate itself as it brings benefits for both parts, the animal game bankers and politicians.

### 2018-11-08 paper-bicho/response-reviewers.tex
add

@@
 
-Thank you very much for another valuable recommendation. The last three paragraphs of section 4.2 frame the state as an embedded entity as suggested by the reviewer.
+Thank you very much for another valuable recommendation.
+The last three paragraphs of section 4.2 frame the state as an embedded entity as suggested by the reviewer, and I believe the framing is also visible in other parts of the manuscript (although not so clearly as in section 4.2).
+However, I was not able to make
 

### 2018-11-08 paper-bicho/bicho.tex
add

@@
 The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citep{gaspari2002ditadura,jupiara2015poroes}.
-The regime effectively dismantled the few checks and balances implemented in the Second Republic (1945--1964), and paramilitaries and police forces had considerable discretion to repress potential dissidents.
+The regime effectively dismantled the few checks and balances implemented in the Second Republic (1945--1964), and paramilitaries and police forces had considerable discretion to repress political dissidents.
 Extortion of civilians was also widespread \citep{magalhaes1997logica,misse2009acumulaccao}.
@@
 The agreement between \emph{bicheiros} and corrupt members of the military was the ultimate responsible for the transformation of the \emph{jogo do bicho} into a `coercive oligopoly' \citep{jupiara2015poroes}.
+When the \textit{jogo} transitioned from an informal to an illegal market, the use of violence in the game became more widespread.
 The support of the armed forces meant that new groups would be prohibited from entering the market and that the illegal lottery could operate undisturbed by the government.
@@
 In the military regime, government officials were mainly interested in bribes from the animal game.
-But in the democratic period, votes became a sought-after political resource.
-\emph{Bicheiros} are important in this sense as they have direct influence over a number of poor communities.
+But in the democratic period, votes were the most sought-after political resource.
+\emph{Bicheiros} are important in this sense as they have direct influence over a number of poor communities, either because of their role as patrons, or as reliable sources of governance.
 Their patronage networks ensure that candidates supported by \emph{bicheiros} receive a substantial amount of votes from areas where campaigning is too difficult or too costly \citep[17]{misse2011crime}.
@@
 Individuals may be punished if the candidate does not receive the expected number of votes, and are often compelled to vote for politicians that have only loose connections with their communities.
-Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil’s lower classes.
+Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil's lower classes.
 
-The difficulty in permanently outlawing the \textit{jogo} showcases how the Brazilian state itself is deeply embedded within criminal sectors of the society.
+In a nutshell, the difficulty in permanently outlawing the \textit{jogo} showcases how the Brazilian state itself is deeply embedded within criminal sectors of the society.
 While some sectors of the Brazilian bureaucracy may be described as `islands of excellence' and closely resemble the Weberian ideal of public administration \citep{bersch2017state}, local politicians remain dependent of unstable, and often unlawful, connections with social elites.
-This system of `partial embedded autonomy' provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour.
-In contrast with the  \cite{evans1995embedded}.
-This structure tends to perpetuate itself as it brings benefits for both parts, the animal game bankers and politicians.
-The partial autonomy of state agents also highlight a social dilemma within the Brazilian public law.
-Although federal judges have successfully prosecuted \textit{bicheiros}, local state agents are not willing to comply with judicial decisions.
-The lottery elites are soon able to return to their positions in little time.
-As a result,
+This system of `partial embedded autonomy' \citep{evans1995embedded} provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour.
+This structure tends to perpetuate itself as it brings benefits for both the animal game bankers and local politicians, such as limited competition and the ability to extract resources from poor voters.
 

### 2018-11-08 paper-bicho/response-reviewers.tex
final update! :D

@@
 \usepackage{setspace}
-\usepackage[top=2.3cm,bottom=2.3cm,left=2.3cm,right=2.3cm]{geometry}
+\usepackage[top=2.15cm,bottom=2.15cm,left=2.15cm,right=2.15cm]{geometry}
 \usepackage[breaklinks=true]{hyperref}
@@
 I argue that the \textit{banqueiros} provide a mix of carrots and sticks to enforce compliance from bookmakers, and describe how the \textit{jogo} elite offers cash benefits and risk-sharing contracts as incentives to their employees.
-The parts on the funding of samba schools have been revised and are available in section 4.1.
+The parts on the funding of samba schools have been revised in section 4.1.
 
@@
 	\citet{jupiara2015poroes} argue that Ailton Guimarães Jorge, a former Army officer, tortured and murdered rival lottery bosses in the late 1970s.
-One of his former allies, Army Colonel Paulo Malha ̃es, told to the Rio de Janeiro State Truth Commission that Guimarães `went on a rampage' to consolidate his power \citep{belem2015guimaraes}.
+One of his former allies, Army Colonel Paulo Malhães, told to the Rio de Janeiro State Truth Commission that Guimarães `went on a rampage' to consolidate his power \citep{belem2015guimaraes}.
 Castor de Andrade, Rio's most influential animal game banker, also employed similar methods to run his business.
@@
 See for instance Centeno and Portes 2006.
-Also, informal and extra-legal are not always the same thing, as pointed out by Cross and Peña 2006.
+Also, informal and extra-legal are not always the same thing, as pointed out by Cross and Pena 2006.
 There is a difference in selling an illegal product illegally and selling a legal product illegally.
@@
 The last three paragraphs of section 4.2 frame the state as an embedded entity as suggested by the reviewer, and I believe the framing is also visible in other parts of the manuscript (although not so clearly as in section 4.2).
-However, I was not able to make
+However, I was not able to discuss the links between the \textit{jogo} and national development as Evans does in his famous text.
+Unfortunately, I could not find strong evidence connecting the \textit{jogo} to industrial policies or long-term economic growth.
+In that regard, I mention how the animal game provides opportunities for rent-seeking behaviour -- which might indeed have a negative impact on growth -- but I am a bit more cautious of pushing the argument further.
 

### 2018-11-08 paper-bicho/bicho.tex
final update! :D

@@
 \usepackage{setspace}
-\usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
+\usepackage[top=2.3cm,bottom=2.3cm,left=2.3cm, right=2.3cm]{geometry}
 \usepackage[backref,pagebackref,breaklinks=true]{hyperref}
@@
 But more than a act of civic disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
-A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world.\footnote{See \href{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil_n_8737212.html}{\texttt{https://www.huffpostbrasil.com/2015/12/07/legalizacao-jogos-brasil\_n\_8737212.html}} (in Portuguese).
-Access: August 2018.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce.
-See \href{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}{\texttt{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}}.
-Access: August 2018.}
+A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world \citep{huff2015jogos}.
+\citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}
 
@@
 In the early years of \textit{jogo do bicho}, lottery `bankers' were allowed to operate virtually free from police interference, what surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}.
-Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code, and in 1941 the animal game was banned.\footnote{See \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del3688.htm}} (in Portuguese).
-Access: September 2017.} Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.
-The text can be read at \href{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm}{\texttt{http://www.planalto.gov.br/ccivil\_03/decreto-lei/Del9215.htm}} (in Portuguese).
-Access: September 2018.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning up to BRL 20 billion (USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
+Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code, and in 1941 the animal game was banned.
+Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning up to BRL 20 billion (USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
 
@@
 For instance, the Brazilian official lottery, run by the federal government, are regularly audited by the \textit{Controladoria Geral da União} (Comptroller General of Brazil), the \textit{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
-The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute.\footnote{See \href{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm}{\texttt{https://bit.ly/2q3XnAV}} (in Portuguese).
-Access: December 2016.} A clandestine lottery, in contrast, cannot provide the same signals.
+The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute \citep{uol2016loteria}.
+A clandestine lottery, in contrast, cannot provide the same signals.
 Thus, the game providers need to assure gamblers that their business is honest although it is illegal.

### 2018-11-10 paper-bicho/bicho.tex
more updates

@@
 The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under both military regimes and democratic governments \citep{jupiara2015poroes}.
-But more than a act of civic disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
+But more than an act of civil disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
 A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world \citep{huff2015jogos}.
@@
 The \textit{jogo do bicho's} market formation closely resembles what \citet{fligstein1996markets} calls `markets as politics', in which firms create institutions to restrain the competition and organise the labour force.
-At the formation of the lottery market, competition is fierce and incumbents an challengers are akin to social movements.
+At the formation of the lottery market, competition is fierce and participants are akin to social movements.
 They are constantly trying to convince others of the viability of their ideas.
@@
 The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute \citep{uol2016loteria}.
+
 A clandestine lottery, in contrast, cannot provide the same signals.
@@
 Thus, street bookmakers have fewer incentives to cheat.
-In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}'s donations \citep{bbc2012aniz}.
+In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}' donations \citep{bbc2012aniz}.
 Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation.
@@
 Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}.
-In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}.
+In addition, Brazil uses an open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}.
 These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence.
@@
 
-Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}.
+Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population.
 Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
@@
 Violence is used to punish defectors and to constrain competitors.
-The clientelistic relationship that \textit{bicheiros} have with local politicians have lead to sub-optimal outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
+The clientelistic relationship that \textit{bicheiros} have with local politicians have lead to undemocratic outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
 These negative externalities have long-term effects and still impact the Brazilian public sphere.

### 2018-11-18 paper-bicho/response-reviewers.tex
proofreading done

@@
 In the last 5 paragraphs of section 2.2, I point out that not only the animal game has negatively affected the formal labour market, but it has also reduced the demand for social rights in Brazil.
-The \textit{jogo} provides jobs for those would could not easily enter the labour force, and as result ``[\dots] \textit{poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population}.'' I also mention how the \textit{jogo} has collaborated to increase inequality levels in the country.
-Discussions about how the animal game distorts the allocation of public funds and corrupts political representation are also included in the last two sections.
+The \textit{jogo} provides jobs for those would could not easily enter the labour force, and as result ``[\dots] \textit{poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population}.'' I then mention how the \textit{jogo} has collaborated to increase inequality levels in the country, mainly by transferring resources to rich animal game members and restricting access to newcomers.
+The game elites frequently employ violent methods to secure their rents against demands for redistribution from employees.
+Discussions about how the animal game distorts the allocation of public funds and corrupts political representation are also mentioned in section 2.2.
 
-I believe the text is not as neutralistic as before, and hopefully these changes have illustrated how pervasive are social tensions in the animal game.
+I believe that my analysis of the text is not as neutralistic as before, and hopefully these changes have illustrated how pervasive are social tensions in the \textit{jogo do bicho}.
 
@@
 The reviewer is correct to point out that I should have described those topics should have in further detail.
-I now dedicate the last paragraphs of section 2 to a discussion about the role of gambling rackets in Brazil's formal economy.
-The paragraphs address specifically the issue raised by the reviewer.
+I now provide a detailed discussion about the role of gambling rackets in Brazil's formal economy.
+Some of the paragraphs I have added to the revised manuscript address specifically the issue raised by the reviewer.
 In the case of the political candidates supported by the \textit{jogo}, I note in paragraph 3 of section 4.2 that `[\dots] \textit{Politicians from all spheres of government are involved with jogo do bicho bankers.
@@
 Cachoeira is a well-known \textit{jogo do bicho} boss in his home state of Goiás.
-As mentioned in the previous response, I discuss the overall impact of the \textit{jogo} in the formal economy in the last paragraphs of section 2.2.
+Additionally, as I mentioned in the previous response, I discuss the overall impact of the \textit{jogo} in the formal economy in the last paragraphs of section 2.2.
 
@@
 
-The article has been proofread and the text now uses American spelling as requested.
-With regards to the reference list, I fully agree that it was too extensive.
+The article has been proofread and the text now uses American spelling.
+With regards to the reference list, I fully agree that it was too long.
 The bibliography is now about 50\% shorter, and in my view the text remains as informative as before.
@@
 
-There are a number of similar excerpts in other parts of the manuscript. For instance, I also add more details about how the animal game profits from slot machines in Brazil:
+There are similar excerpts in other parts of the manuscript. For instance, I add some information about the commercialisation of slot machines by \textit{jogo do bicho} members:
 
@@
 
-I have mentioned in the introduction that the article uses secondary sources.
-Indeed this was not obvious to readers and I thank the reviewer for suggesting the clarification.
+Thank you very much for your comments.
+The introduction now states that the article uses secondary sources exclusively.
+Indeed this was not obvious to readers and the reviewer was correct to ask for further clarifications.
 The introduction now reads as follows:
@@
 This is an important point.
-The article deals with a number of topics that seemed only loosely associated.
-The section on repugnance, for instance, added little to the main ideas of the text and is not required for readers to understand the remaining of the article.
-That section has been removed from the manuscript and the subsections on costly signalling and reputation strategies now appear more prominently in the text.
+The article deals with a number of topics that at first seemed only loosely associated.
+The section on repugnance, for instance, added little to the main ideas of the text and is not required to understand the remaining of the article.
+That section has been removed and the subsections on costly signalling and reputation strategies are longer.
 The article is, in my view, more focused and concise as a result.
@@
 
-The referee rightly points out that I did not emphasise strongly enough the power structures of the animal game.
+The referee rightly points out that I did not emphasise the power structures of the animal game strongly enough.
 This criticism also appeared in other referee reports and I am happy to address it.
-As previously mentioned, section 2.2 has been largely rewritten and it now includes new paragraphs stressing the social conflicts within the \textit{jogo}.
-I have also changed the section title to `A Hierarchical Organisational Structure' to make this point clear.
+As previously mentioned, section 2.2 has been largely rewritten and it now includes new paragraphs on the social conflicts within the \textit{jogo}.
+I have also changed the section title to `A Hierarchical Organisational Structure' to make this point clearer.
 The indication of Neil Fligstein's work is also very appropriate and I sincerely thank the referee for it.
@@
 In contrast, I have added examples of how the \textit{jogo} bosses employ violence, or the threat thereof, against their own employees to enforce compliance to their decisions.
-,
-Nevertheless, I argue that violence can be used in rational terms, and I believe it is the case with the \textit{jogo}.
+
+Nevertheless, in the article I argue that violence can be used rationally, and I believe it is the case with the \textit{jogo}.
 However, I adopt a model of rationality which also assumes humans are embedded in a myriad of formal and informal institutions that design and shape their choices.
-It is unrealistic to assume otherwise.
-Apart from institutional constrains, individuals have limited access to information, what limits their abilities to correctly evaluate alternative scenarios.
-But in the \textit{jogo}, individuals generally behave in ways that maximise their own interest.
-This `thin rationality' conforms well to a wide range of the literature on crime and violence, such as \citet{balcells2010rivalry}, \citet{gambetta1996sicilian,gambetta2009codes}, \citet{kalyvas2006logic}, \citet{leeson2010pirational}, or \citet{skarbek2011governance,skarbek2012prison}.
+I concur with the reviewer that no individual is free of social constrains.
+It is indeed unrealistic to assume otherwise.
+Apart from institutional limits, individuals have limited access to information, what reduces their abilities to correctly evaluate alternative scenarios.
+But in the \textit{jogo}, individuals generally behave in ways that maximise their own interest, so I believe rationally is not fully absent from the dynamics of the game.
+This idea of `thin rationality' conforms well to a wide range of the literature on crime and violence, such as \citet{balcells2010rivalry}, \citet{gambetta1996sicilian,gambetta2009codes}, \citet{kalyvas2006logic}, \citet{leeson2010pirational}, or \citet{skarbek2011governance,skarbek2012prison}.
 
 I did not include a lengthier discussion about the rational choice sense of agency because I thought it would be slightly too abstract for my purposes.
-The article is mainly focused on the bankers' strategies in the long run, and I believe readers will be more interested in more concrete aspects of the animal game.
+The article is mainly focused on the bankers' strategies for long-term survival, and I believe readers will be more interested in more concrete aspects of the animal game.
 But I did add a footnote to the introduction describing the notion of rationality I employ in the text.
@@
 This also returns us to point 3, whereby clarifying extra-legal economy would help to understand extra-legal actors.}
-
 \vspace{.25cm}
@@
 I think this is an important criticism.
-Due to poor phrasing, the sentence highlighted by the referee does suggest a moralistic undertone.
+Due to my poor phrasing, the sentence highlighted by the referee does suggest a moralistic undertone.
 Yet I would like to affirm that it was not the case.
-My intention was not to use a moral judgements to describe the character or motivation of individuals, but only to draw attention to potential agent-principal dilemmas.
-This problem is widely discussed in the economics literature and it not ascribed only to informal or illegal activities, but pervasive in firms big or small and in government agencies \citep[e.g.,][]{grossman1983analysis,laffont2009theory,sappington1991incentives}.
-I sincerely apologise for the confusion.
+My intention was not to use a moral judgements to describe the character or motivation of individuals, but only to draw attention to potential agent-principal dilemmas in the game.
+This problem is widely discussed in the economics literature and it is not ascribed only to informal or illegal activities.
+Agency dilemmas are pervasive in firms big or small and in government agencies too \citep[e.g.,][]{grossman1983analysis,laffont2009theory,sappington1991incentives}.
+I sincerely apologise for the confusion my statement has caused.
 
 I have removed the mention to Arrow's work to avoid further misunderstandings.
-As I have mentioned previously, I believe the revised manuscript exposes more clearly the social tensions of the animal game, and this has improved the quality of the text.
+I believe the revised manuscript exposes more clearly the social tensions of the animal game, and this has improved the quality of the text.
 

### 2018-11-18 paper-bicho/bicho.tex
proofreading done

@@
 \noindent
-This article presents a rational choice account of Brazil's \textit{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world.
-Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and the game has played a major role in Brazil's cultural and political life.
+This article presents a rational choice account of Brazil's \textit{jogo do bicho} (``animal game''), possibly the largest illegal lottery game in the world.
+In over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and the game has played a major role in Brazil's cultural and political life.
 My analysis of the \textit{jogo do bicho} has three goals.
-First, I examine the mechanisms that fostered the lottery's notable growth outside of the boundaries of Brazilian law.
-Second, I investigate how the animal game financiers combine costly signals and selective incentives to induce cooperation from members of community.
+First, I examine the mechanisms that fostered the lottery's notable growth outside the boundaries of Brazilian law.
+Second, I investigate how the animal game financiers combine costly signals and selective incentives to induce cooperation from members of the community.
 Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of Brazil's political system to advance their long-term interests.
@@
 In 1892, Baron João Batista de Viana Drummond came up with a new idea to fund his cash-strapped zoo.
-Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \textit{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city.
-But it lacked visitors.
+Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \textit{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city, but it lacked visitors.
 An experienced businessman, Drummond realised the zoo would have to provide other kinds of entertainment to keep itself afloat.
@@
 The rules were straightforward.
-In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
+In the morning, the baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
 Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, so the game consisted in a simple raffle.
@@
 This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.} The lucky winner would take home a prize worth 20 times the ticket price, an amount higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.
-The Baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
+The baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
 Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in many stores across Rio de Janeiro.
@@
 A network of street bookmakers, called \textit{bicheiros}, made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the numbers game \citep[37]{chazkel2011laws}.
-The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today $[1985]$ in Rio de Janeiro, the game is everything.
+The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: ``Today $[1985]$ in Rio de Janeiro, the game is everything.
 $[\dots]$  Nobody works!
-Everybody plays' \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.}
+Everybody plays''  \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.}
 
 But this tolerant state of affairs did not last.
-Civil servants and police officers criminalised the \textit{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}.
+Civil servants and police officers criminalised the \textit{jogo do bicho} on the grounds of public safety, and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}.
 The campaign extended for several decades and received considerable support from the \textit{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}.
-The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and the lottery remains illegal until this day.
+The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and the lottery remains illegal to this day.
 
@@
 But more than an act of civil disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
-A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world \citep{huff2015jogos}.
-\citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}
+A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world \citep{huff2015jogos}.
+\citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in Rio de Janeiro, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was ``the largest single industry in Latin America'' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}
 
@@
 In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information at the moment of their choices.
-Thus, my analysis employs a `thin' notion of rationality and a `thick' description of social institutions, where individual action can only be understood with reference to the social environment \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, commonly called \textit{bicheiros},\footnote{As I describe in the next section, the \textit{jogo do bicho} structure can be broadly divided into three occupations: the rich game financiers (\textit{banqueiros}), mid-level managers (\textit{gerentes}), and street bookmakers (\textit{bicheiros}).
+Thus, my analysis employs a ``thin'' notion of rationality and a `thick' description of social institutions, where individual action can only be understood with reference to the social environment \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, commonly called \textit{bicheiros},\footnote{As I describe in the next section, the \textit{jogo do bicho} structure can be broadly divided into three occupations: the rich game financiers (\textit{banqueiros}), mid-level managers (\textit{gerentes}), and street bookmakers (\textit{bicheiros}).
 While there are important hierarchical differences among these groups, all members of the \textit{jogo} are collectively called \textit{bicheiros} in Brazil.
-I follow the same practice here, and only refer to their specific role within the animal game structure when necessary for clarification.} have developed unique strategies to solve collective action problems and maximise their political strategies.
-This does not mean that their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals.
+I follow the same practice here, and only refer to their specific role within the animal game structure when necessary for clarification.} have developed unique strategies to solve collective action problems and to maximise their political strategies.
+This does not mean their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals.
 Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive, they do address the long-term needs of the \textit{jogo do bicho} financiers.
@@
 In particular, \textit{bicheiros} face two main challenges to keep the lottery running.
-First, they need to gather public support so that gamblers are not discouraged to engage in the lottery despite it being illegal.
-Second, \textit{bicheiros} also have to ensure that the state repression is not prohibitively costly to their business, otherwise they would be better off by shutting it down.
+First, they need to gather public support so gamblers are not discouraged to engage in the lottery despite it being illegal.
+Second, \textit{bicheiros} also have to ensure the state repression is not prohibitively costly to their business; otherwise they would be better off shutting it down.
 I argue below that the \textit{bicheiros} have succeeded in both by using carefully-designed reputation strategies and employing costly signals to the communities they serve.
@@
 I use the case of Rio de Janeiro to illustrate how the \textit{jogo do bicho} has overcome the obstacles to its expansion.
-Rio is a particularly interesting case because in no other part of Brazil the game financiers established such an effective patronage network.
+Rio is a particularly interesting case because in no other part of Brazil is the game the financiers established such an effective patronage network.
 \textit{Bicheiros} have sponsored political campaigns, financed cultural activities and football teams, and sometimes even run in local elections themselves.
@@
 \citet{leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations.
-\citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking and coordinating productive activities in California prison gangs.
-I argue that \textit{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and foster trust in the community.
+\citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking behaviour and in coordinating productive activities in California prison gangs.
+I argue that \textit{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and to foster trust in the community.
 
 Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,spence1973job}.
-I provide evidence that \textit{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and reduce the uncertainty associated with clandestine markets.
+I provide evidence that \textit{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and to reduce the uncertainty associated with clandestine markets.
 Their main tool to increase credibility was costly signalling \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}.
@@
 More specifically, I use the Brazilian case to illustrate how politicians and civil servants are co-opted by criminal groups, and how this collusion distorts the electoral process and benefits wealthy members of the \textit{jogo do bicho} network.
-\citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities.
-\citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested that the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s.
+\citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed this influence gave them leverage over political authorities.
+\citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s.
 In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \textit{jogo do bicho} and the military regime in Brazil (1964--1985).
@@
 
-The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}.
-They are: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
+The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
 I discuss each of these elements below.
@@
 The profession requires little technical skills and has low barriers of entry, but it can be profitable if vendors are able to quickly identify a growing demand for a particular product.
-The \textit{jogo do bicho} was one such case; not only the game had a surge in popularity in the 1890s, but it also offered a high rate of return.
+The \textit{jogo do bicho} was one such case; the game had a surge in popularity in the 1890s, and it also offered a high rate of return.
 As the demand for lottery tickets grew, the \textit{jogo} comprised an important share of the Brazilian extra-legal economy.
@@
 Immigration also influenced the \textit{jogo do bicho} via social ties.
-Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}.
-Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}.
+Most foreigners who moved to Brazil came from countries such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}.
+Family and neighbourhood networks created incentives for immigrants to establish trade relations and to enforce cooperation through community responsibility systems \citep{roth2014prison}.
 Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil} and in the \textit{jogo do bicho} in particular \citep{magalhaes2005ganhou, villar2008contravencao}.
@@
 The abolition of slavery and the growing industrialisation of Brazil increased the amount of capital available in the country \citep{franco1987reformas, schulz2008financial}.
-Moreover, the 1888 Banking Act gave extra liquidity to local financial markets, what made credit more widely available in cities like São Paulo and Rio de Janeiro.
+Moreover, the 1888 Banking Act gave extra liquidity to local financial markets, which made credit more widely available in cities like São Paulo and Rio de Janeiro.
 Individuals received a temporary boost in personal income, a part of which they spent on leisure activities such as the \textit{jogo do bicho}.
@@
 \citet[69--100]{chazkel2011laws} notes that until the 1940s, police district chiefs operated within a large margin of discretion, so official repression against bookmakers was notably inconsistent.
-In the early years of \textit{jogo do bicho}, lottery `bankers' were allowed to operate virtually free from police interference, what surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}.
+In the early years of \textit{jogo do bicho}, lottery ``bankers'' were allowed to operate virtually free from police interference, which surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}.
 Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code, and in 1941 the animal game was banned.
-Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.} Recent estimations show that the prohibition of the \textit{jogo do bicho} have prevented the state from earning up to BRL 20 billion (USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
+Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.} Recent estimations show that the prohibition of the \textit{jogo do bicho} prevented the state from earning up to BRL 20 billion (USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
 
 Since the mid-twentieth century, the \textit{jogo do bicho} has been outlawed but it continues to be ubiquitous in Brazil.
-There is virtually no Brazilian city which does not have its local group of \textit{bicheiros}.
-\citet{cross2006risk} suggest a distinction between informal and illegal markets that is useful to understand the current `semi-legal' status of the \textit{jogo}.
+There is virtually no Brazilian city that does not have its local group of \textit{bicheiros}.
+\citet{cross2006risk} suggest a distinction between informal and illegal markets that is useful to understand the current ``semi-legal'' status of the \textit{jogo}.
 The animal game started as an informal activity, in which the Baron of Drummond and his associates sold lottery tickets to the public.
-Although the state did not regulate the lottery market, the product itself was not illegal, allegedly not even immoral.
+Although the state did not regulate the lottery market, the product itself was not illegal or allegedly immoral.
 In fact, many organisations, including the Catholic Church, sold raffles and lottery tickets to sponsor their activities \citep[49]{torcato2011repressao}.
@@
 After 1941, however, selling \textit{jogo do bicho} tickets became a legal offence after a long campaign asking for its criminalisation.
-Since then, the \textit{jogo} has moved to the underground and consciously evaded state regulations.
+Since then, the \textit{jogo} has moved underground and consciously evaded state regulations.
 This shift from informal to illegal has brought important changes to the whole dynamics of the \textit{jogo}.
-As an example, wealthy animal game bosses have colluded with shady sectors of the armed forces for protection, and this violence has also used against potential competitors.
+As an example, wealthy animal game bosses have colluded with shady sectors of the armed forces for protection, and this violence has also been used against potential competitors.
 Moreover, this move to illegality enabled the game to exploit and subsidise large sectors of Brazil's formal economy, mainly in poor urban areas.
-I describe the animal game organisation structure and their impact in the Brazilian economic activity in further detail below.
+I describe the animal game organisation structure and its impact in the Brazilian economic activity in further detail below.
 
@@
 At the bottom level are the \textit{bicheiros}, those in charge of selling \textit{jogo do bicho} tickets \citep{chazkel2007beyond, da1999aguias}.
-\textit{Bicheiros} are the most visible part of the \textit{jogo do bicho}, and the name loosely describes all those involved in the lottery organisation.
-Yet their meaning in the \textit{jogo do bicho} structure is more particular and refers to street-level ticker sellers.
+\textit{Bicheiros} are the most visible part of the \textit{jogo do bicho}.
+The name loosely describes all those involved in the lottery organisation, yet their meaning in the \textit{jogo do bicho} structure is more particular and refers to street-level ticker sellers.
 The \textit{bicheiros}  usually build their vending stands inside the premises of a local shop, such as a small grocery store, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}.
-The street bookmakers usually work alone, but may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
+The street bookmakers usually work alone but may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
 
@@
 
-The \textit{banqueiros}, or the Portuguese for bankers, occupy the top position in the \textit{jogo do bicho} hierarchy.
+The \textit{banqueiros} (Portuguese for bankers) occupy the top position in the \textit{jogo do bicho} hierarchy.
 They comprise the small financial elite of the game.
@@
 
-The lottery bosses run their businesses from fortified houses in unknown locations, called the \textit{fortalezas} (`forts').
+The lottery bosses run their businesses from fortified houses in unknown locations called \textit{fortalezas} (``forts'').
 The first \textit{fortalezas} likely appeared in the 1950s, when the animal game was already well-established across the Brazilian territory.
 The period coincides with a time when the \textit{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}.
-Due to the growing scope of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and make coordination easier.
+Due to the growing scope of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and to make coordination easier.
 
@@
 Thus, the game bankers have built an extensive network of gunmen and bribed police officers to protect their employees from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
-`Zé' (Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \textit{jogo do bicho} informal security personnel:
+``Zé'' (Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \textit{jogo do bicho} informal security personnel:
 
 \begin{quote}
- [\dots] bums are scared and they don't mess around with us; they think there's an officer nearby or something like that.
+Bums are scared and they don't mess around with us; they think there's an officer nearby or something like that.
 Look at all this money here!
-[shows the interviewer a handful of cash] It's not ours [referring to street-corner bookmakers].
+[Shows the interviewer a handful of cash.] It's not ours [referring to street-corner bookmakers].
 And if it's not ours, it's someone else's.
@@
 
-But the \textit{banqueiros} do not use violence only against other criminals.
+The \textit{banqueiros} do not use violence only against other criminals.
 They often employ violent methods against competitors and their own staff, too.
 \citet{jupiara2015poroes} argue that Ailton Guimarães Jorge, a former Army officer, tortured and murdered rival lottery bosses in the late 1970s.
-One of his former allies, Army Colonel Paulo Malhães, told to the Rio de Janeiro State Truth Commission that Guimarães `went on a rampage' to consolidate his power \citep{belem2015guimaraes}.
+One of his former allies, Army Colonel Paulo Malhães, told the Rio de Janeiro State Truth Commission that Guimarães ``went on a rampage'' to consolidate his power \citep{belem2015guimaraes}.
 Castor de Andrade, Rio's most influential animal game banker, also employed similar methods to run his business.
 Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors.
-In a famous case, Andrade shot Euclides Ponar, an old \textit{jogo do bicho} boss known as `Grey-Headed Chinese' (\textit{China Cabe\c{c}a Branca}) after Ponar denounced a fraud in lottery draws in 1976.
+In a famous case, Andrade shot Euclides Ponar, an old \textit{jogo do bicho} boss known as the ``Grey-Headed Chinese'' (\textit{China Cabe\c{c}a Branca}) after Ponar denounced a fraud in lottery draws in 1976.
 Andrade was likely involved in other assassination plots in the 1990s  \citep{globo2017castor}.
@@
 This is a good strategy for the \textit{banqueiros}.
-The fact that they had committed violent crimes in the past reduces the need to commit them in the present, as and a result the bosses can spend less money on security and increase profits.
+The fact that they had committed violent crimes in the past reduces the need to commit them in the present, and as a result the bosses can spend less money on security and increase profits.
 As it happens in many traditional markets, if a group is able to form a cartel, they can increase the price of their services without fearing immediate competition.
@@
 
-The most important financial mechanism implemented by bankers to help \textit{bicheiros} is the \textit{descarga}, loosely translated as `the unloading'.
+The most important financial mechanism implemented by bankers to help \textit{bicheiros} is the \textit{descarga}, loosely translated as ``the unloading''.
 The \textit{descarga} is the \textit{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}}.
 Booking agents are sometimes unable to honour expensive bets.
-The top prize in the animal game pays up to 4,000 times the amount invested, thus \textit{bicheiros} may have to raise thousands of Brazilian Reals in a single day to pay the lucky winners.
-To prevent the \textit{quebra da banca} (`bust of the bank'), \textit{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
+The top prize in the animal game pays up to 4,000 times the amount invested; thus, \textit{bicheiros} may have to raise thousands of Brazilian Reals in a single day to pay the lucky winners.
+To prevent the \textit{quebra da banca} (``bust of the bank''), \textit{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
 The \textit{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \textit{jogo do bicho}.
@@
 Over time, there is a transfer of income from the bottom to the top of the animal game structure due to the fees.
-This accumulation of capital is one of the reasons why in the 1990s bankers started offering other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
+This accumulation of capital is one of the reasons in the 1990s bankers started offering other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
 They simply had more capital to invest, and this eventually helped to compensate the downturn in the \textit{jogo do bicho} markets in the last years \citep{globo2017castor}.
@@
 This has brought broad consequences to the Brazilian illegal economy.
-As the number of slot machines increased in Brazil, the opportunity for criminals to use them for money laundering are higher as well.
+As the number of slot machines increased in Brazil, the opportunity for criminals to use them for money laundering increased as well.
 The scheme is easy to carry out.
@@
 Then he can legally use that money for any purpose without raising suspicion from the authorities.
-The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called `the slot machine mafia'.
+The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called ``slot machine mafia''.
 Relatives of Castor de Andrade were also involved in the scheme \citep{estado2011cacaniquel}.
-In 2012, President Dilma Rousseff sanctioned a law that considered slot machines and the \textit{jogo do bicho} as money laundering \citep{agenciabrasil2012dilma}.
+In 2012, President Dilma Rousseff sanctioned a law that considered slot machines and the \textit{jogo do bicho} as money laundering crimes \citep{agenciabrasil2012dilma}.
 
@@
 For instance, the \textit{jogo do bicho} gives a boost to the Brazilian economy by providing jobs for unskilled workers who cannot easily join the labour market.
-By doing so, the \textit{jogo} prevents some of poorest members of the Brazilian society from demanding more inclusive government policies, although they are the ones who would benefit the most from public assistance.
-Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population.
+By doing so, the \textit{jogo} prevents some of the poorest members of the Brazilian society from demanding more inclusive government policies, although they are the ones who would benefit the most from public assistance.
+Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds on other, probably wealthier, sectors of the population.
 
@@
 The threat of violence reduces the space for collective bargaining with the lottery bosses, and higher profits at the top of the \textit{jogo do bicho} structure do not trickle down to those at the bottom.
-This is one of the reasons why the game bankers turned the game into an oligopoly: the use of violence guarantees not only that new entrants will not be allowed to join the market, but also that profits are concentrated in the hands of very few individuals.
+This is one of the reasons why the game bankers turned the game into an oligopoly: the use of violence guarantees new entrants will not be allowed to join the market and profits will be concentrated in the hands of very few individuals.
 
-The \textit{jogo do bicho's} market formation closely resembles what \citet{fligstein1996markets} calls `markets as politics', in which firms create institutions to restrain the competition and organise the labour force.
+The \textit{jogo do bicho's} market formation closely resembles what \citet{fligstein1996markets} calls ``markets as politics'', in which firms create institutions to restrain the competition and organise the labour force.
 At the formation of the lottery market, competition is fierce and participants are akin to social movements.
@@
 It is puzzling because \textit{bicheiros} do not use violence to induce individuals to play the lottery, nor have they ever clashed with the Brazilian government.
-Precisely because violence could shun off profits, \textit{bicheiros} devised other mechanisms to create a friendly environment for the illegal lottery.
-I investigate two of them, costly signalling and reputation building.
+Precisely because violence could shun profits, \textit{bicheiros} devised other mechanisms to create a friendly environment for the illegal lottery.
+I investigate two of them: costly signalling and reputation building.
 
@@
 Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical case studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible even in difficult situations.
-The main requirement for sustained cooperation is that players believe that future pay-offs will be higher than present ones.
+The main requirement for sustained cooperation is that players believe future pay-offs will be higher than present ones.
 If that condition is true, fear of retaliation will induce individuals not to cheat.
@@
 In theory, the same should also apply to illegal organisations.
-Yet in practice we see that criminal groups are generally short-term oriented, that is, they tend to discount the future more heavily that most people.
-This makes cooperative behaviour among criminal rather uncommon, and there is substantial evidence suggesting that illegal groups face serious collective action problems \citep[e.g.,][]{gambetta2009codes, leeson2010pirational, skarbek2011governance, skarbek2012prison, varese2001russian}.
+Yet in practice we see that criminal groups are generally short-term oriented, that is, they tend to discount the future more heavily than most people.
+This makes cooperative behaviour among criminal rather uncommon, and there is substantial evidence suggesting illegal groups face serious collective action problems \citep[e.g.,][]{gambetta2009codes, leeson2010pirational, skarbek2011governance, skarbek2012prison, varese2001russian}.
 
 The \textit{jogo do bicho} is an exception to this rule.
-The game has been running for more than a century without considerable interruption, what attests that \textit{bicheiros} have managed to solve collective action issues in one way or another.
+The game has been running for more than a century without considerable interruption, attesting to the fact that \textit{bicheiros} have managed to solve collective action issues in one way or another.
 Importantly, the \textit{jogo} involves moderately low levels of violence -- at least when compared to other illegal activities such as drug trafficking.
 Moreover, Brazilians widely consider the \textit{jogo do bicho} an honest lottery, and reports of cheating are surprisingly uncommon.
-The popular motto associated with the game testifies in its favour: `\textit{vale o escrito; ganhou, leva}', or `what is written is what counts; if you win it, you take it' \citep{magalhaes2005ganhou}.
-Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community, costly signalling and reputation strategies.
+The popular motto associated with the game testifies in its favour: ``\textit{Vale o escrito; ganhou, leva}'', or ``What is written is what counts; if you win it, you take it'' \citep{magalhaes2005ganhou}.
+Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community: costly signalling and reputation strategies.
 
@@
 
-Legal lotteries employ many techniques to show that this is the case.
+Legal lotteries employ many techniques to show this is the case.
 For instance, the Brazilian official lottery, run by the federal government, are regularly audited by the \textit{Controladoria Geral da União} (Comptroller General of Brazil), the \textit{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
@@
 The fixed-multiplier formula alleviates such problems of adverse selection \citep{akerlof1970market, cohen2010testing, levin2001information}.
-As players and vendors known the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \textit{bicheiros} to a contract that can be easily enforced.
+As players and vendors know the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \textit{bicheiros} to a contract that can be easily enforced.
 This technique offers buyers a simple yet effective screening strategy that induces \textit{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.
@@
 The media broadcasts the draws on radio and TV, so any interested player can verify the selected numbers.
-The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public; hence, \textit{bicheiros} can free ride on the lottery's long-standing reputation of credibility.
+The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public.
+Hence, \textit{bicheiros} can free ride on the lottery's long-standing reputation of credibility.
 
@@
 This ability to elicit cooperation from external actors caused the \textit{jogo do bicho} to last longer than collection action theories would anticipate.
-In that regard, the bankers' investment in costly signalling and reputation has largely paid off: the lottery's continuous profits are a proof of their success.
+In that regard, the bankers' investments in costly signalling and reputation has largely paid off: the lottery's continuous profits are a proof of their success.
 With that continuous flow of income, \textit{banqueiros} were able to extend their influence well beyond the poor communities in which they operate.
@@
 The impact of the \emph{jogo do bicho} is not restricted to the Brazilian economy.
-Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural festivity, the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,queiroz1992carnaval}.
-The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that `without the \emph{jogo do bicho} the Carnival would have ended' \citep{odia2016aniz}.
+Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural festivity: the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,queiroz1992carnaval}.
+The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that ``without the \emph{jogo do bicho} the Carnival would have ended'' \citep{odia2016aniz}.
 Based on that support, \emph{bicheiros} have established an extensive patronage network over samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}.
-Such clientelistic network brings large material benefits to their members, yet it has created perverse incentives for government officials and causes several distortions to the Brazilian democratic system.
+Such client network brings large material benefits to their members, yet it has created perverse incentives for government officials and has caused several distortions to the Brazilian democratic system.
 
@@
 
-\subsection{The `Medici of Samba': \emph{Bicheiros} as Patrons of Carnival}
+\subsection{The Medici of Samba: \emph{Bicheiros} as Patrons of Carnival}
 \label{sub:patrons}
@@
 In 1930, opposition leader Getúlio Vargas led a bloodless coup d'état that brought Brazil's First Republic to an end.
-During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by dismantling effectively federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,fausto1972revoluccao,skidmore1967politics}.
+During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by effectively dismantling federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,fausto1972revoluccao,skidmore1967politics}.
 In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,williams2001culture}.
-Popular music, in turn, occupied an important place in Vargas's project of `brazilianing Brazil'.
+Popular music, in turn, occupied an important place in Vargas's project of ``Brazilianing Brazil''.
 Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{mccann2004hello,stockler2011samba}.
@@
 In a period when civil and political rights were limited, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}.
-Patriotic sambas exalted the country's natural beauties and the figure of the `friendly, happy, cordial and industrious' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
+Patriotic sambas exalted the country's natural beauties and the figure of the ``friendly, happy, cordial and industrious'' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
 The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of a male donkey and a female horse.
@@
 
-However, the samba groups were not passive members in this process.
-Since the 1960s, the Rio Carnival expanded in scope and, stimulated by growing numbers of spectators, the parades became more elaborate \citetext{\citealp{cabral2016escolas}; \citealp[214]{chinelli1993vazio}; \citealp[240]{hertzman2013making}}.
-Unable to cope with the rising costs of the show, the `samba schools', which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}.
+The samba groups were not passive members in this process.
+Since the 1960s, the Rio Carnival has expanded in scope.
+Stimulated by growing numbers of spectators, the parades have become more elaborate \citetext{\citealp{cabral2016escolas}; \citealp[214]{chinelli1993vazio}; \citealp[240]{hertzman2013making}}.
+Unable to cope with the rising costs of the show, the samba schools, which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}.
 This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,misse2011crime,queiroz1992carnaval}.
@@
 
-The funding of the samba schools had an indirect effect to the animal game.
+The funding of the samba schools had an indirect effect on the animal game.
 The patronage also reduced agent-principal problems within the \emph{jogo do bicho}.
 \emph{Bicheiros} donate to samba school to gather support of the communities, and by doing so they gain access to local information on their business.
-Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring is cost-effective for animal game managers.
+Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring the cost-effectiveness for animal game managers.
 Thus, street bookmakers have fewer incentives to cheat.
 In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}' donations \citep{bbc2012aniz}.
-Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation.
-The patronage can be interpreted as an illegal version of `profit-sharing', a mechanism which has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.
+Hence, funds donated to samba schools and other charitable organisations help align the interests of different members of the \emph{jogo do bicho} organisation.
+The patronage can be interpreted as an illegal version of ``profit-sharing,'' a mechanism that has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.
 
-The samba schools have profited from this association too.
+Samba schools have profited from this association too.
 First, they have gained autonomy from the government.
-The samba schools do not need to rely exclusively on public funds to organise the parade, and money from the \emph{jogo do bicho} permitted the schools to act independently \citep[209]{chinelli1993vazio}.
+Samba schools do not need to rely exclusively on public funds to organise the parade, and money from the \emph{jogo do bicho} has permitted the schools to act independently \citep[209]{chinelli1993vazio}.
 Second, the support of the \emph{jogo do bicho} has increased the political and social clout of the samba schools.
@@
 Although the \emph{banqueiros} are interested in weak law enforcement against the animal game, politicians have resorted to samba schools to contact \emph{bicheiros} and use their financial and electoral influence in the shanty towns \citep[17]{misse2011crime}.
-The samba schools, therefore, have increased their bargaining power in the political sphere and extended their reach within Rio's poor communities \citep[215]{chinelli1993vazio}.
+The samba schools, therefore, have increased their bargaining power in the political sphere and have extended their reach within Rio's poor communities \citep[215]{chinelli1993vazio}.
 
@@
 But \emph{bicheiros} saw the corruption of some members of the military as an opportunity to increase profits.
-Wealthy \emph{jogo do bicho} bankers hired rogue police officers not only to work as security guards but to threaten eventual competitors in their regions of influence.
-The agreement between \emph{bicheiros} and corrupt members of the military was the ultimate responsible for the transformation of the \emph{jogo do bicho} into a `coercive oligopoly' \citep{jupiara2015poroes}.
+Wealthy \emph{jogo do bicho} bankers hired rogue police officers to work as security guards and to threaten eventual competitors in their regions of influence.
+The agreement between \emph{bicheiros} and corrupt members of the military was ultimately responsible for the transformation of the \emph{jogo do bicho} into a coercive oligopoly \citep{jupiara2015poroes}.
 When the \textit{jogo} transitioned from an informal to an illegal market, the use of violence in the game became more widespread.
-The support of the armed forces meant that new groups would be prohibited from entering the market and that the illegal lottery could operate undisturbed by the government.
+The support of the armed forces meant new groups would be prohibited from entering the market and the illegal lottery could operate undisturbed by the government.
 
@@
 Carlinhos Cachoeira, a famous animal game banker from the state of Goiás established a large patronage network that included mayors, deputies, senators, judges, and businessmen, many of them linked to the federal government.
-His brother Marcos ironically noted that Carlinhos was `too dedicated to politics' and that he illegally donated about USD 300 million to political candidates in his home state \citep{estadao2012gene}.
+His brother Marcos ironically noted that Carlinhos was ``too dedicated to politics'' and he illegally donated about USD 300 million to political candidates in his home state \citep{estadao2012gene}.
 
-The Brazilian political system is particularly conductive to clientelistic practices.
+The Brazilian political system is particularly conductive to client practices.
 Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}.
-In addition, Brazil uses an open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}.
+In addition, Brazil uses an open-list proportional representation electoral system; each of the 27 states of the federation are considered at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}.
 These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence.
@@
 As candidates do not know whether their competitors receive funding from the \emph{jogo do bicho} nor the amount each one was paid, their dominant position is to contact the \emph{bicheiros} and join their networks.
-The situation is a prisoner's dilemma in which candidates would be better off by running cheaper campaigns and not being dependent of \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching a cheaper solution.
+The situation is a prisoner's dilemma in which candidates would be better off running cheaper campaigns and not being dependent on \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching a cheaper solution.
 
@@
 Consequently, votes have high marginal utility for politicians.
-As elections may be decided by a small difference, the \emph{bicheiros}' clientelistic ties guarantee a minimum number of votes that politicians can rely upon on election day.
+As elections may be decided by a small difference, the \emph{bicheiros}' client ties guarantee a minimum number of votes that politicians can rely upon on election day.
 Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process \textit{per se}.
-Individuals may be punished if the candidate does not receive the expected number of votes, and are often compelled to vote for politicians that have only loose connections with their communities.
+Individuals may be punished if the candidate does not receive the expected number of votes, and they are often compelled to vote for politicians who have only loose connections with their communities.
 Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil's lower classes.
@@
 In a nutshell, the difficulty in permanently outlawing the \textit{jogo} showcases how the Brazilian state itself is deeply embedded within criminal sectors of the society.
-While some sectors of the Brazilian bureaucracy may be described as `islands of excellence' and closely resemble the Weberian ideal of public administration \citep{bersch2017state}, local politicians remain dependent of unstable, and often unlawful, connections with social elites.
-This system of `partial embedded autonomy' \citep{evans1995embedded} provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour.
+While some sectors of the Brazilian bureaucracy may be described as ``islands of excellence'' and closely resemble the Weberian ideal of public administration \citep{bersch2017state}, local politicians remain dependent of unstable, and often unlawful, connections with social elites.
+This system of ``partial embedded autonomy'' \citep{evans1995embedded} provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour.
 This structure tends to perpetuate itself as it brings benefits for both the animal game bankers and local politicians, such as limited competition and the ability to extract resources from poor voters.
@@
 
-Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population.
+Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and to establish close ties with the population.
 Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
 \textit{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers.
-Furthermore, by investing in the Carnival parade \textit{bicheiros} have been able to gather popular and government support.
-Poor communities have associated with the \textit{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.
+Furthermore, by investing in the Carnival parade, \textit{bicheiros} have been able to gather popular and government support.
+Poor communities have associated with the \textit{bicheiros} to receive welfare provisions, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.
 
@@
 Violence is used to punish defectors and to constrain competitors.
-The clientelistic relationship that \textit{bicheiros} have with local politicians have lead to undemocratic outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
+The client relationship \textit{bicheiros} have with local politicians has lead to undemocratic outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
 These negative externalities have long-term effects and still impact the Brazilian public sphere.
 
-Although the \textit{jogo do bicho} has received an increasing attention from scholars, much of its inner workings remain poorly understood.
+Although the \textit{jogo do bicho} has received increasing attention from scholars, much of its inner workings remain poorly understood.
 First, the relationship between \textit{bicheiros} and drug dealers is a topic that deserves attention.
@@
 The question whether \textit{bicheiros} collaborated or opposed the emergent drug dealing business is still unclear.
-Second, the extent to which \textit{bicheiros} use other businesses, such as hotels or factories, to laundry money has been mentioned by members of the Brazilian judiciary \citep{globo2012bicheiro,globo2015cacaniquel}; however, there is no reliable estimate on its size.
+Second, the extent to which \textit{bicheiros} use other businesses, such as hotels or factories, to launder money has been mentioned by members of the Brazilian judiciary \citep{globo2012bicheiro,globo2015cacaniquel}; however, there is no reliable estimate on its size.
 Lastly, more research is required to clarify how \textit{bicheiros} from different parts of Brazil coordinate their activities and prevent large-scale conflicts.

### 2018-11-18 chapters/03-bicho.tex
update
add bicho

--- 2017-04-09 03-bicho.tex
@@
 \section{Introduction}
-\label{sec:intro}
+\label{sec:intro3}
 
 In 1892, Baron João Batista de Viana Drummond came up with a new idea to fund his cash-strapped zoo.
-Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \emph{Jardim Zoológico,} or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city.
-But it lacked visitors.
-As an experienced businessman, Drummond soon realised the zoo would have to provide other kinds of entertainment to keep itself afloat.
-Among his suggestions, one seemed particularly promising: a lottery raffle.
+Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \textit{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city, but it lacked visitors.
+An experienced businessman, Drummond realised the zoo would have to provide other kinds of entertainment to keep itself afloat.
+One of his plans seemed particularly promising: a lottery raffle.
 
 The rules were straightforward.
-In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
-Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, making the game similar to a common raffle.
-But it did not take long until visitors could name their animals of choice.
-This change made the game considerably more appealing and lasts until this day \citep[71--74]{da1999aguias}.} At five in the afternoon, Drummond opened the box, showed the picture to the public, and paid to every winner a cash prize worth 20 times the zoo's admission fee.\footnote{The amount was higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.} The lottery was labelled as the \emph{jogo do bicho}, or the animal game, and it was immediately adopted by the public.
-Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but in stores across Rio de Janeiro.
-He rightly predicted that this small change would increase profits, but there was one thing the Baron did not foresee.
-He unleashed a new gambling market.
+In the morning, the baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance.
+Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, so the game consisted in a simple raffle.
+However, the zoo soon allowed participants to choose the animals they preferred.
+This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.} The lucky winner would take home a prize worth 20 times the ticket price, an amount higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}.
+The baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public.
+Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in many stores across Rio de Janeiro.
+What was once a small raffle soon became a large gambling market of its own.
 
-The \emph{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace \citep{magalhaes2005ganhou, soares1993jogo}.
-A network of street bookmakers, called \emph{bicheiros}, expanded spontaneously the original game in innovative ways.
-Evading state regulations, \emph{bicheiros} made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the clandestine numbers game  \citep[37]{chazkel2011laws}.
-The \emph{jogo do bicho} became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today {[}1895{]}, in Rio de Janeiro, the game is everything.
-{[}\ldots{}{]} Nobody works!
-Everybody plays' \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.} But this tolerant state of affairs did not last.
-Civil servants and police officers criminalised the \emph{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos, krelling2014jogos, villar2008contravencao}.\footnote{The National Lottery Company (\emph{Companhia das Loterias Nacionaes do Brazil}), a public-private partnership founded four years after the creation of the \emph{jogo do bicho}, also lobbied actively for a hard-line stance against the \emph{bicheiros} \citep[82]{da1999aguias}.}
+A \textit{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace.
+A network of street bookmakers, called \textit{bicheiros}, made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the numbers game \citep[37]{chazkel2011laws}.
+The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: ``Today $[1985]$ in Rio de Janeiro, the game is everything.
+$[\dots]$  Nobody works!
+Everybody plays''  \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.}
+
+But this tolerant state of affairs did not last.
+Civil servants and police officers criminalised the \textit{jogo do bicho} on the grounds of public safety, and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}.
+The campaign extended for several decades and received considerable support from the \textit{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}.
+The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and the lottery remains illegal to this day.
 
 Yet the game has survived.
-The animal game has outlasted more than 30 Brazilian presidents and thrived under military regimes and democratic governments alike \citep{gaspari2002ditadura, jupiara2015poroes}.
-But more than a act of defiance, the \emph{jogo do bicho} is a successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
-A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \emph{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world.\footnote{See \url{http://goo.gl/9kNeX8} and \url{http://goo.gl/8FSAZl} (in Portuguese).
-Access: December 2016.} \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost the same number of employees that the oil giant Petrobras had in 2011 \citep{exame2013petrobras}.\footnote{In 1966, Time Magazine wrote that the \emph{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the Brazilian workforce.
-See \url{http://content.time.com/time/magazine/article/0,9171,842527-1,00.html}.
-Access: December 2016.}
+The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under both military regimes and democratic governments \citep{jupiara2015poroes}.
+But more than an act of civil disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}.
+A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world \citep{huff2015jogos}.
+\citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in Rio de Janeiro, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was ``the largest single industry in Latin America'' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}
 
-\begin{figure}[!htbp]
-	\centering
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho01.jpg}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho02.jpg}
-	\end{minipage}
-	\caption{Left: cartoon of the Baron of Drummond and the animals of the \emph{jogo do bicho} (1896).
-Right: entry ticket to Rio's Zoological Garden that allowed the bearer to join the raffle.
-Sources: Instituto Histórico e Geográfico Brasileiro, Rio de Janeiro, Revista Illustrada, ano 21, no.
-718 (1896) and Museu da Imagem e do Som, Rio de Janeiro.
-Reproduced in \citet[35--36]{chazkel2011laws}.}
-	\label{fig:barao}
-\end{figure}
+In this article I offer a rational choice interpretation of the \textit{jogo do bicho}.\footnote{I adopt a very broad definition of rationality in this paper.
+In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information at the moment of their choices.
+Thus, my analysis employs a ``thin'' notion of rationality and a `thick' description of social institutions, where individual action can only be understood with reference to the social environment \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, commonly called \textit{bicheiros},\footnote{As I describe in the next section, the \textit{jogo do bicho} structure can be broadly divided into three occupations: the rich game financiers (\textit{banqueiros}), mid-level managers (\textit{gerentes}), and street bookmakers (\textit{bicheiros}).
+While there are important hierarchical differences among these groups, all members of the \textit{jogo} are collectively called \textit{bicheiros} in Brazil.
+I follow the same practice here, and only refer to their specific role within the animal game structure when necessary for clarification.} have developed unique strategies to solve collective action problems and to maximise their political strategies.
+This does not mean their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals.
+Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive, they do address the long-term needs of the \textit{jogo do bicho} financiers.
 
-Moreover, the animal game plays a crucial role in the expansion of Rio de Janeiro's Carnival Parade, a popular festivity synonymous with Brazil at home and abroad \citep{araujo2003carnaval,costa2001100,da1973carnaval, da1979carnavais,vianna1995misterio}.
-\emph{Bicheiros} donate hefty sums to `samba schools' to gather support of poor communities and, no less importantly, to co-opt local politicians attracted by the financial and electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.
-This patron-client relationship has been proven effective: in 2016, the Carnival generated about USD 900 million in revenue and the Rio de Janeiro state received more than one million tourists.\footnote{Data provided by the Brazilian government.
-See \url{https://goo.gl/XMcbTM} (in Portuguese).
-Access: December 2016.}
+Like any business manager, \textit{bicheiros} have to run their firm with low costs to increase profits.
+However, the fact that the \textit{jogo do bicho} is clandestine imposes additional difficulties to its managers.
+In particular, \textit{bicheiros} face two main challenges to keep the lottery running.
+First, they need to gather public support so gamblers are not discouraged to engage in the lottery despite it being illegal.
+Second, \textit{bicheiros} also have to ensure the state repression is not prohibitively costly to their business; otherwise they would be better off shutting it down.
+I argue below that the \textit{bicheiros} have succeeded in both by using carefully-designed reputation strategies and employing costly signals to the communities they serve.
 
-In this article I offer a rational choice interpretation of the \emph{jogo do bicho} and discuss how \emph{bicheiros} promote social order, solve information asymmetries, and reduce negative externalities.
+I use the case of Rio de Janeiro to illustrate how the \textit{jogo do bicho} has overcome the obstacles to its expansion.
+Rio is a particularly interesting case because in no other part of Brazil is the game the financiers established such an effective patronage network.
+\textit{Bicheiros} have sponsored political campaigns, financed cultural activities and football teams, and sometimes even run in local elections themselves.
+I discuss the ways by which the \textit{bicheiros} have exploited fragilities of the Brazilian political system to their advantage and how those practices have weakened Brazilian democracy.
+
 My analysis discusses three strands of academic literature.
@@
 For instance, \citet{gambetta1996sicilian} examines the strategies used by the Sicilian Mafia to settle disputes among their members and enforce rules in the areas they exercise control.
-\citet{leeson2007arrgh,leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations.
-\citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking and coordinating productive activities in California prison gangs.
-I argue that \emph{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and foster trust among criminals.
-Moreover, I also describe how \emph{bicheiros} have developed sophisticated financial mechanisms, such as informal hedging operations and risk-sharing contracts, to prevent predatory behaviour in their community.
+\citet{leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations.
+\citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking behaviour and in coordinating productive activities in California prison gangs.
+I argue that \textit{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and to foster trust in the community.
 
-Second, this work relates to the literature on repugnant transactions and the relationship between morality and the market \citep{boettke1995morality, roth2007repugnance, sandel2012money, satz2010some, simmel1900geldes, zelizer1979morals}.
-In the following sections, I claim that the Brazilian elites have attached pejorative meaning to the \emph{jogo do bicho} to constrain the gambling market.
-I provide evidence that \emph{bicheiros} were aware of this problem, and as a response, they devised a series of rules aimed at reducing the costs associated with repugnance \citep{labronici2014sorteio, magalhaes2005ganhou}.
-\emph{Bicheiros} have made considerable efforts to increase the levels of trust in the system and distance themselves from other types of illegal activities.
-Their main tool to increase credibility was costly signalling, that is, the \emph{bicheiros} hoped the public would see them as credible brokers by sacrificing their immediate interests  \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}.
+Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,spence1973job}.
+I provide evidence that \textit{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and to reduce the uncertainty associated with clandestine markets.
+Their main tool to increase credibility was costly signalling \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}.
+\textit{Bicheiros} believed that by sacrificing their immediate interests they could gain a reputation of honesty that would benefit them in the long run.
 
-Lastly, this work connects to the literature on state capture, which is among the most important topics in public choice theory \citep{hellman2003seize, rose1978corruption, rose1999corruption, shleifer2002grabbing, tollison1982rent}.
-More specifically, I use the Brazilian case to illustrate how politicians and civil servants can be co-opted by criminal groups and produce sub-optimal social outcomes.
-\citet{queiroz1992carnaval} explored why \emph{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities.
-\citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested that the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s.
-In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \emph{jogo do bicho} and the military regime in Brazil (1964--1985).
-I supplement this literature by highlighting how asymmetrical information, agency dilemmas, and rent-seeking behaviour offer convincing explanations to the issues presented above.
-Although those concepts have a long tradition in public choice, scholars have not applied those ideas thus far to understand the dynamics of the \emph{jogo do bicho}.
+Lastly, this work connects to the literature on state capture, which is one of the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}.
+More specifically, I use the Brazilian case to illustrate how politicians and civil servants are co-opted by criminal groups, and how this collusion distorts the electoral process and benefits wealthy members of the \textit{jogo do bicho} network.
+\citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed this influence gave them leverage over political authorities.
+\citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s.
+In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \textit{jogo do bicho} and the military regime in Brazil (1964--1985).
+I supplement this literature by highlighting how asymmetrical information and rent-seeking behaviour offer convincing explanations to the issues presented above.
+Although those concepts have a long tradition in public choice, scholars have not applied those ideas to understand the dynamics of the \textit{jogo do bicho}.
 By doing so, I integrate seemingly contradictory historical facts into a single narrative that connects micro-level decisions to macro-level outcomes.
 
-The remainder of this article proceeds as follows.
-Section \ref{sec:organisation} presents a brief historical overview of the \emph{jogo do bicho}.
-It describes the necessary conditions for the emergence of the game and presents its basic organisational structure.
-Section \ref{sec:governance} details the \emph{jogo do bicho}'s governance mechanisms, particularly the strategies employed by vendors to increase trust in markets that operate at the margins of the law.
-Section \ref{sec:capture} discusses the links between illegal gambling markets, samba schools and the Brazilian state.
-Section \ref{sec:conclusion} offers some concluding remarks.
+\section{An Overview of the \textit{Jogo do Bicho}}%
+\label{sec:overview}
 
-\section{\emph{Jogo do Bicho} as an Emergent Institution}
-\label{sec:organisation}
+\subsection{Historical Background: How the \textit{Bicheiros} Avoided Extinction}%
+\label{sub:historical_background}
 
-\subsection{Historical Background}
-\label{sub:background}
-
-The early history of the \emph{jogo do bicho} is a textbook example of spontaneous order.
-Spontaneous orders are emergent macro-level phenomena that result from voluntary actions of purposive, self-interested individuals utilising their contextual knowledge \citep{boettke1990theory, boettke2005methodological, hayek1945use, hayek1960constitution, hayek1973law, leeson2008coordination, menger1871grundsatze, polanyi1948planning, polanyi1951logic}.
-Drummond, the Zoological Garden's original owner, designed the basic framework for the \emph{jogo do bicho}; but independent bookmakers were the ones who popularised the game \citep[77]{magalhaes2005ganhou}.
-Ticket sellers could quickly respond to market signals and then allocate their products where they were more valuable because of the lack of central coordination.
-Moreover, competition among sellers fostered innovation, and the \emph{bicheiros} invented new game rules to make the lottery more appealing to their customers \citep[61]{mello1989historia}.
-In this sense, the animal game is the materialisation of an evolutionary process of entrepreneurial discovery in which the interactions that provided the highest value to consumers were preserved over time \citep{boettke2008gordon, boettke2014entrepreneurship, buchanan1964should, hayek1978competition, kirzner1997entrepreneurial}.
-
-However, the \emph{jogo do bicho} only emerged because of historically contingent circumstances.
-The late nineteenth-century Brazil had four characteristics that explain how the animal game came to being: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
-Figure \ref{fig:dag} presents a simple directed acyclic graph (DAG) \citep{pearl2009causality} that shows the relationships between these explanatory variables and the development of the \emph{jogo do bicho}.\footnote{The main purpose of direct acyclic graphs is to graphically display the possible links among the exposure variables, confounders, and outcomes \citep{morgan2014counterfactuals, pearl2009causality}.
-DAGs are transparent by definition, as all theoretical choices made by the researcher are stated explicitly in the model.
-Each single-headed arrow in a DAG indicates that the variable at the origin causes the variable at the end of the directed edge.
-Dashed edges suggest that two variables are jointly dependent on unobserved common causes.
-There are no assumptions regarding the functional form of the relationships, and unless mentioned otherwise, the arrows represent fully non-parametric associations.
-Variables between two nodes are mediators, and variables pointed at by two or more factors have multiple causes.
-They are called \emph{colliders}.
-For the sake of clarity, errors are assumed independent and often excluded from the graphs.
-For an accessible introduction to DAGs see \citet[chap.
-3--4]{morgan2014counterfactuals} and \citet{pearl2016causal}.}
-
-\begin{figure}[!htbp]
-	\begin{tikzpicture}[->,>=stealth',shorten >=4pt,auto,node distance=3cm, semithick][h]
-		% nodes %
-		\node[circle,fill,inner sep=3pt,label=above right: \shortstack{Urban \\ Poverty}] (p) {};
-		\node[circle,fill,inner sep=3pt,label=right: \shortstack{Jogo \\ do Bicho}, right = 8 of p] (y) {};
-		\node[circle,fill,inner sep=3pt,label=below left: {Immigration}, below = 3 of p] (i) {};
-		\node[circle,fill,inner sep=3pt,label=above:{Money Supply}, above = 3 of p] (m) {};
-		\node[circle,fill,inner sep=3pt,label=below right:{Social Ties}, right = 3 of i] (s) {};
-		\node[circle,fill,inner sep=3pt,label=above:{Weak Law Enforcement}, above = 3 of y] (l) {};
-		\node[circle,fill=white, draw, outer sep=0pt, inner sep=3pt,label= left: \shortstack{Correlated \\ Background Factors}, above left = 2 of p] (u) {};
-		% edges %
-		\draw[->, line width = 1.2] (p) -- (y);
-		\draw[->, line width = 1] (s) -- (y);
-		\draw[->, line width = 1] (i) -- (s);
-		\draw[->, line width = 1.2] (l) -- (y);
-		\draw[->, line width = 1] (m) -- (y);
-		\draw[->, line width = 1.2] (i) -- (p);
-		\draw[->, dashed, line width = 1] (u) -- (p);
-		\draw[->, dashed, line width = 1] (u) -- (m);
-	\end{tikzpicture}
-	\caption{Directed Acyclic Graph -- Explanatory Variables for the \emph{Jogo do Bicho}}
-	\label{fig:dag}
-\end{figure}
+The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.
+I discuss each of these elements below.
 
 I start with the impact of urban poverty on the animal game.
-Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, fausto2014concise, naro1992revision, skidmore1993black}.
-The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration, smith1979ethnic}.
-Nevertheless, the job market tightened considerably after the \emph{Encilhamento} financial crisis of 1891 \citep{topik2014political, triner2005baring}.
-During the economic downturn, the informal economy was an obvious destination for the urban poor.
-Given its widespread popularity, the \emph{jogo do bicho} attracted hopeful entrepreneurs, either Brazilian or foreign-born, who could not enter the formal labour force.\footnote{The underground economy was also more democratic than the formal sector.
-As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor.
-These vendors used to sell different types of merchandise and many of them would later offer \emph{jogo do bicho} tickets.}
+Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising.
+Cities like São Paulo and Rio de Janeiro offered a number of occupations for former slaves who wanted to move away from their former rural masters \citep{andrews1991blacks, skidmore1993black}.
+Increasing numbers of Asian and European immigrants also moved to large cities after arriving in Brazil, as urban areas usually provided better standards of living than the countryside \citep{hall1969origins, lesser2013immigration}.
+However, the hopes of the African-Brazilians and the new foreign settlers would be frustrated by a series of downturns in the Brazilian economy.
+Brazil's labour markets suffered a severe contraction in the wake of the \textit{Encilhamento} financial crisis of 1891, and the economic instability aggravated the already difficult conditions of the working classes \citep{topik2014political, triner2005baring}.
 
-The immigration also influenced the \emph{jogo do bicho} via social ties.
-Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{klein1994imigraccao, lobo2001imigraccao, trento1989outro}.
-Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}.
-Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, torcato2011repressao, villar2008contravencao}.
-Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \emph{jogo do bicho}'s formative years.
+As a result, large swathes of the urban population turned to the informal economy.
+As \citet[115]{chazkel2011laws} observes, there were few occupations available to lower-class women and foreigners in the 1890s, and a large number of poor workers became street vendors.
+The profession requires little technical skills and has low barriers of entry, but it can be profitable if vendors are able to quickly identify a growing demand for a particular product.
+The \textit{jogo do bicho} was one such case; the game had a surge in popularity in the 1890s, and it also offered a high rate of return.
+As the demand for lottery tickets grew, the \textit{jogo} comprised an important share of the Brazilian extra-legal economy.
+
+Immigration also influenced the \textit{jogo do bicho} via social ties.
+Most foreigners who moved to Brazil came from countries such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}.
+Family and neighbourhood networks created incentives for immigrants to establish trade relations and to enforce cooperation through community responsibility systems \citep{roth2014prison}.
+Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil} and in the \textit{jogo do bicho} in particular \citep{magalhaes2005ganhou, villar2008contravencao}.
+Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \textit{jogo do bicho}'s formative years.
 
@@
 The abolition of slavery and the growing industrialisation of Brazil increased the amount of capital available in the country \citep{franco1987reformas, schulz2008financial}.
-The 1888 Banking Act gave extra liquidity to local financial markets, and the \emph{jogo do bicho} entrepreneurs utilised that increase in the monetary base to extend the scope of their business.
-Some years later, the animal game would be available not only across the city of Rio de Janeiro but throughout Brazil \citep[76]{da1999aguias}.
+Moreover, the 1888 Banking Act gave extra liquidity to local financial markets, which made credit more widely available in cities like São Paulo and Rio de Janeiro.
+Individuals received a temporary boost in personal income, a part of which they spent on leisure activities such as the \textit{jogo do bicho}.
+Moreover, the lottery attracted new entrants as it became more profitable, and in only a few years similar versions of the animal game were available throughout Brazil \citep[79]{da1999aguias}.
 
-The country's lax financial policy might correlate with poverty through unspecified factors.
-For instance, political decisions may have caused inadvertently both poverty and the expansion of the monetary base \citep{mattos2013shantytown, schmidt1982modernization}; alternatively, external events such as institutional instability \citep{costantini2014index, fausto2014concise, luna2014economic} or commodity shocks \citep{musacchio2014colonial} could be the cause of those two variables.
-There is not enough evidence to discard such scenarios.
-To illustrate this uncertainty, the two nodes, namely, money supply and urban poverty, are connected with a dashed edge in Figure \ref{fig:dag}.
+The last necessary condition for the emergence of the \textit{jogo do bicho} is weak law enforcement.
+\citet[69--100]{chazkel2011laws} notes that until the 1940s, police district chiefs operated within a large margin of discretion, so official repression against bookmakers was notably inconsistent.
+In the early years of \textit{jogo do bicho}, lottery ``bankers'' were allowed to operate virtually free from police interference, which surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}.
+Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code, and in 1941 the animal game was banned.
+Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.} Recent estimations show that the prohibition of the \textit{jogo do bicho} prevented the state from earning up to BRL 20 billion (USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
 
-The last necessary condition for the emergence of the \emph{jogo do bicho} is weak law enforcement.
-\citet[69--100]{chazkel2011laws} notes that until the 1940s police district chiefs operated within a large margin of discretion and repression against bookmakers was idiosyncratic.
-Prosecution against the \emph{bicheiros} had hardened in 1917, but only in 1946, when the federal government banned all gambling activities in the country \citep[155--156]{magalhaes2005ganhou}, the law was consistently enforced.
+Since the mid-twentieth century, the \textit{jogo do bicho} has been outlawed but it continues to be ubiquitous in Brazil.
+There is virtually no Brazilian city that does not have its local group of \textit{bicheiros}.
+\citet{cross2006risk} suggest a distinction between informal and illegal markets that is useful to understand the current ``semi-legal'' status of the \textit{jogo}.
+The animal game started as an informal activity, in which the Baron of Drummond and his associates sold lottery tickets to the public.
+Although the state did not regulate the lottery market, the product itself was not illegal or allegedly immoral.
+In fact, many organisations, including the Catholic Church, sold raffles and lottery tickets to sponsor their activities \citep[49]{torcato2011repressao}.
+Moreover, the state initially found no contradiction between the animal game and its own lottery, and both coexisted for about 15 years \citep[559]{chazkel2007beyond}.
+After 1941, however, selling \textit{jogo do bicho} tickets became a legal offence after a long campaign asking for its criminalisation.
+Since then, the \textit{jogo} has moved underground and consciously evaded state regulations.
+This shift from informal to illegal has brought important changes to the whole dynamics of the \textit{jogo}.
+As an example, wealthy animal game bosses have colluded with shady sectors of the armed forces for protection, and this violence has also been used against potential competitors.
+Moreover, this move to illegality enabled the game to exploit and subsidise large sectors of Brazil's formal economy, mainly in poor urban areas.
+I describe the animal game organisation structure and its impact in the Brazilian economic activity in further detail below.
 
-\subsection{Organisational Structure}
+\subsection{A Hierarchical Organisational Structure}%
 \label{sub:organisation}
 
-The animal game has three levels of hierarchy.
-At the bottom are the \emph{bicheiros}, who are those in charge of selling \emph{jogo do bicho} tickets \citep{chazkel2007beyond, chazkel2011laws, da1999aguias, labronici2014sorteio, magalhaes2005ganhou, misse2007illegal}.
-\emph{Bicheiros} are the most visible part of the \emph{jogo do bicho} structure.
-The bookmakers often build their vending stands inside the premises of a local shop, such as a small grocery store or a pub, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}.
-\emph{Bicheiros} usually work alone, but they may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
+The animal game operates with three levels of hierarchy.
+At the bottom level are the \textit{bicheiros}, those in charge of selling \textit{jogo do bicho} tickets \citep{chazkel2007beyond, da1999aguias}.
+\textit{Bicheiros} are the most visible part of the \textit{jogo do bicho}.
+The name loosely describes all those involved in the lottery organisation, yet their meaning in the \textit{jogo do bicho} structure is more particular and refers to street-level ticker sellers.
+The \textit{bicheiros}  usually build their vending stands inside the premises of a local shop, such as a small grocery store, and are recognisable by their chairs facing the street, stamps and blocks of paper \citep[259]{chazkel2011laws}.
+The street bookmakers usually work alone but may employ up to 10 people depending on how busy their betting site is \citep[69]{labronici2014sorteio}.
 
-The \emph{gerentes} (managers) oversee all \emph{jogo do bicho} stands in a given area.
+The \textit{gerentes} (managers) oversee all \textit{jogo do bicho} stands in a given area.
 Their task is akin to that of a firm accountant.
-Gerentes control the cash flow between the \emph{bicheiros} and the bankers, manage the payroll of the employees, and provide financial information to the top members of the organisation.
+Gerentes control the cash flow between the \textit{bicheiros} and the bankers, manage the payroll of the employees, and provide financial information to the top members of the organisation.
 They also supervise individuals who carry menial tasks in the business, transfer money to other gambling branches and double-check the balance sheets of the betting sites \citetext{\citealp[71]{labronici2012paratodos}; \citealp[142]{misse2007illegal}}.
 
-\begin{figure}[!htbp]
-	\centering
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho03.jpg}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
-		\includegraphics[width=\textwidth, height=6cm]{images/bicho04.jpg}
-	\end{minipage}
-	\caption{\emph{Jogo do bicho} betting sites in 1917 and in 2011.
-The picture on the right shows a \emph{bicheiro}, a street-corner vendor.
-Sources: \citet{alecrim2012bicho} and \citet{ferrarini2011bicho}.}
-	\label{fig:banca1917}
-\end{figure}
+The \textit{banqueiros} (Portuguese for bankers) occupy the top position in the \textit{jogo do bicho} hierarchy.
+They comprise the small financial elite of the game.
+A 2012 report by the Brazilian Federal Police affirmed that 10 \textit{banqueiros} controlled the market throughout the country, and five of them were based in the state of Rio de Janeiro \citep{globo2012contraventores}.
+Apart from funding the game, the bankers provide support for the employees to undertake their activities.
+The \textit{banqueiros}' attributions include paying bribes to police personnel, bailing out sellers arrested by security forces, and offering judicial assistance to employees in case of legal persecution \citep[75]{labronici2012paratodos}.
 
-The \emph{banqueiros}, or the Portuguese for bankers, occupy the top position in the \emph{jogo do bicho} hierarchy, comprising the small financial elite of the game.
-A 2012 report by the Brazilian Federal Police affirmed that 10 \emph{banqueiros} controlled the market throughout the country; five of them based in the state of Rio de Janeiro \citep{globo2012contraventores}.
-Apart from funding the game, the bankers provide support for the employees to undertake their activities.
-The \emph{banqueiros}' main attributions include paying bribes to police personnel, bailing out sellers arrested by security forces, and offering judicial assistance to employees in case of legal persecution \citep[75]{labronici2012paratodos}.
+The lottery bosses run their businesses from fortified houses in unknown locations called \textit{fortalezas} (``forts'').
+The first \textit{fortalezas} likely appeared in the 1950s, when the animal game was already well-established across the Brazilian territory.
+The period coincides with a time when the \textit{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}.
+Due to the growing scope of the \textit{jogo do bicho} economy, \textit{banqueiros} decided to move their operations away from the public to avoid police persecution and to make coordination easier.
 
-\emph{Banqueiros} run their businesses from fortified houses in unknown locations, the \emph{fortalezas} (`forts').
-The first \emph{fortalezas} likely appeared in the 1950s, when the animal game was already well-established across the Brazilian territory.
-The period coincides with a time when the \emph{jogo do bicho} finances had become increasingly concentrated in fewer hands \citep[259]{chazkel2011laws}.
-Due to the growing size of the \emph{jogo do bicho} economy, \emph{banqueiros} decided to move their operations away from the public to avoid police persecution and reduce coordination costs.
+\textit{Banqueiros} solve problems of internal cooperation by providing club goods \citep{buchanan1965economic} while simultaneously shunning cheaters through selective punishments \citep{bo2005cooperation, roth1978equilibrium}.
+The first club good offered to \textit{bicheiros} by their bosses is private security.
+As the game is illegal, street sellers cannot rely on official institutions to protect themselves.
+Thus, the game bankers have built an extensive network of gunmen and bribed police officers to protect their employees from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
+``Zé'' (Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \textit{jogo do bicho} informal security personnel:
 
-Although the forts provided safety to the bankers, the existence of those hideouts posed a challenge to the organisation.
-Bankers removed from the public view are not accountable to players and booking agents.
-Similarly, bankers and managers working in the \emph{fortalezas} cannot oversee their employees as effectively as before.
-Considering that the animal game itself is illegal and the amount of money involved in the bets is often substantial, both \emph{banqueiros} and booking agents have strong incentives to defect.
-Players, in turn, have no evident reason to trust \emph{banqueiros} or \emph{bicheiros}.
-How do \emph{jogo do bicho} agents overcome trust issues and cooperate under uncertainty?
+\begin{quote}
+Bums are scared and they don't mess around with us; they think there's an officer nearby or something like that.
+Look at all this money here!
+[Shows the interviewer a handful of cash.] It's not ours [referring to street-corner bookmakers].
+And if it's not ours, it's someone else's.
+When I worked in Penha\footnote{Penha is a low middle-class neighbourhood in the city of Rio de Janeiro.}, the owner of a pub close by always asked me to stay at the front door of his pub.
+People know that bums are afraid of \textit{bicheiros}.
+\end{quote}
 
-I argue that the \emph{jogo do bicho} solves problems of internal cooperation by providing club goods \citep{buchanan1965economic, berman2008religion, berman2009radical, leeson2011government, roth2014prison} while simultaneously shunning cheaters through punishments and appeals to `the shadow of the future' \citep{axelrod1984evolution, axelrod1985achieving, bo2005cooperation, roth1978equilibrium}.
-Clients and \emph{bicheiros} cooperate based on trust-enhancing mechanisms, most of them devised specifically for the \emph{jogo do bicho} \citep{da1999aguias, magalhaes2005ganhou}.
-Such mechanisms are relevant because they have allowed the \emph{jogo do bicho} to distance itself from other shadow markets and become a profitable enterprise in the long run.
+The \textit{banqueiros} do not use violence only against other criminals.
+They often employ violent methods against competitors and their own staff, too.
+\citet{jupiara2015poroes} argue that Ailton Guimarães Jorge, a former Army officer, tortured and murdered rival lottery bosses in the late 1970s.
+One of his former allies, Army Colonel Paulo Malhães, told the Rio de Janeiro State Truth Commission that Guimarães ``went on a rampage'' to consolidate his power \citep{belem2015guimaraes}.
+Castor de Andrade, Rio's most influential animal game banker, also employed similar methods to run his business.
+Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors.
+In a famous case, Andrade shot Euclides Ponar, an old \textit{jogo do bicho} boss known as the ``Grey-Headed Chinese'' (\textit{China Cabe\c{c}a Branca}) after Ponar denounced a fraud in lottery draws in 1976.
+Andrade was likely involved in other assassination plots in the 1990s  \citep{globo2017castor}.
 
-\section{Governance of the \emph{Jogo do Bicho}}
-\label{sec:governance}
+Despite these serious events, killings are rare in the animal game.
+Since the lottery bosses can credibly indicate that violence is a low-cost option for them, the mere threat of punishment is enough to discourage defectors.
+This is a good strategy for the \textit{banqueiros}.
+The fact that they had committed violent crimes in the past reduces the need to commit them in the present, and as a result the bosses can spend less money on security and increase profits.
+As it happens in many traditional markets, if a group is able to form a cartel, they can increase the price of their services without fearing immediate competition.
+The same logic is valid for the \textit{jogo do bicho}, although through unconventional means.
 
-\subsection{Gambling Markets and Repugnant Transactions}
-\label{sub:repugnance}
+The threat of violence is not the only tool the \textit{bicheiros} have at their disposal.
+They balance the use of violence with financial benefits to low-rank members of the organisation.
+For instance, street \textit{bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.
 
-The \emph{jogo do bicho} is a repugnant market.
-Individuals that like to gamble cannot do so because of strong moral objections from outsiders \citep{brisset2016marche, roth2007repugnance, satz2010some, zelizer1979morals}.
-As early as in 1890, Brazilian public authorities positioned themselves against the \emph{jogo do bicho} arguing that `{[}\ldots{}{]} this type of amusement is prejudicial to the interests of the unwise, who are naively seduced by the deceptive hope of uncertain lucre' \citep[544]{chazkel2007beyond}.
-In 1941, the government banned the animal game;\footnote{See: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del3688.htm} (in Portuguese).
-Access: December 2016.} five years later, it prohibited all games of chance.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.
-The text can be read at: \url{http://www.planalto.gov.br/ccivil_03/decreto-lei/Del9215.htm} (in Portuguese).
-Access: December 2016.} The \emph{jogo do bicho}, casinos and bingos remain illegal in the country.
-Recent estimations show that the prohibition of the \emph{jogo do bicho} have prevented the state from earning BRL 15 to BRL 20 billion (USD 4.5 to USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players.
-\citep{congressoemfoco2015bicho, fsp2016legalizarbicho}.
+The most important financial mechanism implemented by bankers to help \textit{bicheiros} is the \textit{descarga}, loosely translated as ``the unloading''.
+The \textit{descarga} is the \textit{jogo do bicho}'s main hedging technique and its purpose is to insure small bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}}.
+Booking agents are sometimes unable to honour expensive bets.
+The top prize in the animal game pays up to 4,000 times the amount invested; thus, \textit{bicheiros} may have to raise thousands of Brazilian Reals in a single day to pay the lucky winners.
+To prevent the \textit{quebra da banca} (``bust of the bank''), \textit{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
+The \textit{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \textit{jogo do bicho}.
 
-In contrast with the official statements, the noxious element of the \emph{jogo do bicho} does not come from its inherent randomness.
-The game is `repugnant' precisely because it is \emph{a market}, a setting in which individuals can monetise the entertainment for private profit \citep{chazkel2007beyond, chazkel2011laws}.
-The Brazilian state has never seen any contradiction between banning games of chance and running a national lottery company of its own; even the Catholic Church, which has long condemned the \emph{jogo do bicho}, frequently organises raffles to fund its activities \citep[49]{abreu1996imperio, magalhaes2005ganhou}.
-Only after the introduction of private money that Brazilians objected the idea of benefiting from someone else's bad luck.
+The \textit{descarga} has significantly changed the distribution of resources in the \textit{jogo do bicho}, and the richest bankers benefited the most from it.
+Simple probability dictates that a booking agent rarely pays the highest lottery prize, and yet the bankers receive a commission for \textit{every game} they hedge.
+Over time, there is a transfer of income from the bottom to the top of the animal game structure due to the fees.
+This accumulation of capital is one of the reasons in the 1990s bankers started offering other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
+They simply had more capital to invest, and this eventually helped to compensate the downturn in the \textit{jogo do bicho} markets in the last years \citep{globo2017castor}.
+In sum, while the \textit{descarga} has made the game more resilient at the aggregated level, it increased profits for the richest financiers at the expense of small bookmakers.
 
-In this sense, the main obstacle that confronted \emph{bicheiros} was to convince others that the animal game would not cause the Brazilian society `to slide down a slippery slope to genuinely repugnant transactions' \citep[45]{roth2007repugnance} such as prostitution or debt bondage.
-As the century-old history of the game can attest, \emph{bicheiros} have succeeded in this task.
-But how?
-The literature on repugnant costs tell us little about how markets transition from noxious to tolerated.
-Here I posit two mechanisms that reduced the stigma associated with the game: 1) \emph{a strong reputation of honesty} expressed by costly signals from sellers, and 2) the provision of \emph{selective incentives} for both clients and booking agents.
-Below, I offer evidence that these two factors allowed the animal game to reach its current semi-legal status in Brazil.
+This has brought broad consequences to the Brazilian illegal economy.
+As the number of slot machines increased in Brazil, the opportunity for criminals to use them for money laundering increased as well.
+The scheme is easy to carry out.
+The owner of a slot machine issues a ticket with a winning prize, and the criminal declares the prize as his legitimate wealth.
+Then he can legally use that money for any purpose without raising suspicion from the authorities.
+The practice have become more widespread in the last decades, and in 2009 a Federal Police task force arrested about a dozen \textit{jogo do bicho} bosses involved in the so-called ``slot machine mafia''.
+Relatives of Castor de Andrade were also involved in the scheme \citep{estado2011cacaniquel}.
+In 2012, President Dilma Rousseff sanctioned a law that considered slot machines and the \textit{jogo do bicho} as money laundering crimes \citep{agenciabrasil2012dilma}.
 
-\subsection{External Cooperation}
-\label{sub:external}
+Money from the \textit{jogo} also has other effects on the economy.
+Some of these effects are indirect.
+For instance, the \textit{jogo do bicho} gives a boost to the Brazilian economy by providing jobs for unskilled workers who cannot easily join the labour market.
+By doing so, the \textit{jogo} prevents some of the poorest members of the Brazilian society from demanding more inclusive government policies, although they are the ones who would benefit the most from public assistance.
+Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds on other, probably wealthier, sectors of the population.
 
-Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and experimental studies \citep{dawes1977behavior, isaac1984divergent, kim1984free, marwell1981economists} have both demonstrated that long-term cooperation is possible whenever players expect future pay-offs to be higher than present ones.
-Fear of retaliation induces individuals not to cheat.
-Nevertheless, illegal organisations tend to discount the future even more heavily than the other groups, what makes cooperative behaviour among criminals uncommon \citep{gambetta2009codes, skarbek2011governance,skarbek2012prison,skarbek2014social}.
-The \emph{jogo do bicho} is an exception to this rule.
-The market properties of the game and inconsistent repression by Brazilian authorities have permitted \emph{bicheiros} to overcome the stigma of repugnance and improve the game's long-term profitability.
+Another indirect economic effect of the \textit{jogo} is the rise in inequality.
+While formal workers can -- at least in theory -- demand higher compensations during a market upswing, the same is not true for the animal game employees.
+The threat of violence reduces the space for collective bargaining with the lottery bosses, and higher profits at the top of the \textit{jogo do bicho} structure do not trickle down to those at the bottom.
+This is one of the reasons why the game bankers turned the game into an oligopoly: the use of violence guarantees new entrants will not be allowed to join the market and profits will be concentrated in the hands of very few individuals.
 
-The \emph{jogo do bicho} entrepreneurs have made considerable efforts to present themselves as honest brokers.
-The first trust-enhancing mechanism they have employed to foster external cooperation was the use of a \emph{fixed-multiplier formula} for pay-outs.
+The \textit{jogo do bicho's} market formation closely resembles what \citet{fligstein1996markets} calls ``markets as politics'', in which firms create institutions to restrain the competition and organise the labour force.
+At the formation of the lottery market, competition is fierce and participants are akin to social movements.
+They are constantly trying to convince others of the viability of their ideas.
+When markets stabilise, however, incumbents collude to impose their conditions of control to other players and to workers.
+In the case of the \textit{jogo}, this implies in a mix of financial incentives and violence threats carried by corrupt state agents or private bodyguards.
+
+A more difficult question is how the bankers elicit cooperation from \textit{external} members, such as gamblers, community leaders, or public officers.
+It is puzzling because \textit{bicheiros} do not use violence to induce individuals to play the lottery, nor have they ever clashed with the Brazilian government.
+Precisely because violence could shun profits, \textit{bicheiros} devised other mechanisms to create a friendly environment for the illegal lottery.
+I investigate two of them: costly signalling and reputation building.
+
+\section{Winning Hearts, Minds, and Pockets: Illegal Market Dynamics}%
+\label{sub:pockets}
+
+Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical case studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible even in difficult situations.
+The main requirement for sustained cooperation is that players believe future pay-offs will be higher than present ones.
+If that condition is true, fear of retaliation will induce individuals not to cheat.
+
+In theory, the same should also apply to illegal organisations.
+Yet in practice we see that criminal groups are generally short-term oriented, that is, they tend to discount the future more heavily than most people.
+This makes cooperative behaviour among criminal rather uncommon, and there is substantial evidence suggesting illegal groups face serious collective action problems \citep[e.g.,][]{gambetta2009codes, leeson2010pirational, skarbek2011governance, skarbek2012prison, varese2001russian}.
+
+The \textit{jogo do bicho} is an exception to this rule.
+The game has been running for more than a century without considerable interruption, attesting to the fact that \textit{bicheiros} have managed to solve collective action issues in one way or another.
+Importantly, the \textit{jogo} involves moderately low levels of violence -- at least when compared to other illegal activities such as drug trafficking.
+Moreover, Brazilians widely consider the \textit{jogo do bicho} an honest lottery, and reports of cheating are surprisingly uncommon.
+The popular motto associated with the game testifies in its favour: ``\textit{Vale o escrito; ganhou, leva}'', or ``What is written is what counts; if you win it, you take it'' \citep{magalhaes2005ganhou}.
+Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community: costly signalling and reputation strategies.
+
+Signalling theory predicts that when someone cannot easily observe a characteristic she is interested in, she searches for signals that credibly conveys pieces of that required information.
+In the case of an illegal lottery, the main signal a gambler is looking for is \textit{honesty}, that is, that she will receive her prize if she buys the winning ticket.
+
+Legal lotteries employ many techniques to show this is the case.
+For instance, the Brazilian official lottery, run by the federal government, are regularly audited by the \textit{Controladoria Geral da União} (Comptroller General of Brazil), the \textit{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
+The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute \citep{uol2016loteria}.
+
+A clandestine lottery, in contrast, cannot provide the same signals.
+Thus, the game providers need to assure gamblers that their business is honest although it is illegal.
+\textit{Bicheiros} addressed this issue using a traditional commercial practice: they invested on their reputations.
+
+The \textit{jogo do bicho} entrepreneurs have made considerable efforts to present themselves as honest brokers.
+The first trust-enhancing mechanism they have employed to foster external cooperation was the use of a \textit{fixed-multiplier formula} for pay-outs.
 It works as follows.
@@
 The fixed-multiplier formula alleviates such problems of adverse selection \citep{akerlof1970market, cohen2010testing, levin2001information}.
-As players and vendors known the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \emph{bicheiros} to a contract that can be easily enforced.
-This technique offers buyers a simple yet effective screening strategy that induces \emph{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.
+As players and vendors know the prize value beforehand, the method provides consumers with complete information about their individual prizes while also binding the \textit{bicheiros} to a contract that can be easily enforced.
+This technique offers buyers a simple yet effective screening strategy that induces \textit{bicheiros} to provide honest information about the game \citep{spence1973job, stiglitz1981credit}.
 
-\emph{Bicheiros} have addressed information asymmetries in another ways.
-Since the 1950s, when the \emph{jogo do bicho} bankers had moved their operations to the \emph{fortalezas}, the public could not oversee the lottery draws \citep[259]{chazkel2011laws}.
+\textit{Bicheiros} have addressed information asymmetries in another ways.
+Since the 1950s, when the \textit{jogo do bicho} bankers had moved their operations to the \textit{fortalezas}, the public could not oversee the lottery draws \citep[259]{chazkel2011laws}.
 This could lead to a decline in trust among buyers and vendors of lottery tickets and, as a result, to reduced profits.
-\emph{Bicheiros} have mitigated this problem with a two-pronged strategy.
-First, they started to utilise the winning numbers from the licit government-run lottery, the \emph{Loteria Federal}, instead of their own draws \citetext{\citealp[546]{chazkel2007beyond}; \citealp[89]{labronici2012paratodos}; \citealp[39-40]{mello1989historia}}.
+\textit{Bicheiros} have mitigated this problem with a two-pronged strategy.
+First, they started to utilise the winning numbers from the licit government-run lottery, the \textit{Loteria Federal}, instead of their own draws \citetext{\citealp[546]{chazkel2007beyond}; \citealp[89]{labronici2012paratodos}; \citealp[39-40]{mello1989historia}}.
 The federal lottery numbers are public information.
 The media broadcasts the draws on radio and TV, so any interested player can verify the selected numbers.
-The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public; hence, \emph{bicheiros} can free ride on the lottery's long-standing reputation of credibility.\footnote{As of April 2016, the lottery was audited by the \emph{Controladoria Geral da União} (Comptroller General of Brazil), the \emph{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young.
-The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute.
-See \url{http://noticias.uol.com.br/cotidiano/ultimas-noticias/2016/04/08/auditoria-dos-sorteios-da-caixa-e-confiavel-veja-como-e-o-processo.htm} (in Portuguese).
-Access: December 2016.}
+The Loteria Federal is also audited by two independent state institutions, a private accounting firm, and voluntary members of the public.
+Hence, \textit{bicheiros} can free ride on the lottery's long-standing reputation of credibility.
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=\textwidth, height=6cm]{images/bicho06.jpg}
-	\caption{Results of a \emph{jogo do bicho} draw from 09 January 2016.
-`Federal' means that the winning numbers were drawn by the federal government lottery.
-The first prize was group 14, the cat.
-Source: Unknown.
-Available at: \url{https://goo.gl/6PHV8u}.
-Access: December 2016.}
-	\label{fig:federal}
-\end{figure}
-
-Second, they included representatives of all major \emph{jogo do bicho} bankers in every draw and independently publicise the game results.
-Certain \emph{bicheiros} went as far as publishing the numbers in Rio's newspapers.
+Second, they included representatives of all major \textit{jogo do bicho} bankers in every draw and independently publicise the game results.
+Certain \textit{bicheiros} went as far as publishing the numbers in Rio's newspapers.
 In the early twentieth century, some tabloids were entirely dedicated to the game \citep[60]{magalhaes2005ganhou}.
@@
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=\textwidth, height=6cm]{images/bicho05.jpg}
-	\caption{\emph{Jogo do bicho} results are fixed on light poles in Rio de Janeiro. Source: \citet{gomes1998bicho}.}
-	\label{fig:poste}
-\end{figure}
+These efforts have proved popular with the game enthusiasts.
+Such mutual confidence reduces the potential for conflict in the game.
+As the public does not see the \textit{jogo do bicho} as violent or harmful, the stigma of repugnance associated with gambling becomes less pervasive.
+By reducing the possibilities of cheating and putting long-term interests first, the \textit{jogo do bicho} bankers have avoided the fate of other repugnant markets \citep[20]{da1999aguias}.
 
-These efforts have proved popular with the game enthusiasts.
-One often-repeated saying about the \emph{jogo do bicho} is that `in the \emph{jogo do bicho}, what is written down counts' \citep[159]{chazkel2011laws}, that is, buyers and sellers do fulfill their informal obligations without third-party enforcement.
-Such mutual confidence reduces the potential for conflict in the game.
-As the public does not see the \emph{jogo do bicho} as violent or harmful, the stigma of repugnance associated with gambling becomes less pervasive.
-By reducing the possibilities of cheating and putting long-term interests first, the \emph{jogo do bicho} bankers have avoided the fate of other repugnant markets and run their business relatively undisturbed for decades \citep[20]{da1999aguias}.
-
-\subsection{Internal Governance}
-\label{sub:internal}
-
-Individuals working at different levels of hierarchy often have non-aligned interests.
-As a result, it may occur that one party (the agent) behaves rationally in a manner that maximises his/her benefits, but that is contrary to the interests of his/her superior (the principal).
-This dilemma is pervasive in formal organisations \citep{holmstrom1979moral,jensen1976theory,moe1984new,shapiro2005agency,spence1971insurance}; in illegal markets perhaps it is even more so \citep{campana2013cooperation,gambetta2009codes,skarbek2011governance,skarbek2014social}.
-As monitoring costs in criminal businesses are higher than in formal ones, principals face considerable difficulties to induce cooperation from agents.
-Moreover, criminals often engage in opportunistic behaviour and `hidden actions', that is, they do not put the required levels of effort if they know they are not being monitored \citep[38--42]{arrow1985agency}.
-
-In the \emph{jogo do bicho} setting, one such problems concerns the trade-off between short- and long-term incentives for managers and bankers on the one side and bookmakers on the other.
-Managers have a permanent interest in the long-run profitability of the game, whereas street-corner booking agents tend to discount the future more heavily because their financial gains are small compared to that of their superiors.
-Additionally, bookmakers may denounce their employers to the police if they feel threatened.
-
-One way by which the \emph{jogo do bicho} principals solve the agency dilemma is by supplying club goods and selective incentives for low-tier members.
-Club goods are goods that can be simultaneously enjoyed by more than one individual but where exclusion mechanisms prevent consumption by non-members \citep{buchanan1965economic, cornes1996theory, olson1965logic, sandler1980economic, sandler1997club}.
-Basically, club goods are `public goods \emph{sans} non-excludability' \citep[928]{mcnutt1999public}.
-The first club good offered to \emph{bicheiros} by their bosses is private security.
-The game bankers have built an extensive network of gunmen and bribed police officers to protect their employees (and their profits) from other criminals \citetext{\citealp[48]{chinelli1993vazio}; \citealp[51]{labronici2012paratodos}}.
-The \emph{jogo do bicho} network has a powerful deterrence effect and lethal force is rarely employed.
-However, threats are constant.
-`Zé' (Little Joe), a bicheiro interviewed by \citet[52]{labronici2012paratodos}, described eloquently the deterring effect of the \emph{jogo do bicho} informal security personnel:
-
-\begin{quote}
-	[\dots] bums are scared and they don't mess around with us; they think there's a guard nearby or something like that.
-Look at all this money here!
-[shows the interviewer a handful of cash] It's not ours [referring to street-corner bookmakers].
-And if it's not ours, it's someone else's.
-When I worked in Penha (\emph{a low middle-class neighbourhood in the city of Rio de Janeiro -- translator's note}), the owner of a pub close to where I used to work always asked me to stay at the front door of his pub.
-People know that bums are afraid of \emph{bicheiros}.
-\end{quote}
-
-Apart from guaranteeing the physical integrity of the \emph{bicheiros}, bankers and managers also provide financial incentives for the bookmakers.
-\emph{Bicheiros} are allowed to receive tips from players, often have small expenses paid by managers, and may even request interest-free loans to cover unexpected costs such as illness-related expenses \citep{labronici2012paratodos}.
-
-However, the most important financial mechanism implemented by bankers to help \emph{bicheiros} is the \emph{descarga}, which is loosely translated as `the unloading'.
-The descarga is the \emph{jogo do bicho}'s main hedging technique and its purpose is to insure bookmakers against credit risk \citetext{\citealp[59]{labronici2012paratodos}; \citealp[178]{magalhaes2005ganhou}; \citealp[16]{misse2007illegal}; \citealp[75]{soares1993jogo}}.
-Booking agents are sometimes unable to honour expensive bets.
-As mentioned above, the top prize in the animal game pays up to 4,000 times the amount invested, thus \emph{bicheiros} may have to raise thousands of Brazilian Reals in a single day.
-To prevent the \emph{quebra da banca} (`bust of the bank'), \emph{bicheiros} and small bankers buy an insurance from wealthier financiers, who offer this service for a fee that ranges from 20\% to 25\% of the total selling amount \citep{fsp2006descarga}.
-The \emph{descarga} guarantees that small bookmakers will not have liquidity problems, thus permitting bookmakers to continue investing in the \emph{jogo do bicho}.
-
-The descarga has played an important role in reducing individual risk; nevertheless, it has also changed the distribution of resources in the \emph{jogo do bicho}.
-Simple probability dictates that a booking agent rarely pays the highest prize in the \emph{jogo do bicho}; in contrast, the bankers receive a commission for \emph{every game} they hedge.
-Over time, there is a transfer of income from the bottom to the top of the animal game structure led by this constant inflow of fees.
-This accumulation of capital is probably one of the reasons why bankers were able to diversify their businesses and offer other types of entertainment such as slot machines and sports lotteries \citep{estado2006cacaniquel,globo2015cacaniquel,terra2011cacaniquel}.
-The descarga has made the game more resilient at the aggregated level, although it increased profits for the richest financiers at the expense of small bookmakers.
+This ability to elicit cooperation from external actors caused the \textit{jogo do bicho} to last longer than collection action theories would anticipate.
+In that regard, the bankers' investments in costly signalling and reputation has largely paid off: the lottery's continuous profits are a proof of their success.
+With that continuous flow of income, \textit{banqueiros} were able to extend their influence well beyond the poor communities in which they operate.
+One of these areas is politics.
 
@@
 The impact of the \emph{jogo do bicho} is not restricted to the Brazilian economy.
-Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural and social festivity, the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,chinelli1993vazio,queiroz1992carnaval}.
-The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that `without the \emph{jogo do bicho} the Carnival would have ended' \citep{odia2016aniz}.
-Owing to that support, \emph{bicheiros} have established an extensive patronage network with samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}.
-Although that network brings large material benefits to their members, the patronage system has created perverse incentives for government officials.
+Since the 1960s, \emph{bicheiros} have been the key sponsors of the country's most important cultural festivity: the Rio de Janeiro Carnival parade \citep{bezerra2009mecenato,cavalcanti2006carnaval,queiroz1992carnaval}.
+The \emph{jogo do bicho} accounts for such large share of the funding of the parade that a famous \emph{banqueiro} once remarked that ``without the \emph{jogo do bicho} the Carnival would have ended'' \citep{odia2016aniz}.
+Based on that support, \emph{bicheiros} have established an extensive patronage network over samba schools and local politicians \citetext{\citealp[4641]{arguello2012criminalizaccao}; \citealp{congressoemfoco2007bicho}; \citealp{jornaldobrasil2011bicho}; \citealp[16]{misse2011crime}}.
+Such client network brings large material benefits to their members, yet it has created perverse incentives for government officials and has caused several distortions to the Brazilian democratic system.
 
-The \emph{jogo do bicho}'s clientelism is more evident in the state of Rio de Janeiro than in other parts of the country.
+The \emph{jogo do bicho}'s clientelism is most evident in the state of Rio de Janeiro.
 Historical factors explain why this is the case.
-Firstly, Rio de Janeiro city was the capital of Brazil for almost 200 years; despite losing the position to Brasília in 1960, it remains one of the country's main cultural and financial centres.
-Secondly, \emph{jogo do bicho} operators had historical ties with popular movements, which they eventually exploited to their advantage.
+First, Rio de Janeiro city was the capital of Brazil for almost 200 years, and despite losing the position to Brasília in 1960, it has remained one of the country's main cultural and financial centres.
+Secondly, \emph{jogo do bicho} operators had historical ties with popular movements, especially samba groups, and the animal game elites eventually exploited these connections to their advantage.
 Thirdly, the emergence of state-sponsored Carnival parades created a window of opportunity for \emph{bicheiros} to expand their influence over public authorities, either via bribing or by funding political campaigns.
@@
 In 1930, opposition leader Getúlio Vargas led a bloodless coup d'état that brought Brazil's First Republic to an end.
-During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by dismantling effectively federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,desouza1983estado,fausto1972revoluccao,fausto2014concise,skidmore1967politics}.
-In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,nava1998lessons,williams2001culture}.
-Popular music, in turn, occupied an important place in Vargas's project of `brazilianing Brazil'.
-Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{avelar2011brazilian,mccann2004hello,stockler2011samba,vassberg1969villa,vassberg1975villa}.
+During his first presidency (1930--1945), Vargas promoted a radical shift in Brazilian politics by effectively dismantling federalism in favour of a powerful executive branch and an expanded federal bureaucracy \citep[e.g.][]{bethell2008politicsvargas,fausto1972revoluccao,skidmore1967politics}.
+In terms of ideology, Vargas's authoritarian-corporatist \emph{Estado Novo} (``New State'') promoted a politicised nationalism designed to transcend the regional aspects of Brazilian culture \citep{lauerhass1972getulio,williams2001culture}.
+Popular music, in turn, occupied an important place in Vargas's project of ``Brazilianing Brazil''.
+Created in the late 1920s in the shanty towns of Rio de Janeiro, modern samba embodied the idea of the multicultural, racially-tolerant country the government aspired to forge \citep{mccann2004hello,stockler2011samba}.
 
 By the late 1930s, samba reached a unique position in Brazil's cultural identity.
-In a period when civil and political rights were limited \citep{de2001cidadania,duarte1993vicissitudes}, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}.
-Patriotic sambas exalted the country's natural beauties and the figure of the `friendly, happy, cordial and industrious' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
-The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of the male donkey and a female horse.
+In a period when civil and political rights were limited, Vargas used samba as a means to incorporate ethnic minorities and the new urban classes into the Brazilian mainstream \citep[213]{chinelli1993vazio}.
+Patriotic sambas exalted the country's natural beauties and the figure of the ``friendly, happy, cordial and industrious'' mulatto\footnote{A mulatto is a person of mixed white and black ancestry.
+The etymology of the word is originally derogatory as it alludes to `mule' (Latin: \emph{mulus}), the infertile offspring of a male donkey and a female horse.
 However, in the 1930s the word loses its pejorative connotation in Brazil.
-Mainly due to the work of sociologist \citet{freyre1933casa}, the idea of a racial democracy becomes pervasive in the government discourse, and as a result the word gains a positive tone \citep[4]{reiter2009brazil}.} \citetext{\citealp[47]{dangelo2016samba}; \citealp[51]{vianna1995misterio}}.
-The institutionalisation of the Carnival parade in 1935, and the subsequent increases in public funding to the festival, cemented the relationship between politicians and samba groups \citep{almeida2017carnaval,cabral2016escolas,soihet1998subversao}.
+Mainly due to the work of sociologist \citet{freyre1933casa}, the idea of a racial democracy becomes pervasive in the government discourse, and as a result the word gains a positive tone \citep[4]{reiter2009brazil}.} \citep[51]{vianna1995misterio}.
+The institutionalisation of the Carnival parade in 1935, and the subsequent increases in public funding to the festival, cemented the relationship between politicians and samba groups \citep{cabral2016escolas,soihet1998subversao}.
 
-However, the samba groups were not passive members in this process.
-Since the 1960s, the Rio Carnival expanded in scope and, stimulated by growing numbers of spectators, the parades became more elaborate \citetext{\citealp{cabral2016escolas}; \citealp[214]{chinelli1993vazio}; \citealp[240]{hertzman2013making}}.
-Unable to cope with the rising costs of the show, the `samba schools', which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}.
-This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,farias2013carnival,misse2011crime,queiroz1992carnaval}.
+The samba groups were not passive members in this process.
+Since the 1960s, the Rio Carnival has expanded in scope.
+Stimulated by growing numbers of spectators, the parades have become more elaborate \citetext{\citealp{cabral2016escolas}; \citealp[214]{chinelli1993vazio}; \citealp[240]{hertzman2013making}}.
+Unable to cope with the rising costs of the show, the samba schools, which are large samba groups that compete in the Carnival, resorted to the \emph{jogo do bicho} financiers to fund their activities \citep{misse2007illegal}.
+This informal agreement between samba school organisers and wealthy \emph{bicheiros} remains effective to this day, and many of Rio's most famous samba schools are officially presided by high-profile members of the \emph{jogo do bicho} elite \citep{bezerra2009mecenato,cavalcanti2006carnaval,misse2011crime,queiroz1992carnaval}.
 
-As I have mentioned in the previous section, the animal game at times faced opposition by the local population.
+The animal game at times faced opposition by the local population.
 The public often perceived the game as immoral and repugnant.
-Moreover, even after the bicho was well-established in Rio de Janeiro, the transition from a competitive betting market to an oligopoly involved the threat and often the use of physical violence against bookmakers who resisted the change \citetext{\citealp[143]{bezerra2009mecenato}, \citealp[52]{labronici2012paratodos}}.
-\emph{Bicheiros} were aware of the reputation costs their strategy entailed.
-They decided to finance samba schools hoping to win `the hearts and minds' of the population and attach a more positive image of the game among urban classes.
-Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but only as individuals who had a private interest in samba \citep[209]{chinelli1993vazio}.
-In 1984, a group of rich \emph{jogo do bicho} financiers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
-The LIESA marked a shift in the Carnival.
-For the first time, \emph{bicheiros} decided to act as a group rather than individuals.
-The organisation consolidated the power of \emph{bicheiros} over the parade and provided a formal mechanism to solve disputes among the samba school patrons \citetext{\citealp[43]{cavalcanti2006carnaval}; \citealp[171]{farias2013carnival}; \citealp[55]{labronici2012paratodos}}.
+\emph{Bicheiros} were aware of that problem.
+They decided to finance samba schools hoping to win the support of the population and attach a more positive image of the game among urban classes.
+Members of the \emph{jogo do bicho} had been involved in the Carnival since the early 1920s, but in 1984, a group of rich bankers founded collectively the LIESA (\emph{Liga Independente das Escolas de Samba}, Independent League of the Samba Schools), a civil association intended to direct and sponsor the Carnival parade in Rio de Janeiro.
+The LIESA marked a shift in the history of the Carnival Parade.
+For the first time, the \emph{bicheiros} decided to act as a group rather than individuals who shared an interest in popular festivals.
+The institution was very effective in expanding the parade, but it also provided other benefits to the \textit{jogo do bicho} bankers.
+The organisation consolidated their power over the Carnival and provided a formal mechanism for the \textit{banqueiros} to solve disputes \citetext{\citealp[43]{cavalcanti2006carnaval}; \citealp[171]{farias2013carnival}; \citealp[55]{labronici2012paratodos}}.
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=.8\textwidth, height=8cm]{images/bicho07.jpg}
-	\caption{Castor de Andrade is shown celebrating after the samba school he sponsored, Mocidade Independente, won the Carnival parade in 1996.
-He was surrounded by colleagues and police officers.
-Andrade was the founding president of LIESA (1984--1985) and the wealthiest \emph{bicheiro} of Rio de Janeiro at the time.
-Source: Folha Imagem.
-Reproduced in \citet[139]{misse2007illegal}.}
-	\label{fig:castor}
-\end{figure}
-
-The funding of the samba schools had an indirect effect to the animal game.
+The funding of the samba schools had an indirect effect on the animal game.
 The patronage also reduced agent-principal problems within the \emph{jogo do bicho}.
 \emph{Bicheiros} donate to samba school to gather support of the communities, and by doing so they gain access to local information on their business.
-Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring is cost-effective for animal game managers.
+Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring the cost-effectiveness for animal game managers.
 Thus, street bookmakers have fewer incentives to cheat.
-In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}'s donations \citep{bbc2012aniz}.
-Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation.
-The patronage can be interpreted as an illegal version of `profit-sharing', a mechanism which has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.
+In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}' donations \citep{bbc2012aniz}.
+Hence, funds donated to samba schools and other charitable organisations help align the interests of different members of the \emph{jogo do bicho} organisation.
+The patronage can be interpreted as an illegal version of ``profit-sharing,'' a mechanism that has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.
 
-The samba schools have profited from this association too.
+Samba schools have profited from this association too.
 First, they have gained autonomy from the government.
-The samba schools do not need to rely exclusively on public funds to organise the parade, and money from the \emph{jogo do bicho} permitted the schools to act independently \citep[209]{chinelli1993vazio}.
+Samba schools do not need to rely exclusively on public funds to organise the parade, and money from the \emph{jogo do bicho} has permitted the schools to act independently \citep[209]{chinelli1993vazio}.
 Second, the support of the \emph{jogo do bicho} has increased the political and social clout of the samba schools.
-In a country where the state is not present throughout the territory and human right abuses are frequent \citep{ahnen2003,odonnell1993state,pinheiro2000,pinheiro2001}, \emph{jogo do bicho} bankers, and more recently drug traffickers, have provided private governance to poor areas of Rio de Janeiro by enforcing property rights, mediating disputes, and preventing police abuse in the favelas \citep{arias2006dynamics,goldstein2013laughter,leeds1996cocaine}.
+In a country where the state is not present throughout the territory and human right abuses are frequent \citep{odonnell1993state,pinheiro2000}, \emph{jogo do bicho} bankers, and more recently drug traffickers, have provided private governance to poor areas of Rio de Janeiro by enforcing property rights, mediating disputes, and preventing police abuse in the favelas \citep{arias2006dynamics,goldstein2013laughter}.
 In return for funds and protection from the \emph{bicheiros}, samba schools have served as intermediaries between the underworld and the political system.
 Although the \emph{banqueiros} are interested in weak law enforcement against the animal game, politicians have resorted to samba schools to contact \emph{bicheiros} and use their financial and electoral influence in the shanty towns \citep[17]{misse2011crime}.
-The samba schools, therefore, have increased their bargaining power in the political sphere and extended their reach within Rio's poor communities \citep[215]{chinelli1993vazio}.
+The samba schools, therefore, have increased their bargaining power in the political sphere and have extended their reach within Rio's poor communities \citep[215]{chinelli1993vazio}.
 
 \subsection{Political Support}
+\label{sub:support}
 
-If politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, their relationship with the animal game bankers have become more ambivalent in the last decades.
-The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citetext{\citealp{gaspari2002ditadura}; \citealp{jupiara2015poroes}; \citealp[39]{zaluar2007democratizaccao}}.
-Given the absence of democratic checks and balances, paramilitaries and police forces colluded to repress potential dissidents of the regime and, frequently, to extort civilians \citep{gorender1999combate,magalhaes1997logica,misse2009acumulaccao,skidmore1990politics}.
-\emph{Bicheiros} saw the corruption of some members of the military as an opportunity.
-Wealthy \emph{jogo do bicho} bankers hired rogue police officers not only to work as security guards but to threaten eventual competitors in their regions of influence.
-The agreement between \emph{bicheiros} and corrupt members of the military was the ultimate responsible for the transformation of the \emph{jogo do bicho} into a `coercive oligopoly' \citep{jupiara2015poroes}.
-The support of the armed forces meant that new groups would be prohibited from entering the market and that the illegal lottery could operate undisturbed by the government.
+Politicians were opposed to the \emph{jogo do bicho} in the early twentieth century, but their relationship with the animal game bankers later became more ambivalent.
+The collaboration between public authorities and \emph{bicheiros} gained prominence during the military dictatorship (1964--1985) \citep{gaspari2002ditadura,jupiara2015poroes}.
+The regime effectively dismantled the few checks and balances implemented in the Second Republic (1945--1964), and paramilitaries and police forces had considerable discretion to repress political dissidents.
+Extortion of civilians was also widespread \citep{magalhaes1997logica,misse2009acumulaccao}.
+But \emph{bicheiros} saw the corruption of some members of the military as an opportunity to increase profits.
+Wealthy \emph{jogo do bicho} bankers hired rogue police officers to work as security guards and to threaten eventual competitors in their regions of influence.
+The agreement between \emph{bicheiros} and corrupt members of the military was ultimately responsible for the transformation of the \emph{jogo do bicho} into a coercive oligopoly \citep{jupiara2015poroes}.
+When the \textit{jogo} transitioned from an informal to an illegal market, the use of violence in the game became more widespread.
+The support of the armed forces meant new groups would be prohibited from entering the market and the illegal lottery could operate undisturbed by the government.
 
@@
 In the military regime, government officials were mainly interested in bribes from the animal game.
-But in the democratic period, votes became a sought-after political resource.
-\emph{Bicheiros} are important in this sense as they have direct influence over a number of poor communities.
+But in the democratic period, votes were the most sought-after political resource.
+\emph{Bicheiros} are important in this sense as they have direct influence over a number of poor communities, either because of their role as patrons, or as reliable sources of governance.
 Their patronage networks ensure that candidates supported by \emph{bicheiros} receive a substantial amount of votes from areas where campaigning is too difficult or too costly \citep[17]{misse2011crime}.
 
-The Brazilian political system is particularly conductive to clientelistic practices.
+Politicians from all spheres of government are involved with \textit{jogo do bicho} bankers.
+Recent investigations have shown that from local representatives to senators, politicians of every level receive illegal money to fund their campaigns.
+Carlinhos Cachoeira, a famous animal game banker from the state of Goiás established a large patronage network that included mayors, deputies, senators, judges, and businessmen, many of them linked to the federal government.
+His brother Marcos ironically noted that Carlinhos was ``too dedicated to politics'' and he illegally donated about USD 300 million to political candidates in his home state \citep{estadao2012gene}.
+
+The Brazilian political system is particularly conductive to client practices.
 Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}.
-In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp{mainwaring1992brazilian}; \citealp[483]{samuels2000ambition}}.
+In addition, Brazil uses an open-list proportional representation electoral system; each of the 27 states of the federation are considered at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}.
 These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence.
-Nevertheless, that independence means candidates rely mostly on themselves to raise funds and establish communication with potential voters.
+Nevertheless, that independence means candidates rely mostly on themselves to raise funds and mobilise potential voters.
 Hence, political campaigns in Brazil tend to be expensive and personality-centred.
@@
 The support from the \emph{jogo do bicho} mitigates both problems.
-With respect to the financial costs of campaigns, illegal donations from \emph{bicheiros} help to cover advertising expenses while having the additional benefit of not appearing in the official records of the candidates \citep{congressoemfoco2007bicho,gazetadopovo2007bicho,globo2012bicheiro}.
+With respect to the financial costs of campaigns, illegal donations from \emph{bicheiros} help to cover advertising expenses while having the additional benefit of not appearing in the official records of the candidates \citep{congressoemfoco2007bicho,globo2012bicheiro}.
 This suggests that \emph{jogo do bicho}-funded politicians can circumvent spending limits and have an electoral advantage over their competitors.
 As candidates do not know whether their competitors receive funding from the \emph{jogo do bicho} nor the amount each one was paid, their dominant position is to contact the \emph{bicheiros} and join their networks.
-The situation is a prisoner's dilemma in which candidates would be better off by running cheaper campaigns and not being dependent of \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching an optimal solution.
+The situation is a prisoner's dilemma in which candidates would be better off running cheaper campaigns and not being dependent on \emph{jogo do bicho} bankers, but asymmetric information prevents them from reaching a cheaper solution.
 
-\begin{figure}[!htbp]
-	\centering
-	\includegraphics[width=.6\textwidth, height=8cm]{images/bicho08.jpg}
-	\caption{Political advertising for Abraãozinho David (right), nephew of the \emph{jogo do bicho} banker Aniz Abraão David (third from left to right).
-The banner reads: `The Candidates of the Abraão Family: Fighting for the People is a Family Heritage'.
-Source: \citet{extra2012aniz}.}
-	\label{fig:aniz}
-\end{figure}
-
-The votes from poor communities are instrumental for aspiring politicians.
+Votes from poor communities are instrumental for aspiring politicians.
 Brazil has an enforced compulsory voting system; therefore, turnout rates tend to be higher than in other democracies.
 Consequently, votes have high marginal utility for politicians.
-As elections may be decided by a small difference, the \emph{bicheiros}' clientelistic ties guarantee a minimum number of votes that politicians can rely upon on election day.
-Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process per se.
-Individuals may be punished if the candidate does not receive the expected number of votes, and are often compelled to vote for politicians that have only loose connections with their communities.
-Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil’s lower classes.
+As elections may be decided by a small difference, the \emph{bicheiros}' client ties guarantee a minimum number of votes that politicians can rely upon on election day.
+Nonetheless, the patronage subverts the preferences of the public and, as such, the democratic process \textit{per se}.
+Individuals may be punished if the candidate does not receive the expected number of votes, and they are often compelled to vote for politicians who have only loose connections with their communities.
+Therefore, although voters have the right to choose their representatives, in practice the suffrage is limited for a share of Brazil's lower classes.
 
-Finally, the \emph{jogo do bicho} patronage highlights a crucial social dilemma within the Brazilian public law.
-Even though federal judges have prosecuted \emph{bicheiros}, politicians and police forces have no incentives to enforce the punishment.
-Although Brazilian judges enjoy job stability, the latter groups constantly require local-level support from the \emph{bicheiros}.
-Politicians and police officers may have accurate information on \emph{jogo do bicho} operations and \emph{bicheiros}' whereabouts, but the federal government cannot rely upon their cooperation.
-That can be one of the reasons why even after many attempts to arrest \emph{bicheiros}, there has been little progress in that regard in Brazil's latest democratic period (1985--present).
+In a nutshell, the difficulty in permanently outlawing the \textit{jogo} showcases how the Brazilian state itself is deeply embedded within criminal sectors of the society.
+While some sectors of the Brazilian bureaucracy may be described as ``islands of excellence'' and closely resemble the Weberian ideal of public administration \citep{bersch2017state}, local politicians remain dependent of unstable, and often unlawful, connections with social elites.
+This system of ``partial embedded autonomy'' \citep{evans1995embedded} provides the required stability for the political system to operate, yet it offers significant opportunities for rent-seeking behaviour.
+This structure tends to perpetuate itself as it brings benefits for both the animal game bankers and local politicians, such as limited competition and the ability to extract resources from poor voters.
 
-\section{Conclusion}
-\label{sec:conclusion}
+\section{Concluding Remarks}
+\label{sec:conclusion3}
 
-Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}.
-Yet, the \emph{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
-\emph{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers.
-Furthermore, by investing in the Carnival parade \emph{bicheiros} have been able to gather popular and government support.
-Poor communities have associated with the \emph{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \emph{jogo do bicho}'s networks can provide.
+Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and to establish close ties with the population.
+Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years.
+\textit{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers.
+Furthermore, by investing in the Carnival parade, \textit{bicheiros} have been able to gather popular and government support.
+Poor communities have associated with the \textit{bicheiros} to receive welfare provisions, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.
 
-Nevertheless, the \emph{jogo do bicho} has also created negative externalities.
+Nevertheless, the \textit{jogo do bicho} has also created negative externalities.
 Violence is used to punish defectors and to constrain competitors.
-The clientelistic relationship that \emph{bicheiros} have with local politicians have lead to sub-optimal outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
+The client relationship \textit{bicheiros} have with local politicians has lead to undemocratic outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations.
 These negative externalities have long-term effects and still impact the Brazilian public sphere.
 
-Although the \emph{jogo do bicho} has received an increasing attention from scholars, much of its inner workings remain poorly understood.
-First, the relationship between \emph{bicheiros} and drug dealers is a topic that deserves attention.
+Although the \textit{jogo do bicho} has received increasing attention from scholars, much of its inner workings remain poorly understood.
+First, the relationship between \textit{bicheiros} and drug dealers is a topic that deserves attention.
 Brazil has become one of the world's largest consumers of illicit drugs and South America's principal drug trafficking transit route \citep{miraglia2015drugs,misse2011crime}.
-The question whether \emph{bicheiros} collaborated or opposed the emergent drug dealing business is still unclear.
-Second, the extent to which \emph{bicheiros} use other businesses, such as hotels or factories, to laundry money has been mentioned by members of the Brazilian judiciary \citep{globo2012bicheiro,globo2015cacaniquel}; however, there is no reliable estimate on its size.
-Lastly, more research is required to clarify how \emph{bicheiros} from different parts of Brazil coordinate their activities and prevent large-scale conflicts.
-Cases studies are usually focused on Rio de Janeiro's \emph{bicheiros}, but scholars would benefit from comparative analyses with a larger number of states.
-This is an important step to elucidate how \emph{bicheiros} continue to influence politics and the public across Brazil.
+The question whether \textit{bicheiros} collaborated or opposed the emergent drug dealing business is still unclear.
+Second, the extent to which \textit{bicheiros} use other businesses, such as hotels or factories, to launder money has been mentioned by members of the Brazilian judiciary \citep{globo2012bicheiro,globo2015cacaniquel}; however, there is no reliable estimate on its size.
+Lastly, more research is required to clarify how \textit{bicheiros} from different parts of Brazil coordinate their activities and prevent large-scale conflicts.
+Cases studies are usually focused on Rio de Janeiro's \textit{bicheiros}, but scholars would benefit from comparative analyses with a larger number of states.
+This is an important step to elucidate how \textit{bicheiros} continue to influence politics and the public throughout Brazil.

### 2018-11-29 paper-killings/appendix/appendix.tex
add killings

--- 2018-05-07 chapters/04-killings.tex
@@
-\chapter{What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models}
-\label{chap:killings}
+\documentclass[a4paper,12pt]{article}
+\usepackage[utf8]{inputenc}
+\usepackage{rotating}
+\usepackage{lscape}
+\usepackage{amssymb,amsmath,amssymb}
+\usepackage[stable]{footmisc}
+\usepackage{lmodern}
+\usepackage[T1]{fontenc}
+\usepackage[ttscale=.875]{libertine}
+\usepackage[libertine]{newtxmath}
+\usepackage[authoryear]{natbib}
+\usepackage{babelbib}
+\usepackage{rotating}
+\usepackage{adjustbox}
+\usepackage[usenames,dvipsnames]{xcolor}
+\definecolor{darkblue}{rgb}{0.0,0.0,0.55}
+\setcitestyle{aysep={}} 
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd{\NAT@citex}
+  {\@citea\NAT@hyper@{%
+	 \NAT@nmfmt{\NAT@nm}%
+	 \hyper@natlinkbreak{\NAT@aysep\NAT@spacechar}{\@citeb\@extra@b@citeb}%
+	 \NAT@date}}
+  {\@citea\NAT@nmfmt{\NAT@nm}%
+   \NAT@aysep\NAT@spacechar\NAT@hyper@{\NAT@date}}{}{}
+\patchcmd{\NAT@citex}
+  {\@citea\NAT@hyper@{%
+	 \NAT@nmfmt{\NAT@nm}%
+	 \hyper@natlinkbreak{\NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi}%
+   {\@citeb\@extra@b@citeb}%
+	 \NAT@date}}
+  {\@citea\NAT@nmfmt{\NAT@nm}%
+   \NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi\NAT@hyper@{\NAT@date}}
+  {}{}
+\makeatother
+\usepackage{setspace}
+\usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
+\usepackage{hyperref}
+\usepackage{graphicx}
+\usepackage{dcolumn}
+\usepackage{float}
+\usepackage{pgf}
+\usepackage{tikz}
+\usetikzlibrary{arrows}
+\usetikzlibrary{positioning}
+\usepackage{mathtools}
+\usepackage{caption}
+\usepackage[UKenglish]{babel}
+\usepackage[UKenglish]{isodate}
+\cleanlookdateon
+\exhyphenpenalty=1000
+\hyphenpenalty=1000
+\widowpenalty=10000
+\clubpenalty=1000
+\onehalfspacing
+
+\hypersetup{pdftitle={What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models},
+		pdfauthor={Danilo Freire and Gary Uzonyi},
+		pdfborder={0 0 0},
+		breaklinks=true,
+		linkcolor=Mahogany,
+		citecolor=Mahogany,
+		urlcolor=darkblue,
+		colorlinks=true}
+
+\title{Appendix for \textit{What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models}\thanks{Freire: PhD candidate, Department of Political Economy, King's College London.
+Email address: \href{mailto:danilofreire@gmail.com}{\texttt{danilofreire@gmail.com}}.
+Uzonyi: Assistant Professor, Department of Political Science; Research Fellow, Howard H.
+Baker Jr.
+Center for Public Policy, University of Tennessee.
+Email address: \href{mailto:guzonyi@utk.edu}{\texttt{guzonyi@utk.edu}}.
+All replication materials are available at \href{https://github.com/danilofreire/mass-killings}{\texttt{https://github.com/danilofreire/mass-killings}}.}}
+
+\author{Danilo Freire \and Gary Uzonyi}
+
+\date{\today}
+
+\begin{document}
+\maketitle
+
+{\hypersetup{linkcolor=black}
+\tableofcontents
+}
+
+\newpage
 
 \section{Introduction}
-\label{sec:intro4}
-
-Since the end of World War II, mass killings, genocides, and politicides have claimed over 34.5 million lives \citep{marshall2017pitf}.\footnote{Genocide and politicide are the attempted intentional destruction of communal or political groups, respectively \citep[see][]{harff1988toward}.
-Mass killing includes these atrocities, as well as attacks against civilians that result in at least 1,000 deaths but are not intended to destroy a particular group \citep[see][]{ulfelder2008assessing}.
-While some conflate the logic of these types of atrocities \citep[e.g.,][]{stanton2015regulating, valentino2004draining}, others claim genocide and politicide follow a different logic from other forms of government violence \citep{kalyvas2006logic,stanton2015regulating}.
-Note that these arguments focus on the logic of the violence, not legal definitions that separate and prioritise genocide as the crime of crimes \citep[see][]{schabas2000genocide}.} The international community has responded with an effort to prevent further state-sponsored mass murder by strengthening laws against crimes against humanity, genocide, and war crimes.
-It formed \textit{ad hoc}, hybrid, and a permanent criminal court to punish and deter future atrocities.
-Furthermore, the United Nations established a Special Adviser on the Prevention of Genocide and recognised its members’ responsibility to protect civilian populations within and outside their own borders.
-Yet, such atrocities still occur.
-Recently, President al-Assad of Syria has massacred tens of thousands of civilians during the Syrian Civil War \citep{goldman2017nyt}.
-Similarly, South Sudan's President Kiir is actively starving and killing civilians from dissident and rival tribes \citep{nichols2017reuters}.
-While there is some evidence that such atrocities may be declining since the Cold War \citep{valentino2014we}, the international community has been far from successful in realising slogans like ``Never Again'' and ``Not on My Watch'' \citep{cheadle2007not}.
-
-Ultimately, successful prevention requires us to understand why these atrocities occur.
-In this vein, the academic community has laboured tremendously to establish evidence-based theories as to why governments engage in brutality against their civilian populations.
-Indeed, since 1995, there have been over 45 quantitative political science articles focused on explaining government-sponsored killing of civilians.
-Note, this number does not consider the myriad other books, qualitative articles, and non-political science research conducted into this matter.
-Overall, the mass violence literature generally agrees that as threat increases, so does the likelihood of atrocity, if the costs to such violence are not prohibitive.
-However, there is little consensus on what factors influence the level of threat or costs a regime faces.
-Part of the reason for this uncertainty is that scholars use very different model specifications when testing their arguments.
-Examining the quantitative literature on government violence against civilians, we found that in 45 studies, scholars used nearly 180 measurements to capture roughly 30 key concepts related to threat and costs.
-While models should be constructed specifically to test particular arguments, one concern is that small changes in model specification could in influence the robustness of empirical results and the inferences that we can draw from these results.
-
-To overcome these limitations and provide a better understanding of government atrocity, we employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities.
-Our approach is similar to \cite{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but we provide additional tests to check which variables are able to predict out-of-sample cases of mass violence during both peacetime and wartime.
-In conducting this analysis, we address three debates in the mass violence literature:
-
-\begin{enumerate}
-    \item Why do some governments engage in mass killings, genocides, or politicides?
-This is the primary question asked by advocates, policymakers, and scholars in this field of research.
-It is also the question that the 45 studies mentioned above each attempt to answer.
-    \item Does the logic underpinning government decision-making follow different patterns during peacetime and wartime.
-Recent research suggests that government atrocity occurs predominantly during periods of civil war \citep{harff2003no} which has led some scholars to restrict their analyses to only periods of civil war \citep[e.g.,][]{colaresi2008kill, valentino2004draining} or concentrate on predicting both the onset of civil war and atrocity \citep{goldsmith2013forecasting}.
-Yet, others estimate models of all country-year \citep[e.g.,][]{krain1997state, montalvo2008discrete}, raising questions of how well these studies speak to each other.
-    \item Is there a difference in logic between those atrocities labelled as genocide or politicide, compared to other mass killings?
-While the Political Instability Task Force \citep{marshall2017pitf} provides the most widely used data on government atrocity, these data are limited only to cases of genocide and politicide.
-Others provide data with much more lenient inclusion criteria \citep[e.g.,][]{eck2007one, rummel1995democracy, stanton2015regulating, ulfelder2012forecasting}.
-These differences in definition of atrocity have led to divergent results, raising questions about important determinants of government behaviour \citep[for discussion, see][]{uzonyi2016domestic, wayman2010explaining}.
-\end{enumerate}
-
-Our analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
-We find that GDP per capita, stable political regimes, and the post-Cold War period are negatively associated with mass killings.
-Conversely, previous political turmoil, ethnic diversity, and civil wars increase the risk of such violence.
-The distributed random forest algorithm indicates that years since the last episode of mass violence, GDP per capita, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to our models' out-of-sample predictive power.
-While we discuss the importance of these findings in more detail later in this article, a few notes are imperative to make here.
-First, in some ways, these findings confirm previous research–unstable countries are more likely to witness the regime employ atrocity \citep[e.g.,]{goldsmith2013forecasting,harff2003no,krain1997state}.
-However, many of the factors scholars often cite as observable indicators of such instability--regime transitions, coups d'état, the presence of militias, etc.--are not good proxies for instability.
-Thus, policymakers may be looking for the incorrect signs of impending atrocity when seeking to prevent its onset.
-Second, these findings raise concerns about policy options for preventing violence against civilians.
-If our conclusion is that unstable countries are violent, then preventing atrocity likely requires significant investments of time and resources in state-building, which is often politically and practically unfeasible \citep{doyle2006making}.
-Lastly, our findings are robust to a myriad of specifications, functional forms, and additional tests as we describe below.
-This analysis contributes significantly to the political violence literature by highlighting the parsimonious nature of the logic behind government atrocity and clearing away much of the empirical clutter surrounding this conclusion.
-
-\section{Empirical Methods}
-\label{sec:methods4}
-
-We employ two methods to test the robustness of the potential determinants of state-led violence.
-First, we use a variant of extreme bounds analysis (EBA) to check the statistical significance of 35 explanatory variables cited in the academic literature.\footnote{The complete list of variables included in this article is available in appendix \ref{sec:mk-appendix}.
-Our selection procedure was as follows.
-We included all variables that appeared in at least two two quantitative papers on mass killings which employed logit or probit models in a global sample.
-We considered only published articles in our list.} Researchers have employed the EBA to assess the sensitivity of the determinants of civil war \citep{hegre2006sensitivity}, coups d'état \citep{gassebner2016expect}, democratisation \citep{gassebner2013extreme}, economic growth \citep{levine1992sensitivity, sala1997just, sturm2005determinants}, nuclear deterrence \citep{bell2015examining}, political repression \citep{hafner2005right}, and state corruption \citep{serra2006empirical}.
-The method is particularly useful when there is no consensus about which covariates belong in the ``true'' regression model \citep[178]{sala1997just} and scholars worry that omitted or unnecessary predictors could bias the parameter estimates \citep[60]{angrist2008mostly, clarke2005phantom, elwert2014endogenous, spector2011methodological}.
-By combining a large set of regression models into a single posterior distribution, the EBA reduces the uncertainty associated with the \textit{ad hoc} selection of predictors and provides an intuitive way of assessing the relative strength of the statistical findings \citep{leamer1985sensitivity, sala1997just}.
-
-Second, we employ distributed random forest (DRF) \citep{breiman2001random, h2o2017} to measure the predictive power of our set of independent variables.
-Random forest has been widely used in the machine learning community and consistently ranks among the best algorithms for predicting structured data.\footnote{Random forests, gradient boosting machines, and neural networks have won the highest number of competitions on Kaggle, a crowd-sourced platform for predictive modelling \citep{carpenter2011may}.
-While data analysts prefer neural networks for problems with a high number of hidden features such as image recognition, random forest and gradient boosting are the top choices for dealing with tabular data.
-See: \href{https://www.kaggle.com/antgoldbloom/what-algorithms-are-most-successful-on-kaggle}{https://www.kaggle.com/antgoldbloom/what-algorithms-are-most-successful-on-kaggle} (access: December 2017).} DRF does not require distribution assumptions, can be used with any type of response variables, and are able to automatically detect nonlinear relationships between correlates \citep{fernandez2014we, hill2014empirical, jones2015exploratory, muchlinski2015comparing}.
-Moreover, unlike neural networks and other deep learning algorithms \citep{castelvecchi2016can,rojas2013neural,shwartz2017opening}, the output of DRF models can be interpreted meaningfully.
-DRF offers precise estimates of the predictive ability of each individual variable, and it also allows researchers to graphically visualise the marginal effect of the covariates on the response variable \citep{friedman2001greedy,friedman2001elements,goldstein2015peeking}.
-
-\subsection{Extreme Bounds Analysis}
-\label{subsec:eba}
-
-The main purpose of the extreme bounds analysis is to estimate the distribution of coefficients of each predictor $x$ in an exhaustive combination of regression models with $y$ as a dependent variable.
-\cite[308]{leamer1985sensitivity} argued that the EBA could certify scholars that ``minor changes in the list of variables do not alter fundamentally the conclusions, nor does a slight re-weighting of observations, nor correction for dependence among observations, etcetera, etcetera.'' He proposed that ``sturdy'' variables are those whose minimum and maximum of their coefficient distribution have the same sign and are situated at a distance from zero.
-If we are to use the conventional value of $p < 0.05$, the mean of the variable coefficients' distribution should be located at least $1.96$ standard deviations away from zero.
-
-Leamer's criterion is intuitive, but other authors contend it is too strict for most social science applications.
-\cite{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response.
-If a given variable of interest appears as both positive and negative in the literature, it is likely that after a large a number of regressions the predictor will change signs and therefore be deemed fragile \citep[179]{sala1997just}.
-Sala-i-Martin then offered a less stringent version of the EBA in which researchers analyse the full distribution of point estimates instead of relying only on the coefficients' extreme bounds.
-
-In this paper, we follow his advice and consider the whole range of values of $CDF(0)$.
-We choose to use the whole distribution because the aggregate $CDF(0)$ allows us to move away from a binary indicator of robustness, the one we obtain with Leamer's extreme bounds, and present the estimations with their appropriate degrees of confidence \citep[179]{sala1997just}\footnote{As \citet[179]{sala1997just} describes, ``if 95 percent of the density function for the estimates of $\beta_{1}$ lies to the right of zero and only 52 percent of the density function for $\beta_{2}$  lies to the right of zero, one will probably think of variable 1 as being more likely to be correlated with [a dependent variable $Y$] than variable 2.''}.
-Our main focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero.
-We do not assume that the CDF has a normal distribution: Many of our coefficients do not follow a standard distribution, so the generic model provides a better fit to our data.\footnote{Histograms for all coefficients are available in appendix \ref{sec:mk-appendix}.} We specify our models as follows:
-
-\begin{equation}
-\text{\textit{Mass Killing Onset}}_{it} = \beta_{M}M_{it} + \beta_{F}F_{it} + \beta_{Z}Z_{it} + v_{it}
-\end{equation}
-
-Our main dependent variable is \textit{Mass Killing Onset}, which denotes the onset of government-sponsored killings.
-It was coded by \cite{ulfelder2008assessing}.
-The authors define a mass killing as ``any event in which the actions of state agents result in the intentional death of at least 1,000 noncombatants from a discrete group in a period of sustained violence'' \citep[2]{ulfelder2008assessing}.
-In our additional tests, we also estimate the models with an indicator of Genocide/Politicide Onset by \cite{harff2003no}.
-We prefer the first measure not only because it provides a clear numerical threshold that separates mass killings from low-level violence, but also because it focuses on the killing of members of a given community, not on the absolute number of deaths.
-
-The indices $i$ and $t$ indicate country and year, respectively.
-$M$ is a set of three covariates that are included in every model due to their prominence in the literature \citep{levine1992vale}.
-In our analysis, we have added the natural logarithm of the GDP per capita to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence.
-$F$ denotes a vector of variables of interest, and $Z$ is a vector of other control variables in addition to those included in $M$.
-Our first model includes the variable of interest, 3 additional control variables, and 3 covariates that appear in every regression.
-$v$ is the error term.
-In practice, however, since we are interested in the effect of all variables in the data set and we do not have true control variables except from $M$, in our case $F$ and $Z$ are interchangeable.
-Following \citet[514]{hegre2006sensitivity}, the independent variables have been lagged one year to reduce the risk of endogeneity.
-Although our dependent variable is dichotomous, we use linear probability models in our main analysis.
-As argued by \citet[298]{gassebner2016expect}, linear probability models are less prone to convergence problems, their estimation is faster, and their results can be readily interpreted.
-While we do employ logistic and probit models as robustness tests, we report the coefficients of the linear probability models as our primary estimates.
-Moreover, since the data are grouped into countries, we use cluster-robust standard errors to control for within-cluster error correlation.
-
-One of the shortcomings of our data is that some variables are strongly correlated.
-For instance, we include the Correlates of War's Composite Index of National Capability (\textit{CINC}), which is composed by six different variables \citep{cow2017cinc,singer1988reconstructing}.\footnote{For more information about the CINC index, please refer to the Correlates of War website: \href{http://www.correlatesofwar.org/data-sets/national-material-capabilities/national-material-capabilities-v4-0}{http://www.correlatesofwar.org/data-sets/national-material-capabilities/national-material-capabilities-v4-0} (access: December 2017).} Multicollinearity would induce bias to our estimates, so we add CINC and three of its components\footnote{The variables correspond to the ratio between the national and the global values.} -- military expenditure, military personnel, and total population -- as mutually exclusive variables.
-These predictors are not included in the same regression model to avoid collinearity issues.
-
-We add another set of mutually exclusive variables to reduce collinearity.
-There are three covariates that denote ongoing civil conflicts: one measured by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, another coded by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}.
-We include only one of these measures at a time.
-
-As a last precaution against collinearity, we place a limit on the Variance Inflation Factor (\textit{VIF}) of all regression coefficients.
-The VIF estimates how much of the variance of each predictor is dependent on the other covariates in a model.
-A VIF of 1 indicates that the predictor is uncorrelated with the remaining covariates.
-The VIF limits are often arbitrary \citep{bell2015examining,o2007caution}, thus we use a moderately conservative VIF of 7 in our estimates.
-As robustness tests, we run the same models with different VIF cut-offs and without restriction.
-
-Two variables were omitted from the EBA models but included in the machine learning models below.
-The first is \textit{democracy}, a dummy variable that indicates whether the country is a has a Polity IV score equal or higher than 5.
-The second is \textit{interstate war}, a binary covariate measuring if the country is at war in a given year \citep{sarkees2010resort}.
-We have decided to omit democracy because of its obvious correlation with the Polity measure and interstate war due to its correlation with our dependent variables.
-The EBA models do not converge if those variables are added.
-Since this problem does not affect machine learning algorithms, the two variables were included in the second set of estimations.
-
-Lastly, we depart slightly from Sala-i-Martin's suggested method and do not assign weights to the EBA.
-Although he recommends using goodness-of-fit measures to construct regression weights, we agree with \citet{sturm2002robust} and \citet[299]{gassebner2016expect} and use the unweighted version of the CDF instead.
-Goodness-of-fit indicators are not equivalent to the probability of a given model being true \citep{achen1977measuring,anscombe1973graphs,king1986not}, and the weights constructed this way are not invariant to transformations in the dependent variable.
-Moreover, our data set has a number of missing observations, so that model comparison would be misleading \citep{lall2016multiple}.
-Thus, the results below are for unweighted estimations.
-
-\subsection{Distributed Random Forest}
-\label{sub:drf}
-
-Random forest is a machine learning algorithm that consists of a combination of individual decision trees.
-In a classification problem, each decision tree uses a vector of covariates to split the dependent variable into two increasingly homogeneous parts \citep{breiman2001statistical}.
-However, decision trees are prone to overfitting, i.e., they match the original data set so closely that they tend to perform poorly with new data \citep{dietterich1995comparison,ho1998random}.
-Random forest, in contrast, avoids this issue by growing a decision tree only to a bootstrap sample of the original data, selecting random features at each split, then aggregating the different trees into a single prediction.
-If the independent variable is continuous, the algorithm will simply choose the average value of the predictions as the best candidate; if the covariate is discrete, the majority class will be employed.
-The simple procedure of leaving out some data points and growing separate trees with a random subset of covariates is sufficient to eliminate the risk of overfitting \citep[9-10]{jones2015exploratory}.
-
-Random forest has many desirable properties, such as ``highly accurate predictions, robustness to noise and outliers, internally unbiased estimate of the generalisation error, efficient computation, and the ability to handle large dimensions and many predictors'' \citep[7]{muchlinski2015comparing}.
-Thus, random forest allows the researcher to estimate very flexible models with minimal assumptions.
-Unlike parametric methods such as ordinary least squares or logistic regressions, the analyst does not have to impose any distributional form to the data-generating process.
-As a result, random forest is able to effectively uncover complex, nonlinear interaction effects in the data without prespecification \citep{jones2015exploratory,strobl2007bias}.
-
-The algorithm can also deal with heavily imbalanced data.
-Researchers can balance the classes of their response variable by assigning different sampling probability to each category.
-This enables the algorithm to make accurate prediction for rare events without adding prior distributions or resorting to subjective modelling choices \citep{chen2004using,del2014use,muchlinski2015comparing}.
-
-In this paper we use distributed random forest (DRF) to model our data \citep{h2o2017}.
-The DRF is essentially identical to original random forest algorithm, but it has two additional features that are useful for our purposes.
-Firstly, DRF is optimised for big data, as it grows decision trees on separate cores to speed up computation time.
-Secondly, in DRF, non-observed cases are not assumed to be missing at random, but rather as values that contain information in themselves.
-When building decision trees, DRF treats the missing observations as a separate category that can go either left or right.
-This is a more conservative approach than assuming that missing cases fit into an underlying parametric distribution.\footnote{For more information about how the distributed random forest algorithm deals with missing observations, please refer to: \href{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html}{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html} (access: December 2017).}
-
-The DRF has a series of hyperparameters that can be tuned to improve the algorithm's predictive performance.
-For instance, users can control the number of decision trees in each iteration, how deep trees should grow, which performance metric to adopt, whether to balance classes or to use cross-validation, and many other options.
-The interaction between parameters is generally complex and may involve thousands of potential combinations.
-As an example, a researcher interested in four parameters with 10 possible values each would have to estimate 10,000 models before deciding which is the most efficient one.
-Also, machine learning parameters are sensitive to the data at hand; so what seems an optimal solution for one problem cannot be readily implemented in another data set \citep{genuer2008random,goldstein2010application,jones2015exploratory}.
-
-To address this issue, we have adopted an automated procedure to select the model parameters.
-We perform a grid search where the algorithm starts with a random combination of parameters, jumps to another randomly-chosen set, and then stops after it reaches a certain threshold \citep[123]{cook2017h2o}.
-We follow the literature on predictive political science and use the area under the ROC curve (AUC) as our model evaluation metric \citep[e.g.,][]{clayton2014will,hill2014empirical,ward2010perils,ward2013learning,weidmann2010predicting}.
-We set the metric as follows: If five random models had not increased the AUC by at least in 0.1\% comparing to the previous ones, the algorithm considers the result to be optimal.
-We set the maximum number of models to 1,000.
-
-We have added several parameters to the grid search.
-The first is the number of independent trees to grow in each random forest model.
-The starting values are 256, 512, and 1024 trees.
-The machine learning literature does not provide a heuristic on how large a random forest should be, but \citet[166]{oshiro2012many} affirm that ``from 128 trees there is no more significant difference between the forests using 256, 512, 1024, 2048 and 4096 trees.'' We employ a more conservative approach and start from a higher value that the authors suggest as adding more trees do not reduce prediction accuracy \citep[7]{breiman2001statistical}.
-However, we find little evidence that the number of trees makes a noticeable difference to the estimations.
-
-The depth of each decision tree also influences the algorithm performance.
-Deeper trees indicate more complex models, and in general they provide a better fit to the data.
-Nevertheless, this complexity comes at the risk of overfitting, so deeper trees are not necessarily the most adequate solution for every model \citep[596]{friedman2001greedy,segal2004machine}.
-In this article, we let the algorithm decide among using 10, 20, or 40 levels for each tree.
-In our view, these number offer a good balance between parsimony and complexity.
-
-We test whether having balanced classes of our dependent variable (mass killing onset) affects the predictive ability of the model.
-Since the response measure is heavily imbalanced, oversampling the positive responses could potentially improve our results \citep{chawla2004special,del2014use,japkowicz2002class}.
-Somewhat unexpectedly, DRF had a better fit using unbalanced classes, which is the result we report in the next section.
-
-We also vary how many variables should be considered for each split in the data.
-The DRF's default is to use $\sqrt{p}$, where p is the number of columns in the data set.
-As we have 34 covariates of interest, we have selected 5, 6 and 7 variables per split.
-The DRF uses a majority voting procedure to select which variable is most important.
-Additionally, the algorithm chooses the percentage of the training set to be modelled by each tree.
-The default option is 63.2\%, but we include the options of using 50\% and 100\% of the data.
-Similarly, we give a range of options for choosing how many columns will be included in each tree.
-The algorithm can randomly choose among 50\%, 90\% or 100\% of the independent variables when estimating a decision tree.
-
-Finally, we use four types of histogram to find optimal split points for each independent variable.
-Decision trees consider every value of a given independent variable as a potential candidate for a split in the training data.
-This process is notably time-consuming, and computation time can be significantly reduced at little loss of precision by taking discrete values of the predictor distribution.
-The DRF algorithm offers four choices of histogram selection and we include all of them in our estimations.
-
-\section{Results}
-\label{sec:results4}
-
-Table \ref{tab:eba1} summarises our main EBA results with Ulfelder and Valentino's \citeyear{ulfelder2008assessing} \textit{Mass Killing Onset} as the dependent variable.
-The table shows the average coefficient estimate of all regressions for each robust variable along with their mean standard deviations.\footnote{A list of all independent variables and coding rules are available in the online appendix.} The table also displays the percentage of regressions that are statistically significant at the 90\% level.
-$CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{We show whichever area is the largest.
-The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is our main statistic of interest, and we consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}.
-Lastly, we report the number of estimated regressions models which included each variable.
-
-\vspace{1cm}
-
-\begin{table}[H]
-\centering
-\begin{tabular}{lrrrrr}
-\hline
-\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
-\textit{Base variables} &  &  &  &  &  \\
-Log GDP per capita & -0.0091 & 0.0052 & 76.055 & 0.9335 & 226707 \\
- &  &  &  &  &  \\
-\textit{Additional variables} &  &  &  &  &  \\
-Post-Cold War years & -0.0133 & 0.0085 & 72.845 & 0.9472 & 35614 \\
-UCDP civil war onset & 0.0529 & 0.0321 & 52.378 & 0.9441 & 20854 \\
-Previous riots & 0.0140 & 0.0100 & 56.242 & 0.9216 & 35614 \\
-UCDP ongoing civil war & 0.0172 & 0.0115 & 65.652 & 0.9092 & 20854 \\
-Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.674 & 0.9050 & 35614 \\
-Polity IV squared & -0.0002 & 0.0001 & 61.206 & 0.9031 & 35614 \\ \hline
-\end{tabular}
-\caption{Extreme Bounds Analysis -- Mass Killings (Robust Variables Only)}
-\label{tab:eba1}
-\end{table}
-
-Seven variables pass our EBA criteria.
-Three variables decrease the likelihood of mass killings.
-First, as widely suggested in the literature, the natural logarithm of GDP per capita is robustly associated with the onset of mass killings
-\citep[e.g.,][]{besanccon2005relative, easterly2006development,esteban2015strategic}.
-In our models, GDP per capita has a negative effect on onset risk is statistically significant in 76\% of the models.
-Second, the post-Cold War years are correlated with lower levels of government violence.
-Despite eventual outbreaks of large-scale attacks against civilians \citep[e.g.,][]{prunier2005darfur,tatum2010genocide,waldorf2009revisiting}, after 1991 these events have become less prevalent.
-Indeed, this finding is in line with several studies that point to a general decline in violence over the last decades, including riots, civil wars, and urban crime \citep{eisner2003long,pinker2011better}.
-The third variable negatively correlated with mass killings is the squared term of the Polity IV political regime index.
-Although the linear Polity IV indicator does not reach our threshold for significance, about 90\% of the distribution area of Polity IV squared is located below zero.
-This finding points to a nonlinear relationship between political regime and mass killings, thus providing further evidence that democracy reduces state-sponsored violence \citep{rost2013will,rummel1995democracy} and that regimes that mix democratic with autocratic features have the highest risk of conflict \citep{hegre2001toward,mitchell2013domestic,muchlinski2014grievances,regan2010changing}.
-
-Four variables are positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of state-sponsored violence.
-The onset and continuation of civil wars are correlated with mass killings, but only when we employ the UCDP measures of violent conflict.
-We find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}.
-Coding choices likely account for this difference, and the result is robust to a myriad of specifications.
-Former instances of political turmoil also have a positive coefficient in our models.
-Countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent
-\citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}.
-Our results also show that a higher levels of ethnic diversity increase the likelihood of atrocities against civilians.
-Nevertheless, the variable does not pass all additional tests we implement below and the sturdiness of this finding remains open to question.
-The uncertainty about the impact of ELF on mass killings is also reflected in the literature.
-As \citet[237]{hoeffler2016development} points out, \citet{rummel1995democracy} reports that ELF is not statistically significant in his models, \citet{wood2014opportunities} finds a negative relationship and \citet{querido2009state} finds a positive one.
-Our analysis does not offer a definitive answer for this question either.
-
-A different set of variables reach significance if we only consider mass killings that occur during civil wars.
-We ran the models on three indicators of violent armed conflict.
-The first is the civil war measure by the Uppsala Conflict Database Program  (\citeyear{allansson2017organized,gleditsch2002armed}).
-Here, two variables meet our criteria for robustness, Post-Cold War years and conflicts with territorial aims, both negatively correlated with mass killing onset.
-When we limit our analysis to cases marked as civil wars by the Correlates of War project \citep{sarkees2010resort}, years since the last mass killing episode, previous riots, former human rights abuses, and ethnic diversity appear have a positive effect on the outcome.
-Rather surprisingly, the presence of militias has a negative effect on genocides and politicides.
-Finally, our model using ethnic civil war \citep{cederman2010ethnic} shows similar results: Militias have a negative impact on the likelihood of mass killings during ethnic conflict, as do conflicts over territorial aims.
-This finding is in contradiction with previous research \citep{koren2017means} and is deserving of further exploration.
-Overall, these results suggest two takeaways.
-First, the determinants of mass killing onset during peacetime and wartime are different from each other.
-While there are more restraints on atrocities during peacetime, these restraints fail once fighting begins.
-This raises particularly interesting questions about the pacifying nature of democracy.
-Second, how civil wars are classified significantly alters the results produced.
-This raises concerns for scholars about making claims based on only one data set.
-
-\vspace{1cm}
-
-\begin{table}[H]
-\centering
-\begin{tabular}{lrrrrr}
-\hline
-\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
-\textit{UCDP data} &  &  &  &  &  \\
-Territory aims & -0.044 & 0.019 & 74.997 & 0.9804 & 17902 \\
-Post-Cold War years & -0.038 & 0.019 & 66.574 & 0.9222 & 17902 \\
- &  &  &  &  &  \\
-\textit{COW data} &  &  &  &  &  \\
-Physical integrity & 0.024 & 0.013 & 66.674 & 0.9564 & 17902 \\
-Militias & -0.099 & 0.048 & 73.104 & 0.9490 & 17902 \\
-Years since last mass killing & 0.006 & 0.002 & 88.208 & 0.9472 & 101583 \\
-Previous riots & 0.078 & 0.041 & 65.412 & 0.9348 & 17902 \\
-Ethnic diversity (ELF) & 0.095 & 0.062 & 48.615 & 0.9000 & 17902 \\
- &  &  &  &  &  \\
-\textit{Cederman et al. data} &  &  &  &  &  \\
-Territory aims & -0.051 & 0.026 & 74.288 & 0.9167 & 17902 \\
-Militias & -0.050 & 0.035 & 52.240 & 0.9101 & 17902 \\ \hline
-\end{tabular}
-\caption{EBA -- Mass Killings during Civil Wars (Robust Variables Only)}
-\label{tab:ucdp}
-\end{table}
-
-The random forest models confirm some of the main findings of the EBA, yet they also show some interesting prediction patterns.
-Figure \ref{fig:drfuv} presents the six most important predictors of state-sponsored violence in out-of-sample tests.
-Overall, the machine learning estimations have a good fit, with an AUC of about 0.8 in the test samples.
-
-However, the findings should be interpreted with caution since variable importance metrics are sensitive to the choice of random seed numbers.
-But despite that expected variation, the results in the random forest models are quite stable even when estimated with different random seeds.
-
-\begin{figure}[h]
-\includegraphics[width=.9\textwidth, height=8cm]{images/drf.pdf}
-\caption{Distributed Random Forest -- Variable Importance (Scaled)}
-\label{fig:drfuv}
-\end{figure}
-
-We see that the most important predictor in our models is the natural log of GDP per capita.
-This confirms the findings of the EBA model.
-Years since the last mass killing episode has a nonlinear relationship with large-scale violence: The likelihood of renewed genocides is small in the first years, although it increases slightly over the next decades.
-The percentage of urban population exhibits a similar pattern.
-Rural areas are less prone to mass killings, yet the probability of such events rise according to a country’s level of urbanisation.
-We understand both GDP per capita and percentage of urban population as proxies for state capacity.
-Conducting a mass killing campaign demands some minimal level of military hierarchy and resources, and countries that are overwhelmingly rural are probably less able to amass the human and physical capital required to conduct such task.
-Furthermore, urban population tend to be less dispersed, what facilitates a large attack by government forces.
-
-\vspace{.5cm}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp.pdf}
-\caption{Distributed Random Forest -- Partial Dependence Plots}
-\label{fig:drfdpp}
-\end{center}
-\end{figure}
-
-The ethnic polarisation variable indicates that countries in which a large share of the population is excluded from politically relevant positions, the lower the probability of a mass killing.
-Groups that manage to control the state apparatus rarely face dangerous opposition movements, and as such they rarely resort to mass violence to impose their rule over the population.
-Next, the size of the military personnel substantially increases the likelihood of state-sponsored violence.
-While small and weak states tend to be more prone to civil war and other types of conflict \citep{collier2004greed,fearon2003ethnicity}, they are less likely to experience one-sided state violence.
-Lastly, the Polity IV index shows that mixed regimes have a higher likelihood of atrocity, and democracies tend to have lower levels of state-led violence.
-
-With respect to mass atrocities that occur during armed conflicts, we see that log GDP per capita and years since mass killings are again relevant predictors.
-Additionally, both covariates are among the 10 most important predictors in the three models.
-The results obtained with the UCDP civil war data are broadly similar to those of the previous model, but it highlights that total battle deaths is the predictor that has a noticeable impact on mass killings onset.
-The partial dependence graph shows a large positive effect along the curve, with a pronounced jump at around 200,000 deaths.
-
-\vspace{1cm}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp2.pdf}
-\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (UCDP data)}
-\label{fig:drfdpp2}
-\end{center}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp3.pdf}
-\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (COW data)}
-\label{fig:drfdpp3}
-\end{center}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp4.pdf}
-\caption{Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. data)}
-\label{fig:drfdpp4}
-\end{center}
-\end{figure}
-
-\newpage
-
-When using the COW data, we find that GDP per capita and total battle deaths have little influence in the likelihood of mass killings, and years since the last state-led atrocity increases the chances of a future event, a result we have also seen in the EBA models.
-As expected, higher levels of abuse against the physical integrity of citizens are associated with mass killings.
-Ethnic polarisation and the threshold of excluded population have a somewhat constant effect in our estimations, yet both show a sharp decline in probability when they reach their maximum levels.
-This provides more evidence to the idea that when certain groups are able to concentrate power, they tend not to resort to mass-scale violence as a political means.
-
-Our last model concerns the onset of ethnic conflicts as defined by \citet{cederman2010ethnic}.
-The graphs show ethnic wars operate with a different dynamics than other armed conflicts.
-We see that an increase in all the most important predictors have a large negative impact on the likelihood of mass killings, thus we conclude that poor, unstable countries are more prone mass killings during ethnic conflicts.
-
-\section{Additional Tests}
-\label{sec:additional-tests4}
-
-We estimate a set of additional regressions to assess the robustness of our main findings.
-In regard to the EBA, we include 10 variants of our original model.
-All of them largely confirm our prior results.
-First, we varied the number of covariates included in each regression to 3 and 5 while keeping the $M$ set of 3 variables included in every model, that is, the natural logarithm of GDP per capita, the Polity IV index, and a linear time trend since the last episode of state-sponsored violence.
-The results are identical to the main model when we add 3 control variables at a time.
-The same covariates discussed above remain significant and with the same sign: log of GDP per capita (negative), post-Cold War period (negative), previous riots (positive), civil war onset and ongoing indicators as measured by the UCDP data set (both positive), Polity IV squared (negative), and ethnic diversity (negative).
-When 5 control variables are added to every model, three variables have a CDF(0) of 0.9 or higher.
-They are the natural logarithm of GDP per capita, civil war onset (UCDP), previous riots, and the dummy indicating the post-Cold War period.
-Ethnic fractionalisation and Polity IV squared become marginally significant with a CDF(0) of about 0.88.
-
-Second, we place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving our results.
-The test is similar to that employed by Bell (2015).
-While in the main analysis we use a VIF of 7, in the additional analyses we include a conservative VIF of 2.5, a less strict value of 10, and a model without any restriction.
-The results are unchanged.
-The two models with different values of VIF show identical results to the results displayed in table \ref{tab:eba1}.
-In the model with no VIF restriction, however, ethnic fractionalisation fails to meet our threshold by a very small margin.
-The CDF(0) of that covariate is 0.897, very close to the required value of 0.9.
-
-We also estimate our models using logit and probit regressions.
-In order to deal with the issue of complete separation \citep{bell2015questioning,zorn2005solution} we follow \citet{gelman2008weakly} and add a weakly informative prior distribution to the coefficients.
-We first scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then we added a Cauchy distribution with center 0 and scale 2.5.
-The probit regressions use a scale of $2.5 \times 1.6$, which is also recommended by the authors \citep{arm2017rpackage}.
-
-In both cases, the logarithm of GDP per capita, post-Cold War period, previous riots, and Polity IV squared remain significant.
-In contrast, ethnic fractionalisation and former conflict as measured by the UCDP data set are again close to the 0.9 significance level.
-
-In regard to random forests, grid searches are themselves a data-driven selection of many possible machine learning models, thus it is not strictly necessary to run a batch of additional tests.
-Nevertheless, we performed a series of grid searches using three different seeds obtained from \texttt{Random.Org} to estimate how the choice of different starting numbers would influence the model outcomes.
-The output of those models are largely comparable.
-
-As a last set of robustness models, we estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide.
-No variable appear as significant in our EBA models for genocide or politicide onset in peacetime.
-When we limit our sample to civil war years, the Post-Cold War period is again negatively correlated with the outcome when using the Correlates of War data set.
-Additionally, excluded population has a negative sign in more than 90\% of the models using both Correlates of War’s and Cederman et al's (\citeyear{cederman2010ethnic}) indicators of conflict.
-Displaced population also has a negative effect in the Correlates of War data set.
-During ethnic conflicts, our dummy variable for political assassinations has a negative impact on the onset of genocides.
-Overall, we conclude then that the significant covariate of genocide and politicide onset differ significantly from those of more general forms of government mass violence.
-Finally, the machine learning models are comparable to the ones we present above, with a similar set of variables appearing in the random forest estimations.
-The results are available in appendix \ref{sec:mk-appendix}.
-
-\section{Conclusion}
-\label{sec:conclusion4}
-
-In this paper, we apply extreme bounds analysis and distributed random forests to estimate the robustness and predictive ability of 40 variables that have been pointed out as potential determinants of mass killings.
-We find that from all variables we employ, GDP per capita is the covariate that appears statistically significant more often in our models.
-In the EBA estimations, we see that the level of democracy, and the post-Cold War period are negatively associated with our dependent variable, whereas ethnic diversity, civil wars, and previous political turmoils increase the likelihood of those events.
-When we run a machine learning algorithm to estimate out-of-sample predictive performance, GDP per capita and years since last mass killings again appear as important predictors, and variables related to military capabilities and the composition of the population are included in most models.
-In a nutshell, our tests provide strong evidence in favour of two well-established findings in the specialised literature: Mass killings are unlikely to happen in rich, stable countries and there is a positive association between several types of political conflict, such as previous riots, civil wars and government atrocities.
-
-Nevertheless, there is considerable heterogeneity in some of our findings.
-Mass killings are rare outcomes, so it is possible that our analysis does not capture all variables that explain the onset of large-scale violence, or that the effect of predictors vary across time and space \citep[8]{bell2015examining}.
-Additionally, the findings point out that mass killings may have different causes according to the context in which they erupt, so a general theory of state atrocities may obscure important details in our understanding of the root causes of state killings.
-
-Yet we see this diversity of outcomes under a positive light.
-The findings presented in this text suggest new avenues for research, and we believe they also highlight the importance of scholars moving from purely correlational, cross-country regressions to other methods that can yield more robust predictions and causal explanations.
-For instance, why are mass killings in ethnic conflicts correlated with a different set of variables than in armed conflicts in general?
-Would the results remain robust had scholars decided to code ethnic conflicts in another way?
-More theoretical advancement would also be welcome.
-Given that GDP per capita is negatively correlated to state atrocities in virtually every model, it would be interesting to unpack the causal mechanisms by which it operates by testing more specific mechanisms.
-Disaggregated data can also be of great help to further current research on genocides, politicide, and mass killings.
-
-In terms of practical implications, the results indicate that regime change and pro-growth economic policies are the most efficient ways to prevent mass killings.
-The international community can therefore play a role in deterring leaders from using force against their own population, either by offering support for domestic opposition groups, intervening, or by fostering economic development.
-Although costly in the short run, these measures would substantially decrease the likelihood of state violence by breaking the ``conflict trap'' in which past conflicts create the condition for new ones \citep{collier2003breaking}.
-
-\newpage
-
-\section{Appendix} 
-\label{sec:mk-appendix}
-
-This appendix contains all required information to replicate the numerical analyses presented in sections \ref{sec:results4} and \ref{sec:additional-tests4}.
-\textt{R} code can be found in subsection \ref{sec:mk-code} and the data are available on the following GitHub repository: \href{https://github.com/danilofreire/mass-killings}{https://github.com/danilofreire/mass-killings}.
+\label{sec:intro}
+
+\doublespacing
+
+This appendix contains all required information to replicate the numerical analyses presented in ``What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models.'' \texttt{R} code can be found in subsection \ref{sec:mk-code} and the data are available on the following GitHub repository: \href{https://github.com/danilofreire/mass-killings}{\texttt{https://github.com/danilofreire/mass-killings}}.
 We used \texttt{R} version 3.4.4 (15-03-2018) and Ubuntu 16.04.4 LTS to perform all statistical calculations.
@@
 The covariates are summarised in table \ref{tab:mk-vs}.
-A complete list of variables is available at \href{https://github.com/danilofreire/mass-killings}{https://github.com/danilofreire/mass-killings}.
+A complete list of variables is available at \href{https://github.com/danilofreire/mass-killings/data}{\texttt{https://github.com/danilofreire/mass-killings/data}}.
 
@@
 \citet[819]{salaimartin2004determinants} argue that random sampling produces unbiased estimates of the regression coefficients with low computational time.
-The models presented in section \ref{sec:results4}, however, include the full set of possible regressions.
+The models presented in the article, however, include the full set of possible regressions.
 
@@
 
-As noted in section \ref{sec:methods4}, we perform a grid search to optimise the hyperparameters of the random forest models.
+As noted in the article, we perform a grid search to optimise the hyperparameters of the random forest models.
 The grid search evaluates a wide range of parameter values at once, therefore it is generally unnecessary to run additional tests to assess the robustness of the results.
@@
 The variable that came closest to significance was a dummy indicator of coups d'état, which has a CDF(0) of 0.897 and, as expected, is positively correlated with the onset of genocides.
-The distibution of the covariates' coefficients are available in figure \ref{fig:uamk}.
+The distribution of the covariates' coefficients are available in figure \ref{fig:uamk}.
 
@@
 
-The \texttt{R} code below reproduces the analyses presented in chapter \ref{chap:killings}.
+The \texttt{R} code below reproduces the analyses presented in the article.
 
 \singlespacing
-\small
+\footnotesize
 \begin{verbatim}
@@
 
-\doublespacing
-\normalsize
+\newpage	
+\bibliography{references}
+\bibliographystyle{apalike}
+\end{document}

### 2018-11-29 paper-killings/manuscript.tex
add killings

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{rotating}
\usepackage{lscape}
\usepackage{amssymb,amsmath,amssymb}
\usepackage[stable]{footmisc}
\usepackage{lmodern}
\usepackage{libertine}
\usepackage[libertine]{newtxmath}
\setmonofont[Scale=0.825]{FiraCode.ttf}
\usepackage[authoryear]{natbib}
\usepackage{babelbib}
\usepackage{booktabs, makecell, longtable}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{darkblue}{rgb}{0.0,0.0,0.55}
\setcitestyle{aysep={}} 
\usepackage{etoolbox}
\makeatletter
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{%
	 \NAT@nmfmt{\NAT@nm}%
	 \hyper@natlinkbreak{\NAT@aysep\NAT@spacechar}{\@citeb\@extra@b@citeb}%
	 \NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}%
   \NAT@aysep\NAT@spacechar\NAT@hyper@{\NAT@date}}{}{}
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{%
	 \NAT@nmfmt{\NAT@nm}%
	 \hyper@natlinkbreak{\NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi}%
   {\@citeb\@extra@b@citeb}%
	 \NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}%
   \NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi\NAT@hyper@{\NAT@date}}
  {}{}
\makeatother
\usepackage{setspace}
\usepackage[top=2.3cm,bottom=2.3cm,left=2.4cm,right=2.4cm]{geometry}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usepackage{mathtools}
\usepackage{caption}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\cleanlookdateon
\exhyphenpenalty=1000
\hyphenpenalty=1000
\widowpenalty=10000
\clubpenalty=1000
\onehalfspacing
\usepackage[backref,pagebackref]{hyperref}
\renewcommand*{\backref}[1]{}
\renewcommand*{\backrefalt}[4]{%
	\ifcase #1 (Not cited.)%
	\or        Cited on page~#2.%
	\else      Cited on pages~#2.%
	\fi}
\renewcommand{\backreftwosep}{ and~}
\renewcommand{\backreflastsep}{ and~}

\hypersetup{pdftitle={What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models},
        pdfauthor={Danilo Freire and Gary Uzonyi},
        pdfsubject={social sciences},
        pdfkeywords={extreme bounds analysis, genocide, mass killings, random forest, state-sponsored violence},
		pdfborder={0 0 0},
		breaklinks=true,
		linkcolor=Mahogany,
		citecolor=Mahogany,
		urlcolor=darkblue,
		colorlinks=true}

\title{What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models\thanks{We thank Toke Aidt, Robert McDonnell, David Skarbek, and Graham Denyer Willis for their helpful suggestions and comments. We are also grateful to Mark S. Bell for sharing his \texttt{R} code to estimate penalised-likelihood models. All replication materials are available at \href{https://github.com/danilofreire/mass-killings}{\texttt{https://github.com/danilofreire/mass-killings}}.}}

\author{Danilo Freire\thanks{Postdoctoral Research Associate, The Political Theory Project, Brown University, 8 Fones Alley, Providence, RI 02912,  \href{mailto:danilofreire@gmail.com}{\texttt{danilofreire@gmail.com}},  \href{http://danilofreire.com}{\texttt{http://danilofreire.com}}. Corresponding author.} \and Gary Uzonyi\thanks{Assistant Professor, Department of Political Science; Research Fellow, Howard H. Baker Jr. Center for Public Policy, University of Tennessee, 1640 Cumberland Ave, Knoxville, TN 37996, \href{mailto:guzonyi@utk.edu}{\texttt{guzonyi@utk.edu}}, \href{https://sites.google.com/site/uzonyigary/}{\texttt{https://sites.google.com/site/uzonyigary}}.}}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
 \noindent
The literature on state-sponsored violence has grown significantly over the last decades. Although scholars have suggested a number of potential correlates of mass killings, it remains unclear whether the estimates are robust to different model specifications, or which variables accurately predict the onset of large-scale violence. We employ extreme bounds analysis and random forests to test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013. The results help clear the brush around mass killings, as few variables in this literature are robust determinants of atrocity. However, support for an opportunity logic persists as greater constraints on a government limit its ability to employ barbarous tactics. It appears that the Conflict Trap applies to government atrocity. Atrocity breeds atrocity, while wealthy stable democracies tend to avoid episodes of mass killing.

 \vspace{.75cm}
 \noindent
 \textbf{Keywords}: extreme bounds analysis, genocide, mass killings, random forests, state-sponsored violence

 \vspace{.25cm}
 \noindent
 \textbf{JEL Classification}: C52, C53, D74, H56, K10
\end{abstract}

\newpage

\section{Introduction}
\label{sec:intro}

\doublespacing

Since the end of World War II, mass killings, genocides, and politicides have claimed over 34.5 million lives \citep{marshall2017pitf}.\footnote{Genocide and politicide are the attempted intentional destruction of communal or political groups, respectively \citep[see][]{harff1988toward}. Mass killing includes these atrocities, as well as attacks against civilians that result in at least 1,000 deaths but are not intended to destroy a particular group \citep[see][]{ulfelder2008assessing}. While some conflate the logic of these types of atrocities \citep[e.g.,][]{rummel1995democracy, valentino2004draining}, others claim genocide and politicide follow a different logic from other forms of government violence \citep{kalyvas2006logic,stanton2015regulating}. For discussion on these important differences in conceptualisation see \citep[]{straus2007second, finkel2012macro}} The international community has responded with an effort to prevent further state-sponsored mass murder by strengthening laws against war crimes, genocide, and crimes against humanity. Furthermore, the United Nations established a Special Adviser on the Prevention of Genocide and recognised its members' responsibility to protect civilian populations within and outside their own borders. Yet, such atrocities still occur. Recently, President al-Assad of Syria has massacred tens of thousands of civilians during the Syrian Civil War \citep{goldman2017nyt}. Similarly, South Sudan's President Kiir is actively starving and killing civilians from dissident and rival tribes \citep{nichols2017reuters}. While there is some evidence that such atrocities may be declining since the Cold War \citep{valentino2014we}, the international community has been far from successful in realising slogans like ``Never Again'' and ``Not on My Watch'' \citep{cheadle2007not}.

Ultimately, effective prevention requires us to understand why these atrocities occur. In this vein, the academic community has laboured tremendously to establish evidence-based theories as to why governments engage in brutality against their civilian populations. Indeed, since 1995, there have been over 45 quantitative political science articles focused on explaining government-sponsored killing of civilians. Overall, the mass violence literature agrees that government atrocity follows an opportunity logic: as threat increases, so does the likelihood of atrocity, if the costs to such violence are not prohibitive. However, there is little consensus on what factors influence the level of threat or costs a regime faces. Part of the reason for this uncertainty is that scholars use very different model specifications when testing their arguments, thus small changes in model parameters could influence the robustness of empirical results and the inferences we draw from these findings.

To overcome these limitations and provide a better understanding of government atrocity, I employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities. Our approach is similar to \cite{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but we provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results \citep{bell2015examining,jones2015exploratory,jones2018there,muchlinski2015comparing}. In conducting this analysis, we address three debates in the mass violence literature:

\begin{enumerate}
 \item Why do some governments engage in mass killings, genocides, or politicides? This is the primary question asked by activists, policymakers, and scholars in this field of research. 
 \item Does the logic underpinning government decision-making follow different patterns during peacetime and wartime? Recent research suggests that government atrocity occurs predominantly during periods of civil unrest \citep{harff2003no} which has led some scholars to restrict their analyses to only periods of civil war \citep[e.g.,][]{colaresi2008kill, valentino2004draining} or concentrate on predicting both the onset of civil war and atrocity \citep{goldsmith2013forecasting}. Yet, others estimate models of all country-years \citep[e.g.,][]{krain1997state, montalvo2008discrete}, raising questions of how well these studies speak to each other.
 \item Is there a difference in logic between those atrocities labelled as genocide or politicide, compared to other mass killings? While the Political Instability Task Force \citep{marshall2017pitf} provides the most widely used data on government atrocity, others provide data with much more lenient inclusion criteria \citep[e.g.,][]{stanton2015regulating, ulfelder2012forecasting}. These differences in definition of atrocity have led to divergent results, raising questions about important determinants of government behaviour \citep[for discussion, see][]{straus2007second, uzonyi2016domestic, wayman2010explaining}.
\end{enumerate}

Our analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013. Our findings partially confirm previous research -- unstable countries are more likely to witness the regime employ atrocity \citep[e.g.,][]{goldsmith2013forecasting,harff2003no,krain1997state}. However, many of the factors scholars often cite as observable indicators of such instability -- regime transitions, coups d'état, the presence of militias, etc. -- are not good proxies for instability. Thus, policymakers may be looking for the incorrect signs of impending atrocity when seeking to prevent its onset. Furthermore, we find that the conclusions scholars draw regarding the likelihood of government atrocity largely depend on whether they combine peace and war years or just analyse periods of civil wars, as patterns in mass killings differ dramatically across these contexts. Lastly, we find that genocide and politicide follow vastly different patterns of onset than other forms of state-sponsored mass murder. This is further evidence that different logics govern different forms of political violence \citep{stanton2013terrorism}.

Overall, these findings raise concerns about policy options for preventing violence against civilians. If our conclusion is that unstable countries are violent, then preventing atrocity likely requires significant investments of time and resources in state-building, which is often politically and practically unfeasible \citep{doyle2006making}.  This analysis contributes significantly to the political violence literature by highlighting the parsimonious nature of the logic behind government atrocity and clearing away much of the empirical clutter surrounding this conclusion.

\section{Empirical Methods}
\label{sec:methods}

To conduct our analysis, we began by surveying the quantitative political science literature on the causes of government mass killing since Rummel's (\citeyear{rummel1995democracy}) seminal work on the subject. Counting only published works, we identified 45 articles which employed logit or probit models of mass killing onset in a global sample.  We then included all variables that appeared in at least two of these papers in our data set at the country-year unit of analysis for all years from 1945 to 2013. The appendix provides a complete list of the articles we considered and a complete list of the variables we included in our models. Next, we estimated an extreme bounds analysis to determine which variables were the most robust in explaining the onset of government atrocity. Then, we estimated a distributed random forest analysis to see which of the variables best predicted the onset of these atrocities. In this section, we provide more detail on each of these estimation procedures before turning to the results of both analyses in the next section.

\subsection{Extreme Bounds Analysis}
\label{subsec:eba}

Th first method we employ to test the robustness of the potential determinants of state-led violence is the extreme bounds analysis (EBA). Researchers have employed EBA to assess the sensitivity of the determinants of civil war \citep{hegre2006sensitivity}, coups d'état \citep{gassebner2016expect}, democratisation \citep{gassebner2013extreme}, economic growth \citep{levine1992sensitivity, sala1997just}, nuclear deterrence \citep{bell2015examining}, and political repression \citep{hafner2005right}. The method is particularly useful when there is no consensus about which covariates belong in the ``true'' regression model \citep[178]{sala1997just} and scholars worry that omitted or unnecessary predictors can bias the model estimates \citep[60]{angrist2008mostly, clarke2005phantom, elwert2014endogenous, spector2011methodological}. 
More specifically, the main purpose of EBA is to estimate the distribution of coefficients of each predictor $x$ in an exhaustive combination of regression models with $y$ as a dependent variable. \cite[308]{leamer1985sensitivity} proposed that ``sturdy'' variables are those whose minimum and maximum of their coefficient distribution have the same sign and are situated at a distance from zero. If we are to use the conventional value of $p < 0.05$, the mean of the variable coefficients' distribution should be located at least $1.96$ standard deviations away from zero.

Leamer's criterion is intuitive, but other authors contend it is too strict for most social science applications. \cite{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response. In this paper, we use Sala-i-Martin's \citeyear{sala1997just} more flexible version of EBA and consider the whole range of values of $CDF(0)$. We choose to use the whole distribution because the aggregate $CDF(0)$ allows us to move away from a binary indicator of robustness and present the estimations with their appropriate degrees of confidence. Our focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero. We do not assume that the CDFs are normally distributed and use Sala-i-Martin's generic model instead.\footnote{The generic model provides a better fit to our data. Histograms for all coefficients are available in online appendix.} We specify our models as follows:

\begin{equation}
\text{\textit{Mass Killing Onset}}_{it} = \beta_{M}M_{it} + \beta_{F}F_{it} + \beta_{Z}Z_{it} + v_{it}
\end{equation}

Our main dependent variable is \textit{Mass Killing Onset}, which denotes the onset of government-sponsored killings. \citet[2]{ulfelder2008assessing} define a mass killing as ``any event in which the actions of state agents result in the intentional death of at least 1,000 noncombatants from a discrete group in a period of sustained violence''. Respectively, $i$ and $t$ indicate country and year. $M$ is a set of three covariates that are included in every model due to their prominence in the literature \citep{levine1992vale}. In our analysis, $M$ includes the natural logarithm of the GDP per capita to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence. $F$ denotes a vector of variables of interest, and $Z$ is a vector of other control variables in addition to those included in $M$. $v$ is the error term. In practice, however, since we are interested in the effect of all variables in the data set and do not have true control variables, except from $M$, $F$ and $Z$ are interchangeable. We thus only use this notation to help clarify the connection of our analysis to previous conflict scholars who employed similar extreme bounds analysis \citep[e.g.,][]{hegre2006sensitivity,gassebner2016expect}. Following \citet[514]{hegre2006sensitivity}, we lagged the independent variables one year to reduce the risk of endogeneity.

Although our dependent variable is dichotomous, we use linear probability models in our main analysis. \citet[298]{gassebner2016expect} argue that linear probability models are less prone to convergence problems and their results can be readily interpreted. Since the data are grouped into countries, we also use cluster-robust standard errors.

As a precaution against collinearity, we place a limit on the Variance Inflation Factor (\textit{VIF}) of all regression coefficients. The VIF estimates how much of the variance of each predictor is dependent on the other covariates in a model. A VIF of 1 indicates that the predictor is uncorrelated with the remaining covariates. VIF limits are often arbitrary \citep{bell2015examining,o2007caution}, thus here we use a moderately conservative VIF of 7. As robustness tests, we run the same models without restriction and with different cut-offs.

Two variables were omitted from EBA models but included in the machine learning estimations. The first is \textit{democracy}, a dummy variable that indicates whether the country has a Polity IV score equal or higher than 5. The second is \textit{interstate war}, a binary covariate measuring if the country is at war in a given year \citep{sarkees2010resort}. We have decided to omit democracy because of its evident correlation with the Polity measure and interstate war due to its correlation with our dependent variables. EBA models do not converge otherwise.\footnote{This is a typical case of multicollinearity and conceptual overlap. \citet{hlavac2016eba} suggests specifying a set of mutually exclusive variables to avoid the issue. However, as the Polity index is one of our core variables, we decided to drop the binary democracy indicator and use the continuous measure as it provides more details about the effect of political regimes on mass killings. More information available in the appendix.}  Since this problem does not affect machine learning algorithms, the two variables were included in the second set of estimations.

Lastly, we depart slightly from Sala-i-Martin's suggested method and do not assign weights to EBA. Although he recommends using goodness-of-fit measures to construct regression weights, we agree with \citet{sturm2002robust} and \citet[299]{gassebner2016expect} and use the unweighted version of the CDF instead. Goodness-of-fit indicators are not equivalent to the probability of a given model being true \citep{anscombe1973graphs,king1986not}, and the weights constructed this way are not invariant to transformations in the dependent variable. Moreover, our data set has a number of missing observations, so model comparison measures could be misleading \citep{lall2016multiple}.

\subsection{Random Forests}
\label{sub:drf}

We also make use of random forest (RF) \citep{breiman2001random} to evaluate whether the empirical results discussed in the mass killings literature are not driven by rigid parametric assumptions and model specifications. Random forest is a machine learning algorithm that consists of a combination of individual decision trees. In a classification problem, each decision tree uses a vector of covariates to split the dependent variable into two increasingly homogeneous parts \citep{breiman2001statistical}. However, decision trees are prone to overfitting, i.e., they match the original data set so closely that they tend to perform poorly with new data \citep{dietterich1995comparison,ho1998random}. Random forest, in contrast, avoids this issue by growing a decision tree only to a bootstrap sample of the original data, selecting random features at each split, then aggregating the different trees into a single prediction. If the independent variable is continuous, the algorithm will simply choose the average value of the predictions as the best candidate; if the covariate is discrete, the majority class will be employed. The simple procedure of leaving out some data points and growing separate trees with a random subset of covariates is sufficient to eliminate the risk of overfitting \citep[9-10]{jones2015exploratory}.

Random forest has many desirable properties, such as ``highly accurate predictions, robustness to noise and outliers, internally unbiased estimate of the generalisation error, efficient computation, and the ability to handle large dimensions and many predictors'' \citep[7]{muchlinski2015comparing}. Thus, random forest allows the researcher to estimate very flexible models with minimal assumptions. Unlike parametric methods such as ordinary least squares or logistic regressions, the analyst does not have to impose any distributional form to the data-generating process. As a result, random forest is able to effectively uncover complex, nonlinear interaction effects in the data without prespecification \citep{jones2015exploratory,jones2018there}.

In this paper we use distributed random forest (DRF) to model our data, a slightly modified version of the original random forest algorithm \citep{h2o2017}. The DRF has two additional features that are useful for our purposes. Firstly, DRF is optimised for big data, as it grows decision trees on separate cores to speed up computation time. Secondly, in DRF, non-observed cases are not assumed to be missing at random, but rather as values that contain information in themselves. The algorithm assumes that observations are missing for a reason, what is most likely the case with social science data \citep{lall2016multiple}. This is a more conservative approach than assuming that missing cases fit into an underlying parametric distribution.\footnote{For more information about how the distributed random forest algorithm deals with missing observations, please refer to: \href{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html}{\texttt{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html}} (access: December 2017).}

The DRF has a series of hyperparameters that can be tuned to improve its predictive performance. For instance, users can control the number of decision trees in each iteration, how deep trees should grow, and many other options. The interaction between parameters is generally complex and may involve thousands of potential combinations. As an example, a researcher interested in four parameters with 10 values each would have to estimate 10,000 models before deciding which is the most efficient one. Also, machine learning parameters are sensitive to the data at hand, that is, an optimal solution for one problem cannot be readily implemented in another data set \citep{genuer2008random,goldstein2010application,jones2015exploratory}.

To address these issues, we perform a grid search to select the most accurate random forest model \citep[123]{cook2017h2o}. We estimate a model for every possible combination of the hyperparameter space to make sure our results are robust to different specifications. For model selection, we follow the literature on predictive political science and use the area under the ROC curve (AUC) as our model evaluation metric \citep[e.g.,][]{hill2014empirical,ward2010perils,ward2013learning}. Models with higher AUC values are considered more accurate.

We add several parameters to the grid search. The first is the number of independent trees to grow in each forest. The starting values are 256, 512, and 1024 trees. The machine learning literature does not provide a heuristic on how large a random forest should be, but \citet[166]{oshiro2012many} affirm that ``from 128 trees there is no more significant difference between the forests using 256, 512, 1024, 2048 and 4096 trees.'' We employ a more conservative approach and start from a higher value that the authors suggest as adding more trees do not reduce prediction accuracy \citep[7]{breiman2001statistical}.

The depth of each decision tree also influences the algorithm performance. Deeper trees indicate more complex models, and in general they provide a better fit to the data. Nevertheless, this complexity comes at the risk of overfitting, so deeper trees are not necessarily the most adequate solution for every model \citep[596]{friedman2001greedy,segal2004machine}. In this article, we let the algorithm decide among using 10, 20, or 40 levels for each tree.

We test whether having balanced classes of our dependent variable (mass killing onset) affects the predictive ability of the model. Since the response measure is heavily imbalanced, oversampling the positive responses could potentially improve our results \citep{chawla2004special,del2014use,japkowicz2002class}. 
We also vary how many variables should be considered for each split in the data. The default option is to use $\sqrt{p}$, where $p$ is the number of columns in the data set. As we have 40 covariates of interest, we have selected 5, 6 and 7 variables per split. The DRF uses a majority voting procedure to select which variable is most important. Additionally, the algorithm chooses the percentage of the training set to be modelled by each tree. The default option is 63.2\%, but we include the options of using 50\% and 100\% of the data. Similarly, we give a range of options for choosing how many columns will be included in each tree. The algorithm can randomly choose among 50\%, 90\% or 100\% of the independent variables when estimating a decision tree.

Finally, we use three types of histogram to find optimal split points for each independent variable. Decision trees consider every value of a given independent variable as a potential candidate for a split in the training data. This process is notably time-consuming, and computation time can be significantly reduced at little loss of precision by taking discrete values of the predictor distribution. The DRF algorithm also offers the choice of randomly cycling through all histogram types, including one type in each tree estimation. We adopt this ``round robin'' arrangement as it is both efficient and parsimonious.

\section{Results}
\label{sec:results}

We endeavour to answer three questions in our analysis: (1) what are the robust predictors of government mass killing, (2) do these predictors differ when considering only cases of civil war, and (3) are genocide and politicide different than other forms of atrocity? Table \ref{tab:eba01} summarises our main EBA results in answering Question 1. The table shows the average coefficient estimate of all regressions for each robust variable along with their mean standard deviations.\footnote{A list of all independent variables and coding rules are available in the appendix.} The table also displays the percentage of regressions that are statistically significant at the 90\% level. $CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{We show whichever area is the largest. The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is our main statistic of interest, and we consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}. Lastly, we report the number of estimated regressions models which included each variable.

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{Base variables} &  &  &  &  &  \\
Log GDP per capita & -0.0091 & 0.0052 & 76.055 & 0.9335 & 226707 \\
 &  &  &  &  &  \\
\textit{Additional variables} &  &  &  &  &  \\
Post-Cold War years & -0.0133 & 0.0085 & 72.845 & 0.9472 & 35614 \\
UCDP civil war onset & 0.0529 & 0.0321 & 52.378 & 0.9441 & 20854 \\
Previous riots & 0.0140 & 0.0100 & 56.242 & 0.9216 & 35614 \\
UCDP ongoing civil war & 0.0172 & 0.0115 & 65.652 & 0.9092 & 20854 \\
Ethnic diversity (ELF) & 0.0184 & 0.0137 & 56.674 & 0.9050 & 35614 \\
Polity IV squared & -0.0002 & 0.0001 & 61.206 & 0.9031 & 35614 \\ \hline
\end{tabular}
\caption{Extreme Bounds Analysis -- Mass Killings (Robust Variables Only)}
\label{tab:eba01}
\end{table}

Seven variables pass our EBA criteria and three of them decrease the likelihood of mass killings. First, as widely suggested in the literature, the natural logarithm of GDP per capita is negatively associated with the onset of mass killings \citep[e.g.,][]{besanccon2005relative, easterly2006development,esteban2015strategic}. Second, the post-Cold War years are correlated with lower levels of government violence. Indeed, this finding is in line with several studies that point to a general decline in violence over the last decades, including riots, civil wars, and urban crime \citep{pinker2011better,straus2012wars,valentino2014we}. The third robust variable is the squared term of the Polity IV political regime index. This finding points to a nonlinear relationship between political regime and mass killings, thus providing further evidence that democracy reduces state-sponsored violence \citep{rost2013will,rummel1995democracy} and that regimes that mix democratic with autocratic features have the highest risk of conflict \citep{hegre2001toward,muchlinski2014grievances}.

Four variables are robustly and positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of government-sponsored violence. Onset and continuation of civil wars are correlated with mass killings, but only when we employ the UCDP measures of violent conflict. We find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}. Former instances of political turmoil also have a positive coefficient in our models. Moreover, countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent \citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}. Our results also show that higher levels of ethnic diversity increase the likelihood of atrocities against civilians. Nevertheless, ethnic diversity does not pass all additional tests we implement below and the sturdiness of this finding remains open to question.

Overall, the EBA indicates two patterns in answer to our first question on the causes of mass killing. Atrocity is (1) more likely when violence is already present, reducing the costs of escalating brutality and (2) is less likely as domestic and international constraints increase, increasing the costs of escalating violence further. These patterns support the dominant opportunity narrative in the literature. However, several of the variables commonly used to proxy opportunity, such as military size or regime change, are not robust predictors of atrocity. Thus, this analysis helps clear away much of the brush around the opportunity argument.

Figure \ref{fig:drfuv} presents the ten most important predictors of state-sponsored violence in our random forest models. Overall, the machine learning estimations have a good fit, with an AUC of about 0.83 in the validation sample. The algorithm confirms some of the main findings of EBA, yet they also show some interesting patterns. Only democracy and state capacity appear to be robust explanators of mass killing in both EBA and random forest models. These patterns further support the refrain that stable states tend to stay stable states.

\begin{figure}[ht]
    \centering
    \includegraphics[width=.8\textwidth, height=8.25cm]{images/rf-mk.pdf}
    \caption{Distributed Random Forest -- Variable Importance (Scaled)}
    \label{fig:drfuv}
\end{figure}

Interestingly, several variables that are not robust explanators of mass killing in the EBA, have large importance in the machine learning estimates. Variables related to characteristics of the military forces are a good examples. As we see below, parametrisation and interactions likely account for this difference. The linear model imposes a parametric structure to the covariate, and the relationship between an independent variable and the response may be a nonlinear one. Also, variables can be relevant predictors only when in interaction with each other. In both cases, those relationships will be captured in the machine learning estimations but not in the extreme bounds analysis.

% Moreover, one should have in mind that the EBA and DRF play two distinct, albeit complementary, academic goals: explanation and prediction \citep{bell2015examining,breiman2001statistical,ward2010perils}. In the first case, the main purpose is to understand historical patterns, the latter method tests the predictive power of the covariates using out-of-sample data. There is no reason to suspect the results obtained in the EBA, but the DRF provides a stricter test to the covariates as it uses data that the training model had not seen. This provides a better safeguard against overfitting and model mispecification.

Figure \ref{fig:drfdpp} displays the partial dependence plots for the ten variables that the distributed random forests highlight as the most important explanators of mass killing onset. These graphs are akin to marginal effect plots in correlation models and help clarify the directional effects of these variables over their entire range. For example, we are able to see that the effects of Years since Last Mass Killings and Polity IV are highly nonlinear, or that Log GDP per capita does not decrease the likelihood of mass killings after it reaches a value of 10.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth, height=8cm]{images/rf-mk-pd.pdf}
    \caption{Distributed Random Forest -- Partial Dependence Plots}
    \label{fig:drfdpp}
\end{figure}

% One paragraph on the results

Table \ref{tab:ucdp} presents our EBA results when we restrict our analysis to only civil war years to answer Question 2. We consider three different codings of civil war: (1) the Uppsala Conflict Database Program  (\citeyear{allansson2017organized,gleditsch2002armed}), (2) the Correlates of War project \citep{sarkees2010resort}, and (3) ethnic civil war from \citet{cederman2010ethnic}. We find two important patterns. First, considering only civil war years provides a very different understanding of atrocity. Across these models, the only similarity with the full analysis is that mass killing is less likely post-Cold War. Instead, military factors, such military size and militias, and territorial war aims are the most robust predictors of atrocity once war begins. However, contrary to past expectation \citep{koren2017means}, militias have a negative impact on the likelihood of mass killings. Second, there is wide variation in which variables are robust depending on how we code civil war. Across our three codings, no variable is robust to all codings and only territorial aims and militias are robust to more than one coding. These results are concerning for scholars using correlation models, as they indicate that our understanding of atrocity, from null hypothesis testing, is largely dependent on which coding of civil war we use. For example, only the UCDP data suggests that the post-Cold War years see less barbarism than during the Cold War.

\newpage

\begin{table}[H]
\centering
\begin{tabular}{lrrrrr}
\hline
\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
\textit{UCDP Data} &  &  &  &  &  \\
Territory aims & -0.044 & 0.019 & 74.997 & 0.9804 & 17902 \\
Post-Cold War years & -0.038 & 0.019 & 66.574 & 0.9222 & 17902 \\
 &  &  &  &  &  \\
\textit{COW Data} &  &  &  &  &  \\
Physical integrity & 0.024 & 0.013 & 66.674 & 0.9564 & 17902 \\
Militias & -0.099 & 0.048 & 73.104 & 0.9490 & 17902 \\
Years since last mass killing & 0.006 & 0.002 & 88.208 & 0.9472 & 101583 \\
Previous riots & 0.078 & 0.041 & 65.412 & 0.9348 & 17902 \\
Ethnic diversity (ELF) & 0.095 & 0.062 & 48.615 & 0.9000 & 17902 \\
& & & & \\
\textit{Cederman et al. Data} &  &  &  &  &  \\
Territory aims & -0.051 & 0.026 & 74.288 & 0.9167 & 17902 \\
Militias & -0.050 & 0.035 & 52.240 & 0.9101 & 17902 \\ \hline
\end{tabular}
\caption{EBA -- Mass Killings during Civil Wars (Robust Variables Only)}
\label{tab:ucdp}
\end{table}

When we analyse the three codings of civil war using our random forest analysis, we find further intricacies in the patterns of mass killing. First, the machine learning estimates highlight a different set of variables than the EBA when analysing the UCDP and Ethnic War data. However, the COW EBA and machine learning analyses both highlight the importance of human rights, previous riots, and the time since the state last engaged in mass killing. Thus, the COW analysis provides the most stable picture of atrocity during civil war. It again highlights the important pattern of the Conflict Trap: violence breeds violence. Second, though, the three codings of civil war each highlight a very similar set of strong predictors of atrocity during conflict. Therefore, the machine learning estimates are not as dependent on the data set employed as are the EBA results. This is good news for scholars of mass killing because it indicates that while our parametric models do not produce robust findings across different civil war data sets, our nonlinear models are able to given us a consistent and clear picture of which factors place a country at the greatest risk for atrocity during civil war.

Figures \ref{fig:drfdpp2}--\ref{fig:drfdpp4} display the partial dependence plots for the variables with the highest impact in each of the three civil war data sets.

\vspace{1cm}

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth, height=8cm]{images/rf-ucdp-pd.pdf}
    \caption{Partial Dependence Plots -- Mass Killings during Civil Wars (UCDP Data)}
    \label{fig:drfdpp2}
\end{figure}

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=\textwidth, height=8cm]{images/rf-cow-pd.pdf}
    \caption{Partial Dependence Plots -- Mass Killings during Civil Wars (COW Data)}
    \label{fig:drfdpp3}
    \end{center}
\end{figure}

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=\textwidth, height=8cm]{images/rf-cow-pd.pdf}
    \caption{Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. Data)}
    \label{fig:drfdpp4}
    \end{center}
\end{figure}

To answer Question 3,  we estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide. We find that no variable appears significant in our EBA models for genocide or politicide onset in the full data set. When we limit our sample to civil war years, the Post-Cold War period is negatively correlated with the outcome when using the Correlates of War data set. Excluded population has a negative sign in more than 90\% of the models using both Correlates of War's and Cederman et al's (\citeyear{cederman2010ethnic}) indicators of conflict. Displaced population also has a negative effect in the Correlates of War data set. During ethnic conflicts, our dummy variable for political assassinations has a negative impact on the onset of genocides. Overall, from these EBA analyses, we conclude then that the significant covariates of genocide and politicide onset differ significantly from those of more general forms of government mass violence. Though, the opportunity story still receives some limited support in these models. However, the machine learning models using Harff's genocide and politicide data are comparable to the ones we present above, with a similar set of variables appearing in the random forest estimations. These results once more highlight that while the mass killing literature struggles to identify correlates of atrocity that are robust across model specification, scholars have done a much better job at identifying variables that help predict both the onset of genocide/politicide and mass killings, more broadly.

\section{Additional Tests}
\label{sec:additional-tests}

We estimate a set of additional regressions to assess the robustness of our main findings. In regard to EBA, we include 10 variants of our original model. They largely confirm our prior results. First, we varied the number of covariates included in each regression to 3 and 5 while keeping the $M$ set of 3 control variables. The results are the same as those of the main model, except that ethnic fractionalisation and Polity IV squared become marginally significant with a CDF(0) of about 0.88. Second, we place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving our results. The two models with different values of VIF are identical to the model reported here, while in the model with no VIF restriction ethnic fractionalisation again fails to meet our threshold by a very small margin.

We also reestimate the models using logit and probit regressions. In order to deal with the issue of complete separation \citep{bell2015questioning,zorn2005solution} we follow \citet{gelman2008weakly} and add a weakly informative prior distribution to the coefficients. In both cases, the logarithm of GDP per capita, post-Cold War period, previous riots, and Polity IV squared remain significant.

In regard to random forests, grid searches are themselves a data-driven selection of many possible machine learning models, thus it is not strictly necessary to run a batch of additional tests. Nevertheless, we performed a series of grid searches using three different seeds obtained from \href{https://random.org}{\texttt{http://random.org}} to estimate how different starting numbers influence the model outcomes. The output of those models are largely comparable.

The results of each of these analyses are available in online appendix.

\section{Conclusion}
\label{sec:conclusion}

In this paper, we apply extreme bounds analysis and distributed random forests to estimate the robustness and predictive ability of 40 variables that have been pointed out as potential determinants of mass killings. We find strong evidence that mass killings are unlikely to happen in rich, stable countries. Nevertheless, there is considerable heterogeneity in some of our results. The findings point out that mass killings have different causes according to the context in which they erupt, so a general theory of state atrocities may obscure important details in our understanding of state killings. Moreover, mass killings are rare events, so local factors likely play an important role in their onset \citep{straus2007second,straus2012destroy}.

Yet we see this diversity of outcomes under a positive light. Our results suggest new avenues for research, and we believe they also highlight the importance of scholars moving from simple cross-country regressions to methods that can yield more robust predictions. For instance, why are mass killings in ethnic conflicts correlated with a different set of variables than in armed conflicts in general? Would the results remain robust had scholars decided to code ethnic conflicts in another way? More theoretical advancement would also be welcome. Given that GDP per capita is negatively correlated to state atrocities in virtually every model, it would be interesting to unpack the causal mechanisms by which it operates by testing more specific mechanisms.

In terms of practical implications, the results indicate that democratisation and pro-growth economic policies are the most efficient ways to prevent mass killings. The international community can therefore play a role in deterring leaders from using force against their own population, either by offering support for domestic opposition groups, intervening, or by fostering economic development. Although costly in the short run, and sometimes violent during the transition, these measures would substantially decrease the likelihood of state violence by breaking the ``conflict trap'' in which past conflicts create the condition for new ones \citep{collier2003breaking}.

\newpage

\bibliography{references}
\bibliographystyle{apalike}
\end{document}

### 2018-11-29 paper-killings/R/data-wrangling.R

@@ -1,0 +1,102 @@
+##########################################################
+### The scripts included in this folder replicate      ###
+### the analyses contained in the paper                ###
+### "What Drives State-Sponsored Violence?: Evidence   ###
+### from Extreme Bounds Analysis and Ensemble          ###
+### Learning Models", by Danilo Freire and Gary Uzonyi ###
+###                                                    ### 
+### Please contact me at danilofreire@gmail.com if you ###
+### encounter any difficulties                         ###
+##########################################################
+
+### Load required packages
+if (!require("tidyverse")) {
+        install.packages("tidyverse")
+}
+if (!require("data.table")) {
+        install.packages("data.table")
+}
+if (!require("ExtremeBounds")) {
+        install.packages("ExtremeBounds")
+}
+if (!require("h2o")) {
+        install.packages("h2o")
+}
+if (!require("sandwich")) {
+        install.packages("sandwich")
+}
+if (!require("arm")) {
+        install.packages("arm")
+}
+
+### Load data
+setwd("~/Documents/mk/") # set the working directory
+df <- haven::read_dta("data/base variables.dta") %>% setDT()
+
+### Select and lag variables
+sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
+             "COWcivilwarongoing", "ethnowarstart", "ethnowarongoing",
+             "assdummy", "demdummy", "elf", "lmtnest", "pop", "realgdp",
+             "rgdppc", "polity2", "exclpop", "discpop", "polrqnew",
+             "poltrqnew", "egiptpolrqnew", "egippolrqnew", "discrim",
+             "elf2", "interstatewar", "milex", "milper", "percentpopurban",
+             "postcoldwar", "coupdummy", "riotdummy", "territoryaims",
+             "totaltrade", "tradedependence", "militias", "physint", "cinc",
+             "totalbeaths", "change", "guerrilladummy", "sf", "regtrans")
+
+df1 <- cbind(df, df[, shift(.SD, 1, give.names = TRUE),
[…]

### 2018-11-29 paper-killings/R/extreme-bounds.R

@@ -1,0 +1,319 @@
+
+################################
+### Extreme bounds analysis  ###
+################################
+
+# Classifying a few variables as mutually exclusive variables.
+# "Change" was removed because it was correlated at 0.99 with "regtrans". 
+# don't forget to add CINC
+free.variables <- c("logrgdppc", "polity2", "mksyr")
+civilwar.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart")
+doubtful.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart", "assdummy",
+                        "totaltrade", "tradedependence", "milper", "milex",
+                        "pop", "totalbeaths", "guerrilladummy", "regtrans",
+                        "riotdummy", "territoryaims", "militias",
+                        "physint", "percentpopurban", "coupdummy",
+                        "postcoldwar",  "lmtnest", "realgdp", "discrim",
+                        "exclpop", "discpop", "elf",  "polrqnew",
+                        "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                        "polity2sq")
+
+# Cluster-robust standard errors
+se.clustered.robust <- function(model.object){
+        model.fit <- vcovHC(model.object, type = "HC", cluster = "country")
+        out <- sqrt(diag(model.fit))
+        return(out)
+}
+
+### Models
+
+# Main
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, vif = 7, level = 0.9,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/mk.rda")
+
+# 3 vars at a time
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:3,
+          data = df2, vif = 7, level = 0.9, draws = 10000,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/mk-3vars.rda")
+
+# 5 vars at a time
[…]

### 2018-11-29 paper-killings/R/graphs.R

@@ -1,0 +1,639 @@
+################
+# EBA Graphs ###
+################
+
+# Main models
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
+                       "UCDPcivilwarongoing",
+                       "UCDPcivilwarstart", "COWcivilwarongoing",
+                       "COWcivilwarstart", "ethnowarongoing", "ethnowarstart",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias", "physint",
+                       "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
+                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
+                       "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last genocide",
+              "UCDP ongoing", "UCDP onset", "COW ongoing", "COW onset", 
+              "Ethnic ongoing", "Ethnic onset", "Assassination", "Total trade", 
+              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+              "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+# Round
+m1$coefficients$mean$beta2 <- round(as.numeric(m1$coefficients$mean$beta),4)
+m1$coefficients$mean$se2 <- round(as.numeric(m1$coefficients$mean$se),4)
+m1$coefficients$mean
+
+## Models including only mass killings during civil wars
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
+                       "assdummy", "totaltrade", "tradedependence", "milper",
+                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
[…]

### 2018-11-29 paper-killings/R/random-forests.R

@@ -1,0 +1,514 @@
+######################
+### Random forests ###
+######################
+
+# The following two commands remove any previously installed H2O packages for R.
+if ("package:h2o" %in% search()) { detach("package:h2o", unload=TRUE) }
+if ("h2o" %in% rownames(installed.packages())) { remove.packages("h2o") }
+
+# Next, we download packages that H2O depends on.
+pkgs <- c("RCurl","jsonlite")
+for (pkg in pkgs) {
+        if (! (pkg %in% rownames(installed.packages()))) { install.packages(pkg) }
+}
+install.packages("h2o", type="source", repos="http://h2o-release.s3.amazonaws.com/h2o/rel-wolpert/4/R")
+
+
+# Load package
+library(h2o)
+h2o.init(nthreads = -1)
+
+df2a <- as.h2o(df2)
+
+df2a$MKstart <- as.factor(df2a$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df2a$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2a, 
+                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
+                         seed = 42)  # reproducibility
+
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex") 
+test <- h2o.assign(splits[[3]], "test.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df2), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq")) 
+
+##########################
+### Running the models ###
+##########################
+
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, nfolds = 5, grid_id = "gridrf01",
+               fold_assignment = "Stratified",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = c("UniformAdaptive",
+                                                      "Random",
+                                                      "QuantilesGlobal",
+                                                      "RoundRobin")),
[…]

### 2018-11-30 paper-killings/manuscript.tex
first edits, continue

@@
 \usepackage[libertine]{newtxmath}
-\setmonofont[Scale=0.825]{FiraCode.ttf}
+\usepackage[FiraMono]
 \usepackage[authoryear]{natbib}
@@
 \hypersetup{pdftitle={What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models},
-        pdfauthor={Danilo Freire and Gary Uzonyi},
+        pdfauthor={Danilo Freire},
         pdfsubject={social sciences},
@@
 
-\title{What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models\thanks{We thank Toke Aidt, Robert McDonnell, David Skarbek, and Graham Denyer Willis for their helpful suggestions and comments.
-We are also grateful to Mark S.
-Bell for sharing his \texttt{R} code to estimate penalised-likelihood models.
-All replication materials are available at \href{https://github.com/danilofreire/mass-killings}{\texttt{https://github.com/danilofreire/mass-killings}}.}}
+\title{What Drives State-Sponsored Violence?: Evidence from Extreme Bounds Analysis and Ensemble Learning Models}
 
-\author{Danilo Freire\thanks{Postdoctoral Research Associate, The Political Theory Project, Brown University, 8 Fones Alley, Providence, RI 02912,  \href{mailto:danilofreire@gmail.com}{\texttt{danilofreire@gmail.com}},  \href{http://danilofreire.com}{\texttt{http://danilofreire.com}}.
-Corresponding author.} \and Gary Uzonyi\thanks{Assistant Professor, Department of Political Science; Research Fellow, Howard H.
-Baker Jr.
-Center for Public Policy, University of Tennessee, 1640 Cumberland Ave, Knoxville, TN 37996, \href{mailto:guzonyi@utk.edu}{\texttt{guzonyi@utk.edu}}, \href{https://sites.google.com/site/uzonyigary/}{\texttt{https://sites.google.com/site/uzonyigary}}.}}
+\author{Danilo Freire}
 
@@
 Although scholars have suggested a number of potential correlates of mass killings, it remains unclear whether the estimates are robust to different model specifications, or which variables accurately predict the onset of large-scale violence.
-We employ extreme bounds analysis and random forests to test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+I employ extreme bounds analysis and random forests to test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
 The results help clear the brush around mass killings, as few variables in this literature are robust determinants of atrocity.
@@
 However, there is little consensus on what factors influence the level of threat or costs a regime faces.
-Part of the reason for this uncertainty is that scholars use very different model specifications when testing their arguments, thus small changes in model parameters could influence the robustness of empirical results and the inferences we draw from these findings.
+Part of the reason for this uncertainty is that scholars use very different model specifications when testing their arguments, thus small changes in model parameters could influence the robustness of empirical results and the inferences I draw from these findings.
 
 To overcome these limitations and provide a better understanding of government atrocity, I employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities.
-Our approach is similar to \cite{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but we provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results \citep{bell2015examining,jones2015exploratory,jones2018there,muchlinski2015comparing}.
-In conducting this analysis, we address three debates in the mass violence literature:
+My approach is similar to Hegre and Sambanis' \citeyear{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results \citep{bell2015examining,jones2015exploratory,jones2018there,muchlinski2015comparing}.
+In conducting this analysis, I address three debates in the mass violence literature:
 
@@
 
-Our analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
-Our findings partially confirm previous research -- unstable countries are more likely to witness the regime employ atrocity \citep[e.g.,][]{goldsmith2013forecasting,harff2003no,krain1997state}.
+My analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+My findings partially confirm previous research -- poor, unstable countries are more likely to witness the regime employ atrocity \citep[e.g.,][]{goldsmith2013forecasting,harff2003no,krain1997state}.
 However, many of the factors scholars often cite as observable indicators of such instability -- regime transitions, coups d'état, the presence of militias, etc.
@@
 Thus, policymakers may be looking for the incorrect signs of impending atrocity when seeking to prevent its onset.
-Furthermore, we find that the conclusions scholars draw regarding the likelihood of government atrocity largely depend on whether they combine peace and war years or just analyse periods of civil wars, as patterns in mass killings differ dramatically across these contexts.
-Lastly, we find that genocide and politicide follow vastly different patterns of onset than other forms of state-sponsored mass murder.
+Furthermore, I find that the conclusions scholars draw regarding the likelihood of government atrocity largely depend on whether they combine peace and war years or just analyse periods of civil wars, as patterns in mass killings differ dramatically across these contexts.
+Lastly, I find that genocide and politicide follow vastly different patterns of onset than other forms of state-sponsored mass murder.
 This is further evidence that different logics govern different forms of political violence \citep{stanton2013terrorism}.
@@
 Overall, these findings raise concerns about policy options for preventing violence against civilians.
-If our conclusion is that unstable countries are violent, then preventing atrocity likely requires significant investments of time and resources in state-building, which is often politically and practically unfeasible \citep{doyle2006making}.
+If my conclusion is that poor and unstable countries are violent, then preventing atrocity likely requires significant investments of time and resources in state-building, which is often politically and practically infeasible \citep{doyle2006making}.
 This analysis contributes significantly to the political violence literature by highlighting the parsimonious nature of the logic behind government atrocity and clearing away much of the empirical clutter surrounding this conclusion.
@@
 
-To conduct our analysis, we began by surveying the quantitative political science literature on the causes of government mass killing since Rummel's (\citeyear{rummel1995democracy}) seminal work on the subject.
-Counting only published works, we identified 45 articles which employed logit or probit models of mass killing onset in a global sample.
-We then included all variables that appeared in at least two of these papers in our data set at the country-year unit of analysis for all years from 1945 to 2013.
-The appendix provides a complete list of the articles we considered and a complete list of the variables we included in our models.
-Next, we estimated an extreme bounds analysis to determine which variables were the most robust in explaining the onset of government atrocity.
-Then, we estimated a distributed random forest analysis to see which of the variables best predicted the onset of these atrocities.
-In this section, we provide more detail on each of these estimation procedures before turning to the results of both analyses in the next section.
+To conduct my analysis, I began by surveying the quantitative political science literature on the causes of government mass killing since Rummel's (\citeyear{rummel1995democracy}) seminal work on the subject.
+Counting only published works, I identified 45 articles which employed logit or probit models of mass killing onset in a global sample.
+I then included all variables that appeared in at least two of these papers in the data set at the country-year unit of analysis for all years from 1945 to 2013.
+The appendix provides a complete list of the articles I considered and a complete list of the variables I included in my models.
+Next, I estimated an extreme bounds analysis to determine which variables were the most robust in explaining the onset of government atrocity.
+Then, I estimated a distributed random forest analysis to see which of the variables best predicted the onset of these atrocities.
+In this section, I provide more detail on each of these estimation procedures before turning to the results of both analyses in the next section.
 
@@
 
-Th first method we employ to test the robustness of the potential determinants of state-led violence is the extreme bounds analysis (EBA).
+The first method I employ to test the robustness of the potential determinants of state-led violence is the extreme bounds analysis (EBA).
 Researchers have employed EBA to assess the sensitivity of the determinants of civil war \citep{hegre2006sensitivity}, coups d'état \citep{gassebner2016expect}, democratisation \citep{gassebner2013extreme}, economic growth \citep{levine1992sensitivity, sala1997just}, nuclear deterrence \citep{bell2015examining}, and political repression \citep{hafner2005right}.
 The method is particularly useful when there is no consensus about which covariates belong in the ``true'' regression model \citep[178]{sala1997just} and scholars worry that omitted or unnecessary predictors can bias the model estimates \citep[60]{angrist2008mostly, clarke2005phantom, elwert2014endogenous, spector2011methodological}.
+
 More specifically, the main purpose of EBA is to estimate the distribution of coefficients of each predictor $x$ in an exhaustive combination of regression models with $y$ as a dependent variable.
@@
 \cite{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response.
-In this paper, we use Sala-i-Martin's \citeyear{sala1997just} more flexible version of EBA and consider the whole range of values of $CDF(0)$.
-We choose to use the whole distribution because the aggregate $CDF(0)$ allows us to move away from a binary indicator of robustness and present the estimations with their appropriate degrees of confidence.
-Our focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero.
-We do not assume that the CDFs are normally distributed and use Sala-i-Martin's generic model instead.\footnote{The generic model provides a better fit to our data.
-Histograms for all coefficients are available in online appendix.} We specify our models as follows:
+In this paper, I use Sala-i-Martin's \citeyear{sala1997just} more flexible version of EBA and consider the whole range of values of $CDF(0)$.
+I choose to use the whole distribution because the aggregate $CDF(0)$ allows researchers to move away from a binary indicator of robustness and present the estimations with their appropriate degrees of confidence.
+My focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero.
+I do not assume that the CDFs are normally distributed and use Sala-i-Martin's generic model instead.\footnote{The generic model provides a better fit to the data.
+Histograms for all coefficients are available in online appendix.} I specify the models as follows:
 
@@
 
-Our main dependent variable is \textit{Mass Killing Onset}, which denotes the onset of government-sponsored killings.
+My main dependent variable is \textit{Mass Killing Onset}, which denotes the onset of government-sponsored killings.
 \citet[2]{ulfelder2008assessing} define a mass killing as ``any event in which the actions of state agents result in the intentional death of at least 1,000 noncombatants from a discrete group in a period of sustained violence''.
@@
 $M$ is a set of three covariates that are included in every model due to their prominence in the literature \citep{levine1992vale}.
-In our analysis, $M$ includes the natural logarithm of the GDP per capita to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence.
+In my analysis, $M$ includes the natural logarithm of the GDP per capita to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence.
 $F$ denotes a vector of variables of interest, and $Z$ is a vector of other control variables in addition to those included in $M$.
 $v$ is the error term.
-In practice, however, since we are interested in the effect of all variables in the data set and do not have true control variables, except from $M$, $F$ and $Z$ are interchangeable.
-We thus only use this notation to help clarify the connection of our analysis to previous conflict scholars who employed similar extreme bounds analysis \citep[e.g.,][]{hegre2006sensitivity,gassebner2016expect}.
-Following \citet[514]{hegre2006sensitivity}, we lagged the independent variables one year to reduce the risk of endogeneity.
+In practice, however, since this research is interested in the effect of all variables in the data set and do not have true control variables, except from $M$, $F$ and $Z$ are interchangeable.
+I thus only use this notation to help clarify the connection of my analysis to previous conflict scholars who employed similar extreme bounds analysis \citep[e.g.,][]{hegre2006sensitivity,gassebner2016expect}.
+Following \citet[514]{hegre2006sensitivity}, I lagged the independent variables one year to reduce the risk of endogeneity.
 
-Although our dependent variable is dichotomous, we use linear probability models in our main analysis.
+Although the dependent variable is dichotomous, I use linear probability models in my main analysis.
 \citet[298]{gassebner2016expect} argue that linear probability models are less prone to convergence problems and their results can be readily interpreted.
-Since the data are grouped into countries, we also use cluster-robust standard errors.
+Since the data are grouped into countries, I also use cluster-robust standard errors.
 
-As a precaution against collinearity, we place a limit on the Variance Inflation Factor (\textit{VIF}) of all regression coefficients.
+As a precaution against collinearity, I place a limit on the Variance Inflation Factor (\textit{VIF}) of all regression coefficients.
 The VIF estimates how much of the variance of each predictor is dependent on the other covariates in a model.
 A VIF of 1 indicates that the predictor is uncorrelated with the remaining covariates.
-VIF limits are often arbitrary \citep{bell2015examining,o2007caution}, thus here we use a moderately conservative VIF of 7.
-As robustness tests, we run the same models without restriction and with different cut-offs.
+VIF limits are often arbitrary \citep{bell2015examining,o2007caution}, thus here I use a moderately conservative VIF of 7.
+As robustness tests, I run the same models without restriction and with different cut-offs.
 
@@
 The second is \textit{interstate war}, a binary covariate measuring if the country is at war in a given year \citep{sarkees2010resort}.
-We have decided to omit democracy because of its evident correlation with the Polity measure and interstate war due to its correlation with our dependent variables.
+I have decided to omit democracy because of its evident correlation with the Polity measure and interstate war due to its correlation with the dependent variables.
 EBA models do not converge otherwise.\footnote{This is a typical case of multicollinearity and conceptual overlap.
 \citet{hlavac2016eba} suggests specifying a set of mutually exclusive variables to avoid the issue.
-However, as the Polity index is one of our core variables, we decided to drop the binary democracy indicator and use the continuous measure as it provides more details about the effect of political regimes on mass killings.
+However, as the Polity index is one of the core variables, I decided to drop the binary democracy indicator and use the continuous measure as it provides more details about the effect of political regimes on mass killings.
 More information available in the appendix.}  Since this problem does not affect machine learning algorithms, the two variables were included in the second set of estimations.
 
-Lastly, we depart slightly from Sala-i-Martin's suggested method and do not assign weights to EBA.
-Although he recommends using goodness-of-fit measures to construct regression weights, we agree with \citet{sturm2002robust} and \citet[299]{gassebner2016expect} and use the unweighted version of the CDF instead.
+Lastly, I depart slightly from Sala-i-Martin's suggested method and do not assign weights to EBA.
+Although he recommends using goodness-of-fit measures to construct regression weights, I follow \citet{sturm2002robust} and \citet[299]{gassebner2016expect} and use the unweighted version of the CDF instead.
 Goodness-of-fit indicators are not equivalent to the probability of a given model being true \citep{anscombe1973graphs,king1986not}, and the weights constructed this way are not invariant to transformations in the dependent variable.
-Moreover, our data set has a number of missing observations, so model comparison measures could be misleading \citep{lall2016multiple}.
+Moreover, the data set has a number of missing observations, so model comparison measures could be misleading \citep{lall2016multiple}.
 
@@
 
-We also make use of random forest (RF) \citep{breiman2001random} to evaluate whether the empirical results discussed in the mass killings literature are not driven by rigid parametric assumptions and model specifications.
+I also make use of random forest \citep{breiman2001random} to evaluate whether the empirical results discussed in the mass killings literature are not driven by rigid parametric assumptions and model specifications.
 Random forest is a machine learning algorithm that consists of a combination of individual decision trees.
@@
 
-In this paper we use distributed random forest (DRF) to model our data, a slightly modified version of the original random forest algorithm \citep{h2o2017}.
-The DRF has two additional features that are useful for our purposes.
+In this paper I use distributed random forest (DRF) to model the data, a slightly modified version of the original random forest algorithm \citep{h2o2017}.
+The DRF has two additional features that are useful for my purposes.
 Firstly, DRF is optimised for big data, as it grows decision trees on separate cores to speed up computation time.
@@
 
-To address these issues, we perform a grid search to select the most accurate random forest model \citep[123]{cook2017h2o}.
-We estimate a model for every possible combination of the hyperparameter space to make sure our results are robust to different specifications.
-For model selection, we follow the literature on predictive political science and use the area under the ROC curve (AUC) as our model evaluation metric \citep[e.g.,][]{hill2014empirical,ward2010perils,ward2013learning}.
+To address these issues, I perform a grid search to select the most accurate random forest model \citep[123]{cook2017h2o}.
+I estimate a model for every possible combination of the hyperparameter space to make sure the model results are robust to different specifications.
+For model selection, I follow the literature on predictive political science and use the area under the ROC curve (AUC) as the model evaluation metric \citep[e.g.,][]{hill2014empirical,ward2010perils,ward2013learning}.
 Models with higher AUC values are considered more accurate.
 
-We add several parameters to the grid search.
+I add several parameters to the grid search.
 The first is the number of independent trees to grow in each forest.
 The starting values are 256, 512, and 1024 trees.
-The machine learning literature does not provide a heuristic on how large a random forest should be, but \citet[166]{oshiro2012many} affirm that ``from 128 trees there is no more significant difference between the forests using 256, 512, 1024, 2048 and 4096 trees.'' We employ a more conservative approach and start from a higher value that the authors suggest as adding more trees do not reduce prediction accuracy \citep[7]{breiman2001statistical}.
+The machine learning literature does not provide a heuristic on how large a random forest should be, but \citet[166]{oshiro2012many} affirm that ``from 128 trees there is no more significant difference between the forests using 256, 512, 1024, 2048 and 4096 trees.'' I employ a more conservative approach and start from a higher value that the authors suggest as adding more trees do not reduce prediction accuracy \citep[7]{breiman2001statistical}.
 
@@
 Nevertheless, this complexity comes at the risk of overfitting, so deeper trees are not necessarily the most adequate solution for every model \citep[596]{friedman2001greedy,segal2004machine}.
-In this article, we let the algorithm decide among using 10, 20, or 40 levels for each tree.
+In this article, I let the algorithm decide among using 10, 20, or 40 levels for each tree.
 
-We test whether having balanced classes of our dependent variable (mass killing onset) affects the predictive ability of the model.
-Since the response measure is heavily imbalanced, oversampling the positive responses could potentially improve our results \citep{chawla2004special,del2014use,japkowicz2002class}.
-We also vary how many variables should be considered for each split in the data.
+I test whether having balanced classes of the dependent variable (mass killing onset) affects the predictive ability of the model.
+Since the response measure is heavily imbalanced, oversampling the positive responses could potentially improve the results \citep{chawla2004special,del2014use,japkowicz2002class}.
+I also vary how many variables should be considered for each split in the data.
 The default option is to use $\sqrt{p}$, where $p$ is the number of columns in the data set.
-As we have 40 covariates of interest, we have selected 5, 6 and 7 variables per split.
+As I have 40 covariates of interest, I have selected 5, 6 and 7 variables per split.
 The DRF uses a majority voting procedure to select which variable is most important.
 Additionally, the algorithm chooses the percentage of the training set to be modelled by each tree.
-The default option is 63.2\%, but we include the options of using 50\% and 100\% of the data.
-Similarly, we give a range of options for choosing how many columns will be included in each tree.
+The default option is 63.2\%, but I include the options of using 50\% and 100\% of the data.
+Similarly, I give a range of options for choosing how many columns will be included in each tree.
 The algorithm can randomly choose among 50\%, 90\% or 100\% of the independent variables when estimating a decision tree.
 
-Finally, we use three types of histogram to find optimal split points for each independent variable.
+Finally, I use three types of histogram to find optimal split points for each independent variable.
 Decision trees consider every value of a given independent variable as a potential candidate for a split in the training data.
@@
 The DRF algorithm also offers the choice of randomly cycling through all histogram types, including one type in each tree estimation.
-We adopt this ``round robin'' arrangement as it is both efficient and parsimonious.
+I adopt this ``round robin'' arrangement as it is both efficient and parsimonious.
 
@@
 
-We endeavour to answer three questions in our analysis: (1) what are the robust predictors of government mass killing, (2) do these predictors differ when considering only cases of civil war, and (3) are genocide and politicide different than other forms of atrocity?
+I endeavour to answer three questions in this analysis: (1) what are the robust predictors of government mass killing, (2) do these predictors differ when considering only cases of civil war, and (3) are genocide and politicide different than other forms of atrocity?
 Table \ref{tab:eba01} summarises our main EBA results in answering Question 1.
 The table shows the average coefficient estimate of all regressions for each robust variable along with their mean standard deviations.\footnote{A list of all independent variables and coding rules are available in the appendix.} The table also displays the percentage of regressions that are statistically significant at the 90\% level.
-$CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{We show whichever area is the largest.
-The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is our main statistic of interest, and we consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}.
-Lastly, we report the number of estimated regressions models which included each variable.
+$CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{I show whichever area is the largest.
+The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is our main statistic of interest, and I consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}.
+Lastly, I report the number of estimated regressions models which included each variable.
 

### 2018-11-30 paper-killings/manuscript.tex
change pronouns

@@
 \usepackage[libertine]{newtxmath}
-\usepackage[FiraMono]
+\usepackage{FiraMono}
 \usepackage[authoryear]{natbib}
@@
 Ultimately, effective prevention requires us to understand why these atrocities occur.
-In this vein, the academic community has laboured tremendously to establish evidence-based theories as to why governments engage in brutality against their civilian populations.
+In this vein, the academic community has laboured tremendously to establish empirically-based theories as to why governments engage in brutality against their civilian populations.
 Indeed, since 1995, there have been over 45 quantitative political science articles focused on explaining government-sponsored killing of civilians.
@@
 To overcome these limitations and provide a better understanding of government atrocity, I employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities.
-My approach is similar to Hegre and Sambanis' \citeyear{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results \citep{bell2015examining,jones2015exploratory,jones2018there,muchlinski2015comparing}.
+My approach is similar to Hegre and Sambanis' \citeyear{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but like \citet{bell2015examining}, \citep{hill
+} and \citet{jones2018there}, I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results.
 In conducting this analysis, I address three debates in the mass violence literature:
@@
 
+More broadly, the goal of this study is to examine whether the quantitative scholarship is able to identify the main explanators of mass atrocities and, if positive, which explanators are more robust relative to others \citep[2]{bell2015examining}.
 My analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
@@
 As a result, random forest is able to effectively uncover complex, nonlinear interaction effects in the data without prespecification \citep{jones2015exploratory,jones2018there}.
+Random forest models complement the extreme bounds analysis in two ways; first, by providing robust additional tests to the parametric estimations, and also by pointing out eventual limitations of the widely-employed modelling techniques in the mass violence literature.
 
 In this paper I use distributed random forest (DRF) to model the data, a slightly modified version of the original random forest algorithm \citep{h2o2017}.
-The DRF has two additional features that are useful for my purposes.
+The DRF has two additional features that are useful for the purposes of this chapter.
 Firstly, DRF is optimised for big data, as it grows decision trees on separate cores to speed up computation time.
@@
 This process is notably time-consuming, and computation time can be significantly reduced at little loss of precision by taking discrete values of the predictor distribution.
-The DRF algorithm also offers the choice of randomly cycling through all histogram types, including one type in each tree estimation.
-I adopt this ``round robin'' arrangement as it is both efficient and parsimonious.
+The DRF algorithm also offers the choice of randomly cycling through all histogram types, including one of the types in each tree estimation.
+I adopt this ``round robin'' arrangement as it is both computationally efficient and methodologically parsimonious.
 
@@
 I endeavour to answer three questions in this analysis: (1) what are the robust predictors of government mass killing, (2) do these predictors differ when considering only cases of civil war, and (3) are genocide and politicide different than other forms of atrocity?
-Table \ref{tab:eba01} summarises our main EBA results in answering Question 1.
+Table \ref{tab:eba01} summarises the main EBA results in answering Question 1.
 The table shows the average coefficient estimate of all regressions for each robust variable along with their mean standard deviations.\footnote{A list of all independent variables and coding rules are available in the appendix.} The table also displays the percentage of regressions that are statistically significant at the 90\% level.
 $CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{I show whichever area is the largest.
-The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is our main statistic of interest, and I consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}.
+The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is my main statistic of interest, and I consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}.
 Lastly, I report the number of estimated regressions models which included each variable.
@@
 
-Seven variables pass our EBA criteria and three of them decrease the likelihood of mass killings.
+Seven variables pass the EBA criterion and three of them decrease the likelihood of mass killings.
 First, as widely suggested in the literature, the natural logarithm of GDP per capita is negatively associated with the onset of mass killings \citep[e.g.,][]{besanccon2005relative, easterly2006development,esteban2015strategic}.
@@
 Four variables are robustly and positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of government-sponsored violence.
-Onset and continuation of civil wars are correlated with mass killings, but only when we employ the UCDP measures of violent conflict.
-We find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}.
+Onset and continuation of civil wars are correlated with mass killings, but only when I employ the UCDP measures of violent conflict.
+I find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}.
 Former instances of political turmoil also have a positive coefficient in our models.
 Moreover, countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent \citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}.
-Our results also show that higher levels of ethnic diversity increase the likelihood of atrocities against civilians.
-Nevertheless, ethnic diversity does not pass all additional tests we implement below and the sturdiness of this finding remains open to question.
+The results also show that higher levels of ethnic diversity increase the likelihood of atrocities against civilians.
+Nevertheless, ethnic diversity does not pass all additional tests I implement below and the sturdiness of this finding remains open to question.
 
@@
 Figure \ref{fig:drfuv} presents the ten most important predictors of state-sponsored violence in our random forest models.
-Overall, the machine learning estimations have a good fit, with an AUC of about 0.83 in the validation sample.
+In general, the machine learning estimations have a good fit, with an AUC of about 0.83 in the validation sample.
 The algorithm confirms some of the main findings of EBA, yet they also show some interesting patterns.
@@
 Interestingly, several variables that are not robust explanators of mass killing in the EBA, have large importance in the machine learning estimates.
-Variables related to characteristics of the military forces are a good examples.
+Variables related to characteristics of the military forces are a good example.
 As we see below, parametrisation and interactions likely account for this difference.
@@
 In both cases, those relationships will be captured in the machine learning estimations but not in the extreme bounds analysis.
+This provides evidence that model specification is driving some of the results in the EBA.
 
@@
 These graphs are akin to marginal effect plots in correlation models and help clarify the directional effects of these variables over their entire range.
-For example, we are able to see that the effects of Years since Last Mass Killings and Polity IV are highly nonlinear, or that Log GDP per capita does not decrease the likelihood of mass killings after it reaches a value of 10.
+For example, one can see that the effects of Years since Last Mass Killings is highly nonlinear, or that Log GDP per capita does not decrease the likelihood of mass killings after it reaches values close to 9.
 
@@
 
-% One paragraph on the results
+One can also infer that authoritarian and mixed political regime are more likely to engage in mass violence than democratic countries, a result that is also supported by both the EBA and the specialised literature.
+The number of military personnel positively affects the likelihood of mass killings, yet this increase is counterbalanced by military expenditures.
+Together these results indicate that countries with large and poorly-funded armed forces have higher risks of mass violence.
 
@@
     \begin{center}
-    \includegraphics[width=\textwidth, height=8cm]{images/rf-cow-pd.pdf}
+    \includegraphics[width=\textwidth, height=8cm]{images/rf-eth-pd.pdf}
     \caption{Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. Data)}
@@
 
-To answer Question 3,  we estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide.
-We find that no variable appears significant in our EBA models for genocide or politicide onset in the full data set.
-When we limit our sample to civil war years, the Post-Cold War period is negatively correlated with the outcome when using the Correlates of War data set.
+To answer Question 3, I estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide.
+No variable appears significant in the EBA models for genocide or politicide onset in the full data set.
+When I limit the sample to civil war years, the Post-Cold War period is negatively correlated with the outcome when using the Correlates of War data set.
 Excluded population has a negative sign in more than 90\% of the models using both Correlates of War's and Cederman et al's (\citeyear{cederman2010ethnic}) indicators of conflict.
 Displaced population also has a negative effect in the Correlates of War data set.
-During ethnic conflicts, our dummy variable for political assassinations has a negative impact on the onset of genocides.
-Overall, from these EBA analyses, we conclude then that the significant covariates of genocide and politicide onset differ significantly from those of more general forms of government mass violence.
+During ethnic conflicts, the dummy variable for political assassinations has a negative impact on the onset of genocides.
+Overall, from these EBA analyses, one can conclude then that the significant covariates of genocide and politicide onset differ significantly from those of more general forms of government mass violence.
 Though, the opportunity story still receives some limited support in these models.
-However, the machine learning models using Harff's genocide and politicide data are comparable to the ones we present above, with a similar set of variables appearing in the random forest estimations.
+However, the machine learning models using Harff's genocide and politicide data are comparable to the ones I present above, with a similar set of variables appearing in the random forest estimations.
 These results once more highlight that while the mass killing literature struggles to identify correlates of atrocity that are robust across model specification, scholars have done a much better job at identifying variables that help predict both the onset of genocide/politicide and mass killings, more broadly.
@@
 
-We estimate a set of additional regressions to assess the robustness of our main findings.
-In regard to EBA, we include 10 variants of our original model.
-They largely confirm our prior results.
-First, we varied the number of covariates included in each regression to 3 and 5 while keeping the $M$ set of 3 control variables.
+I estimate a set of additional regressions to assess the robustness of the main findings.
+In regard to EBA, I include 10 variants of the original model.
+They largely confirm the prior results.
+First, I varied the number of covariates included in each regression to 3 and 5 while keeping the $M$ set of 3 control variables.
 The results are the same as those of the main model, except that ethnic fractionalisation and Polity IV squared become marginally significant with a CDF(0) of about 0.88.
-Second, we place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving our results.
-The two models with different values of VIF are identical to the model reported here, while in the model with no VIF restriction ethnic fractionalisation again fails to meet our threshold by a very small margin.
+Second, we place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving the results.
+The two models with different values of VIF are identical to the model reported here, while in the model with no VIF restriction ethnic fractionalisation again fails to meet the threshold by a very small margin.
 
-We also reestimate the models using logit and probit regressions.
-In order to deal with the issue of complete separation \citep{bell2015questioning,zorn2005solution} we follow \citet{gelman2008weakly} and add a weakly informative prior distribution to the coefficients.
+I also reestimate the models using logit and probit regressions.
+In order to deal with the issue of complete separation \citep{bell2015questioning,zorn2005solution} I follow \citet{gelman2008weakly} and add a weakly informative prior distribution to the coefficients.
 In both cases, the logarithm of GDP per capita, post-Cold War period, previous riots, and Polity IV squared remain significant.
@@
 In regard to random forests, grid searches are themselves a data-driven selection of many possible machine learning models, thus it is not strictly necessary to run a batch of additional tests.
-Nevertheless, we performed a series of grid searches using three different seeds obtained from \href{https://random.org}{\texttt{http://random.org}} to estimate how different starting numbers influence the model outcomes.
+Nevertheless, I performed a series of grid searches using three different seeds obtained from \href{https://random.org}{\texttt{http://random.org}} to estimate how different starting numbers influence the model outcomes.
 The output of those models are largely comparable.
-
-The results of each of these analyses are available in online appendix.
+The results of each of these analyses are available in the appendix.
 
@@
 
-In this paper, we apply extreme bounds analysis and distributed random forests to estimate the robustness and predictive ability of 40 variables that have been pointed out as potential determinants of mass killings.
-We find strong evidence that mass killings are unlikely to happen in rich, stable countries.
-Nevertheless, there is considerable heterogeneity in some of our results.
+In this chapter, I apply extreme bounds analysis and distributed random forests to estimate the robustness and predictive ability of 40 variables that have been pointed out as potential determinants of mass killings.
+I find strong evidence that mass killings are unlikely to happen in rich, stable countries.
+Nevertheless, there is considerable heterogeneity in some of the results.
 The findings point out that mass killings have different causes according to the context in which they erupt, so a general theory of state atrocities may obscure important details in our understanding of state killings.
@@
 
-Yet we see this diversity of outcomes under a positive light.
-Our results suggest new avenues for research, and we believe they also highlight the importance of scholars moving from simple cross-country regressions to methods that can yield more robust predictions.
+Yet one can see this diversity of outcomes under a positive light.
+The results above suggest new avenues for research, and they also highlight the importance of scholars moving from simple cross-country regressions to methods that can yield more robust predictions.
 For instance, why are mass killings in ethnic conflicts correlated with a different set of variables than in armed conflicts in general?

### 2018-11-30 paper-killings/manuscript.tex
small update

@@
 
-One can also infer that authoritarian and mixed political regime are more likely to engage in mass violence than democratic countries, a result that is also supported by both the EBA and the specialised literature.
+One can also infer that authoritarian and mixed political regimes are more likely to engage in mass violence than democratic countries, a result that is also supported by both the EBA and the specialised literature.
 The number of military personnel positively affects the likelihood of mass killings, yet this increase is counterbalanced by military expenditures.
-Together these results indicate that countries with large and poorly-funded armed forces have higher risks of mass violence.
+Taken together, these results indicate that countries with large and poorly-funded armed forces have higher risks of mass violence.
 
@@
 
+% Add a footnote about Sala-i-Martin's claim that the sample is unbiased. 
 I estimate a set of additional regressions to assess the robustness of the main findings.
@@
 The results are the same as those of the main model, except that ethnic fractionalisation and Polity IV squared become marginally significant with a CDF(0) of about 0.88.
-Second, we place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving the results.
+Second, I place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving the results.
 The two models with different values of VIF are identical to the model reported here, while in the model with no VIF restriction ethnic fractionalisation again fails to meet the threshold by a very small margin.

### 2018-11-30 paper-killings/manuscript.tex
upload

@@
 To overcome these limitations and provide a better understanding of government atrocity, I employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities.
-My approach is similar to Hegre and Sambanis' \citeyear{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but like \citet{bell2015examining}, \citep{hill
-} and \citet{jones2018there}, I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results.
+My approach is similar to Hegre and Sambanis' \citeyear{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but like \citet{bell2015examining}, \citep{hill2014empirical} and \citet{jones2018there}, I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results.
+More broadly, my goal is to examine whether the quantitative scholarship is able to identify robust explanators of mass atrocities and, if positive, how they compare to each other in explaining power.
+
+While recent studies have progressively adopted stronger identification strategies, the majority of the literature on mass killings is comprised of cross-country regressions.
+Large-$n$ analyses have benefits that are often overlooked.
+For instance, quantitative studies allow scholars to assess how generalisable are the results of specific mechanisms.
+These studies provide a safeguard against the perils of sampling on the dependent variable, and they increase the external validity of micro-level statistical or qualitative findings \citep{bell2015examining,king1994designing}.
+Additionally, cross-national comparisons show how structural factors condition immediate causes of mass violence.
+Atrocities are multicausal phenomena, and large-$n$ studies can point to important interactions that scholars would otherwise miss.
+In this specific study, there is the additional advantage of running all the models with the same dependent variable, what constitutes an effective method of replication of the original findings.
+
 In conducting this analysis, I address three debates in the mass violence literature:
@@
 
-More broadly, the goal of this study is to examine whether the quantitative scholarship is able to identify the main explanators of mass atrocities and, if positive, which explanators are more robust relative to others \citep[2]{bell2015examining}.
 My analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
@@
 
+\subsection{Main Model}%
+\label{sub:main_model}
+
 I endeavour to answer three questions in this analysis: (1) what are the robust predictors of government mass killing, (2) do these predictors differ when considering only cases of civil war, and (3) are genocide and politicide different than other forms of atrocity?
@@
 
+\subsection{Mass Killings during Civil Wars}%
+\label{sub:mass_killings_during_civil_wars}
+
 Table \ref{tab:ucdp} presents our EBA results when we restrict our analysis to only civil war years to answer Question 2.
@@
 For example, only the UCDP data suggests that the post-Cold War years see less barbarism than during the Cold War.
-
-\newpage
 
@@
 
+\subsection{Mass Killings in and after the Cold War}%
+\label{sub:mass_killings_in_and_after_the_cold_war}
+
+\subsection{Genocides and Politicides}%
+\label{sub:genocides_and_politicides}
+
 To answer Question 3, I estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide.

### 2018-12-03 paper-killings/manuscript.tex
add files

--- 2018-11-30 paper-killings/manuscript.tex
@@
 \renewcommand*{\backref}[1]{}
-\renewcommand*{\backrefalt}[4]{%
+\renewcommand*{\backrefalt}[4]{
 	\ifcase #1 (Not cited.)%
@@
 While some conflate the logic of these types of atrocities \citep[e.g.,][]{rummel1995democracy, valentino2004draining}, others claim genocide and politicide follow a different logic from other forms of government violence \citep{kalyvas2006logic,stanton2015regulating}.
-For discussion on these important differences in conceptualisation see \citep[]{straus2007second, finkel2012macro}} The international community has responded with an effort to prevent further state-sponsored mass murder by strengthening laws against war crimes, genocide, and crimes against humanity.
+For discussion on these important differences in conceptualisation see \citet[]{straus2007second} and \citet{finkel2012macro}.} The international community has responded with an effort to prevent further state-sponsored mass murder by strengthening laws against war crimes, genocide, and crimes against humanity.
 Furthermore, the United Nations established a Special Adviser on the Prevention of Genocide and recognised its members' responsibility to protect civilian populations within and outside their own borders.
@@
 To overcome these limitations and provide a better understanding of government atrocity, I employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities.
-My approach is similar to Hegre and Sambanis' \citeyear{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but like \citet{bell2015examining}, \citep{hill
-} and \citet{jones2018there}, I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results.
+My approach is similar to Hegre and Sambanis' (\citeyear{hegre2006sensitivity}) seminal analysis on the causes of civil war onset, but like \citet{bell2015examining}, \citet{hill2014empirical} and \citet{jones2018there}, I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results.
+More broadly, my goals are: 1) to examine whether the current quantitative scholarship is able to identify robust explanators of mass atrocities; 2) to evaluate how those variables compare to each other in explaining power.
+
+While recent studies of mass killings have progressively adopted stronger identification strategies, the majority of the literature consists of cross-country regressions.
+While the importance of micro-level causal designs have been largely discussed \citep[e.g.][]{angrist2008mostly,imbens2015causal}, large-$n$ analyses also have strong benefits that are often overlooked.
+For instance, quantitative studies allow scholars to assess the external validity of specific explanatory mechanisms.
+Additionally, these studies provide a safeguard against the perils of selecting on the dependent variable, a bias that can severely distort regression results \citep{bell2015examining,king1994designing}.
+Cross-national comparisons also show how structural factors condition immediate causes of mass violence.
+Atrocities are multicausal phenomena, and large-$n$ studies can point to interactions that scholars would otherwise miss.
+In this specific study, there is the additional advantage of running all the models with the same dependent variable, what constitutes an effective method of replication of the original findings.
+
 In conducting this analysis, I address three debates in the mass violence literature:
@@
 Leamer's criterion is intuitive, but other authors contend it is too strict for most social science applications.
-\cite{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response.
-In this paper, I use Sala-i-Martin's \citeyear{sala1997just} more flexible version of EBA and consider the whole range of values of $CDF(0)$.
+\citet{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response.
+In this paper, I use Sala-i-Martin's more flexible version of EBA and consider the whole range of values of $CDF(0)$.
 I choose to use the whole distribution because the aggregate $CDF(0)$ allows researchers to move away from a binary indicator of robustness and present the estimations with their appropriate degrees of confidence.
 My focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero.
-I do not assume that the CDFs are normally distributed and use Sala-i-Martin's generic model instead.\footnote{The generic model provides a better fit to the data.
+I do not assume that the CDFs are normally distributed and use Sala-i-Martin's generic model.\footnote{The generic model provides a better fit to the data.
 Histograms for all coefficients are available in online appendix.} I specify the models as follows:
@@
 
-I also make use of random forest \citep{breiman2001random} to evaluate whether the empirical results discussed in the mass killings literature are not driven by rigid parametric assumptions and model specifications.
+I also make use of random forest \citep{breiman2001random} to evaluate whether the empirical results in the mass killings literature are driven by parametric assumptions and model specifications.
 Random forest is a machine learning algorithm that consists of a combination of individual decision trees.
@@
 If the independent variable is continuous, the algorithm will simply choose the average value of the predictions as the best candidate; if the covariate is discrete, the majority class will be employed.
-The simple procedure of leaving out some data points and growing separate trees with a random subset of covariates is sufficient to eliminate the risk of overfitting \citep[9-10]{jones2015exploratory}.
+The simple procedure of leaving out some data points and growing separate trees with a random subset of covariates is sufficient to eliminate overfitting \citep[9-10]{jones2015exploratory}.
 
@@
 
-I endeavour to answer three questions in this analysis: (1) what are the robust predictors of government mass killing, (2) do these predictors differ when considering only cases of civil war, and (3) are genocide and politicide different than other forms of atrocity?
+\subsection{Main Model}%
+\label{sub:main_model}
+
+I endeavour to answer three questions in this analysis: 1) what are the robust predictors of government mass killing, 2) do these predictors differ when considering only cases of civil war, and 3) are genocide and politicide different than other forms of atrocity?
 Table \ref{tab:eba01} summarises the main EBA results in answering Question 1.
@@
 
+\subsection{Mass Killings in and after the Cold War}%
+\label{sub:mass_killings_in_and_after_the_cold_war}
+
+Lastly, I test the heterogeneity of the main findings with three sets of models.
+First, I analyse which factors increase the risk of mass killings during and after the Cold War period.
+Global dynamics may influence the cost-benefit calculus of state leaders, and consequently affect the likelihood of large-scale responses to internal threats.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{Cold War Period} &  &  &  &  &  \\
+Log GDP per capita & -0.018 & 0.009 & 83.204 & 0.9678 & 50000 \\
+Previous riots & 0.022 & 0.014 & 62.457 & 0.9031 & 8278 \\
+ &  &  &  &  &  \\
+\textit{Post-Cold War Period} &  &  &  &  &  \\
+Ethnic war onset & -0.024 & 0.011 & 89.608 & 0.9823 & 4850 \\
+Coup d'état & -0.022 & 0.011 & 89.200 & 0.9822 & 8602 \\ 
+Territory aims & -0.027 & 0.014 & 81.083 & 0.9653 & 8775 \\ 
+Displaced Population & -0.048 & 0.027 & 58.689 & 0.9392 & 8695 \\ \hline
+\end{tabular}
+\caption{EBA -- Mass Killings in and after the Cold War Period (Robust Variables Only)}
+\label{tab:coldwar}
+\end{table}
+
+Indeed we see different patterns in both periods.
+In the Cold War years, Log GDP per capita has a negative impact on mass killings, while instances of previous riots increase the likelihood of state-led atrocities.
+The results are in line with those of the pooled model.
+However, mass killings seem to follow a separate logic in the post-Cold War years.
+Four independent variables lower the risk of mass killings: ethnic war onset, wars fought for territorial aims, coups d'état, and the share of discriminated population.
+I interpret the results as showing that ethnic wars are fought by groups with similar capabilities, thus large-scale, one-sided violence is relatively rare.
+This also explains why atrocities are more likely to occur when the share of discriminated population is not very large.
+The models show that territorial wars lead to more mass killings than governmental conflicts, a findings which has been previously described in the literature \citep[240]{eck2007one}.
+Coups d'état are correlated with fewer atrocities as well, what stands in contrast with previous research \citep[10]{wayman2010explaining}.
+
+\subsection{Genocides and Politicides}%
+\label{sub:genocides_and_politicides}
+
 To answer Question 3, I estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide.

### 2018-12-04 paper-killings/manuscript.tex
add files

@@
 I find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}.
-Former instances of political turmoil also have a positive coefficient in our models.
+Former instances of political turmoil also have a positive coefficient in the models.
 Moreover, countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent \citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}.
@@
 
-Overall, the EBA indicates two patterns in answer to our first question on the causes of mass killing.
+Overall, the EBA indicates two patterns in answer to my first question on the causes of mass killing.
 Atrocity is (1) more likely when violence is already present, reducing the costs of escalating brutality and (2) is less likely as domestic and international constraints increase, increasing the costs of escalating violence further.
@@
 
-Figure \ref{fig:drfuv} presents the ten most important predictors of state-sponsored violence in our random forest models.
+Figure \ref{fig:drfuv} presents the ten most important predictors of state-sponsored violence in the random forest models.
 In general, the machine learning estimations have a good fit, with an AUC of about 0.83 in the validation sample.
@@
 Variables related to characteristics of the military forces are a good example.
-As we see below, parametrisation and interactions likely account for this difference.
+As seen below, parametrisation and interactions likely account for this difference.
 The linear model imposes a parametric structure to the covariate, and the relationship between an independent variable and the response may be a nonlinear one.
@@
 
-Table \ref{tab:ucdp} presents our EBA results when we restrict our analysis to only civil war years to answer Question 2.
-We consider three different codings of civil war: (1) the Uppsala Conflict Database Program  (\citeyear{allansson2017organized,gleditsch2002armed}), (2) the Correlates of War project \citep{sarkees2010resort}, and (3) ethnic civil war from \citet{cederman2010ethnic}.
-We find two important patterns.
+Table \ref{tab:ucdp} presents the EBA results when I restrict the analysis to only civil war years to answer Question 2.
+I consider three different codings of civil war: 1) the Uppsala Conflict Database Program \citeyear{allansson2017organized,gleditsch2002armed}), 2) the Correlates of War project \citep{sarkees2010resort}, and 3) ethnic civil war from \citet{cederman2010ethnic}.
+I find two important patterns.
 First, considering only civil war years provides a very different understanding of atrocity.
@@
 However, contrary to past expectation \citep{koren2017means}, militias have a negative impact on the likelihood of mass killings.
-Second, there is wide variation in which variables are robust depending on how we code civil war.
-Across our three codings, no variable is robust to all codings and only territorial aims and militias are robust to more than one coding.
-These results are concerning for scholars using correlation models, as they indicate that our understanding of atrocity, from null hypothesis testing, is largely dependent on which coding of civil war we use.
+Second, there is wide variation in which variables are robust depending on how scholars code civil war.
+Across the three codings I use here, no variable is robust to all codings and only territorial aims and militias are robust to more than one coding.
+These results are concerning for scholars using correlation models, as they indicate that our understanding of atrocity, from null hypothesis testing, is largely dependent on which coding of civil war researchers use.
 For example, only the UCDP data suggests that the post-Cold War years see less barbarism than during the Cold War.
@@
 
-When we analyse the three codings of civil war using our random forest analysis, we find further intricacies in the patterns of mass killing.
+When I analyse the three codings of civil war using random forest analysis, I find further intricacies in the patterns of mass killing.
 First, the machine learning estimates highlight a different set of variables than the EBA when analysing the UCDP and Ethnic War data.
@@
 Therefore, the machine learning estimates are not as dependent on the data set employed as are the EBA results.
-This is good news for scholars of mass killing because it indicates that while our parametric models do not produce robust findings across different civil war data sets, our nonlinear models are able to given us a consistent and clear picture of which factors place a country at the greatest risk for atrocity during civil war.
+This is good news for scholars of mass killing because it indicates that while the parametric models do not produce robust findings across different civil war data sets, the nonlinear models are able to given us a consistent and clear picture of which factors place a country at the greatest risk for atrocity during civil war.
 
@@
 
-Indeed we see different patterns in both periods.
+Indeed there are different patterns in both periods.
 In the Cold War years, Log GDP per capita has a negative impact on mass killings, while instances of previous riots increase the likelihood of state-led atrocities.
@@
 I interpret the results as showing that ethnic wars are fought by groups with similar capabilities, thus large-scale, one-sided violence is relatively rare.
-This also explains why atrocities are more likely to occur when the share of discriminated population is not very large.
+This also explains why atrocities are more likely to occur in countries where the share of discriminated population is not very large.
 The models show that territorial wars lead to more mass killings than governmental conflicts, a findings which has been previously described in the literature \citep[240]{eck2007one}.
 Coups d'état are correlated with fewer atrocities as well, what stands in contrast with previous research \citep[10]{wayman2010explaining}.
+
+The random forest estimations
 
@@
 
+As a last robustness test for the EBA, I ran the main model with peace years only; that is, only country-years in which the UCDP, COW and Cederman et al's measures of civil conflicts had a value of zero.
+The results are very similar to the original model, what indicates that the difference in the results is conditional on civil war years.
+
 In regard to random forests, grid searches are themselves a data-driven selection of many possible machine learning models, thus it is not strictly necessary to run a batch of additional tests.

### 2018-12-04 chapters/02-synth.tex
beta 2! :)
maybe final version

@@
-\chapter{Evaluating the Effect of Homicide Prevention Strategies in São Paulo, Brazil: A Synthetic Control Approach\footnote{Forthcoming in \textit{Latin American Research Review}, 53(2), 2018.
-All data, code, and information required to replicate this study are available at \href{https://github.com/danilofreire/homicides-sp-synth}{https://github.com/danilofreire/homicides-sp-synth}.}}
+\chapter{Evaluating the Effect of Homicide Prevention Strategies in São Paulo, Brazil: A Synthetic Control Approach}
 \label{chap:synth}
@@
 New research could provide insights into how public policies work and, hopefully, help public authorities to design more effective policies against crime.
+
+\newpage
 

### 2018-12-04 chapters/04-killings.tex
maybe final version

@@
 Mass killing includes these atrocities, as well as attacks against civilians that result in at least 1,000 deaths but are not intended to destroy a particular group \citep[see][]{ulfelder2008assessing}.
-While some conflate the logic of these types of atrocities \citep[e.g.,][]{stanton2015regulating, valentino2004draining}, others claim genocide and politicide follow a different logic from other forms of government violence \citep{kalyvas2006logic,stanton2015regulating}.
-Note that these arguments focus on the logic of the violence, not legal definitions that separate and prioritise genocide as the crime of crimes \citep[see][]{schabas2000genocide}.} The international community has responded with an effort to prevent further state-sponsored mass murder by strengthening laws against crimes against humanity, genocide, and war crimes.
-It formed \textit{ad hoc}, hybrid, and a permanent criminal court to punish and deter future atrocities.
-Furthermore, the United Nations established a Special Adviser on the Prevention of Genocide and recognised its members’ responsibility to protect civilian populations within and outside their own borders.
+While some conflate the logic of these types of atrocities \citep[e.g.,][]{rummel1995democracy, valentino2004draining}, others claim genocide and politicide follow a different logic from other forms of government violence \citep{kalyvas2006logic,stanton2015regulating}.
+For discussion on these important differences in conceptualisation see \citet[]{straus2007second} and \citet{finkel2012macro}.} The international community has responded with an effort to prevent further state-sponsored mass murder by strengthening laws against war crimes, genocide, and crimes against humanity.
+Furthermore, the United Nations established a Special Adviser on the Prevention of Genocide and recognised its members' responsibility to protect civilian populations within and outside their own borders.
 Yet, such atrocities still occur.
@@
 
-Ultimately, successful prevention requires us to understand why these atrocities occur.
-In this vein, the academic community has laboured tremendously to establish evidence-based theories as to why governments engage in brutality against their civilian populations.
+Ultimately, effective prevention requires us to understand why these atrocities occur.
+In this vein, the academic community has laboured tremendously to establish empirically-based theories as to why governments engage in brutality against their civilian populations.
 Indeed, since 1995, there have been over 45 quantitative political science articles focused on explaining government-sponsored killing of civilians.
-Note, this number does not consider the myriad other books, qualitative articles, and non-political science research conducted into this matter.
-Overall, the mass violence literature generally agrees that as threat increases, so does the likelihood of atrocity, if the costs to such violence are not prohibitive.
+Overall, the mass violence literature agrees that government atrocity follows an opportunity logic: as threat increases, so does the likelihood of atrocity, if the costs to such violence are not prohibitive.
 However, there is little consensus on what factors influence the level of threat or costs a regime faces.
-Part of the reason for this uncertainty is that scholars use very different model specifications when testing their arguments.
-Examining the quantitative literature on government violence against civilians, we found that in 45 studies, scholars used nearly 180 measurements to capture roughly 30 key concepts related to threat and costs.
-While models should be constructed specifically to test particular arguments, one concern is that small changes in model specification could in influence the robustness of empirical results and the inferences that we can draw from these results.
-
-To overcome these limitations and provide a better understanding of government atrocity, we employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities.
-Our approach is similar to \cite{hegre2006sensitivity} seminal analysis on the causes of civil war onset, but we provide additional tests to check which variables are able to predict out-of-sample cases of mass violence during both peacetime and wartime.
-In conducting this analysis, we address three debates in the mass violence literature:
+Part of the reason for this uncertainty is that scholars use very different model specifications when testing their arguments, thus small changes in model parameters could influence the robustness of empirical results and the inferences I draw from these findings.
+
+To overcome these limitations and provide a better understanding of government atrocity, I employ extreme bounds analysis and random forests to identify the most robust determinants of state-sponsored atrocities.
+My approach is similar to Hegre and Sambanis' (\citeyear{hegre2006sensitivity}) seminal analysis on the causes of civil war onset, but like \citet{bell2015examining}, \citet{hill2014empirical} and \citet{jones2018there}, I provide additional tests to verify whether complex interactions and nonlinearities are driving the statistical results.
+More broadly, my goals are: 1) to examine whether the current quantitative scholarship is able to identify robust explanators of mass atrocities; 2) to evaluate how those variables compare to each other in explaining power.
+
+While recent studies of mass killings have progressively adopted stronger identification strategies, the majority of the literature consists of cross-country regressions.
+While the importance of micro-level causal designs have been largely discussed \citep[e.g.][]{angrist2008mostly,imbens2015causal}, large-$n$ analyses also have strong benefits that are often overlooked.
+For instance, quantitative studies allow scholars to assess the external validity of specific explanatory mechanisms.
+Additionally, these studies provide a safeguard against the perils of selecting on the dependent variable, a bias that can severely distort regression results \citep{bell2015examining,king1994designing}.
+Cross-national comparisons also show how structural factors condition immediate causes of mass violence.
+Atrocities are multicausal phenomena, and large-$n$ studies can point to interactions that scholars would otherwise miss.
+In this specific study, there is the additional advantage of running all the models with the same dependent variable, what constitutes an effective method of replication of the original findings.
+
+In conducting this analysis, I address three debates in the mass violence literature:
 
 \begin{enumerate}
-    \item Why do some governments engage in mass killings, genocides, or politicides?
-This is the primary question asked by advocates, policymakers, and scholars in this field of research.
-It is also the question that the 45 studies mentioned above each attempt to answer.
-    \item Does the logic underpinning government decision-making follow different patterns during peacetime and wartime.
-Recent research suggests that government atrocity occurs predominantly during periods of civil war \citep{harff2003no} which has led some scholars to restrict their analyses to only periods of civil war \citep[e.g.,][]{colaresi2008kill, valentino2004draining} or concentrate on predicting both the onset of civil war and atrocity \citep{goldsmith2013forecasting}.
-Yet, others estimate models of all country-year \citep[e.g.,][]{krain1997state, montalvo2008discrete}, raising questions of how well these studies speak to each other.
+ \item Why do some governments engage in mass killings, genocides, or politicides? This is the primary question asked by activists, policymakers, and scholars in this field of research. 
+ \item Does the logic underpinning government decision-making follow different patterns during peacetime and wartime?
+Recent research suggests that government atrocity occurs predominantly during periods of civil unrest \citep{harff2003no} which has led some scholars to restrict their analyses to only periods of civil war \citep[e.g.,][]{colaresi2008kill, valentino2004draining} or concentrate on predicting both the onset of civil war and atrocity \citep{goldsmith2013forecasting}.
+Yet, others estimate models of all country-years \citep[e.g.,][]{krain1997state, montalvo2008discrete}, raising questions of how well these studies speak to each other.
  \item Is there a difference in logic between those atrocities labelled as genocide or politicide, compared to other mass killings?
-While the Political Instability Task Force \citep{marshall2017pitf} provides the most widely used data on government atrocity, these data are limited only to cases of genocide and politicide.
-Others provide data with much more lenient inclusion criteria \citep[e.g.,][]{eck2007one, rummel1995democracy, stanton2015regulating, ulfelder2012forecasting}.
-These differences in definition of atrocity have led to divergent results, raising questions about important determinants of government behaviour \citep[for discussion, see][]{uzonyi2016domestic, wayman2010explaining}.
+While the Political Instability Task Force \citep{marshall2017pitf} provides the most widely used data on government atrocity, others provide data with much more lenient inclusion criteria \citep[e.g.,][]{stanton2015regulating, ulfelder2012forecasting}.
+These differences in definition of atrocity have led to divergent results, raising questions about important determinants of government behaviour \citep[for discussion, see][]{straus2007second, uzonyi2016domestic, wayman2010explaining}.
 \end{enumerate}
 
-Our analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
-We find that GDP per capita, stable political regimes, and the post-Cold War period are negatively associated with mass killings.
-Conversely, previous political turmoil, ethnic diversity, and civil wars increase the risk of such violence.
-The distributed random forest algorithm indicates that years since the last episode of mass violence, GDP per capita, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to our models' out-of-sample predictive power.
-While we discuss the importance of these findings in more detail later in this article, a few notes are imperative to make here.
-First, in some ways, these findings confirm previous research–unstable countries are more likely to witness the regime employ atrocity \citep[e.g.,]{goldsmith2013forecasting,harff2003no,krain1997state}.
-However, many of the factors scholars often cite as observable indicators of such instability--regime transitions, coups d'état, the presence of militias, etc.--are not good proxies for instability.
+My analysis tests the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+My findings partially confirm previous research -- poor, unstable countries are more likely to witness the regime employ atrocity \citep[e.g.,][]{goldsmith2013forecasting,harff2003no,krain1997state}.
+However, many of the factors scholars often cite as observable indicators of such instability -- regime transitions, coups d'état, the presence of militias, etc.
+-- are not good proxies for instability.
 Thus, policymakers may be looking for the incorrect signs of impending atrocity when seeking to prevent its onset.
-Second, these findings raise concerns about policy options for preventing violence against civilians.
-If our conclusion is that unstable countries are violent, then preventing atrocity likely requires significant investments of time and resources in state-building, which is often politically and practically unfeasible \citep{doyle2006making}.
-Lastly, our findings are robust to a myriad of specifications, functional forms, and additional tests as we describe below.
+Furthermore, I find that the conclusions scholars draw regarding the likelihood of government atrocity largely depend on whether they combine peace and war years or just analyse periods of civil wars, as patterns in mass killings differ dramatically across these contexts.
+Lastly, I find that genocide and politicide follow vastly different patterns of onset than other forms of state-sponsored mass murder.
+This is further evidence that different logics govern different forms of political violence \citep{stanton2013terrorism}.
+
+Overall, these findings raise concerns about policy options for preventing violence against civilians.
+If my conclusion is that poor and unstable countries are violent, then preventing atrocity likely requires significant investments of time and resources in state-building, which is often politically and practically infeasible \citep{doyle2006making}.
 This analysis contributes significantly to the political violence literature by highlighting the parsimonious nature of the logic behind government atrocity and clearing away much of the empirical clutter surrounding this conclusion.
@@
 \section{Empirical Methods}
-\label{sec:methods4}
-
-We employ two methods to test the robustness of the potential determinants of state-led violence.
-First, we use a variant of extreme bounds analysis (EBA) to check the statistical significance of 35 explanatory variables cited in the academic literature.\footnote{The complete list of variables included in this article is available in appendix \ref{sec:mk-appendix}.
-Our selection procedure was as follows.
-We included all variables that appeared in at least two two quantitative papers on mass killings which employed logit or probit models in a global sample.
-We considered only published articles in our list.} Researchers have employed the EBA to assess the sensitivity of the determinants of civil war \citep{hegre2006sensitivity}, coups d'état \citep{gassebner2016expect}, democratisation \citep{gassebner2013extreme}, economic growth \citep{levine1992sensitivity, sala1997just, sturm2005determinants}, nuclear deterrence \citep{bell2015examining}, political repression \citep{hafner2005right}, and state corruption \citep{serra2006empirical}.
-The method is particularly useful when there is no consensus about which covariates belong in the ``true'' regression model \citep[178]{sala1997just} and scholars worry that omitted or unnecessary predictors could bias the parameter estimates \citep[60]{angrist2008mostly, clarke2005phantom, elwert2014endogenous, spector2011methodological}.
-By combining a large set of regression models into a single posterior distribution, the EBA reduces the uncertainty associated with the \textit{ad hoc} selection of predictors and provides an intuitive way of assessing the relative strength of the statistical findings \citep{leamer1985sensitivity, sala1997just}.
-
-Second, we employ distributed random forest (DRF) \citep{breiman2001random, h2o2017} to measure the predictive power of our set of independent variables.
-Random forest has been widely used in the machine learning community and consistently ranks among the best algorithms for predicting structured data.\footnote{Random forests, gradient boosting machines, and neural networks have won the highest number of competitions on Kaggle, a crowd-sourced platform for predictive modelling \citep{carpenter2011may}.
-While data analysts prefer neural networks for problems with a high number of hidden features such as image recognition, random forest and gradient boosting are the top choices for dealing with tabular data.
-See: \href{https://www.kaggle.com/antgoldbloom/what-algorithms-are-most-successful-on-kaggle}{https://www.kaggle.com/antgoldbloom/what-algorithms-are-most-successful-on-kaggle} (access: December 2017).} DRF does not require distribution assumptions, can be used with any type of response variables, and are able to automatically detect nonlinear relationships between correlates \citep{fernandez2014we, hill2014empirical, jones2015exploratory, muchlinski2015comparing}.
-Moreover, unlike neural networks and other deep learning algorithms \citep{castelvecchi2016can,rojas2013neural,shwartz2017opening}, the output of DRF models can be interpreted meaningfully.
-DRF offers precise estimates of the predictive ability of each individual variable, and it also allows researchers to graphically visualise the marginal effect of the covariates on the response variable \citep{friedman2001greedy,friedman2001elements,goldstein2015peeking}.
+\label{sec:methods}
+
+To conduct my analysis, I began by surveying the quantitative political science literature on the causes of government mass killing since Rummel's (\citeyear{rummel1995democracy}) seminal work on the subject.
+Counting only published works, I identified 45 articles which employed logit or probit models of mass killing onset in a global sample.
+I then included all variables that appeared in at least two of these papers in the data set at the country-year unit of analysis for all years from 1945 to 2013.
+The appendix provides a complete list of the articles I considered and a complete list of the variables I included in my models.
+Next, I estimated an extreme bounds analysis to determine which variables were the most robust in explaining the onset of government atrocity.
+Then, I estimated a distributed random forest analysis to see which of the variables best predicted the onset of these atrocities.
+In this section, I provide more detail on each of these estimation procedures before turning to the results of both analyses in the next section.
 
@@
 
-The main purpose of the extreme bounds analysis is to estimate the distribution of coefficients of each predictor $x$ in an exhaustive combination of regression models with $y$ as a dependent variable.
-\cite[308]{leamer1985sensitivity} argued that the EBA could certify scholars that ``minor changes in the list of variables do not alter fundamentally the conclusions, nor does a slight re-weighting of observations, nor correction for dependence among observations, etcetera, etcetera.'' He proposed that ``sturdy'' variables are those whose minimum and maximum of their coefficient distribution have the same sign and are situated at a distance from zero.
+The first method I employ to test the robustness of the potential determinants of state-led violence is the extreme bounds analysis (EBA).
+Researchers have employed EBA to assess the sensitivity of the determinants of civil war \citep{hegre2006sensitivity}, coups d'état \citep{gassebner2016expect}, democratisation \citep{gassebner2013extreme}, economic growth \citep{levine1992sensitivity, sala1997just}, nuclear deterrence \citep{bell2015examining}, and political repression \citep{hafner2005right}.
+The method is particularly useful when there is no consensus about which covariates belong in the ``true'' regression model \citep[178]{sala1997just} and scholars worry that omitted or unnecessary predictors can bias the model estimates \citep[60]{angrist2008mostly, clarke2005phantom, elwert2014endogenous, spector2011methodological}.
+
+More specifically, the main purpose of EBA is to estimate the distribution of coefficients of each predictor $x$ in an exhaustive combination of regression models with $y$ as a dependent variable.
+\cite[308]{leamer1985sensitivity} proposed that ``sturdy'' variables are those whose minimum and maximum of their coefficient distribution have the same sign and are situated at a distance from zero.
 If we are to use the conventional value of $p < 0.05$, the mean of the variable coefficients' distribution should be located at least $1.96$ standard deviations away from zero.
@@
 Leamer's criterion is intuitive, but other authors contend it is too strict for most social science applications.
-\cite{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response.
-If a given variable of interest appears as both positive and negative in the literature, it is likely that after a large a number of regressions the predictor will change signs and therefore be deemed fragile \citep[179]{sala1997just}.
-Sala-i-Martin then offered a less stringent version of the EBA in which researchers analyse the full distribution of point estimates instead of relying only on the coefficients' extreme bounds.
-
-In this paper, we follow his advice and consider the whole range of values of $CDF(0)$.
-We choose to use the whole distribution because the aggregate $CDF(0)$ allows us to move away from a binary indicator of robustness, the one we obtain with Leamer's extreme bounds, and present the estimations with their appropriate degrees of confidence \citep[179]{sala1997just}\footnote{As \citet[179]{sala1997just} describes, ``if 95 percent of the density function for the estimates of $\beta_{1}$ lies to the right of zero and only 52 percent of the density function for $\beta_{2}$  lies to the right of zero, one will probably think of variable 1 as being more likely to be correlated with [a dependent variable $Y$] than variable 2.''}.
-Our main focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero.
-We do not assume that the CDF has a normal distribution: Many of our coefficients do not follow a standard distribution, so the generic model provides a better fit to our data.\footnote{Histograms for all coefficients are available in appendix \ref{sec:mk-appendix}.} We specify our models as follows:
+\citet{sala1997just} argued that Leamer's EBA would increase the number of false negatives; in other words, it would classify as fragile covariates that are truly associated with the response.
+In this paper, I use Sala-i-Martin's more flexible version of EBA and consider the whole range of values of $CDF(0)$.
+I choose to use the whole distribution because the aggregate $CDF(0)$ allows researchers to move away from a binary indicator of robustness and present the estimations with their appropriate degrees of confidence.
+My focus is the percentage of the variable's cumulative distribution function that is smaller or greater than zero.
+I do not assume that the CDFs are normally distributed and use Sala-i-Martin's generic model.\footnote{The generic model provides a better fit to the data.
+Histograms for all coefficients are available in online appendix.} I specify the models as follows:
 
@@
 
-Our main dependent variable is \textit{Mass Killing Onset}, which denotes the onset of government-sponsored killings.
-It was coded by \cite{ulfelder2008assessing}.
-The authors define a mass killing as ``any event in which the actions of state agents result in the intentional death of at least 1,000 noncombatants from a discrete group in a period of sustained violence'' \citep[2]{ulfelder2008assessing}.
-In our additional tests, we also estimate the models with an indicator of Genocide/Politicide Onset by \cite{harff2003no}.
-We prefer the first measure not only because it provides a clear numerical threshold that separates mass killings from low-level violence, but also because it focuses on the killing of members of a given community, not on the absolute number of deaths.
-
-The indices $i$ and $t$ indicate country and year, respectively.
+My main dependent variable is \textit{Mass Killing Onset}, which denotes the onset of government-sponsored killings.
+\citet[2]{ulfelder2008assessing} define a mass killing as ``any event in which the actions of state agents result in the intentional death of at least 1,000 noncombatants from a discrete group in a period of sustained violence''.
+Respectively, $i$ and $t$ indicate country and year.
 $M$ is a set of three covariates that are included in every model due to their prominence in the literature \citep{levine1992vale}.
-In our analysis, we have added the natural logarithm of the GDP per capita to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence.
+In my analysis, $M$ includes the natural logarithm of the GDP per capita to control for income, the Polity IV index to control for level of democracy, and a linear time trend since the last episode of government-led atrocity to account for temporal dependence.
 $F$ denotes a vector of variables of interest, and $Z$ is a vector of other control variables in addition to those included in $M$.
-Our first model includes the variable of interest, 3 additional control variables, and 3 covariates that appear in every regression.
 $v$ is the error term.
-In practice, however, since we are interested in the effect of all variables in the data set and we do not have true control variables except from $M$, in our case $F$ and $Z$ are interchangeable.
-Following \citet[514]{hegre2006sensitivity}, the independent variables have been lagged one year to reduce the risk of endogeneity.
-Although our dependent variable is dichotomous, we use linear probability models in our main analysis.
-As argued by \citet[298]{gassebner2016expect}, linear probability models are less prone to convergence problems, their estimation is faster, and their results can be readily interpreted.
-While we do employ logistic and probit models as robustness tests, we report the coefficients of the linear probability models as our primary estimates.
-Moreover, since the data are grouped into countries, we use cluster-robust standard errors to control for within-cluster error correlation.
-
-One of the shortcomings of our data is that some variables are strongly correlated.
-For instance, we include the Correlates of War's Composite Index of National Capability (\textit{CINC}), which is composed by six different variables \citep{cow2017cinc,singer1988reconstructing}.\footnote{For more information about the CINC index, please refer to the Correlates of War website: \href{http://www.correlatesofwar.org/data-sets/national-material-capabilities/national-material-capabilities-v4-0}{http://www.correlatesofwar.org/data-sets/national-material-capabilities/national-material-capabilities-v4-0} (access: December 2017).} Multicollinearity would induce bias to our estimates, so we add CINC and three of its components\footnote{The variables correspond to the ratio between the national and the global values.} -- military expenditure, military personnel, and total population -- as mutually exclusive variables.
-These predictors are not included in the same regression model to avoid collinearity issues.
-
-We add another set of mutually exclusive variables to reduce collinearity.
-There are three covariates that denote ongoing civil conflicts: one measured by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, another coded by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}.
-We include only one of these measures at a time.
-
-As a last precaution against collinearity, we place a limit on the Variance Inflation Factor (\textit{VIF}) of all regression coefficients.
+In practice, however, since this research is interested in the effect of all variables in the data set and do not have true control variables, except from $M$, $F$ and $Z$ are interchangeable.
+I thus only use this notation to help clarify the connection of my analysis to previous conflict scholars who employed similar extreme bounds analysis \citep[e.g.,][]{hegre2006sensitivity,gassebner2016expect}.
+Following \citet[514]{hegre2006sensitivity}, I lagged the independent variables one year to reduce the risk of endogeneity.
+
+Although the dependent variable is dichotomous, I use linear probability models in my main analysis.
+\citet[298]{gassebner2016expect} argue that linear probability models are less prone to convergence problems and their results can be readily interpreted.
+Since the data are grouped into countries, I also use cluster-robust standard errors.
+
+As a precaution against collinearity, I place a limit on the Variance Inflation Factor (\textit{VIF}) of all regression coefficients.
 The VIF estimates how much of the variance of each predictor is dependent on the other covariates in a model.
 A VIF of 1 indicates that the predictor is uncorrelated with the remaining covariates.
-The VIF limits are often arbitrary \citep{bell2015examining,o2007caution}, thus we use a moderately conservative VIF of 7 in our estimates.
-As robustness tests, we run the same models with different VIF cut-offs and without restriction.
-
-Two variables were omitted from the EBA models but included in the machine learning models below.
-The first is \textit{democracy}, a dummy variable that indicates whether the country is a has a Polity IV score equal or higher than 5.
+VIF limits are often arbitrary \citep{bell2015examining,o2007caution}, thus here I use a moderately conservative VIF of 7.
+As robustness tests, I run the same models without restriction and with different cut-offs.
+
+Two variables were omitted from EBA models but included in the machine learning estimations.
+The first is \textit{democracy}, a dummy variable that indicates whether the country has a Polity IV score equal or higher than 5.
 The second is \textit{interstate war}, a binary covariate measuring if the country is at war in a given year \citep{sarkees2010resort}.
-We have decided to omit democracy because of its obvious correlation with the Polity measure and interstate war due to its correlation with our dependent variables.
-The EBA models do not converge if those variables are added.
-Since this problem does not affect machine learning algorithms, the two variables were included in the second set of estimations.
-
-Lastly, we depart slightly from Sala-i-Martin's suggested method and do not assign weights to the EBA.
-Although he recommends using goodness-of-fit measures to construct regression weights, we agree with \citet{sturm2002robust} and \citet[299]{gassebner2016expect} and use the unweighted version of the CDF instead.
-Goodness-of-fit indicators are not equivalent to the probability of a given model being true \citep{achen1977measuring,anscombe1973graphs,king1986not}, and the weights constructed this way are not invariant to transformations in the dependent variable.
-Moreover, our data set has a number of missing observations, so that model comparison would be misleading \citep{lall2016multiple}.
-Thus, the results below are for unweighted estimations.
-
-\subsection{Distributed Random Forest}
+I have decided to omit democracy because of its evident correlation with the Polity measure and interstate war due to its correlation with the dependent variables.
+EBA models do not converge otherwise.\footnote{This is a typical case of multicollinearity and conceptual overlap.
+\citet{hlavac2016eba} suggests specifying a set of mutually exclusive variables to avoid the issue.
+However, as the Polity index is one of the core variables, I decided to drop the binary democracy indicator and use the continuous measure as it provides more details about the effect of political regimes on mass killings.
+More information available in the appendix.}  Since this problem does not affect machine learning algorithms, the two variables were included in the second set of estimations.
+
+Lastly, I depart slightly from Sala-i-Martin's suggested method and do not assign weights to EBA.
+Although he recommends using goodness-of-fit measures to construct regression weights, I follow \citet{sturm2002robust} and \citet[299]{gassebner2016expect} and use the unweighted version of the CDF instead.
+Goodness-of-fit indicators are not equivalent to the probability of a given model being true \citep{anscombe1973graphs,king1986not}, and the weights constructed this way are not invariant to transformations in the dependent variable.
+Moreover, the data set has a number of missing observations, so model comparison measures could be misleading \citep{lall2016multiple}.
+
+\subsection{Random Forests}
 \label{sub:drf}
 
+I also make use of random forest \citep{breiman2001random} to evaluate whether the empirical results in the mass killings literature are driven by parametric assumptions and model specifications.
 Random forest is a machine learning algorithm that consists of a combination of individual decision trees.
@@
 Unlike parametric methods such as ordinary least squares or logistic regressions, the analyst does not have to impose any distributional form to the data-generating process.
-As a result, random forest is able to effectively uncover complex, nonlinear interaction effects in the data without prespecification \citep{jones2015exploratory,strobl2007bias}.
-
-The algorithm can also deal with heavily imbalanced data.
-Researchers can balance the classes of their response variable by assigning different sampling probability to each category.
-This enables the algorithm to make accurate prediction for rare events without adding prior distributions or resorting to subjective modelling choices \citep{chen2004using,del2014use,muchlinski2015comparing}.
-
-In this paper we use distributed random forest (DRF) to model our data \citep{h2o2017}.
-The DRF is essentially identical to original random forest algorithm, but it has two additional features that are useful for our purposes.
+As a result, random forest is able to effectively uncover complex, nonlinear interaction effects in the data without prespecification \citep{jones2015exploratory,jones2018there}.
+Random forest models complement the extreme bounds analysis in two ways; first, by providing robust additional tests to the parametric estimations, and also by pointing out eventual limitations of the widely-employed modelling techniques in the mass violence literature.
+
+In this paper I use distributed random forest (DRF) to model the data, a slightly modified version of the original random forest algorithm \citep{h2o2017}.
+The DRF has two additional features that are useful for the purposes of this chapter.
 Firstly, DRF is optimised for big data, as it grows decision trees on separate cores to speed up computation time.
 Secondly, in DRF, non-observed cases are not assumed to be missing at random, but rather as values that contain information in themselves.
-When building decision trees, DRF treats the missing observations as a separate category that can go either left or right.
-This is a more conservative approach than assuming that missing cases fit into an underlying parametric distribution.\footnote{For more information about how the distributed random forest algorithm deals with missing observations, please refer to: \href{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html}{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html} (access: December 2017).}
-
-The DRF has a series of hyperparameters that can be tuned to improve the algorithm's predictive performance.
-For instance, users can control the number of decision trees in each iteration, how deep trees should grow, which performance metric to adopt, whether to balance classes or to use cross-validation, and many other options.
+The algorithm assumes that observations are missing for a reason, what is most likely the case with social science data \citep{lall2016multiple}.
+This is a more conservative approach than assuming that missing cases fit into an underlying parametric distribution.\footnote{For more information about how the distributed random forest algorithm deals with missing observations, please refer to: \href{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html}{\texttt{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/drf.html}} (access: December 2017).}
+
+The DRF has a series of hyperparameters that can be tuned to improve its predictive performance.
+For instance, users can control the number of decision trees in each iteration, how deep trees should grow, and many other options.
 The interaction between parameters is generally complex and may involve thousands of potential combinations.
-As an example, a researcher interested in four parameters with 10 possible values each would have to estimate 10,000 models before deciding which is the most efficient one.
-Also, machine learning parameters are sensitive to the data at hand; so what seems an optimal solution for one problem cannot be readily implemented in another data set \citep{genuer2008random,goldstein2010application,jones2015exploratory}.
-
-To address this issue, we have adopted an automated procedure to select the model parameters.
-We perform a grid search where the algorithm starts with a random combination of parameters, jumps to another randomly-chosen set, and then stops after it reaches a certain threshold \citep[123]{cook2017h2o}.
-We follow the literature on predictive political science and use the area under the ROC curve (AUC) as our model evaluation metric \citep[e.g.,][]{clayton2014will,hill2014empirical,ward2010perils,ward2013learning,weidmann2010predicting}.
-We set the metric as follows: If five random models had not increased the AUC by at least in 0.1\% comparing to the previous ones, the algorithm considers the result to be optimal.
-We set the maximum number of models to 1,000.
-
-We have added several parameters to the grid search.
-The first is the number of independent trees to grow in each random forest model.
+As an example, a researcher interested in four parameters with 10 values each would have to estimate 10,000 models before deciding which is the most efficient one.
+Also, machine learning parameters are sensitive to the data at hand, that is, an optimal solution for one problem cannot be readily implemented in another data set \citep{genuer2008random,goldstein2010application,jones2015exploratory}.
+
+To address these issues, I perform a grid search to select the most accurate random forest model \citep[123]{cook2017h2o}.
+I estimate a model for every possible combination of the hyperparameter space to make sure the model results are robust to different specifications.
+For model selection, I follow the literature on predictive political science and use the area under the ROC curve (AUC) as the model evaluation metric \citep[e.g.,][]{hill2014empirical,ward2010perils,ward2013learning}.
+Models with higher AUC values are considered more accurate.
+
+I add several parameters to the grid search.
+The first is the number of independent trees to grow in each forest.
 The starting values are 256, 512, and 1024 trees.
-The machine learning literature does not provide a heuristic on how large a random forest should be, but \citet[166]{oshiro2012many} affirm that ``from 128 trees there is no more significant difference between the forests using 256, 512, 1024, 2048 and 4096 trees.'' We employ a more conservative approach and start from a higher value that the authors suggest as adding more trees do not reduce prediction accuracy \citep[7]{breiman2001statistical}.
-However, we find little evidence that the number of trees makes a noticeable difference to the estimations.
+The machine learning literature does not provide a heuristic on how large a random forest should be, but \citet[166]{oshiro2012many} affirm that ``from 128 trees there is no more significant difference between the forests using 256, 512, 1024, 2048 and 4096 trees.'' I employ a more conservative approach and start from a higher value that the authors suggest as adding more trees do not reduce prediction accuracy \citep[7]{breiman2001statistical}.
 
@@
 Nevertheless, this complexity comes at the risk of overfitting, so deeper trees are not necessarily the most adequate solution for every model \citep[596]{friedman2001greedy,segal2004machine}.
-In this article, we let the algorithm decide among using 10, 20, or 40 levels for each tree.
-In our view, these number offer a good balance between parsimony and complexity.
-
-We test whether having balanced classes of our dependent variable (mass killing onset) affects the predictive ability of the model.
-Since the response measure is heavily imbalanced, oversampling the positive responses could potentially improve our results \citep{chawla2004special,del2014use,japkowicz2002class}.
-Somewhat unexpectedly, DRF had a better fit using unbalanced classes, which is the result we report in the next section.
-
-We also vary how many variables should be considered for each split in the data.
-The DRF's default is to use $\sqrt{p}$, where p is the number of columns in the data set.
-As we have 34 covariates of interest, we have selected 5, 6 and 7 variables per split.
+In this article, I let the algorithm decide among using 10, 20, or 40 levels for each tree.
+
+I test whether having balanced classes of the dependent variable (mass killing onset) affects the predictive ability of the model.
+Since the response measure is heavily imbalanced, oversampling the positive responses could potentially improve the results \citep{chawla2004special,del2014use,japkowicz2002class}.
+I also vary how many variables should be considered for each split in the data.
+The default option is to use $\sqrt{p}$, where $p$ is the number of columns in the data set.
+As I have 40 covariates of interest, I have selected 5, 6 and 7 variables per split.
 The DRF uses a majority voting procedure to select which variable is most important.
 Additionally, the algorithm chooses the percentage of the training set to be modelled by each tree.
-The default option is 63.2\%, but we include the options of using 50\% and 100\% of the data.
-Similarly, we give a range of options for choosing how many columns will be included in each tree.
+The default option is 63.2\%, but I include the options of using 50\% and 100\% of the data.
+Similarly, I give a range of options for choosing how many columns will be included in each tree.
 The algorithm can randomly choose among 50\%, 90\% or 100\% of the independent variables when estimating a decision tree.
 
-Finally, we use four types of histogram to find optimal split points for each independent variable.
+Finally, I use three types of histogram to find optimal split points for each independent variable.
 Decision trees consider every value of a given independent variable as a potential candidate for a split in the training data.
 This process is notably time-consuming, and computation time can be significantly reduced at little loss of precision by taking discrete values of the predictor distribution.
-The DRF algorithm offers four choices of histogram selection and we include all of them in our estimations.
+The DRF algorithm also offers the choice of randomly cycling through all histogram types, including one of the types in each tree estimation.
+I adopt this ``round robin'' arrangement as it is both computationally efficient and methodologically parsimonious.
 
@@
 
-Table \ref{tab:eba1} summarises our main EBA results with Ulfelder and Valentino's \citeyear{ulfelder2008assessing} \textit{Mass Killing Onset} as the dependent variable.
-The table shows the average coefficient estimate of all regressions for each robust variable along with their mean standard deviations.\footnote{A list of all independent variables and coding rules are available in the online appendix.} The table also displays the percentage of regressions that are statistically significant at the 90\% level.
-$CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{We show whichever area is the largest.
-The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is our main statistic of interest, and we consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}.
-Lastly, we report the number of estimated regressions models which included each variable.
-
-\vspace{1cm}
+\subsection{Main Model}%
+\label{sub:main_model}
+
+I endeavour to answer three questions in this analysis: 1) what are the robust predictors of government mass killing, 2) do these predictors differ when considering only cases of civil war, and 3) are genocide and politicide different than other forms of atrocity?
+Table \ref{tab:eba01} summarises the main EBA results in answering Question 1.
+The table shows the average coefficient estimate of all regressions for each robust variable along with their mean standard deviations.\footnote{A list of all independent variables and coding rules are available in the appendix.} The table also displays the percentage of regressions that are statistically significant at the 90\% level.
+$CDF(0)$ represents the cumulative distribution function, which is the area of the distribution that falls above or below zero.\footnote{I show whichever area is the largest.
+The sign of the average $\beta$ coefficient indicates if most of the cumulative distribution is located above or below zero.} This is my main statistic of interest, and I consider a covariate to be robust if it has a $CDF(0)$ of 0.9 or higher \citep[181]{sala1997just}.
+Lastly, I report the number of estimated regressions models which included each variable.
 
@@
 \caption{Extreme Bounds Analysis -- Mass Killings (Robust Variables Only)}
-\label{tab:eba1}
+\label{tab:eba01}
 \end{table}
 
-Seven variables pass our EBA criteria.
-Three variables decrease the likelihood of mass killings.
-First, as widely suggested in the literature, the natural logarithm of GDP per capita is robustly associated with the onset of mass killings
-\citep[e.g.,][]{besanccon2005relative, easterly2006development,esteban2015strategic}.
-In our models, GDP per capita has a negative effect on onset risk is statistically significant in 76\% of the models.
+Seven variables pass the EBA criterion and three of them decrease the likelihood of mass killings.
+First, as widely suggested in the literature, the natural logarithm of GDP per capita is negatively associated with the onset of mass killings \citep[e.g.,][]{besanccon2005relative, easterly2006development,esteban2015strategic}.
 Second, the post-Cold War years are correlated with lower levels of government violence.
-Despite eventual outbreaks of large-scale attacks against civilians \citep[e.g.,][]{prunier2005darfur,tatum2010genocide,waldorf2009revisiting}, after 1991 these events have become less prevalent.
-Indeed, this finding is in line with several studies that point to a general decline in violence over the last decades, including riots, civil wars, and urban crime \citep{eisner2003long,pinker2011better}.
-The third variable negatively correlated with mass killings is the squared term of the Polity IV political regime index.
-Although the linear Polity IV indicator does not reach our threshold for significance, about 90\% of the distribution area of Polity IV squared is located below zero.
-This finding points to a nonlinear relationship between political regime and mass killings, thus providing further evidence that democracy reduces state-sponsored violence \citep{rost2013will,rummel1995democracy} and that regimes that mix democratic with autocratic features have the highest risk of conflict \citep{hegre2001toward,mitchell2013domestic,muchlinski2014grievances,regan2010changing}.
-
-Four variables are positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of state-sponsored violence.
-The onset and continuation of civil wars are correlated with mass killings, but only when we employ the UCDP measures of violent conflict.
-We find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}.
-Coding choices likely account for this difference, and the result is robust to a myriad of specifications.
-Former instances of political turmoil also have a positive coefficient in our models.
-Countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent
-\citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}.
-Our results also show that a higher levels of ethnic diversity increase the likelihood of atrocities against civilians.
-Nevertheless, the variable does not pass all additional tests we implement below and the sturdiness of this finding remains open to question.
-The uncertainty about the impact of ELF on mass killings is also reflected in the literature.
-As \citet[237]{hoeffler2016development} points out, \citet{rummel1995democracy} reports that ELF is not statistically significant in his models, \citet{wood2014opportunities} finds a negative relationship and \citet{querido2009state} finds a positive one.
-Our analysis does not offer a definitive answer for this question either.
-
-A different set of variables reach significance if we only consider mass killings that occur during civil wars.
-We ran the models on three indicators of violent armed conflict.
-The first is the civil war measure by the Uppsala Conflict Database Program  (\citeyear{allansson2017organized,gleditsch2002armed}).
-Here, two variables meet our criteria for robustness, Post-Cold War years and conflicts with territorial aims, both negatively correlated with mass killing onset.
-When we limit our analysis to cases marked as civil wars by the Correlates of War project \citep{sarkees2010resort}, years since the last mass killing episode, previous riots, former human rights abuses, and ethnic diversity appear have a positive effect on the outcome.
-Rather surprisingly, the presence of militias has a negative effect on genocides and politicides.
-Finally, our model using ethnic civil war \citep{cederman2010ethnic} shows similar results: Militias have a negative impact on the likelihood of mass killings during ethnic conflict, as do conflicts over territorial aims.
-This finding is in contradiction with previous research \citep{koren2017means} and is deserving of further exploration.
-Overall, these results suggest two takeaways.
-First, the determinants of mass killing onset during peacetime and wartime are different from each other.
-While there are more restraints on atrocities during peacetime, these restraints fail once fighting begins.
-This raises particularly interesting questions about the pacifying nature of democracy.
-Second, how civil wars are classified significantly alters the results produced.
-This raises concerns for scholars about making claims based on only one data set.
-
-\vspace{1cm}
+Indeed, this finding is in line with several studies that point to a general decline in violence over the last decades, including riots, civil wars, and urban crime \citep{pinker2011better,straus2012wars,valentino2014we}.
+The third robust variable is the squared term of the Polity IV political regime index.
+This finding points to a nonlinear relationship between political regime and mass killings, thus providing further evidence that democracy reduces state-sponsored violence \citep{rost2013will,rummel1995democracy} and that regimes that mix democratic with autocratic features have the highest risk of conflict \citep{hegre2001toward,muchlinski2014grievances}.
+
+Four variables are robustly and positively associated with Ulfelder and Valentino's (\citeyear{ulfelder2008assessing}) indicator of government-sponsored violence.
+Onset and continuation of civil wars are correlated with mass killings, but only when I employ the UCDP measures of violent conflict.
+I find no effect for the variables compiled by the Correlates of War project or \cite{cederman2010ethnic}.
+Former instances of political turmoil also have a positive coefficient in the models.
+Moreover, countries with a previous history of riots are more prone to state violence, which suggests that government repression is path dependent \citep[e.g.,][]{gurr2000peoples,harff2003no,krain1997state,nyseth2017re}.
+The results also show that higher levels of ethnic diversity increase the likelihood of atrocities against civilians.
+Nevertheless, ethnic diversity does not pass all additional tests I implement below and the sturdiness of this finding remains open to question.
+
+Overall, the EBA indicates two patterns in answer to my first question on the causes of mass killing.
+Atrocity is (1) more likely when violence is already present, reducing the costs of escalating brutality and (2) is less likely as domestic and international constraints increase, increasing the costs of escalating violence further.
+These patterns support the dominant opportunity narrative in the literature.
+However, several of the variables commonly used to proxy opportunity, such as military size or regime change, are not robust predictors of atrocity.
+Thus, this analysis helps clear away much of the brush around the opportunity argument.
+
+Figure \ref{fig:drfuv} presents the ten most important predictors of state-sponsored violence in the random forest models.
+In general, the machine learning estimations have a good fit, with an AUC of about 0.83 in the validation sample.
+The algorithm confirms some of the main findings of EBA, yet they also show some interesting patterns.
+Only democracy and state capacity appear to be robust explanators of mass killing in both EBA and random forest models.
+These patterns further support the refrain that stable states tend to stay stable states.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=.8\textwidth, height=8.25cm]{images/rf-mk.pdf}
+    \caption{Distributed Random Forest -- Variable Importance (Scaled)}
+    \label{fig:drfuv}
+\end{figure}
+
+Interestingly, several variables that are not robust explanators of mass killing in the EBA, have large importance in the machine learning estimates.
+Variables related to characteristics of the military forces are a good example.
+As seen below, parametrisation and interactions likely account for this difference.
+The linear model imposes a parametric structure to the covariate, and the relationship between an independent variable and the response may be a nonlinear one.
+Also, variables can be relevant predictors only when in interaction with each other.
+In both cases, those relationships will be captured in the machine learning estimations but not in the extreme bounds analysis.
+This provides evidence that model specification is driving some of the results in the EBA.
+
+Figure \ref{fig:drfdpp} displays the partial dependence plots for the ten variables that the distributed random forests highlight as the most important explanators of mass killing onset.
+These graphs are akin to marginal effect plots in correlation models and help clarify the directional effects of these variables over their entire range.
+For example, one can see that the effects of Years since Last Mass Killings is highly nonlinear, or that Log GDP per capita does not decrease the likelihood of mass killings after it reaches values close to 9.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=8cm]{images/rf-mk-pd.pdf}
+    \caption{Distributed Random Forest -- Partial Dependence Plots}
+    \label{fig:drfdpp}
+\end{figure}
+
+One can also infer that authoritarian and mixed political regimes are more likely to engage in mass violence than democratic countries, a result that is also supported by both the EBA and the specialised literature.
+The number of military personnel positively affects the likelihood of mass killings, yet this increase is counterbalanced by military expenditures.
+Taken together, these results indicate that countries with large and poorly-funded armed forces have higher risks of mass violence.
+
+\subsection{Mass Killings during Civil Wars}%
+\label{sub:mass_killings_during_civil_wars}
+
+Table \ref{tab:ucdp} presents the EBA results when I restrict the analysis to only civil war years to answer Question 2.
+I consider three different codings of civil war: 1) the Uppsala Conflict Database Program \citeyear{allansson2017organized,gleditsch2002armed}), 2) the Correlates of War project \citep{sarkees2010resort}, and 3) ethnic civil war from \citet{cederman2010ethnic}.
+I find two important patterns.
+First, considering only civil war years provides a very different understanding of atrocity.
+Across these models, the only similarity with the full analysis is that mass killing is less likely post-Cold War.
+Instead, military factors, such military size and militias, and territorial war aims are the most robust predictors of atrocity once war begins.
+However, contrary to past expectation \citep{koren2017means}, militias have a negative impact on the likelihood of mass killings.
+Second, there is wide variation in which variables are robust depending on how scholars code civil war.
+Across the three codings I use here, no variable is robust to all codings and only territorial aims and militias are robust to more than one coding.
+These results are concerning for scholars using correlation models, as they indicate that our understanding of atrocity, from null hypothesis testing, is largely dependent on which coding of civil war researchers use.
+For example, only the UCDP data suggests that the post-Cold War years see less barbarism than during the Cold War.
 
@@
 Ethnic diversity (ELF) & 0.095 & 0.062 & 48.615 & 0.9000 & 17902 \\
- &  &  &  &  &  \\
+& & & & \\
 \textit{Cederman et al. Data} &  &  &  &  &  \\
@@
 
-The random forest models confirm some of the main findings of the EBA, yet they also show some interesting prediction patterns.
-Figure \ref{fig:drfuv} presents the six most important predictors of state-sponsored violence in out-of-sample tests.
-Overall, the machine learning estimations have a good fit, with an AUC of about 0.8 in the test samples.
-
-However, the findings should be interpreted with caution since variable importance metrics are sensitive to the choice of random seed numbers.
-But despite that expected variation, the results in the random forest models are quite stable even when estimated with different random seeds.
-
-\begin{figure}[h]
-\includegraphics[width=.9\textwidth, height=8cm]{images/drf.pdf}
-\caption{Distributed Random Forest -- Variable Importance (Scaled)}
-\label{fig:drfuv}
-\end{figure}
-
-We see that the most important predictor in our models is the natural log of GDP per capita.
-This confirms the findings of the EBA model.
-Years since the last mass killing episode has a nonlinear relationship with large-scale violence: The likelihood of renewed genocides is small in the first years, although it increases slightly over the next decades.
-The percentage of urban population exhibits a similar pattern.
-Rural areas are less prone to mass killings, yet the probability of such events rise according to a country’s level of urbanisation.
-We understand both GDP per capita and percentage of urban population as proxies for state capacity.
-Conducting a mass killing campaign demands some minimal level of military hierarchy and resources, and countries that are overwhelmingly rural are probably less able to amass the human and physical capital required to conduct such task.
-Furthermore, urban population tend to be less dispersed, what facilitates a large attack by government forces.
-
-\vspace{.5cm}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp.pdf}
-\caption{Distributed Random Forest -- Partial Dependence Plots}
-\label{fig:drfdpp}
-\end{center}
-\end{figure}
-
-The ethnic polarisation variable indicates that countries in which a large share of the population is excluded from politically relevant positions, the lower the probability of a mass killing.
-Groups that manage to control the state apparatus rarely face dangerous opposition movements, and as such they rarely resort to mass violence to impose their rule over the population.
-Next, the size of the military personnel substantially increases the likelihood of state-sponsored violence.
-While small and weak states tend to be more prone to civil war and other types of conflict \citep{collier2004greed,fearon2003ethnicity}, they are less likely to experience one-sided state violence.
-Lastly, the Polity IV index shows that mixed regimes have a higher likelihood of atrocity, and democracies tend to have lower levels of state-led violence.
-
-With respect to mass atrocities that occur during armed conflicts, we see that log GDP per capita and years since mass killings are again relevant predictors.
-Additionally, both covariates are among the 10 most important predictors in the three models.
-The results obtained with the UCDP civil war data are broadly similar to those of the previous model, but it highlights that total battle deaths is the predictor that has a noticeable impact on mass killings onset.
-The partial dependence graph shows a large positive effect along the curve, with a pronounced jump at around 200,000 deaths.
+When I analyse the three codings of civil war using random forest analysis, I find further intricacies in the patterns of mass killing.
+First, the machine learning estimates highlight a different set of variables than the EBA when analysing the UCDP and Ethnic War data.
+However, the COW EBA and machine learning analyses both highlight the importance of human rights, previous riots, and the time since the state last engaged in mass killing.
+Thus, the COW analysis provides the most stable picture of atrocity during civil war.
+It again highlights the important pattern of the Conflict Trap: violence breeds violence.
+Second, though, the three codings of civil war each highlight a very similar set of strong predictors of atrocity during conflict.
+Therefore, the machine learning estimates are not as dependent on the data set employed as are the EBA results.
+This is good news for scholars of mass killing because it indicates that while the parametric models do not produce robust findings across different civil war data sets, the nonlinear models are able to given us a consistent and clear picture of which factors place a country at the greatest risk for atrocity during civil war.
+
+Figures \ref{fig:drfdpp2}--\ref{fig:drfdpp4} display the partial dependence plots for the variables with the highest impact in each of the three civil war data sets.
 
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp2.pdf}
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=8cm]{images/rf-ucdp-pd.pdf}
     \caption{Partial Dependence Plots -- Mass Killings during Civil Wars (UCDP Data)}
     \label{fig:drfdpp2}
-\end{center}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}[H]
     \begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp3.pdf}
+    \includegraphics[width=\textwidth, height=8cm]{images/rf-cow-pd.pdf}
     \caption{Partial Dependence Plots -- Mass Killings during Civil Wars (COW Data)}
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
     \begin{center}
-\includegraphics[width=\textwidth, height=9cm]{images/drfdpp4.pdf}
+    \includegraphics[width=\textwidth, height=8cm]{images/rf-eth-pd.pdf}
     \caption{Partial Dependence Plots -- Mass Killings during Civil Wars (Cederman et al. Data)}
@@
 
-\newpage
-
-When using the COW data, we find that GDP per capita and total battle deaths have little influence in the likelihood of mass killings, and years since the last state-led atrocity increases the chances of a future event, a result we have also seen in the EBA models.
-As expected, higher levels of abuse against the physical integrity of citizens are associated with mass killings.
-Ethnic polarisation and the threshold of excluded population have a somewhat constant effect in our estimations, yet both show a sharp decline in probability when they reach their maximum levels.
-This provides more evidence to the idea that when certain groups are able to concentrate power, they tend not to resort to mass-scale violence as a political means.
-
-Our last model concerns the onset of ethnic conflicts as defined by \citet{cederman2010ethnic}.
-The graphs show ethnic wars operate with a different dynamics than other armed conflicts.
-We see that an increase in all the most important predictors have a large negative impact on the likelihood of mass killings, thus we conclude that poor, unstable countries are more prone mass killings during ethnic conflicts.
+\subsection{Mass Killings in and after the Cold War}%
+\label{sub:mass_killings_in_and_after_the_cold_war}
+
+Lastly, I test the heterogeneity of the main findings with three sets of models.
+First, I analyse which factors increase the risk of mass killings during and after the Cold War period.
+Global dynamics may influence the cost-benefit calculus of state leaders, and consequently affect the likelihood of large-scale responses to internal threats.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{Cold War Period} &  &  &  &  &  \\
+Log GDP per capita & -0.018 & 0.009 & 83.204 & 0.9678 & 50000 \\
+Previous riots & 0.022 & 0.014 & 62.457 & 0.9031 & 8278 \\
+ &  &  &  &  &  \\
+\textit{Post-Cold War Period} &  &  &  &  &  \\
+Ethnic war onset & -0.024 & 0.011 & 89.608 & 0.9823 & 4850 \\
+Coup d'état & -0.022 & 0.011 & 89.200 & 0.9822 & 8602 \\ 
+Territory aims & -0.027 & 0.014 & 81.083 & 0.9653 & 8775 \\ 
+Displaced Population & -0.048 & 0.027 & 58.689 & 0.9392 & 8695 \\ \hline
+\end{tabular}
+\caption{EBA -- Mass Killings in and after the Cold War Period (Robust Variables Only)}
+\label{tab:coldwar}
+\end{table}
+
+The EBA models show different patterns for both periods.
+In the Cold War years, Log GDP per capita has a negative impact on mass killings, while instances of previous riots increase the likelihood of state-led atrocities.
+The results are in line with those of the pooled model.
+However, mass killings seem to follow a separate logic in the post-Cold War years.
+Four independent variables lower the risk of mass killings: ethnic war onset, wars fought for territorial aims, coups d'état, and the share of discriminated population.
+I interpret the results as showing that ethnic wars are fought by groups with similar capabilities, thus large-scale, one-sided violence is relatively rare.
+This also explains why atrocities are more likely to occur in countries where the share of discriminated population is not very large.
+The models show that territorial wars lead to more mass killings than governmental conflicts, a findings which has been previously described in the literature \citep[240]{eck2007one}.
+Coups d'état are correlated with fewer atrocities as well, what stands in contrast with previous research \citep[10]{wayman2010explaining}.
+
+The random forests models also display some difference between the two periods, yet several variables appear in both estimations and in the main models presented above.
+The results for the Cold War model also classify Log GDP per capita and previous riots as important predictors of mass killings.
+As expected, the effect is negative for income and positive for past social upheavals.
+The variables that appear in the main model have similar distribution, such as the inverted-U relationship between the Polity IV index and mass killings, and the shape decline in atrocity risk when Log GDP per capita has a value of of 10.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth, height=8cm]{images/rf-coldwar-pd.pdf}
+    \caption{Partial Dependence Plots -- Mass Killings during the Cold War Period}
+    \label{fig:drfdpp2}
+\end{figure}
+
+\begin{figure}[H]
+    \begin{center}
+    \includegraphics[width=\textwidth, height=8cm]{images/rf-postcoldwar-pd.pdf}
+    \caption{Partial Dependence Plots -- Mass Killings after the Cold War Period}
+    \label{fig:drfdpp3}
+    \end{center}
+\end{figure}
+
+\subsection{Genocides and Politicides}%
+\label{sub:genocides_and_politicides}
+
+To answer Question 3, I estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide.
+No variable appears significant in the EBA models for genocide or politicide onset in the full data set.
+When I limit the sample to civil war years, the Post-Cold War period is negatively correlated with the outcome when using the Correlates of War data set.
+Excluded population has a negative sign in more than 90\% of the models using both Correlates of War's and Cederman et al's (\citeyear{cederman2010ethnic}) indicators of conflict.
+Displaced population also has a negative effect in the Correlates of War data set.
+During ethnic conflicts, the dummy variable for political assassinations has a negative impact on the onset of genocides.
+Overall, from these EBA analyses, one can conclude then that the significant covariates of genocide and politicide onset differ significantly from those of more general forms of government mass violence.
+Though, the opportunity story still receives some limited support in these models.
+However, the machine learning models using Harff's genocide and politicide data are comparable to the ones I present above, with a similar set of variables appearing in the random forest estimations.
+These results once more highlight that while the mass killing literature struggles to identify correlates of atrocity that are robust across model specification, scholars have done a much better job at identifying variables that help predict both the onset of genocide/politicide and mass killings, more broadly.
 
@@
 
-We estimate a set of additional regressions to assess the robustness of our main findings.
-In regard to the EBA, we include 10 variants of our original model.
-All of them largely confirm our prior results.
-First, we varied the number of covariates included in each regression to 3 and 5 while keeping the $M$ set of 3 variables included in every model, that is, the natural logarithm of GDP per capita, the Polity IV index, and a linear time trend since the last episode of state-sponsored violence.
-The results are identical to the main model when we add 3 control variables at a time.
-The same covariates discussed above remain significant and with the same sign: log of GDP per capita (negative), post-Cold War period (negative), previous riots (positive), civil war onset and ongoing indicators as measured by the UCDP data set (both positive), Polity IV squared (negative), and ethnic diversity (negative).
-When 5 control variables are added to every model, three variables have a CDF(0) of 0.9 or higher.
-They are the natural logarithm of GDP per capita, civil war onset (UCDP), previous riots, and the dummy indicating the post-Cold War period.
-Ethnic fractionalisation and Polity IV squared become marginally significant with a CDF(0) of about 0.88.
-
-Second, we place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving our results.
-The test is similar to that employed by Bell (2015).
-While in the main analysis we use a VIF of 7, in the additional analyses we include a conservative VIF of 2.5, a less strict value of 10, and a model without any restriction.
-The results are unchanged.
-The two models with different values of VIF show identical results to the results displayed in table \ref{tab:eba1}.
-In the model with no VIF restriction, however, ethnic fractionalisation fails to meet our threshold by a very small margin.
-The CDF(0) of that covariate is 0.897, very close to the required value of 0.9.
-
-We also estimate our models using logit and probit regressions.
-In order to deal with the issue of complete separation \citep{bell2015questioning,zorn2005solution} we follow \citet{gelman2008weakly} and add a weakly informative prior distribution to the coefficients.
-We first scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then we added a Cauchy distribution with center 0 and scale 2.5.
-The probit regressions use a scale of $2.5 \times 1.6$, which is also recommended by the authors \citep{arm2017rpackage}.
-
+I estimate a set of additional regressions to assess the robustness of the main findings.\footnote{For computational purposes, I conducted all additional tests on 50,000 random draws from EBA's posterior distribution.
+\citet[819]{salaimartin2004determinants} argue that random draws from the full EBA models are unbiased.} In regard to EBA, I include 10 variants of the original model.
+They largely confirm the prior results.
+First, I varied the number of covariates included in each regression to 3 and 5 while keeping the $M$ set of 3 control variables.
+The results are the same as those of the main model, except that ethnic fractionalisation and Polity IV squared become marginally significant with a CDF(0) of about 0.88.
+Second, I place different restrictions on the variance inflation factor (VIF) to test whether multicollinearity is driving the results.
+The two models with different values of VIF are identical to the model reported here, while in the model with no VIF restriction ethnic fractionalisation again fails to meet the threshold by a very small margin.
+
+I also reestimate the models using logit and probit regressions.
+In order to deal with the issue of complete separation \citep{bell2015questioning,zorn2005solution} I follow \citet{gelman2008weakly} and add a weakly informative prior distribution to the coefficients.
 In both cases, the logarithm of GDP per capita, post-Cold War period, previous riots, and Polity IV squared remain significant.
-In contrast, ethnic fractionalisation and former conflict as measured by the UCDP data set are again close to the 0.9 significance level.
+
+As a last robustness test for the EBA, I ran the main model with peace years only; that is, only country-years in which the UCDP, COW and Cederman et al's dichotomous measures of civil conflicts are equal to zero.
+Despite some issues of multicollinearity,\footnote{Some independent variables were dropped from the models due to problems of collinearity.
+The bounds for my indicator of ``wars fought over territory'' could not be estimated, and the coefficients for ``number of battle deaths'' and ``presence of guerrillas'' are unreliable due to their sample sample size.
+The appendix contains the distribution of the coefficients.} the results are similar to the original model, what indicates that the difference in the estimations is conditional on civil war years.
 
 In regard to random forests, grid searches are themselves a data-driven selection of many possible machine learning models, thus it is not strictly necessary to run a batch of additional tests.
-Nevertheless, we performed a series of grid searches using three different seeds obtained from \texttt{Random.Org} to estimate how the choice of different starting numbers would influence the model outcomes.
+Nevertheless, I performed a series of grid searches using three different seeds obtained from \href{https://random.org}{\texttt{http://random.org}} to estimate how different starting numbers influence the model outcomes.
 The output of those models are largely comparable.
-
-As a last set of robustness models, we estimate the same regressions using Harff's (\citeyear{harff2003no}) indicator of genocide and politicide.
-No variable appear as significant in our EBA models for genocide or politicide onset in peacetime.
-When we limit our sample to civil war years, the Post-Cold War period is again negatively correlated with the outcome when using the Correlates of War data set.
-Additionally, excluded population has a negative sign in more than 90\% of the models using both Correlates of War’s and Cederman et al's (\citeyear{cederman2010ethnic}) indicators of conflict.
-Displaced population also has a negative effect in the Correlates of War data set.
-During ethnic conflicts, our dummy variable for political assassinations has a negative impact on the onset of genocides.
-Overall, we conclude then that the significant covariate of genocide and politicide onset differ significantly from those of more general forms of government mass violence.
-Finally, the machine learning models are comparable to the ones we present above, with a similar set of variables appearing in the random forest estimations.
-The results are available in appendix \ref{sec:mk-appendix}.
+The results of each of these analyses are available in the appendix.
 
 \section{Conclusion}
-\label{sec:conclusion4}
-
-In this paper, we apply extreme bounds analysis and distributed random forests to estimate the robustness and predictive ability of 40 variables that have been pointed out as potential determinants of mass killings.
-We find that from all variables we employ, GDP per capita is the covariate that appears statistically significant more often in our models.
-In the EBA estimations, we see that the level of democracy, and the post-Cold War period are negatively associated with our dependent variable, whereas ethnic diversity, civil wars, and previous political turmoils increase the likelihood of those events.
-When we run a machine learning algorithm to estimate out-of-sample predictive performance, GDP per capita and years since last mass killings again appear as important predictors, and variables related to military capabilities and the composition of the population are included in most models.
-In a nutshell, our tests provide strong evidence in favour of two well-established findings in the specialised literature: Mass killings are unlikely to happen in rich, stable countries and there is a positive association between several types of political conflict, such as previous riots, civil wars and government atrocities.
-
-Nevertheless, there is considerable heterogeneity in some of our findings.
-Mass killings are rare outcomes, so it is possible that our analysis does not capture all variables that explain the onset of large-scale violence, or that the effect of predictors vary across time and space \citep[8]{bell2015examining}.
-Additionally, the findings point out that mass killings may have different causes according to the context in which they erupt, so a general theory of state atrocities may obscure important details in our understanding of the root causes of state killings.
-
-Yet we see this diversity of outcomes under a positive light.
-The findings presented in this text suggest new avenues for research, and we believe they also highlight the importance of scholars moving from purely correlational, cross-country regressions to other methods that can yield more robust predictions and causal explanations.
+\label{sec:conclusion}
+
+In this chapter, I apply extreme bounds analysis and distributed random forests to estimate the robustness and predictive ability of 40 variables that have been pointed out as potential determinants of mass killings.
+I find strong evidence that mass killings are unlikely to happen in rich, stable countries.
+Nevertheless, there is considerable heterogeneity in some of the results.
+The findings point out that mass killings have different causes according to the context in which they erupt, so a general theory of state atrocities may obscure important details in our understanding of state killings.
+Moreover, mass killings are rare events, so local factors likely play an important role in their onset \citep{straus2007second,straus2012destroy}.
+
+Yet one can see this diversity of outcomes under a positive light.
+The results above suggest new avenues for research, and they also highlight the importance of scholars moving from simple cross-country regressions to methods that can yield more robust predictions.
 For instance, why are mass killings in ethnic conflicts correlated with a different set of variables than in armed conflicts in general?
@@
 Given that GDP per capita is negatively correlated to state atrocities in virtually every model, it would be interesting to unpack the causal mechanisms by which it operates by testing more specific mechanisms.
-Disaggregated data can also be of great help to further current research on genocides, politicide, and mass killings.
-
-In terms of practical implications, the results indicate that regime change and pro-growth economic policies are the most efficient ways to prevent mass killings.
+
+In terms of practical implications, the results indicate that democratisation and pro-growth economic policies are the most efficient ways to prevent mass killings.
 The international community can therefore play a role in deterring leaders from using force against their own population, either by offering support for domestic opposition groups, intervening, or by fostering economic development.
-Although costly in the short run, these measures would substantially decrease the likelihood of state violence by breaking the ``conflict trap'' in which past conflicts create the condition for new ones \citep{collier2003breaking}.
+Although costly in the short run, and sometimes violent during the transition, these measures would substantially decrease the likelihood of state violence by breaking the ``conflict trap'' in which past conflicts create the condition for new ones \citep{collier2003breaking}.
 
@@
 \textt{R} code can be found in subsection \ref{sec:mk-code} and the data are available on the following GitHub repository: \href{https://github.com/danilofreire/mass-killings}{https://github.com/danilofreire/mass-killings}.
-We used \texttt{R} version 3.4.4 (15-03-2018) and Ubuntu 16.04.4 LTS to perform all statistical calculations.
+I used \texttt{R} version 3.4.4 (15-03-2018) and Ubuntu 16.04.4 LTS to perform all statistical calculations.
 
@@
 
-We employ some criteria to select our explanatory variables.
-First, we included only published articles in our sample.
-Although working papers and policy may also provide important insights about the onset of mass killings, we believe that peer-reviewed research is probably better suited for our purposes.
-Also, we included only papers that use regression methods on a global sample and were published from 1995 to 2015.
-Our final sample comprises 45 articles: \citet{anderton2015new}, \citet{balcells2010rivalry, balcells2011continuation}, \citet{besanccon2005relative}, \citet{bulutgil2015social}, \citet{bundervoet2009livestock}, \citet{clayton2016civilianizing}, \citet{colaresi2008kill}, \citet{downes2006desperate, downes2007restraint},  \citet{easterly2006development}, \citet{eck2007one}, \citet{esteban2015strategic}, \citet{fazal2015particular}, \citet{fjelde2014weakening}, \citet{goldsmith2013forecasting}, \citet{harff2003no}, \citet{joshi2017kills}, \citet{kim2010makes}, \citet{kim2016revolutionary}, \citet{kisangani2007political}, \citet{koren2017means}, \citet{krain1997state}, \citet{manekin2013violence}, \citet{mcdoom2013killed,mcdoom2014predicting}, \citet{melander2009new}, \citet{montalvo2008discrete}, \citet{pilster2016differentiation}, \citet{querido2009state}, \citet{raleigh2012violence}, \citet{rost2013will}, \citet{rummel1995democracy}, \citet{schneider2013accounting}, \citet{siroky2015empire}, \citet{stanton2015regulating}, \citet{sullivan2012blood}, \citet{tir2008domestic}, \citet{ulfelder2008assessing}, \citet{ulfelder2012forecasting}, \citet{uzonyi2015civil, uzonyi2016domestic} \citet{valentino2004draining}, \citet{valentino2006covenants}, \citet{verpoorten2012leave}, \citet{wayman2010explaining}, \citet{wig2016local}, and \citet{yanagizawa2014propaganda}.
-
-We find that in those 45 studies scholars made use of nearly 180 measurements to capture roughly 30 key concepts related to threat and costs of mass killings.
+I employ some criteria to select our explanatory variables.
+First, I included only published articles in the sample.
+Although working papers and policy may also provide important insights about the onset of mass killings, peer-reviewed research is probably better suited for our purposes.
+Also, I included only papers that use regression methods on a global sample and were published from 1995 to 2015.
+The final sample comprises 45 articles: \citet{anderton2015new}, \citet{balcells2010rivalry, balcells2011continuation}, \citet{besanccon2005relative}, \citet{bulutgil2015social}, \citet{bundervoet2009livestock}, \citet{clayton2016civilianizing}, \citet{colaresi2008kill}, \citet{downes2006desperate, downes2007restraint},  \citet{easterly2006development}, \citet{eck2007one}, \citet{esteban2015strategic}, \citet{fazal2015particular}, \citet{fjelde2014weakening}, \citet{goldsmith2013forecasting}, \citet{harff2003no}, \citet{joshi2017kills}, \citet{kim2010makes}, \citet{kim2016revolutionary}, \citet{kisangani2007political}, \citet{koren2017means}, \citet{krain1997state}, \citet{manekin2013violence}, \citet{mcdoom2013killed,mcdoom2014predicting}, \citet{melander2009new}, \citet{montalvo2008discrete}, \citet{pilster2016differentiation}, \citet{querido2009state}, \citet{raleigh2012violence}, \citet{rost2013will}, \citet{rummel1995democracy}, \citet{schneider2013accounting}, \citet{siroky2015empire}, \citet{stanton2015regulating}, \citet{sullivan2012blood}, \citet{tir2008domestic}, \citet{ulfelder2008assessing}, \citet{ulfelder2012forecasting}, \citet{uzonyi2015civil, uzonyi2016domestic} \citet{valentino2004draining}, \citet{valentino2006covenants}, \citet{verpoorten2012leave}, \citet{wayman2010explaining}, \citet{wig2016local}, and \citet{yanagizawa2014propaganda}.
+
+In those 45 studies, scholars made use of nearly 180 measurements to capture roughly 30 key concepts related to threat and costs of mass killings.
 To be added to our models, a variable should appear in at least two articles.
@@
 
-We present a series of histograms with the coefficients' distribution of all variables in the main EBA model.
+I present a series of histograms with the coefficients' distribution of all variables in the main EBA model.
 There are 36 variables in total, seven of which are robust: Log GDP per capita, post-Cold War period, onset and ongoing civil wars (measured by the UCDP), previous riots, ethnic diversity and the squared term of the Polity IV index.
@@
 
-Next, we discuss genocides that occur during wartime.
-We use three covariates that denote ongoing civil conflicts: one by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, another by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}.
+Next, I discuss genocides that occur during wartime.
+I use three covariates that denote ongoing civil conflicts: one by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, another by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}.
 The variables that reach significance in this set of models below are notably different from those obtained in the main estimation.
@@
 Table \ref{tab:mk-5vars} presents the results for models with up 5 variables in each regressions.
-In contrast with our main EBA model, the indicators of UCDP ongoing civil wars, ethnic diversity, and Polity IV square drop out of significance.
+In contrast with the main EBA model, the indicators of UCDP ongoing civil wars, ethnic diversity, and Polity IV square drop out of significance.
 Their individual CDFs(0) are about 0.88, just marginally below our specified threshold of 0.9.
@@
 
-In this subsection, we estimate EBA models with different values of Variance Inflation Factor (VIF), which is a measure of multicollinearity.
+In this subsection, I estimate EBA models with different values of Variance Inflation Factor (VIF), which is a measure of multicollinearity.
 There is no standard definition about what constitutes an acceptable VIF value, although researchers often use 10 as rule of thumb to indicate strong multicollinearity \citep[674]{o2007caution}.
-Our original model used a slightly more conservative value of 7 as a cutoff.
-Here, we test the same model with VIF $=$ 10 (less strict), 2.5 (more conservative), and a model without VIF restrictions.
+My original model used a slightly more conservative value of 7 as a cutoff.
+Here, I test the same model with VIF $=$ 10 (less strict), 2.5 (more conservative), and a model without VIF restrictions.
 The results are essentially identical to those of the main model.
@@
 
-We reestimate the main EBA model with logit and probit models.
+I reestimate the main EBA model with logit and probit models.
 Nevertheless, logistic and probit regressions may have issues of complete separation, that is, some covariates may perfectly separate zeros and ones in the outcome variable.
 In that case, the estimations fail to converge.
-We address this problem by adding a weak prior to the regression coefficients as suggested by \citet{gelman2008weakly}.\footnote{We thank Mark Bell for sharing \texttt{R} code to estimate penalised-likelihood models.} First, we scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then added a Cauchy distribution with centre 0 and scale 2.5.
+We address this problem by adding a weak prior to the regression coefficients as suggested by \citet{gelman2008weakly}.\footnote{I thank Mark Bell for sharing \texttt{R} code to estimate penalised-likelihood models.} First, we scaled the non-binary variables to have a mean of 0 and a standard deviation of 0.5, then added a Cauchy distribution with centre 0 and scale 2.5.
 The probit regressions use a scale of $2.5 \times 1.6$, which is also recommended by the authors \citep{arm2017rpackage}.
@@
 
-\subsection{Random Forest Extensions}
-\label{sec:mk-rfe}
-
-\subsubsection{Alternative Random Seeds}
-
-As noted in section \ref{sec:methods4}, we perform a grid search to optimise the hyperparameters of the random forest models.
-The grid search evaluates a wide range of parameter values at once, therefore it is generally unnecessary to run additional tests to assess the robustness of the results.
-Nevertheless, as random forests themselves are an approximation to a number of possible parameter combinations, changes in seed numbers may influence the output.
-Thus, we start the models with different random seed numbers to evaluate how sturdy are our original results.\footnote{The numbers were generated at \href{https://www.random.org/}{https://www.random.org/}.} The findings holds quite well: Although variable importance changes from one model to another, the most significant variables appear repeatedly in the estimations.
-The marginal plots also show that their effects on the outcome variable remains similar despite eventual nonlinearities.
-We show the six most significant predictors of mass killings and their respective partial dependence plots.
-
-\begin{figure}[H]
-    \centering
-    \includegraphics{images/drf-mk2.pdf}
-    \caption{Variable Importance -- Seed 44849999}
-    \label{fig:my_label}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp2.pdf}
-    \caption{Partial Plots -- Seed 44849999}
-    \label{fig:my_label}
-\end{figure}
-
-\newpage
-
-Second model -- seed 1502436:
-
-\begin{figure}[H]
-    \centering
-    \includegraphics{images/drf-mk3.pdf}
-    \caption{Variable Importance -- Seed 1502436}
-    \label{fig:my_label}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp3a.pdf}
-    \caption{Partial Plots -- Seed 1502436}
-    \label{fig:my_label}
-\end{figure}
-
-\newpage
-
-\subsection{Genocides/Politicides}
-\label{sec:mk-other-variable}
+\subsubsection{Mass Killings in and after the Cold War}%
+\label{sub:mass_killings_in_and_after_the_cold_war}
+
+I also test the heterogeneity of the findings by running one model including only the Cold War years (1945--1991) and another with the post-Cold War period (1991--2013).
+The results vary in both periods, and there is no overlap between significant variables.
+
+\vspace{.5cm}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{Cold War Period} &  &  &  &  &  \\
+Log GDP per capita & -0.018 & 0.009 & 83.204 & 0.9678 & 50000 \\
+Previous riots & 0.022 & 0.014 & 62.457 & 0.9031 & 8278 \\
+ &  &  &  &  &  \\
+\textit{Post-Cold War Period} &  &  &  &  &  \\
+Ethnic war onset & -0.024 & 0.011 & 89.608 & 0.9823 & 4850 \\
+Coup d'état & -0.022 & 0.011 & 89.200 & 0.9822 & 8602 \\ 
+Territory aims & -0.027 & 0.014 & 81.083 & 0.9653 & 8775 \\ 
+Displaced Population & -0.048 & 0.027 & 58.689 & 0.9392 & 8695 \\ \hline
+\end{tabular}
+\caption{EBA -- Mass Killings in and after the Cold War Period (Robust Variables Only)}
+\label{tab:coldwar2}
+\end{table}
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=.9\textwidth]{images/mk-coldwar.pdf}
+    \caption{EBA -- Cold War Period}
+    \label{fig:mk-coldwar}
+\end{sidewaysfigure}
+\clearpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=.9\textwidth]{images/mk-postcoldwar.pdf}
+    \caption{EBA -- Post-Cold War Period}
+    \label{fig:mk-postcoldwar}
+\end{sidewaysfigure}
+\clearpage
+
+\subsubsection{Mass Killings during Peacetime}
+
+I have also tested whether the main EBA findings differ if the sample is restricted to peace years.
+That is, all observations denoted as 1 in the three conflict indicators mentioned above (UCDP, COW, Cederman et al.) were removed from the dataset.
+The results are similar to the main model, yet ``wars fought over territory'' was removed from the EBA due to multicollinearity issues.
+Moreover, two other variables (total battle deaths and presence of guerrillas) have very small sample sizes and their estimates were should not be interpreted as reliable.
+The significant variables are presented below.
+
+\vspace{1cm}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Variable} & \textbf{Avg. $\beta$} & \textbf{Avg. SE} & \textbf{$\%$ Sig.} & \textbf{CDF(0)} & \textbf{Models} \\ \hline
+\textit{Base variables} &  &  &  &  &  \\
+Log GDP per capita & -0.004 & 0.002 & 86.608 & 0.9748 & 26620 \\
+ &  &  &  &  &  \\
+\textit{Additional variables} &  &  &  &  &  \\
+Post Cold War & -0.011 & 0.003 & 100 & 0.9984 & 5459\\
+Polity IV squared & -1.63e-04 & 6.27e-05 & 98.309 & 0.9914 & 5441 \\
+Discriminated pop & 0.009 & 0.005 & 92.516 & 0.9750 & 5438 \\
+Mountainous terrain & 0.002 & 0.001 & 74.171 & 0.9718 & 5428\\
+Population & -9.89e-09 & 6.20e-09 & 61.776 & 0.9280 & 5418 \\
+Previous riots  & 0.008 & 0.006 & 28.315 & 0.9119 & 5400\\ \hline
+\end{tabular}
+\caption{EBA -- VIF 10}
+\label{tab:mk-high-vif}
+\end{table}
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/mk-nowar.pdf}
+    \caption{EBA -- Peace Years}
+    \label{fig:mk-postcoldwar}
+\end{sidewaysfigure}
+\clearpage
+
+\subsection{Harff's Genocides and Politicides Data}
+\label{sec:harff}
+
+\subsubsection{Main Model}
 
@@
 That is, none of the 36 predictors reached the threshold of CDF(0) $> 0.9$.
+Thus, we do not present a table with the results.
 The variable that came closest to significance was a dummy indicator of coups d'état, which has a CDF(0) of 0.897 and, as expected, is positively correlated with the onset of genocides.
-The distibution of the covariates' coefficients are available in figure \ref{fig:uamk}.
+The distribution of the covariates' coefficients are available in figure \ref{fig:uamk}.
+
+\newpage
 
@@
     \includegraphics[width=\textwidth]{images/uamk.pdf}
-    \caption{EBA -- Genocides/Politicides}
+    \caption{EBA -- Genocides and Politicides}
     \label{fig:uamk}
@@
 
-\subsubsection{Genocides/Politicides during Civil Wars}
+\subsubsection{Genocides and Politicides during Civil Wars}
 
@@
 Excluded population & -0.057 & 0.031 & 65.703 & 0.9570 & 8820 \\
-Discriminated population & -0.050 & 0.029 & 53.850 & 0.93.67 & 8767 \\
+Discriminated population & -0.050 & 0.029 & 53.850 & 0.9367 & 8767 \\
 Post-Cold War years & -0.019 & 0.013 & 42.531 & 0.9203 & 8904 \\
@@
 
-\newpage
 \clearpage
@@
 
+\subsection{Random Forest}
+\label{sec:mk-rfe}
+
+\subsubsection{Main Model}
+
+We employed the \texttt{H2O} machine learning platform \citep{h2o2017} to estimate the models.
+\texttt{H20} is open-source, optimised for big data and estimates a large number of models with only a few lines of code.
+We run the algorithms on 75\% of our dataset, and use the remaining 25\% as a validation set.
+That is, we use a percentage of the data to assess the main model's accuracy.\footnote{For more information about training and validation samples, please refer to \href{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/algo-params/validation_frame.html}{\texttt{http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science/algo-params/validation\_frame.html}}.} Our measure of accuracy is the area under the curve (AUC).
+All models score well in that regard, and measures of about 0.8 accuracy in our validation sample are common.
+
+The next two plots show the results of the main random forest models.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.85\textwidth]{images/rf-mk.pdf}
+    \caption{Variable Importance -- Main Model}
+    \label{fig:rf-mk}
+\end{figure}
+
 \newpage
-\subsection{Genocides/Politicides -- Random Forests}
-
-Lastly, we present three models using the distributed random forest algorithm \citep{h2o2017}.
-The results are in line with those obtained with the mass killing variable by  \citet{ulfelder2008assessing}.
-Again, we see that CINC, the percentage of urban population, and variables concerning the military are some of the most important predictors of state-led violence.
-The results confirm the overall finding of the chapter: Poor countries are more likely to experience mass killings, and states with a stronger army see a significant upward shift in genocide risk.
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=.85\textwidth]{images/rf-mk-pd.pdf}
+    \caption{Partial Dependence Plot -- Main Model}
+    \label{fig:rf-mk-pd}
+\end{sidewaysfigure}
+\clearpage
+
+\subsubsection{Mass Killings During Civil Wars}
+
+The following graphs display the most important predictors of mass killings when we restrict our sample to cases that occur during civil wars.
+As we note in section \ref{sec:civil-wars}, we employ three different measures of civil conflicts.
+The first one is provided by the Uppsala Conflict Data Program \citep{allansson2017organized,gleditsch2002armed}, the second is offered by the Correlates of War \citep{sarkees2010resort}, and a third indicating the onset of ethnic conflict as coded by \citet{cederman2010ethnic}.
 
@@
     \centering
-    \includegraphics{images/drf-gp.pdf}
-    \caption{Variable Importance -- Genocides/Politicides}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-ucdp.pdf}
+    \caption{Variable Importance -- Mass Killings during Civil Wars (UCDP Data)}
+    \label{fig:rf-mk-ucdp}
 \end{figure}
 
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-ucdp-pd.pdf}
+    \caption{Partial Dependence Plot -- Mass Killings during Civil Wars (UCDP Data)}
+    \label{fig:rf-mk-ucdp-pd}
+\end{sidewaysfigure}
+\clearpage
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.85\textwidth]{images/rf-cow.pdf}
+    \caption{Variable Importance -- Mass Killings during Civil Wars (COW Data)}
+    \label{fig:rf-mk-ucdp}
+\end{figure}
+
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-cow-pd.pdf}
+    \caption{Partial Dependence Plot -- Mass Killings during Civil Wars (COW Data)}
+    \label{fig:rf-mk-ucdp-pd}
+\end{sidewaysfigure}
+\clearpage
+
 \begin{figure}[H]
     \centering
-    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp4a.pdf}
-    \caption{Partial Plots -- Genocides/Politicides}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-eth.pdf}
+    \caption{Variable Importance -- Mass Killings during Civil Wars (Cederman et al. Data)}
+    \label{fig:rf-mk-ucdp}
 \end{figure}
@@
 
-The next graphs show the most important predictors of genocides that occur during civil wars and their respective partial dependence plots.
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=.85\textwidth]{images/rf-eth-pd.pdf}
+    \caption{Partial Dependence Plot -- Mass Killings during Civil Wars (Cederman et al. Data)}
+    \label{fig:rf-mk-ucdp-pd}
+\end{sidewaysfigure}
+\clearpage
+
+\newpage
+
+\subsubsection{Alternative Random Seeds}
+
+As random forests themselves are an approximation to a number of possible parameter combinations, changes in seed numbers may influence the model output.
+Thus, we start the main model with two different random seed numbers to check if the results are robust.\footnote{The numbers were generated at \href{https://www.random.org/}{\texttt{https://www.random.org/}}.} The main findings hold well; although variable importance changes from one model to another, the most significant variables appear repeatedly in the estimations.
+The marginal plots also show that the effect of the independent variables remain roughly similar despite the nonlinearities.
+The graphs below display the ten most significant predictors of mass killings and their respective partial dependence plots.
 
@@
     \centering
-    \includegraphics{images/drf-gp1.pdf}
-    \caption{Variable Importance -- Genocides/Politicides (UCDP Data)}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-mk-4363.pdf}
+    \caption{Variable Importance -- Seed 4363}
+    \label{fig:rf-mk-4363}
 \end{figure}
 
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \includegraphics[width=\textheight]{images/rf-mk-4363-pd.pdf}
+    \caption{Partial Dependence Plot -- Seed 4363}
+    \label{fig:rf-mk-4363}
+\end{sidewaysfigure}
+\clearpage
+
+\newpage
+
+\begin{figure}
+    \includegraphics[width=.85\textwidth]{images/rf-mk-7015.pdf}
+    \caption{Variable Importance -- Seed 7015}
+    \label{fig:rf-mk-4363}
+\end{figure}
+
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-mk-7015-pd.pdf}
+    \caption{Partial Dependence Plot -- Seed 7015}
+    \label{fig:rf-mk-4363}
+\end{sidewaysfigure}
+
+\newpage
+
+\subsubsection{Mass Killings in and after the Cold War}
+
+This last set of models splits the sample into two periods, the Cold War years and the post-Cold War years.
+A similar set of variables are significant in both periods, and most of them also appear in the main model shown above.
+
 \begin{figure}[H]
     \centering
-    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp5a.pdf}
-    \caption{Partial Plots -- Genocides/Politicides (UCDP Data)}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-coldwar.pdf}
+    \caption{Variable Importance -- Cold War Period}
+    \label{fig:rf-coldwar}
 \end{figure}
 
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \includegraphics[width=\textwidth]{images/rf-coldwar-pd.pdf}
+    \caption{Partial Dependence Plot -- Cold War Period}
+    \label{fig:rf-coldwar2}
+\end{sidewaysfigure}
+\clearpage
+
+\newpage
+
 \begin{figure}[H]
     \centering
-    \includegraphics{images/drf-gp2.pdf}
-    \caption{Variable Importance -- Genocides/Politicides (COW Data)}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-postcoldwar.pdf}
+    \caption{Variable Importance -- Post-Cold War Period}
+    \label{fig:rf-mk-4363}
 \end{figure}
 
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-postcoldwar-pd.pdf}
+    \caption{Partial Dependence Plot -- Post-Cold War Period}
+    \label{fig:rf-mk-4363}
+\end{sidewaysfigure}
+\clearpage
+
+\subsubsection{Mass Killings in Peacetime}
+
+The figures below describe the results of the random forest estimations when I restrict the sample to only peace years.
+All cases coded as conflicts by the UCDP, COW or Cederman and his colleagues were removed from the data, and I estimate the model only with observations where the three sourced coded as peace years.
+The results are almost identical to the main model, with only small variations in the importance of the explanator variables.
+
 \begin{figure}[H]
     \centering
-    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp6a.pdf}
-    \caption{Partial Plots -- Genocides/Politicides (COW Data)}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-nowar.pdf}
+    \caption{Variable Importance -- Mass Killings during Peacetime}
+    \label{fig:rf-mk-4363}
 \end{figure}
 
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-nowar-pd.pdf}
+    \caption{Partial Dependence Plot -- Mass Killings during Peacetime}
+    \label{fig:rf-mk-4363}
+\end{sidewaysfigure}
+\clearpage
+
+\newpage
+
+\subsection{Harff's Genocides and Politicides Data}
+
+\subsubsection{Main Model}
+
+We replicate the same analysis using Harff's \citeyear{harff2003no} data. The results are comparable to the ones presented above. A similar set of variables appear in this model.
+
 \begin{figure}[H]
     \centering
-    \includegraphics{images/drf-gp3.pdf}
-    \caption{Variable Importance -- Genocides/Politicides (Cederman et al. Data)}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-uamk.pdf}
+    \caption{Variable Importance -- Genocides and Politicides}
+    \label{fig:rf-mk-4363}
 \end{figure}
 
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-uamk-pd.pdf}
+    \caption{Partial Dependence Plot -- Genocides and Politicides}
+    \label{fig:rf-mk-4363}
+\end{sidewaysfigure}
+\clearpage
+
+\newpage
+
+\subsubsection{Genocides and Politicides during Civil Wars}
+
+Lastly, the graphs below show the results of the grid search when we only include civil war years.
+
 \begin{figure}[H]
     \centering
-    \includegraphics[width=\textwidth, height=9cm]{images/drfdpp7a.pdf}
-    \caption{Partial Plots -- Genocides/Politicides (Cederman et al. Data)}
-    \label{fig:my_label}
+    \includegraphics[width=.85\textwidth]{images/rf-uamk-ucdp.pdf}
+    \caption{Variable Importance -- Genocides and Politicides during Civil Wars (UCDP Data)}
+    \label{fig:rf-mk-ucdp}
 \end{figure}
@@
 
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-uamk-ucdp-pd.pdf}
+    \caption{Partial Dependence Plot -- Genocides and Politicides during Civil Wars (UCDP Data)}
+    \label{fig:rf-mk-ucdp-pd}
+\end{sidewaysfigure}
+\clearpage
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-uamk-cow.pdf}
+    \caption{Variable Importance -- Genocides and Politicides during Civil Wars (COW Data)}
+    \label{fig:rf-mk-ucdp}
+\end{figure}
+
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-uamk-cow-pd.pdf}
+    \caption{Partial Dependence Plot -- Genocides and Politicides during Civil Wars (COW Data)}
+    \label{fig:rf-mk-ucdp-pd}
+\end{sidewaysfigure}
+\clearpage
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.85\textwidth]{images/rf-uamk-eth.pdf}
+    \caption{Variable Importance -- Genocides and Politicides during Civil Wars (Cederman et al. Data)}
+    \label{fig:rf-mk-ucdp}
+\end{figure}
+
+\newpage
+
+\clearpage
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{images/rf-uamk-eth-pd.pdf}
+    \caption{Partial Dependence Plot -- Genocides and Politicides during Civil Wars (Cederman et al. Data)}
+    \label{fig:rf-mk-ucdp-pd}
+\end{sidewaysfigure}
+\clearpage
+
 \subsection{\texttt{R} Code}
-\label{sec:mk-code}
-
-The \texttt{R} code below reproduces the analyses presented in chapter \ref{chap:killings}.
+
+The \texttt{R} code below replicates all statistical analyses and graphs included in this chapter.
 
@@
 \begin{verbatim}
-######################
+
+#######################
 ### Data Wrangling ###
@@
 
-## Install and load required packages 
+### Load required packages
 if (!require("tidyverse")) {
@@
 }
+if (!require("sandwich")) {
+        install.packages("sandwich")
+}
 if (!require("h2o")) {
   install.packages("h2o")
-}
-if (!require("sandwich")) {
-        install.packages("sandwich")
 }
@@
 }
-if (!require("stargazer")) {
-        install.packages("stargazer")
-}
-
-## Load dataset
+
+### Load data
+setwd("~/Documents/GitHub/mass-killings-8k/") # set the working directory
 df <- haven::read_dta("data/base variables.dta") %>% setDT()
 
-## Select and lag variables
+### Select and lag variables
+sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
+             "COWcivilwarongoing", "ethnowarstart", "ethnowarongoing",
+             "assdummy", "demdummy", "elf", "lmtnest", "pop", "realgdp",
+             "rgdppc", "polity2", "exclpop", "discpop", "polrqnew",
+             "poltrqnew", "egiptpolrqnew", "egippolrqnew", "discrim",
+             "elf2", "interstatewar", "milex", "milper", "percentpopurban",
+             "postcoldwar", "coupdummy", "riotdummy", "territoryaims",
+             "totaltrade", "tradedependence", "militias", "physint", "cinc",
+             "totalbeaths", "guerrilladummy", "change", "sf", "regtrans")
+
+df1 <- cbind(df, df[, shift(.SD, 1, give.names = TRUE),
+                    by = ccode, .SDcols = sd.cols])
+
+# Remove the second `ccode` variable
+df1 <- as.data.frame(df1[, -c(70)])
+
+# Add new variables
+df1$logrgdppc_lag_1 <- log(df1$rgdppc_lag_1)
+df1$polity2sq_lag_1 <- df1$polity2_lag_1^2
+
+# UCDP civil war == 1
+df.ucdp <- df1 %>% filter(UCDPcivilwarongoing == 1)
+df.ucdp <- as.data.frame(df.ucdp[, c(1:7, 76:111)])
+names(df.ucdp) <- sub("_.*","", names(df.ucdp))
+
+# COW civil war == 1
+df.cow <- df1 %>% filter(COWcivilwarongoing == 1)
+df.cow <- as.data.frame(df.cow[, c(1:7, 76:111)])
+names(df.cow) <- sub("_.*","", names(df.cow))
+
+# Ethnic civil war == 1
+df.eth <- df1 %>% filter(ethnowarongoing == 1)
+df.eth <- as.data.frame(df.eth[, c(1:7, 75:110)])
+names(df.eth) <- sub("_.*","", names(df.eth))
+
+# Regular model
+df2 <- as.data.frame(df1[, c(1:7, 70:111)])
+names(df2) <- sub("_.*","", names(df2))
+
+# Cold War period
+df2.coldwar <- df2 %>% filter(year <= 1991)
+df2.postcoldwar <- df2 %>% filter(year > 1991)
+
+# Countries without civil wars
+df.nowar <- df2 %>% filter(COWcivilwarongoing == 0 & UCDPcivilwarongoing == 0, ethnowarongoing == 0)
+
+#### Same procedure with the uamkstart variable
+
+# Preparing the dataset
+df3 <- haven::read_dta("data/uamkstart.dta") %>% setDT()
 sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
@@
 
-df1 <- cbind(df, df[, shift(.SD, 1, give.names = TRUE),
+df4 <- cbind(df3, df3[, shift(.SD, 1, give.names = TRUE),
                     by = ccode, .SDcols = sd.cols])
@@
 # Remove the second `ccode` variable
-df1 <- as.data.frame(df1[, -c(70)])
+df4 <- as.data.frame(df4[, -c(75)])
 
 # Add new variables
-df1$logrgdppc_lag_1 <- log(df1$rgdppc_lag_1)
-df1$polity2sq_lag_1 <- df1$polity2_lag_1^2
+df4$logrgdppc_lag_1 <- log(df4$rgdppc_lag_1)
+df4$polity2sq_lag_1 <- df4$polity2_lag_1^2
+
+# Renaming variables
+df5 <- as.data.frame(df4[, c(1:4, 72:116)])
+names(df5) <- sub("_.*","", names(df5))
 
 # UCDP civil war == 1
-df.ucdp <- df1 %>% filter(UCDPcivilwarongoing == 1)
-df.ucdp <- as.data.frame(df.ucdp[, c(1:7, 76:111)])
-names(df.ucdp) <- sub("_.*","", names(df.ucdp))
+df.ucdp2 <- df5 %>% filter(UCDPcivilwarongoing == 1)
+df.ucdp2 <- as.data.frame(df.ucdp2[, c(1:7, 14:49)])
+names(df.ucdp2) <- sub("_.*","", names(df.ucdp2))
 
 # COW civil war == 1
-df.cow <- df1 %>% filter(COWcivilwarongoing == 1)
-df.cow <- as.data.frame(df.cow[, c(1:7, 76:111)])
-names(df.cow) <- sub("_.*","", names(df.cow))
+df.cow2 <- df5 %>% filter(COWcivilwarongoing == 1)
+df.cow2 <- as.data.frame(df.cow2[, c(1:7, 14:49)])
+names(df.cow2) <- sub("_.*","", names(df.cow2))
 
 # Ethnic civil war == 1
-df.eth <- df1 %>% filter(ethnowarongoing == 1)
-df.eth <- as.data.frame(df.eth[, c(1:7, 76:111)])
-names(df.eth) <- sub("_.*","", names(df.eth))
-
-# Regular model
-df2 <- as.data.frame(df1[, c(1:7, 70:111)])
-names(df2) <- sub("_.*","", names(df2))
-
-###############################
-### Extreme Bounds Analyses ###
-###############################
-
-## Classifying a few variables as mutually exclusive.
-## "Change" was removed because it was correlated at 0.99 with "regtrans". 
+df.eth2 <- df5 %>% filter(ethnowarongoing == 1)
+df.eth2 <- as.data.frame(df.eth2[, c(1:7, 14:49)])
+names(df.eth2) <- sub("_.*","", names(df.eth2))
+
+# Cold War period
+df5.coldwar <- df5 %>% filter(year <= 1991)
+df5.postcoldwar <- df5 %>% filter(year > 1991)
+
+################################
+### Extreme bounds analysis  ###
+################################
+
+# Classifying a few variables as mutually exclusive variables.
+# "Change" was removed because it was correlated at 0.99 with "regtrans". 
+# don't forget to add CINC
 free.variables <- c("logrgdppc", "polity2", "mksyr")
@@
 
-# Main model
+### Models
+
+# Main
 m1 <- eba(y = "MKstart", free = free.variables,
@@
           se.fun = se.clustered.robust)
-
-summary(m1)
-hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
-                       "UCDPcivilwarongoing",
-                       "UCDPcivilwarstart", "COWcivilwarongoing",
-                       "COWcivilwarstart", "ethnowarongoing", "ethnowarstart",
-                       "assdummy", "totaltrade", "tradedependence", "milper",
-                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
-                       "riotdummy", "territoryaims", "militias", "physint",
-                       "percentpopurban", "coupdummy", "postcoldwar",
-                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
-                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
-                       "egiptpolrqnew"),
-     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
-              "UCDP ongoing", "UCDP onset", "COW ongoing", "COW onset", 
-              "Ethnic ongoing", "Ethnic onset", "Assassination", "Total trade", 
-              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
-              "Total deaths", "Guerrilla", "Regime transition", "Riots",
-              "Territory Aims", "Militias", "Physical integrity", "% Urban",
-              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
-              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
-              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
-     density.col = "black", mu.col = "red3")
-
-# Mass killings during civil war
-# UCDP civil conflicts == 1
+save(m1, file = "~/Documents/mk/mk.rda")
+
+# 3 vars at a time
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:3,
+          data = df2, vif = 7, level = 0.9, draws = 10000,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/mk-3vars.rda")
+
+# 5 vars at a time
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:5,
+          data = df2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/mk-5vars.rda")
+
+# Low VIF
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, vif = 2.5, level = 0.9,
+          draws = 50000,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/mk-low-vif.rda")
+
+# High VIF
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, vif = 10, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/mk-high-vif.rda")
+
+# No VIF
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/mk-no-vif.rda")
+
+# Logit
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, level = 0.9, vif = 7, draws = 50000,
+          reg.fun = bayesglm, family = binomial(link = "logit"))
+save(m1, file = "~/Documents/mk/mk-logit.rda")
+
+# Probit
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, level = 0.9, vif = 7, draws = 50000,
+          reg.fun = bayesglm, family = binomial(link="probit"))
+save(m1, file = "~/Documents/mk/mk-probit.rda")
+
+# CINC
+doubtful.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart", "assdummy",
+                        "totaltrade", "tradedependence", "cinc",
+                        "totalbeaths", "guerrilladummy", "regtrans",
+                        "riotdummy", "territoryaims", "militias",
+                        "physint", "percentpopurban", "coupdummy",
+                        "postcoldwar",  "lmtnest", "realgdp", "discrim",
+                        "exclpop", "discpop", "elf",  "polrqnew",
+                        "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                        "polity2sq")
+
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2, vif = 7, level = 0.9, 
+          se.fun = se.clustered.robust, draws = 50000)
+save(m1, file = "~/Documents/mk/mk-cinc.rda")
+
+# Cold War Period
+civilwar.variables <- c("UCDPcivilwarstart","COWcivilwarstart","ethnowarstart")
+m1 <- doubtful.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                              "COWcivilwarongoing", "COWcivilwarstart",
+                              "ethnowarongoing", "ethnowarstart", "assdummy",
+                              "totaltrade", "tradedependence", "cinc",
+                              "totalbeaths", "guerrilladummy", "regtrans",
+                              "riotdummy", "territoryaims", "militias",
+                              "physint", "percentpopurban", "coupdummy",
+                              "lmtnest", "realgdp", "discrim",
+                              "exclpop", "discpop", "elf",  "polrqnew",
+                              "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                              "polity2sq")
+
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2.coldwar, vif = 7, level = 0.9, 
+          se.fun = se.clustered.robust, draws = 50000)
+save(m1, file = "data/mk-coldwar.rda")
+
+# Post-Cold War
+m1 <- doubtful.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                              "COWcivilwarongoing", "COWcivilwarstart",
+                              "ethnowarongoing", "ethnowarstart", "assdummy",
+                              "totaltrade", "tradedependence", "cinc",
+                              "totalbeaths", "guerrilladummy", "regtrans",
+                              "riotdummy", "territoryaims", "militias",
+                              "physint", "percentpopurban", "coupdummy",
+                              "lmtnest", "realgdp", "discrim",
+                              "exclpop", "discpop", "elf",  "polrqnew",
+                              "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                              "polity2sq")
+
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df2.postcoldwar, vif = 7, level = 0.9, 
+          se.fun = se.clustered.robust, draws = 50000)
+save(m1, file = "data/mk-postcoldwar.rda")
+
+## Countries with no civil wars
+free.variables <- c("logrgdppc", "polity2", "mksyr")
+civilwar.variables <- c("UCDPcivilwarstart", "COWcivilwarstart",
+                        "ethnowarstart")
+m1 <- doubtful.variables <- c("UCDPcivilwarstart","COWcivilwarstart",
+                              "ethnowarstart", "assdummy",
+                              "totaltrade", "tradedependence", "cinc",
+                              "totalbeaths","guerrilladummy",
+                              "riotdummy", "territoryaims", "militias",
+                              "physint", "percentpopurban", "coupdummy",
+                              "postcoldwar", "lmtnest", "realgdp", "discrim",
+                              "exclpop", "discpop", "elf",  "polrqnew",
+                              "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                              "polity2sq")
+
+m1 <- eba(y = "MKstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.nowar, vif = 7, level = 0.9, 
+          se.fun = se.clustered.robust, draws = 50000)
+save(m1, file = "data/mk-nowar.rda")
+
+### Ongoing Civil Wars
+
+# UCDPcivilwarongoing == 1
 doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
@@
           doubtful = doubtful.variables, k = 0:4,
-          data = df.ucdp, vif = 7,
+          data = df.ucdp, vif = 7, draws = 50000,
           level = 0.9, se.fun = se.clustered.robust)
-
-summary(m1)
-hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
-                       "assdummy", "totaltrade", "tradedependence", "milper",
-                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
-                       "riotdummy", "territoryaims", "militias", "physint",
-                       "percentpopurban", "coupdummy", "postcoldwar",
-                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
-                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
-                       "egiptpolrqnew"),
-     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
-              "Assassination", "Total trade", 
-              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
-              "Total deaths", "Guerrilla", "Regime transition", "Riots",
-              "Territory Aims", "Militias", "Physical integrity", "% Urban",
-              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
-              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
-              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
-     density.col = "black", mu.col = "red3")
-
-# COW civil wars == 1
+save(m1, file = "~/Documents/mk/mk-ucdp.rda")
+
+# COWcivilwarongoing == 1
 doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
@@
           doubtful = doubtful.variables, k = 0:4,
-          data = df.cow, vif = 7,
+          data = df.cow, vif = 7, draws = 50000,
           level = 0.9, se.fun = se.clustered.robust)
-
-summary(m1)
-hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
-                       "assdummy", "totaltrade", "tradedependence", "milper",
-                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
-                       "riotdummy", "territoryaims", "militias", "physint",
-                       "percentpopurban", "coupdummy", "postcoldwar",
-                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
-                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
-                       "egiptpolrqnew"),
-     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
-              "Assassination", "Total trade", 
-              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
-              "Total deaths", "Guerrilla", "Regime transition", "Riots",
-              "Territory Aims", "Militias", "Physical integrity", "% Urban",
-              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
-              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
-              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
-     density.col = "black", mu.col = "red3")
-
-# Ethnic civil war == 1
+save(m1, file = "~/Documents/mk/mk-cow.rda")
+
+# Ethnic conflict == 1
 doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
@@
           doubtful = doubtful.variables, k = 0:4,
-          data = df.eth, vif = 7,
+          data = df.eth, vif = 7, draws = 50000,
           level = 0.9, se.fun = se.clustered.robust)
-
-summary(m1)
-hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr",
-                       "assdummy", "totaltrade", "tradedependence", "milper",
-                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
-                       "riotdummy", "territoryaims", "militias", "physint",
-                       "percentpopurban", "coupdummy", "postcoldwar",
-                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
-                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
-                       "egiptpolrqnew"),
-     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last mass killing",
-              "Assassination", "Total trade", 
-              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
-              "Total deaths", "Guerrilla", "Regime transition", "Riots",
-              "Territory Aims", "Militias", "Physical integrity", "% Urban",
-              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
-              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
-              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
-     density.col = "black", mu.col = "red3")
-
-## Different values of k
-## Code for the histogram not included as it is the same as that of the main model.
-
-# 3 variables per model
-m1 <- eba(y = "MKstart", free = free.variables,
-          exclusive = list(civilwar.variables),
-          doubtful = doubtful.variables, k = 0:3,
-          data = df2, vif = 7, level = 0.9, draws = 50000,
-          se.fun = se.clustered.robust)
-
-# 5 variables per model
-m1 <- eba(y = "MKstart", free = free.variables,
-          exclusive = list(civilwar.variables),
-          doubtful = doubtful.variables, k = 0:5,
-          data = df2, vif = 7, draws = 50000,
-          level = 0.9, se.fun = se.clustered.robust)
-
-## Alternative VIFs
-
-# VIF = 10
-# Low VIF
-m1 <- eba(y = "MKstart", free = free.variables,
-          exclusive = list(civilwar.variables),
-          doubtful = doubtful.variables, k = 0:4,
-          data = df2, vif = 10, level = 0.9, draws = 50000,
-          se.fun = se.clustered.robust)
-
-# VIF = 2.5
-m1 <- eba(y = "MKstart", free = free.variables,
-          exclusive = list(civilwar.variables),
-          doubtful = doubtful.variables, k = 0:4,
-          data = df2, vif = 2.5, level = 0.9, draws = 50000,
-          se.fun = se.clustered.robust)
-
-# No VIF
-m1 <- eba(y = "MKstart", free = free.variables,
-          exclusive = list(civilwar.variables),
-          doubtful = doubtful.variables, k = 0:4,
-          data = df2, level = 0.9, draws = 50000,
-          se.fun = se.clustered.robust)
-
-## Generalised linear models
-
-# Logit
-m1 <- eba(y = "MKstart", free = free.variables,
-          exclusive = list(civilwar.variables),
-          doubtful = doubtful.variables, k = 0:4,
-          data = df2, level = 0.9, vif = 7, draws = 50000,
-          reg.fun = bayesglm, family = binomial(link = "logit"))
-
-# Probit
-m1 <- eba(y = "MKstart", free = free.variables,
-          exclusive = list(civilwar.variables),
-          doubtful = doubtful.variables, k = 0:4,
-          data = df2, level = 0.9, vif = 7, draws = 50000,
-          reg.fun = bayesglm, family = binomial(link = "probit"))
-
-##################################
-### Distributed Random Forests ###
-##################################
-
-## Random Forests
-
-## Prepare the data set
-h2o.init(nthreads = -1, max_mem_size = "20g") # memory size
-
-df2a <- as.h2o(df2)
-
-df2a$MKstart <- as.factor(df2a$MKstart)  # encode the binary response as a factor
-h2o.levels(df2a$MKstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df2a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "MKstart"
-x <- setdiff(names(df2), c(y, "ccode", "year", "rgdppc",
-                           "mksyr2", "mksyr3", "sf", "country",
-                           "elf2", "polity2sq"))
-
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf01",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-# Saving the most accurate model
-rf.grid <- h2o.getGrid(grid_id = "gridrf01",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-# Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf01_model_21")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Polity IV", 
-                      "Military personnel",
-                      "Ethnic polarisation", 
-                      "% Urban pop.",
-                      "Years mass killing",
-                      "Log GDP per capita"),
-        main = "")
-
-logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
-p1 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Log GDP per capita") + ylab("Mean response")
-
-mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
-p2 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Years since mass killing") +  ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p3 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        theme_classic() + ylim(0, 0.05) + xlab("% Urban") + ylab("Mean response")
-
-egiptpolrqnew <- h2o.partialPlot(object = a, data = train, cols = c("egiptpolrqnew"))
-p4 <- qplot(egiptpolrqnew$egiptpolrqnew, egiptpolrqnew$mean_response) + geom_line() +
-        theme_classic() + ylim(0, 0.05) + xlab("Ethnic polarisation") + ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p5 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Military personnel") + ylab("Mean response")
-
-polity2 <- h2o.partialPlot(object = a, data = train, cols = c("polity2"))
-p6 <- qplot(polity2$polity2, polity2$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Polity IV") + ylab("Mean response")
-
-# Multiplot function: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
-multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
-        library(grid)
-
-        # Make a list from the ... arguments and plotlist
-        plots <- c(list(...), plotlist)
-
-        numPlots = length(plots)
-
-        # If layout is NULL, then use 'cols' to determine layout
-        if (is.null(layout)) {
-                # Make the panel
-                # ncol: Number of columns of plots
-                # nrow: Number of rows needed, calculated from # of cols
-                layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-                                 ncol = cols, nrow = ceiling(numPlots/cols))
-        }
-
-        if (numPlots==1) {
-                print(plots[[1]])
-
-        } else {
-                # Set up the page
-                grid.newpage()
-                pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
-
-                # Make each plot, in the correct location
-                for (i in 1:numPlots) {
-                        # Get the i,j matrix positions of the regions that contain this subplot
-                        matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-
-                        print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-                                                        layout.pos.col = matchidx$col))
-                }
-        }
-}
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-## Different seeds
-
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf01b",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 44849999))
-
-# Saving the most accurate model
-rf.grid <- h2o.getGrid(grid_id = "gridrf01b",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-# Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf01b_model_8")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Population", 
-                      "Military personnel",
-                      "Trade dependence", 
-                      "Log GDP per capita",
-                      "% Urban pop.",
-                      "Years mass killing"),
-        main = "")
-
-mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
-p1 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Years since mass killing") +  ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        theme_classic() + ylim(0, 0.05) + xlab("% Urban") + ylab("Mean response")
-
-logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
-p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Log GDP per capita") + ylab("Mean response")
-
-tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
-p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
-        theme_classic() + ylim(0, 0.05) + xlab("Trade dependence") + ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p5 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Military personnel") + ylab("Mean response")
-
-pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"))
-p6 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Population") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf01c",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 1502436))
-
-# Saving the most accurate model
-rf.grid <- h2o.getGrid(grid_id = "gridrf01c",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-# Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf01c_model_58")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Population", 
-                      "Military personnel",
-                      "Trade dependence", 
-                      "Log GDP per capita",
-                      "% Urban pop.",
-                      "Years mass killing"),
-        main = "")
-
-mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
-p1 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Years since mass killing") +  ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        theme_classic() + ylim(0, 0.05) + xlab("% Urban") + ylab("Mean response")
-
-logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
-p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Log GDP per capita") + ylab("Mean response")
-
-tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
-p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
-        theme_classic() + ylim(0, 0.05) + xlab("Trade dependence") + ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p5 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Military personnel") + ylab("Mean response")
-
-pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"))
-p6 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() + ylim(0, 0.05) +
-        xlab("Population) + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-## Genocides in civil wars
-
-## UCDP data
-df.ucdpa <- as.h2o(df.ucdp)
-
-df.ucdpa$MKstart <- as.factor(df.ucdpa$MKstart)  # encode the binary response as a factor
-h2o.levels(df.ucdpa$MKstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df.ucdpa, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "MKstart"
-x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
-                           "mksyr2", "mksyr3", "sf", "country",
-                           "elf2", "polity2sq"))
-
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf02",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf02",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-# Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf02_model_34")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Total battle deaths", 
-                      "Log GDP per capita",
-                       "Trade dependence",
-                      "Military personnel",
-                      "% Urban pop.",
-                      "Years mass killing"),
-        main = "")
-
-mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
-p1 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + 
-        ylim(0, 0.1) + xlab("Years since mass killing") +  ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        ylim(0, 0.1) + theme_classic() + xlab("% Urban") + ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p3 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
-        ylim(0, 0.25) + xlab("Military personnel") + ylab("Mean response")
-
-tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
-p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() + theme_classic() +
-        ylim(0, 0.1) + xlab("Trade dependence") + ylab("Mean response")
-
-logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
-p5 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
-        ylim(0, 0.1) + xlab("Log GDP per capita") + ylab("Mean response")
-
-totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"))
-p6 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() + theme_classic() +
-        ylim(0, 0.35) + xlab("Total battle deaths") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-## COW data
-df.cowa <- as.h2o(df.cow)
-
-df.cowa$MKstart <- as.factor(df.cowa$MKstart)
-h2o.levels(df.cowa$MKstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df.cowa, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "MKstart"
-x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
-                               "mksyr2", "mksyr3", "sf", "country",
-                               "elf2", "polity2sq"))
-
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf03",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf03",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## Graphs
-a <- h2o.loadModel("/root/Documents/mk/gridrf03_model_3")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6)) 
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Total battle deaths", 
-                      "Excluded population",
-                      "Yrs since mass killing",
-                      "Log GDP per capita",
-                      "Ethnic polarisation",
-                      "Physical integrity"),
-        main = "")
-
-physint <- h2o.partialPlot(object = a, data = train, cols = c("physint"))
-p1 <- qplot(physint$physint, physint$mean_response) + geom_line() + theme_classic() + 
-        xlab("Physical integrity") +  ylab("Mean response")
-
-egiptpolrqnew <- h2o.partialPlot(object = a, data = train, cols = c("egiptpolrqnew"))
-p2 <- qplot(egiptpolrqnew$egiptpolrqnew, egiptpolrqnew$mean_response) + geom_line() +
-        theme_classic() + xlab("Ethnic polarisation") + ylab("Mean response")
-
-logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
-p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
-        ylim(0, 0.1) + xlab("Log GDP per capita") + ylab("Mean response")
-
-mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"))
-p4 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + 
-        ylim(0, 0.1) + xlab("Years since mass killing") +  ylab("Mean response")
-
-exclpop <- h2o.partialPlot(object = a, data = train, cols = c("exclpop"))
-p5 <- qplot(exclpop$exclpop, exclpop$mean_response) + geom_line() + theme_classic() +
-        ylim(0, 0.1) + xlab("Excluded population") + ylab("Mean response")
-
-totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"))
-p6 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() + theme_classic() +
-        ylim(0, 0.1) + xlab("Total battle deaths") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-## Ethnic war
-df.etha <- as.h2o(df.eth)
-
-df.etha$MKstart <- as.factor(df.etha$MKstart) 
-h2o.levels(df.etha$MKstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df.etha, 
-                         ratios = c(0.7, 0.15), 
-                         seed = 42)
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "MKstart"
-x <- setdiff(names(df.eth), c(y, "ccode", "year", "rgdppc",
-                               "mksyr2", "mksyr3", "sf", "country",
-                               "elf2", "polity2sq"))
-
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf04",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf04",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-# Graphs
-a <- h2o.loadModel("/root/Documents/mk/gridrf04_model_14")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Military expenditure",
-                      "Total trade", 
-                      "% Urban",
-                      "Trade dependence",
-                      "Military personnel",
-                      "CINC"),
-        main = "")
-
-cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"))
-p1 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
-        xlab("CINC") +  ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p2 <- qplot(milper$milper, milper$mean_response) + geom_line() +
-        theme_classic() + xlab("Military personnel") + ylab("Mean response")
-
-tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
-p3 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() + theme_classic() +
-        xlab("Trade dependence") + ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p4 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-       theme_classic() + xlab("% Urban") + ylab("Mean response")
-
-totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"))
-p5 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() +
-        theme_classic() + xlab("Total trade") + ylab("Mean response")
-
-milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
-p6 <- qplot(milex$milex, milex$mean_response) + geom_line() +
-        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-####################################
-### Genocide/Politicide Variable ###
-####################################
-
-## The code below replicates the same analyses presented above
-## but using a measure of genocide/politicide coded by Harff (2003).
-
-## Data wrangling
-df3 <- haven::read_dta("data/uamkstart.dta") %>% setDT()
-
-sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
-             "COWcivilwarongoing", "ethnowarstart", "ethnowarongoing",
-             "assdummy", "demdummy", "elf", "lmtnest", "pop", "realgdp",
-             "rgdppc", "polity2", "exclpop", "discpop", "polrqnew",
-             "poltrqnew", "egiptpolrqnew", "egippolrqnew", "discrim",
-             "elf2", "interstatewar", "milex", "milper", "percentpopurban",
-             "postcoldwar", "coupdummy", "riotdummy", "territoryaims",
-             "totaltrade", "tradedependence", "militias", "physint", "cinc",
-             "totalbeaths", "change", "guerrilladummy", "sf", "regtrans")
-
-df4 <- cbind(df3, df3[, shift(.SD, 1, give.names = TRUE),
-                      by = ccode, .SDcols = sd.cols])
-
-# Remove the second `ccode` variable
-df4 <- as.data.frame(df4[, -c(75)])
-
-# Add new variables
-df4$logrgdppc_lag_1 <- log(df4$rgdppc_lag_1)
-df4$polity2sq_lag_1 <- df4$polity2_lag_1^2
-
-# Renaming variables
-df5 <- as.data.frame(df4[, c(1:4, 72:116)])
-names(df5) <- sub("_.*","", names(df5))
-
-#################################################
-### Distributed Random Forests - Harff (2003) ###
-#################################################
-
-df5a <- as.h2o(df5)
-
-df5a$uamkstart <- as.factor(df5a$uamkstart)  #encode the binary repsonse as a factor
-h2o.levels(df5a$uamkstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df5a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "uamkstart"
-x <- setdiff(names(df5), c(y, "ccode", "year", "rgdppc",
-                           "uamkyr2", "uamkyr3", "sf", "country",
-                           "elf2", "polity2sq")) 
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, 
-               grid_id = "gridrf05",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-# Saving the most accurate model
-rf.grid <- h2o.getGrid(grid_id = "gridrf05",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf05_model_27")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Real GDP", 
-                      "Military expenditure",
-                      "Total trade", 
-                      "Military personnel",
-                      "CINC", 
-                      "% Urban"),
-        main = "")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p1 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        theme_classic() + xlab("% Urban") + ylab("Mean response")
-
-cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"))
-p2 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
-        xlab("CINC") +  ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p3 <- qplot(milper$milper, milper$mean_response) + geom_line() +
-        theme_classic() + xlab("Military personnel") + ylab("Mean response")
-
-totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"))
-p4 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
-        xlab("Total trade") + ylab("Mean response")
-
-milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
-p5 <- qplot(milex$milex, milex$mean_response) + geom_line() +
-        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
-
-realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"))
-p6 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() +
-        theme_classic() + xlab("Real GDP") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-## UCDP == 1
-df.ucdp2 <- df5 %>% filter(UCDPcivilwarongoing == 1)
-df.ucdp2a <- as.h2o(df.ucdp2)
-
-df.ucdp2a$uamkstart <- as.factor(df.ucdp2a$uamkstart)  #encode the binary repsonse as a factor
-h2o.levels(df.ucdp2a$uamkstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df.ucdp2a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "uamkstart"
-x <- setdiff(names(df.ucdp2), c(y, "ccode", "year", "rgdppc",
-                           "uamkyr2", "uamkyr3", "sf", "country",
-                           "elf2", "polity2sq"))
-
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf06",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf06",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf06_model_43")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Excluded population",
-                      "% Urban pop.", 
-                      "Trade dependence",
-                      "Log GDP per capita",
-                      "Years since genocide", 
-                      "Military personnel"),
-        main = "")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p1 <- qplot(milper$milper, milper$mean_response) + geom_line() +
-        theme_classic() + xlab("Military personnel") + ylab("Mean response")
-
-uamkyr <- h2o.partialPlot(object = a, data = train, cols = c("uamkyr"))
-p2 <- qplot(uamkyr$uamkyr, uamkyr$mean_response) + geom_line() + theme_classic() + 
-        xlab("Years since genocide") +  ylab("Mean response")
-
-logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
-p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
-        xlab("Log GDP per capita") + ylab("Mean response")
-
-tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
-p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
-        theme_classic() + xlab("Trade dependence") + ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p5 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        theme_classic() + xlab("% Urban") + ylab("Mean response")
-
-exclpop <- h2o.partialPlot(object = a, data = train, cols = c("exclpop"))
-p6 <- qplot(exclpop$exclpop, exclpop$mean_response) + geom_line() +
-        theme_classic() + xlab("Excluded population") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-## COW == 1
-df.cow2 <- df5 %>% filter(COWcivilwarongoing == 1)
-
-df.cow2a$uamkstart <- as.factor(df.cow2a$uamkstart)  #encode the binary repsonse as a factor
-h2o.levels(df.cow2a$uamkstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df.cow2a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "uamkstart"
-x <- setdiff(names(df.cow2), c(y, "ccode", "year", "rgdppc",
-                               "uamkyr2", "uamkyr3", "sf", "country",
-                               "elf2", "polity2sq"))
-
-## Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf07",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf07",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf07_model_27")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Military expenditure",
-                      "% Urban pop.", 
-                      "Total trade",
-                      "Years since genocide", 
-                      "Military personnel", 
-                      "Trade dependence"),
-        main = "")
-
-uamkyr <- h2o.partialPlot(object = a, data = train, cols = c("uamkyr"))
-p3 <- qplot(uamkyr$uamkyr, uamkyr$mean_response) + geom_line() + theme_classic() + 
-        xlab("Years since genocide") +  ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p2 <- qplot(milper$milper, milper$mean_response) + geom_line() +
-        theme_classic() + xlab("Military personnel") + ylab("Mean response")
-
-totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"))
-p4 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
-        xlab("Total trade") + ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p5 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        theme_classic() + xlab("% Urban") + ylab("Mean response")
-
-tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"))
-p1 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
-        theme_classic() + xlab("Trade dependence") + ylab("Mean response")
-
-milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
-p6 <- qplot(milex$milex, milex$mean_response) + geom_line() +
-        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-## ETHONSET == 1
-df.eth2 <- df5 %>% filter(ethnowarongoing == 1)
-
-df.eth2a <- as.h2o(df.eth2)
-
-df.eth2a$uamkstart <- as.factor(df.eth2a$uamkstart)  #encode the binary repsonse as a factor
-h2o.levels(df.eth2a$uamkstart)
-
-# Partition the data into training, validation and test sets
-splits <- h2o.splitFrame(data = df.eth2a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
-
-train <- h2o.assign(splits[[1]], "train.hex")   
-valid <- h2o.assign(splits[[2]], "valid.hex") 
-test <- h2o.assign(splits[[3]], "test.hex")
-
-y <- "uamkstart"
-x <- setdiff(names(df.eth2), c(y, "ccode", "year", "rgdppc",
-                               "uamkyr2", "uamkyr3", "sf", "country",
-                               "elf2", "polity2sq"))
-
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf08",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 1000, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf08",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## Graphs
-a <- h2o.loadModel("/home/sussa/Documents/GitHub/mk/gridrf08_model_55")
-print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 6))
-
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
-barplot(va$scaled_importance[6:1],
-        horiz = TRUE, las = 1, cex.names=0.9,
-        names.arg = c("Total battle deaths", 
-                      "Military personnel",
-                      "Log GDP per capita", 
-                      "CINC",
-                      "% Urban",
-                      "Military expenditure"),
-        main = "")
-
-cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"))
-p3 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
-        xlab("CINC") +  ylab("Mean response")
-
-milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"))
-p5 <- qplot(milper$milper, milper$mean_response) + geom_line() +
-        theme_classic() + xlab("Military personnel") + ylab("Mean response")
-
-logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"))
-p4 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
-        xlab("Log GDP per capita") + ylab("Mean response")
-
-percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"))
-p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
-        theme_classic() + xlab("% Urban") + ylab("Mean response")
-
-totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"))
-p6 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() +
-        theme_classic() + xlab("Total battle deaths") + ylab("Mean response")
-
-milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"))
-p1 <- qplot(milex$milex, milex$mean_response) + geom_line() +
-        theme_classic() + xlab("Military expenditure") + ylab("Mean response")
-
-multiplot(p1, p4, p2, p5, p3, p6, cols = 3)
-
-##############################################
-### Genocides and Politicides (Harff 2003) ###
-##############################################
-
-# Preparing the dataset
-df3 <- haven::read_dta("data/uamkstart.dta") %>% setDT()
-sd.cols <- c("UCDPcivilwarstart", "UCDPcivilwarongoing", "COWcivilwarstart",
-             "COWcivilwarongoing", "ethnowarstart", "ethnowarongoing",
-             "assdummy", "demdummy", "elf", "lmtnest", "pop", "realgdp",
-             "rgdppc", "polity2", "exclpop", "discpop", "polrqnew",
-             "poltrqnew", "egiptpolrqnew", "egippolrqnew", "discrim",
-             "elf2", "interstatewar", "milex", "milper", "percentpopurban",
-             "postcoldwar", "coupdummy", "riotdummy", "territoryaims",
-             "totaltrade", "tradedependence", "militias", "physint", "cinc",
-             "totalbeaths", "change", "guerrilladummy", "sf", "regtrans")
-
-df4 <- cbind(df3, df3[, shift(.SD, 1, give.names = TRUE),
-                    by = ccode, .SDcols = sd.cols])
-
-# Remove the second `ccode` variable
-df4 <- as.data.frame(df4[, -c(75)])
-
-# Add new variables
-df4$logrgdppc_lag_1 <- log(df4$rgdppc_lag_1)
-df4$polity2sq_lag_1 <- df4$polity2_lag_1^2
-
-# Renaming variables
-df5 <- as.data.frame(df4[, c(1:4, 72:116)])
-names(df5) <- sub("_.*","", names(df5))
-
-## Extreme Bounds
+save(m1, file = "~/Documents/mk/mk-eth.rda")
+
+# Main
 free.variables <- c("logrgdppc", "polity2", "uamkyr")
@@
           se.fun = se.clustered.robust)
-
+save(m1, file = "~/Documents/mk/uamk.rda")
+
+# 3 vars at a time
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:3,
+          data = df5, vif = 7, level = 0.9,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-3vars.rda")
+
+# 5 vars at a time
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:5,
+          data = df5, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-5vars.rda")
+
+# Low VIF
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df5, vif = 2.5, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-low-vif.rda")
+
+# High VIF
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df5, vif = 10, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-high-vif.rda")
+
+# No VIF
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df5, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-no-vif.rda")
+
+# Logit
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df5, level = 0.9, vif = 7, draws = 50000,
+          reg.fun = bayesglm, family = binomial(link = "logit"))
+save(m1, file = "~/Documents/mk/uamk-logit.rda")
+
+# Probit
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df5, level = 0.9, vif = 7, draws = 50000,
+          reg.fun = bayesglm, family = binomial(link="probit"))
+save(m1, file = "~/Documents/mk/uamk-probit.rda")
+
+# CINC
+doubtful.variables <- c("UCDPcivilwarongoing", "UCDPcivilwarstart",
+                        "COWcivilwarongoing", "COWcivilwarstart",
+                        "ethnowarongoing", "ethnowarstart", "assdummy",
+                        "totaltrade", "tradedependence", "cinc",
+                        "totalbeaths", "guerrilladummy", "regtrans",
+                        "riotdummy", "territoryaims", "militias",
+                        "physint", "percentpopurban", "coupdummy",
+                        "postcoldwar",  "lmtnest", "realgdp", "discrim",
+                        "exclpop", "discpop", "elf",  "polrqnew",
+                        "egippolrqnew", "poltrqnew", "egiptpolrqnew",
+                        "polity2sq")
+
+m1 <- eba(y = "uamkstart", free = free.variables,
+          exclusive = list(civilwar.variables),
+          doubtful = doubtful.variables, k = 0:4,
+          data = df5, vif = 7, level = 0.9, draws = 50000,
+          se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-cinc.rda")
+
+### Ongoing Civil Wars
+
+# UCDPcivilwarongoing == 1
+df.ucdp2 <- df5 %>% filter(UCDPcivilwarongoing == 1)
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop",
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "uamkstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.ucdp2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-ucdp.rda")
+
+# COWcivilwarongoing == 1
+df.cow2 <- df5 %>% filter(COWcivilwarongoing == 1)
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop",
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "uamkstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.cow2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-cow.rda")
+
+# Ethnic conflict == 1
+df.eth2 <- df5 %>% filter(ethnowarongoing == 1)
+doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
+                        "milper", "milex", "pop", "totalbeaths",
+                        "guerrilladummy", "regtrans", "riotdummy",
+                        "territoryaims", "militias", "physint",
+                        "percentpopurban", "coupdummy", "postcoldwar",
+                        "lmtnest", "realgdp", "discrim", "exclpop", 
+                        "discpop", "elf",  "polrqnew", "egippolrqnew",
+                        "poltrqnew", "egiptpolrqnew", "polity2sq")
+
+m1 <- eba(y = "uamkstart", free = free.variables,
+          doubtful = doubtful.variables, k = 0:4,
+          data = df.eth2, vif = 7, draws = 50000,
+          level = 0.9, se.fun = se.clustered.robust)
+save(m1, file = "~/Documents/mk/uamk-eth.rda")
+
+######################
+### Random forests ###
+######################
+
+# Load required package
+library(h2o)
+h2o.init(nthreads = -1, max_mem_size = "6G") # change min RAM size if necessary
+
+df2a <- as.h2o(df2)
+
+df2a$MKstart <- as.factor(df2a$MKstart)  #encode the binary response as a factor
+h2o.levels(df2a$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2a, 
+                         ratios = 0.75, # train, validation
+                         seed = 1234)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df2), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+##########################
+### Running the models ###
+##########################
+
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "grid01",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "grid01",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+h2o.varimp(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+# Second model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf01b",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 4363))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf01b",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+
+# Third model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf01c",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 7015))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf01c",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+##########################
+### Ongoing civil wars ###
+##########################
+
+# UCDP == 1
+df.ucdpa <- as.h2o(df.ucdp)
+
+df.ucdpa$MKstart <- as.factor(df.ucdpa$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df.ucdpa$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.ucdpa, 
+                         ratios = 0.75, 
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid,  grid_id = "grid02",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "grid02",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+h2o.varimp_plot(rf2)
+
+# COW == 1
+df.cowa <- as.h2o(df.cow)
+
+df.cowa$MKstart <- as.factor(df.cowa$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df.cowa$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.cowa, 
+                         ratios = 0.75,  
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
+                               "mksyr2", "mksyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf03",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf03",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+# Ethnic conflict == 1
+df.etha <- as.h2o(df.eth)
+
+df.etha$MKstart <- as.factor(df.etha$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df.etha$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.etha, 
+                         ratios = 0.75,  
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.eth), c(y, "ccode", "year", "rgdppc",
+                               "mksyr2", "mksyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf04",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf04",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+#######################
+### Cold War Period ###
+#######################
+df2.coldwar2 <- as.h2o(df2.coldwar)
+
+df2.coldwar2$MKstart <- as.factor(df2.coldwar2$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df2.coldwar2$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2.coldwar2, 
+                         ratios = 0.75,  
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.eth), c(y, "ccode", "year", "rgdppc",
+                              "mksyr2", "mksyr3", "sf", "country",
+                              "elf2", "polity2sq"))
+
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf04cw",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf04cw",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+
+############################
+### Post Cold War Period ###
+############################
+df2.postcoldwar2 <- as.h2o(df2.postcoldwar)
+
+df2.postcoldwar2$MKstart <- as.factor(df2.postcoldwar2$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df2.postcoldwar2$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2.postcoldwar2, 
+                         ratios = 0.75,  
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.eth), c(y, "ccode", "year", "rgdppc",
+                              "mksyr2", "mksyr3", "sf", "country",
+                              "elf2", "polity2sq"))
+
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf04pcw",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf04pcw",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+
+########################
+### Only Peace Years ###
+########################
+df2.nowar <- as.h2o(df.nowar)
+
+df2.nowar$MKstart <- as.factor(df2.nowar$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df2.nowar$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2.nowar, 
+                         ratios = 0.75,  
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.eth), c(y, "ccode", "year", "rgdppc",
+                              "mksyr2", "mksyr3", "sf", "country",
+                              "elf2", "polity2sq"))
+
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf04nowar",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf04nowar",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+
+###########################################################
+## Same models with Genocide/Politicide variable (Harf) ###
+###########################################################
+df5a <- as.h2o(df5)
+
+df5a$uamkstart <- as.factor(df5a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df5a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df5a, 
+                         ratios = 0.75, 
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df5), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Main model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf05",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+# Saving the most accurate model
+rf.grid <- h2o.getGrid(grid_id = "gridrf05",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+# UCDP == 1
+df.ucdp2a <- as.h2o(df.ucdp2)
+
+df.ucdp2a$uamkstart <- as.factor(df.ucdp2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.ucdp2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.ucdp2a, 
+                         ratios = 0.75, 
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.ucdp2), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf06",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf06",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+# COW == 1
+df.cow2a <- as.h2o(df.cow2)
+
+df.cow2a$uamkstart <- as.factor(df.cow2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.cow2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.cow2a, 
+                         ratios = 0.75,  
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.cow2), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf07",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf07",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+# Ethnic conflict == 1
+df.eth2a <- as.h2o(df.eth2)
+
+df.eth2a$uamkstart <- as.factor(df.eth2a$uamkstart)  #encode the binary repsonse as a factor
+h2o.levels(df.eth2a$uamkstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.eth2a, 
+                         ratios = 0.75, 
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "uamkstart"
+x <- setdiff(names(df.eth2), c(y, "ccode", "year", "rgdppc",
+                           "uamkyr2", "uamkyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Running the model
+rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
+               validation_frame = valid, grid_id = "gridrf08",
+               hyper_params = list(ntrees = c(256, 512, 1024),
+                                   max_depth = c(10, 20, 40),
+                                   mtries = c(5, 6, 7),
+                                   balance_classes = c(TRUE, FALSE),
+                                   sample_rate = c(0.5, 0.632, 0.95),
+                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
+                                   histogram_type = "RoundRobin",
+                                   seed = 1234))
+
+rf.grid <- h2o.getGrid(grid_id = "gridrf08",
+                       sort_by = "auc",
+                       decreasing = TRUE)
+rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
+h2o.saveModel(rf2, path = "/Users/politicaltheory/Documents/GitHub/mass-killings-8k/data/")
+summary(rf2)
+varimp <- as.data.frame(h2o.varimp(rf2))
+h2o.varimp_plot(rf2)
+
+##############
+### Graphs ###
+##############
+
+################
+# EBA Graphs ###
+################
+
+# Main models
 hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
@@
 
-### Ongoing Civil Wars
-
-# UCDPcivilwarongoing == 1
-df.ucdp2 <- df5 %>% filter(UCDPcivilwarongoing == 1)
-doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
-                        "milper", "milex", "pop", "totalbeaths",
-                        "guerrilladummy", "regtrans", "riotdummy",
-                        "territoryaims", "militias", "physint",
-                        "percentpopurban", "coupdummy", "postcoldwar",
-                        "lmtnest", "realgdp", "discrim", "exclpop",
-                        "discpop", "elf",  "polrqnew", "egippolrqnew",
-                        "poltrqnew", "egiptpolrqnew", "polity2sq")
-
-m1 <- eba(y = "uamkstart", free = free.variables,
-          doubtful = doubtful.variables, k = 0:4,
-          data = df.ucdp2, vif = 7, draws = 50000,
-          level = 0.9, se.fun = se.clustered.robust)
-
+# Round
+m1$coefficients$mean$beta2 <- round(as.numeric(m1$coefficients$mean$beta),4)
+m1$coefficients$mean$se2 <- round(as.numeric(m1$coefficients$mean$se),4)
+m1$coefficients$mean
+
+## Models including only mass killings during civil wars
 hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
@@
 
-# COWcivilwarongoing == 1
-df.cow2 <- df5 %>% filter(COWcivilwarongoing == 1)
-doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
-                        "milper", "milex", "pop", "totalbeaths",
-                        "guerrilladummy", "regtrans", "riotdummy",
-                        "territoryaims", "militias", "physint",
-                        "percentpopurban", "coupdummy", "postcoldwar",
-                        "lmtnest", "realgdp", "discrim", "exclpop",
-                        "discpop", "elf",  "polrqnew", "egippolrqnew",
-                        "poltrqnew", "egiptpolrqnew", "polity2sq")
-
-m1 <- eba(y = "uamkstart", free = free.variables,
-          doubtful = doubtful.variables, k = 0:4,
-          data = df.cow2, vif = 7, draws = 50000,
-          level = 0.9, se.fun = se.clustered.robust)
-
-hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
-                       "assdummy", "totaltrade", "tradedependence", "milper",
-                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
-                       "riotdummy", "territoryaims", "militias", "physint",
-                       "percentpopurban", "coupdummy", "postcoldwar",
-                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
-                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
-                       "egiptpolrqnew"),
-     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last genocide",
-              "Assassination", "Total trade", 
-              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
+# Cold War and Post-Cold War Periods
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "UCDPcivilwarongoing",
+                       "UCDPcivilwarstart",
+                       "COWcivilwarongoing", "COWcivilwarstart",
+                       "ethnowarongoing", "ethnowarstart", "assdummy",
+                       "totaltrade", "tradedependence", "cinc",
+                       "totalbeaths", "guerrilladummy", "regtrans",
+                       "riotdummy", "territoryaims", "militias",
+                       "physint", "percentpopurban", "coupdummy",
+                       "lmtnest", "realgdp", "discrim",
+                       "exclpop", "discpop", "elf",  "polrqnew",
+                       "egippolrqnew", "poltrqnew", "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years mass killings",
+              "UCDP ongoing", "UCDP onset", "COW ongoing", "COW onset", 
+              "Ethnic ongoing", "Ethnic onset", "Assassination", "Total trade", 
+              "Trade dependence", "CINC", 
               "Total deaths", "Guerrilla", "Regime transition", "Riots",
+              "Territory Aims", "Militias", "Physical integrity", "% Urban",
+              "Coups", "Mountainous terrain", "Real GDP",
+              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+     density.col = "black", mu.col = "red3")
+
+#### Peacetime
+hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "mksyr", "assdummy",
+                       "totaltrade", "tradedependence", "milper", "milex", "pop",
+                       "totalbeaths", "guerrilladummy",
+                       "riotdummy", "territoryaims", "militias",
+                       "physint", "percentpopurban", "coupdummy", "postcoldwar",
+                       "lmtnest", "realgdp", "discrim",
+                       "exclpop", "discpop", "elf",  "polrqnew",
+                       "egippolrqnew", "poltrqnew", "egiptpolrqnew"),
+     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years mass killings","Assassination", "Total trade", 
+              "Trade dependence", "Military Personnel", "Military Expenditure", "Population",
+              "Total deaths", "Guerrilla", "Previous riots",
               "Territory Aims", "Militias", "Physical integrity", "% Urban",
@@
               "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
-              "Groups/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
+              "Group/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
      density.col = "black", mu.col = "red3")
 
-# ethnic civil war == 1
-df.eth2 <- df5 %>% filter(ethnowarongoing == 1)
-doubtful.variables <- c("assdummy", "totaltrade", "tradedependence",
-                        "milper", "milex", "pop", "totalbeaths",
-                        "guerrilladummy", "regtrans", "riotdummy",
-                        "territoryaims", "militias", "physint",
-                        "percentpopurban", "coupdummy", "postcoldwar",
-                        "lmtnest", "realgdp", "discrim", "exclpop", 
-                        "discpop", "elf",  "polrqnew", "egippolrqnew",
-                        "poltrqnew", "egiptpolrqnew", "polity2sq")
-
-m1 <- eba(y = "uamkstart", free = free.variables,
-          doubtful = doubtful.variables, k = 0:4,
-          data = df.eth2, vif = 7, draws = 50000,
-          level = 0.9, se.fun = se.clustered.robust)
-
-hist(m1, variables = c("logrgdppc", "polity2", "polity2sq", "uamkyr",
-                       "assdummy", "totaltrade", "tradedependence", "milper",
-                       "milex","pop", "totalbeaths", "guerrilladummy", "regtrans",
-                       "riotdummy", "territoryaims", "militias", "physint",
-                       "percentpopurban", "coupdummy", "postcoldwar",
-                       "lmtnest", "realgdp", "discrim", "exclpop", "discpop",
-                       "elf", "polrqnew", "egippolrqnew", "poltrqnew",
-                       "egiptpolrqnew"),
-     main = c("Log GDP capita", "Polity IV", "Polity IV^2", "Years last genocide",
-              "Assassination", "Total trade", 
-              "Trade dependence", "Military personnel", "Military expenditure", "Population", 
-              "Total deaths", "Guerrilla", "Regime transition", "Riots",
-              "Territory Aims", "Militias", "Physical integrity", "% Urban",
-              "Coups", "Post-Cold War", "Mountainous terrain", "Real GDP",
-              "Discrimination", "Excl pop", "Discrim pop", "ELF", "Groups/Eth relevant", 
-              "Groups/Tot pop", "Inc groups/Eth relevant", "Inc groups/Tot pop"),
-     density.col = "black", mu.col = "red3")
-
-########################################################
-### Genocide/Politicide -- Distributed Random Forest ###
-########################################################
+######################
+### Random forests ###
+######################
+
+# Main model
+library(h2o)
+h2o.init(nthreads = -1, max_mem_size = "6G") 
+a <- h2o.loadModel("grid01_model_197")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+df2a <- as.h2o(df2)
+
+df2a$MKstart <- as.factor(df2a$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df2a$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2a, 
+                         ratios = 0.75,  # 70%, 15%, 15%
+                         seed = 1234)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df2), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Variable Importance
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Real GDP",
+                      "Total trade",
+                      "Polity IV",
+                      "Population", 
+                      "Military expenditure",
+                      "Military personnel",
+                      "Trade dependence", 
+                      "% Urban pop.",
+                      "Log GDP per capita",
+                      "Years mass killing"),
+        main = "")
+
+# Partial dependence plots
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p1 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() +
+        xlab("Years mass killing") +  ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p2 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p3 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + xlab("% Urban pop.") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+        theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p5 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+        xlab("Military personnel") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p6 <- qplot(milex$milex, milex$mean_response) + geom_line() + theme_classic() +
+  xlab("Military expenditure") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p7 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() +
+        xlab("Population") + ylab("Mean response")
+
+polity2 <- h2o.partialPlot(object = a, data = train, cols = c("polity2"), plot_stddev = F)
+p8 <- qplot(polity2$polity2, polity2$mean_response) + geom_line() + theme_classic() +
+  xlab("Polity IV") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p9 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+  xlab("Total trade") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p10 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() + theme_classic() +
+  xlab("Real GDP") + ylab("Mean response")
+
+# Multiplot function: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+        library(grid)
+
+        # Make a list from the ... arguments and plotlist
+        plots <- c(list(...), plotlist)
+
+        numPlots = length(plots)
+
+        # If layout is NULL, then use 'cols' to determine layout
+        if (is.null(layout)) {
+                # Make the panel
+                # ncol: Number of columns of plots
+                # nrow: Number of rows needed, calculated from # of cols
+                layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                                 ncol = cols, nrow = ceiling(numPlots/cols))
+        }
+
+        if (numPlots==1) {
+                print(plots[[1]])
+
+        } else {
+                # Set up the page
+                grid.newpage()
+                pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+
+                # Make each plot, in the correct location
+                for (i in 1:numPlots) {
+                        # Get the i,j matrix positions of the regions that contain this subplot
+                        matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+
+                        print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                                        layout.pos.col = matchidx$col))
+                }
+        }
+}
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.1x5.14 in
+
+#######################################
+### Mass killings during civil wars ###
+#######################################
+
+# UCDP == 1
+a <- h2o.loadModel("grid02_model_349")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Real GDP",
+                      "Military personnel",
+                      "Population",
+                      "Military expenditure",
+                      "Total trade",
+                      "Log GDP per capita",
+                      "CINC",
+                      "Years mass killing",
+                      "Trade dependence",
+                      "% Urban pop."),
+        main = "")
+
+df.ucdpa <- as.h2o(df.ucdp)
+
+df.ucdpa$MKstart <- as.factor(df.ucdpa$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df.ucdpa$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.ucdpa, 
+                         ratios = 0.75,  # 70%, 15%, 15%
+                         seed = 1234)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
+                               "mksyr2", "mksyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p1 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p2 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p3 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + 
+  xlab("Years since mass killing") +  ylab("Mean response")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"), plot_stddev = F)
+p4 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
+  xlab("CINC") +  ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p5 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p6 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+        xlab("Total trade") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p7 <- qplot(milex$milex, milex$mean_response) + geom_line() + theme_classic() +
+  xlab("Military expenditure") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p8 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() +
+  xlab("Population") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p9 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+  xlab("Military personnel") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p10 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() + theme_classic() +
+  xlab("Real GDP") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+# COW == 1
+a <- h2o.loadModel("gridrf03_model_41")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Polarisation",
+                      "Military expenditure",
+                      "Military personnel",
+                      "Excluded population", 
+                      "% Urban",
+                      "Previous riots",
+                      "Total battle deaths", 
+                      "Log GDP per capita",
+                      "Years mass killing",
+                      "Physical integrity"),
+        main = "")
+
+df.cowa <- as.h2o(df.cow)
+
+df.cowa$MKstart <- as.factor(df.cowa$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df.cowa$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.cowa, 
+                         ratios = 0.75,  
+                         seed = 1234)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
+                               "mksyr2", "mksyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+physint <- h2o.partialPlot(object = a, data = train, cols = c("physint"), plot_stddev = F)
+p1 <- qplot(physint$physint, physint$mean_response) + geom_line() + theme_classic() + 
+        xlab("Physical integrity") +  ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p2 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + 
+  xlab("Years mass killing") +  ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p3 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"), plot_stddev = F)
+p4 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() + theme_classic() +
+  xlab("Total battle deaths") + ylab("Mean response")
+
+riotdummy <- h2o.partialPlot(object = a, data = train, cols = c("riotdummy"), plot_stddev = F)
+p5 <- qplot(riotdummy$riotdummy, riotdummy$mean_response) + geom_line() + theme_classic() +
+        xlab("Previous riots") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p6 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+exclpop <- h2o.partialPlot(object = a, data = train, cols = c("exclpop"), plot_stddev = F)
+p7 <- qplot(exclpop$exclpop, exclpop$mean_response) + geom_line() +
+  theme_classic() + xlab("Excluded population") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p8 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+  xlab("Military personnel") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p9 <- qplot(milex$milex, milex$mean_response) + geom_line() + theme_classic() +
+  xlab("Military expenditure") + ylab("Mean response")
+
+egiptpolrqnew <- h2o.partialPlot(object = a, data = train, cols = c("egiptpolrqnew"), plot_stddev = F)
+p10 <- qplot(egiptpolrqnew$egiptpolrqnew, egiptpolrqnew$mean_response) + geom_line() + theme_classic() +
+  xlab("Polarisation") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+# Ethnic conflict == 1
+a <- h2o.loadModel("gridrf04_model_52")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Log GDP per capita",
+                      "Democracy",
+                      "Years mass killing",
+                      "Polarisation",
+                      "% Urban",
+                      "Territorial aims", 
+                      "Trade dependence",
+                      "Excluded population",
+                      "Military personnel",
+                      "Polity IV"),
+        main = "")
+
+df.etha <- as.h2o(df.eth)
+
+df.etha$MKstart <- as.factor(df.etha$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df.etha$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df.etha, 
+                         ratios = 0.75, 
+                         seed = 42)
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df.ucdp), c(y, "ccode", "year", "rgdppc",
+                               "mksyr2", "mksyr3", "sf", "country",
+                               "elf2", "polity2sq"))
+
+polity2 <- h2o.partialPlot(object = a, data = train, cols = c("polity2"), plot_stddev = F)
+p1 <- qplot(polity2$polity2, polity2$mean_response) + geom_line() + theme_classic() + 
+        xlab("Polity IV") +  ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p2 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+  xlab("Military personnel") + ylab("Mean response")
+
+exclpop <- h2o.partialPlot(object = a, data = train, cols = c("exclpop"), plot_stddev = F)
+p3 <- qplot(exclpop$exclpop, exclpop$mean_response) + geom_line() +
+  theme_classic() + xlab("Excluded population") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+territoryaims <- h2o.partialPlot(object = a, data = train, cols = c("territoryaims"), plot_stddev = F)
+p5 <- qplot(territoryaims$territoryaims, territoryaims$mean_response) + geom_line() +
+  theme_classic() + xlab("Territory aims") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p6 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+egiptpolrqnew <- h2o.partialPlot(object = a, data = train, cols = c("egiptpolrqnew"), plot_stddev = F)
+p7 <- qplot(egiptpolrqnew$egiptpolrqnew, egiptpolrqnew$mean_response) + geom_line() + theme_classic() +
+  xlab("Polarisation") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p8 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() + 
+  xlab("Years mass killing") +  ylab("Mean response")
+
+demdummy <- h2o.partialPlot(object = a, data = train, cols = c("demdummy"), plot_stddev = F)
+p9 <- qplot(demdummy$demdummy, demdummy$mean_response) + geom_line() + theme_classic() + 
+  xlab("Democracy") +  ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p10 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+#######################
+### Different seeds ###
+#######################
+
+## Seed 4363
+a <- h2o.loadModel("gridrf01b_model_73")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+df2a <- as.h2o(df2)
+
+df2a$MKstart <- as.factor(df2a$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df2a$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2a, 
+                         ratios = 0.75,  # 70%, 15%, 15%
+                         seed = 1234)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df2), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Variable Importance
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Total trade",
+                      "Real GDP",
+                      "CINC",
+                      "Population", 
+                      "Military personnel",
+                      "Military expenditure",
+                      "Years mass killing",
+                      "% Urban pop.",
+                      "Trade dependence", 
+                      "Log GDP per capita"),
+        main = "")
+
+# Partial dependence plots
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p1 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") +  ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p2 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p3 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban pop.") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p4 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() +
+  xlab("Years mass killing") +  ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p5 <- qplot(milex$milex, milex$mean_response) + geom_line() + theme_classic() +
+  xlab("Military expenditure") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p6 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+  xlab("Military personnel") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p7 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() +
+  xlab("Population") + ylab("Mean response")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"), plot_stddev = F)
+p8 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() +
+  xlab("CINC") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p9 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() + theme_classic() +
+  xlab("Real GDP") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p10 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+  xlab("Total trade") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+## Seed 7015
+
+a <- h2o.loadModel("gridrf01c_model_409")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+df2a <- as.h2o(df2)
+
+df2a$MKstart <- as.factor(df2a$MKstart)  #encode the binary repsonse as a factor
+h2o.levels(df2a$MKstart)
+
+# Partition the data into training, validation and test sets
+splits <- h2o.splitFrame(data = df2a, 
+                         ratios = 0.75,  # 70%, 15%, 15%
+                         seed = 1234)  # reproducibility
+
+train <- h2o.assign(splits[[1]], "train.hex")   
+valid <- h2o.assign(splits[[2]], "valid.hex")
+
+y <- "MKstart"
+x <- setdiff(names(df2), c(y, "ccode", "year", "rgdppc",
+                           "mksyr2", "mksyr3", "sf", "country",
+                           "elf2", "polity2sq"))
+
+# Variable Importance
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Total trade",
+                      "Real GDP",
+                      "Military personnel",
+                      "CINC",
+                      "Population", 
+                      "Military expenditure",
+                      "% Urban pop.",
+                      "Years mass killing",
+                      "Trade dependence", 
+                      "Log GDP per capita"),
+        main = "")
+
+# Partial dependence plots
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p1 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") +  ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p2 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p3 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() +
+  xlab("Years mass killing") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p4 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban pop.") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p5 <- qplot(milex$milex, milex$mean_response) + geom_line() + theme_classic() +
+  xlab("Military expenditure") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p6 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() +
+  xlab("Population") + ylab("Mean response")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"), plot_stddev = F)
+p7 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() +
+  xlab("CINC") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p8 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+  xlab("Military personnel") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p9 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() + theme_classic() +
+  xlab("Real GDP") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p10 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+  xlab("Total trade") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+#### Cold War Period
+# Variable Importance
+a <- h2o.loadModel("data/gridrf04cw_model_100")
+va <- h2o.varimp(a)
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Population", 
+                      "Trade dependence",
+                      "Polarisation",
+                      "Previous riots",
+                      "Years mass killing",
+                      "Ethnic frac.", 
+                      "Excluded pop.",
+                      "Mountainous terrain",
+                      "Log GDP per capita",
+                      "Polity IV"),
+        main = "")
+
+# Partial dependence plots
+
+polity2 <- h2o.partialPlot(object = a, data = train, cols = c("polity2"), plot_stddev = F)
+p1 <- qplot(polity2$polity2, polity2$mean_response) + geom_line() + theme_classic() +
+  xlab("Polity IV") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p6 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() +
+  xlab("Years mass killing") +  ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p2 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") +  ylab("Mean response")
+
+lmtnest <- h2o.partialPlot(object = a, data = train, cols = c("lmtnest"), plot_stddev = F)
+p3 <- qplot(lmtnest$lmtnest, lmtnest$mean_response) + geom_line() +
+  theme_classic() + xlab("Mountainous terrain") + ylab("Mean response")
+
+exclpop <- h2o.partialPlot(object = a, data = train, cols = c("exclpop"), plot_stddev = F)
+p4 <- qplot(exclpop$exclpop, exclpop$mean_response) + geom_line() +
+  theme_classic() + xlab("Excluded pop.") + ylab("Mean response")
+
+elf <- h2o.partialPlot(object = a, data = train, cols = c("elf"), plot_stddev = F)
+p5 <- qplot(elf$elf, elf$mean_response) + geom_line() + theme_classic() +
+  xlab("Military personnel") + ylab("Mean response")
+
+riotdummy <- h2o.partialPlot(object = a, data = train, cols = c("riotdummy"), plot_stddev = F)
+p7 <- qplot(riotdummy$riotdummy, riotdummy$mean_response) + geom_line() + theme_classic() +
+  xlab("Previous riots") + ylab("Mean response")
+
+polrqnew <- h2o.partialPlot(object = a, data = train, cols = c("polrqnew"), plot_stddev = F)
+p8 <- qplot(polrqnew$polrqnew, polrqnew$mean_response) + geom_line() + theme_classic() +
+  xlab("Polarisation") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p10 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() +
+  xlab("Population") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p9 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() + theme_classic() +
+  xlab("Trade dependence") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+#### Post-Cold War Period
+# Variable Importance
+a <- h2o.loadModel("data/gridrf04pcw_model_459")
+va <- h2o.varimp(a)
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Trade dependence",
+                      "Military expenditures",
+                      "Polity IV",
+                      "Military personnel", 
+                      "Mountainous terrain",
+                      "Polarisation",
+                      "Log GDP per capita",
+                      "Years mass killing",
+                      "Total battle deaths",
+                      "% Urban"),
+        main = "")
+
+# Partial dependence plots
+
+polity2 <- h2o.partialPlot(object = a, data = train, cols = c("polity2"), plot_stddev = F)
+p8 <- qplot(polity2$polity2, polity2$mean_response) + geom_line() + theme_classic() +
+  xlab("Polity IV") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p3 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() +
+  xlab("Years mass killing") +  ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p4 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p1 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"), plot_stddev = F)
+p2 <- qplot(totalbeaths$totalbeaths,totalbeaths$mean_response) + geom_line() +
+  theme_classic() + xlab("Total battle deaths") + ylab("Mean response")
+
+egippolrqnew <- h2o.partialPlot(object = a, data = train, cols = c("egippolrqnew"), plot_stddev = F)
+p5 <- qplot(egippolrqnew$egippolrqnew, egippolrqnew$mean_response) + geom_line() + theme_classic() +
+  xlab("Polarisation") + ylab("Mean response")
+
+lmtnest <- h2o.partialPlot(object = a, data = train, cols = c("lmtnest"), plot_stddev = F)
+p6 <- qplot(lmtnest$lmtnest, lmtnest$mean_response) + geom_line() +
+  theme_classic() + xlab("Mountainous terrain") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p7 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+  theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p9 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+  theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p10 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() + theme_classic() +
+  xlab("Trade dependence") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+### Mass Killings during Peacetime
+
+# Variable Importance
+a <- h2o.loadModel("data/gridrf04nowar_model_391")
+va <- h2o.varimp(a)
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Total trade",
+                      "Military personnel",
+                      "CINC",
+                      "Military expenditure",
+                      "% Urban pop.",
+                      "Real GDP",
+                      "Trade dependence", 
+                      "Population", 
+                      "Years mass killing",
+                      "Log GDP per capita"),
+        main = "")
+
+# Partial dependence plots
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p1 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") +  ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p4 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p6 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban pop.") + ylab("Mean response")
+
+mksyr <- h2o.partialPlot(object = a, data = train, cols = c("mksyr"), plot_stddev = F)
+p2 <- qplot(mksyr$mksyr, mksyr$mean_response) + geom_line() + theme_classic() +
+  xlab("Years mass killing") +  ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p7 <- qplot(milex$milex, milex$mean_response) + geom_line() + theme_classic() +
+  xlab("Military expenditure") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p9 <- qplot(milper$milper, milper$mean_response) + geom_line() + theme_classic() +
+  xlab("Military personnel") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p3 <- qplot(pop$pop, pop$mean_response) + geom_line() + theme_classic() +
+  xlab("Population") + ylab("Mean response")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"), plot_stddev = F)
+p8 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() +
+  xlab("CINC") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p5 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() + theme_classic() +
+  xlab("Real GDP") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p10 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+  xlab("Total trade") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+####################################################
+### Same models with Genocide/Politicide (Harff) ###
+####################################################
+
+# Main model
+a <- h2o.loadModel("gridrf05_model_79")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Population",
+                      "Polity IV",
+                      "Real GDP", 
+                      "Trade dependence", 
+                      "Log GDP per capita",
+                      "Total trade",
+                      "Military expenditure", 
+                      "Military personnel",
+                      "% Urban",
+                      "CINC"),
+        main = "")
 
 df5a <- as.h2o(df5)
-
 df5a$uamkstart <- as.factor(df5a$uamkstart)  #encode the binary repsonse as a factor
@@
 splits <- h2o.splitFrame(data = df5a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
+                         ratios = 0.75, 
+                         seed = 1234)
 
@@
 valid <- h2o.assign(splits[[2]], "valid.hex")
-test <- h2o.assign(splits[[3]], "test.hex")
 
@@
 
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, 
-               grid_id = "gridrf05",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 500, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-# Saving the most accurate model
-rf.grid <- h2o.getGrid(grid_id = "gridrf05",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## UCDP == 1
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"), plot_stddev = F)
+p1 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() + theme_classic() + 
+        xlab("CINC") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p2 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p3 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+        theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p4 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+  theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p5 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() + theme_classic() +
+        xlab("Total trade") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p6 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") +  ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p7 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p8 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() +
+        theme_classic() + xlab("Real GDP") + ylab("Mean response")
+
+polity2 <- h2o.partialPlot(object = a, data = train, cols = c("polity2"), plot_stddev = F)
+p9 <- qplot(polity2$polity2, polity2$mean_response) + geom_line() +
+  theme_classic() + xlab("Polity IV") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p10 <- qplot(pop$pop, pop$mean_response) + geom_line() +
+  theme_classic() + xlab("Population") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+# UCDP == 1
+df.ucdp2 <- df5 %>% filter(UCDPcivilwarongoing == 1)
 df.ucdp2a <- as.h2o(df.ucdp2)
@@
 splits <- h2o.splitFrame(data = df.ucdp2a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
+                         ratios = .75,  
+                         seed = 1234)
 
@@
 
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf06",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 100, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf06",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## COW == 1
+a <- h2o.loadModel("gridrf06_model_275")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Military expenditure",
+                      "Population",
+                      "Regime change",
+                      "Physical integrity",
+                      "Total battle deaths",
+                      "Trade dependence",
+                      "% Urban", 
+                      "Years since genocide", 
+                      "Log GDP per capita",
+                      "Military personnel"),
+        main = "")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p1 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+        theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p2 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+        xlab("Log GDP per capita") + ylab("Mean response")
+
+uamkyr <- h2o.partialPlot(object = a, data = train, cols = c("uamkyr"), plot_stddev = F)
+p3 <- qplot(uamkyr$uamkyr, uamkyr$mean_response) + geom_line() + theme_classic() + 
+  xlab("Years since genocide") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p4 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+        theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p5 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+        theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"), plot_stddev = F)
+p6 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() +
+        theme_classic() + xlab("Total battle deaths") + ylab("Mean response")
+
+physint <- h2o.partialPlot(object = a, data = train, cols = c("physint"), plot_stddev = F)
+p7 <- qplot(physint$physint, physint$mean_response) + geom_line() +
+  theme_classic() + xlab("Physical integrity") + ylab("Mean response")
+
+change <- h2o.partialPlot(object = a, data = train, cols = c("change"), plot_stddev = F)
+p8 <- qplot(change$change, change$mean_response) + geom_line() +
+  theme_classic() + xlab("Regime change") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p9 <- qplot(pop$pop, pop$mean_response) + geom_line() +
+  theme_classic() + xlab("Population") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p10 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+  theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+# COW == 1
+df.cow2 <- df5 %>% filter(COWcivilwarongoing == 1)
 df.cow2a <- as.h2o(df.cow2)
-
 df.cow2a$uamkstart <- as.factor(df.cow2a$uamkstart)  #encode the binary repsonse as a factor
@@
 splits <- h2o.splitFrame(data = df.cow2a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
+                         ratios = .75,  # 70%, 15%, 15%
+                         seed = 1234)  # reproducibility
 
@@
 
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf07",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 100, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf07",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
-## ETHONSET == 1
+a <- h2o.loadModel("gridrf07_model_413")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Trade dependence",
+                      "Real GDP",
+                      "Population",
+                      "CINC",
+                      "Total trade",
+                      "Total battle deaths",
+                      "Log GDP per capita",
+                      "% Urban", 
+                      "Years since genocide", 
+                      "Military personnel"),
+        main = "")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p1 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+  theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+uamkyr <- h2o.partialPlot(object = a, data = train, cols = c("uamkyr"), plot_stddev = F)
+p2 <- qplot(uamkyr$uamkyr, uamkyr$mean_response) + geom_line() + theme_classic() + 
+  xlab("Years since genocide") +  ylab("Mean response")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p3 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p4 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") + ylab("Mean response")
+
+totalbeaths <- h2o.partialPlot(object = a, data = train, cols = c("totalbeaths"), plot_stddev = F)
+p5 <- qplot(totalbeaths$totalbeaths, totalbeaths$mean_response) + geom_line() +
+  theme_classic() + xlab("Total battle deaths") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p6 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"), plot_stddev = F)
+p7 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() +
+  theme_classic() + xlab("CINC") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p8 <- qplot(pop$pop, pop$mean_response) + geom_line() +
+  theme_classic() + xlab("Population") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p9 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() +
+  theme_classic() + xlab("Real GDP") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p10 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
+
+# Ethnic conflict == 1
+df.eth2 <- df5 %>% filter(ethnowarongoing == 1)
+
 df.eth2a <- as.h2o(df.eth2)
@@
 splits <- h2o.splitFrame(data = df.eth2a, 
-                         ratios = c(0.7, 0.15),  # 70%, 15%, 15%
-                         seed = 42)  # reproducibility
+                         ratios = .75,  # 70%, 15%, 15%
+                         seed = 1234)  # reproducibility
 
@@
 
-# Running the model
-rf <- h2o.grid("randomForest", x = x, y = y, training_frame = train, 
-               validation_frame = valid, nfolds = 5, grid_id = "gridrf08",
-               fold_assignment = "Stratified",
-               hyper_params = list(ntrees = c(256, 512, 1024),
-                                   max_depth = c(10, 20, 40),
-                                   mtries = c(5, 6, 7),
-                                   balance_classes = c(TRUE, FALSE),
-                                   sample_rate = c(0.5, 0.632, 0.95),
-                                   col_sample_rate_per_tree = c(0.5, 0.9, 1.0),
-                                   histogram_type = c("UniformAdaptive",
-                                                      "Random",
-                                                      "QuantilesGlobal",
-                                                      "RoundRobin")),
-               search_criteria = list(strategy = "RandomDiscrete", 
-                                      max_models = 100, 
-                                      stopping_metric = "auc", 
-                                      stopping_tolerance = 0.01, 
-                                      stopping_rounds = 5, 
-                                      seed = 26227709))
-
-rf.grid <- h2o.getGrid(grid_id = "gridrf08",
-                       sort_by = "auc",
-                       decreasing = TRUE)
-rf2 <- h2o.getModel(rf.grid@model_ids[[1]])
-h2o.saveModel(rf2, path = "/root/Documents/mk/")
-summary(rf2)
-varimp <- as.data.frame(h2o.varimp(rf2))
-h2o.varimp_plot(rf2)
-h2o.performance(rf2, newdata = test)
-
+a <- h2o.loadModel("gridrf08_model_173")
+print(va <- a %>% h2o.varimp() %>% as.data.frame() %>% head(., 10))
+
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,7.5,1,1))
+barplot(va$scaled_importance[10:1],
+        horiz = TRUE, las = 1, cex.names=0.9,
+        names.arg = c("Real GDP",
+                      "Trade dependence",
+                      "Physical Integrity",
+                      "Log GDP per capita",
+                      "Population",
+                      "Total trade",
+                      "Military personnel", 
+                      "CINC",
+                      "Military expenditure",
+                      "% Urban"),
+        main = "")
+
+percentpopurban <- h2o.partialPlot(object = a, data = train, cols = c("percentpopurban"), plot_stddev = F)
+p1 <- qplot(percentpopurban$percentpopurban, percentpopurban$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+milex <- h2o.partialPlot(object = a, data = train, cols = c("milex"), plot_stddev = F)
+p2 <- qplot(milex$milex, milex$mean_response) + geom_line() +
+  theme_classic() + xlab("Military expenditure") + ylab("Mean response")
+
+cinc <- h2o.partialPlot(object = a, data = train, cols = c("cinc"), plot_stddev = F)
+p3 <- qplot(cinc$cinc, cinc$mean_response) + geom_line() +
+  theme_classic() + xlab("CINC") + ylab("Mean response")
+
+milper <- h2o.partialPlot(object = a, data = train, cols = c("milper"), plot_stddev = F)
+p4 <- qplot(milper$milper, milper$mean_response) + geom_line() +
+  theme_classic() + xlab("Military personnel") + ylab("Mean response")
+
+totaltrade <- h2o.partialPlot(object = a, data = train, cols = c("totaltrade"), plot_stddev = F)
+p5 <- qplot(totaltrade$totaltrade, totaltrade$mean_response) + geom_line() +
+  theme_classic() + xlab("% Urban") + ylab("Mean response")
+
+pop <- h2o.partialPlot(object = a, data = train, cols = c("pop"), plot_stddev = F)
+p6 <- qplot(pop$pop, pop$mean_response) + geom_line() +
+  theme_classic() + xlab("Population") + ylab("Mean response")
+
+logrgdppc <- h2o.partialPlot(object = a, data = train, cols = c("logrgdppc"), plot_stddev = F)
+p7 <- qplot(logrgdppc$logrgdppc, logrgdppc$mean_response) + geom_line() + theme_classic() +
+  xlab("Log GDP per capita") + ylab("Mean response")
+
+physint <- h2o.partialPlot(object = a, data = train, cols = c("physint"), plot_stddev = F)
+p5 <- qplot(physint$physint, physint$mean_response) + geom_line() +
+  theme_classic() + xlab("Physical integrity") + ylab("Mean response")
+
+tradedependence <- h2o.partialPlot(object = a, data = train, cols = c("tradedependence"), plot_stddev = F)
+p9 <- qplot(tradedependence$tradedependence, tradedependence$mean_response) + geom_line() +
+  theme_classic() + xlab("Trade dependence") + ylab("Mean response")
+
+realgdp <- h2o.partialPlot(object = a, data = train, cols = c("realgdp"), plot_stddev = F)
+p10 <- qplot(realgdp$realgdp, realgdp$mean_response) + geom_line() +
+  theme_classic() + xlab("Real GDP") + ylab("Mean response")
+
+multiplot(p1,p5,p8,p2,p6,p9,p3,p7,p10,p4, cols = 4)  # 11.09x5.14 in
 \end{verbatim}

### 2018-12-04 dissertation.tex
maybe final version

@@
 % DATE OF SUBMISSION
-May 2018
+December 2018
 \end{center}	
@@
 In the third essay, I employ extreme bounds analysis and distributed random forests to identify the key determinants of state-sponsored violence.
-Although the literature on mass killings has grown significantly over the last decades, it remains unclear whether estimates are robust to different model specifications or which variables accurately predict the onset of genocides and politicides.
-To address these issues, I test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
-The results show that GDP per capita, the post-Cold War period, and stable political regimes are negatively associated with mass killings.
-In contrast, ethnic diversity, civil wars, and previous political turmoils increase the risk of state-led violence.
-Mass killings that occur during civil wars are correlated with a different set of covariates, and I find that territorial conflicts and the use of militias are negatively associated with the outcome variable.
-The random forests estimations indicate that years since the last episode of mass violence, GDP per capita, urban population, ethnic polarisation, the number of military personnel, and democracy make the greatest contribution to the models' out-of-sample predictive power.
+Although scholars have suggested a number of potential correlates of mass killings, it remains unclear whether the estimates are robust to different model specifications, or which variables accurately predict the onset of large-scale violence.
+I employ extreme bounds analysis and random forests to test the sensitivity of 40 variables on a sample of 177 countries from 1945 to 2013.
+The results help clear the brush around mass killings, as few variables in this literature are robust determinants of atrocity.
+However, support for an opportunity logic persists as greater constraints on a government limit its ability to employ barbarous tactics.
+It appears that the Conflict Trap applies to government atrocity.
+Atrocity breeds atrocity, while wealthy stable democracies tend to avoid episodes of mass killing.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
